### 2014-02-10 paper/bare_conf.tex
Add paper template

%% bare_conf.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%
%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[conference]{IEEEtran}
% Add the compsoc option for Computer Society conferences.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

% Some very useful LaTeX packages include:
% (uncomment the ones you want to load)

% *** MISC UTILITY PACKAGES ***
%
%\usepackage{ifpdf}
% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
% compilation based on whether the output is pdf or dvi.
% usage:
% \ifpdf
%   % pdf code
% \else
%   % dvi code
% \fi
% The latest version of ifpdf.sty can be obtained from:
% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
% Also, note that IEEEtran.cls V1.7 and later provides a builtin
% \ifCLASSINFOpdf conditional that works the same way.
% When switching from latex to pdflatex and vice-versa, the compiler may
% have to be run twice to clear warning/error messages.

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
% not currently provide for hyperlinked citations.
% The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
% The documentation is contained in the cite.sty file itself.

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

% *** MATH PACKAGES ***
%
%\usepackage[cmex10]{amsmath}
% A popular package from the American Mathematical Society that provides
% many useful and powerful commands for dealing with mathematics. If using
% it, be sure to load this package with the cmex10 option to ensure that
% only type 1 fonts will utilized at all point sizes. Without this option,
% it is possible that some math symbols, particularly those within
% footnotes, will be rendered in bitmap form which will result in a
% document that can not be IEEE Xplore compliant!
%
% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
% thus preventing page breaks from occurring within multiline equations. Use:
%\interdisplaylinepenalty=2500
% after loading amsmath to restore such page breaks as IEEEtran.cls normally
% does. amsmath.sty is already installed on most LaTeX systems. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

% *** ALIGNMENT PACKAGES ***
%
%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty patches and improves
% the standard LaTeX2e array and tabular environments to provide better
% appearance and additional user controls. As the default LaTeX2e table
% generation code is lacking to the point of almost being broken with
% respect to the quality of the end results, all users are strongly
% advised to use an enhanced (at the very least that provided by array.sty)
% set of table tools. array.sty is already installed on most systems. The
% latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

% IEEEtran contains the IEEEeqnarray family of commands that can be used to
% generate multiline equations as well as matrices, tables, etc., of high
% quality.

%\usepackage{eqparbox}
% Also of notable interest is Scott Pakin's eqparbox package for creating
% (automatically sized) equal width boxes - aka "natural width parboxes".
% Available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

% *** FLOAT PACKAGES ***
%
%\usepackage{fixltx2e}
% fixltx2e, the successor to the earlier fix2col.sty, was written by
% Frank Mittelbach and David Carlisle. This package corrects a few problems
% in the LaTeX2e kernel, the most notable of which is that in current
% LaTeX2e releases, the ordering of single and double column floats is not
% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
% single column figure to be placed prior to an earlier double column
% figure. The latest version and documentation can be found at:
% http://www.ctan.org/tex-archive/macros/latex/base/

%\usepackage{stfloats}
% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
% the ability to do double column floats at the bottom of the page as well
% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
% LaTeX2e). It also provides a command:
%\fnbelowfloat
% to enable the placement of footnotes below bottom floats (the standard
% LaTeX2e kernel puts them above bottom floats). This is an invasive package
% which rewrites many portions of the LaTeX2e float routines. It may not work
% with other packages that modify the LaTeX2e float routines. The latest
% version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
% Documentation is contained in the stfloats.sty comments as well as in the
% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
% does not allow \baselineskip to stretch. Authors submitting work to the
% IEEE should note that IEEE rarely uses double column equations and
% that authors should try to avoid such use. Do not be tempted to use the
% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
% not format its papers in such ways.

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
%\usepackage{url}
% url.sty was written by Donald Arseneau. It provides better support for
% handling and breaking URLs. url.sty is already installed on most LaTeX
% systems. The latest version can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
% Read the url.sty source comments for usage information. Basically,
% \url{my_url_here}.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.
% (Unless specifically asked to do so by the journal or conference you plan
% to submit to, of course. )

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{An Optimizing Compiler for Finite Element Assembly}

\author{\IEEEauthorblockN{Fabio Luporini}
\IEEEauthorblockA{Department of Computing\\
Imperial College London\\
London, UK\\
Email: f.luporini12@imperial.ac.uk}
\and
\IEEEauthorblockN{Paul H. J. Kelly}
\IEEEauthorblockA{Department of Computing\\
Imperial College London\\
London, UK\\
Email: p.kelly@imperial.ac.uk}
\and
\IEEEauthorblockN{David Ham}
\IEEEauthorblockA{Department of Computing\\
Imperial College London\\
London, UK\\
Email: d.ham@imperial.ac.uk}}

% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

% for over three affiliations, or if they all won't fit within the width
% of the page, use this alternative format:
% 
%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
%Homer Simpson\IEEEauthorrefmark{2},
%James Kirk\IEEEauthorrefmark{3}, 
%Montgomery Scott\IEEEauthorrefmark{3} and
%Eldon Tyrell\IEEEauthorrefmark{4}}
%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
%Georgia Institute of Technology,
%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
%Email: homer@thesimpsons.com}
%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
%Telephone: (800) 555--1212, Fax: (888) 555--1212}
%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

\begin{abstract}
%\boldmath
The abstract goes here.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

% no keywords

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}
% no \IEEEPARstart
This demo file is intended to serve as a ``starter file''
for IEEE conference papers produced under \LaTeX\ using
IEEEtran.cls version 1.7 and later.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)
I wish you the best of success.

\hfill mds

\hfill January 11, 2007

\subsection{Subsection Heading Here}
Subsection text here.

\subsubsection{Subsubsection Heading Here}
Subsubsection text here.

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.

\section{Conclusion}
The conclusion goes here.

% conference papers do not normally have an appendix

% use section* for acknowledgement
\section*{Acknowledgment}

The authors would like to thank...

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\begin{thebibliography}{1}

\bibitem{IEEEhowto:kopka}
H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.

\end{thebibliography}

% that's all folks
\end{document}

### 2014-02-11 paper/bare_conf.tex
Add skeleton of the paper

@@
-%% bare_conf.tex
-%% V1.3
-%% 2007/01/11
-%% by Michael Shell
-%% See:
-%% http://www.michaelshell.org/
-%% for current contact information.
-%%
-%% This is a skeleton file demonstrating the use of IEEEtran.cls
-%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
-%%
-%% Support sites:
-%% http://www.michaelshell.org/tex/ieeetran/
-%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
-%% and
-%% http://www.ieee.org/
-
-%%*************************************************************************
-%% Legal Notice:
-%% This code is offered as-is without any warranty either expressed or
-%% implied; without even the implied warranty of MERCHANTABILITY or
-%% FITNESS FOR A PARTICULAR PURPOSE! 
-%% User assumes all risk.
-%% In no event shall IEEE or any contributor to this code be liable for
-%% any damages or losses, including, but not limited to, incidental,
-%% consequential, or any other damages, resulting from the use or misuse
-%% of any information contained here.
-%%
-%% All comments are the opinions of their respective authors and are not
-%% necessarily endorsed by the IEEE.
-%%
-%% This work is distributed under the LaTeX Project Public License (LPPL)
-%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
-%% distributed and modified. A copy of the LPPL, version 1.3, is included
-%% in the base LaTeX documentation of all distributions of LaTeX released
-%% 2003/12/01 or later.
-%% Retain all contribution notices and credits.
-%% ** Modified files should be clearly indicated as such, including  **
-%% ** renaming them and changing author support contact information. **
-%%
-%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
-%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
-%%*************************************************************************
-
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. IEEE's font choices can trigger bugs that do  ***
-% *** not appear when using other class files.                            ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
-% Note that the a4paper option is mainly intended so that authors in
-% countries using A4 can easily print to A4 and see how their papers will
-% look in print - the typesetting of the document will not typically be
-% affected with changes in paper size (but the bottom and side margins will).
-% Use the testflow package mentioned above to verify correct handling of
-% both paper sizes by the user's LaTeX system.
-%
-% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
-% should be used if it is desired that the figures are to be displayed in
-% draft mode.
-%
 \documentclass[conference]{IEEEtran}
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{An Optimizing Compiler for Finite Element Assembly}
+\title{A Compiler-Driven Optimization Approach\\to Solving Partial Differential Equations}
 
@@
 \section{Introduction}
-% no \IEEEPARstart
-This demo file is intended to serve as a ``starter file''
-for IEEE conference papers produced under \LaTeX\ using
-IEEEtran.cls version 1.7 and later.
-% You must have at least 2 lines in the paragraph with the drop letter
-% (should never be an issue)
-I wish you the best of success.
+What we do in computational science.
+Applications motivating our study.
+The programming model of lizst and op2: ``simple'' kernels applied to sets of mesh elements.
+The need for optimizing such kernels.
+Contributions of the paper:
+\begin{itemize}
+\item A \textbf{compiler-driven} optimisation approach that can be generalised to a wide variety of kernels employed in many computational science codes.
+The key feature common to these kernels is that they usually fit the L1 cache of commodity processors.
+\item Automation of such code optimizations in a framework for the resoloution of partial differential equations through the Finite Element Method, called Firedrake.
+This allows us to perform an in-depth performance evaluation of the proposed optimization strategy in real-world computations.
+\end{itemize}
 
-\hfill mds
+\section{Background}
+Here we discuss about the computational characteristics of computational science kernels.
+Briefly cite op2 kernels.
+Emphasis on Finite Element Assembly.
+Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.
 
-\hfill January 11, 2007
+\section{A Compiler-driven Approach to Code Optimization}
+All the work done in the polyhedral model should be useful, but our kernels fit L1, so they need special attention for what concerns key aspects like managing registers, exploiting ILP in its various forms, and reducing loop overhead.
+Follows a description of code transformations using our compiler-driven approach.
 
-\subsection{Subsection Heading Here}
-Subsection text here.
+\section{The PyOP2 Compiler}
+Design and structure of the PyOP2 Compiler.
+Show the steps through which the IR is transformed (need to cite Firedrake here).
+Briefly describe a simple cost model that allows us to prune the space of code transformations.
 
-\subsubsection{Subsubsection Heading Here}
-Subsubsection text here.
+\section{Performance Evaluation}
+Performance evaluation.
+We show all of the results got so far and we explain them.
+We demonstrate the claim that a compiler-driven approach is necessary to obtain maximum performance.
+Also, our results prove that current vendor compilers still have notable limitations, despite optimizing loop nests that are 1) affine, 2) have no stencil, and 3) ?
+
+\section{Related Work}
+Show other work of people working on optimizations for computational science kernels.
+Inter-kernel vectorisation paper from Istvan.
+Previous work in FFC.
+Spencer et-al + Shin et.
+al.
+attempt to optimize computations that could benefit from using BLAS, but in practice they don't, due to the very small dgemm operations employed.
+Saday's model-driven SIMD code vectorisation for the tensor contraction engine.
+
+\section{Conclusions}
+
+%\subsubsection{Subsubsection Heading Here}
+%Subsubsection text here.
 
@@
 
-\section{Conclusion}
-The conclusion goes here.
-
 % conference papers do not normally have an appendix

### 2014-02-12 paper/bare_conf.tex
Start writing the introduction

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{A Compiler-Driven Optimization Approach\\to Solving Partial Differential Equations}
+\title{A Model-Driven Optimization Approach\\to Solving Partial Differential Equations}
 
@@
 \section{Introduction}
-What we do in computational science.
+In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, programs are required that model phenomena.
+A feature common to these scientific computations is that they involve the resolution of partial differential equations (PDEs).
+Numerical techniques, like Finite Volume Method (FVM) and Finite Element Method (FEM), are widely employed in real-world applications to approximate solutions of PDEs.
+Unstructured meshes are often used to discretize the PDE domain, since their geometric flexibility allows such PDE solvers to be extrimely effective.
+However, they are also the main responsible for performance degradation due to the introduction of indirect accesses (e.g.
+\texttt{A[B[i]]}) to access mesh components.
+A typical problem, therefore, is that implementing a scientific application requires a great effort, other than domain knowledge, and even more time is needed to optimize their execution time.
+
+An unstructured mesh represents a discretization of the equation domain into two- or three-dimensional cells (or ``elements''), which define the smallest area within which the solution is sought.
+As the number of elements can be of the order of millions, a major issue is the time required to perform the numerical method, which can be of the order of hours or days.
+OP2~\cite{pyop2ics} and Lizst~\cite{lizst} are Domain Specific Languages that target mesh-based computations by allowing programmers to achieve highly-efficient parallel execution.
+The successful porting of Hydra, a CFD industrial application devised by Rolls Royce for turbomachinery design (based on FVM, roughly 50000 lines of code and a mesh size that can be over 100 millions edges), to OP2, demonstrates the effectiveness of this framework for implementing PDEs solvers~\cite{OP2-hydra}.
+Both OP2 and Lizst adopt a kernel-oriented programming model: the computation semantics is expressed through self-contained functions (or ``kernels''), each function applied in parallel to all items in a specific set of mesh components (e.g.
+edges, vertices, elements), with an implicit synchronization between the application of two different functions.
+Together with the indirections problem, having systematic optimizations for the kernels is the major concern in unstructured mesh applications.
+
+In this paper, we describe a model-driven optimization strategy for a class of kernels used in Finite Element computations.
+The working set of these kernels is usually small enough to fit the L1 cache, and definitely fits the L2 cache when high-order methods are employed to improve the accuracy of the solution.
+However, the latter case is not considered in our work.
+This implies that main actors of our study will be register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation.
+
+%the focus of optimization should be on managing registers, exploiting ILP in its various forms, and reducing loop overhead
+
+% systematic optimization strategy -> discuss the importance of domain knowledge
+
+% A python implementation of OP2, namely PyOP2, is used in the Firedrake project as an intermediate representation for FEM computations.
+In Firedrake all computational kernels are generated by a compiler, (the Fenics Form Compiler, FFC), whereas in OP2 applications kernels are provided by users.
+
+%What we do in computational science.
 Applications motivating our study.
@@
 Contributions of the paper:
-\begin{itemize}
-\item A \textbf{compiler-driven} optimisation approach that can be generalised to a wide variety of kernels employed in many computational science codes.
+%\begin{itemize}
+%\item A \textbf{compiler-driven} optimisation approach that can be generalised to a wide variety of kernels employed in many computational science codes.
 The key feature common to these kernels is that they usually fit the L1 cache of commodity processors.
-\item Automation of such code optimizations in a framework for the resoloution of partial differential equations through the Finite Element Method, called Firedrake.
+%\item Automation of such code optimizations in a framework for the resoloution of partial differential equations through the Finite Element Method, called Firedrake.
 This allows us to perform an in-depth performance evaluation of the proposed optimization strategy in real-world computations.
-\end{itemize}
+%\end{itemize}
 
 \section{Background}
-Here we discuss about the computational characteristics of computational science kernels.
+
+%Here we discuss about the computational characteristics of computational science kernels.
 Briefly cite op2 kernels.
@@
 
-\section{A Compiler-driven Approach to Code Optimization}
-All the work done in the polyhedral model should be useful, but our kernels fit L1, so they need special attention for what concerns key aspects like managing registers, exploiting ILP in its various forms, and reducing loop overhead.
-Follows a description of code transformations using our compiler-driven approach.
+\section{A Model-driven Approach to Code Optimization}
+%All the work done in the polyhedral model should be useful, but our kernels fit L1, so they need special attention for what concerns key aspects like managing registers, exploiting ILP in its various forms, and reducing loop overhead.
+%Follows a description of code transformations using our compiler-driven approach.
 
 \section{The PyOP2 Compiler}
-Design and structure of the PyOP2 Compiler.
+%Design and structure of the PyOP2 Compiler.
 Show the steps through which the IR is transformed (need to cite Firedrake here).
@@
 \section{Performance Evaluation}
-Performance evaluation.
+%Performance evaluation.
 We show all of the results got so far and we explain them.
@@
 \section{Related Work}
-Show other work of people working on optimizations for computational science kernels.
+%Show other work of people working on optimizations for computational science kernels.
 Inter-kernel vectorisation paper from Istvan.
@@
 %
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-\begin{thebibliography}{1}
+%% <OR> manually copy in the resultant .bbl file
+%% set second argument of \begin to the number of references
+%% (used to reserve space for the reference number labels box)
+%\begin{thebibliography}{1}
+%
+%\bibitem{IEEEhowto:kopka}
+%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
+%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
+%
+%\end{thebibliography}
+%
 
-\bibitem{IEEEhowto:kopka}
-H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
-  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-
-\end{thebibliography}
+\bibliographystyle{plain}	% (uses file "plain.bst")
+\bibliography{biblio}		% expects file "myrefs.bib"
 

### 2014-02-13 paper/bare_conf.tex
Add compiler algorithms

@@
 %
-%\usepackage{algorithmic}
+\usepackage[longend,linesnumbered]{algorithm2e}
+\usepackage{algorithmic}
+
+\newcommand{\code}[1]{\lstset{basicstyle=\small\tt}\lstinline£#1£\lstset
+{basicstyle=\footnotesize\tt}}
+
 % algorithmic.sty was written by Peter Williams and Rogerio Brito.
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{A Model-Driven Optimization Approach\\to Solving Partial Differential Equations}
+\title{Domain-Driven Code Transformation\\for Solving Partial Differential Equations}
 
@@
 
-\section{A Model-driven Approach to Code Optimization}
+\section{A Domain-Driven Approach to Code Optimization}
 %All the work done in the polyhedral model should be useful, but our kernels fit L1, so they need special attention for what concerns key aspects like managing registers, exploiting ILP in its various forms, and reducing loop overhead.
 %Follows a description of code transformations using our compiler-driven approach.
+
+\clearpage
 
@@
 Briefly describe a simple cost model that allows us to prune the space of code transformations.
+
+The compiler structure
+
+\begin{algorithm}[h]
+\caption{The PyOP2 compiler.}
+\label{algo:PyOP2Compiler}
+  \textbf{The PyOP2 Compiler}\\
+  \KwIn{ast, wrapper, isa}
+  \KwOut{code}
+// Analyze ast and build optimization plan \\
+it\_space = analyze(ast) \\
+\If{\KwSty{not} it\_space}{
+  ast.apply\_inter\_kernel\_vectorization(wrapper)
+  \Return{ast.from\_ast\_to\_c()}
+}
+plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs) \\
+~~\\
+// Optimize ast based on plan \\
+ast.apply\_licm() \\
+\If{plan.sz\_split}{
+  ast.split(plan.sz\_split)
+}
+\If{plan.uaj\_factor}{
+  ast.vr\_tile(plan.uaj\_factor)
+}
+ast.padding() \\
+ast.data\_align() \\
+\Return{ast.from\_ast\_to\_c()}
+\end{algorithm}
+
+The compiler cost model. Used to find out tile size and split size.
+
+\begin{algorithm}[h]
+\caption{Procedure to estimate the most suitable unroll-and-jam factor and/or split size.}
+\label{algo:applyCostModel}
+  \textbf{Algorithm: ApplyCostModel}\\
+  \KwIn{n\_inner\_arrays, n\_regs}
+  \KwOut{uaj\_factor, sz\_split}
+
+%// roundup(A, B) is a function that rounds A up to the nearest multiple of B \\
+
+sz\_split = 0 \\
+// Compute spltting factor \\
+\While{n\_inner\_arrays $>$ n\_regs}{
+  n\_inner\_arrays = n\_inner\_arrays / 2 \\
+  sz\_split = sz\_split + 1 
+}
+
+n\_inner\_regs = n\_regs / 2 \\
+\If{n\_inner\_arrays $>$ n\_inner\_regs \KwSty{or} sz\_split $>$ 0}{
+  // Rely on autovectorization, as there might be not enough registers to improve performance by tiling \\
+  \Return{$<$sz\_split, 0$>$}
+}
+
+// Compute unroll-and-jam factor for vector-register tiling \\
+n\_regs\_avail = n\_regs - n\_inner\_arrays \\
+uaj\_factor = $\lceil$n\_reg\_avail / n\_inner\_arrays$\rceil$ \\
+\Return{$<$sz\_split, uaj\_factor$>$}
+\end{algorithm}
+
+\clearpage
 

### 2014-02-14 paper/bare_conf.tex
Write out a first version of the introduction

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{Domain-Driven Code Transformation\\for Solving Partial Differential Equations}
+\title{A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations}
 
@@
 \section{Introduction}
-In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, programs are required that model phenomena.
-A feature common to these scientific computations is that they involve the resolution of partial differential equations (PDEs).
+In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, programs model phenomena by means of partial differential equations (PDEs).
 Numerical techniques, like Finite Volume Method (FVM) and Finite Element Method (FEM), are widely employed in real-world applications to approximate solutions of PDEs.
-Unstructured meshes are often used to discretize the PDE domain, since their geometric flexibility allows such PDE solvers to be extrimely effective.
-However, they are also the main responsible for performance degradation due to the introduction of indirect accesses (e.g.
+Unstructured meshes are often used to discretize the domain of a PDE, since their geometric flexibility allows solvers to be extrimely effective.
+However, they are also one of the main responsible for performance degradation, due to the introduction of indirect accesses (e.g.
 \texttt{A[B[i]]}) to access mesh components.
@@
 
-An unstructured mesh represents a discretization of the equation domain into two- or three-dimensional cells (or ``elements''), which define the smallest area within which the solution is sought.
-As the number of elements can be of the order of millions, a major issue is the time required to perform the numerical method, which can be of the order of hours or days.
-OP2~\cite{pyop2ics} and Lizst~\cite{lizst} are Domain Specific Languages that target mesh-based computations by allowing programmers to achieve highly-efficient parallel execution.
-The successful porting of Hydra, a CFD industrial application devised by Rolls Royce for turbomachinery design (based on FVM, roughly 50000 lines of code and a mesh size that can be over 100 millions edges), to OP2, demonstrates the effectiveness of this framework for implementing PDEs solvers~\cite{OP2-hydra}.
-Both OP2 and Lizst adopt a kernel-oriented programming model: the computation semantics is expressed through self-contained functions (or ``kernels''), each function applied in parallel to all items in a specific set of mesh components (e.g.
-edges, vertices, elements), with an implicit synchronization between the application of two different functions.
-Together with the indirections problem, having systematic optimizations for the kernels is the major concern in unstructured mesh applications.
+An unstructured mesh discretizes the domain of the equation into two- or three-dimensional cells (or ``elements''), within which the solution is sought.
+As the number of elements can be of the order of millions, a major issue is the time required to perform the numerical method, which can be hours or days.
+To address this issue, domain-specific languages (DSLs), which ease the programming burden and enable highly-efficient parallel execution, have been developed.
+The successful porting of Hydra, a CFD industrial application devised by Rolls Royce for turbomachinery design (based on FVM, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to the OP2 DSL~\cite{pyop2ics}, demonstrates the effectiveness of the DSL approach to implementing PDEs solvers~\cite{OP2-hydra}.
 
-In this paper, we describe a model-driven optimization strategy for a class of kernels used in Finite Element computations.
-The working set of these kernels is usually small enough to fit the L1 cache, and definitely fits the L2 cache when high-order methods are employed to improve the accuracy of the solution.
-However, the latter case is not considered in our work.
-This implies that main actors of our study will be register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation.
+OP2, as well as Lizst~\cite{lizst}, adopt a kernel-oriented programming model, in which the computation semantics is expressed through self-contained functions (or ``kernels'').
+A kernel is applied to all elements in a set of mesh components (e.g.
+edges, vertices, elements), with an implicit synchronization between the application of two consecutive kernels.
+On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread.
+Together with the indirections problem, kernel optimization is the major concern in unstructured mesh applications.
+In this paper, we tackle this problem by proposing a domain-driven optimization strategy for a class of kernels used in FEM.
+
+We focus on Finite Element Local Assembly (``assembly'', in the following), a fundamental step of FEM that covers an important fraction of the overall computation run-time, typically in the range 30$\%$-60$\%$.
+During the assembly phase, the solution of the PDE by executing a suitable kernel over all elements in the discretized domain.
+A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when (very) high-order methods are employed to improve the accuracy of the solution.
+However, we do not consider the latter case.
+An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 100, with a minimum value of 3, depending on the order of the method).
+With such small kernels, our work focuses on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation.
+
+Our study is conducted in the context of Firedrake, a multilayered system for the automated solution of PDEs through FEM.
+PDEs are written in Firedrake using a domain-specific language called Unified Form Language (UFL).
+The high-level UFL notation is used by the Fenics Form Compiler (FFC) to produce assembly kernels.
+At the core of Firedrake~\cite{firedrake} there is PyOP2~\cite{pyop2ics} - a python implementation of the OP2 abstraction - which triggers compilation of kernels using an available vendor compiler, and eventually manages parallel execution over the unstructured mesh.
+We exploit the domain knowledge and the structure inherent to FFC's assembly kernels to develop an optimizing compiler for all problems that are expressible in Firedrake.
+
+The compiler we developed is integrated with PyOP2, and already used in the release of Firedrake~\cite{firedrake-code}.
+This allows us to evaluate the proposed code transformations in a range of real-world problems, employing different low-order methods (from $p=1$ to $p=4$; the higher, the more expensive a kernel is, but better is the accuracy), and varying the topology of the mesh (2D, 3D, 3D hybrid structured-unstructured, which in turn affect the kernel cost).
+Performance results show that compiler-generated code for non-trivial assembly kernels, if not adequately transformed, is sub-optimal.
+A cost-model-driven sequence of source-to-source code transformations, aimed at improving register allocation and register data locality, can result in performance improvements up to 33$\%$ over ``softly-optimized'' code (i.e.
+where only basic transformations are performed, such as auto-vectorizable loop-invariant code motion, padding and data alignment), and up to 77$\%$ over the original FFC kernels.
+The contribution of this paper is therefore twofold:
+\begin{itemize}
+\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely Finite Element Local Assembly kernels.
+Our approach exploits domain knowledge to go beyond the limits of both vendor (e.g.
+$icc$) and research (e.g.
+those based on the polyhedral model) compilers.
+\item Automation of such code optimizations in Firedrake, a framework for the resoloution of PDEs through FEMs.
+This enables performing an in-depth performance evaluation of the proposed optimization strategy in real-world simulations.
+\end{itemize}
+
+The paper is organized as follows. ...
 
@@
 
-The compiler cost model. Used to find out tile size and split size.
+The compiler cost model. Used to find out suitable unroll-and-jam factor and split size.
 

### 2014-02-18 paper/bare_conf.tex
Add pictures of normalized runtimes
Start writing Background section

@@
 \ifCLASSINFOpdf
-  % \usepackage[pdftex]{graphicx}
+  \usepackage[pdftex]{graphicx}
   % declare the path(s) where your graphic files are
@@
 We focus on Finite Element Local Assembly (``assembly'', in the following), a fundamental step of FEM that covers an important fraction of the overall computation run-time, typically in the range 30$\%$-60$\%$.
-During the assembly phase, the solution of the PDE by executing a suitable kernel over all elements in the discretized domain.
+During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain.
 A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when (very) high-order methods are employed to improve the accuracy of the solution.
 However, we do not consider the latter case.
-An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 100, with a minimum value of 3, depending on the order of the method).
+An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 60, with a minimum value of 3, depending on the order of the method).
 With such small kernels, our work focuses on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation.
@@
 \section{Background}
-
 %Here we discuss about the computational characteristics of computational science kernels.
@@
 Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.
+
+Local assembly consists of evaluating so called element stiffness matrix and vector; in this work, we focus on computation of matrices, which is the costly part of the process.
+A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell of the discretized domain.
+Quadrature algorithms (based on numerical integration), which follow on from the mathematical description of a stiffness matrix, are widely-used in scientific codes to implement assembly~\cite{quadrature1, quadrature 2}.
+The FFC compiler used by Firedrake and Fenics can generate assembly kernels using optimized quadrature schemes.
+It is worth noticing, however, that problems and considerations similar to that exposed below can be found also in non-FFC assembly codes.
+
+The complexity of FFC-generated kernels depends on the mathematical problem being solved.
+In simpler cases, the loop nest (always characterized by affine memory accesses) is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few, small outer-product vector-operations.
+An example is provided in Figure~\ref{helmholtz-code}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$.
+In other situations, for instance when solving a non-linear problem like Burgers (see Figure~\ref{burgers-code}), the number of arrays involved in the computation of the stiffness matrix can be much larger (14 in the case of Burgers using $p=1$ Lagrange tetrahedrons).
+Also, constants evaluated in outer loops, acting as scaling factors of arrays, may be required; trip counts can be larger (the increment is proportional to the order of the method); arrays may be block-sparse.
+Despite such a big space of possible assembly implementations, it is still possible to identify a domain-specific structure common to all kernels, which can be turned into effective code transformations and simd-vectorization.
+
+%
 
@@
 Also, our results prove that current vendor compilers still have notable limitations, despite optimizing loop nests that are 1) affine, 2) have no stencil, and 3) ?
+\begin{figure}[h]
+\centerline{\includegraphics[scale=0.7]{Pictures/helmholtz-normalized.pdf}
+\label{fig_first_case}}
+\caption{Helmholtz}
+\end{figure}
+
+\begin{figure}[h]
+\centerline{\includegraphics[scale=0.7]{Pictures/advdiff-normalized.pdf}
+\label{fig_first_case}}
+\caption{Advection-Diffusion}
+\end{figure}
+
+\begin{figure}[h]
+\centerline{\includegraphics[scale=0.7]{Pictures/burgers-normalized.pdf}
+\label{fig_first_case}}
+\caption{Burgers}
+\end{figure}
+
+%\begin{figure*}[!t]
+%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{Pictures/helmholtz-normalized.pdf}%
+%\label{fig_first_case}}
+%\hfil
+%\subfloat[Case II]{\includegraphics[width=2.5in]{Pictures/advdiff-normalized.pdf}%
+%\label{fig_second_case}}}
+%\caption{Simulation results}
+%\label{fig_sim}
+%\end{figure*}
 

### 2014-02-18 paper/bare_conf.tex
Add code snippets: helmholtz and burgers

@@
 Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.
+\begin{algorithm}[t]
+\small
+\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).}
+\label{code:helmholtz}
+\KwSty{void} helmholtz(double A[3][3], double **coords) $\lbrace$\\
+~~// K, det = Compute Jacobian (coords) \\
+~~\\
+~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
+~~\KwSty{static const double} X$\_$D10[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
+~~\KwSty{static const double} X$\_$D01[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
+~~\\
+~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) \\
+~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
+~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$3; k++) \\
+~~~~~~~~A[j][k] += ((Y[i][k]*Y[i][j]+\\
+~~~~~~~~~~~+((K[1]*X$\_$D10[i][k]+K[3]*X$\_$D01[i][k])*\\
+~~~~~~~~~~~*(K[1]*X$\_$D10[i][j]+K[3]*X$\_$D01[i][j]))+\\
+~~~~~~~~~~~+((K[0]*X$\_$D10[i][k]+K[2]*X$\_$D01[i][k])*\\
+~~~~~~~~~~~*(K[0]*X$\_$D10[i][j]+K[2]*X$\_$D01[i][j])))*\\
+~~~~~~~~~~~*det*W3[i]);\\
+$\rbrace$
+\end{algorithm}
+
+\begin{algorithm}[t]
+\small
+\caption{Local assembly code generated by FFC for a Burgers problem (3D mesh, Lagrange $p=1$ elements).}
+\label{code:burgers}
+\KwSty{void} burgers(double A[12][12], double **c, double **w) $\lbrace$\\
+~~// K, det = Compute Jacobian (c) \\
+~~\\
+~~\KwSty{static const double} W5[5] = $\lbrace$...$\rbrace$\\
+~~\KwSty{static const double} X1$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
+~~\KwSty{static const double} X2$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
+~~//It follows 11 other basis functions definitions.\\
+~~...\\
+~~\\
+~~\KwSty{for} (\KwSty{int} i = 0; i$<$5; i++) $\lbrace$\\
+~~~~\KwSty{double} F0 = 0.0;\\
+~~~~//It follows 10 other declarations (F1, F2,...)\\
+~~~~...\\
+~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$12; r++) $\lbrace$\\
+~~~~~~F0 += (w[r][0]*X1$\_$D100[ip][r]);\\
+~~~~~~//It follows 10 analogous statements (F1, F2, ...)\\
+~~~~$\rbrace$\\
+~~~~...\\
+~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$12; j++) \\
+~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$12; k++) \\
+~~~~~~~~A[j][k] += (..(K[5]*F9)+(K[8]*F10))*Y1[i][j])+\\
+~~~~~~~~~+(((K[0]*X1$\_$D100[i][k])+(K[3]*X1$\_$D010[i][k])+\\
+~~~~~~~~~+(K[6]*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
+~~~~~~~~~+(..((K[2]*X2$\_$D100[i][k])+...+(K[8]*X2$\_$D001[i][k])*\\
+~~~~~~~~~*(K[2]*X2$\_$D100[i][j])+...+(K[8]*X2$\_$D001[i][j]))..)+\\
+~~~~~~~~~+ $<$a hundred of sum/muls go here$>$)..)*\\
+~~~~~~~~~*det*W5[i]);\\
+~~$\rbrace$ \\
+$\rbrace$
+\end{algorithm}
 
@@
 The complexity of FFC-generated kernels depends on the mathematical problem being solved.
-In simpler cases, the loop nest (always characterized by affine memory accesses) is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few, small outer-product vector-operations.
-An example is provided in Figure~\ref{helmholtz-code}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$.
-In other situations, for instance when solving a non-linear problem like Burgers (see Figure~\ref{burgers-code}), the number of arrays involved in the computation of the stiffness matrix can be much larger (14 in the case of Burgers using $p=1$ Lagrange tetrahedrons).
-Also, constants evaluated in outer loops, acting as scaling factors of arrays, may be required; trip counts can be larger (the increment is proportional to the order of the method); arrays may be block-sparse.
+In simpler cases, the loop nest, always characterized by affine 1-stride memory accesses, is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few, small outer-product vector-operations.
+An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$.
+The outer-product along the $j$ and $k$ dimensions computes the stiffness matrix $A$.
+In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of $A$ can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
+Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse.
 Despite such a big space of possible assembly implementations, it is still possible to identify a domain-specific structure common to all kernels, which can be turned into effective code transformations and simd-vectorization.
-
-%
 
@@
 
-\begin{algorithm}[h]
+\begin{algorithm}[t]
 \caption{The PyOP2 compiler.}
@@
 
-\begin{algorithm}[h]
+\begin{algorithm}[t]
 \caption{Procedure to estimate the most suitable unroll-and-jam factor and/or split size.}

### 2014-02-18 paper/bare_conf.tex
Add explanation about why we do not show blas results

@@
 Unstructured meshes are often used to discretize the domain of a PDE, since their geometric flexibility allows solvers to be extrimely effective.
-However, they are also one of the main responsible for performance degradation, due to the introduction of indirect accesses (e.g.
+However, they are also one of the main responsible for performance degradation, due to the need for indirect accesses (e.g.
 \texttt{A[B[i]]}) to access mesh components.
@@
 On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread.
-Together with the indirections problem, kernel optimization is the major concern in unstructured mesh applications.
+Together with the indirections problem, kernel optimization is one of the major concerns in unstructured mesh applications.
 In this paper, we tackle this problem by proposing a domain-driven optimization strategy for a class of kernels used in FEM.
@@
 
-The compiler we developed is integrated with PyOP2, and already used in the release of Firedrake~\cite{firedrake-code}.
-This allows us to evaluate the proposed code transformations in a range of real-world problems, employing different low-order methods (from $p=1$ to $p=4$; the higher, the more expensive a kernel is, but better is the accuracy), and varying the topology of the mesh (2D, 3D, 3D hybrid structured-unstructured, which in turn affect the kernel cost).
+The compiler we have developed is integrated with PyOP2, and already used in the released version of Firedrake~\cite{firedrake-code}.
+This allows us to evaluate the proposed code transformations in a range of real-world problems, employing different low-order methods (from $p=1$ to $p=4$; the higher, the more expensive is a kernel, but better is the accuracy), and varying the topology of the mesh (2D, 3D, 3D hybrid structured-unstructured, which in turn affect the kernel cost).
 Performance results show that compiler-generated code for non-trivial assembly kernels, if not adequately transformed, is sub-optimal.
@@
 
-%the focus of optimization should be on managing registers, exploiting ILP in its various forms, and reducing loop overhead
-
 % systematic optimization strategy -> discuss the importance of domain knowledge
@@
 In Firedrake all computational kernels are generated by a compiler, (the Fenics Form Compiler, FFC), whereas in OP2 applications kernels are provided by users.
-
-%What we do in computational science.
-Applications motivating our study.
-The programming model of lizst and op2: ``simple'' kernels applied to sets of mesh elements.
-The need for optimizing such kernels.
-Contributions of the paper:
-%\begin{itemize}
-%\item A \textbf{compiler-driven} optimisation approach that can be generalised to a wide variety of kernels employed in many computational science codes.
-The key feature common to these kernels is that they usually fit the L1 cache of commodity processors.
-%\item Automation of such code optimizations in a framework for the resoloution of partial differential equations through the Finite Element Method, called Firedrake.
-This allows us to perform an in-depth performance evaluation of the proposed optimization strategy in real-world computations.
-%\end{itemize}
 
@@
 The complexity of FFC-generated kernels depends on the mathematical problem being solved.
-In simpler cases, the loop nest, always characterized by affine 1-stride memory accesses, is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few, small outer-product vector-operations.
+In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few, small outer-product vector-operations.
 An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$.
-The outer-product along the $j$ and $k$ dimensions computes the stiffness matrix $A$.
 In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of $A$ can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
@@
 
-\section{A Domain-Driven Approach to Code Optimization}
+The class of kernels we are considering has, in particular, some peculiarities.
+1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest; this is not true in general, since the unstructured mesh might have bended elements that require the Jacobian be re-computed every $i$ iteration.
+2) The stiffness matrix $A$ is always the result of an outer-product computation along the $j$ and $k$ dimensions.
+3) Memory accesses along the three loop dimensions are always 1-stride .4) The $j$ and $k$ loops are interchangable, whereas to permute $i$ it might be necessary to precompute $F$ values and to introduce temporaries.
+5) Basis functions arrays (denoted by $X$, $Y$, ...) are constants, and many sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
+In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.
+
+One attractive possibility is to transform the loop nest into a sequence of calls to BLAS routines.
+There are several issues that make achieving this difficult and, potentially, not effective.
+The first problem concerns automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of the memory access pattern.
+Related to this problem is the fact that, in general, many sub-expressions should be pre-computed (i.e.
+computed once before the loop nest), which leads to using temporaries and, possibly, to working sets that do not fit the L1 cache (for $p>2$).
+In addition, using BLAS routines may impact data locality, since the same matrix can appear in multiple products, implying, at least, a certain degree of register spilling.
+Finally, it is known that BLAS routines perform far from peak performance when the dimension of the matrices is rather small~\cite{nek5000}.
+Hand-made implementations of the simple Helmholtz problem ($p=1,2,3,4$) as a sequence of BLAS ($dgemm$) routines have demonstrated that performance is worse than that shown in Section~\ref{sec:perf-results}.
+A comprehensive study concerning BLAS optimization is therefore left as further work.
+
+\section{Code Transformations}
+\label{sec:code-transf}
 %All the work done in the polyhedral model should be useful, but our kernels fit L1, so they need special attention for what concerns key aspects like managing registers, exploiting ILP in its various forms, and reducing loop overhead.
@@
 \section{Performance Evaluation}
+\label{sec:perf-results}
 %Performance evaluation.

### 2014-02-19 paper/bare_conf.tex
Add a bit of code transformations

@@
 \and
-\IEEEauthorblockN{Paul H. J. Kelly}
+\IEEEauthorblockN{Florian Rathgeber}
 \IEEEauthorblockA{Department of Computing\\
@@
 London, UK\\
-Email: p.kelly@imperial.ac.uk}
+Email: f.rathgeber10@imperial.ac.uk}
 \and
@@
 Email: d.ham@imperial.ac.uk}}
+
+%\and
+%\IEEEauthorblockN{Paul H. J. Kelly}
+%\IEEEauthorblockA{Department of Computing\\
+%Imperial College London\\
+%London, UK\\
+%Email: p.kelly@imperial.ac.uk}
 
@@
 ~~~~~~~~~*(K[2]*X2$\_$D100[i][j])+...+(K[8]*X2$\_$D001[i][j]))..)+\\
-~~~~~~~~~+ $<$a hundred of sum/muls go here$>$)..)*\\
+~~~~~~~~~+ $<$roughly a hundred of sum/muls go here$>$)..)*\\
 ~~~~~~~~~*det*W5[i]);\\
@@
 
-\section{Code Transformations}
+\section{Domain-Driven Code Transformations}
 \label{sec:code-transf}
@@
 %Follows a description of code transformations using our compiler-driven approach.
+
+% Autovectorizable LICM -> show code snippet?
+From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only.
+Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion ($licm$).
+Vendor compilers apply $licm$, although not in the systematic way we need in our assembly kernels.
+We want to overcome two deficiencies that both $icc$ and $gcc$ have.
+First, these compilers can identify sub-expressions that are invariant with respect to the innermost loop only.
+This is an issue for sub-expressions depending on $i$-$k$ only, which are not automatically lifted.
+Second, the hoisted code is scalar, i.e.
+it is not subjected to auto-vectorization.
+We work around these limitations by doing source-level $licm$.
+In addition, we pre-compute all values that an invariant sub-expression assumes along the fastest varying dimension.
+This is implemented by introducing temporary arrays and new loops in the nest.
+At the price of both extra memory for storing temporaries and reduced register locality, the gain is that lifted terms can be auto-vectorized, because part of an inner loop.
+Note that given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which can be otherwise significant.
+
+% Padding and Data Alignment
+Auto-vectorization of assembly code that computes $A$ can be less effective if data are not aligned and if the length of the innermost loop is small compared to the hardware vector length ($vl$).
+Data alignment is enforced in two steps.
+Initially, both arrays and matrices are allocated to addresses that are multiples of $vl$.
+Then, matrices are padded by rounding the number of columns to the nearest multiple of $vl$.
+For example, assume the original size of a matrix is 3$\times$3, and the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use double-precision floating-point values (64 bits).
+Then, the padded matrix on this architecture is 3$\times$4.
+Padding of all matrices involved in the evaluation of the stiffness matrix allows us to safely round the loop trip count to the nearest multiple of $vl$.
+This avoids the introduction of a remainder (scalar) loop from the compiler, which would be responsible for inefficient vectorization.
+
+% Register tiling in its two forms
+One notable problem assembly kernels are exposed to concerns register allocation and locality.
+There exist two critical situations that can negatively impact performance.
+The first occurs when trip counts are sufficiently large to prevent register reuse along the innermost dimension ($k$).
+A combination of the two scenarios is also possible.
+We want to exploit the small size of the iteration space and the structure of the computation to optimize register management.
+
+% remember: op vectorization PLUS unroll and jam !!
+
+% Splitting
+
+% mention (obvious) composability
 
@@
 
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{Pictures/helmholtz-normalized.pdf}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{Pictures/advdiff-normalized.pdf}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
+% We fix a specific loop permutation (i esterno perche minimizza.... swapping j or k, after LICM, is equivalent)
 

### 2014-02-19 paper/bare_conf.tex
Refine padding/data alignment, start register tiling

@@
 Then, the padded matrix on this architecture is 3$\times$4.
+The compiler is explicitly informed about data alignment using a suitable pragma.
 Padding of all matrices involved in the evaluation of the stiffness matrix allows us to safely round the loop trip count to the nearest multiple of $vl$.
@@
 % Register tiling in its two forms
-One notable problem assembly kernels are exposed to concerns register allocation and locality.
-There exist two critical situations that can negatively impact performance.
-The first occurs when trip counts are sufficiently large to prevent register reuse along the innermost dimension ($k$).
+One notable problem assembly kernels are exposed to concerns register allocation and register locality.
+The critical situation occurs when loop trip counts and number of accessed matrices along the $j$ and $k$ dimensions are such that available vector registers are under-utilized.
+Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management in order to maximize performance.
+Canonical optimizations (i.e.
+non domain-driven), such as loop interchange, unroll, unroll-and-jam and register tiling are typically employed to deal with this problem.
+In our compiler we support these optimizations either by means of explicit source code transformation (interchange, register tiling, unroll-and-jam) or indirectly advising the compiler through standard pragmas (unroll).
+With register tiling, in particular, we slice the innermost loop into rectangular blocks of identical size (except the ``reminder'' block), and we apply unrolling to individual blocks.
+
+%are sufficiently large to prevent register reuse along the innermost dimension $k$.
 A combination of the two scenarios is also possible.

### 2014-02-20 paper/bare_conf.tex
Improve code transformation description. Add splitting and snippets

@@
 A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell of the discretized domain.
-Quadrature algorithms (based on numerical integration), which follow on from the mathematical description of a stiffness matrix, are widely-used in scientific codes to implement assembly~\cite{quadrature1, quadrature 2}.
+Quadrature algorithms (based on numerical integration), which follow on from the mathematical description of a stiffness matrix, are widely-used in scientific codes to implement assembly~\cite{quadrature1, quadrature2}.
 The FFC compiler used by Firedrake and Fenics can generate assembly kernels using optimized quadrature schemes.
@@
 
+\begin{algorithm}[t]
+\small
+\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).
+The padding, data alignment and $licm$ optimizations are applied.
+Data alignment and padding relate to an AVX machine.
+In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$; in general, this could not be the case.}
+\label{code:helmholtz-licm}
+\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
+~~// K, det = Compute Jacobian (coords) \\
+~~\\
+~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
+~~\KwSty{static const double} X$\_$D10[3][4] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
+~~\KwSty{static const double} X$\_$D01[3][4] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
+~~\\
+~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
+~~~~double LI$\_$0[4];\\
+~~~~double LI$\_$1[4];\\
+~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
+~~~~~~LI$\_$0[r] = ((K[1]*X$\_$D10[ip][r])+(K[3]*X$\_$D01[ip][r]));\\
+~~~~~~LI$\_$1[r] = ((K[0]*X$\_$D10[ip][r])+(K[2]*X$\_$D01[ip][r]));\\
+~~~~$\rbrace$\\
+~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
+~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
+~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+\\
+~~~~~~~~~~~~~~~~~~~~+LI$\_$1[k]*LI$\_$1[j])*det*W3[i]);\\
+~~$\rbrace$\\
+$\rbrace$
+\end{algorithm}
+
+\begin{algorithm}[t]
+\small
+\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).
+The padding, data alignment, $licm$ and $split$ optimizations are applied.
+Data alignment and padding relate to an AVX machine.
+In this specific case, the split factor is 2.}
+\label{code:helmholtz-split}
+\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
+~~// K, det = Compute Jacobian (coords) \\
+~~// Declaration of basis function matrices \\
+~~\\
+~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
+~~~~double LI$\_$0[4];\\
+~~~~double LI$\_$1[4];\\
+~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
+~~~~~~LI$\_$0[r] = ((K[1]*X$\_$D10[ip][r])+(K[3]*X$\_$D01[ip][r]));\\
+~~~~~~LI$\_$1[r] = ((K[0]*X$\_$D10[ip][r])+(K[2]*X$\_$D01[ip][r]));\\
+~~~~$\rbrace$\\
+~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
+~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
+~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
+~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
+~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
+~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
+~~$\rbrace$\\
+$\rbrace$
+\end{algorithm}
+
 % Autovectorizable LICM -> show code snippet?
@@
 
-% Register tiling in its two forms
+% Register allocation and register locality
 One notable problem assembly kernels are exposed to concerns register allocation and register locality.
-The critical situation occurs when loop trip counts and number of accessed matrices along the $j$ and $k$ dimensions are such that available vector registers are under-utilized.
+The critical situation occurs when loop trip counts and number of accessed matrices along the inner dimensions are such that available vector registers are under-utilized.
 Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management in order to maximize performance.
-Canonical optimizations (i.e.
-non domain-driven), such as loop interchange, unroll, unroll-and-jam and register tiling are typically employed to deal with this problem.
+Canonical optimizations, such as loop interchange, unroll, unroll-and-jam and register tiling are typically employed to deal with this problem.
 In our compiler we support these optimizations either by means of explicit source code transformation (interchange, register tiling, unroll-and-jam) or indirectly advising the compiler through standard pragmas (unroll).
-With register tiling, in particular, we slice the innermost loop into rectangular blocks of identical size (except the ``reminder'' block), and we apply unrolling to individual blocks.
+With register tiling, in particular, we slice the innermost loop into rectangular blocks of identical size (except the ``reminder'' block), and we apply unrolling to individual blocks to expose ILP.
 
@@
 
+% Splitting
+In complex kernels, like Burgers in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
+If the number of variables independent of the innermost-loop dimension, i.e.
+basis function matrices, temporaries introduced by $licm$ and constants, is larger than the available registers on the CPU, it is likely to obtain poor register reuse.
+For example, applying $licm$ to Burgers on a 3D mesh needs 33 temporaries for the $ijk$ loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers.
+One potential solution to this problem consists of suitably ``splitting'' the computation of the stiffness matrix $A$ into multiple sub-expressions; an example, for the simpler Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}.
+Splitting an expression has, in general, several drawbacks.
+Firstly, it increases the number of accesses to $A$ proportially to the ``split factor'', which is the number of sub-expressions produced.
+Secondly, depending on how the split is executed, it can lead to redundant computation (e.g.
+the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}).
+Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same matrix could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
+Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself are based upon heuristics that aim at minimizing the aforementioned issues.
+
 % remember: op vectorization PLUS unroll and jam !!
 
-% Splitting
+\begin{algorithm}[t]
+\small
+\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=2$ elements).
+The padding, data alignment, $licm$ and $op$-$vect$ optimizations are applied.
+Data alignment and padding relate to an AVX machine.
+In this specific case, the unroll-and-jam factor is 1.}
+\label{code:helmholtz-split}
+\KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
+~~// K, det = Compute Jacobian (coords) \\
+~~// Declaration of basis function matrices \\
+~~\\
+~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
+~~~~// Do loop-invariant code motion \\
+~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
+~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
+~~~~~~~~// Call $Load$ and $set$ intrinsics \\
+~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
+~~~~~~~~// One $permute\_pd$ intrinsics per $k$-loop $load$\\
+~~~~~~~~// Compute A[1,2],A[2,1],A[3,4],A[4,3] \\
+~~~~~~~~// One $permute2f128\_pd$ intrinsics per $k$-loop $load$\\
+~~~~~~~~// ...\\
+~~~~~~$\rbrace$\\
+~~~~// Do Remainder loop (from $j=4$ to $j=6$)\\
+~~$\rbrace$\\
+~~// Restore the storage layout:\\
+~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
+~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
+~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
+~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
+~~~~~~// Load A[j+1][k], A[j+2][k], A[j+3][k]\\
+~~~~~~r4 = $\_$mm256$\_$unpackhi$\_$pd (r1, r0);\\
+~~~~~~r5 = $\_$mm256$\_$unpacklo$\_$pd (r0, r1);\\
+~~~~~~r6 = $\_$mm256$\_$unpackhi$\_$pd (r2, r3);\\
+~~~~~~r7 = $\_$mm256$\_$unpacklo$\_$pd (r3, r2);\\
+~~~~~~r0 = $\_$mm256$\_$permute2f128$\_$pd (r5, r7, 32);\\
+~~~~~~r1 = $\_$mm256$\_$permute2f128$\_$pd (r4, r6, 32);\\
+~~~~~~r2 = $\_$mm256$\_$permute2f128$\_$pd (r7, r5, 49);\\
+~~~~~~r3 = $\_$mm256$\_$permute2f128$\_$pd (r6, r4, 49);\\
+~~~~~~$\_$mm256$\_$store$\_$pd ($\&$A[j+0][k], r0);\\
+~~~~~~// Store A[j+1][k], A[j+2][k], A[j+3][k]\\
+~~~~$\rbrace$\\
+~~$\rbrace$\\
+$\rbrace$
+\end{algorithm}
 

### 2014-02-21 paper/bare_conf.tex
Add outer-product vectorization subsection

@@
 %
-%\usepackage[cmex10]{amsmath}
+\usepackage[cmex10]{amsmath}
 % A popular package from the American Mathematical Society that provides
@@
 % Autovectorizable LICM -> show code snippet?
+\subsection{Auto-vectorizable Loop-invariant Code Motion}
 From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only.
@@
 % Padding and Data Alignment
+\subsection{Padding and Data Alignment}
 Auto-vectorization of assembly code that computes $A$ can be less effective if data are not aligned and if the length of the innermost loop is small compared to the hardware vector length ($vl$).
@@
 % Register allocation and register locality
+\subsection{Toward Improving Register Locality}
 One notable problem assembly kernels are exposed to concerns register allocation and register locality.
@@
 
+% remember: op vectorization PLUS unroll and jam !!
+\subsection{Outer-product Vectorization}
+Register tiles along both inner dimensions can be determined as well.
+Our domain-driven strategy, based on the observation that the evaluation of the stiffness matrix can be reduced to a summation of outer products along the $j$ and $k$ dimensions, aims at optimizing register reuse, in terms of both minimization of load operations and achievement of larger tiles, over a canonical implementation.
+If we consider the code snippet in Figure~\ref{code:helmholtz-licm}, we can notice that the computation of $A$ is abstractly expressible as
+\begin{equation}
+\label{outer-product}
+A_{jk} = \sum_{\substack{
+  x \in B' \subseteq B \\ 
+  y \in B'' \subseteq B}} 
+x_j\cdot y_k ~~~~~~ j,k = 0,...,4
+\end{equation}
+where $B$ is the set of all matrices (or temporaries) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets.
+Without loss of generality, the presence of other constant or $i$-dependent variables can be momentarily neglected.
+Note that regardless of the specific input problem, the assembly computation is always reducible to this form.
+Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix with just 2 vector registers.
+Once the first four entries are evaluated, values in a register are shuffled and the same product is performed again.
+Standard compiler auto-vectorization ($gcc$ and $icc$), instead, would employ 5 registers, 4 of them target of a broadcast operation (i.e.
+``splatting'' a single value over the four locations of an AVX register) along the outer dimension, and 1 of a load operation along the inner dimension.
+In our outer-product vectorization ($op$-$vect$), the 4 broadcast accesses to the L1 cache are avoided by an initial load and by shuffling.
+More importantly, we avoid using $vf$-1 registers per each outer-loop-dependent variable (i.e.
+all $x$ terms in Equation~\ref{outer-product}), with $vf$ being the vector length.
+This is a considerable gain, which allows us to slice the iteration space into bigger tiles, implemented directly by vector registers.
+
+\begin{figure}[h]
+\centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect.pdf}
+\label{fig:vect-by-vect}}
+\caption{Outer-product vectorization by permuting values in a vector register.}
+\end{figure}
+
+The storage layout of $A$, however, is incorrect after the application of $op$-$vect$.
+We efficiently restore it with a sequence of vector shuffle following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed just once outside of the $ijk$ loop nest.
+The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.
+
+\begin{figure}[h]
+\centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}
+\label{fig:vect-restore}}
+\caption{Restoring the storage layout after $op$-$vect$.}
+\end{figure}
+
 %are sufficiently large to prevent register reuse along the innermost dimension $k$.
@@
 % Splitting
+\subsection{Assembly splitting}
 In complex kernels, like Burgers in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
@@
 Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same matrix could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
-Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself are based upon heuristics that aim at minimizing the aforementioned issues.
-
-% remember: op vectorization PLUS unroll and jam !!
+Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.
 
@@
 In this specific case, the unroll-and-jam factor is 1.}
-\label{code:helmholtz-split}
+\label{code:helmholtz-opvect}
 \KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
@@
 \end{algorithm}
+
+Table~\ref{} summarizes ...
 

### 2014-02-21 paper/bare_conf.tex
Add PyOP2 Compiler description (not finished though)

@@
 % \documentclass[conference]{../sty/IEEEtran}
+
+% TODO: ok?
+\usepackage{tabulary}
 
@@
 
-\author{\IEEEauthorblockN{Fabio Luporini}
-\IEEEauthorblockA{Department of Computing\\
-Imperial College London\\
-London, UK\\
-Email: f.luporini12@imperial.ac.uk}
-\and
-\IEEEauthorblockN{Florian Rathgeber}
-\IEEEauthorblockA{Department of Computing\\
-Imperial College London\\
-London, UK\\
-Email: f.rathgeber10@imperial.ac.uk}
-\and
-\IEEEauthorblockN{David Ham}
-\IEEEauthorblockA{Department of Computing\\
-Imperial College London\\
-London, UK\\
-Email: d.ham@imperial.ac.uk}}
-
+%\author{\IEEEauthorblockN{Fabio Luporini}
+%\IEEEauthorblockA{Department of Computing\\
+%Imperial College London\\
+%London, UK\\
+%Email: f.luporini12@imperial.ac.uk}
+%\and
+%\IEEEauthorblockN{Florian Rathgeber}
+%\IEEEauthorblockA{Department of Computing\\
+%Imperial College London\\
+%London, UK\\
+%Email: f.rathgeber10@imperial.ac.uk}
+%\and
+%\IEEEauthorblockN{Ana Lucia Varbanescu}
+%\IEEEauthorblockA{Department of Computing\\
+%Imperial College London\\
+%London, UK\\
+%Email: a.l.varbanescu@uva.nl}
+%\and
+%\IEEEauthorblockN{Gheorghe-Teodor Bercea}
+%\IEEEauthorblockA{Department of Computing\\
+%Imperial College London\\
+%London, UK\\
+%Email: gb308@doc.ic.ac.uk}
 %\and
@@
 %Email: p.kelly@imperial.ac.uk}
+%\and
+%\IEEEauthorblockN{David Ham}
+%\IEEEauthorblockA{Department of Computing\\
+%Imperial College London\\
+%London, UK\\
+%Email: d.ham@imperial.ac.uk}}
+
+\author{\IEEEauthorblockN{Fabio Luporini\IEEEauthorrefmark{1},
+Florian Rathgeber\IEEEauthorrefmark{1},
+Ana Lucia Varbanescu\IEEEauthorrefmark{2}, 
+Gheorghe-Teodor Bercea\IEEEauthorrefmark{1}, \\
+Paul H. J. Kelly\IEEEauthorrefmark{1},
+and
+David Ham\IEEEauthorrefmark{1}}
+\IEEEauthorblockA{\IEEEauthorrefmark{1}Imperial College London, Department of Computing,\\
+\IEEEauthorrefmark{2}TU Delft, Faculty of Engineering, Mathematics and Computer Science (EWI)\\
+\IEEEauthorrefmark{1}\{f.luporini12$|$f.rathgeber10$|$gheorghe-teodor.bercea08$|$p.kelly$|$d.ham@imperial.ac.uk\}@imperial.ac.uk,\\
+\IEEEauthorrefmark{2}a.l.varbanescu@uva.nl
+}
+}
 
@@
 During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain.
-A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when (very) high-order methods are employed to improve the accuracy of the solution.
+A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution.
 However, we do not consider the latter case.
@@
 \subsection{Padding and Data Alignment}
-Auto-vectorization of assembly code that computes $A$ can be less effective if data are not aligned and if the length of the innermost loop is small compared to the hardware vector length ($vl$).
+Auto-vectorization of assembly code that computes $A$ can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length ($vl$).
 Data alignment is enforced in two steps.
@@
 For example, assume the original size of a matrix is 3$\times$3, and the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use double-precision floating-point values (64 bits).
-Then, the padded matrix on this architecture is 3$\times$4.
+Then, a padded matrix on this architecture will be 3$\times$4.
 The compiler is explicitly informed about data alignment using a suitable pragma.
@@
 In our outer-product vectorization ($op$-$vect$), the 4 broadcast accesses to the L1 cache are avoided by an initial load and by shuffling.
-More importantly, we avoid using $vf$-1 registers per each outer-loop-dependent variable (i.e.
-all $x$ terms in Equation~\ref{outer-product}), with $vf$ being the vector length.
+More importantly, we avoid using $vl$-1 registers per each outer-loop-dependent variable (i.e.
+all $x$ terms in Equation~\ref{outer-product}), with $vl$ being the vector length.
 This is a considerable gain, which allows us to slice the iteration space into bigger tiles, implemented directly by vector registers.
@@
 \centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}
-\label{fig:vect-restore}}
+\label{fig:restore-layout}}
 \caption{Restoring the storage layout after $op$-$vect$.}
@@
 
-Table~\ref{} summarizes ...
+Table~\ref{table:code-transformations} summarizes the code transformations described so far.
+Given that many of these transformations depend on some parameters (e.g.
+tile size), we need a mechanism to prune such a large space of optimization.
+This aspect is treated in Section~\ref{sec:pyop2-compiler}
+
+\begin{table}[h]
+\begin{tabulary}{1.0\columnwidth}{|C|C|}
+\hline
+Name (Abbreviation) & Parameter \\\hline\hline
+Auto-vectorizable Loop-invariant code motion ($licm$) &   \\ \hline
+Padding &  \\ \hline
+Data Alignment & \\ \hline
+Loop interchange      & loops  \\ \hline
+Loop unrolling  & unroll factor \\ \hline
+Register tiling & tile size \\ \hline
+Outer-product vectorization ($op$-$vect$) & tile size \\ \hline
+Assembly splitting ($split$) & split point, split factor \\ \hline
+\end{tabulary}
+\caption{Overview of code transformations for FFC-generated assembly kernels.}
+\label{table:code-transformations}
+\end{table}
 
@@
 
-\clearpage
-
 \section{The PyOP2 Compiler}
+\label{sec:pyop2-compiler}
 %Design and structure of the PyOP2 Compiler.
@@
 
-The compiler structure
+Our approach to solving partial differential equations is based on domain-specific languages and libraries.
+Fenics and Firedrake provide the high-level Unified Form Language (UFL) to write problems in a notation resembling mathematical equations (``forms'').
+In Fenics, this specification is translated by a form compiler (FFC) into C code, which is subsequently just-in-time compiled and executed on the target architecture.
+In Firedrake we adopt a completely different approach.
+We think that FFC should produce a high-level representation of a Finite Element assembly kernel, and should abstract from the specific details of the underlying architecture.
+We have modified FFC, therefore, to build Abstract Syntax Trees (AST), rather than C code.
+ASTs are then passed to PyOP2, which implements the core of Firedrake, where parallel execution over the unstructured mesh is managed.
+We have developed a compiler that, given an AST conformed to the PyOP2 specification (such as those produced by the new FFC), can apply the transformations described in Section~\ref{sec:code-transf} and, subsequently, generate C code.
+The C representation is eventually just-in-time compiled for one of the backends that are currently supported by PyOP2, namely OpenMP, MPI, OpenMP+MPI, CUDA, OpenCL.
+Our compiler is now fully integrated with PyOP2, so it supports any kind of problem expressible in UFL.
+
+The PyOP2 compiler has been thought to support a range of higher-level languages and applications that concern the resolution of partial differential equations; Firedrake, in terms of its components UFL and FFC, is one of them.
+We believe, however, that many of the discussed optimizations are generalizable and applicable to non-Firedrake kernels.
+For example, it is always the case that Finite Element Local Assembly using a quadrature strategy leads to a (more or less complex) loop nest and outer-product-like computations.
+
+The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}.
+The AST is initially inspected, looking for the presence of an iteration space and other domain-specific information provided by the upper layer.
+If the kernel is just a sequence of statement, then so-called inter-kernel vectorization, in which a non-affine loop over mesh elements is vectorized, can be applied.
+This might be useful in a Finite Volume context, where kernels often lack an iteration space representing the bulk of the computation.
+Although not implemented yet, Reguly et.
+al.
+already proved the profitability of this technique in the OP2 framework~\cite{reguly-inter-kernel}.
+If, instead, an iteration space is found and the underlying platform is a GPU, the compiler seeks fully-parallel loops (i.e.
+loops whose iterations can be executed in any order without need for synchronization) and hoist them out at the level of the ``wrapper'' (i.e.
+the function that iterates over mesh elements and invokes the kernel); this is aimed at optimizing register management on GPUs.
+
+If the backend is a CPU and the AST comes from Firedrake, then a sequence of optimization steps are executed.
+A cost model, currently under development, is queried to estimate the potential gain (or loss) coming from the $split$ and $op$-$vect$ transformations.
+The model suggests a split factor and a tile size.
+Firstly, the compiler applies $licm$, which introduces temporary arrays, and on top of this padding and data alignment are enforced.
+These three transformations are always applied, since they usually provide a significant performance improvement (this will be shown in Section~\ref{sec:perf-results}).
+Finally, $split$ and/or $op$-$vect$ take place.
 
@@
   ast.apply\_inter\_kernel\_vectorization(wrapper)
-  \Return{ast.from\_ast\_to\_c()}
+  \Return{wrapper + ast.from\_ast\_to\_c()}
+}
+\If{isa.backend == gpu}{
+  \If{it\_space}{ast.extract\_iteration\_space(wrapper)}
+  \Return{wrapper + ast.from\_ast\_to\_c()}
 }
 plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs) \\
-~~\\
 // Optimize ast based on plan \\
 ast.apply\_licm() \\
+ast.padding() \\
+ast.data\_align() \\
 \If{plan.sz\_split}{
@@
 }
-ast.padding() \\
-ast.data\_align() \\
-\Return{ast.from\_ast\_to\_c()}
+\Return{wrapper + ast.from\_ast\_to\_c()}
 \end{algorithm}
 
-The compiler cost model. Used to find out suitable unroll-and-jam factor and split size.
+The cost model ..... TODO
 

### 2014-02-22 paper/bare_conf.tex
Huge refactor of intro, background, compiler

@@
 In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, programs model phenomena by means of partial differential equations (PDEs).
-Numerical techniques, like Finite Volume Method (FVM) and Finite Element Method (FEM), are widely employed in real-world applications to approximate solutions of PDEs.
-Unstructured meshes are often used to discretize the domain of a PDE, since their geometric flexibility allows solvers to be extrimely effective.
-However, they are also one of the main responsible for performance degradation, due to the need for indirect accesses (e.g.
-\texttt{A[B[i]]}) to access mesh components.
-A typical problem, therefore, is that implementing a scientific application requires a great effort, other than domain knowledge, and even more time is needed to optimize their execution time.
+Numerical techniques, like Finite Volume Method and Finite Element Method, are widely employed in real-world applications to approximate solutions of PDEs.
+Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective.
+The solution is sought in each cell of the discretized domain by applying suitable numerical kernels.
+Iteration over mesh components is also expensive, due to the presence of indirect memory accesses.
+As the number of cells can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days.
+To address this issue, domain-specific languages (DSLs) have been developed.
+The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2ics}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{OP2-hydra}.
 
-An unstructured mesh discretizes the domain of the equation into two- or three-dimensional cells (or ``elements''), within which the solution is sought.
-As the number of elements can be of the order of millions, a major issue is the time required to perform the numerical method, which can be hours or days.
-To address this issue, domain-specific languages (DSLs), which ease the programming burden and enable highly-efficient parallel execution, have been developed.
-The successful porting of Hydra, a CFD industrial application devised by Rolls Royce for turbomachinery design (based on FVM, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to the OP2 DSL~\cite{pyop2ics}, demonstrates the effectiveness of the DSL approach to implementing PDEs solvers~\cite{OP2-hydra}.
-
-OP2, as well as Lizst~\cite{lizst}, adopt a kernel-oriented programming model, in which the computation semantics is expressed through self-contained functions (or ``kernels'').
+OP2 adopts a kernel-oriented programming model, in which the computation semantics is expressed through self-contained functions (``kernels'').
 A kernel is applied to all elements in a set of mesh components (e.g.
@@
 Together with the indirections problem, kernel optimization is one of the major concerns in unstructured mesh applications.
-In this paper, we tackle this problem by proposing a domain-driven optimization strategy for a class of kernels used in FEM.
+In this paper, we tackle this problem by proposing a domain-driven optimization strategy for a class of kernels used in Finite Element Methods.
 
-We focus on Finite Element Local Assembly (``assembly'', in the following), a fundamental step of FEM that covers an important fraction of the overall computation run-time, typically in the range 30$\%$-60$\%$.
+We focus on Local Assembly (``assembly'', in the following), a fundamental step of a Finite Element Method that covers an important fraction of the overall computation run-time, typically in the range 30$\%$-60$\%$.
 During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain.
@@
 However, we do not consider the latter case.
-An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 60, with a minimum value of 3, depending on the order of the method).
-With such small kernels, our work focuses on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation.
+An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 40, with a minimum value of 3, depending on the order of the method).
+With such small kernels, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation.
 
-Our study is conducted in the context of Firedrake, a multilayered system for the automated solution of PDEs through FEM.
-PDEs are written in Firedrake using a domain-specific language called Unified Form Language (UFL).
-The high-level UFL notation is used by the Fenics Form Compiler (FFC) to produce assembly kernels.
-At the core of Firedrake~\cite{firedrake} there is PyOP2~\cite{pyop2ics} - a python implementation of the OP2 abstraction - which triggers compilation of kernels using an available vendor compiler, and eventually manages parallel execution over the unstructured mesh.
-We exploit the domain knowledge and the structure inherent to FFC's assembly kernels to develop an optimizing compiler for all problems that are expressible in Firedrake.
+Since loop nests are affine, one possibility is to rely on polyhedral compilation for improving performance.
+There are, however, several limitations.
+The first problem is that a polyhedral compiler reasons on iteration spaces, whereas some of the transformations we propose rely on analysis of the data space.
+Our attempts at using PoCC's optimizations~\cite{PoCC}, especially register tiling, were unsuccessful.
+A domain-specific transformation of the iteration space, which is not expressible in the polyhedral model because of its non-affine nature, also showed a significant performance gain due to improved register locality.
 
-The compiler we have developed is integrated with PyOP2, and already used in the released version of Firedrake~\cite{firedrake-code}.
-This allows us to evaluate the proposed code transformations in a range of real-world problems, employing different low-order methods (from $p=1$ to $p=4$; the higher, the more expensive is a kernel, but better is the accuracy), and varying the topology of the mesh (2D, 3D, 3D hybrid structured-unstructured, which in turn affect the kernel cost).
-Performance results show that compiler-generated code for non-trivial assembly kernels, if not adequately transformed, is sub-optimal.
+Assembly kernels could be turned into a sequence of calls to BLAS routines, although it is complicated to achieve and, potentially, ineffective.
+One issue concerns automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of the memory access pattern.
+In addition, register locality would be affected, since the same matrix can appear in a number of products.
+Finally, it is known that BLAS routines perform far from peak performance when the dimension of the matrices is small~\cite{nek5000}.
+Hand-made implementations of the simple Helmholtz assembly kernel as a sequence of BLAS $dgemm$ calls, for a range of polynomial orders ($p=1,2,3,4$), has showed that the achieved performance is worse than that exposed in Section~\ref{sec:perf-results}.
+
+Given the constraints on polyhedral compilation and linear-algebra-specialized libraries, we address assembly optimization by studying a set of domain-specific code transformations, applicable to a wide class of problems.
+We have developed a domain-specific optimizing compiler and we have integrated it with Firedrake, a system for solving PDEs through the Finite Element Method~\cite{firedrake} based on the OP2 abstraction.
+This allows us to evaluate our code transformations in a range of real-world problems, varying parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the mesh type (2D, 3D, 3D hybrid structured-unstructured).
+
+Performance results show that Firedrake-generated code for non-trivial assembly kernels, if not adequately optimized, is sub-optimal.
 A cost-model-driven sequence of source-to-source code transformations, aimed at improving register allocation and register data locality, can result in performance improvements up to 33$\%$ over ``softly-optimized'' code (i.e.
-where only basic transformations are performed, such as auto-vectorizable loop-invariant code motion, padding and data alignment), and up to 77$\%$ over the original FFC kernels.
+where only basic transformations are performed, such as auto-vectorizable loop-invariant code motion, padding and data alignment), and up to 77$\%$ over the original kernels.
 The contribution of this paper is therefore twofold:
 \begin{itemize}
-\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely Finite Element Local Assembly kernels.
+\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely Local Assembly in the context of the Finite Element Method.
 Our approach exploits domain knowledge to go beyond the limits of both vendor (e.g.
@@
 those based on the polyhedral model) compilers.
-\item Automation of such code optimizations in Firedrake, a framework for the resoloution of PDEs through FEMs.
-This enables performing an in-depth performance evaluation of the proposed optimization strategy in real-world simulations.
+\item Automation of such code optimizations in Firedrake. This enables performing an in-depth performance evaluation of the proposed optimization strategy in real-world simulations.
 \end{itemize}
@@
 \small
-\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).}
+\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).}
 \label{code:helmholtz}
@@
 A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell of the discretized domain.
-Quadrature algorithms (based on numerical integration), which follow on from the mathematical description of a stiffness matrix, are widely-used in scientific codes to implement assembly~\cite{quadrature1, quadrature2}.
-The FFC compiler used by Firedrake and Fenics can generate assembly kernels using optimized quadrature schemes.
-It is worth noticing, however, that problems and considerations similar to that exposed below can be found also in non-FFC assembly codes.
+Numerical integration algorithms are widely-used in scientific codes to implement assembly~\cite{quadrature1, quadrature2}.
 
-The complexity of FFC-generated kernels depends on the mathematical problem being solved.
+Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm.
+At the core of Firedrake~\cite{firedrake} lies PyOP2~\cite{pyop2ics}, which triggers compilation of such kernels using an available vendor compiler, and eventually manages parallel execution over the mesh.
+In our work, we have enhanced this execution model by adding an optimization stage prior to the generation of C code.
+The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used.
+Notable examples are~\cite{uflacs, fluidity, mcfc}, in which the kernel's loop nest and memory access pattern are reducible to that generated by Firedrake.
+
+The complexity of Firedrake-generated kernels depends on the mathematical problem being solved.
 In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few, small outer-product vector-operations.
@@
 
-One attractive possibility is to transform the loop nest into a sequence of calls to BLAS routines.
-There are several issues that make achieving this difficult and, potentially, not effective.
-The first problem concerns automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of the memory access pattern.
-Related to this problem is the fact that, in general, many sub-expressions should be pre-computed (i.e.
-computed once before the loop nest), which leads to using temporaries and, possibly, to working sets that do not fit the L1 cache (for $p>2$).
-In addition, using BLAS routines may impact data locality, since the same matrix can appear in multiple products, implying, at least, a certain degree of register spilling.
-Finally, it is known that BLAS routines perform far from peak performance when the dimension of the matrices is rather small~\cite{nek5000}.
-Hand-made implementations of the simple Helmholtz problem ($p=1,2,3,4$) as a sequence of BLAS ($dgemm$) routines have demonstrated that performance is worse than that shown in Section~\ref{sec:perf-results}.
-A comprehensive study concerning BLAS optimization is therefore left as further work.
-
 \section{Domain-Driven Code Transformations}
@@
 \small
-\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).
+\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).
 The padding, data alignment and $licm$ optimizations are applied.
@@
 \small
-\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).
+\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).
 The padding, data alignment, $licm$ and $split$ optimizations are applied.
@@
 \small
-\caption{Local assembly code generated by FFC for a Helmholtz problem (2D mesh, Lagrange $p=2$ elements).
+\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=2$ elements).
 The padding, data alignment, $licm$ and $op$-$vect$ optimizations are applied.
@@
 \end{tabulary}
-\caption{Overview of code transformations for FFC-generated assembly kernels.}
+\caption{Overview of code transformations for Firedrake-generated assembly kernels.}
 \label{table:code-transformations}
@@
 
-Our approach to solving partial differential equations is based on domain-specific languages and libraries.
-Fenics and Firedrake provide the high-level Unified Form Language (UFL) to write problems in a notation resembling mathematical equations (``forms'').
-In Fenics, this specification is translated by a form compiler (FFC) into C code, which is subsequently just-in-time compiled and executed on the target architecture.
-In Firedrake we adopt a completely different approach.
-We think that FFC should produce a high-level representation of a Finite Element assembly kernel, and should abstract from the specific details of the underlying architecture.
-We have modified FFC, therefore, to build Abstract Syntax Trees (AST), rather than C code.
-ASTs are then passed to PyOP2, which implements the core of Firedrake, where parallel execution over the unstructured mesh is managed.
-We have developed a compiler that, given an AST conformed to the PyOP2 specification (such as those produced by the new FFC), can apply the transformations described in Section~\ref{sec:code-transf} and, subsequently, generate C code.
-The C representation is eventually just-in-time compiled for one of the backends that are currently supported by PyOP2, namely OpenMP, MPI, OpenMP+MPI, CUDA, OpenCL.
-Our compiler is now fully integrated with PyOP2, so it supports any kind of problem expressible in UFL.
-
-The PyOP2 compiler has been thought to support a range of higher-level languages and applications that concern the resolution of partial differential equations; Firedrake, in terms of its components UFL and FFC, is one of them.
-We believe, however, that many of the discussed optimizations are generalizable and applicable to non-Firedrake kernels.
-For example, it is always the case that Finite Element Local Assembly using a quadrature strategy leads to a (more or less complex) loop nest and outer-product-like computations.
+Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations.
+This high-level specification is translated by the domain-specific Fenics Form Compiler~\cite{ffc} into an Abstract Syntax Tree representation of a Finite Element assembly kernel.
+ASTs are then passed to PyOP2, where parallel execution over the unstructured mesh is managed.
+Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it can manipulate ASTs introducing optimizations, and then it generates C code, which is eventually just-in-time compiled on the underlying architecture.
 
 The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}.
-The AST is initially inspected, looking for the presence of an iteration space and other domain-specific information provided by the upper layer.
-If the kernel is just a sequence of statement, then so-called inter-kernel vectorization, in which a non-affine loop over mesh elements is vectorized, can be applied.
-This might be useful in a Finite Volume context, where kernels often lack an iteration space representing the bulk of the computation.
-Although not implemented yet, Reguly et.
-al.
-already proved the profitability of this technique in the OP2 framework~\cite{reguly-inter-kernel}.
-If, instead, an iteration space is found and the underlying platform is a GPU, the compiler seeks fully-parallel loops (i.e.
-loops whose iterations can be executed in any order without need for synchronization) and hoist them out at the level of the ``wrapper'' (i.e.
-the function that iterates over mesh elements and invokes the kernel); this is aimed at optimizing register management on GPUs.
+Initially, an AST is inspected, looking for the presence of iteration spaces and other domain-specific information provided by the higher layer.
+If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied.
+This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{reguly-inter-kernel}.
+The second transformation step is applied if the backend is a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{quadrature-cuda}.
+Then, a sequence of optimization steps are executed.
+A cost model is queried to estimate the potential gain if the $split$ and $op$-$vect$ transformations are applied.
+Optimization passes must follow a specific order.
+Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays.
+Then, $split$ might take place, depending on the outcome of the cost model.
+The $op$-$vect$ optimization is introduced lastly.
+Note that the compiler can handle all possible corner cases: for example, if $op$-$vect$ is to be applied but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.
 
-If the backend is a CPU and the AST comes from Firedrake, then a sequence of optimization steps are executed.
-A cost model, currently under development, is queried to estimate the potential gain (or loss) coming from the $split$ and $op$-$vect$ transformations.
-The model suggests a split factor and a tile size.
-Firstly, the compiler applies $licm$, which introduces temporary arrays, and on top of this padding and data alignment are enforced.
-These three transformations are always applied, since they usually provide a significant performance improvement (this will be shown in Section~\ref{sec:perf-results}).
-Finally, $split$ and/or $op$-$vect$ take place.
+Our experience... unrolling/unroll and jam
 

### 2014-02-25 paper/bare_conf.tex
Re-factor lots of paragraphs

@@
 The solution is sought in each cell of the discretized domain by applying suitable numerical kernels.
-Iteration over mesh components is also expensive, due to the presence of indirect memory accesses.
 As the number of cells can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days.
-To address this issue, domain-specific languages (DSLs) have been developed.
+To address this problem, domain-specific languages (DSLs) have been developed.
 The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2ics}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{OP2-hydra}.
@@
 On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread.
-Together with the indirections problem, kernel optimization is one of the major concerns in unstructured mesh applications.
+Similar programming and execution models are adopted by~\cite{fenics, fluidity, lizst}.
+Kernel optimization is one of the major concerns in unstructured mesh applications.
 In this paper, we tackle this problem by proposing a domain-driven optimization strategy for a class of kernels used in Finite Element Methods.
@@
 With such small kernels, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation.
+Our study is conducted in the context of Firedrake, a system for solving PDEs through the Finite Element Method based on the OP2 abstraction~\cite{firedrake}.
 
-Since loop nests are affine, one possibility is to rely on polyhedral compilation for improving performance.
-There are, however, several limitations.
-The first problem is that a polyhedral compiler reasons on iteration spaces, whereas some of the transformations we propose rely on analysis of the data space.
-Our attempts at using PoCC's optimizations~\cite{PoCC}, especially register tiling, were unsuccessful.
-A domain-specific transformation of the iteration space, which is not expressible in the polyhedral model because of its non-affine nature, also showed a significant performance gain due to improved register locality.
+Optimization of assembly kernels is non-trivial.
+Polyhedral compilation could be used to improve performance; however, given the exceptionally small size of the kernels, effective register locality can be obtained only by an in-depth analysis of the data space, rather than reasoning on the iteration space (our attempts to use PoCC's optimizations~\cite{PoCC}, especially unrolling and register tiling, were unsuccessful).
+BLAS routines could be employed as well, although fairly complicated code manipulations would be required, for a performance gain potentially far from the expectations.
+The major issue concerns automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of the memory access pattern.
+Register locality would also be affected, since the same matrix could appear in a number of products.
+BLAS libraries are also known to perform distant from peak performance when the dimension of the matrices is small~\cite{nek5000}.
+Hand-made BLAS-based implementations of the Helmholtz assembly kernel (illustrated later), varying the polynomial order from $p=1$ to $p=4$, has showed that the performance is worse than that achieved with our optimization strategy.
 
-Assembly kernels could be turned into a sequence of calls to BLAS routines, although it is complicated to achieve and, potentially, ineffective.
-One issue concerns automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of the memory access pattern.
-In addition, register locality would be affected, since the same matrix can appear in a number of products.
-Finally, it is known that BLAS routines perform far from peak performance when the dimension of the matrices is small~\cite{nek5000}.
-Hand-made implementations of the simple Helmholtz assembly kernel as a sequence of BLAS $dgemm$ calls, for a range of polynomial orders ($p=1,2,3,4$), has showed that the achieved performance is worse than that exposed in Section~\ref{sec:perf-results}.
+%Also, tiling at the level of vector register, achievable through a non-affine transformation of the iteration space, has showed a significant performance improvement in a range of problems. .
 
 Given the constraints on polyhedral compilation and linear-algebra-specialized libraries, we address assembly optimization by studying a set of domain-specific code transformations, applicable to a wide class of problems.
-We have developed a domain-specific optimizing compiler and we have integrated it with Firedrake, a system for solving PDEs through the Finite Element Method~\cite{firedrake} based on the OP2 abstraction.
+We have developed an optimizing compiler and we have integrated it with Firedrake.
 This allows us to evaluate our code transformations in a range of real-world problems, varying parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the mesh type (2D, 3D, 3D hybrid structured-unstructured).
@@
 A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell of the discretized domain.
-Numerical integration algorithms are widely-used in scientific codes to implement assembly~\cite{quadrature1, quadrature2}.
+Numerical integration algorithms are widely-used in assembly codes to implement evaluation of stiffness matrices~\cite{quadrature1, quadrature2}.
 
 Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm.
-At the core of Firedrake~\cite{firedrake} lies PyOP2~\cite{pyop2ics}, which triggers compilation of such kernels using an available vendor compiler, and eventually manages parallel execution over the mesh.
+It then triggers compilation of such kernels using an available vendor compiler, and eventually manages parallel execution over the mesh.
 In our work, we have enhanced this execution model by adding an optimization stage prior to the generation of C code.
@@
 The complexity of Firedrake-generated kernels depends on the mathematical problem being solved.
-In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few, small outer-product vector-operations.
+In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products.
 An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$.
+The stiffness matrix in the code is called $A$.
 In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of $A$ can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
 Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse.
-Despite such a big space of possible assembly implementations, it is still possible to identify a domain-specific structure common to all kernels, which can be turned into effective code transformations and simd-vectorization.
+Note that in addition to a larger number of operations to compute the stiffness matrix, the Burgers case shows a register pressure higher than that in Helmholtz.
+Despite assembly kernels being problem-dependent, and so infinite is the space of codes that Firedrake can generate, it is still possible to identify common domain-specific traits, which can be exploited for effective code transformations and SIMD vectorization.
 
 The class of kernels we are considering has, in particular, some peculiarities.
-1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest; this is not true in general, since the unstructured mesh might have bended elements that require the Jacobian be re-computed every $i$ iteration.
-2) The stiffness matrix $A$ is always the result of an outer-product computation along the $j$ and $k$ dimensions.
-3) Memory accesses along the three loop dimensions are always 1-stride .4) The $j$ and $k$ loops are interchangable, whereas to permute $i$ it might be necessary to precompute $F$ values and to introduce temporaries.
-5) Basis functions arrays (denoted by $X$, $Y$, ...) are constants, and many sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
+1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest; this is not true in general, since the unstructured mesh might use bended elements that would require the Jacobian be re-computed every $i$ iteration.
+2) Memory accesses along the three loop dimensions are always 1-stride .3) The $j$ and $k$ loops are interchangeable, whereas permutation of $i$ is might be subjected to pre-computation of values and introduction of temporaries (e.g.
+the $F$ values in Burgers).
+4) So called basis functions arrays (denoted by $X$, $Y$, ...) are constants, and most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
 In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.
@@
 Register tiles along both inner dimensions can be determined as well.
-Our domain-driven strategy, based on the observation that the evaluation of the stiffness matrix can be reduced to a summation of outer products along the $j$ and $k$ dimensions, aims at optimizing register reuse, in terms of both minimization of load operations and achievement of larger tiles, over a canonical implementation.
+Based on the observation that the evaluation of the stiffness matrix can be reduced to a summation of outer products along the $j$ and $k$ dimensions, aims at optimizing register reuse, in terms of both minimization of load operations and achievement of larger tiles, over a canonical implementation.
 If we consider the code snippet in Figure~\ref{code:helmholtz-licm}, we can notice that the computation of $A$ is abstractly expressible as

### 2014-02-25 paper/bare_conf.tex
Minor changes

@@
 
-% Autovectorizable LICM -> show code snippet?
-\subsection{Auto-vectorizable Loop-invariant Code Motion}
-From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only.
-Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion ($licm$).
-Vendor compilers apply $licm$, although not in the systematic way we need in our assembly kernels.
-We want to overcome two deficiencies that both $icc$ and $gcc$ have.
-First, these compilers can identify sub-expressions that are invariant with respect to the innermost loop only.
-This is an issue for sub-expressions depending on $i$-$k$ only, which are not automatically lifted.
-Second, the hoisted code is scalar, i.e.
-it is not subjected to auto-vectorization.
-We work around these limitations by doing source-level $licm$.
-In addition, we pre-compute all values that an invariant sub-expression assumes along the fastest varying dimension.
-This is implemented by introducing temporary arrays and new loops in the nest.
-At the price of both extra memory for storing temporaries and reduced register locality, the gain is that lifted terms can be auto-vectorized, because part of an inner loop.
-Note that given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which can be otherwise significant.
-
 % Padding and Data Alignment
 \subsection{Padding and Data Alignment}
-Auto-vectorization of assembly code that computes $A$ can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length ($vl$).
+Auto-vectorization of assembly code computing the stiffness matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length ($vl$).
 Data alignment is enforced in two steps.
@@
 Then, matrices are padded by rounding the number of columns to the nearest multiple of $vl$.
-For example, assume the original size of a matrix is 3$\times$3, and the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use double-precision floating-point values (64 bits).
-Then, a padded matrix on this architecture will be 3$\times$4.
+For example, assume the original size of a matrix is 3$\times$3 and that the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values.
+Then, a padded matrix on this architecture will have size 3$\times$4.
 The compiler is explicitly informed about data alignment using a suitable pragma.
@@
 
+% Autovectorizable LICM -> show code snippet?
+\subsection{Auto-vectorizable Loop-invariant Code Motion}
+From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only.
+Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion.
+Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels.
+We want to overcome two deficiencies that both $icc$ and $gcc$ have.
+First, these compilers can identify sub-expressions that are invariant with respect to the innermost loop only.
+This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted.
+Second, the hoisted code is scalar, i.e.
+it is not subjected to auto-vectorization.
+We work around these limitations with source-level loop-invariant code motion.
+In particular, we pre-compute all values that an invariant sub-expression assumes along the fastest varying dimension.
+This is implemented by introducing a temporary array (per invariant sub-expression) and by adding a new loop to the nest.
+At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop.
+Note that given the short trip counts of our loops, it is important to achieve auto-vectorization of all hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant.
+
+Figure~\ref{code:helmholtz-licm} shows the Helmholtz assembly code after the application of loop-invariant code motion, padding, and data alignment.
+
 % Register allocation and register locality
-\subsection{Toward Improving Register Locality}
+\subsection{Optimizing Register Management}
 One notable problem assembly kernels are exposed to concerns register allocation and register locality.
-The critical situation occurs when loop trip counts and number of accessed matrices along the inner dimensions are such that available vector registers are under-utilized.
+The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high.
 Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management in order to maximize performance.
 Canonical optimizations, such as loop interchange, unroll, unroll-and-jam and register tiling are typically employed to deal with this problem.
-In our compiler we support these optimizations either by means of explicit source code transformation (interchange, register tiling, unroll-and-jam) or indirectly advising the compiler through standard pragmas (unroll).
-With register tiling, in particular, we slice the innermost loop into rectangular blocks of identical size (except the ``reminder'' block), and we apply unrolling to individual blocks to expose ILP.
-
-% remember: op vectorization PLUS unroll and jam !!
-\subsection{Outer-product Vectorization}
+In the compiler we have developed, we support these optimizations either by means of explicit source code transformation (interchange, unroll-and-jam) or indirectly advising the compiler through standard pragmas (unroll).
+In this context, register tiling is a critical optimization.
+A way of obtaining register tiles consists of slicing the innermost loop into rectangular blocks of identical size (except the ``reminder'' block), and applying unrolling to individual blocks in order to expose ILP.
+This is automated by some vendor compilers~\cite{comdex}.
 Register tiles along both inner dimensions can be determined as well.
-Based on the observation that the evaluation of the stiffness matrix can be reduced to a summation of outer products along the $j$ and $k$ dimensions, aims at optimizing register reuse, in terms of both minimization of load operations and achievement of larger tiles, over a canonical implementation.
-If we consider the code snippet in Figure~\ref{code:helmholtz-licm}, we can notice that the computation of $A$ is abstractly expressible as
+Based on the observation that the evaluation of the stiffness matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, we implement a domain-specific vector-register tiling strategy that leads to larger tiles, enhancing reuse.
+If we consider the code snippet in Figure~\ref{code:helmholtz-licm} (Helmholtz after loop-invariant code motion), we can notice that the computation of $A$ is abstractly expressible as
 \begin{equation}
@@
 where $B$ is the set of all matrices (or temporaries) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets.
-Without loss of generality, the presence of other constant or $i$-dependent variables can be momentarily neglected.
-Note that regardless of the specific input problem, the assembly computation is always reducible to this form.
-Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix with just 2 vector registers.
-Once the first four entries are evaluated, values in a register are shuffled and the same product is performed again.
-Standard compiler auto-vectorization ($gcc$ and $icc$), instead, would employ 5 registers, 4 of them target of a broadcast operation (i.e.
-``splatting'' a single value over the four locations of an AVX register) along the outer dimension, and 1 of a load operation along the inner dimension.
-In our outer-product vectorization ($op$-$vect$), the 4 broadcast accesses to the L1 cache are avoided by an initial load and by shuffling.
-More importantly, we avoid using $vl$-1 registers per each outer-loop-dependent variable (i.e.
+Without loss of generality, the presence of constants or other variables independent of both $j$ and $k$ can be momentarily neglected.
+Note that regardless of the specific input problem, the stiffness matrix computation is always reducible to this kind of form.
+Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix using just 2 vector registers (a 4$\times$4 tile), assuming $\vert B' \vert = \vert B'' \vert = 1$.
+Values in a register are shuffled each time a product is performed.
+Standard compiler auto-vectorization ($gcc$ and $icc$), instead, executes 4 broadcast operations (i.e.
+``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation, and would also need to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations if unroll-and-jam is found to be useful, with $f$ being the unroll-and-jam factor.
+
+%More importantly, we avoid using $vl$-1 registers per each outer-loop-dependent variable (i.e.
 all $x$ terms in Equation~\ref{outer-product}), with $vl$ being the vector length.
@@
 \begin{figure}[h]
+\normalsize
 \centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect.pdf}
@@
 
-The storage layout of $A$, however, is incorrect after the application of $op$-$vect$.
-We efficiently restore it with a sequence of vector shuffle following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed just once outside of the $ijk$ loop nest.
+The storage layout of $A$, however, is incorrect after the application of this ``outer-product vectorization'' ($op$-$vect$).
+We efficiently restore it with a sequence of vector shuffle following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the $ijk$ loop nest.
 The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.
 
-\begin{figure}[h]
+\begin{figure}[t]
 \centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}
 \label{fig:restore-layout}}
-\caption{Restoring the storage layout after $op$-$vect$.}
+\caption{Restoring the storage layout after $op$-$vect$. The figure shows how 4$\times$4 elements in the top-left block of the stiffness matrix $A$ can be brought to their correct positions.}
 \end{figure}
@@
 If the number of variables independent of the innermost-loop dimension, i.e.
-basis function matrices, temporaries introduced by $licm$ and constants, is larger than the available registers on the CPU, it is likely to obtain poor register reuse.
-For example, applying $licm$ to Burgers on a 3D mesh needs 33 temporaries for the $ijk$ loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers.
+basis function matrices, temporaries introduced by loop-invariant code motion and constants, is even close to the number of available CPU registers, it is likely to obtain poor register reuse.
+For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the $ijk$ loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers.
 One potential solution to this problem consists of suitably ``splitting'' the computation of the stiffness matrix $A$ into multiple sub-expressions; an example, for the simpler Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}.
@@
 Data alignment and padding relate to an AVX machine.
-In this specific case, the unroll-and-jam factor is 1.}
+The original size of the $j$-$k$ iteration space (i.e.
+before padding was applied) was 6$\times$6.
+In this example, the unroll-and-jam factor is 1.}
 \label{code:helmholtz-opvect}
@@
 This high-level specification is translated by the domain-specific Fenics Form Compiler~\cite{ffc} into an Abstract Syntax Tree representation of a Finite Element assembly kernel.
-ASTs are then passed to PyOP2, where parallel execution over the unstructured mesh is managed.
-Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it can manipulate ASTs introducing optimizations, and then it generates C code, which is eventually just-in-time compiled on the underlying architecture.
+ASTs are then passed to PyOP2~\cite{pyop2isc}, which lies at the core of Firedrake, where parallel execution over the unstructured mesh is managed.
+Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it receives FFC's ASTs as input, it introduces optimizations, and it generates C code as output, which is eventually just-in-time compiled on the underlying architecture.
 
@@
 
-Our experience... unrolling/unroll and jam
+Our experience... unrolling/unroll and jam/interchange
 

### 2014-02-26 paper/bare_conf.tex
Add stuff to the PyOP2 compiler section

@@
 Optimization of assembly kernels is non-trivial.
-Polyhedral compilation could be used to improve performance; however, given the exceptionally small size of the kernels, effective register locality can be obtained only by an in-depth analysis of the data space, rather than reasoning on the iteration space (our attempts to use PoCC's optimizations~\cite{PoCC}, especially unrolling and register tiling, were unsuccessful).
-BLAS routines could be employed as well, although fairly complicated code manipulations would be required, for a performance gain potentially far from the expectations.
-The major issue concerns automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of the memory access pattern.
-Register locality would also be affected, since the same matrix could appear in a number of products.
+Polyhedral compilation could be used to improve performance; however, given the exceptionally small size of the kernels, effective register locality can be obtained only through an in-depth analysis of the data space, rather than by reasoning on the iteration space (our attempts to use PoCC's optimizations~\cite{PoCC}, especially unrolling and register tiling, did not improve performance over $icc$ compilation).
+BLAS routines could be employed as well, although fairly complicated code manipulation would be required, for a performance gain potentially far from the expectation.
+The major issue would concern automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of control- and data-flow.
 BLAS libraries are also known to perform distant from peak performance when the dimension of the matrices is small~\cite{nek5000}.
+In addition, register locality would be affected, since the same matrix could appear in a number of products.
 Hand-made BLAS-based implementations of the Helmholtz assembly kernel (illustrated later), varying the polynomial order from $p=1$ to $p=4$, has showed that the performance is worse than that achieved with our optimization strategy.
@@
 Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations.
-This high-level specification is translated by the domain-specific Fenics Form Compiler~\cite{ffc} into an Abstract Syntax Tree representation of a Finite Element assembly kernel.
+This high-level specification is translated by the Fenics Form Compiler~\cite{ffc} into an Abstract Syntax Tree representation of a Finite Element assembly kernel.
 ASTs are then passed to PyOP2~\cite{pyop2isc}, which lies at the core of Firedrake, where parallel execution over the unstructured mesh is managed.
 Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it receives FFC's ASTs as input, it introduces optimizations, and it generates C code as output, which is eventually just-in-time compiled on the underlying architecture.
+
+TODO: say there is a very huge space of implementation opportunities; in this section, we say how to prune it (without autotuning, which would be prohibitive for our jit-compilation).
 
@@
 This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{reguly-inter-kernel}.
-The second transformation step is applied if the backend is a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{quadrature-cuda}.
+The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{quadrature-cuda}.
 Then, a sequence of optimization steps are executed.
-A cost model is queried to estimate the potential gain if the $split$ and $op$-$vect$ transformations are applied.
+A cost model is queried to estimate the potential gain if the $split$ and $op$-$vect$ transformations were applied.
 Optimization passes must follow a specific order.
@@
 
-Our experience... unrolling/unroll and jam/interchange
+Loop unroll and unroll-and-jam of outer loops are fundamental to expose ILP and data reuse, and so tuning critical parameters such as the unroll factor becomes of great importance.
+Autotuning might be used, although we avoid it to minimize the overhead of Firedrake's just-in-time compilation.
+It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for our assembly kernels, where the loop nest is affine, bounds are known at compile-time, and memory accesses are unit-stride, newest versions of a vendor compiler like $icc$ employ cost models capable of estimating close-to-optimal values for such parameters.
+We therefore leave the backend compiler in charge to select unroll and unroll-and-jam factors.
+This choice also simplifies the PyOP2 compiler's cost model.
+The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $icc$ compiler from applying any other unroll-based optimizations.
+
+All loops are interchangeable provided that temporaries are introduced if the nest is not perfect.
+It would be possible, therefore, to permute loops in order to improve data movement, although we discovered that a specific loop order often leads to the best run-time, for the storage layout we employ.
+The loop order $ijk$ is likely to maximize performance.
+Conceptually, this is motivated by the following considerations: 1) all of the accessed arrays depend on the $i$ loop, and having it in one of the inner positions would mean a significantly higher number of load operations, with a performance loss greater than the reduction of store operations into the stiffness matrix; 2) if loop-invariant code motion is applied, then the number of loads along the $j$ and $k$ dimensions becomes identical (this property derives from the nature of assembly), which implies the loop order $ijk$ is performance-wise identical to $ikj$.
+These observations, which allow us to fix a specific loop order, contribute to relieve the decisions the PyOP2 compiler has to make.
 

### 2014-03-05 paper/bare_conf.tex
General refactoring

@@
 An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 40, with a minimum value of 3, depending on the order of the method).
-With such small kernels, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation.
+With such small kernels, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.
 Our study is conducted in the context of Firedrake, a system for solving PDEs through the Finite Element Method based on the OP2 abstraction~\cite{firedrake}.
@@
 In addition, register locality would be affected, since the same matrix could appear in a number of products.
-Hand-made BLAS-based implementations of the Helmholtz assembly kernel (illustrated later), varying the polynomial order from $p=1$ to $p=4$, has showed that the performance is worse than that achieved with our optimization strategy.
+Hand-made BLAS-based implementations of the Helmholtz assembly kernel (illustrated later), varying the polynomial order from $p=1$ to $p=4$ and on a 3D mesh, have showed run-times worse than those achieved with our optimization strategy.
 
@@
 
-Performance results show that Firedrake-generated code for non-trivial assembly kernels, if not adequately optimized, is sub-optimal.
-A cost-model-driven sequence of source-to-source code transformations, aimed at improving register allocation and register data locality, can result in performance improvements up to 33$\%$ over ``softly-optimized'' code (i.e.
-where only basic transformations are performed, such as auto-vectorizable loop-invariant code motion, padding and data alignment), and up to 77$\%$ over the original kernels.
+Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal.
+Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 33$\%$ over ``softly-optimized'' code (i.e.
+where only basic transformations are performed, such as auto-vectorizable loop-invariant code motion, padding and data alignment), and up to 77$\%$ over original kernels.
 The contribution of this paper is therefore twofold:
@@
 $icc$) and research (e.g.
-those based on the polyhedral model) compilers.
+polyhedral) compilers.
 \item Automation of such code optimizations in Firedrake. This enables performing an in-depth performance evaluation of the proposed optimization strategy in real-world simulations.
@@
 \section{Background}
+\label{sec:background}
 %Here we discuss about the computational characteristics of computational science kernels.
@@
 ~~~~~~~~A[j][k] += ((Y[i][k]*Y[i][j]+\\
-~~~~~~~~~~~+((K[1]*X$\_$D10[i][k]+K[3]*X$\_$D01[i][k])*\\
-~~~~~~~~~~~*(K[1]*X$\_$D10[i][j]+K[3]*X$\_$D01[i][j]))+\\
-~~~~~~~~~~~+((K[0]*X$\_$D10[i][k]+K[2]*X$\_$D01[i][k])*\\
-~~~~~~~~~~~*(K[0]*X$\_$D10[i][j]+K[2]*X$\_$D01[i][j])))*\\
+~~~~~~~~~~~+((K1*X$\_$D10[i][k]+K3*X$\_$D01[i][k])*\\
+~~~~~~~~~~~*(K1*X$\_$D10[i][j]+K3*X$\_$D01[i][j]))+\\
+~~~~~~~~~~~+((K0*X$\_$D10[i][k]+K2*X$\_$D01[i][k])*\\
+~~~~~~~~~~~*(K0*X$\_$D10[i][j]+K2*X$\_$D01[i][j])))*\\
 ~~~~~~~~~~~*det*W3[i]);\\
@@
 ~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$12; r++) $\lbrace$\\
-~~~~~~F0 += (w[r][0]*X1$\_$D100[ip][r]);\\
+~~~~~~F0 += (w[r][0]*X1$\_$D100[i][r]);\\
 ~~~~~~//It follows 10 analogous statements (F1, F2, ...)\\
@@
 ~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$12; k++) \\
-~~~~~~~~A[j][k] += (..(K[5]*F9)+(K[8]*F10))*Y1[i][j])+\\
-~~~~~~~~~+(((K[0]*X1$\_$D100[i][k])+(K[3]*X1$\_$D010[i][k])+\\
-~~~~~~~~~+(K[6]*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
-~~~~~~~~~+(..((K[2]*X2$\_$D100[i][k])+...+(K[8]*X2$\_$D001[i][k])*\\
-~~~~~~~~~*(K[2]*X2$\_$D100[i][j])+...+(K[8]*X2$\_$D001[i][j]))..)+\\
+~~~~~~~~A[j][k] += (..(K5*F9)+(K8*F10))*Y1[i][j])+\\
+~~~~~~~~~+(((K0*X1$\_$D100[i][k])+(K3*X1$\_$D010[i][k])+\\
+~~~~~~~~~+(K6*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
+~~~~~~~~~+(..((K2*X2$\_$D100[i][k])+...+(K8*X2$\_$D001[i][k])*\\
+~~~~~~~~~*(K2*X2$\_$D100[i][j])+...+(K8*X2$\_$D001[i][j]))..)+\\
 ~~~~~~~~~+ $<$roughly a hundred of sum/muls go here$>$)..)*\\
@@
 
-Local assembly consists of evaluating so called element stiffness matrix and vector; in this work, we focus on computation of matrices, which is the costly part of the process.
-A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell of the discretized domain.
+Local assembly consists of evaluating so called element stiffness matrix and element stiffness vector; in this work, we focus on computation of matrices, which is the costly part of the process.
+A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell (or element) of the discretized domain.
 Numerical integration algorithms are widely-used in assembly codes to implement evaluation of stiffness matrices~\cite{quadrature1, quadrature2}.
@@
 Note that in addition to a larger number of operations to compute the stiffness matrix, the Burgers case shows a register pressure higher than that in Helmholtz.
-Despite assembly kernels being problem-dependent, and so infinite is the space of codes that Firedrake can generate, it is still possible to identify common domain-specific traits, which can be exploited for effective code transformations and SIMD vectorization.
+Despite assembly kernels being problem-dependent, meaning that infinite is the space of codes that Firedrake can generate, it is still possible to identify common domain-specific traits, which can be exploited for effective code transformations and SIMD vectorization.
 
 The class of kernels we are considering has, in particular, some peculiarities.
-1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest; this is not true in general, since the unstructured mesh might use bended elements that would require the Jacobian be re-computed every $i$ iteration.
-2) Memory accesses along the three loop dimensions are always 1-stride .3) The $j$ and $k$ loops are interchangeable, whereas permutation of $i$ is might be subjected to pre-computation of values and introduction of temporaries (e.g.
-the $F$ values in Burgers).
-4) So called basis functions arrays (denoted by $X$, $Y$, ...) are constants, and most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
+1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest.
+This is not true in general, since the unstructured mesh might use bended elements that would require the Jacobian be re-computed every $i$ iteration; 2) memory accesses along the three loop dimensions are always 1-stride; 3) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g.
+the $F$ values in Burgers) and introduction of temporary arrays; 4) the $j$ and $k$ loops iterate over the same iteration space; 5) so called basis functions arrays (denoted by $X$, $Y$, ...) are constants, and most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
 In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.
@@
 \KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
+~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
 ~~// K, det = Compute Jacobian (coords) \\
 ~~\\
-~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
-~~\KwSty{static const double} X$\_$D10[3][4] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
-~~\KwSty{static const double} X$\_$D01[3][4] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
+~~\KwSty{static const double} W3[3] ALIGN = $\lbrace$...$\rbrace$\\
+~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
+~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
 ~~\\
@@
 ~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
-~~~~~~LI$\_$0[r] = ((K[1]*X$\_$D10[ip][r])+(K[3]*X$\_$D01[ip][r]));\\
-~~~~~~LI$\_$1[r] = ((K[0]*X$\_$D10[ip][r])+(K[2]*X$\_$D01[ip][r]));\\
+~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
+~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
 ~~~~$\rbrace$\\
 ~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
+~~~~~~\KwSty{$\#$pragma vector aligned}\\
 ~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
@@
 ~~~~~~~~~~~~~~~~~~~~+LI$\_$1[k]*LI$\_$1[j])*det*W3[i]);\\
-~~$\rbrace$\\
-$\rbrace$
-\end{algorithm}
-
-\begin{algorithm}[t]
-\small
-\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).
-The padding, data alignment, $licm$ and $split$ optimizations are applied.
-Data alignment and padding relate to an AVX machine.
-In this specific case, the split factor is 2.}
-\label{code:helmholtz-split}
-\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
-~~// K, det = Compute Jacobian (coords) \\
-~~// Declaration of basis function matrices \\
-~~\\
-~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
-~~~~double LI$\_$0[4];\\
-~~~~double LI$\_$1[4];\\
-~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
-~~~~~~LI$\_$0[r] = ((K[1]*X$\_$D10[ip][r])+(K[3]*X$\_$D01[ip][r]));\\
-~~~~~~LI$\_$1[r] = ((K[0]*X$\_$D10[ip][r])+(K[2]*X$\_$D01[ip][r]));\\
-~~~~$\rbrace$\\
-~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
-~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
-~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
-~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
-~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
-~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
 ~~$\rbrace$\\
@@
 At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop.
-Note that given the short trip counts of our loops, it is important to achieve auto-vectorization of all hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant.
+Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant.
+It is also worth noting that, in some problems, invariant sub-expressions along $j$ are identical to those along $k$ (e.g.
+in Helmholtz).
+In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.
 
@@
 The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high.
-Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management in order to maximize performance.
+Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management.
 Canonical optimizations, such as loop interchange, unroll, unroll-and-jam and register tiling are typically employed to deal with this problem.
@@
 Standard compiler auto-vectorization ($gcc$ and $icc$), instead, executes 4 broadcast operations (i.e.
-``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation, and would also need to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations if unroll-and-jam is found to be useful, with $f$ being the unroll-and-jam factor.
+``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation, and would also need to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.
 
@@
 The storage layout of $A$, however, is incorrect after the application of this ``outer-product vectorization'' ($op$-$vect$).
-We efficiently restore it with a sequence of vector shuffle following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the $ijk$ loop nest.
+We efficiently restore it with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the $ijk$ loop nest.
 The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.
@@
 We want to exploit the small size of the iteration space and the structure of the computation to optimize register management.
-
-% Splitting
-\subsection{Assembly splitting}
-In complex kernels, like Burgers in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
-If the number of variables independent of the innermost-loop dimension, i.e.
-basis function matrices, temporaries introduced by loop-invariant code motion and constants, is even close to the number of available CPU registers, it is likely to obtain poor register reuse.
-For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the $ijk$ loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers.
-One potential solution to this problem consists of suitably ``splitting'' the computation of the stiffness matrix $A$ into multiple sub-expressions; an example, for the simpler Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}.
-Splitting an expression has, in general, several drawbacks.
-Firstly, it increases the number of accesses to $A$ proportially to the ``split factor'', which is the number of sub-expressions produced.
-Secondly, depending on how the split is executed, it can lead to redundant computation (e.g.
-the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}).
-Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same matrix could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
-Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.
 
@@
 
+% Splitting
+\subsection{Assembly splitting}
+In complex kernels, like Burgers in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
+If the number of variables independent of the innermost-loop dimension, i.e.
+basis function matrices, temporaries introduced by loop-invariant code motion and constants, is even close to the number of available CPU registers, it is likely to obtain poor register reuse.
+For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the $ijk$ loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers.
+One potential solution to this problem consists of suitably ``splitting'' the computation of the stiffness matrix $A$ into multiple sub-expressions; an example, for the simpler Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}.
+Splitting an expression has, in general, several drawbacks.
+Firstly, it increases the number of accesses to $A$ proportially to the ``split factor'', which is the number of sub-expressions produced.
+Secondly, depending on how the split is executed, it can lead to redundant computation (e.g.
+the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}).
+Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same matrix could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
+Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.
+
+\begin{algorithm}[t]
+\small
+\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).
+The padding, data alignment, $licm$ and $split$ optimizations are applied.
+Data alignment and padding relate to an AVX machine.
+In this specific case, the split factor is 2.}
+\label{code:helmholtz-split}
+\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
+~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
+~~// K, det = Compute Jacobian (coords) \\
+~~// Declaration of basis function matrices \\
+~~\\
+~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
+~~~~double LI$\_$0[4];\\
+~~~~double LI$\_$1[4];\\
+~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
+~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
+~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
+~~~~$\rbrace$\\
+~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
+~~~~~~\KwSty{$\#$pragma vector aligned}\\
+~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
+~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
+~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
+~~~~~~\KwSty{$\#$pragma vector aligned}\\
+~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
+~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
+~~$\rbrace$\\
+$\rbrace$
+\end{algorithm}
+
 Table~\ref{table:code-transformations} summarizes the code transformations described so far.
@@
 Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it receives FFC's ASTs as input, it introduces optimizations, and it generates C code as output, which is eventually just-in-time compiled on the underlying architecture.
-
-TODO: say there is a very huge space of implementation opportunities; in this section, we say how to prune it (without autotuning, which would be prohibitive for our jit-compilation).
+Because of the large number of (possibly parametric) transformations, we need a mechanism to select the most suitable optimization strategy for a given problem.
+Auto-tuning is not an attractive approach since it would be performed at run-time when kernel code is available.
+The optimization strategy is described in the following, along with an overview of our compiler.
 
@@
 The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{quadrature-cuda}.
-Then, a sequence of optimization steps are executed.
-A cost model is queried to estimate the potential gain if the $split$ and $op$-$vect$ transformations were applied.
-Optimization passes must follow a specific order.
+Then, an ordered sequence of optimization steps are executed.
 Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays.
-Then, $split$ might take place, depending on the outcome of the cost model.
-The $op$-$vect$ optimization is introduced lastly.
-Note that the compiler can handle all possible corner cases: for example, if $op$-$vect$ is to be applied but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.
+Based on a cost model, the $split$, and subsequently the $op$-$vect$, transformations may be introduced.
+The compiler handles any possible corner cases: for example, if $op$-$vect$ is to be applied but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.
 
@@
 We therefore leave the backend compiler in charge to select unroll and unroll-and-jam factors.
-This choice also simplifies the PyOP2 compiler's cost model.
+This choice also simplifies the compiler's cost model.
 The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $icc$ compiler from applying any other unroll-based optimizations.
@@
 All loops are interchangeable provided that temporaries are introduced if the nest is not perfect.
-It would be possible, therefore, to permute loops in order to improve data movement, although we discovered that a specific loop order often leads to the best run-time, for the storage layout we employ.
-The loop order $ijk$ is likely to maximize performance.
-Conceptually, this is motivated by the following considerations: 1) all of the accessed arrays depend on the $i$ loop, and having it in one of the inner positions would mean a significantly higher number of load operations, with a performance loss greater than the reduction of store operations into the stiffness matrix; 2) if loop-invariant code motion is applied, then the number of loads along the $j$ and $k$ dimensions becomes identical (this property derives from the nature of assembly), which implies the loop order $ijk$ is performance-wise identical to $ikj$.
-These observations, which allow us to fix a specific loop order, contribute to relieve the decisions the PyOP2 compiler has to make.
+For the employed storage layout, the loop permutations $ijk$ and $ikj$ are likely to maximize performance.
+Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration.
+Our experiments showed that the performance loss is greater than the gain due to accumulating increments in a register along the $i$ loop.
+The choice between $ijk$ and $ikj$ depends on the number of load instructions that can be hoisted out of the innermost dimension.
+Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers are availabe to carry out the computation of the stiffness matrix.
+
+%TODO: NOT TRUE that kj == jk. You can have less matrices depending on j than those depending on k, meaning that less loads are needed along the j dimension.
 
@@
 \end{algorithm}
+
+Mention heuristic used by our compiler for splitting.. TODO
 

### 2014-03-07 paper/bare_conf.tex
Start perf evaluation section. Add francis' table

@@
 % TODO: ok?
+
+\usepackage{tabularx}
 \usepackage{tabulary}
@@
 Optimization of assembly kernels is non-trivial.
-Polyhedral compilation could be used to improve performance; however, given the exceptionally small size of the kernels, effective register locality can be obtained only through an in-depth analysis of the data space, rather than by reasoning on the iteration space (our attempts to use PoCC's optimizations~\cite{PoCC}, especially unrolling and register tiling, did not improve performance over $icc$ compilation).
-BLAS routines could be employed as well, although fairly complicated code manipulation would be required, for a performance gain potentially far from the expectation.
-The major issue would concern automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of control- and data-flow.
+Polyhedral compilation could be used to improve performance; however, given the structure and the exceptionally small size of the kernels, effective SIMD vectorization can be obtained only through an in-depth analysis of both data and iteration space, which is not supported by state-of-the-art polyhedral compilers.
+BLAS routines could be employed as well, although fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplications.
 BLAS libraries are also known to perform distant from peak performance when the dimension of the matrices is small~\cite{nek5000}.
-In addition, register locality would be affected, since the same matrix could appear in a number of products.
-Hand-made BLAS-based implementations of the Helmholtz assembly kernel (illustrated later), varying the polynomial order from $p=1$ to $p=4$ and on a 3D mesh, have showed run-times worse than those achieved with our optimization strategy.
+As detailed in Section~\ref{sec:perf-results}, hand-made BLAS implementations of the Helmholtz assembly kernel (illustrated later) have run-times worse than those achieved with our optimization strategy.
+
+%(our attempts to use PoCC's optimizations~\cite{PoCC}, especially unrolling and register tiling, did not improve performance over $icc$ compilation)
 
@@
 This choice also simplifies the compiler's cost model.
-The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $icc$ compiler from applying any other unroll-based optimizations.
+The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $icc$ compiler from applying loop unrolling, even if specific pragmas are added.
 
@@
 Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers are availabe to carry out the computation of the stiffness matrix.
-
-%TODO: NOT TRUE that kj == jk. You can have less matrices depending on j than those depending on k, meaning that less loads are needed along the j dimension.
 
@@
 // Optimize ast based on plan \\
-ast.apply\_licm() \\
+ast.licm() \\
 ast.padding() \\
 ast.data\_align() \\
+\If{plan.permute}{
+  ast.permute\_assembly\_loops()
+}
 \If{plan.sz\_split}{
@@
 \If{plan.uaj\_factor}{
-  ast.vr\_tile(plan.uaj\_factor)
+  ast.op\_vect(plan.uaj\_factor)
 }
@@
 
-\clearpage
-
 \section{Performance Evaluation}
 \label{sec:perf-results}
-%Performance evaluation.
-We show all of the results got so far and we explain them.
-We demonstrate the claim that a compiler-driven approach is necessary to obtain maximum performance.
-Also, our results prove that current vendor compilers still have notable limitations, despite optimizing loop nests that are 1) affine, 2) have no stencil, and 3) ?
+
+Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Phi.
+The \texttt{icc 2013} compiler was used, with optimization level \texttt{-O2} and with auto-vectorization enabled (\texttt{-xAVX} on the Sandy Bridge, and \texttt{TODO} on the Phi).
+Other optimization levels performed, in general, slightly worse than \texttt{-O2}.
+Our code transformations were evaluated in three real-world problems expressible with Firedrake:
+\begin{itemize}
+\item Helmholtz
+\item Advection-Diffusion
+\item Burgers
+\end{itemize}
+The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}.
+For Advection-Diffusion the ``Diffusion'' equation is considered, which uses the same differential operators as Helmholtz.
+In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the stiffness matrix $A$.
+Burgers is a time-dependent problem (i.e.
+the assembly is recalculated every time step based on the result of previous iterations) employing differential operators different than those of Helmholtz and Advection-Diffusion.
+The assembly code, partly exposed in Figure~\ref{code:burgers}, is therefore fairly different, with a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) employed.
+
+These problems were studied varying both the shape of mesh elements and the polynomial order $p$ of the method.
+Intuitively, the larger the element shape, the bigger is the iteration space.
+Triangles (2D), tetrahedron (3D), and prisms (3D) were tested.
+For instance, in the case of Helmholtz with $p=1$, the size of the $j$ and $k$ loops for the three kind of elements is, respectively, $3$, $4$, and $6$.
+Moving from 2D to 3D also increases the number of basis function arrays, since conceptually the behaviour of the equation has to be approximated also along the $z$ dimension.
+On the other hand, the polynomial order affects only the problem size (the three loops $i$, $j$, and $k$, and, as a consequence, the size of $X$ and $Y$ arrays).
+A range of polynomial orders, from $p=1$ to $p=4$, are tested; higher polynomial orders are excluded from the study because of current Firedrake limitations.
+In such a large space of problems, the size of the stiffness matrix rarely excees 20$\times$20, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
+
+For the Helmholtz 3D problem, manual implementations based on MKL BLAS were tested on Sandy Bridge.
+This particular kernel can be easily reduced to a sequence of four matrix-matrix multiplications that can be computed via calls to BLAS \texttt{dgemm}.
+In the case of $p=4$, where the stiffness matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only $licm$, data alignment and padding were used.
+As anticipated, both extraction of matrix-matrix multiplications from analysis of the kernel's AST and invasive re-design of the Fenics Form Compiler to explicitly expose these operations will be addressed in further work.
+However, these experiments justify that there's a set of problems for which turning to BLAS is not beneficial in terms of performance.
+It is possible that employing BLAS is useful for particularly large problems, for instance Burgers on 3D meshes with $p \geq 3$, although the loss in data locality due to re-loading matrices appearing in multiple products might inhibit the performance gain.
+
+Table~\ref{table:perf-results}
+%In addition, register locality would be affected, since the same matrix could appear in a number of products.
+
+%The problem is time-dependent, so , which intuitively increaes the computation cost.
+
+%Also, our results prove that current vendor compilers still have notable limitations, despite optimizing loop nests that are 1) affine, 2) have no stencil, and 3) ?
+
+%\begin{table}[h]
+%\begin{tabulary}{1.0\columnwidth}{|C|C|}
+%\hline
+%Problem & shape & p & original & $licm$ & $licm$ \\\hline\hline
+%Auto-vectorizable Loop-invariant code motion ($licm$) &   \\ \hline
+%Padding &  \\ \hline
+%Data Alignment & \\ \hline
+%Loop interchange      & loops  \\ \hline
+%Loop unrolling  & unroll factor \\ \hline
+%Register tiling & tile size \\ \hline
+%Outer-product vectorization ($op$-$vect$) & tile size \\ \hline
+%Assembly splitting ($split$) & split point, split factor \\ \hline
+%\end{tabulary}
+%\caption{.}
+%\label{table:perf-results}
+%\end{table}
+
+%\newcommand{\massmatrixtwodresultsnorms}{
+%\begin{tabularx}{1.0\textwidth}{c|XX|XX|XX|XX|}
+%\cline{2-9}
+%& \multicolumn{2}{c|}{$n_f=1$} & \multicolumn{2}{c|}{$n_f=2$} & \multicolumn{2}{c|}{$n_f=3$} & \multicolumn{2}{c|}{$n_f=4$} \\
+%\cline{2-9}
+%& A & Tensor& A & Tensor& A & Tensor& A & Tensor\\
+%\hline
+%\multicolumn{1}{|c|}{$p=1,q=1$} & $6.89e{-}17$ & $9.30e{-}17$ & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+%\multicolumn{1}{|c|}{$p=1,q=2$} & $6.80e{-}17$ & $1.56e{-}16$ & $1.22e{-}16$ & $7.14e{-}17$ & $1.17e{-}17$ & $1.27e{-}17$ & $8.05e{-}18$ & $\mathbf{9.77e{-}5}$ \\
+%\multicolumn{1}{|c|}{$p=1,q=3$} & $8.16e{-}17$ & $8.94e{-}17$ & $6.21e{-}17$ & $\mathbf{6.77e{-}5}$ & $2.44e{-}17$ & $\mathbf{7.69e{-}4}$ & $8.61e{-}17$ & $\mathbf{5.02e{-}4}$ \\
+%\multicolumn{1}{|c|}{$p=1,q=4$} & $3.70e{-}16$ & $\mathbf{1.83e{-}4}$ & $7.12e{-}16$ & $\mathbf{4.11e{-}4}$ & $2.65e{-}16$ & $\mathbf{5.42e{-}4}$ & $5.03e{-}16$ & $\mathbf{4.36e{-}4}$ \\
+%\hline
+%\multicolumn{1}{|c|}{$p=2,q=1$} & $2.42e{-}16$ & $2.34e{-}16$ & $1.18e{-}16$ & $3.42e{-}17$ & $2.78e{-}17$ & $\mathbf{1.30e{-}4}$ & $1.22e{-}16$ & $\mathbf{3.06e{-}4}$ \\
+%\multicolumn{1}{|c|}{$p=2,q=2$} & $4.50e{-}16$ & $1.47e{-}16$ & $7.47e{-}17$ & $\mathbf{7.04e{-}5}$ & $4.05e{-}16$ & $\mathbf{6.47e{-}4}$ & - & - \\
+%\multicolumn{1}{|c|}{$p=2,q=3$} & $1.85e{-}16$ & $\mathbf{1.65e{-}4}$ & $7.76e{-}16$ & $\mathbf{5.57e{-}4}$ & $1.88e{-}15$ & $\mathbf{1.47e{-}3}$ & - & - \\
+%\multicolumn{1}{|c|}{$p=2,q=4$} & $2.22e{-}15$ & $\mathbf{2.90e{-}4}$ & $4.01e{-}15$ & $\mathbf{1.42e{-}3}$ & - & - & - & - \\
+%\hline
+%\multicolumn{1}{|c|}{$p=3,q=1$} & $6.54e{-}17$ & $5.96e{-}17$ & $8.13e{-}17$ & $\mathbf{9.70e{-}5}$ & - & - & - & - \\
+%\multicolumn{1}{|c|}{$p=3,q=2$} & $9.43e{-}17$ & $1.26e{-}16$ & $8.99e{-}16$ & $\mathbf{5.23e{-}4}$ & - & - & - & - \\
+%\multicolumn{1}{|c|}{$p=3,q=3$} & $2.27e{-}16$ & $\mathbf{2.86e{-}4}$ & $5.77e{-}15$ & $\mathbf{1.81e{-}3}$ & - & - & - & - \\
+%\multicolumn{1}{|c|}{$p=3,q=4$} & $2.45e{-}15$ & $\mathbf{5.31e{-}4}$ & $1.62e{-}14$ & $\mathbf{2.54e{-}3}$ & - & - & - & - \\
+%\hline
+%\end{tabularx}
+%}
+%\begin{figure*}
+%\massmatrixtwodresultsnorms
+%\end{figure*}
+
 \begin{figure}[h]

### 2014-03-10 paper/bare_conf.tex
Add cost model description. Change cost model algo. Minor refactoring.

@@
 However, we do not consider the latter case.
-An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 40, with a minimum value of 3, depending on the order of the method).
+An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 30, with a minimum value of 3, depending on the order of the method).
 With such small kernels, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.
@@
 
-\section{Domain-Driven Code Transformations}
+\section{Code Transformations}
 \label{sec:code-transf}
@@
 % Autovectorizable LICM -> show code snippet?
-\subsection{Auto-vectorizable Loop-invariant Code Motion}
+\subsection{Domain-driven Loop-invariant Code Motion}
 From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only.
@@
 % Register allocation and register locality
-\subsection{Optimizing Register Management}
+\subsection{Domain-drive Optimization of Register Management}
 One notable problem assembly kernels are exposed to concerns register allocation and register locality.
@@
 
-Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations.
-This high-level specification is translated by the Fenics Form Compiler~\cite{ffc} into an Abstract Syntax Tree representation of a Finite Element assembly kernel.
-ASTs are then passed to PyOP2~\cite{pyop2isc}, which lies at the core of Firedrake, where parallel execution over the unstructured mesh is managed.
-Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it receives FFC's ASTs as input, it introduces optimizations, and it generates C code as output, which is eventually just-in-time compiled on the underlying architecture.
-Because of the large number of (possibly parametric) transformations, we need a mechanism to select the most suitable optimization strategy for a given problem.
-Auto-tuning is not an attractive approach since it would be performed at run-time when kernel code is available.
-The optimization strategy is described in the following, along with an overview of our compiler.
-
-The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}.
-Initially, an AST is inspected, looking for the presence of iteration spaces and other domain-specific information provided by the higher layer.
-If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied.
-This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{reguly-inter-kernel}.
-The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{quadrature-cuda}.
-Then, an ordered sequence of optimization steps are executed.
-Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays.
-Based on a cost model, the $split$, and subsequently the $op$-$vect$, transformations may be introduced.
-The compiler handles any possible corner cases: for example, if $op$-$vect$ is to be applied but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.
-
-Loop unroll and unroll-and-jam of outer loops are fundamental to expose ILP and data reuse, and so tuning critical parameters such as the unroll factor becomes of great importance.
-Autotuning might be used, although we avoid it to minimize the overhead of Firedrake's just-in-time compilation.
-It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for our assembly kernels, where the loop nest is affine, bounds are known at compile-time, and memory accesses are unit-stride, newest versions of a vendor compiler like $icc$ employ cost models capable of estimating close-to-optimal values for such parameters.
-We therefore leave the backend compiler in charge to select unroll and unroll-and-jam factors.
-This choice also simplifies the compiler's cost model.
-The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $icc$ compiler from applying loop unrolling, even if specific pragmas are added.
-
-All loops are interchangeable provided that temporaries are introduced if the nest is not perfect.
-For the employed storage layout, the loop permutations $ijk$ and $ikj$ are likely to maximize performance.
-Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration.
-Our experiments showed that the performance loss is greater than the gain due to accumulating increments in a register along the $i$ loop.
-The choice between $ijk$ and $ikj$ depends on the number of load instructions that can be hoisted out of the innermost dimension.
-Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers are availabe to carry out the computation of the stiffness matrix.
-
 \begin{algorithm}[t]
@@
 
-The cost model ..... TODO
+Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations.
+This high-level specification is translated by the Fenics Form Compiler~\cite{ffc} into an Abstract Syntax Tree representation of a Finite Element assembly kernel.
+ASTs are then passed to PyOP2~\cite{pyop2isc}, which lies at the core of Firedrake, where parallel execution over the unstructured mesh is managed.
+Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it receives FFC's ASTs as input, it introduces optimizations, and it generates C code as output, which is eventually just-in-time compiled on the underlying architecture.
+Because of the large number of (possibly parametric) transformations, we need a mechanism to select the most suitable optimization strategy for a given problem.
+Auto-tuning is not an attractive approach since it would be performed at run-time when kernel code is available.
+The optimization strategy is described in the following, along with an overview of our compiler.
+
+The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}.
+Initially, an AST is inspected, looking for the presence of iteration spaces and other domain-specific information provided by the higher layer.
+If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied.
+This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{reguly-inter-kernel}.
+The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{quadrature-cuda}.
+Then, an ordered sequence of optimization steps are executed.
+Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays.
+Based on a cost model, the $split$ and $op$-$vect$ transformations may be introduced; their implementation is based on analysis and trasformation of the AST.
+The compiler handles any possible corner cases: for example, if $op$-$vect$ is to be applied but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.
+
+The cost model is shown in Figure~\ref{algo:applyCostModel}.
+It takes into account the number of available vector registers ($n\_regs$) and the number of variables iterating along the $j$ and $k$ dimensions ($n\_consts$ for independent variables, $n\_outer\_arrays$ for $j$ variables, and $n\_inner\_arrays$ for $k$ variables, assuming the $ijk$ loop order) to estimate unroll-and-jam and split factors when, respectively, $op$-$vect$ and $split$ are used (if a factor is 0, then the corresponding transformation is not applied).
+The $split$ transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 2).
+A split factor of $n$ means that the assembly expression should be ``cut'' into $n$ sub-expressions.
+Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size.
+Finally, the unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of ``free'' registers, i.e.
+those not used for keeping hoisted terms (line 9-11).
+
+Loop unroll and unroll-and-jam of outer loops are fundamental to expose ILP and data reuse, and so tuning critical parameters such as the unroll factor becomes of great importance.
+Autotuning might be used, although we avoid it to minimize the overhead of Firedrake's just-in-time compilation.
+It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for our assembly kernels, where the loop nest is affine, bounds are known at compile-time, and memory accesses are unit-stride, newest versions of a vendor compiler like $icc$ employ cost models capable of estimating close-to-optimal values for such parameters.
+We therefore leave the backend compiler in charge to select unroll and unroll-and-jam factors.
+This choice also simplifies the compiler's cost model.
+The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $icc$ compiler from applying unroll-based optimizations, even if specific pragmas are added.
+
+All loops are interchangeable provided that temporaries are introduced if the nest is not perfect.
+For the employed storage layout, the loop permutations $ijk$ and $ikj$ are likely to maximize performance.
+Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration.
+Our experiments showed that the performance loss is greater than the gain due to accumulating increments in a register along the $i$ loop.
+The choice between $ijk$ and $ikj$ depends on the number of load instructions that can be hoisted out of the innermost dimension.
+Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers are availabe to carry out the computation of the stiffness matrix.
 
 \begin{algorithm}[t]
-\caption{Procedure to estimate the most suitable unroll-and-jam factor and/or split size.}
+\caption{The cost model is employed by the compiler to estimate the most suitable unroll-and-jam (when $op$-$vect$ is used) and split factors, avoiding the overhead of auto-tuning.}
 \label{algo:applyCostModel}
-  \textbf{Algorithm: ApplyCostModel}\\
-  \KwIn{n\_inner\_arrays, n\_regs}
-  \KwOut{uaj\_factor, sz\_split}
-
-%// roundup(A, B) is a function that rounds A up to the nearest multiple of B \\
-
-sz\_split = 0 \\
+  \textbf{Cost Model}\\
+  \KwIn{n\_outer\_arrays, n\_inner\_arrays, n\_consts, n\_regs}
+  \KwOut{uaj\_factor, split\_factor}
+
+n\_outer\_regs = n\_regs / 2 \\
+split\_factor = 0 \\ 
 // Compute spltting factor \\
-\While{n\_inner\_arrays $>$ n\_regs}{
-  n\_inner\_arrays = n\_inner\_arrays / 2 \\
-  sz\_split = sz\_split + 1 
+\While{n\_outer\_arrays $>$ n\_outer\_regs}{
+  n\_outer\_arrays = n\_outer\_arrays / 2 \\
+  split\_factor = split\_factor + 1 
 }
 
-n\_inner\_regs = n\_regs / 2 \\
-\If{n\_inner\_arrays $>$ n\_inner\_regs \KwSty{or} sz\_split $>$ 0}{
-  // Rely on autovectorization, as there might be not enough registers to improve performance by tiling \\
-  \Return{$<$sz\_split, 0$>$}
-}
-
-// Compute unroll-and-jam factor for vector-register tiling \\
-n\_regs\_avail = n\_regs - n\_inner\_arrays \\
+// Compute unroll-and-jam factor for $op$-$vect$ \\
+n\_regs\_avail = n\_regs - (n\_outer\_arrays + n\_consts)\\
 uaj\_factor = $\lceil$n\_reg\_avail / n\_inner\_arrays$\rceil$ \\
-\Return{$<$sz\_split, uaj\_factor$>$}
+\Return{$<$split\_factor, uaj\_factor$>$}
 \end{algorithm}
-
-Mention heuristic used by our compiler for splitting.. TODO
 
@@
 
-%\newcommand{\massmatrixtwodresultsnorms}{
-%\begin{tabularx}{1.0\textwidth}{c|XX|XX|XX|XX|}
-%\cline{2-9}
-%& \multicolumn{2}{c|}{$n_f=1$} & \multicolumn{2}{c|}{$n_f=2$} & \multicolumn{2}{c|}{$n_f=3$} & \multicolumn{2}{c|}{$n_f=4$} \\
-%\cline{2-9}
-%& A & Tensor& A & Tensor& A & Tensor& A & Tensor\\
-%\hline
-%\multicolumn{1}{|c|}{$p=1,q=1$} & $6.89e{-}17$ & $9.30e{-}17$ & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-%\multicolumn{1}{|c|}{$p=1,q=2$} & $6.80e{-}17$ & $1.56e{-}16$ & $1.22e{-}16$ & $7.14e{-}17$ & $1.17e{-}17$ & $1.27e{-}17$ & $8.05e{-}18$ & $\mathbf{9.77e{-}5}$ \\
-%\multicolumn{1}{|c|}{$p=1,q=3$} & $8.16e{-}17$ & $8.94e{-}17$ & $6.21e{-}17$ & $\mathbf{6.77e{-}5}$ & $2.44e{-}17$ & $\mathbf{7.69e{-}4}$ & $8.61e{-}17$ & $\mathbf{5.02e{-}4}$ \\
-%\multicolumn{1}{|c|}{$p=1,q=4$} & $3.70e{-}16$ & $\mathbf{1.83e{-}4}$ & $7.12e{-}16$ & $\mathbf{4.11e{-}4}$ & $2.65e{-}16$ & $\mathbf{5.42e{-}4}$ & $5.03e{-}16$ & $\mathbf{4.36e{-}4}$ \\
-%\hline
-%\multicolumn{1}{|c|}{$p=2,q=1$} & $2.42e{-}16$ & $2.34e{-}16$ & $1.18e{-}16$ & $3.42e{-}17$ & $2.78e{-}17$ & $\mathbf{1.30e{-}4}$ & $1.22e{-}16$ & $\mathbf{3.06e{-}4}$ \\
-%\multicolumn{1}{|c|}{$p=2,q=2$} & $4.50e{-}16$ & $1.47e{-}16$ & $7.47e{-}17$ & $\mathbf{7.04e{-}5}$ & $4.05e{-}16$ & $\mathbf{6.47e{-}4}$ & - & - \\
-%\multicolumn{1}{|c|}{$p=2,q=3$} & $1.85e{-}16$ & $\mathbf{1.65e{-}4}$ & $7.76e{-}16$ & $\mathbf{5.57e{-}4}$ & $1.88e{-}15$ & $\mathbf{1.47e{-}3}$ & - & - \\
-%\multicolumn{1}{|c|}{$p=2,q=4$} & $2.22e{-}15$ & $\mathbf{2.90e{-}4}$ & $4.01e{-}15$ & $\mathbf{1.42e{-}3}$ & - & - & - & - \\
-%\hline
-%\multicolumn{1}{|c|}{$p=3,q=1$} & $6.54e{-}17$ & $5.96e{-}17$ & $8.13e{-}17$ & $\mathbf{9.70e{-}5}$ & - & - & - & - \\
-%\multicolumn{1}{|c|}{$p=3,q=2$} & $9.43e{-}17$ & $1.26e{-}16$ & $8.99e{-}16$ & $\mathbf{5.23e{-}4}$ & - & - & - & - \\
-%\multicolumn{1}{|c|}{$p=3,q=3$} & $2.27e{-}16$ & $\mathbf{2.86e{-}4}$ & $5.77e{-}15$ & $\mathbf{1.81e{-}3}$ & - & - & - & - \\
-%\multicolumn{1}{|c|}{$p=3,q=4$} & $2.45e{-}15$ & $\mathbf{5.31e{-}4}$ & $1.62e{-}14$ & $\mathbf{2.54e{-}3}$ & - & - & - & - \\
-%\hline
-%\end{tabularx}
-%}
-%\begin{figure*}
-%\massmatrixtwodresultsnorms
-%\end{figure*}
+\newcommand{\massmatrixtwodresultsnorms}{
+\begin{tabularx}{1.0\textwidth}{c|XX|XX|XX|XX|}
+\cline{2-9}
+& \multicolumn{2}{c|}{$n_f=1$} & \multicolumn{2}{c|}{$n_f=2$} & \multicolumn{2}{c|}{$n_f=3$} & \multicolumn{2}{c|}{$n_f=4$} \\
+\cline{2-9}
+& A & Tensor& A & Tensor& A & Tensor& A & Tensor\\
+\hline
+\multicolumn{1}{|c|}{$p=1,q=1$} & $6.89e{-}17$ & $9.30e{-}17$ & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+\multicolumn{1}{|c|}{$p=1,q=2$} & $6.80e{-}17$ & $1.56e{-}16$ & $1.22e{-}16$ & $7.14e{-}17$ & $1.17e{-}17$ & $1.27e{-}17$ & $8.05e{-}18$ & $\mathbf{9.77e{-}5}$ \\
+\multicolumn{1}{|c|}{$p=1,q=3$} & $8.16e{-}17$ & $8.94e{-}17$ & $6.21e{-}17$ & $\mathbf{6.77e{-}5}$ & $2.44e{-}17$ & $\mathbf{7.69e{-}4}$ & $8.61e{-}17$ & $\mathbf{5.02e{-}4}$ \\
+\multicolumn{1}{|c|}{$p=1,q=4$} & $3.70e{-}16$ & $\mathbf{1.83e{-}4}$ & $7.12e{-}16$ & $\mathbf{4.11e{-}4}$ & $2.65e{-}16$ & $\mathbf{5.42e{-}4}$ & $5.03e{-}16$ & $\mathbf{4.36e{-}4}$ \\
+\hline
+\multicolumn{1}{|c|}{$p=2,q=1$} & $2.42e{-}16$ & $2.34e{-}16$ & $1.18e{-}16$ & $3.42e{-}17$ & $2.78e{-}17$ & $\mathbf{1.30e{-}4}$ & $1.22e{-}16$ & $\mathbf{3.06e{-}4}$ \\
+\multicolumn{1}{|c|}{$p=2,q=2$} & $4.50e{-}16$ & $1.47e{-}16$ & $7.47e{-}17$ & $\mathbf{7.04e{-}5}$ & $4.05e{-}16$ & $\mathbf{6.47e{-}4}$ & - & - \\
+\multicolumn{1}{|c|}{$p=2,q=3$} & $1.85e{-}16$ & $\mathbf{1.65e{-}4}$ & $7.76e{-}16$ & $\mathbf{5.57e{-}4}$ & $1.88e{-}15$ & $\mathbf{1.47e{-}3}$ & - & - \\
+\multicolumn{1}{|c|}{$p=2,q=4$} & $2.22e{-}15$ & $\mathbf{2.90e{-}4}$ & $4.01e{-}15$ & $\mathbf{1.42e{-}3}$ & - & - & - & - \\
+\hline
+\multicolumn{1}{|c|}{$p=3,q=1$} & $6.54e{-}17$ & $5.96e{-}17$ & $8.13e{-}17$ & $\mathbf{9.70e{-}5}$ & - & - & - & - \\
+\multicolumn{1}{|c|}{$p=3,q=2$} & $9.43e{-}17$ & $1.26e{-}16$ & $8.99e{-}16$ & $\mathbf{5.23e{-}4}$ & - & - & - & - \\
+\multicolumn{1}{|c|}{$p=3,q=3$} & $2.27e{-}16$ & $\mathbf{2.86e{-}4}$ & $5.77e{-}15$ & $\mathbf{1.81e{-}3}$ & - & - & - & - \\
+\multicolumn{1}{|c|}{$p=3,q=4$} & $2.45e{-}15$ & $\mathbf{5.31e{-}4}$ & $1.62e{-}14$ & $\mathbf{2.54e{-}3}$ & - & - & - & - \\
+\hline
+\end{tabularx}
+}
+\begin{figure*}
+\massmatrixtwodresultsnorms
+\end{figure*}
 

### 2014-03-11 paper/bare_conf.tex
Adjust table for performance results

@@
 For the employed storage layout, the loop permutations $ijk$ and $ikj$ are likely to maximize performance.
-Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration.
-Our experiments showed that the performance loss is greater than the gain due to accumulating increments in a register along the $i$ loop.
+Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to accumulating increments in a register along the $i$ loop.
 The choice between $ijk$ and $ikj$ depends on the number of load instructions that can be hoisted out of the innermost dimension.
@@
 The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}.
-For Advection-Diffusion the ``Diffusion'' equation is considered, which uses the same differential operators as Helmholtz.
+For Advection-Diffusion, the ``Diffusion'' equation, which uses the same differential operators as Helmholtz, is considered.
 In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the stiffness matrix $A$.
 Burgers is a time-dependent problem (i.e.
-the assembly is recalculated every time step based on the result of previous iterations) employing differential operators different than those of Helmholtz and Advection-Diffusion.
+the assembly is recalculated every time step based on the result of previous iterations) employing differential operators different than those of Helmholtz.
 The assembly code, partly exposed in Figure~\ref{code:burgers}, is therefore fairly different, with a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) employed.
@@
 In the case of $p=4$, where the stiffness matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only $licm$, data alignment and padding were used.
-As anticipated, both extraction of matrix-matrix multiplications from analysis of the kernel's AST and invasive re-design of the Fenics Form Compiler to explicitly expose these operations will be addressed in further work.
+As anticipated, extraction of matrix-matrix multiplications from analysis of the kernel's AST or re-design of the Fenics Form Compiler to explicitly expose these operations will be addressed in further work.
 However, these experiments justify that there's a set of problems for which turning to BLAS is not beneficial in terms of performance.
-It is possible that employing BLAS is useful for particularly large problems, for instance Burgers on 3D meshes with $p \geq 3$, although the loss in data locality due to re-loading matrices appearing in multiple products might inhibit the performance gain.
+It is possible that employing BLAS is useful for particularly large problems, for instance Burgers on 3D meshes with $p \geq 3$, although the loss in data locality due to re-loading matrices appearing in multiple products might limit the performance gain.
 
@@
 \newcommand{\massmatrixtwodresultsnorms}{
-\begin{tabularx}{1.0\textwidth}{c|XX|XX|XX|XX|}
-\cline{2-9}
-& \multicolumn{2}{c|}{$n_f=1$} & \multicolumn{2}{c|}{$n_f=2$} & \multicolumn{2}{c|}{$n_f=3$} & \multicolumn{2}{c|}{$n_f=4$} \\
-\cline{2-9}
-& A & Tensor& A & Tensor& A & Tensor& A & Tensor\\
-\hline
-\multicolumn{1}{|c|}{$p=1,q=1$} & $6.89e{-}17$ & $9.30e{-}17$ & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-\multicolumn{1}{|c|}{$p=1,q=2$} & $6.80e{-}17$ & $1.56e{-}16$ & $1.22e{-}16$ & $7.14e{-}17$ & $1.17e{-}17$ & $1.27e{-}17$ & $8.05e{-}18$ & $\mathbf{9.77e{-}5}$ \\
-\multicolumn{1}{|c|}{$p=1,q=3$} & $8.16e{-}17$ & $8.94e{-}17$ & $6.21e{-}17$ & $\mathbf{6.77e{-}5}$ & $2.44e{-}17$ & $\mathbf{7.69e{-}4}$ & $8.61e{-}17$ & $\mathbf{5.02e{-}4}$ \\
-\multicolumn{1}{|c|}{$p=1,q=4$} & $3.70e{-}16$ & $\mathbf{1.83e{-}4}$ & $7.12e{-}16$ & $\mathbf{4.11e{-}4}$ & $2.65e{-}16$ & $\mathbf{5.42e{-}4}$ & $5.03e{-}16$ & $\mathbf{4.36e{-}4}$ \\
-\hline
-\multicolumn{1}{|c|}{$p=2,q=1$} & $2.42e{-}16$ & $2.34e{-}16$ & $1.18e{-}16$ & $3.42e{-}17$ & $2.78e{-}17$ & $\mathbf{1.30e{-}4}$ & $1.22e{-}16$ & $\mathbf{3.06e{-}4}$ \\
-\multicolumn{1}{|c|}{$p=2,q=2$} & $4.50e{-}16$ & $1.47e{-}16$ & $7.47e{-}17$ & $\mathbf{7.04e{-}5}$ & $4.05e{-}16$ & $\mathbf{6.47e{-}4}$ & - & - \\
-\multicolumn{1}{|c|}{$p=2,q=3$} & $1.85e{-}16$ & $\mathbf{1.65e{-}4}$ & $7.76e{-}16$ & $\mathbf{5.57e{-}4}$ & $1.88e{-}15$ & $\mathbf{1.47e{-}3}$ & - & - \\
-\multicolumn{1}{|c|}{$p=2,q=4$} & $2.22e{-}15$ & $\mathbf{2.90e{-}4}$ & $4.01e{-}15$ & $\mathbf{1.42e{-}3}$ & - & - & - & - \\
-\hline
-\multicolumn{1}{|c|}{$p=3,q=1$} & $6.54e{-}17$ & $5.96e{-}17$ & $8.13e{-}17$ & $\mathbf{9.70e{-}5}$ & - & - & - & - \\
-\multicolumn{1}{|c|}{$p=3,q=2$} & $9.43e{-}17$ & $1.26e{-}16$ & $8.99e{-}16$ & $\mathbf{5.23e{-}4}$ & - & - & - & - \\
-\multicolumn{1}{|c|}{$p=3,q=3$} & $2.27e{-}16$ & $\mathbf{2.86e{-}4}$ & $5.77e{-}15$ & $\mathbf{1.81e{-}3}$ & - & - & - & - \\
-\multicolumn{1}{|c|}{$p=3,q=4$} & $2.45e{-}15$ & $\mathbf{5.31e{-}4}$ & $1.62e{-}14$ & $\mathbf{2.54e{-}3}$ & - & - & - & - \\
-\hline
+\begin{tabularx}{1.0\textwidth}{cccc|cc|cc|}
+\cline{5-8}
+& & & & \multicolumn{2}{c|}{$split$} & \multicolumn{2}{c|}{$op$-$vect$} \\
+\cline{2-8}
+& \multicolumn{1}{|c|}{original} & \multicolumn{1}{|c|}{$licm$} & $licm$+$ap$ & Best & Cost Model & Best & Cost Model\\
+\cline{1-8}
+\multicolumn{1}{|c|}{Helmholtz$\_$tri$\_$p1} & & \multicolumn{1}{|c|}{$5.49e{-}17$} & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+\cline{1-8}
 \end{tabularx}

### 2014-03-14 paper/bare_conf.tex
Minor fixes

@@
 % Register allocation and register locality
-\subsection{Domain-drive Optimization of Register Management}
+\subsection{Domain-driven Vector-register Tiling}
 One notable problem assembly kernels are exposed to concerns register allocation and register locality.
@@
 Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management.
-Canonical optimizations, such as loop interchange, unroll, unroll-and-jam and register tiling are typically employed to deal with this problem.
-In the compiler we have developed, we support these optimizations either by means of explicit source code transformation (interchange, unroll-and-jam) or indirectly advising the compiler through standard pragmas (unroll).
-In this context, register tiling is a critical optimization.
-A way of obtaining register tiles consists of slicing the innermost loop into rectangular blocks of identical size (except the ``reminder'' block), and applying unrolling to individual blocks in order to expose ILP.
-This is automated by some vendor compilers~\cite{comdex}.
-Register tiles along both inner dimensions can be determined as well.
-Based on the observation that the evaluation of the stiffness matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, we implement a domain-specific vector-register tiling strategy that leads to larger tiles, enhancing reuse.
+Canonical optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem.
+In the compiler we have developed, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll).
+Tiling at the level of vector registers can be introduced as well.
+Based on the observation that the evaluation of the stiffness matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a domain-specific vector-register tiling strategy can be implemented.
 If we consider the code snippet in Figure~\ref{code:helmholtz-licm} (Helmholtz after loop-invariant code motion), we can notice that the computation of $A$ is abstractly expressible as
@@
 Because of the large number of (possibly parametric) transformations, we need a mechanism to select the most suitable optimization strategy for a given problem.
-Auto-tuning is not an attractive approach since it would be performed at run-time when kernel code is available.
-The optimization strategy is described in the following, along with an overview of our compiler.
+Autotuning might be used, although at the moment we avoid it to minimize the run-time overhead.
+Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of our compiler.
 
@@
 Based on a cost model, the $split$ and $op$-$vect$ transformations may be introduced; their implementation is based on analysis and trasformation of the AST.
-The compiler handles any possible corner cases: for example, if $op$-$vect$ is to be applied but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.
+When $op$-$vect$ is selected, the compiler outputs proper AVX (or LRbi) intrinsics code.
+Any possible corner cases are handled: for example, if $op$-$vect$ is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.
 
@@
 Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size.
-Finally, the unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of ``free'' registers, i.e.
-those not used for keeping hoisted terms (line 9-11).
-
-Loop unroll and unroll-and-jam of outer loops are fundamental to expose ILP and data reuse, and so tuning critical parameters such as the unroll factor becomes of great importance.
-Autotuning might be used, although we avoid it to minimize the overhead of Firedrake's just-in-time compilation.
-It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for our assembly kernels, where the loop nest is affine, bounds are known at compile-time, and memory accesses are unit-stride, newest versions of a vendor compiler like $icc$ employ cost models capable of estimating close-to-optimal values for such parameters.
+Finally, the unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of the available registers, i.e.
+those not used for storing hoisted terms (line 9-11).
+
+Loop unroll and unroll-and-jam of outer loops are fundamental to expose ILP and data reuse, and so tuning critical parameters, like the unroll factor, becomes of great importance.
+It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for assembly kernels, where the loop nest is affine, bounds are known at compile-time, and memory accesses are unit-stride, newest versions of a vendor compiler like $icc$ employ cost models capable of estimating close-to-optimal values for such parameters.
 We therefore leave the backend compiler in charge to select unroll and unroll-and-jam factors.
@@
 For the employed storage layout, the loop permutations $ijk$ and $ikj$ are likely to maximize performance.
-Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to accumulating increments in a register along the $i$ loop.
+Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop.
 The choice between $ijk$ and $ikj$ depends on the number of load instructions that can be hoisted out of the innermost dimension.

### 2014-03-24 paper/bare_conf.tex
Integrate Paul's suggestions

@@
 \section{Introduction}
-In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, programs model phenomena by means of partial differential equations (PDEs).
-Numerical techniques, like Finite Volume Method and Finite Element Method, are widely employed in real-world applications to approximate solutions of PDEs.
+In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by means of partial differential equations (PDEs).
+Numerical techniques, like finite volume method and finite element method, are widely employed to approximate solutions of PDEs.
 Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective.
@@
 
-We focus on Local Assembly (``assembly'', in the following), a fundamental step of a Finite Element Method that covers an important fraction of the overall computation run-time, typically in the range 30$\%$-60$\%$.
+We focus on Local Assembly (``assembly'', in the following), a fundamental step of a Finite Element Method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$.
 During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain.
@@
 Optimization of assembly kernels is non-trivial.
-Polyhedral compilation could be used to improve performance; however, given the structure and the exceptionally small size of the kernels, effective SIMD vectorization can be obtained only through an in-depth analysis of both data and iteration space, which is not supported by state-of-the-art polyhedral compilers.
-BLAS routines could be employed as well, although fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplications.
-BLAS libraries are also known to perform distant from peak performance when the dimension of the matrices is small~\cite{nek5000}.
+Given their structure and the exceptionally small size, assembly kernels benefit from transformations like generalized loop-invariant code motion, vector-register tiling and code splitting, which are not supported by state-of-the-art polyhedral and vendor compilers.
+BLAS routines could be theoretically employed, although fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplications.
+BLAS libraries are also known to perform far from peak performance when the dimension of the matrices is small~\cite{nek5000}.
 As detailed in Section~\ref{sec:perf-results}, hand-made BLAS implementations of the Helmholtz assembly kernel (illustrated later) have run-times worse than those achieved with our optimization strategy.
@@
 Given the constraints on polyhedral compilation and linear-algebra-specialized libraries, we address assembly optimization by studying a set of domain-specific code transformations, applicable to a wide class of problems.
-We have developed an optimizing compiler and we have integrated it with Firedrake.
+We have developed an optimizing compiler and we have integrated it with Firedrake, a system for the resolution of PDEs based on the finite element method~\cite{firedrake}.
 This allows us to evaluate our code transformations in a range of real-world problems, varying parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the mesh type (2D, 3D, 3D hybrid structured-unstructured).
@@
 Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 33$\%$ over ``softly-optimized'' code (i.e.
-where only basic transformations are performed, such as auto-vectorizable loop-invariant code motion, padding and data alignment), and up to 77$\%$ over original kernels.
-The contribution of this paper is therefore twofold:
+where only basic transformations are performed, such as loop-invariant code motion, padding and data alignment), and up to 77$\%$ over original kernels.
+The contribution of this paper is threefold
 \begin{itemize}
-\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely Local Assembly in the context of the Finite Element Method.
+\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely local assembly in the context of the finite element method.
 Our approach exploits domain knowledge to go beyond the limits of both vendor (e.g.
@@
 polyhedral) compilers.
-\item Automation of such code optimizations in Firedrake. This enables performing an in-depth performance evaluation of the proposed optimization strategy in real-world simulations.
+\item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
+\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements.
 \end{itemize}
 
-The paper is organized as follows. ...
+The paper is organized as follows.
+In Section~\ref{sec:background} we provide some background on local assembly, showing code generated by Firedrake and emphasizing important computational aspects.
+Section~\ref{sec:code-transf} describes in detail the various code transformations, highlighting when and how domain-knowledge has been exploited.
+The design and implementation of our compiler is discussed in Section~\ref{sec:pyop2-compiler}.
+Section~\ref{sec:perf-results} shows performance results.
+Related work are illustrated in Section~\ref{sec:related-work}, while Section~\ref{sec:conclusions} concludes the paper.
 
@@
 ~~\KwSty{static const double} X2$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
-~~//It follows 11 other basis functions definitions.\\
+~~//11 other basis functions definitions.\\
 ~~...\\
@@
 ~~~~\KwSty{double} F0 = 0.0;\\
-~~~~//It follows 10 other declarations (F1, F2,...)\\
+~~~~//10 other declarations (F1, F2,...)\\
 ~~~~...\\
@@
 ~~~~~~F0 += (w[r][0]*X1$\_$D100[i][r]);\\
-~~~~~~//It follows 10 analogous statements (F1, F2, ...)\\
+~~~~~~//10 analogous statements (F1, F2, ...)\\
 ~~~~$\rbrace$\\
@@
 ~~~~~~~~~*(K2*X2$\_$D100[i][j])+...+(K8*X2$\_$D001[i][j]))..)+\\
-~~~~~~~~~+ $<$roughly a hundred of sum/muls go here$>$)..)*\\
+~~~~~~~~~+ $<$roughly a hundred sum/muls go here$>$)..)*\\
 ~~~~~~~~~*det*W5[i]);\\
@@
 
-Local assembly consists of evaluating so called element stiffness matrix and element stiffness vector; in this work, we focus on computation of matrices, which is the costly part of the process.
+Local assembly consists of evaluating so called element stiffness matrix (just stiffness matrix in the following) and element stiffness vector; in this work, we focus on computation of stiffness matrices, which is the costly part of the process.
 A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell (or element) of the discretized domain.
-Numerical integration algorithms are widely-used in assembly codes to implement evaluation of stiffness matrices~\cite{quadrature1, quadrature2}.
+Numerical integration algorithms are widely-used in assembly codes to evaluate stiffness matrices~\cite{quadrature1, quadrature2}.
 
@@
 It then triggers compilation of such kernels using an available vendor compiler, and eventually manages parallel execution over the mesh.
-In our work, we have enhanced this execution model by adding an optimization stage prior to the generation of C code.
+The subject of this papaer is to enhance this execution model by adding an optimization stage prior to the generation of C code.
 The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used.
@@
 
-The complexity of Firedrake-generated kernels depends on the mathematical problem being solved.
+The complexity of Firedrake-generated kernels depends on the finite element problem being solved.
 In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products.
@@
 Note that in addition to a larger number of operations to compute the stiffness matrix, the Burgers case shows a register pressure higher than that in Helmholtz.
-Despite assembly kernels being problem-dependent, meaning that infinite is the space of codes that Firedrake can generate, it is still possible to identify common domain-specific traits, which can be exploited for effective code transformations and SIMD vectorization.
+Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits, which can be exploited for effective code transformations and SIMD vectorization.
 
@@
 1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest.
-This is not true in general, since the unstructured mesh might use bended elements that would require the Jacobian be re-computed every $i$ iteration; 2) memory accesses along the three loop dimensions are always 1-stride; 3) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g.
-the $F$ values in Burgers) and introduction of temporary arrays; 4) the $j$ and $k$ loops iterate over the same iteration space; 5) so called basis functions arrays (denoted by $X$, $Y$, ...) are constants, and most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
+This is not true in general, since bent elements might be used in the unstructured mesh, which would require the Jacobian be re-computed at every $i$ iteration; 2) memory accesses along the three loop dimensions are always stride-1; 3) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g.
+the $F$ values in Burgers) and introduction of temporary arrays; 4) the $j$ and $k$ loops iterate over the same iteration space; 5) arrays like $X$, $Y$, ..., which represent so called basis functions and their derivatives, are constants; 6) most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
 In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.
@@
 Data alignment and padding relate to an AVX machine.
-In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$; in general, this could not be the case.}
+In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$; in general, this might not be the case.}
 \label{code:helmholtz-licm}
@@
 % Autovectorizable LICM -> show code snippet?
-\subsection{Domain-driven Loop-invariant Code Motion}
+\subsection{Generalized Loop-invariant Code Motion}
 From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only.
@@
 \subsection{Domain-driven Vector-register Tiling}
-One notable problem assembly kernels are exposed to concerns register allocation and register locality.
+One notable problem of assembly kernels concerns register allocation and register locality.
 The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high.
-Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management.
+Since the kernel's working set fits the L1 cache, it is remarkably important to optimize register management.
 Canonical optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem.
@@
 % Splitting
-\subsection{Assembly splitting}
-In complex kernels, like Burgers in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
-If the number of variables independent of the innermost-loop dimension, i.e.
-basis function matrices, temporaries introduced by loop-invariant code motion and constants, is even close to the number of available CPU registers, it is likely to obtain poor register reuse.
+\subsection{Code splitting}
+In complex kernels, like Burgers' in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
+If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, it is likely to obtain poor register reuse.
+This usually happens when the number of basis function matrices, temporaries introduced by loop-invariant code motion, and problem constants is large.
 For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the $ijk$ loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers.
@@
 tile size), we need a mechanism to prune such a large space of optimization.
-This aspect is treated in Section~\ref{sec:pyop2-compiler}
+This aspect is treated in Section~\ref{sec:pyop2-compiler}.
 
@@
 Loop unroll and unroll-and-jam of outer loops are fundamental to expose ILP and data reuse, and so tuning critical parameters, like the unroll factor, becomes of great importance.
-It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for assembly kernels, where the loop nest is affine, bounds are known at compile-time, and memory accesses are unit-stride, newest versions of a vendor compiler like $icc$ employ cost models capable of estimating close-to-optimal values for such parameters.
+It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for assembly kernels, where the loop nest is affine, bounds are known at compile-time, and memory accesses are unit-stride, recent versions of a vendor compiler like $intel$'s (or $icc$) employ cost models capable of estimating close-to-optimal values for such parameters.
 We therefore leave the backend compiler in charge to select unroll and unroll-and-jam factors.
@@
 In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the stiffness matrix $A$.
-Burgers is a time-dependent problem (i.e.
-the assembly is recalculated every time step based on the result of previous iterations) employing differential operators different than those of Helmholtz.
-The assembly code, partly exposed in Figure~\ref{code:burgers}, is therefore fairly different, with a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) employed.
+Burgers is a time-dependent problem, i.e.
+the assembly is recalculated every time step based on the result of previous iterations.
+It employs differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are accessed.
 
@@
 As anticipated, extraction of matrix-matrix multiplications from analysis of the kernel's AST or re-design of the Fenics Form Compiler to explicitly expose these operations will be addressed in further work.
-However, these experiments justify that there's a set of problems for which turning to BLAS is not beneficial in terms of performance.
+However, these experiments justify that there is a set of problems for which turning to BLAS is not beneficial in terms of performance.
 It is possible that employing BLAS is useful for particularly large problems, for instance Burgers on 3D meshes with $p \geq 3$, although the loss in data locality due to re-loading matrices appearing in multiple products might limit the performance gain.
@@
 Table~\ref{table:perf-results}
-%In addition, register locality would be affected, since the same matrix could appear in a number of products.
-
-%The problem is time-dependent, so , which intuitively increaes the computation cost.
-
-%Also, our results prove that current vendor compilers still have notable limitations, despite optimizing loop nests that are 1) affine, 2) have no stencil, and 3) ?
-
-%\begin{table}[h]
-%\begin{tabulary}{1.0\columnwidth}{|C|C|}
-%\hline
-%Problem & shape & p & original & $licm$ & $licm$ \\\hline\hline
-%Auto-vectorizable Loop-invariant code motion ($licm$) &   \\ \hline
-%Padding &  \\ \hline
-%Data Alignment & \\ \hline
-%Loop interchange      & loops  \\ \hline
-%Loop unrolling  & unroll factor \\ \hline
-%Register tiling & tile size \\ \hline
-%Outer-product vectorization ($op$-$vect$) & tile size \\ \hline
-%Assembly splitting ($split$) & split point, split factor \\ \hline
-%\end{tabulary}
-%\caption{.}
-%\label{table:perf-results}
-%\end{table}
 
@@
 
-% We fix a specific loop permutation (i esterno perche minimizza.... swapping j or k, after LICM, is equivalent)
-
 \section{Related Work}
+\label{sec:related-work}
 %Show other work of people working on optimizations for computational science kernels.
@@
 \section{Conclusions}
-
-%\subsubsection{Subsubsection Heading Here}
-%Subsubsection text here.
-
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation Results}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command, the
-% \label for the overall figure must come after \caption.
-% \hfil must be used as a separator to get equal spacing.
-% The subfigure.sty package works much the same way, except \subfigure is
-% used instead of \subfloat.
-%
-%\begin{figure*}[!t]
-%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
-%\label{fig_second_case}}}
-%\caption{Simulation results}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-
-% An example of a floating table. Note that, for IEEE style tables, the 
-% \caption command should come BEFORE the table. Table text will default to
-% \footnotesize as IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that IEEE does not put floats in the very first column - or typically
-% anywhere on the first page for that matter. Also, in-text middle ("here")
-% positioning is not used. Most IEEE journals/conferences use top floats
-% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
-% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
-% command of the stfloats package.
-
-% conference papers do not normally have an appendix
+\label{sec:conclusions}
 
@@
 
-The authors would like to thank...
+This research is partly funded by the MAPDES project and by the Department of Computing at Imperial College London.
+The authors would like to thank Prof.
+J.
+Ramanujam for his invaluable suggestions, and Dr.
+Lawrence Mitchell and Dr.
+Francis Russel for their contributions to the PyOP2 projects.
 

### 2014-03-24 paper/bare_conf.tex
Modifications + related work

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations}
-
-%\author{\IEEEauthorblockN{Fabio Luporini}
-%\IEEEauthorblockA{Department of Computing\\
-%Imperial College London\\
-%London, UK\\
-%Email: f.luporini12@imperial.ac.uk}
-%\and
-%\IEEEauthorblockN{Florian Rathgeber}
-%\IEEEauthorblockA{Department of Computing\\
-%Imperial College London\\
-%London, UK\\
-%Email: f.rathgeber10@imperial.ac.uk}
-%\and
-%\IEEEauthorblockN{Ana Lucia Varbanescu}
-%\IEEEauthorblockA{Department of Computing\\
-%Imperial College London\\
-%London, UK\\
-%Email: a.l.varbanescu@uva.nl}
-%\and
-%\IEEEauthorblockN{Gheorghe-Teodor Bercea}
-%\IEEEauthorblockA{Department of Computing\\
-%Imperial College London\\
-%London, UK\\
-%Email: gb308@doc.ic.ac.uk}
-%\and
-%\IEEEauthorblockN{Paul H. J. Kelly}
-%\IEEEauthorblockA{Department of Computing\\
-%Imperial College London\\
-%London, UK\\
-%Email: p.kelly@imperial.ac.uk}
-%\and
-%\IEEEauthorblockN{David Ham}
-%\IEEEauthorblockA{Department of Computing\\
-%Imperial College London\\
-%London, UK\\
-%Email: d.ham@imperial.ac.uk}}
+\title{FEA-Comp: an Optimizing Compiler\\for Solving Partial Differential Equations}
 
@@
 Gheorghe-Teodor Bercea\IEEEauthorrefmark{1}, \\
-Paul H. J. Kelly\IEEEauthorrefmark{1},
+David Ham\IEEEauthorrefmark{1},
 and
-David Ham\IEEEauthorrefmark{1}}
+Paul H. J. Kelly\IEEEauthorrefmark{1}}
 \IEEEauthorblockA{\IEEEauthorrefmark{1}Imperial College London, Department of Computing,\\
@@
 \subsection{Generalized Loop-invariant Code Motion}
+\label{sec:code-transf-licm}
 From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only.
@@
 
-Table~\ref{table:perf-results}
+Tables~\ref{table:perf-results-licm-helmholtz},~\ref{table:perf-results-licm-advdiff}, and~\ref{table:perf-results-licm-burgers} illustrate the speed-ups obtained on the Sandy Bridge and the Phi machines when $licm$, data alignment, and padding are used, over non-transformed code.
+Inspection of assembly code generated by $icc$ confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant with respect to outer loops are hoisted and, interestingly, not vectorized.
+This motivates the usually significant gain.
+Padding and data alignment enhance, in general, the quality of SIMD auto-vectorization.
+Sometimes the value of $licm$-$ap$ is similar to that of $licm$ because the stiffness matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned.
+Occasionally $licm$-$ap$ is slower than $licm$ (e.g.
+in Burgers 3D p3).
+This is due to the large number of aligned temporaries introduced by $licm$, which probably lead to cache associativity conflicts.
 
 \newcommand{\massmatrixtwodresultsnorms}{
-\begin{tabularx}{1.0\textwidth}{cccc|cc|cc|}
-\cline{5-8}
-& & & & \multicolumn{2}{c|}{$split$} & \multicolumn{2}{c|}{$op$-$vect$} \\
-\cline{2-8}
-& \multicolumn{1}{|c|}{original} & \multicolumn{1}{|c|}{$licm$} & $licm$+$ap$ & Best & Cost Model & Best & Cost Model\\
-\cline{1-8}
-\multicolumn{1}{|c|}{Helmholtz$\_$tri$\_$p1} & & \multicolumn{1}{|c|}{$5.49e{-}17$} & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-\cline{1-8}
+\begin{tabularx}{1.0\textwidth}{p{3.5cm}|p{1.98cm}p{1.98cm}p{1.98cm}|p{1.98cm}p{1.98cm}p{1.98cm}|}
+\cline{2-7}
+& \multicolumn{3}{c|}{\texttt{AVX}} & \multicolumn{3}{c|}{\texttt{PHI}} \\
+\cline{2-7}
+& \texttt{original} & \texttt{licm} & \texttt{licm+ap} & \texttt{original} & \texttt{licm} & \texttt{licm+ap}\\
+\cline{1-7}
+\multicolumn{1}{|c|}{\texttt{triangle p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+\multicolumn{1}{|c|}{\texttt{triangle p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+\multicolumn{1}{|c|}{\texttt{triangle p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+\multicolumn{1}{|c|}{\texttt{triangle p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+\cline{1-7}
+\multicolumn{1}{|c|}{\texttt{tetrahedron p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+\multicolumn{1}{|c|}{\texttt{tetrahedron p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+\multicolumn{1}{|c|}{\texttt{tetrahedron p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+\multicolumn{1}{|c|}{\texttt{tetrahedron p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+\cline{1-7}
+\multicolumn{1}{|c|}{\texttt{prism p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+\multicolumn{1}{|c|}{\texttt{prism p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+\multicolumn{1}{|c|}{\texttt{prism p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+\multicolumn{1}{|c|}{\texttt{prism p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+\cline{1-7}
+
 \end{tabularx}
 }
-\begin{figure*}
+
+\begin{table*}[t]\normalsize
 \massmatrixtwodresultsnorms
-\end{figure*}
+\label{table:perf-results-licm-helmholtz}
+\caption{Impact of generalized loop-invariant code motion ($licm$ column) on the Helmholtz problem, for three kind of elements belonging to the Conjugate Gradient family (triangle, tetrahedron, prism) and for the range of polynomial orders $p$ $\in$ $\lbrace$1, 4$\rbrace$.
+The column $licm$-$ap$ illustrates the combination of $licm$ with data alignment and padding.}
+\end{table*}
 
@@
 \label{sec:related-work}
+The finite element method is used in the most disparate contexts to approximate solutions of PDEs.
+Well-known frameworks and applications include~\cite{nek5000-web-page},~\cite{Fenics},~\cite{fluidity_manual_v4}, and of course Firedrake; this is not an exhaustive list, though.
+Different techniques and algorithms are used depending on the level of accuracy that needs to be reached.
+When low-order polynomials are used, standard numerical integration is usually employed.
+The recent introduction of DSLs to decouple the finite element specification from its underlying implementation facilitated, however, the development of novel approaches.
+For instance, methods based on tensor contraction~\cite{FFC-TC} and symbolic manipulation~\cite{Francis} have been implemented, although numerical integration remains the optimal choice for a wide class of problems~\cite{quadrature1}.
+
+Optimization of local assembly using numerical integration for CPU platforms has been tackled in~\cite{quadrature1}.
+However, to the best of our knowledge, ours is the first work targeting low-level compiler optimizations.
+In~\cite{Markall20101815}, and more recently in~\cite{petsc-integration-gpu}, the problem has been studied for GPUs.
+In~\cite{assembly-opencl}, variants of the standard numerical integration algorithm are specialized for the PowerXCell processor and evaluated; it lacks, however, an exhaustive study from the compiler viewpoint as we did, and none of the optimizations presented in Section~\ref{sec:code-transf} is discussed.
+
+Domain-specific languages and automated code generation have been adopted with success in different fields.
+Spiral~\cite{Pueschel:05} automates generation of highly-optimized platform-specific digital signal processing numerical algorithms.
+Similar ideas are adopted in~\cite{Spampinato:14}, which proposes DSLs and compilers for dense linear algebra kernels.
+OP2~\cite{op2-main}, and its python implementation~\cite{pyop2isc}, which is also used by Firedrake, aim at achieving performance portability for scientific computations based on unstructured meshes.
+
+- finding the best problem implementation: cost model based vs auto-tuning vs hybrid
+
+%(i.e. given a mesh, those that lead to ``less accurate'' solutions than when using high-order polynomials, at a lower computational cost)
+
 %Show other work of people working on optimizations for computational science kernels.

### 2014-03-25 paper/bare_conf.tex
Add conclusions

@@
 %
-%\usepackage{url}
+\usepackage{url}
 % url.sty was written by Donald Arseneau. It provides better support for
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{FEA-Comp: an Optimizing Compiler\\for Solving Partial Differential Equations}
+\title{XY: an Optimizing Compiler\\for Solving Partial Differential Equations}
 
@@
 On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread.
-Similar programming and execution models are adopted by~\cite{fenics, fluidity, lizst}.
+Similar programming and execution models are adopted by~\cite{Fenics, fluidity, lizst}.
 Kernel optimization is one of the major concerns in unstructured mesh applications.
@@
 \begin{table}[h]
+\caption{Overview of code transformations for Firedrake-generated assembly kernels.}
+\label{table:code-transformations}
+\begin{center}
 \begin{tabulary}{1.0\columnwidth}{|C|C|}
@@
 Name (Abbreviation) & Parameter \\\hline\hline
-Auto-vectorizable Loop-invariant code motion ($licm$) &   \\ \hline
+Generalized loop-invariant code motion ($licm$) &   \\ \hline
 Padding &  \\ \hline
@@
 \end{tabulary}
-\caption{Overview of code transformations for Firedrake-generated assembly kernels.}
-\label{table:code-transformations}
+\end{center}
 \end{table}
@@
 Other optimization levels performed, in general, slightly worse than \texttt{-O2}.
-Our code transformations were evaluated in three real-world problems expressible with Firedrake:
+Our code transformations were evaluated in three real-world problems based on the following PDEs:
 \begin{itemize}
@@
 \end{itemize}
+The code was written in UFL and then executed over real unstructured meshes through Firedrake.
 The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}.
@@
 The finite element method is used in the most disparate contexts to approximate solutions of PDEs.
-Well-known frameworks and applications include~\cite{nek5000-web-page},~\cite{Fenics},~\cite{fluidity_manual_v4}, and of course Firedrake; this is not an exhaustive list, though.
-Different techniques and algorithms are used depending on the level of accuracy that needs to be reached.
-When low-order polynomials are used, standard numerical integration is usually employed.
+Well-known frameworks and applications include nek5000~\cite{nek5000-web-page}, the Fenics project~\cite{Fenics}, Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake; this is not an exhaustive list, though.
+Numerical integration is usually employed to implement the local assembly phase.
 The recent introduction of DSLs to decouple the finite element specification from its underlying implementation facilitated, however, the development of novel approaches.
-For instance, methods based on tensor contraction~\cite{FFC-TC} and symbolic manipulation~\cite{Francis} have been implemented, although numerical integration remains the optimal choice for a wide class of problems~\cite{quadrature1}.
+Methods based on tensor contraction~\cite{FFC-TC} and symbolic manipulation~\cite{Francis} have been developed, although it has been demonstrated that numerical integration remains the optimal choice for a wide class of problems~\cite{quadrature1}.
 
 Optimization of local assembly using numerical integration for CPU platforms has been tackled in~\cite{quadrature1}.
-However, to the best of our knowledge, ours is the first work targeting low-level compiler optimizations.
-In~\cite{Markall20101815}, and more recently in~\cite{petsc-integration-gpu}, the problem has been studied for GPUs.
-In~\cite{assembly-opencl}, variants of the standard numerical integration algorithm are specialized for the PowerXCell processor and evaluated; it lacks, however, an exhaustive study from the compiler viewpoint as we did, and none of the optimizations presented in Section~\ref{sec:code-transf} is discussed.
+However, to the best of our knowledge, ours is the first work targeting low-level optimizations and adopting a real compiler approach.
+In~\cite{Markall20101815}, and more recently in~\cite{petsc-integration-gpu}, the problem has been studied for GPU architectures.
+In~\cite{assembly-opencl}, variants of the standard numerical integration algorithm have been specialized for the PowerXCell processor and evaluated; The paper lacks, however, an exhaustive study from the compiler viewpoint as we did, and none of the optimizations presented in Section~\ref{sec:code-transf} are mentioned.
 
-Domain-specific languages and automated code generation have been adopted with success in different fields.
+Domain-specific languages and automated code generation have been adopted with success in different areas.
 Spiral~\cite{Pueschel:05} automates generation of highly-optimized platform-specific digital signal processing numerical algorithms.
-Similar ideas are adopted in~\cite{Spampinato:14}, which proposes DSLs and compilers for dense linear algebra kernels.
-OP2~\cite{op2-main}, and its python implementation~\cite{pyop2isc}, which is also used by Firedrake, aim at achieving performance portability for scientific computations based on unstructured meshes.
+Similar ideas are adopted in~\cite{Spampinato:14}, where a DSL and a compiler for dense linear algebra kernels are proposed.
+OP2~\cite{op2-main}, and its python implementation~\cite{pyop2isc}, which is used by Firedrake to express iteration over meshes, aim at achieving performance portability for scientific codes based on unstructured meshes.
+Stencil DSLs and compilers, such as Pochoir~\cite{pochoir} and SDSL~\cite{stencil-compiler}, have been developed to support development and high performance in fields like image processing and scientific computations over structured grids.
 
-- finding the best problem implementation: cost model based vs auto-tuning vs hybrid
+XY currently uses a simple cost model and heuristics to steer the optimization process.
+Many code generators, like those based on the Polyhedral model~\cite{PLUTO} and those driven by domain-knowledge~\cite{modeldriven}, make use of cost models.
+The other extreme consists of relying on auto-tuning to select the best implementation for a given problem on a certain platform.
+Notable examples include tuning of small matrix-matrix multiplies in nek5000~\cite{nek5000}, the ATLAS library~\cite{ATLAS}, and FFTW~\cite{FFTW} for fast fourier transforms.
+In both cases, pruning the implementation space, as we did in Section~\ref{sec:pyop2-compiler}, is fundamental to mitigate complexity and overhead.
 
@@
 
+In this paper we have presented design, optimizations and systematic performance evaluation of XY, a compiler for finite element local assembly.
+In this context, to the best of our knowledge, this is the first compiler oriented towards the introduction of low-level optimizations to maximize instruction-level parallelism, register locality and SIMD vectorization.
+Assembly kernels have peculiar characteristics.
+Their iteration space is usually very small, with the size depending on aspects like the degree of accuracy one wants to reach (polynomial order of the method) and the mesh discretization employed.
+The data space, in terms of number of matrices and scalar, grows proportionally to the complexity of the finite element problem.
+XY has been developed taking into account all of these aspects, knowing that some transformations are useful only in a subset of all possible problems.
+The various optimizations overcome limitations of current vendor and research compilers.
+The exploitation of domain-knowledge allows some of them to be particularly effective, as demonstrated by our experiments on two state-of-the-art Intel platforms.
+Further work include a comprehensive study about feasibility and constraints of transforming the kernel into a sequence of BLAS calls.
+XY supports all of the problems expressible in Firedrake, and it is already integrated with this framework.
+
 % use section* for acknowledgement
@@
 Lawrence Mitchell and Dr.
-Francis Russel for their contributions to the PyOP2 projects.
+Francis Russell for their contribution to the PyOP2 project.
 

### 2014-03-26 paper/bare_conf.tex
Various additions. New plots.

@@
 %
-\usepackage[longend,linesnumbered]{algorithm2e}
+\usepackage[longend,linesnumbered,figure]{algorithm2e}
 \usepackage{algorithmic}
@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{XY: an Optimizing Compiler\\for Solving Partial Differential Equations}
+\title{XY: an Optimizing Compiler\\for Finite Element Local Assembly}
 
@@
 %\boldmath
-The abstract goes here.
+The finite element method is widely-employed to determine approximated solutions of partial differential equations.
+Local assembly is one of its fundamental steps covering a considerable fraction of the overall run-time.
+In this paper we present design, implementation and systematic performance evaluation of XY, an optimizing compiler for local assembly numerical kernels.
+An abstract syntax tree of such kernels is provided to XY by the higher layer of abstractions, which make extensive use of domain-specific languages.
+XY manipulates this abstract representation by introducing composable optimizations aimed at improving instruction-level parallelism, register locality and SIMD vectorization.
+It then generates C code, possibly including AVX or LBRi intrinsics, which is eventually just-in-time compiled on the underlying architecture and executed.
+Performance evaluation using a suite of examples of real-world importance showed that speed-ups over non-optimized kernels between 1.32$\times$ and 4.44$\times$ can be achieved.
 \end{abstract}
+
 % IEEEtran.cls defaults to using nonbold math in the Abstract.
@@
 
-% no keywords
+\begin{IEEEkeywords}
+Finite element method, compilers, optimizations, simd vectorization 
+\end{IEEEkeywords}
 
@@
 To address this problem, domain-specific languages (DSLs) have been developed.
-The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2ics}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{OP2-hydra}.
+The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2isc}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{OP2-hydra}.
 
@@
 On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread.
-Similar programming and execution models are adopted by~\cite{Fenics, fluidity, lizst}.
+Similar programming and execution models are adopted in~\cite{Fenics},~\cite{fluidity_manual_v4},~\cite{lizst}.
 Kernel optimization is one of the major concerns in unstructured mesh applications.
@@
 
-We focus on Local Assembly (``assembly'', in the following), a fundamental step of a Finite Element Method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$.
+We focus on Local Assembly (``assembly'', in the following), a fundamental step of a finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$.
 During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain.
@@
 With such small kernels, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.
-Our study is conducted in the context of Firedrake, a system for solving PDEs through the Finite Element Method based on the OP2 abstraction~\cite{firedrake}.
+Our study is conducted in the context of Firedrake, a system for solving PDEs through the finite element method based on the OP2 abstraction~\cite{firedrake}.
 
@@
 Given the constraints on polyhedral compilation and linear-algebra-specialized libraries, we address assembly optimization by studying a set of domain-specific code transformations, applicable to a wide class of problems.
-We have developed an optimizing compiler and we have integrated it with Firedrake, a system for the resolution of PDEs based on the finite element method~\cite{firedrake}.
+We have developed an optimizing compiler and we have integrated it with Firedrake.
 This allows us to evaluate our code transformations in a range of real-world problems, varying parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the mesh type (2D, 3D, 3D hybrid structured-unstructured).
@@
 Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal.
-Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 33$\%$ over ``softly-optimized'' code (i.e.
-where only basic transformations are performed, such as loop-invariant code motion, padding and data alignment), and up to 77$\%$ over original kernels.
+Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (i.e.
+where only basic transformations are performed, such as loop-invariant code motion, padding and data alignment), and up to 4.44$\times$ over original kernels.
 The contribution of this paper is threefold
@@
 Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.
+
 \begin{algorithm}[t]
 \small
-\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).}
-\label{code:helmholtz}
 \KwSty{void} helmholtz(double A[3][3], double **coords) $\lbrace$\\
@@
 $\rbrace$
+\caption{Local assembly code generated by Firedrake for a Helmholtz problem on a 2D triangular mesh with Lagrange $p=1$ elements.}
+\label{code:helmholtz}
 \end{algorithm}
@@
 \small
-\caption{Local assembly code generated by FFC for a Burgers problem (3D mesh, Lagrange $p=1$ elements).}
-\label{code:burgers}
 \KwSty{void} burgers(double A[12][12], double **c, double **w) $\lbrace$\\
@@
 $\rbrace$
+\caption{Local assembly code generated by Firedrake for a Burgers problem on a 3D tetrahedral mesh with Lagrange $p=1$ elements.}
+\label{code:burgers}
 \end{algorithm}
@@
 A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell (or element) of the discretized domain.
-Numerical integration algorithms are widely-used in assembly codes to evaluate stiffness matrices~\cite{quadrature1, quadrature2}.
+Numerical integration algorithms are widely-used in assembly codes to evaluate stiffness matrices~\cite{quadrature1, fluidity_manual_v4}.
 
@@
 The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used.
-Notable examples are~\cite{uflacs, fluidity, mcfc}, in which the kernel's loop nest and memory access pattern are reducible to that generated by Firedrake.
 
@@
 \small
-\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).
-The padding, data alignment and $licm$ optimizations are applied.
-Data alignment and padding relate to an AVX machine.
-In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$; in general, this might not be the case.}
-\label{code:helmholtz-licm}
 \KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
@@
 $\rbrace$
+\caption{Local assembly code generated by Firedrake when padding, data alignment, and $licm$ are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}.
+Data alignment and padding are for an AVX machine.
+In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$, so they can be precomputed once in a single loop $r$; in general, this might not be the case.}
+\label{code:helmholtz-licm}
 \end{algorithm}
@@
 \begin{figure}[h]
-\normalsize
-\centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect.pdf}
-\label{fig:vect-by-vect}}
+\centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect.pdf}}
 \caption{Outer-product vectorization by permuting values in a vector register.}
+\label{fig:vect-by-vect}
 \end{figure}
@@
 \begin{figure}[t]
-\centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}
-\label{fig:restore-layout}}
+\centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}}
 \caption{Restoring the storage layout after $op$-$vect$. The figure shows how 4$\times$4 elements in the top-left block of the stiffness matrix $A$ can be brought to their correct positions.}
+\label{fig:restore-layout}
 \end{figure}
@@
 \small
-\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=2$ elements).
-The padding, data alignment, $licm$ and $op$-$vect$ optimizations are applied.
-Data alignment and padding relate to an AVX machine.
-The original size of the $j$-$k$ iteration space (i.e.
-before padding was applied) was 6$\times$6.
-In this example, the unroll-and-jam factor is 1.}
-\label{code:helmholtz-opvect}
 \KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
@@
 $\rbrace$
+\caption{Local assembly code generated by Firedrake when padding, data alignment, $licm$ and $op$-$vect$ are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}.
+Here, we assume the polynomial order is $p=2$, since $op$-$vect$ can not be used when an iteration space dimension is smaller than the vector length.
+The original size of the $j$-$k$ iteration space (i.e.
+before padding was applied) was 6$\times$6.
+In this example, the unroll-and-jam factor is 1.}
+\label{code:helmholtz-opvect}
 \end{algorithm}
@@
 % Splitting
-\subsection{Code splitting}
+\subsection{Expression Splitting}
 In complex kernels, like Burgers' in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
@@
 \small
-\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).
-The padding, data alignment, $licm$ and $split$ optimizations are applied.
-Data alignment and padding relate to an AVX machine.
-In this specific case, the split factor is 2.}
-\label{code:helmholtz-split}
 \KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
@@
 $\rbrace$
+\caption{Local assembly code generated by Firedrake when $split$ is applied to the optimized Helmholtz problem given in Figure~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
+\label{code:helmholtz-split}
 \end{algorithm}
@@
 \begin{table}[h]
-\caption{Overview of code transformations for Firedrake-generated assembly kernels.}
-\label{table:code-transformations}
 \begin{center}
@@
 \end{center}
+\caption{Overview of code transformations for Firedrake-generated assembly kernels.}
+\label{table:code-transformations}
 \end{table}
@@
 
-\section{The PyOP2 Compiler}
+\section{Overview of the XY Compiler}
 \label{sec:pyop2-compiler}
@@
 \begin{algorithm}[t]
-\caption{The PyOP2 compiler.}
 \label{algo:PyOP2Compiler}
-  \textbf{The PyOP2 Compiler}\\
+  \textbf{The XY Compiler}\\
   \KwIn{ast, wrapper, isa}
@@
 \If{\KwSty{not} it\_space}{
-  ast.apply\_inter\_kernel\_vectorization(wrapper)
+  ast.apply\_inter\_kernel\_vectorization(wrapper, isa)
   \Return{wrapper + ast.from\_ast\_to\_c()}
@@
 \If{plan.uaj\_factor}{
-  ast.op\_vect(plan.uaj\_factor)
+  uaj = MIN(plan.uaj\_factor, $\lfloor$it\_space.j.size/isa.vf$\rfloor$) \\
+  ast.op\_vect(uaj)
 }
 \Return{wrapper + ast.from\_ast\_to\_c()}
+\caption{Pseudocode of the XY pipeline.}
 \end{algorithm}
@@
 Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations.
-This high-level specification is translated by the Fenics Form Compiler~\cite{ffc} into an Abstract Syntax Tree representation of a Finite Element assembly kernel.
+This high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an Abstract Syntax Tree representation of a Finite Element assembly kernel.
 ASTs are then passed to PyOP2~\cite{pyop2isc}, which lies at the core of Firedrake, where parallel execution over the unstructured mesh is managed.
@@
 If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied.
-This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{reguly-inter-kernel}.
+This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{inter-kernel-vect}.
 The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{quadrature-cuda}.
@@
 The cost model is shown in Figure~\ref{algo:applyCostModel}.
-It takes into account the number of available vector registers ($n\_regs$) and the number of variables iterating along the $j$ and $k$ dimensions ($n\_consts$ for independent variables, $n\_outer\_arrays$ for $j$ variables, and $n\_inner\_arrays$ for $k$ variables, assuming the $ijk$ loop order) to estimate unroll-and-jam and split factors when, respectively, $op$-$vect$ and $split$ are used (if a factor is 0, then the corresponding transformation is not applied).
+It takes into account the number of available logical vector registers ($n\_regs$) and the number of variables iterating along the $j$ and $k$ dimensions ($n\_consts$ for independent variables, $n\_outer\_arrays$ for $j$ variables, and $n\_inner\_arrays$ for $k$ variables, assuming the $ijk$ loop order) to estimate unroll-and-jam and split factors when, respectively, $op$-$vect$ and $split$ are used.
+The $n\_consts$ parameter includes temporary registers to carry out computations, so setting its value is partly driven by heuristics.
+If a factor is 0, then the corresponding transformation is not applied.
 The $split$ transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 2).
@@
 It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for assembly kernels, where the loop nest is affine, bounds are known at compile-time, and memory accesses are unit-stride, recent versions of a vendor compiler like $intel$'s (or $icc$) employ cost models capable of estimating close-to-optimal values for such parameters.
-We therefore leave the backend compiler in charge to select unroll and unroll-and-jam factors.
+We leave therefore the backend compiler in charge to select unroll and unroll-and-jam factors.
 This choice also simplifies the compiler's cost model.
 The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $icc$ compiler from applying unroll-based optimizations, even if specific pragmas are added.
+Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 25 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order $ijk$).
 
@@
 \begin{algorithm}[t]
-\caption{The cost model is employed by the compiler to estimate the most suitable unroll-and-jam (when $op$-$vect$ is used) and split factors, avoiding the overhead of auto-tuning.}
 \label{algo:applyCostModel}
@@
 \Return{$<$split\_factor, uaj\_factor$>$}
+\caption{The cost model is employed by the compiler to estimate the most suitable unroll-and-jam (when $op$-$vect$ is used) and split factors, avoiding the overhead of auto-tuning.}
 \end{algorithm}
@@
 
+\subsection{Experimental Setup}
 Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Phi.
@@
 
+\subsection{Impact of Generalized Loop-invariant Code Motion}
+\newcommand{\massmatrixtwodresultsnorms}{
+\begin{tabularx}{1.0\textwidth}{p{3.5cm}p{3.5cm}|p{2.25cm}p{2.25cm}p{2.25cm}p{2.25cm}|}
+\cline{3-6}
+& & \multicolumn{2}{c|}{\texttt{AVX}} & \multicolumn{2}{c|}{\texttt{Phi}}\\
+\cline{3-6}
+& & \multicolumn{1}{c}{\texttt{licm}} & \multicolumn{1}{c|}{\texttt{licm+ap}} & \multicolumn{1}{c}{\texttt{licm}} & \multicolumn{1}{c|}{\texttt{licm+ap}} \\
+\cline{1-6}
+\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{prism}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
+\cline{1-6}
+\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{triangle}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{prism}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
+\cline{1-6}
+\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{triangle}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{prism}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
+\cline{1-6}
+\end{tabularx}
+}
+
+\begin{table*}[t]\normalsize
+\massmatrixtwodresultsnorms
+\label{table:perf-results-licm-helmholtz}
+\caption{Impact of generalized loop-invariant code motion ($licm$ column) on the Helmholtz, Diffusion and Burgers problems, for three kind of elements belonging to the Lagrange family (triangle, tetrahedron, prism), for the range of polynomial orders $p$ $\in$ $\lbrace$1, 4$\rbrace$.
+Each entry indicates the range of speed-ups obtained over the non-optimized implementation.
+The column $licm$-$ap$ illustrates the combination of $licm$ with data alignment and padding.
+Results are shown for both the Sandy Bridge and the Phi machine.}
+\end{table*}
+
 Tables~\ref{table:perf-results-licm-helmholtz},~\ref{table:perf-results-licm-advdiff}, and~\ref{table:perf-results-licm-burgers} illustrate the speed-ups obtained on the Sandy Bridge and the Phi machines when $licm$, data alignment, and padding are used, over non-transformed code.
@@
 Padding and data alignment enhance, in general, the quality of SIMD auto-vectorization.
-Sometimes the value of $licm$-$ap$ is similar to that of $licm$ because the stiffness matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned.
+As shown in Figure~\ref{algo:PyOP2Compiler}, these transformations are applied on top of $licm$, so we refer to them as $licm$-$ap$.
+Sometimes the run-time of $licm$-$ap$ is similar to that of $licm$ because the stiffness matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned.
 Occasionally $licm$-$ap$ is slower than $licm$ (e.g.
 in Burgers 3D p3).
-This is due to the large number of aligned temporaries introduced by $licm$, which probably lead to cache associativity conflicts.
-
-\newcommand{\massmatrixtwodresultsnorms}{
-\begin{tabularx}{1.0\textwidth}{p{3.5cm}|p{1.98cm}p{1.98cm}p{1.98cm}|p{1.98cm}p{1.98cm}p{1.98cm}|}
-\cline{2-7}
-& \multicolumn{3}{c|}{\texttt{AVX}} & \multicolumn{3}{c|}{\texttt{PHI}} \\
-\cline{2-7}
-& \texttt{original} & \texttt{licm} & \texttt{licm+ap} & \texttt{original} & \texttt{licm} & \texttt{licm+ap}\\
-\cline{1-7}
-\multicolumn{1}{|c|}{\texttt{triangle p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-\multicolumn{1}{|c|}{\texttt{triangle p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-\multicolumn{1}{|c|}{\texttt{triangle p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-\multicolumn{1}{|c|}{\texttt{triangle p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-\cline{1-7}
-\multicolumn{1}{|c|}{\texttt{tetrahedron p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-\multicolumn{1}{|c|}{\texttt{tetrahedron p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-\multicolumn{1}{|c|}{\texttt{tetrahedron p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-\multicolumn{1}{|c|}{\texttt{tetrahedron p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-\cline{1-7}
-\multicolumn{1}{|c|}{\texttt{prism p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-\multicolumn{1}{|c|}{\texttt{prism p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-\multicolumn{1}{|c|}{\texttt{prism p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-\multicolumn{1}{|c|}{\texttt{prism p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-\cline{1-7}
-
-\end{tabularx}
-}
-
-\begin{table*}[t]\normalsize
-\massmatrixtwodresultsnorms
-\label{table:perf-results-licm-helmholtz}
-\caption{Impact of generalized loop-invariant code motion ($licm$ column) on the Helmholtz problem, for three kind of elements belonging to the Conjugate Gradient family (triangle, tetrahedron, prism) and for the range of polynomial orders $p$ $\in$ $\lbrace$1, 4$\rbrace$.
-The column $licm$-$ap$ illustrates the combination of $licm$ with data alignment and padding.}
-\end{table*}
+This is due to the large number of aligned temporaries introduced by $licm$, which probably leads to cache associativity conflicts.
+
+\subsection{Impact of Vector-register Tiling}
+Figures~\ref{fig:opvect-helmholtz-speedup},~\ref{fig:opvect-diffusion-speedup}, and~\ref{fig:opvect-burgers-speedup} show the speed-ups achieved by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmoltz, Diffusion, and Burgers assembly kernels, respectively.
+For each problem instance we report the best run-time obtained by empirically testing a set of different unroll/unroll-and-jam factors, and the one retrieved through the XY's cost model.
+
+% why better results at higher order
+
+\subsection{Impact of Expression Splitting}
+% why split useless in helmholtz/diffusion
+% why vect vect+split useless in burgers
 
 \begin{figure}[h]
-\centerline{\includegraphics[scale=0.7]{Pictures/helmholtz-normalized.pdf}
-\label{fig_first_case}}
-\caption{Helmholtz}
+\centerline{\includegraphics[scale=0.7]{Pictures/helmholtz-normalized-opvect.pdf}
+\label{fig:opvect-helmholtz-speedup}}
+\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmholtz kernel.}
 \end{figure}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[scale=0.7]{Pictures/advdiff-normalized.pdf}
-\label{fig_first_case}}
-\caption{Advection-Diffusion}
+\centerline{\includegraphics[scale=0.7]{Pictures/diffusion-normalized-opvect.pdf}
+\label{fig:opvect-diffusion-speedup}}
+\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Diffusion kernel.}
 \end{figure}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[scale=0.7]{Pictures/burgers-normalized.pdf}
-\label{fig_first_case}}
-\caption{Burgers}
+\centerline{\includegraphics[scale=0.7]{Pictures/burgers-normalized-split.pdf}
+\label{fig:opvect-burgers-speedup}}
+\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Burgers kernel.}
 \end{figure}
@@
 
+%\newcommand{\massmatrixtwodresultsnorms}{
+%\begin{tabularx}{1.0\textwidth}{p{3.5cm}|p{1.98cm}p{1.98cm}p{1.98cm}|p{1.98cm}p{1.98cm}p{1.98cm}|}
+%\cline{2-7}
+%& \multicolumn{3}{c|}{\texttt{AVX}} & \multicolumn{3}{c|}{\texttt{PHI}} \\
+%\cline{2-7}
+%& \texttt{original} & \texttt{licm} & \texttt{licm+ap} & \texttt{original} & \texttt{licm} & \texttt{licm+ap}\\
+%\cline{1-7}
+%\multicolumn{1}{|c|}{\texttt{triangle p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+%\multicolumn{1}{|c|}{\texttt{triangle p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+%\multicolumn{1}{|c|}{\texttt{triangle p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+%\multicolumn{1}{|c|}{\texttt{triangle p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+%\cline{1-7}
+%\multicolumn{1}{|c|}{\texttt{tetrahedron p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+%\multicolumn{1}{|c|}{\texttt{tetrahedron p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+%\multicolumn{1}{|c|}{\texttt{tetrahedron p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+%\multicolumn{1}{|c|}{\texttt{tetrahedron p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+%\cline{1-7}
+%\multicolumn{1}{|c|}{\texttt{prism p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+%\multicolumn{1}{|c|}{\texttt{prism p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+%\multicolumn{1}{|c|}{\texttt{prism p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+%\multicolumn{1}{|c|}{\texttt{prism p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
+%\cline{1-7}
+%
+%\end{tabularx}
+%}
+%
+%\begin{table*}[t]\normalsize
+%\massmatrixtwodresultsnorms
+%\label{table:perf-results-licm-helmholtz}
+%\caption{Impact of generalized loop-invariant code motion ($licm$ column) on the Helmholtz problem, for three kind of elements belonging to the Lagrange family (triangle, tetrahedron, prism) and for the range of polynomial orders $p$ $\in$ $\lbrace$1, 4$\rbrace$.
+The column $licm$-$ap$ illustrates the combination of $licm$ with data alignment and padding.}
+%\end{table*}
+
 % trigger a \newpage just before the given reference

### 2014-03-27 paper/bare_conf.tex
added some paragraphs

@@
 Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.
+
+\begin{algorithm}[t]
+\small
+\KwIn{stiffness matrix A (initialized to 0),\\~~~~~~~~~element coordinates in the domain (array),\\~~~~~~~~~(optional) fields (e.g. velocity, array)}
+\KwOut{stiffness matrix A (computed)}
+\nlset{1}~~- Jacobian computation, using element coordinates \\
+~~- Declaration of constant data, among which basis function\\
+~~- Computation of the stiffness matrix
+\caption{General structure of a local assembly kernel generated by Firedrake.}
+\label{code:general-structure}
+\end{algorithm}
 
@@
 
-The complexity of Firedrake-generated kernels depends on the finite element problem being solved.
+The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}.
+Input is a pointer to a 0-valued bi-dimensional array, which will be used to store the stiffness matrix; a pointer to the element coordinates in the discretized domain; one or more pointers to so called fields, for instance indicating value of velocity or pressure in the element.
+The output is the evaluated element stiffness matrix.
+A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse (referred to as $det$ and $K$ in the following code snippets), 2) declaration of constant arrays (identified by letters $X$, $Y$, ...), 3) evaluation of the stiffness matrix (called $A$) in an affine loop nest.
+In finite element terminology, constant (bi-dimensional) arrays are called ``basis functions'' (or derivatives of basis functions).
+They are identical to all mesh elements.
+
+The actual complexity of a local assembly kernel depends on the finite element problem being solved.
 In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products.
 An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$.
-The stiffness matrix in the code is called $A$.
-In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of $A$ can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
+In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of the stiffness matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
 Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse.
@@
 This is not true in general, since bent elements might be used in the unstructured mesh, which would require the Jacobian be re-computed at every $i$ iteration; 2) memory accesses along the three loop dimensions are always stride-1; 3) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g.
-the $F$ values in Burgers) and introduction of temporary arrays; 4) the $j$ and $k$ loops iterate over the same iteration space; 5) arrays like $X$, $Y$, ..., which represent so called basis functions and their derivatives, are constants; 6) most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
+the $F$ values in Burgers) and introduction of temporary arrays; 4) the $j$ and $k$ loops iterate over the same iteration space; 5) most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
 In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.
@@
 \label{sec:code-transf}
-%All the work done in the polyhedral model should be useful, but our kernels fit L1, so they need special attention for what concerns key aspects like managing registers, exploiting ILP in its various forms, and reducing loop overhead.
-%Follows a description of code transformations using our compiler-driven approach.
+The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake.
+One peculiar characteristic is that they all aim at improving the run-time of the $ijk$ loop nest, where the numerical integration takes place.
+In rare, yet important cases, however, this might not be sufficient to achieve notable speed-ups.
+Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions.
+It is worth noting there are circumnstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that its execution time becomes the dominant factor of the overall local assembly run-time.
+In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure).
+Given its inherent complexity, a comprehensive study of this problem is left as further work.
+
+% Padding and Data Alignment
+\subsection{Padding and Data Alignment}
+Auto-vectorization of assembly code computing the stiffness matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length ($vl$).
+Data alignment is enforced in two steps.
+Initially, both arrays and matrices are allocated to addresses that are multiples of $vl$.
+Then, matrices are padded by rounding the number of columns to the nearest multiple of $vl$.
+For example, assume the original size of a matrix is 3$\times$3 and that the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values.
+Then, a padded matrix on this architecture will have size 3$\times$4.
+The compiler is explicitly informed about data alignment using a suitable pragma.
+Padding of all matrices involved in the evaluation of the stiffness matrix allows us to safely round the loop trip count to the nearest multiple of $vl$.
+This avoids the introduction of a remainder (scalar) loop from the compiler, which would be responsible for inefficient vectorization.
+
+% Autovectorizable LICM -> show code snippet?
+\subsection{Generalized Loop-invariant Code Motion}
+\label{sec:code-transf-licm}
+From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only.
+Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion.
+Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels.
+We want to overcome two deficiencies that both $icc$ and $gcc$ have.
+First, these compilers can identify sub-expressions that are invariant with respect to the innermost loop only.
+This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted.
+Second, the hoisted code is scalar, i.e.
+it is not subjected to auto-vectorization.
+We work around these limitations with source-level loop-invariant code motion.
+In particular, we pre-compute all values that an invariant sub-expression assumes along the fastest varying dimension.
+This is implemented by introducing a temporary array (per invariant sub-expression) and by adding a new loop to the nest.
+At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop.
+Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant.
+It is also worth noting that, in some problems, invariant sub-expressions along $j$ are identical to those along $k$ (e.g.
+in Helmholtz).
+In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.
 
@@
 \end{algorithm}
-
-% Padding and Data Alignment
-\subsection{Padding and Data Alignment}
-Auto-vectorization of assembly code computing the stiffness matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length ($vl$).
-Data alignment is enforced in two steps.
-Initially, both arrays and matrices are allocated to addresses that are multiples of $vl$.
-Then, matrices are padded by rounding the number of columns to the nearest multiple of $vl$.
-For example, assume the original size of a matrix is 3$\times$3 and that the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values.
-Then, a padded matrix on this architecture will have size 3$\times$4.
-The compiler is explicitly informed about data alignment using a suitable pragma.
-Padding of all matrices involved in the evaluation of the stiffness matrix allows us to safely round the loop trip count to the nearest multiple of $vl$.
-This avoids the introduction of a remainder (scalar) loop from the compiler, which would be responsible for inefficient vectorization.
-
-% Autovectorizable LICM -> show code snippet?
-\subsection{Generalized Loop-invariant Code Motion}
-\label{sec:code-transf-licm}
-From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only.
-Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion.
-Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels.
-We want to overcome two deficiencies that both $icc$ and $gcc$ have.
-First, these compilers can identify sub-expressions that are invariant with respect to the innermost loop only.
-This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted.
-Second, the hoisted code is scalar, i.e.
-it is not subjected to auto-vectorization.
-We work around these limitations with source-level loop-invariant code motion.
-In particular, we pre-compute all values that an invariant sub-expression assumes along the fastest varying dimension.
-This is implemented by introducing a temporary array (per invariant sub-expression) and by adding a new loop to the nest.
-At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop.
-Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant.
-It is also worth noting that, in some problems, invariant sub-expressions along $j$ are identical to those along $k$ (e.g.
-in Helmholtz).
-In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.
 

### 2014-03-27 paper/bare_conf.tex
finished writing licm impact

@@
 Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Phi.
-The \texttt{icc 2013} compiler was used, with optimization level \texttt{-O2} and with auto-vectorization enabled (\texttt{-xAVX} on the Sandy Bridge, and \texttt{TODO} on the Phi).
+The \texttt{icc 13.1} compiler was used, with optimization level \texttt{-O2} and with auto-vectorization enabled (\texttt{-xAVX} on the Sandy Bridge, and \texttt{TODO} on the Phi).
 Other optimization levels performed, in general, slightly worse than \texttt{-O2}.
@@
 \massmatrixtwodresultsnorms
-\label{table:perf-results-licm-helmholtz}
+\label{table:perf-results-licm}
 \caption{Impact of generalized loop-invariant code motion ($licm$ column) on the Helmholtz, Diffusion and Burgers problems, for three kind of elements belonging to the Lagrange family (triangle, tetrahedron, prism), for the range of polynomial orders $p$ $\in$ $\lbrace$1, 4$\rbrace$.
@@
 
-Tables~\ref{table:perf-results-licm-helmholtz},~\ref{table:perf-results-licm-advdiff}, and~\ref{table:perf-results-licm-burgers} illustrate the speed-ups obtained on the Sandy Bridge and the Phi machines when $licm$, data alignment, and padding are used, over non-transformed code.
+Table~\ref{table:perf-results-licm} illustrates the speed-ups obtained on the Sandy Bridge and the Phi machines when $licm$, data alignment, and padding are used, over non-transformed code.
 Inspection of assembly code generated by $icc$ confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant with respect to outer loops are hoisted and, interestingly, not vectorized.
-This motivates the usually significant gain.
+This motivates the usually significant gain obtained.
 Padding and data alignment enhance, in general, the quality of SIMD auto-vectorization.
@@
 \subsection{Impact of Vector-register Tiling}
-Figures~\ref{fig:opvect-helmholtz-speedup},~\ref{fig:opvect-diffusion-speedup}, and~\ref{fig:opvect-burgers-speedup} show the speed-ups achieved by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmoltz, Diffusion, and Burgers assembly kernels, respectively.
-For each problem instance we report the best run-time obtained by empirically testing a set of different unroll/unroll-and-jam factors, and the one retrieved through the XY's cost model.
+Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the speed-ups achieved by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmoltz and Diffusion kernels, respectively.
+For each problem instance we report two bars: one indicates the best run-time achieved by autotuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved by the XY's cost model.
+In general, there is no substantial difference between the two.
+This is chiefly because assembly kernels fit the L1 cache, so estimating the run-time based on the quality of register allocation is relatively simple.
+The rationale behind these results is that, for smaller configurations, vector-register tiling is rarely helpful.
+This is due to the fact that in the considered problems the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short.
+On the other hand, as soon as the iteration space becomes too big, vector-register tiles are essential to minimize pipeline stalls.
+A maximum speed-up of 1.40$\times$ over $licm$-$ap$ was observed in the Diffusion kernel when using a prismatic mesh and $p=4$.
+Application of $op$-$vect$ to the Burgers problem induces meaningful slow downs, because of the large number of arrays that, generally, need to be tiled.
+Expression splitting can be used in combination to $op$-$vect$ to alleviate this issue; this is discussed in Section~\ref{sec:perf-results-split}.
 
@@
 \subsection{Impact of Expression Splitting}
+\label{sec:perf-results-split}
+Expression splitting, or loop fission, contributes to relieve the register pressure in loop nests where a large number of arrays is accessed.
+If the stiffness matrix can be evaluated by accessing a few (relatively ) basis functions or temporary values introduced by $licm$,
+
 % why split useless in helmholtz/diffusion
@@
 \begin{figure}[h]
-\centerline{\includegraphics[scale=0.7]{Pictures/helmholtz-normalized-opvect.pdf}
-\label{fig:opvect-helmholtz-speedup}}
-\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmholtz kernel.}
+\includegraphics[scale=0.7]{Pictures/helmholtz-normalized-opvect.pdf}
+\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmholtz kernel.
+Tri, Tet, and Prism stand, respectively, for triangual, tetrahedral and prismatic mesh.
+CM stands for cost model.}
+\label{fig:opvect-helmholtz-speedup}
 \end{figure}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[scale=0.7]{Pictures/diffusion-normalized-opvect.pdf}
-\label{fig:opvect-diffusion-speedup}}
-\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Diffusion kernel.}
+\includegraphics[scale=0.7]{Pictures/diffusion-normalized-opvect.pdf}
+\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Diffusion kernel.
+Tri, Tet, and Prism stand, respectively, for triangual, tetrahedral and prismatic mesh.
+CM stands for cost model.}
+\label{fig:opvect-diffusion-speedup}
 \end{figure}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[scale=0.7]{Pictures/burgers-normalized-split.pdf}
-\label{fig:opvect-burgers-speedup}}
-\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Burgers kernel.}
+\includegraphics[scale=0.7]{Pictures/burgers-normalized-split.pdf}
+\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Burgers kernel.
+Tri, Tet, and Prism stand, respectively, for triangual, tetrahedral and prismatic mesh.
+CM stands for cost model.}
+\label{fig:opvect-burgers-speedup}
 \end{figure}

### 2014-03-28 paper/bare_conf.tex
One day of refactoring...

@@
 % can use linebreaks \\ within to get better formatting as desired
-\title{XY: an Optimizing Compiler\\for Finite Element Local Assembly}
+\title{COFFEE: an Optimizing Compiler\\for Finite Element Local Assembly}
 
@@
 David Ham\IEEEauthorrefmark{1},
+J. Ramanujam\IEEEauthorrefmark{3},
 and
@@
 \IEEEauthorrefmark{2}TU Delft, Faculty of Engineering, Mathematics and Computer Science (EWI)\\
-\IEEEauthorrefmark{1}\{f.luporini12$|$f.rathgeber10$|$gheorghe-teodor.bercea08$|$p.kelly$|$d.ham@imperial.ac.uk\}@imperial.ac.uk,\\
-\IEEEauthorrefmark{2}a.l.varbanescu@uva.nl
+\IEEEauthorrefmark{1}\{f.luporini12$|$f.rathgeber10$|$gheorghe-teodor.bercea08$|$p.kelly$|$d.ham\}@imperial.ac.uk,\\
+\IEEEauthorrefmark{2}a.l.varbanescu@uva.nl,
+\IEEEauthorrefmark{3}jxr@ece.lsu.edu
 }
@@
 The finite element method is widely-employed to determine approximated solutions of partial differential equations.
-Local assembly is one of its fundamental steps covering a considerable fraction of the overall run-time.
-In this paper we present design, implementation and systematic performance evaluation of XY, an optimizing compiler for local assembly numerical kernels.
-An abstract syntax tree of such kernels is provided to XY by the higher layer of abstractions, which make extensive use of domain-specific languages.
-XY manipulates this abstract representation by introducing composable optimizations aimed at improving instruction-level parallelism, register locality and SIMD vectorization.
-It then generates C code, possibly including AVX or LBRi intrinsics, which is eventually just-in-time compiled on the underlying architecture and executed.
+Local assembly is one of its fundamental steps, in which a suitable numerical procedure, or kernel, is applied to all elements in the discretized domain of the equation.
+Since the size of the domain can be huge, a notable problem consists of producing efficient kernels to minimize the cost local assembly.
+In this context, we present design, implementation and systematic performance evaluation of COFFEE, a domain-specific compiler for local assembly kernels.
+COFFEE manipulates kernels' abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially in the form of SIMD vectorization, and register locality.
+It then generates C code, possibly including AVX or LRBNi intrinsics, which is eventually just-in-time compiled on the underlying architecture and executed.
 Performance evaluation using a suite of examples of real-world importance showed that speed-ups over non-optimized kernels between 1.32$\times$ and 4.44$\times$ can be achieved.
@@
 In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by means of partial differential equations (PDEs).
-Numerical techniques, like finite volume method and finite element method, are widely employed to approximate solutions of PDEs.
+Numerical techniques, like finite volume method and finite element method, are widely-employed to approximate solutions of PDEs.
 Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective.
@@
 
-OP2 adopts a kernel-oriented programming model, in which the computation semantics is expressed through self-contained functions (``kernels'').
+OP2 adopts a kernel-oriented programming model, in which the computation semantics is expressed through self-contained functions, or ``kernels''.
 A kernel is applied to all elements in a set of mesh components (e.g.
-edges, vertices, elements), with an implicit synchronization between the application of two consecutive kernels.
+edges, vertices, cells/elements), with an implicit synchronization between the application of two consecutive kernels.
 On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread.
@@
 Kernel optimization is one of the major concerns in unstructured mesh applications.
-In this paper, we tackle this problem by proposing a domain-driven optimization strategy for a class of kernels used in Finite Element Methods.
-
-We focus on Local Assembly (``assembly'', in the following), a fundamental step of a finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$.
+In this paper, we tackle this problem by proposing an optimization strategy for a class of kernels used in finite element methods.
+
+We focus on local assembly (``assembly'', in the following), a fundamental step of a finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$.
 During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain.
@@
 However, we do not consider the latter case.
-An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 30, with a minimum value of 3, depending on the order of the method).
+An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 30, with a minimum value of 3, depending on the order of the method).
 With such small kernels, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.
-Our study is conducted in the context of Firedrake, a system for solving PDEs through the finite element method based on the OP2 abstraction~\cite{firedrake}.
 
@@
 
-Given the constraints on polyhedral compilation and linear-algebra-specialized libraries, we address assembly optimization by studying a set of domain-specific code transformations, applicable to a wide class of problems.
-We have developed an optimizing compiler and we have integrated it with Firedrake.
-This allows us to evaluate our code transformations in a range of real-world problems, varying parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the mesh type (2D, 3D, 3D hybrid structured-unstructured).
+Due to the constraints of available compilers and on linear-algebra specialized libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For Finite ElemEnt local assembly.}, a compiler for optimizing local assembly kernels.
+COFFEE is integrated with Firedrake, a system for solving PDEs through the finite element method based on the OP2 abstraction~\cite{firedrake-code}.
+It supports any problems expressible with this framework.
+This allows us to evaluate our code transformations in a range of real-world problems, varying key parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).
 
@@
 Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (i.e.
-where only basic transformations are performed, such as loop-invariant code motion, padding and data alignment), and up to 4.44$\times$ over original kernels.
+where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels.
 The contribution of this paper is threefold
@@
 The paper is organized as follows.
-In Section~\ref{sec:background} we provide some background on local assembly, showing code generated by Firedrake and emphasizing important computational aspects.
+In Section~\ref{sec:background} we provide some background on local assembly, showing code generated by Firedrake and emphasizing critical computational aspects.
 Section~\ref{sec:code-transf} describes in detail the various code transformations, highlighting when and how domain-knowledge has been exploited.
@@
 
-Local assembly consists of evaluating so called element stiffness matrix (just stiffness matrix in the following) and element stiffness vector; in this work, we focus on computation of stiffness matrices, which is the costly part of the process.
+Local assembly consists of evaluating so called element stiffness matrix (``stiffness matrix'', in the following) and element stiffness vector; in this work, we focus on computation of stiffness matrices, which is usually the costly part of the process.
 A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell (or element) of the discretized domain.
@@
 It then triggers compilation of such kernels using an available vendor compiler, and eventually manages parallel execution over the mesh.
-The subject of this papaer is to enhance this execution model by adding an optimization stage prior to the generation of C code.
+The subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.
 The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used.
@@
 The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}.
-Input is a pointer to a 0-valued bi-dimensional array, which will be used to store the stiffness matrix; a pointer to the element coordinates in the discretized domain; one or more pointers to so called fields, for instance indicating value of velocity or pressure in the element.
+Input are a pointer to a 0-valued bi-dimensional array, which will be used to store the stiffness matrix; a pointer to the element's coordinates in the discretized domain; one or more pointers to so called fields, for instance indicating value of velocity or pressure in the element.
 The output is the evaluated element stiffness matrix.
-A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse (referred to as $det$ and $K$ in the following code snippets), 2) declaration of constant arrays (identified by letters $X$, $Y$, ...), 3) evaluation of the stiffness matrix (called $A$) in an affine loop nest.
-In finite element terminology, constant (bi-dimensional) arrays are called ``basis functions'' (or derivatives of basis functions).
-They are identical to all mesh elements.
+A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant bi-dimensional arrays, called ``basis functions'' (or derivatives of basis functions) in finite element terminology, initialized to problem-specific double-precision floatin point values; 3) evaluation of the stiffness matrix in an affine loop nest.
+All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (i.e.
+using \texttt{static const} in C language).
+Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel's objects.
+
+\begin{table}[h]
+\begin{center}
+\begin{tabulary}{1.0\columnwidth}{|C|C|C|}
+\hline
+Object name & Type & Variable name(s) \\\hline\hline
+Determinant of Jacobian matrix & double & det  \\ \hline
+Inverse of Jacobian matrix & double[] & K \\ \hline
+Coordinates & double** & coords\\ \hline
+Fields & double** & w \\ \hline
+Integration points & double[] & W \\ \hline
+Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ \hline
+Stiffness matrix & double[][] & A\\ \hline
+\end{tabulary}
+\end{center}
+\caption{Type and variable names used in the various code snippets to identify local assembly objects.}
+\label{table:map-name-letters}
+\end{table}
 
@@
 An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$.
-In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of the stiffness matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
+In other scenarios, for instance when solving a non-linear problem like Burgers as in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the stiffness matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
 Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse.
@@
 The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake.
-One peculiar characteristic is that they all aim at improving the run-time of the $ijk$ loop nest, where the numerical integration takes place.
+One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place.
 In rare, yet important cases, however, this might not be sufficient to achieve notable speed-ups.
@@
 \subsection{Padding and Data Alignment}
-Auto-vectorization of assembly code computing the stiffness matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length ($vl$).
+Auto-vectorization of assembly code computing the stiffness matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length $vl$.
 Data alignment is enforced in two steps.
-Initially, both arrays and matrices are allocated to addresses that are multiples of $vl$.
-Then, matrices are padded by rounding the number of columns to the nearest multiple of $vl$.
-For example, assume the original size of a matrix is 3$\times$3 and that the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values.
-Then, a padded matrix on this architecture will have size 3$\times$4.
+Initially, all arrays are allocated to addresses that are multiples of $vl$.
+Then, bi-dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$.
+For instance, assume the original size of a basis function array is 3$\times$3 and that the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values.
+In this example, a padded version of the array will have size 3$\times$4.
 The compiler is explicitly informed about data alignment using a suitable pragma.
-Padding of all matrices involved in the evaluation of the stiffness matrix allows us to safely round the loop trip count to the nearest multiple of $vl$.
-This avoids the introduction of a remainder (scalar) loop from the compiler, which would be responsible for inefficient vectorization.
+Padding of all bi-dimensional arrays involved in the evaluation of the stiffness matrix also allows us to safely round the loop trip count to the nearest multiple of $vl$.
+This avoids the introduction of a remainder (scalar) loop from the compiler, which would render vectorization less efficient.
 
@@
 Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels.
-We want to overcome two deficiencies that both $icc$ and $gcc$ have.
-First, these compilers can identify sub-expressions that are invariant with respect to the innermost loop only.
-This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted.
+We want to overcome two deficiencies that both $intel$ and $gnu$ compilers have.
+First, they can identify sub-expressions that are invariant with respect to the innermost loop only.
+This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted in the default loop order \texttt{ijk}.
 Second, the hoisted code is scalar, i.e.
@@
 We work around these limitations with source-level loop-invariant code motion.
-In particular, we pre-compute all values that an invariant sub-expression assumes along the fastest varying dimension.
-This is implemented by introducing a temporary array (per invariant sub-expression) and by adding a new loop to the nest.
+In particular, we pre-compute all values that an invariant sub-expression assumes along its fastest varying dimension.
+This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest.
 At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop.
@@
 ~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
-~~~~double LI$\_$0[4];\\
-~~~~double LI$\_$1[4];\\
+~~~~double LI$\_$0[4] ALIGN;\\
+~~~~double LI$\_$1[4] ALIGN;\\
 ~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
@@
 % Register allocation and register locality
-\subsection{Domain-driven Vector-register Tiling}
+\subsection{Model-driven Vector-register Tiling}
 One notable problem of assembly kernels concerns register allocation and register locality.
@@
 Canonical optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem.
-In the compiler we have developed, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll).
-Tiling at the level of vector registers can be introduced as well.
-Based on the observation that the evaluation of the stiffness matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a domain-specific vector-register tiling strategy can be implemented.
-If we consider the code snippet in Figure~\ref{code:helmholtz-licm} (Helmholtz after loop-invariant code motion), we can notice that the computation of $A$ is abstractly expressible as
+In COFFEE, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll).
+Tiling at the level of vector registers is an additional feature of COFFEE.
+Based on the observation that the evaluation of the stiffness matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented.
+If we consider the code snippet in Figure~\ref{code:helmholtz-licm} (Helmholtz after loop-invariant code motion) and we ignore the presence of \texttt{det*W3[i]}, we can notice that the computation of the stiffness matrix is abstractly expressible as
 \begin{equation}
@@
 \end{equation}
-where $B$ is the set of all matrices (or temporaries) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets.
-Without loss of generality, the presence of constants or other variables independent of both $j$ and $k$ can be momentarily neglected.
-Note that regardless of the specific input problem, the stiffness matrix computation is always reducible to this kind of form.
-Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix using just 2 vector registers (a 4$\times$4 tile), assuming $\vert B' \vert = \vert B'' \vert = 1$.
+where $B$ is the set of all basis functions (or temporary variables, e.g.
+\texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets.
+Regardless of the specific input problem, and by abstracting from the presence of all variables independent of both $j$ and $k$, the stiffness matrix computation is always reducible to this kind of form.
+Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$.
 Values in a register are shuffled each time a product is performed.
-Standard compiler auto-vectorization ($gcc$ and $icc$), instead, executes 4 broadcast operations (i.e.
-``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation, and would also need to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.
+Standard compiler auto-vectorization ($gnu$ and $intel$), instead, executes 4 broadcast operations (i.e.
+``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation.
+Besides a larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when using unroll-and-jam, with $f$ being the unroll-and-jam factor.
 
@@
 
-The storage layout of $A$, however, is incorrect after the application of this ``outer-product vectorization'' ($op$-$vect$).
-We efficiently restore it with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the $ijk$ loop nest.
+The storage layout of $A$, however, is incorrect after the application of this ``outer-product vectorization'' ($op$-$vect$, in the following).
+We efficiently restore it with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the \texttt{ijk} loop nest.
 The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.
@@
 \centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}}
-\caption{Restoring the storage layout after $op$-$vect$. The figure shows how 4$\times$4 elements in the top-left block of the stiffness matrix $A$ can be brought to their correct positions.}
+\caption{Restoring the storage layout after $op$-$vect$.
+The figure shows how 4$\times$4 elements in the top-left block of the stiffness matrix $A$ can be moved to their correct positions.
+Each rotation, represented by a group of three same-colored arrows, is implemented by a single vector shuffle intrinsics.}
 \label{fig:restore-layout}
@@
 \subsection{Expression Splitting}
+\label{sec:expr-split}
 In complex kernels, like Burgers' in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
 If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, it is likely to obtain poor register reuse.
-This usually happens when the number of basis function matrices, temporaries introduced by loop-invariant code motion, and problem constants is large.
-For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the $ijk$ loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers.
-One potential solution to this problem consists of suitably ``splitting'' the computation of the stiffness matrix $A$ into multiple sub-expressions; an example, for the simpler Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}.
+This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large.
+For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the \texttt{ijk} loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers.
+One potential solution to this problem consists of suitably ``splitting'' the computation of the stiffness matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}.
 Splitting an expression has, in general, several drawbacks.
@@
 the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}).
-Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same matrix could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
+Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
 Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.
@@
 
-\section{Overview of the XY Compiler}
+\section{Overview of COFFEE}
 \label{sec:pyop2-compiler}
@@
 \label{algo:PyOP2Compiler}
-  \textbf{The XY Compiler}\\
+  \textbf{The COFFEE Compiler}\\
   \KwIn{ast, wrapper, isa}
@@
 \Return{wrapper + ast.from\_ast\_to\_c()}
-\caption{Pseudocode of the XY pipeline.}
+\caption{Pseudocode of the COFFEE pipeline.}
 \end{algorithm}
 
+%PyOP2 which lies at the core of Firedrake, where  is managed
+
 Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations.
-This high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an Abstract Syntax Tree representation of a Finite Element assembly kernel.
-ASTs are then passed to PyOP2~\cite{pyop2isc}, which lies at the core of Firedrake, where parallel execution over the unstructured mesh is managed.
-Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it receives FFC's ASTs as input, it introduces optimizations, and it generates C code as output, which is eventually just-in-time compiled on the underlying architecture.
-Because of the large number of (possibly parametric) transformations, we need a mechanism to select the most suitable optimization strategy for a given problem.
+This high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
+ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}.
+The output of COFFEE is C code, which is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and parallel execution over the unstructured mesh take place.
+Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem.
 Autotuning might be used, although at the moment we avoid it to minimize the run-time overhead.
-Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of our compiler.
+Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.
 
 The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}.
-Initially, an AST is inspected, looking for the presence of iteration spaces and other domain-specific information provided by the higher layer.
+Initially, an AST is inspected, looking for the presence of iteration spaces and domain-specific information provided by the higher layer.
 If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied.
 This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{inter-kernel-vect}.
-The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{quadrature-cuda}.
+The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{pyop2isc}.
 Then, an ordered sequence of optimization steps are executed.
 Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays.
-Based on a cost model, the $split$ and $op$-$vect$ transformations may be introduced; their implementation is based on analysis and trasformation of the AST.
-When $op$-$vect$ is selected, the compiler outputs proper AVX (or LRbi) intrinsics code.
+Based on a cost model, loop interchange, $split$, and $op$-$vect$ may be introduced.
+Their implementation is based on analysis and trasformation of the AST.
+When $op$-$vect$ is selected, the compiler outputs proper AVX (or LRBNi) intrinsics code.
 Any possible corner cases are handled: for example, if $op$-$vect$ is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.
 
+All loops are interchangeable provided that temporaries are introduced if the nest is not perfect.
+For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance.
+Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop.
+The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension.
+Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers are available to carry out the computation of the stiffness matrix.
+
+Loop unroll and unroll-and-jam of outer loops are fundamental to expose instruction-level parallelism and register reuse, so tuning critical parameters, like the unroll factor, becomes of great importance.
+It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for assembly kernels, where the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are unit-stride, recent versions of a vendor compiler like $intel$'s employ cost models capable of estimating close-to-optimal values for such parameters.
+We leave therefore the backend compiler in charge to select the unroll factor.
+This choice also simplifies the COFFEE's cost model.
+The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $intel$ compiler from applying this optimization, even if specific pragmas are added.
+Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 25 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order \texttt{ijk}).
+
 The cost model is shown in Figure~\ref{algo:applyCostModel}.
-It takes into account the number of available logical vector registers ($n\_regs$) and the number of variables iterating along the $j$ and $k$ dimensions ($n\_consts$ for independent variables, $n\_outer\_arrays$ for $j$ variables, and $n\_inner\_arrays$ for $k$ variables, assuming the $ijk$ loop order) to estimate unroll-and-jam and split factors when, respectively, $op$-$vect$ and $split$ are used.
+It takes into account the number of available logical vector registers ($n\_regs$) and the number of variables iterating along the $j$ and $k$ dimensions ($n\_consts$ for independent variables, $n\_outer\_arrays$ for $j$ variables, and $n\_inner\_arrays$ for $k$ variables, assuming the \texttt{ijk} loop order) to estimate unroll-and-jam and split factors when, respectively, $op$-$vect$ and $split$ are used.
 The $n\_consts$ parameter includes temporary registers to carry out computations, so setting its value is partly driven by heuristics.
@@
 Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size.
-Finally, the unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of the available registers, i.e.
+The unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of the available registers, i.e.
 those not used for storing hoisted terms (line 9-11).
-
-Loop unroll and unroll-and-jam of outer loops are fundamental to expose ILP and data reuse, and so tuning critical parameters, like the unroll factor, becomes of great importance.
-It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for assembly kernels, where the loop nest is affine, bounds are known at compile-time, and memory accesses are unit-stride, recent versions of a vendor compiler like $intel$'s (or $icc$) employ cost models capable of estimating close-to-optimal values for such parameters.
-We leave therefore the backend compiler in charge to select unroll and unroll-and-jam factors.
-This choice also simplifies the compiler's cost model.
-The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $icc$ compiler from applying unroll-based optimizations, even if specific pragmas are added.
-Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 25 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order $ijk$).
-
-All loops are interchangeable provided that temporaries are introduced if the nest is not perfect.
-For the employed storage layout, the loop permutations $ijk$ and $ikj$ are likely to maximize performance.
-Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop.
-The choice between $ijk$ and $ikj$ depends on the number of load instructions that can be hoisted out of the innermost dimension.
-Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers are availabe to carry out the computation of the stiffness matrix.
+Finally, the profitability of loop interchange is evaluated (line 12-16)
 
@@
 uaj\_factor = $\lceil$n\_reg\_avail / n\_inner\_arrays$\rceil$ \\
-\Return{$<$split\_factor, uaj\_factor$>$}
+\eIf{n\_outer\_arrays $>$ n\_inner\_arrays}{
+  permute = \KwSty{True}
+}{
+  permute = \KwSty{False}
+}
+\Return{$<$permute, split\_factor, uaj\_factor$>$}
 \caption{The cost model is employed by the compiler to estimate the most suitable unroll-and-jam (when $op$-$vect$ is used) and split factors, avoiding the overhead of auto-tuning.}
@@
 \subsection{Experimental Setup}
+
+TODO: This section needs some rewording! And still lacks the Phi discussion!! \\
+
 Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Phi.
@@
 These problems were studied varying both the shape of mesh elements and the polynomial order $p$ of the method.
-Intuitively, the larger the element shape, the bigger is the iteration space.
-Triangles (2D), tetrahedron (3D), and prisms (3D) were tested.
+Intuitively, the bigger the element shape and $p$, the larger is the iteration space.
+Triangles (2D), tetrahedron (3D), and prisms (3D) were tested as element shape.
 For instance, in the case of Helmholtz with $p=1$, the size of the $j$ and $k$ loops for the three kind of elements is, respectively, $3$, $4$, and $6$.
-Moving from 2D to 3D also increases the number of basis function arrays, since conceptually the behaviour of the equation has to be approximated also along the $z$ dimension.
+Moving from 2D to 3D has the effect of increasing the number of basis function arrays, since, conceptually, the behaviour of the equation has now to be approximated also along a third axis.
 On the other hand, the polynomial order affects only the problem size (the three loops $i$, $j$, and $k$, and, as a consequence, the size of $X$ and $Y$ arrays).
-A range of polynomial orders, from $p=1$ to $p=4$, are tested; higher polynomial orders are excluded from the study because of current Firedrake limitations.
-In such a large space of problems, the size of the stiffness matrix rarely excees 20$\times$20, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
-
-For the Helmholtz 3D problem, manual implementations based on MKL BLAS were tested on Sandy Bridge.
-This particular kernel can be easily reduced to a sequence of four matrix-matrix multiplications that can be computed via calls to BLAS \texttt{dgemm}.
-In the case of $p=4$, where the stiffness matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only $licm$, data alignment and padding were used.
-As anticipated, extraction of matrix-matrix multiplications from analysis of the kernel's AST or re-design of the Fenics Form Compiler to explicitly expose these operations will be addressed in further work.
-However, these experiments justify that there is a set of problems for which turning to BLAS is not beneficial in terms of performance.
-It is possible that employing BLAS is useful for particularly large problems, for instance Burgers on 3D meshes with $p \geq 3$, although the loss in data locality due to re-loading matrices appearing in multiple products might limit the performance gain.
+A range of polynomial orders, from $p=1$ to $p=4$, were tested; higher polynomial orders are excluded from the study because of current Firedrake limitations.
+In such a large space of problems, the size of the stiffness matrix rarely excees 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
+
+For the Helmholtz 3D problem, manual implementations based on \texttt{Intel MKL BLAS} were tested on the Sandy Bridge.
+This particular kernel can be easily reduced to a sequence of four matrix-matrix multiplies that can be computed via calls to BLAS \texttt{dgemm}.
+In the case of $p=4$, where the stiffness matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only $licm$, data alignment and padding were used, with the slow down being even worse for smaller problem sizes.
+These experiments justify that there is a set of problems for which turning to MKL is not beneficial in terms of performance.
+It is possible that employing certain BLAS implementations or auto-tuning strategies for particular problem sizes is helpful, although the same problems afflicting $split$ (exposed in Section~\ref{sec:expr-split}) might limit the performance gain.
+However, as anticipated, extraction of matrix-matrix multiplies from analysis of the kernel's AST and a strategy for their execution will be comprehensively addressed in further work.
 
@@
 \cline{3-6}
-& & \multicolumn{2}{c|}{\texttt{AVX}} & \multicolumn{2}{c|}{\texttt{Phi}}\\
+& & \multicolumn{2}{c|}{\texttt{AVX}} & \multicolumn{2}{c|}{\texttt{LRBNi}}\\
 \cline{3-6}
@@
 \cline{1-6}
-\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
+%\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
 \multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
@@
 \label{table:perf-results-licm}
-\caption{Impact of generalized loop-invariant code motion ($licm$ column) on the Helmholtz, Diffusion and Burgers problems, for three kind of elements belonging to the Lagrange family (triangle, tetrahedron, prism), for the range of polynomial orders $p$ $\in$ $\lbrace$1, 4$\rbrace$.
+\caption{Impact of generalized loop-invariant code motion ($licm$ column) on the Helmholtz, Diffusion and Burgers problems, for three kind of elements belonging to the Lagrange family (triangle, tetrahedron, prism), for the range of polynomial orders $p \in \lbrace$1, 4$\rbrace$.
 Each entry indicates the range of speed-ups obtained over the non-optimized implementation.
@@
 Table~\ref{table:perf-results-licm} illustrates the speed-ups obtained on the Sandy Bridge and the Phi machines when $licm$, data alignment, and padding are used, over non-transformed code.
-Inspection of assembly code generated by $icc$ confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant with respect to outer loops are hoisted and, interestingly, not vectorized.
+We distinguish between $licm$ and $licm$-$ap$; the latter makes use of padding and data alignment.
+Inspection of assembly code generated by \texttt{icc} confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant with respect to outer loops are hoisted and, interestingly, not vectorized.
 This motivates the usually significant gain obtained.
 Padding and data alignment enhance, in general, the quality of SIMD auto-vectorization.
-As shown in Figure~\ref{algo:PyOP2Compiler}, these transformations are applied on top of $licm$, so we refer to them as $licm$-$ap$.
 Sometimes the run-time of $licm$-$ap$ is similar to that of $licm$ because the stiffness matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned.
@@
 in Burgers 3D p3).
-This is due to the large number of aligned temporaries introduced by $licm$, which probably leads to cache associativity conflicts.
+This is due to the large number of aligned temporaries introduced by $licm$, which probably induces cache associativity conflicts.
 
@@
 Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the speed-ups achieved by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmoltz and Diffusion kernels, respectively.
-For each problem instance we report two bars: one indicates the best run-time achieved by autotuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved by the XY's cost model.
+For each problem instance we report two bars: one indicates the best run-time achieved by autotuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved by the COFFEE's cost model.
 In general, there is no substantial difference between the two.
@@
 On the other hand, as soon as the iteration space becomes too big, vector-register tiles are essential to minimize pipeline stalls.
-A maximum speed-up of 1.40$\times$ over $licm$-$ap$ was observed in the Diffusion kernel when using a prismatic mesh and $p=4$.
+A maximum speed-up of 1.4$\times$ over $licm$-$ap$ was observed in the Diffusion kernel when using a prismatic mesh and $p=4$.
 Application of $op$-$vect$ to the Burgers problem induces meaningful slow downs, because of the large number of arrays that, generally, need to be tiled.
 Expression splitting can be used in combination to $op$-$vect$ to alleviate this issue; this is discussed in Section~\ref{sec:perf-results-split}.
-
-% why better results at higher order
-
-\subsection{Impact of Expression Splitting}
-\label{sec:perf-results-split}
-Expression splitting, or loop fission, contributes to relieve the register pressure in loop nests where a large number of arrays is accessed.
-If the stiffness matrix can be evaluated by accessing a few (relatively ) basis functions or temporary values introduced by $licm$,
-
-% why split useless in helmholtz/diffusion
-% why vect vect+split useless in burgers
 
@@
 
+\subsection{Impact of Expression Splitting}
+\label{sec:perf-results-split}
+Expression splitting, or loop fission, contributes to relieve register pressure.
+The transformation is useless if the stiffness matrix can be evaluated by accessing a relatively small number of basis functions (or temporary values introduced by $licm$), for which efficient register allocation can be already guaranteed.
+In the cases of Helmholtz and Diffusion, irrespective of split factor and cut, the only consequence of applying $split$ is to augment the number of increments for evaluating the stiffness matrix.
+On the Sandy Bridge machine, slow downs up to 1.4$\times$ were observed in these two problems.
+The cost model, however, prevents the adoption of this transformation, since the \texttt{while} statement at line 5 in Figure~\ref{algo:applyCostModel} is never entered.
+In the Burgers problem, $split$ allows notable speed-ups to be achieved over the best run-time provided by either $licm$ or $licm$-$ap$.
+Results are shown in Figure~\ref{fig:split-burgers-speedup}.
+In almost all problem configurations, a split factor of 1 induced optimal performance, meaning that the original expression was divided into two parts, with one part performing a slightly larger number of floating point operations than the other.
+Register locality could have been affected by requiring the same array be re-loaded twice as before splitting.
+However, in this experiment, this was only partly true, because most of the arrays appear in one sub-expression only.
+
+% why split useless in helmholtz/diffusion
+% why vect vect+split useless in burgers
+
 \begin{figure}[h]
@@
 CM stands for cost model.}
-\label{fig:opvect-burgers-speedup}
+\label{fig:split-burgers-speedup}
 \end{figure}
@@
 
-XY currently uses a simple cost model and heuristics to steer the optimization process.
+COFFEE currently uses a simple cost model and heuristics to steer the optimization process.
 Many code generators, like those based on the Polyhedral model~\cite{PLUTO} and those driven by domain-knowledge~\cite{modeldriven}, make use of cost models.
@@
 
-In this paper we have presented design, optimizations and systematic performance evaluation of XY, a compiler for finite element local assembly.
+In this paper we have presented design, optimizations and systematic performance evaluation of COFFEE, a compiler for finite element local assembly.
 In this context, to the best of our knowledge, this is the first compiler oriented towards the introduction of low-level optimizations to maximize instruction-level parallelism, register locality and SIMD vectorization.
@@
 The data space, in terms of number of matrices and scalar, grows proportionally to the complexity of the finite element problem.
-XY has been developed taking into account all of these aspects, knowing that some transformations are useful only in a subset of all possible problems.
+COFFEE has been developed taking into account all of these aspects, knowing that some transformations are useful only in a subset of all possible problems.
 The various optimizations overcome limitations of current vendor and research compilers.
@@
 Further work include a comprehensive study about feasibility and constraints of transforming the kernel into a sequence of BLAS calls.
-XY supports all of the problems expressible in Firedrake, and it is already integrated with this framework.
+COFFEE supports all of the problems expressible in Firedrake, and it is already integrated with this framework.
 

### 2014-04-01 paper/bare_conf.tex
a LOT of refactoring

@@
 
+%The finite element method is widely-employed to approximate solutions of partial differential equations.
+Local assembly is one of its steps, in which a problem-specific kernel is applied to all elements in the discretized domain of the equation.
+Since the domain size can be huge, executing efficient kernels is fundamental.
+Automating their optimization is, however, a challenging issue.
+Even if an affine loop nest can be generally distinguished, small problem size and complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations.
+In this context, we present COFFEE, a domain-specific compiler for local assembly kernels.
+COFFEE manipulates kernels’ abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality.
+It then generates C code including vector intrinsics.
+Systematic performance evaluation using a suite of real-world examples shows that speed-ups over non-optimized kernels up to 4.44$\times$ are achievable.
+
 \begin{abstract}
 %\boldmath
-The finite element method is widely-employed to determine approximated solutions of partial differential equations.
-Local assembly is one of its fundamental steps, in which a suitable numerical procedure, or kernel, is applied to all elements in the discretized domain of the equation.
-Since the size of the domain can be huge, a notable problem consists of producing efficient kernels to minimize the cost local assembly.
-In this context, we present design, implementation and systematic performance evaluation of COFFEE, a domain-specific compiler for local assembly kernels.
-COFFEE manipulates kernels' abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially in the form of SIMD vectorization, and register locality.
-It then generates C code, possibly including AVX or LRBNi intrinsics, which is eventually just-in-time compiled on the underlying architecture and executed.
-Performance evaluation using a suite of examples of real-world importance showed that speed-ups over non-optimized kernels between 1.32$\times$ and 4.44$\times$ can be achieved.
+The finite element method is widely-employed to approximate solutions of partial differential equations.
+Local assembly is one of its steps, in which a problem-specific kernel is applied to all elements in the discretized domain of the equation.
+Since the domain size can be huge, executing efficient kernels is fundamental.
+Automating their optimization is, however, a challenging issue.
+Even if an affine loop nest can be generally distinguished, small problem size and complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations.
+In this context, we present design and systematic evaluation of COFFEE, a domain-specific compiler for local assembly kernels.
+COFFEE manipulates abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality.
+It then generates C code including vector intrinsics.
+Performance evaluation using a suite of real-world examples shows that signifcant speed-ups up are achievable.
 \end{abstract}
@@
 \begin{IEEEkeywords}
-Finite element method, compilers, optimizations, simd vectorization 
+Finite element integration, local assembly, compilers, optimizations, simd vectorization 
 \end{IEEEkeywords}
@@
 To address this problem, domain-specific languages (DSLs) have been developed.
-The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2isc}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{OP2-hydra}.
+The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2isc}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}.
 
@@
 
-Due to the constraints of available compilers and on linear-algebra specialized libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For Finite ElemEnt local assembly.}, a compiler for optimizing local assembly kernels.
+Due to the constraints of available compilers and on linear-algebra specialized libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for optimizing local assembly kernels.
 COFFEE is integrated with Firedrake, a system for solving PDEs through the finite element method based on the OP2 abstraction~\cite{firedrake-code}.
@@
 The output is the evaluated element stiffness matrix.
-A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant bi-dimensional arrays, called ``basis functions'' (or derivatives of basis functions) in finite element terminology, initialized to problem-specific double-precision floatin point values; 3) evaluation of the stiffness matrix in an affine loop nest.
+A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant bi-dimensional arrays, called ``basis functions'' (or derivatives of basis functions) in finite element terminology, initialized to problem-specific double-precision floating point values; 3) evaluation of the stiffness matrix in an affine loop nest.
 All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (i.e.
@@
 Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions.
-It is worth noting there are circumnstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that its execution time becomes the dominant factor of the overall local assembly run-time.
+It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that its execution time becomes the dominant factor of the overall local assembly run-time.
 In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure).
 Given its inherent complexity, a comprehensive study of this problem is left as further work.
+
+In a local assembly kernel, although the bulk of the computation is carried out by the \texttt{ijk} affine loop nest, the typically small problem size and the complexity of the stiffness matrix evaluation make it hard to determine a single or unique sequence of successful transformations.
+Even though the memory access pattern is fixed, the fact that both data- and iteration-space sizes vary, possibly substantially, among different problems, render the optimization process challenging.
+Relatively to machine peak, while some problems achieve very good performance when applying a specific set of transformations, there is also an ample space of cases for which a more elaborated optimization strategy is required.
+For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, will suffer from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} will not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of data and iteration spaces.
+Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger.
+In this section, we provide a non-exhaustive list of transformations that local assembly kernels could benefit from.
+All of them are automated in COFFEE.
 
@@
 Splitting an expression has, in general, several drawbacks.
-Firstly, it increases the number of accesses to $A$ proportially to the ``split factor'', which is the number of sub-expressions produced.
+Firstly, it increases the number of accesses to $A$ proportionally to the ``split factor'', which is the number of sub-expressions produced.
 Secondly, depending on how the split is executed, it can lead to redundant computation (e.g.
@@
 \KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
-~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
-~~// K, det = Compute Jacobian (coords) \\
-~~// Declaration of basis function matrices \\
-~~\\
-~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
-~~~~double LI$\_$0[4] ALIGN;\\
-~~~~double LI$\_$1[4] ALIGN;\\
-~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
-~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
-~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
-~~~~$\rbrace$\\
+%~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
+~~~~// Same code as in Figure~\ref{code:helmholtz-licm} up to line 15 \\
+%~~// Declaration of basis function matrices \\
+%~~\\
+%~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
+%~~~~double LI$\_$0[4] ALIGN;\\
+%~~~~double LI$\_$1[4] ALIGN;\\
+%~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
+%~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
+%~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
+%~~~~$\rbrace$\\
 ~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
@@
 ~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
-~~$\rbrace$\\
 $\rbrace$
@@
 Loop unrolling  & unroll factor \\ \hline
-Register tiling & tile size \\ \hline
 Outer-product vectorization ($op$-$vect$) & tile size \\ \hline
-Assembly splitting ($split$) & split point, split factor \\ \hline
+Expression splitting ($split$) & split point, split factor \\ \hline
 \end{tabulary}
@@
 Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem.
-Autotuning might be used, although at the moment we avoid it to minimize the run-time overhead.
+Auto-tuning might be used, although at the moment we avoid it to minimize the run-time overhead.
 Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.
@@
 Based on a cost model, loop interchange, $split$, and $op$-$vect$ may be introduced.
-Their implementation is based on analysis and trasformation of the AST.
+Their implementation is based on analysis and transformation of the AST.
 When $op$-$vect$ is selected, the compiler outputs proper AVX (or LRBNi) intrinsics code.
@@
 The cost model is shown in Figure~\ref{algo:applyCostModel}.
-It takes into account the number of available logical vector registers ($n\_regs$) and the number of variables iterating along the $j$ and $k$ dimensions ($n\_consts$ for independent variables, $n\_outer\_arrays$ for $j$ variables, and $n\_inner\_arrays$ for $k$ variables, assuming the \texttt{ijk} loop order) to estimate unroll-and-jam and split factors when, respectively, $op$-$vect$ and $split$ are used.
+It takes into account the number of available logical vector registers, $n\_regs$, and the number of unique variables accessed: $n\_consts$ counts variables independent of both $j$ and $k$ loops, $n\_outer\_arrays$ counts $j$-dependent variables, and $n\_inner\_arrays$ counts $k$-dependent variables, assuming the \texttt{ijk} loop order.
+These are used to estimate unroll-and-jam and split factors when, respectively, $op$-$vect$ and $split$ are used.
 The $n\_consts$ parameter includes temporary registers to carry out computations, so setting its value is partly driven by heuristics.
@@
 A range of polynomial orders, from $p=1$ to $p=4$, were tested; higher polynomial orders are excluded from the study because of current Firedrake limitations.
-In such a large space of problems, the size of the stiffness matrix rarely excees 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
+In such a large space of problems, the size of the stiffness matrix rarely exceeds 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
 
@@
 \subsection{Impact of Vector-register Tiling}
+\label{sec:perf-eval-opvect}
 Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the speed-ups achieved by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmoltz and Diffusion kernels, respectively.
-For each problem instance we report two bars: one indicates the best run-time achieved by autotuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved by the COFFEE's cost model.
+For each problem instance we report two bars: one indicates the best run-time achieved by auto-tuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved by the COFFEE's cost model.
 In general, there is no substantial difference between the two.
@@
 A maximum speed-up of 1.4$\times$ over $licm$-$ap$ was observed in the Diffusion kernel when using a prismatic mesh and $p=4$.
-Application of $op$-$vect$ to the Burgers problem induces meaningful slow downs, because of the large number of arrays that, generally, need to be tiled.
+Application of $op$-$vect$ to the Burgers problem induces meaningful slow downs, because of the large number of temporary arrays that need to be tiled, which exceeds the number of available logical registers on the underlying machine.
 Expression splitting can be used in combination to $op$-$vect$ to alleviate this issue; this is discussed in Section~\ref{sec:perf-results-split}.
+
+We have also used the Intel Architecture Code Analyzer (IACA) tool~\cite{IACA} to demonstrate that the improved run-time is actually due to increased register re-use.
+By applying the tool to the Helmholtz problem on a tetrahedral mesh and $p=4$, it can be noticed that the number of clock cycles to execute one iteration of the $j$ loop decreases by roughly 17$\%$, on the Sandy Bridge, when using $op$-$vect$.
+This is because of less pressure on both data (cache) ports available in the processor.
 
@@
 
-% why split useless in helmholtz/diffusion
-% why vect vect+split useless in burgers
+The $split$ transformation was also tried in combination with $op$-$vect$ ($split$-$op$-$vect$), although the cost model prevents the application of the latter on the Sandy Bridge.
+Despite performance improvements up to 1.22$\times$, $split$-$op$-$vect$ never outperforms $split$.
+This is because of two reasons: for small split factors, like 1 and 2, the data space to be tiled is still too big, and register spilling still affects the run-time; for higher ones, sub-expressions become so small that, as explained in Section~\ref{sec:perf-eval-opvect}, full or partial unrolling of the loop nest allows achieving a certain degree of register re-use.
 
@@
 
+\subsection{Comparison with Fenics Form Compiler's built-in Optimizations}
+We have modified the Fenics Form Compiler (FFC) to make it return an abstract syntax tree representation of a local assembly kernel, rather than plain C code, so as to enable code transformations in COFFEE.
+Besides Firedrake, FFC is used in the FEniCS project~\cite{Fenics}.
+In FEniCS, FFC can apply its own model-driven optimizations to local assembly kernels~\cite{quadrature1}, which mainly consist of loop-invariant code motion and elimination, at code generation time, of floating point operations involving zero-valued columns in basis function arrays.
+The FEniCS' loop-invariant code motion is, however, definitely different from COFFEE's, since it is based on expansion of arithmetic operations to identify terms that are invariant of the whole loop nest.
+Thus, depending on the way expansion is performed, operation count does not necessarily decrease.
+On the other hand, elimination of zero-valued columns, which are due to certain properties of the input finite element problem, has the effect of introducing indirection arrays in the generated code.
+This kind of optimization is currently under development in COFFEE, although it will differ from FEniCS' by avoiding non-contiguous memory accesses, which affect vectorization because of the need for gather operations, at the price of removing only some of the zero-columns.
+
+Table~\ref{table:comparison-to-FFC-opt} summarizes the results of the comparison between FEniCS' (FFC) optimizations and COFFEE on the Sandy Bridge.
+Slow downs in Burgers occur in presence of small iteration spaces (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$), and generally much better than the maximum one reported in table, obtained with a triangular mesh and $p = 1$.
+This is a result of zero-columns removal in FEniCS: indirection arrays prevent auto-vectorization, which significantly impacts performance as soon as the stiffness matrix becomes bigger than 12$\times$12.
+However, with the forthcoming zero-removal optimization in COFFEE, we expect to outperform FEniCS even in these corner cases as well.
+In the cases of Helmholtz and Diffusion, the minimum speed-ups are, respectively, 1.10 and 1.18 (2D mesh, $p=1$), and they keep increasing with polynomial order and element shape up to the values illustrated in the table.
+
+\begin{table}[h]
+\begin{center}
+\begin{tabulary}{1.0\columnwidth}{|C|C|C|}
+\hline
+Problem & Max slow down & Max speed up \\\hline\hline
+Helmholtz & / & 11.72$\times$ \\
+Diffusion & / & 14.68$\times$ \\
+Burgers & 2.24$\times$ & 1.61$\times$ \\\hline
+\end{tabulary}
+\end{center}
+\label{table:comparison-to-FFC-opt}
+\caption{Summary of performance comparison between FEniCS and COFFEE on the Sandy Bridge machine.
+The slow down and speed up columns compare the best COFFEE implementation and the FEniCS implementation with optimizations enabled.}
+\end{table}
+
 \section{Related Work}
@@
 Many code generators, like those based on the Polyhedral model~\cite{PLUTO} and those driven by domain-knowledge~\cite{modeldriven}, make use of cost models.
-The other extreme consists of relying on auto-tuning to select the best implementation for a given problem on a certain platform.
+The opposite extreme consists of relying on auto-tuning to select the best implementation for a given problem on a certain platform.
 Notable examples include tuning of small matrix-matrix multiplies in nek5000~\cite{nek5000}, the ATLAS library~\cite{ATLAS}, and FFTW~\cite{FFTW} for fast fourier transforms.
@@
 The exploitation of domain-knowledge allows some of them to be particularly effective, as demonstrated by our experiments on two state-of-the-art Intel platforms.
-Further work include a comprehensive study about feasibility and constraints of transforming the kernel into a sequence of BLAS calls.
+Further work include a comprehensive study about feasibility and constraints on transforming the kernel into a sequence of BLAS calls.
 COFFEE supports all of the problems expressible in Firedrake, and it is already integrated with this framework.
@@
 This research is partly funded by the MAPDES project and by the Department of Computing at Imperial College London.
-The authors would like to thank Prof.
-J.
-Ramanujam for his invaluable suggestions, and Dr.
+The authors would like to thank Dr.
 Lawrence Mitchell and Dr.
-Francis Russell for their contribution to the PyOP2 project.
-
-%\newcommand{\massmatrixtwodresultsnorms}{
-%\begin{tabularx}{1.0\textwidth}{p{3.5cm}|p{1.98cm}p{1.98cm}p{1.98cm}|p{1.98cm}p{1.98cm}p{1.98cm}|}
-%\cline{2-7}
-%& \multicolumn{3}{c|}{\texttt{AVX}} & \multicolumn{3}{c|}{\texttt{PHI}} \\
-%\cline{2-7}
-%& \texttt{original} & \texttt{licm} & \texttt{licm+ap} & \texttt{original} & \texttt{licm} & \texttt{licm+ap}\\
-%\cline{1-7}
-%\multicolumn{1}{|c|}{\texttt{triangle p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-%\multicolumn{1}{|c|}{\texttt{triangle p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-%\multicolumn{1}{|c|}{\texttt{triangle p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-%\multicolumn{1}{|c|}{\texttt{triangle p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-%\cline{1-7}
-%\multicolumn{1}{|c|}{\texttt{tetrahedron p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-%\multicolumn{1}{|c|}{\texttt{tetrahedron p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-%\multicolumn{1}{|c|}{\texttt{tetrahedron p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-%\multicolumn{1}{|c|}{\texttt{tetrahedron p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-%\cline{1-7}
-%\multicolumn{1}{|c|}{\texttt{prism p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-%\multicolumn{1}{|c|}{\texttt{prism p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-%\multicolumn{1}{|c|}{\texttt{prism p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-%\multicolumn{1}{|c|}{\texttt{prism p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
-%\cline{1-7}
-%
-%\end{tabularx}
-%}
-%
-%\begin{table*}[t]\normalsize
-%\massmatrixtwodresultsnorms
-%\label{table:perf-results-licm-helmholtz}
-%\caption{Impact of generalized loop-invariant code motion ($licm$ column) on the Helmholtz problem, for three kind of elements belonging to the Lagrange family (triangle, tetrahedron, prism) and for the range of polynomial orders $p$ $\in$ $\lbrace$1, 4$\rbrace$.
-The column $licm$-$ap$ illustrates the combination of $licm$ with data alignment and padding.}
-%\end{table*}
-
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-%\bibliographystyle{IEEEtran}
-% argument is your BibTeX string definitions and bibliography database(s)
-%\bibliography{IEEEabrv,../bib/paper}
-%
-%% <OR> manually copy in the resultant .bbl file
-%% set second argument of \begin to the number of references
-%% (used to reserve space for the reference number labels box)
-%\begin{thebibliography}{1}
-%
-%\bibitem{IEEEhowto:kopka}
-%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
-%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.
-%
-%\end{thebibliography}
-%
+Francis Russell for their invaluable suggestions and their contribution to the Firedrake project.
 

### 2014-04-01 paper/bare_conf.tex
Some re-wording

@@
 Given their structure and the exceptionally small size, assembly kernels benefit from transformations like generalized loop-invariant code motion, vector-register tiling and code splitting, which are not supported by state-of-the-art polyhedral and vendor compilers.
-BLAS routines could be theoretically employed, although fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplications.
+BLAS routines could be theoretically employed, although fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies.
 BLAS libraries are also known to perform far from peak performance when the dimension of the matrices is small~\cite{nek5000}.
@@
 Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions.
-It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that its execution time becomes the dominant factor of the overall local assembly run-time.
+It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time become the dominant factor of the overall local assembly run-time.
 In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure).
@@
 
-%PyOP2 which lies at the core of Firedrake, where  is managed
-
 Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations.
@@
 Based on a cost model, loop interchange, $split$, and $op$-$vect$ may be introduced.
-Their implementation is based on analysis and transformation of the AST.
+Their implementation is based on analysis and transformation of the kernel's AST.
 When $op$-$vect$ is selected, the compiler outputs proper AVX (or LRBNi) intrinsics code.
@@
 These are used to estimate unroll-and-jam and split factors when, respectively, $op$-$vect$ and $split$ are used.
-The $n\_consts$ parameter includes temporary registers to carry out computations, so setting its value is partly driven by heuristics.
+The $n\_consts$ parameter includes temporary registers to carry out computations.
 If a factor is 0, then the corresponding transformation is not applied.
@@
 those not used for storing hoisted terms (line 9-11).
-Finally, the profitability of loop interchange is evaluated (line 12-16)
+Finally, the profitability of loop interchange is evaluated (line 12-16).
 
@@
 \subsection{Experimental Setup}
-
-TODO: This section needs some rewording! And still lacks the Phi discussion!! \\
 
@@
 
-For the Helmholtz 3D problem, manual implementations based on \texttt{Intel MKL BLAS} were tested on the Sandy Bridge.
-This particular kernel can be easily reduced to a sequence of four matrix-matrix multiplies that can be computed via calls to BLAS \texttt{dgemm}.
-In the case of $p=4$, where the stiffness matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only $licm$, data alignment and padding were used, with the slow down being even worse for smaller problem sizes.
-These experiments justify that there is a set of problems for which turning to MKL is not beneficial in terms of performance.
-It is possible that employing certain BLAS implementations or auto-tuning strategies for particular problem sizes is helpful, although the same problems afflicting $split$ (exposed in Section~\ref{sec:expr-split}) might limit the performance gain.
-However, as anticipated, extraction of matrix-matrix multiplies from analysis of the kernel's AST and a strategy for their execution will be comprehensively addressed in further work.
-
 \subsection{Impact of Generalized Loop-invariant Code Motion}
@@
 %\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{prism}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.05$\times$-1.68$\times$ & \multicolumn{1}{c|}{~1.32$\times$-4.14$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.36$\times$-2.64$\times$ & \multicolumn{1}{c|}{~1.35$\times$-3.32$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~2.16$\times$-2.45$\times$ & \multicolumn{1}{c|}{~2.42$\times$-2.63$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
 \cline{1-6}
-\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{triangle}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{prism}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.09$\times$-1.64$\times$ & \multicolumn{1}{c|}{~1.37$\times$-4.28$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.30$\times$-3.11$\times$ & \multicolumn{1}{c|}{~1.41$\times$-3.82$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~1.81$\times$-2.70$\times$ & \multicolumn{1}{c|}{~2.55$\times$-3.13$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
 \cline{1-6}
-\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{triangle}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{prism}} & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ & 1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.53$\times$-2.46$\times$ & \multicolumn{1}{c|}{~1.56$\times$-2.77$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.58$\times$-2.24$\times$ & \multicolumn{1}{c|}{~1.59$\times$-2.10$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~1.32$\times$-2.11$\times$ & \multicolumn{1}{c|}{~1.42$\times$-2.31$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
 \cline{1-6}
@@
 On the other hand, elimination of zero-valued columns, which are due to certain properties of the input finite element problem, has the effect of introducing indirection arrays in the generated code.
-This kind of optimization is currently under development in COFFEE, although it will differ from FEniCS' by avoiding non-contiguous memory accesses, which affect vectorization because of the need for gather operations, at the price of removing only some of the zero-columns.
+This kind of optimization is currently under development in COFFEE, although it will differ from FEniCS' by avoiding non-contiguous memory accesses, which affect vectorization because of the need for gather operations.
+The price to pay is that only some of the zero-columns will be removed.
 
 Table~\ref{table:comparison-to-FFC-opt} summarizes the results of the comparison between FEniCS' (FFC) optimizations and COFFEE on the Sandy Bridge.
-Slow downs in Burgers occur in presence of small iteration spaces (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$), and generally much better than the maximum one reported in table, obtained with a triangular mesh and $p = 1$.
+In COFFEE, Burgers' slow downs occur in presence of a small iteration space (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$), although they are generally milder than the maximum one reported in table, obtained with a triangular mesh and $p = 1$.
 This is a result of zero-columns removal in FEniCS: indirection arrays prevent auto-vectorization, which significantly impacts performance as soon as the stiffness matrix becomes bigger than 12$\times$12.
-However, with the forthcoming zero-removal optimization in COFFEE, we expect to outperform FEniCS even in these corner cases as well.
-In the cases of Helmholtz and Diffusion, the minimum speed-ups are, respectively, 1.10 and 1.18 (2D mesh, $p=1$), and they keep increasing with polynomial order and element shape up to the values illustrated in the table.
+However, with the forthcoming zero-removal optimization in COFFEE, we expect to outperform FEniCS even in these cases as well.
+In the cases of Helmholtz and Diffusion, the minimum speed-ups are, respectively, 1.10$\times$ and 1.18$\times$ (2D mesh, $p=1$), which keep increasing with polynomial order and element shape up to the values illustrated in the table.
 
@@
 
+\subsection{Comparison with hand-made BLAS-based implementations}
+For the Helmholtz 3D problem, manual implementations based on \texttt{Intel MKL BLAS} were tested on the Sandy Bridge.
+This particular kernel can be easily reduced to a sequence of four matrix-matrix multiplies that can be computed via calls to BLAS \texttt{dgemm}.
+In the case of $p=4$, where the stiffness matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only $licm$, data alignment and padding were used, with the slow down being even worse for smaller problem sizes.
+These experiments justify that there is a set of problems for which turning to MKL is not beneficial in terms of performance.
+It is possible that employing certain BLAS implementations or auto-tuning strategies for particular problem sizes is helpful, although the same problems afflicting $split$ (exposed in Section~\ref{sec:expr-split}) might limit the performance gain.
+However, as anticipated, extraction of matrix-matrix multiplies from analysis of the kernel's AST and a strategy for their execution will be comprehensively addressed in further work.
+
 \section{Related Work}
@@
 In~\cite{Markall20101815}, and more recently in~\cite{petsc-integration-gpu}, the problem has been studied for GPU architectures.
-In~\cite{assembly-opencl}, variants of the standard numerical integration algorithm have been specialized for the PowerXCell processor and evaluated; The paper lacks, however, an exhaustive study from the compiler viewpoint as we did, and none of the optimizations presented in Section~\ref{sec:code-transf} are mentioned.
+In~\cite{assembly-opencl}, variants of the standard numerical integration algorithm have been specialized for the PowerXCell processor and evaluated; the paper lacks, however, an exhaustive study from the compiler viewpoint as we did, and none of the optimizations presented in Section~\ref{sec:code-transf} are mentioned.
 
@@
 Their iteration space is usually very small, with the size depending on aspects like the degree of accuracy one wants to reach (polynomial order of the method) and the mesh discretization employed.
-The data space, in terms of number of matrices and scalar, grows proportionally to the complexity of the finite element problem.
-COFFEE has been developed taking into account all of these aspects, knowing that some transformations are useful only in a subset of all possible problems.
+The data space, in terms of number of matrices and scalar, grows proportionally with the complexity of the finite element problem.
+COFFEE has been developed taking into account all of these degrees of freedom, based on the idea that reducing the problem of local assembly optimization to a fixed sequence of transformations is simply too superficial if close-to-peak performance needs to be reached.
 The various optimizations overcome limitations of current vendor and research compilers.

### 2014-04-03 paper/bare_conf.tex
Refactor a few sections (msg of the paper targeted)

@@
 Gheorghe-Teodor Bercea\IEEEauthorrefmark{1}, \\
-David Ham\IEEEauthorrefmark{1},
+David A. Ham\IEEEauthorrefmark{4},
 J. Ramanujam\IEEEauthorrefmark{3},
@@
 \IEEEauthorblockA{\IEEEauthorrefmark{1}Imperial College London, Department of Computing,\\
+\IEEEauthorrefmark{4}Imperial College London, Department of Computing and Department of Mathematics,\\
 \IEEEauthorrefmark{2}TU Delft, Faculty of Engineering, Mathematics and Computer Science (EWI)\\
@@
 \maketitle
-
+%OLD 1
 %The finite element method is widely-employed to approximate solutions of partial differential equations.
@@
 
-\begin{abstract}
-%\boldmath
-The finite element method is widely-employed to approximate solutions of partial differential equations.
+%OLD 2 before feedback
+%The finite element method is widely-employed to approximate solutions of partial differential equations.
 Local assembly is one of its steps, in which a problem-specific kernel is applied to all elements in the discretized domain of the equation.
@@
 Performance evaluation using a suite of real-world examples shows that signifcant speed-ups up are achievable.
+
+\begin{abstract}
+%\boldmath
+Numerical solution of partial differential equations using the finite element method is one of the key applications of high performance computing.
+The local assembly is its characteristic operation, which executes a problem-specific kernel for each element in the discretized problem domain.
+Since the domain size can be huge, executing efficient kernels is fundamental.
+Their optimization is, however, a challenging issue.
+Even though affine loop nests are generally present, the short trip counts and the complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations.
+We present, therefore, the design and systematic evaluation of COFFEE, a domain-specific compiler for local assembly kernels.
+COFFEE manipulates abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality.
+It then generates C code including vector intrinsics.
+Experiments using a suite of real-world examples show that significant performance improvement is achieved.
 \end{abstract}
@@
 Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective.
-The solution is sought in each cell of the discretized domain by applying suitable numerical kernels.
-As the number of cells can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days.
+The solution is sought in each element of the discretized domain by applying suitable numerical kernels.
+As the number of elements can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days.
 To address this problem, domain-specific languages (DSLs) have been developed.
@@
 However, we do not consider the latter case.
-An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 30, with a minimum value of 3, depending on the order of the method).
-With such small kernels, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.
-
-Optimization of assembly kernels is non-trivial.
-Given their structure and the exceptionally small size, assembly kernels benefit from transformations like generalized loop-invariant code motion, vector-register tiling and code splitting, which are not supported by state-of-the-art polyhedral and vendor compilers.
-BLAS routines could be theoretically employed, although fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies.
-BLAS libraries are also known to perform far from peak performance when the dimension of the matrices is small~\cite{nek5000}.
-As detailed in Section~\ref{sec:perf-results}, hand-made BLAS implementations of the Helmholtz assembly kernel (illustrated later) have run-times worse than those achieved with our optimization strategy.
-
-%(our attempts to use PoCC's optimizations~\cite{PoCC}, especially unrolling and register tiling, did not improve performance over $icc$ compilation)
-
-%Also, tiling at the level of vector register, achievable through a non-affine transformation of the iteration space, has showed a significant performance improvement in a range of problems. .
+An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small: their trip count rarely exceeds 30, whereas the minimum value is 3.
+In the innermost loop, a problem-specific expression evaluates a bi-dimensional array, representing the result of local assembly in an element of the discretized domain.
+With such a kernel structure, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.
+
+Achieving high-performance is non-trivial.
+The complexity of the mathematical expressions, often characterized by a large number of operations on constants and small matrices, makes it hard to determine a single or specific sequence of transformations that is successfully applicable to all problems.
+The variation in loop trip counts further exacerbates the issue.
+A compiler-based approach is, therefore, the only reasonable option to reach close-to-peak performance.
+Optimizations like generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but they are not supported by state-of-the-art polyhedral and vendor compilers.
+BLAS routines could be theoretically employed, although a fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies.
+In addition, as detailed in Section~\ref{sec:perf-results}, in which manual BLAS-based implementations are evaluated, the small dimension of the involved matrices and loss in data locality can limit the performance gain, if any.
 
@@
 \item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
-\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements.
+\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel machines, a Sandy Bridge and the Phi.
 \end{itemize}
@@
 Note that in addition to a larger number of operations to compute the stiffness matrix, the Burgers case shows a register pressure higher than that in Helmholtz.
-Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits, which can be exploited for effective code transformations and SIMD vectorization.
-
-The class of kernels we are considering has, in particular, some peculiarities.
-1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest.
-This is not true in general, since bent elements might be used in the unstructured mesh, which would require the Jacobian be re-computed at every $i$ iteration; 2) memory accesses along the three loop dimensions are always stride-1; 3) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g.
-the $F$ values in Burgers) and introduction of temporary arrays; 4) the $j$ and $k$ loops iterate over the same iteration space; 5) most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
+Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.
+
+Some peculiarities of assembly kernels are: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g.
+the $F$ values in Burgers) and introduction of temporary arrays; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
 In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.
@@
 
-In a local assembly kernel, although the bulk of the computation is carried out by the \texttt{ijk} affine loop nest, the typically small problem size and the complexity of the stiffness matrix evaluation make it hard to determine a single or unique sequence of successful transformations.
-Even though the memory access pattern is fixed, the fact that both data- and iteration-space sizes vary, possibly substantially, among different problems, render the optimization process challenging.
-Relatively to machine peak, while some problems achieve very good performance when applying a specific set of transformations, there is also an ample space of cases for which a more elaborated optimization strategy is required.
-For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, will suffer from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} will not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of data and iteration spaces.
+As already emphasized, the structure of mathematical expressions evaluating the stiffness matrix and the variation in loop trip counts, although typically limited to the order of tens iterations, render the optimization process challenging.
+Relatively to machine peak, while some problems achieve good performance when applying a very specific set of transformations, there is also an ample space of cases for which a more elaborated optimization strategy is required.
+For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces.
 Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger.
-In this section, we provide a non-exhaustive list of transformations that local assembly kernels could benefit from.
-All of them are automated in COFFEE.
+In this context, our goal is to understand the relationship between distinct code transformations, their impact on local assembly kernels, and to what extent their composability is effective in a class of problems and architectures.
 

### 2014-04-03 paper/bare_conf.tex
Minor rewording of abstract

@@
 \author{\IEEEauthorblockN{Fabio Luporini\IEEEauthorrefmark{1},
+Ana Lucia Varbanescu\IEEEauthorrefmark{2},
 Florian Rathgeber\IEEEauthorrefmark{1},
-Ana Lucia Varbanescu\IEEEauthorrefmark{2}, 
 Gheorghe-Teodor Bercea\IEEEauthorrefmark{1}, \\
+J. Ramanujam\IEEEauthorrefmark{3},
 David A. Ham\IEEEauthorrefmark{4},
-J. Ramanujam\IEEEauthorrefmark{3},
 and
@@
 \IEEEauthorrefmark{2}TU Delft, Faculty of Engineering, Mathematics and Computer Science (EWI)\\
-\IEEEauthorrefmark{1}\{f.luporini12$|$f.rathgeber10$|$gheorghe-teodor.bercea08$|$p.kelly$|$d.ham\}@imperial.ac.uk,\\
+\IEEEauthorrefmark{1}\{f.luporini12$|$f.rathgeber10$|$gheorghe-teodor.bercea08$|$p.kelly$|$david.ham\}@imperial.ac.uk,\\
 \IEEEauthorrefmark{2}a.l.varbanescu@uva.nl,
@@
 %\boldmath
-Numerical solution of partial differential equations using the finite element method is one of the key applications of high performance computing.
-The local assembly is its characteristic operation, which executes a problem-specific kernel for each element in the discretized problem domain.
+The numerical solution of partial differential equations using the finite element method is a key application in scientifc computing.
+Local assembly is its characteristic operation, which executes a problem-specific kernel for each element in the discretized problem domain.
 Since the domain size can be huge, executing efficient kernels is fundamental.
@@
 Even though affine loop nests are generally present, the short trip counts and the complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations.
-We present, therefore, the design and systematic evaluation of COFFEE, a domain-specific compiler for local assembly kernels.
+We present the design and systematic evaluation of COFFEE, a domain-specific compiler for local assembly kernels.
 COFFEE manipulates abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality.
@@
 The complexity of the mathematical expressions, often characterized by a large number of operations on constants and small matrices, makes it hard to determine a single or specific sequence of transformations that is successfully applicable to all problems.
-The variation in loop trip counts further exacerbates the issue.
-A compiler-based approach is, therefore, the only reasonable option to reach close-to-peak performance.
+The variation in loop trip counts and their typically small value, which prevents the adoption of standard techniques for cache locality, further exacerbates the issue.
+A compiler-based approach is, therefore, the only reasonable option to obtain close-to-peak performance in a wide range of different local assembly kernels.
 Optimizations like generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but they are not supported by state-of-the-art polyhedral and vendor compilers.
@@
 where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels.
-The contribution of this paper is threefold
+The contributions of this paper are threefold
 \begin{itemize}
 \item An optimisation strategy for a class of kernels widely-used in scientific applications, namely local assembly in the context of the finite element method.
-Our approach exploits domain knowledge to go beyond the limits of both vendor (e.g.
-$icc$) and research (e.g.
-polyhedral) compilers.
+Our approach, which also exploits domain knowledge, goes beyond the limits of both vendor and research compilers.
 \item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
@@
 Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions.
-It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time become the dominant factor of the overall local assembly run-time.
+It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time.
 In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure).
@@
 First, they can identify sub-expressions that are invariant with respect to the innermost loop only.
-This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted in the default loop order \texttt{ijk}.
+This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted in the loop order \texttt{ijk}.
 Second, the hoisted code is scalar, i.e.
@@
 Based on the observation that the evaluation of the stiffness matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented.
-If we consider the code snippet in Figure~\ref{code:helmholtz-licm} (Helmholtz after loop-invariant code motion) and we ignore the presence of \texttt{det*W3[i]}, we can notice that the computation of the stiffness matrix is abstractly expressible as
+If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, we can notice that the computation of the stiffness matrix is abstractly expressible as
 \begin{equation}
@@
 \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets.
-Regardless of the specific input problem, and by abstracting from the presence of all variables independent of both $j$ and $k$, the stiffness matrix computation is always reducible to this kind of form.
+Regardless of the specific input problem, and by abstracting from the presence of all variables independent of both $j$ and $k$, the stiffness matrix computation is always reducible to this form.
 Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$.
@@
 ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation.
-Besides a larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when using unroll-and-jam, with $f$ being the unroll-and-jam factor.
+Besides a larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.
 
@@
 
-The storage layout of $A$, however, is incorrect after the application of this ``outer-product vectorization'' ($op$-$vect$, in the following).
-We efficiently restore it with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the \texttt{ijk} loop nest.
+The storage layout of $A$, however, is incorrect after the application of this outer-product-based vectorization ($op$-$vect$, in the following).
+It can be efficiently restored with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the \texttt{ijk} loop nest.
 The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.
@@
 Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations.
-This high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
+At run-time, this high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
 ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}.
@@
 Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem.
-Auto-tuning might be used, although at the moment we avoid it to minimize the run-time overhead.
+Auto-tuning might be used, although at the moment we avoid it to minimize the run-time overhead (the generation of the AST occurs at run-time as soon as problem-specific data are available).
 Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.
@@
 If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied.
-This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{inter-kernel-vect}.
+This feature, currently under development, has been proved to be useful in several finite volume applications~\cite{inter-kernel-vect}.
 The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{pyop2isc}.
@@
 Their implementation is based on analysis and transformation of the kernel's AST.
-When $op$-$vect$ is selected, the compiler outputs proper AVX (or LRBNi) intrinsics code.
+When $op$-$vect$ is selected, the compiler outputs proper AVX or LRBNi intrinsics code.
 Any possible corner cases are handled: for example, if $op$-$vect$ is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.
@@
 The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension.
-Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers are available to carry out the computation of the stiffness matrix.
+Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers remain available to carry out the computation of the stiffness matrix.
 
@@
 The cost model is shown in Figure~\ref{algo:applyCostModel}.
-It takes into account the number of available logical vector registers, $n\_regs$, and the number of unique variables accessed: $n\_consts$ counts variables independent of both $j$ and $k$ loops, $n\_outer\_arrays$ counts $j$-dependent variables, and $n\_inner\_arrays$ counts $k$-dependent variables, assuming the \texttt{ijk} loop order.
-These are used to estimate unroll-and-jam and split factors when, respectively, $op$-$vect$ and $split$ are used.
-The $n\_consts$ parameter includes temporary registers to carry out computations.
+It takes into account the number of available logical vector registers, $n\_regs$, and the number of unique variables accessed: $n\_consts$ counts variables independent of both $j$ and $k$ loops and temporary registers, $n\_outer\_arrays$ counts $j$-dependent variables, and $n\_inner\_arrays$ counts $k$-dependent variables, assuming the \texttt{ijk} loop order.
+These values are used to estimate unroll-and-jam and split factors for $op$-$vect$ and $split$.
 If a factor is 0, then the corresponding transformation is not applied.
@@
 Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size.
-The unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of the available registers, i.e.
+The unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of the available logical registers, i.e.
 those not used for storing hoisted terms (line 9-11).
@@
 On the other hand, elimination of zero-valued columns, which are due to certain properties of the input finite element problem, has the effect of introducing indirection arrays in the generated code.
-This kind of optimization is currently under development in COFFEE, although it will differ from FEniCS' by avoiding non-contiguous memory accesses, which affect vectorization because of the need for gather operations.
+This kind of optimization is currently under development in COFFEE, although it will differ from FEniCS' by avoiding non-contiguous memory accesses, which would affect vectorization because of the need for gather operations.
 The price to pay is that only some of the zero-columns will be removed.
@@
 The data space, in terms of number of matrices and scalar, grows proportionally with the complexity of the finite element problem.
-COFFEE has been developed taking into account all of these degrees of freedom, based on the idea that reducing the problem of local assembly optimization to a fixed sequence of transformations is simply too superficial if close-to-peak performance needs to be reached.
+COFFEE has been developed taking into account all of these degrees of freedom, based on the idea that reducing the problem of local assembly optimization to a fixed sequence of transformations is way too superficial if close-to-peak performance needs to be reached.
 The various optimizations overcome limitations of current vendor and research compilers.

### 2014-04-03 paper/bare_conf.tex
Some rewording of intro section

@@
 \section{Introduction}
-In many fields, like computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by means of partial differential equations (PDEs).
-Numerical techniques, like finite volume method and finite element method, are widely-employed to approximate solutions of PDEs.
-Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective.
+In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modeled by means of partial differential equations (PDEs).
+Numerical techniques, such as the finite volume and finite element method, are widely employed to approximate solutions of PDEs.
+Unstructured meshes are often used to discretize the computational domain, since their geometric flexibility allows adapting the mesh to the required resolution.
 The solution is sought in each element of the discretized domain by applying suitable numerical kernels.
-As the number of elements can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days.
+As the number of elements can be of the order of millions, the time required to execute the computation can be hours or days.
 To address this problem, domain-specific languages (DSLs) have been developed.
-The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{pyop2isc}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}.
-
-OP2 adopts a kernel-oriented programming model, in which the computation semantics is expressed through self-contained functions, or ``kernels''.
-A kernel is applied to all elements in a set of mesh components (e.g.
-edges, vertices, cells/elements), with an implicit synchronization between the application of two consecutive kernels.
+The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design to OP2~\cite{pyop2isc} demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}.
+Hydra uses the finite volume method, consists of roughly 50000 lines of code and is able to handle mesh sizes of over 100 millions edges.
+
+% FIXME: that citation is wrong! You're talking about OP2. You MUST maintain that distinction
+
+OP2 adopts a programming model, in which computations are expressed through self-contained functions, referred to as ``kernels''.
+A kernel is applied to all elements in a set of mesh entities such as edges, vertices, or cells, with an implicit synchronization between the application of two consecutive kernels.
 On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread.
@@
 
-We focus on local assembly (``assembly'', in the following), a fundamental step of a finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$.
+% FIXME: citation for the 30-60% figure?
+
+We focus on local assembly (``assembly'' in the following), a fundamental step of the finite element method that covers an important fraction of the overall computation time, often in the range of 30$\%$-60$\%$.
 During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain.
@@
 However, we do not consider the latter case.
-An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small: their trip count rarely exceeds 30, whereas the minimum value is 3.
-In the innermost loop, a problem-specific expression evaluates a bi-dimensional array, representing the result of local assembly in an element of the discretized domain.
-With such a kernel structure, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.
+An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for low-order methods.
+In the innermost loop, a problem-specific expression evaluates a two dimensional array, representing the result of local assembly in an element of the discretized domain.
+With such a kernel structure, we focus on aspects such as minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.
 
@@
 
-Due to the constraints of available compilers and on linear-algebra specialized libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for optimizing local assembly kernels.
-COFFEE is integrated with Firedrake, a system for solving PDEs through the finite element method based on the OP2 abstraction~\cite{firedrake-code}.
+Due to the constraints of available compilers and specialized linear-algebra libraries, we have automated a set of generic and model-driven code transformations in the COmpiler For FinitE Element local assembly (COFFEE), a compiler for optimizing local assembly kernels.
+COFFEE is integrated with Firedrake~\cite{firedrake-code}, a system for solving PDEs through the finite element method based on the PyOP2 abstraction~\cite{pyop2isc,pyop2ws}.
 It supports any problems expressible with this framework.
@@
 
+% FIXME: You're talking about FFC-generated code really
+
 Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal.
-Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (i.e.
-where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels.
-The contributions of this paper are threefold
+Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (that is code where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels.
+The contributions of this paper are threefold:
 \begin{itemize}
-\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely local assembly in the context of the finite element method.
-Our approach, which also exploits domain knowledge, goes beyond the limits of both vendor and research compilers.
+\item An optimization strategy for finite element local assembly, a class of kernels widely-used in scientific applications.
+Our approach exploits domain knowledge and goes beyond the limits of both vendor and research compilers.
 \item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
-\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel machines, a Sandy Bridge and the Phi.
+\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
 \end{itemize}
 
-The paper is organized as follows.
-In Section~\ref{sec:background} we provide some background on local assembly, showing code generated by Firedrake and emphasizing critical computational aspects.
-Section~\ref{sec:code-transf} describes in detail the various code transformations, highlighting when and how domain-knowledge has been exploited.
+This paper is organized as follows: In Section~\ref{sec:background} we provide some background on local assembly, showing code generated by Firedrake and emphasizing important computational aspects.
+Section~\ref{sec:code-transf} describes the various code transformations in detail, highlighting when and how domain knowledge has been exploited.
 The design and implementation of our compiler is discussed in Section~\ref{sec:pyop2-compiler}.

### 2014-04-03 paper/bare_conf.tex
Minor fixes

--- 2014-04-03 paper/bare_conf.tex
@@
 %\boldmath
-The numerical solution of partial differential equations using the finite element method is a key application in scientifc computing.
-Local assembly is its characteristic operation, which executes a problem-specific kernel for each element in the discretized problem domain.
+Numerical solution of partial differential equations using the finite element method is one of the key applications of high performance computing.
+The local assembly is its characteristic operation, which executes a problem-specific kernel for each element in the discretized problem domain.
 Since the domain size can be huge, executing efficient kernels is fundamental.
@@
 Even though affine loop nests are generally present, the short trip counts and the complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations.
-We present the design and systematic evaluation of COFFEE, a domain-specific compiler for local assembly kernels.
+We present, therefore, the design and systematic evaluation of COFFEE, a domain-specific compiler for local assembly kernels.
 COFFEE manipulates abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality.

### 2014-04-05 paper/bare_conf.tex
Strip trailing whitespace
Some rewording of background section

--- 2014-04-03 paper/bare_conf.tex
@@
 \small
-\KwIn{stiffness matrix A (initialized to 0),\\~~~~~~~~~element coordinates in the domain (array),\\~~~~~~~~~(optional) fields (e.g. velocity, array)}
-\KwOut{stiffness matrix A (computed)}
-\nlset{1}~~- Jacobian computation, using element coordinates \\
-~~- Declaration of constant data, among which basis function\\
-~~- Computation of the stiffness matrix
+\KwIn{coordinate field (array), coefficient fields (e.g. velocity, array)}
+\KwOut{element matrix A (2D array)}
+\nlset{1}Compute Jacobian from coordinates \\
+Declare of constant data: basis functions and derivatives\\
+Compute element matrix using numerical quadrature
 \caption{General structure of a local assembly kernel generated by Firedrake.}
@@
 
-Local assembly consists of evaluating so called element stiffness matrix (``stiffness matrix'', in the following) and element stiffness vector; in this work, we focus on computation of stiffness matrices, which is usually the costly part of the process.
-A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell (or element) of the discretized domain.
-Numerical integration algorithms are widely-used in assembly codes to evaluate stiffness matrices~\cite{quadrature1, fluidity_manual_v4}.
+% FIXME: stiffness matrix is only used for certain types of elliptic PDEs
+% FIXME: the fluidity manual isn't a good source for this, cite e.g. Brenner and Scott
+
+Local assembly is the computation of local contributions of a specific cell of the discretized domain to the global sparse matrix or right-hand side vector of the linear system.
+The process consists of numerically evaluating an integral to produce a small dense element matrix or an element vector~\cite{quadrature1, fluidity_manual_v4}.
+In this work we focus on local matrices, which are usually more costly to compute.
+
+% FIXME: you can't attribute the kernel code generation to Firedrake: mention FFC
 
 Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm.
-It then triggers compilation of such kernels using an available vendor compiler, and eventually manages parallel execution over the mesh.
+These kernels are compiled at runtime using a vendor compiler, and Firedrake manages the parallel execution over the mesh.
 The subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.
@@
 The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}.
-Input are a pointer to a 0-valued bi-dimensional array, which will be used to store the stiffness matrix; a pointer to the element's coordinates in the discretized domain; one or more pointers to so called fields, for instance indicating value of velocity or pressure in the element.
-The output is the evaluated element stiffness matrix.
-A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant bi-dimensional arrays, called ``basis functions'' (or derivatives of basis functions) in finite element terminology, initialized to problem-specific double-precision floating point values; 3) evaluation of the stiffness matrix in an affine loop nest.
-All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (i.e.
-using \texttt{static const} in C language).
-Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel's objects.
+Inputs are a pointer to a zero initialized two dimensional array used to store the element matrix, a pointer to the element's coordinates in the discretized domain, and pointers to data of coefficient fields, for instance indicating value of velocity or pressure in the element.
+The output is the evaluated element matrix.
+A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant two dimensional arrays for the finite element basis functions and their derivatives (if needed), initialized to problem-specific double-precision floating point values; 3) evaluation of the element matrix in an affine loop nest.
+All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (using \texttt{static const} in C).
+Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to various objects in the kernel.
 
@@
 Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ \hline
-Stiffness matrix & double[][] & A\\ \hline
+Element matrix & double[][] & A\\ \hline
 \end{tabulary}
@@
 
+% FIXME: You haven't really defined basis functions, so what is a Lagrange basis function?
+
 The actual complexity of a local assembly kernel depends on the finite element problem being solved.
-In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products.
+In simpler cases, the loop nest is perfect: it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products.
 An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$.
-In other scenarios, for instance when solving a non-linear problem like Burgers as in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the stiffness matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
+In other scenarios, for instance when solving a non-linear problem like Burgers as shown in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
 Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse.
-Note that in addition to a larger number of operations to compute the stiffness matrix, the Burgers case shows a register pressure higher than that in Helmholtz.
+Note that in addition to a larger number of operations to compute the element matrix, the Burgers case shows a register pressure higher than that in Helmholtz.
 Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.
 
-Some peculiarities of assembly kernels are: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g.
-the $F$ values in Burgers) and introduction of temporary arrays; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
+Some characteristics of assembly kernels are: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g.
+the $F$ values in Burgers) and introduction of temporary arrays; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
 In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

### 2014-04-05 paper/bare_conf.tex
Some work on code transformations section

@@
 The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake.
-One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place.
+One particular characteristic is that they all aim at improving the run time of the \texttt{ijk} loop nest, where the numerical integration takes place.
 In rare, yet important cases, however, this might not be sufficient to achieve notable speed-ups.
 Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions.
-It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time.
+It is worth noting that there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time.
 In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure).
@@
 
-As already emphasized, the structure of mathematical expressions evaluating the stiffness matrix and the variation in loop trip counts, although typically limited to the order of tens iterations, render the optimization process challenging.
-Relatively to machine peak, while some problems achieve good performance when applying a very specific set of transformations, there is also an ample space of cases for which a more elaborated optimization strategy is required.
+As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens of iterations, render the optimization process challenging.
+While some problems achieve good performance relative to machine peak when applying a very specific set of transformations, there is another space of cases for which a more elaborate optimization strategy is required.
 For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces.
-Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger.
+Furthermore, domain knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger.
 In this context, our goal is to understand the relationship between distinct code transformations, their impact on local assembly kernels, and to what extent their composability is effective in a class of problems and architectures.
@@
 \subsection{Padding and Data Alignment}
-Auto-vectorization of assembly code computing the stiffness matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length $vl$.
+Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length $vl$.
 Data alignment is enforced in two steps.
 Initially, all arrays are allocated to addresses that are multiples of $vl$.
-Then, bi-dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$.
+Then, two dimensional arrays are padded to a number of columns equal to the nearest multiple of $vl$.
 For instance, assume the original size of a basis function array is 3$\times$3 and that the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values.
@@
 The compiler is explicitly informed about data alignment using a suitable pragma.
-Padding of all bi-dimensional arrays involved in the evaluation of the stiffness matrix also allows us to safely round the loop trip count to the nearest multiple of $vl$.
+Padding of all two dimensional arrays involved in the evaluation of the element matrix also allows us to safely round the loop trip count to the nearest multiple of $vl$.
 This avoids the introduction of a remainder (scalar) loop from the compiler, which would render vectorization less efficient.
@@
 Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels.
-We want to overcome two deficiencies that both $intel$ and $gnu$ compilers have.
+We want to overcome two deficiencies that both Intel and GNU compilers have.
 First, they can identify sub-expressions that are invariant with respect to the innermost loop only.
@@
 This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest.
-At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop.
+At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, since they are part of an inner loop.
 Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant.
@@
 Tiling at the level of vector registers is an additional feature of COFFEE.
-Based on the observation that the evaluation of the stiffness matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented.
-If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, we can notice that the computation of the stiffness matrix is abstractly expressible as
+Based on the observation that the evaluation of the element matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented.
+If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, we can notice that the computation of the element matrix is abstractly expressible as
 \begin{equation}
@@
 \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets.
-Regardless of the specific input problem, and by abstracting from the presence of all variables independent of both $j$ and $k$, the stiffness matrix computation is always reducible to this form.
-Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$.
+Regardless of the specific input problem, and by abstracting from the presence of all variables independent of both $j$ and $k$, the element matrix computation is always reducible to this form.
+Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$.
 Values in a register are shuffled each time a product is performed.
-Standard compiler auto-vectorization ($gnu$ and $intel$), instead, executes 4 broadcast operations (i.e.
-``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation.
+Standard compiler auto-vectorization (GNU and Intel), instead, executes 4 broadcast operations (the ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation.
 Besides a larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.
@@
 
-The storage layout of $A$, however, is incorrect after the application of this outer-product-based vectorization ($op$-$vect$, in the following).
+The storage layout of $A$, however, is incorrect after the application of this outer-product-based vectorization (\emph{op-vect}, in the following).
 It can be efficiently restored with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the \texttt{ijk} loop nest.
-The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.
+The generated pseudo-code for the simple Helmholtz problem when using \emph{op-vect} is shown in Figure~\ref{code:helmholtz-opvect}.
 
@@
 \centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}}
-\caption{Restoring the storage layout after $op$-$vect$.
-The figure shows how 4$\times$4 elements in the top-left block of the stiffness matrix $A$ can be moved to their correct positions.
+\caption{Restoring the storage layout after \emph{op-vect}.
+The figure shows how 4$\times$4 elements in the top-left block of the element matrix $A$ can be moved to their correct positions.
 Each rotation, represented by a group of three same-colored arrows, is implemented by a single vector shuffle intrinsics.}
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake when padding, data alignment, $licm$ and $op$-$vect$ are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}.
-Here, we assume the polynomial order is $p=2$, since $op$-$vect$ can not be used when an iteration space dimension is smaller than the vector length.
+\caption{Local assembly code generated by Firedrake when padding, data alignment, \emph{licm} and \emph{op-vect} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}.
+Here, we assume the polynomial order is $p=2$, since \emph{op-vect} can not be used when an iteration space dimension is smaller than the vector length.
 The original size of the $j$-$k$ iteration space (i.e.
@@
 In complex kernels, like Burgers' in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
-If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, it is likely to obtain poor register reuse.
+If the number of variables independent of the innermost loop dimension is close to or greater than the number of available CPU registers, it is likely to obtain poor register reuse.
 This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large.
@@
 Firstly, it increases the number of accesses to $A$ proportionally to the ``split factor'', which is the number of sub-expressions produced.
-Secondly, depending on how the split is executed, it can lead to redundant computation (e.g.
-the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}).
-Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
+Secondly, depending on how the split is executed, it can lead to redundant computation (for example, the product $det*W3[i]$ is performed for each of the sub-expressions in the code of Figure~\ref{code:helmholtz-split}).
+Finally, it might affect register locality, although this is not the case for the Helmholtz example: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
 Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake when $split$ is applied to the optimized Helmholtz problem given in Figure~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
+\caption{Local assembly code generated by Firedrake when \emph{split} is applied to the optimized Helmholtz problem given in Figure~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
 \label{code:helmholtz-split}
@@
 Name (Abbreviation) & Parameter \\\hline\hline
-Generalized loop-invariant code motion ($licm$) &   \\ \hline
+Generalized loop-invariant code motion (\emph{licm}) &   \\ \hline
 Padding &  \\ \hline
@@
 Loop unrolling  & unroll factor \\ \hline
-Outer-product vectorization ($op$-$vect$) & tile size \\ \hline
-Expression splitting ($split$) & split point, split factor \\ \hline
+Outer-product vectorization (\emph{op-vect}) & tile size \\ \hline
+Expression splitting (\emph{split}) & split point, split factor \\ \hline
 \end{tabulary}

### 2014-04-05 paper/bare_conf.tex
Some work on coffee overview section

@@
 
+% FIXME: Firedrake didn't invent UFL, need to credit FEniCS here
+% FIXME: we don't really ``avoid'' autotuning, we haven't implemented it
+
 Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations.
-At run-time, this high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
+At run time, this high-level specification is translated by the FEniCS Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
 ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}.
@@
 Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem.
-Auto-tuning might be used, although at the moment we avoid it to minimize the run-time overhead (the generation of the AST occurs at run-time as soon as problem-specific data are available).
+Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of the AST occurs at run-time as soon as problem-specific data are available.
 Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.
@@
 This feature, currently under development, has been proved to be useful in several finite volume applications~\cite{inter-kernel-vect}.
-The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{pyop2isc}.
+The second transformation step is applied if the backend is a manycore architecture, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{pyop2isc}.
 Then, an ordered sequence of optimization steps are executed.
-Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays.
-Based on a cost model, loop interchange, $split$, and $op$-$vect$ may be introduced.
+Application of \emph{licm} must precede padding and data alignment, due to the introduction of temporary arrays.
+Based on a cost model, loop interchange, \emph{split}, and \emph{op-vect} may be introduced.
 Their implementation is based on analysis and transformation of the kernel's AST.
-When $op$-$vect$ is selected, the compiler outputs proper AVX or LRBNi intrinsics code.
-Any possible corner cases are handled: for example, if $op$-$vect$ is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.
+When \emph{op-vect} is selected, the compiler outputs proper AVX or LRBNi intrinsics code.
+Any possible corner cases are handled: for example, if \emph{op-vect} is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.
 
@@
 For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance.
-Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop.
+Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required for every iteration.
+Experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop.
 The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension.
-Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers remain available to carry out the computation of the stiffness matrix.
-
-Loop unroll and unroll-and-jam of outer loops are fundamental to expose instruction-level parallelism and register reuse, so tuning critical parameters, like the unroll factor, becomes of great importance.
-It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for assembly kernels, where the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are unit-stride, recent versions of a vendor compiler like $intel$'s employ cost models capable of estimating close-to-optimal values for such parameters.
-We leave therefore the backend compiler in charge to select the unroll factor.
-This choice also simplifies the COFFEE's cost model.
-The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $intel$ compiler from applying this optimization, even if specific pragmas are added.
+Our compiler chooses the loop along which the number of invariant loads is smaller as outer, so that more registers remain available to carry out the computation of the stiffness matrix.
+
+Loop unroll and unroll-and-jam of outer loops are fundamental to expose instruction-level parallelism and register reuse, so tuning critical parameters, like the unroll factor, is of great importance.
+However, for assembly kernels the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are unit-stride.
+From our experience of inspecting assembly code and comparing with other hand-written implementations, recent versions of a vendor compiler like Intel's employ cost models capable of estimating close-to-optimal values for such parameters.
+We therefore leave the backend compiler in charge of selecting the unroll factor.
+This choice also simplifies COFFEE's cost model.
+The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code seems to prevent the Intel compiler from applying this optimization, even if specific pragmas are added.
 Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 25 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order \texttt{ijk}).
@@
 It takes into account the number of available logical vector registers, $n\_regs$, and the number of unique variables accessed: $n\_consts$ counts variables independent of both $j$ and $k$ loops and temporary registers, $n\_outer\_arrays$ counts $j$-dependent variables, and $n\_inner\_arrays$ counts $k$-dependent variables, assuming the \texttt{ijk} loop order.
-These values are used to estimate unroll-and-jam and split factors for $op$-$vect$ and $split$.
+These values are used to estimate unroll-and-jam and split factors for \emph{op-vect} and \emph{split}.
 If a factor is 0, then the corresponding transformation is not applied.
-The $split$ transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 2).
+The \emph{split} transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 2).
 A split factor of $n$ means that the assembly expression should be ``cut'' into $n$ sub-expressions.
 Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size.
-The unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of the available logical registers, i.e.
+The unroll-and-jam factor for the \emph{op-vect} transformation is determined as a function of the available logical registers, i.e.
 those not used for storing hoisted terms (line 9-11).
@@
 
-// Compute unroll-and-jam factor for $op$-$vect$ \\
+// Compute unroll-and-jam factor for \emph{op-vect} \\
 n\_regs\_avail = n\_regs - (n\_outer\_arrays + n\_consts)\\
@@
 \Return{$<$permute, split\_factor, uaj\_factor$>$}
-\caption{The cost model is employed by the compiler to estimate the most suitable unroll-and-jam (when $op$-$vect$ is used) and split factors, avoiding the overhead of auto-tuning.}
+\caption{The cost model is employed by the compiler to estimate the most suitable unroll-and-jam (when \emph{op-vect} is used) and split factors, avoiding the overhead of auto-tuning.}
 \end{algorithm}

### 2014-04-07 paper/bare_conf.tex
WIP on performance section

@@
 
-Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Phi.
+% TODO Xeon Phi flags
+% FIXME: cite UFL
+
+Experiments were run on two Intel architectures, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Xeon Phi.
 The \texttt{icc 13.1} compiler was used, with optimization level \texttt{-O2} and with auto-vectorization enabled (\texttt{-xAVX} on the Sandy Bridge, and \texttt{TODO} on the Phi).
@@
 For Advection-Diffusion, the ``Diffusion'' equation, which uses the same differential operators as Helmholtz, is considered.
-In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the stiffness matrix $A$.
-Burgers is a time-dependent problem, i.e.
-the assembly is recalculated every time step based on the result of previous iterations.
-It employs differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are accessed.
+In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$.
+Burgers is a time-dependent problem, which means the assembly is recalculated every time step based on the result of previous iterations.
+It employs differential operators different from those of Helmholtz, which has a major impact on the generated assembly code shown in Figure~\ref{code:burgers}, where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are accessed.
 
@@
 For instance, in the case of Helmholtz with $p=1$, the size of the $j$ and $k$ loops for the three kind of elements is, respectively, $3$, $4$, and $6$.
-Moving from 2D to 3D has the effect of increasing the number of basis function arrays, since, conceptually, the behaviour of the equation has now to be approximated also along a third axis.
+Moving from 2D to 3D has the effect of increasing the number of basis function arrays, since, conceptually, the behavior of the equation has now to be approximated also along a third axis.
 On the other hand, the polynomial order affects only the problem size (the three loops $i$, $j$, and $k$, and, as a consequence, the size of $X$ and $Y$ arrays).
 A range of polynomial orders, from $p=1$ to $p=4$, were tested; higher polynomial orders are excluded from the study because of current Firedrake limitations.
-In such a large space of problems, the size of the stiffness matrix rarely exceeds 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
+In such a large space of problems, the size of the element matrix rarely exceeds 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
+
+% FIXME: which limitations?
 

### 2014-04-07 paper/bare_conf.tex
Incorporate Florian's suggestions

--- 2014-04-05 paper/bare_conf.tex
@@
 \section{Introduction}
-In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modeled by means of partial differential equations (PDEs).
-Numerical techniques, such as the finite volume and finite element method, are widely employed to approximate solutions of PDEs.
-Unstructured meshes are often used to discretize the computational domain, since their geometric flexibility allows adapting the mesh to the required resolution.
-The solution is sought in each element of the discretized domain by applying suitable numerical kernels.
-As the number of elements can be of the order of millions, the time required to execute the computation can be hours or days.
+In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by means of partial differential equations (PDEs).
+Numerical techniques, like finite volume method and finite element method, are widely-employed to approximate solutions of PDEs.
+Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective.
+The solution is sought in each element of the discretized domain by applying suitable numerical operations.
+As the number of elements can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days.
 To address this problem, domain-specific languages (DSLs) have been developed.
-The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design to OP2~\cite{pyop2isc} demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}.
-Hydra uses the finite volume method, consists of roughly 50000 lines of code and is able to handle mesh sizes of over 100 millions edges.
-
-% FIXME: that citation is wrong! You're talking about OP2. You MUST maintain that distinction
-
-OP2 adopts a programming model, in which computations are expressed through self-contained functions, referred to as ``kernels''.
-A kernel is applied to all elements in a set of mesh entities such as edges, vertices, or cells, with an implicit synchronization between the application of two consecutive kernels.
+The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{op2-main}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}.
+
+OP2 adopts a programming model in which computations are expressed through self-contained functions, referred to as ``kernels''.
+A kernel is applied to all elements in a set of mesh components, such as edges, vertices, or cells, with an implicit synchronization between the application of two consecutive kernels.
 On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread.
@@
 
-% FIXME: citation for the 30-60% figure?
-
-We focus on local assembly (``assembly'' in the following), a fundamental step of the finite element method that covers an important fraction of the overall computation time, often in the range of 30$\%$-60$\%$.
+We focus on local assembly (``assembly'', in the following), a fundamental step of a finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$.
 During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain.
@@
 However, we do not consider the latter case.
-An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for low-order methods.
+An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for lower order methods.
 In the innermost loop, a problem-specific expression evaluates a two dimensional array, representing the result of local assembly in an element of the discretized domain.
-With such a kernel structure, we focus on aspects such as minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.
+With such a kernel structure, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.
 
@@
 
-Due to the constraints of available compilers and specialized linear-algebra libraries, we have automated a set of generic and model-driven code transformations in the COmpiler For FinitE Element local assembly (COFFEE), a compiler for optimizing local assembly kernels.
-COFFEE is integrated with Firedrake~\cite{firedrake-code}, a system for solving PDEs through the finite element method based on the PyOP2 abstraction~\cite{pyop2isc,pyop2ws}.
+Due to the constraints of available compilers and linear-algebra specialized libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for optimizing local assembly kernels.
+COFFEE is integrated with Firedrake~\cite{firedrake-code}, a system for solving PDEs through the finite element method based on the PyOP2 abstraction~\cite{pyop2isc}.
 It supports any problems expressible with this framework.
@@
 
-% FIXME: You're talking about FFC-generated code really
-
 Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal.
-Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (that is code where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels.
-The contributions of this paper are threefold:
+Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (i.e.
+code where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels.
+The contributions of this paper are threefold
 \begin{itemize}
-\item An optimization strategy for finite element local assembly, a class of kernels widely-used in scientific applications.
-Our approach exploits domain knowledge and goes beyond the limits of both vendor and research compilers.
+\item An optimisation strategy for finite element local assembly. Our approach exploits domain knowledge and goes beyond the limits of both vendor and research compilers.
 \item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
@@
 
-This paper is organized as follows: In Section~\ref{sec:background} we provide some background on local assembly, showing code generated by Firedrake and emphasizing important computational aspects.
-Section~\ref{sec:code-transf} describes the various code transformations in detail, highlighting when and how domain knowledge has been exploited.
+The paper is organized as follows.
+In Section~\ref{sec:background} we provide some background on local assembly, showing code generated by Firedrake and emphasizing critical computational aspects.
+Section~\ref{sec:code-transf} describes the various code transformations, highlighting when and how domain-knowledge has been exploited.
 The design and implementation of our compiler is discussed in Section~\ref{sec:pyop2-compiler}.
@@
 \small
-\KwIn{coordinate field (array), coefficient fields (e.g. velocity, array)}
-\KwOut{element matrix A (2D array)}
-\nlset{1}Compute Jacobian from coordinates \\
-Declare of constant data: basis functions and derivatives\\
-Compute element matrix using numerical quadrature
+\KwIn{element matrix (2D array, initialized to 0),\\~~~~~~~~~element coordinates (array),\\~~~~~~~~~coefficient fields (array, e.g. velocity)}
+\KwOut{element matrix (2D array)}
+\nlset{1}~~- Compute Jacobian from coordinates \\
+~~- Declare of constant data: basis functions and derivatives\\
+~~- Compute element matrix using numerical quadrature
 \caption{General structure of a local assembly kernel generated by Firedrake.}
@@
 
-% FIXME: stiffness matrix is only used for certain types of elliptic PDEs
-% FIXME: the fluidity manual isn't a good source for this, cite e.g. Brenner and Scott
-
-Local assembly is the computation of local contributions of a specific cell of the discretized domain to the global sparse matrix or right-hand side vector of the linear system.
+Local assembly is the computation of contributions of a specific cell in the discretized domain to the PDE solution.
 The process consists of numerically evaluating an integral to produce a small dense element matrix or an element vector~\cite{quadrature1, fluidity_manual_v4}.
-In this work we focus on local matrices, which are usually more costly to compute.
-
-% FIXME: you can't attribute the kernel code generation to Firedrake: mention FFC
+This operation is applied to all cells in the discretized domain.
+In this work we focus on local matrices, or ``element matrices'', which are more costly to compute.
 
 Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm.
-These kernels are compiled at runtime using a vendor compiler, and Firedrake manages the parallel execution over the mesh.
+It then triggers compilation of such kernels using an available vendor compiler, and eventually manages parallel execution over the mesh.
 The subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.
@@
 The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}.
-Inputs are a pointer to a zero initialized two dimensional array used to store the element matrix, a pointer to the element's coordinates in the discretized domain, and pointers to data of coefficient fields, for instance indicating value of velocity or pressure in the element.
+Input are a pointer to a zero-initialized two dimensional array used to store the element matrix, a pointer to the element's coordinates in the discretized domain, and pointers to data of coefficient fields, for instance indicating value of velocity or pressure in the element.
 The output is the evaluated element matrix.
-A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant two dimensional arrays for the finite element basis functions and their derivatives (if needed), initialized to problem-specific double-precision floating point values; 3) evaluation of the element matrix in an affine loop nest.
-All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (using \texttt{static const} in C).
-Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to various objects in the kernel.
+A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant two dimensional arrays, called ``basis functions'' (and their derivatives) in finite element terminology, initialized to problem-specific double-precision floating point values; 3) evaluation of the element matrix in an affine loop nest.
+All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (i.e.
+using \texttt{static const} in C language).
+Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel's objects.
 
@@
 
-% FIXME: You haven't really defined basis functions, so what is a Lagrange basis function?
-
 The actual complexity of a local assembly kernel depends on the finite element problem being solved.
-In simpler cases, the loop nest is perfect: it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products.
+In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products.
 An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$.
-In other scenarios, for instance when solving a non-linear problem like Burgers as shown in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
+In other scenarios, for instance when solving a non-linear problem, like Burgers in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
 Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse.
@@
 
-Some characteristics of assembly kernels are: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g.
+Some peculiarities of assembly kernels are: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g.
 the $F$ values in Burgers) and introduction of temporary arrays; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
@@
 The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake.
-One particular characteristic is that they all aim at improving the run time of the \texttt{ijk} loop nest, where the numerical integration takes place.
+One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place.
 In rare, yet important cases, however, this might not be sufficient to achieve notable speed-ups.
 Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions.
-It is worth noting that there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time.
+It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time.
 In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure).
@@
 
-As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens of iterations, render the optimization process challenging.
-While some problems achieve good performance relative to machine peak when applying a very specific set of transformations, there is another space of cases for which a more elaborate optimization strategy is required.
+As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens iterations, render the optimization process challenging.
+Relatively to machine peak, while some problems achieve good performance when applying a very specific set of transformations, there is also an ample space of cases for which a more elaborated optimization strategy is required.
 For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces.
-Furthermore, domain knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger.
+Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger.
 In this context, our goal is to understand the relationship between distinct code transformations, their impact on local assembly kernels, and to what extent their composability is effective in a class of problems and architectures.
@@
 Initially, all arrays are allocated to addresses that are multiples of $vl$.
-Then, two dimensional arrays are padded to a number of columns equal to the nearest multiple of $vl$.
+Then, two dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$.
 For instance, assume the original size of a basis function array is 3$\times$3 and that the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values.
@@
 Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels.
-We want to overcome two deficiencies that both Intel and GNU compilers have.
+We want to overcome two deficiencies that both $intel$ and $gnu$ compilers have.
 First, they can identify sub-expressions that are invariant with respect to the innermost loop only.
@@
 This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest.
-At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, since they are part of an inner loop.
+At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop.
 Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant.
@@
 Regardless of the specific input problem, and by abstracting from the presence of all variables independent of both $j$ and $k$, the element matrix computation is always reducible to this form.
-Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$.
+Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries ($j,k=0,...,4$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$.
 Values in a register are shuffled each time a product is performed.
-Standard compiler auto-vectorization (GNU and Intel), instead, executes 4 broadcast operations (the ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation.
+Standard compiler auto-vectorization ($gnu$ and $intel$), instead, executes 4 broadcast operations (i.e.
+``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation.
 Besides a larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.
@@
 
-The storage layout of $A$, however, is incorrect after the application of this outer-product-based vectorization (\emph{op-vect}, in the following).
+The storage layout of $A$, however, is incorrect after the application of this outer-product-based vectorization ($op$-$vect$, in the following).
 It can be efficiently restored with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the \texttt{ijk} loop nest.
-The generated pseudo-code for the simple Helmholtz problem when using \emph{op-vect} is shown in Figure~\ref{code:helmholtz-opvect}.
+The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.
 
@@
 \centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}}
-\caption{Restoring the storage layout after \emph{op-vect}.
+\caption{Restoring the storage layout after $op$-$vect$.
 The figure shows how 4$\times$4 elements in the top-left block of the element matrix $A$ can be moved to their correct positions.
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake when padding, data alignment, \emph{licm} and \emph{op-vect} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}.
-Here, we assume the polynomial order is $p=2$, since \emph{op-vect} can not be used when an iteration space dimension is smaller than the vector length.
+\caption{Local assembly code generated by Firedrake when padding, data alignment, $licm$ and $op$-$vect$ are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}.
+Here, we assume the polynomial order is $p=2$, since $op$-$vect$ can not be used when an iteration space dimension is smaller than the vector length.
 The original size of the $j$-$k$ iteration space (i.e.
@@
 In complex kernels, like Burgers' in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
-If the number of variables independent of the innermost loop dimension is close to or greater than the number of available CPU registers, it is likely to obtain poor register reuse.
+If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, it is likely to obtain poor register reuse.
 This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large.
 For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the \texttt{ijk} loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers.
-One potential solution to this problem consists of suitably ``splitting'' the computation of the stiffness matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}.
+One potential solution to this problem consists of suitably ``splitting'' the computation of the element matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}.
 Splitting an expression has, in general, several drawbacks.
 Firstly, it increases the number of accesses to $A$ proportionally to the ``split factor'', which is the number of sub-expressions produced.
-Secondly, depending on how the split is executed, it can lead to redundant computation (for example, the product $det*W3[i]$ is performed for each of the sub-expressions in the code of Figure~\ref{code:helmholtz-split}).
-Finally, it might affect register locality, although this is not the case for the Helmholtz example: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
+Secondly, depending on how the split is executed, it can lead to redundant computation (e.g.
+the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}).
+Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
 Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake when \emph{split} is applied to the optimized Helmholtz problem given in Figure~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
+\caption{Local assembly code generated by Firedrake when $split$ is applied to the optimized Helmholtz problem given in Figure~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
 \label{code:helmholtz-split}
@@
 Name (Abbreviation) & Parameter \\\hline\hline
-Generalized loop-invariant code motion (\emph{licm}) &   \\ \hline
+Generalized loop-invariant code motion ($licm$) &   \\ \hline
 Padding &  \\ \hline
@@
 Loop unrolling  & unroll factor \\ \hline
-Outer-product vectorization (\emph{op-vect}) & tile size \\ \hline
-Expression splitting (\emph{split}) & split point, split factor \\ \hline
+Outer-product vectorization ($op$-$vect$) & tile size \\ \hline
+Expression splitting ($split$) & split point, split factor \\ \hline
 \end{tabulary}
@@
 The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension.
-Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers remain available to carry out the computation of the stiffness matrix.
+Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers remain available to carry out the computation of the element matrix.
 
@@
 For Advection-Diffusion, the ``Diffusion'' equation, which uses the same differential operators as Helmholtz, is considered.
-In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the stiffness matrix $A$.
+In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$.
 Burgers is a time-dependent problem, i.e.
@@
 A range of polynomial orders, from $p=1$ to $p=4$, were tested; higher polynomial orders are excluded from the study because of current Firedrake limitations.
-In such a large space of problems, the size of the stiffness matrix rarely exceeds 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
+In such a large space of problems, the size of the element matrix rarely exceeds 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
 
@@
 Padding and data alignment enhance, in general, the quality of SIMD auto-vectorization.
-Sometimes the run-time of $licm$-$ap$ is similar to that of $licm$ because the stiffness matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned.
+Sometimes the run-time of $licm$-$ap$ is similar to that of $licm$ because the element matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned.
 Occasionally $licm$-$ap$ is slower than $licm$ (e.g.
@@
 Expression splitting, or loop fission, contributes to relieve register pressure.
-The transformation is useless if the stiffness matrix can be evaluated by accessing a relatively small number of basis functions (or temporary values introduced by $licm$), for which efficient register allocation can be already guaranteed.
-In the cases of Helmholtz and Diffusion, irrespective of split factor and cut, the only consequence of applying $split$ is to augment the number of increments for evaluating the stiffness matrix.
+The transformation is useless if the element matrix can be evaluated by accessing a relatively small number of basis functions (or temporary values introduced by $licm$), for which efficient register allocation can be already guaranteed.
+In the cases of Helmholtz and Diffusion, irrespective of split factor and cut, the only consequence of applying $split$ is to augment the number of increments for evaluating the element matrix.
 On the Sandy Bridge machine, slow downs up to 1.4$\times$ were observed in these two problems.
@@
 In COFFEE, Burgers' slow downs occur in presence of a small iteration space (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$), although they are generally milder than the maximum one reported in table, obtained with a triangular mesh and $p = 1$.
-This is a result of zero-columns removal in FEniCS: indirection arrays prevent auto-vectorization, which significantly impacts performance as soon as the stiffness matrix becomes bigger than 12$\times$12.
+This is a result of zero-columns removal in FEniCS: indirection arrays prevent auto-vectorization, which significantly impacts performance as soon as the element matrix becomes bigger than 12$\times$12.
 However, with the forthcoming zero-removal optimization in COFFEE, we expect to outperform FEniCS even in these cases as well.
@@
 This particular kernel can be easily reduced to a sequence of four matrix-matrix multiplies that can be computed via calls to BLAS \texttt{dgemm}.
-In the case of $p=4$, where the stiffness matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only $licm$, data alignment and padding were used, with the slow down being even worse for smaller problem sizes.
+In the case of $p=4$, where the element matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only $licm$, data alignment and padding were used, with the slow down being even worse for smaller problem sizes.
 These experiments justify that there is a set of problems for which turning to MKL is not beneficial in terms of performance.

### 2014-04-08 paper/bare_conf.tex
Re-factor a few paragraphs

@@
 % Padding and Data Alignment
-\subsection{Padding and Data Alignment}
+\subsection{Prerequisites for Effective Auto-vectorization: Padding and Data Alignment}
 Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length $vl$.
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake when padding, data alignment, and $licm$ are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}.
+\caption{Local assembly code generated by Firedrake when padding, data alignment, and \emph{licm} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}.
 Data alignment and padding are for an AVX machine.
@@
 \centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}}
-\caption{Restoring the storage layout after $op$-$vect$.
+\caption{Restoring the storage layout after \emph{op-vect}.
 The figure shows how 4$\times$4 elements in the top-left block of the element matrix $A$ can be moved to their correct positions.
@@
 Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations.
-At run-time, this high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
+At run-time, the high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
 ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}.
-The output of COFFEE is C code, which is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and parallel execution over the unstructured mesh take place.
+The output of COFFEE, C code, is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and execution over the discretized domain take place.
 Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem.
-Auto-tuning might be used, although at the moment we avoid it to minimize the run-time overhead (the generation of the AST occurs at run-time as soon as problem-specific data are available).
+Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of ASTs occurs at run-time as soon as problem-specific data are available.
 Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.
@@
 For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance.
-Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop.
+Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required for every iteration.
+Experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop.
 The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension.
-Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers remain available to carry out the computation of the element matrix.
-
-Loop unroll and unroll-and-jam of outer loops are fundamental to expose instruction-level parallelism and register reuse, so tuning critical parameters, like the unroll factor, becomes of great importance.
-It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for assembly kernels, where the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are unit-stride, recent versions of a vendor compiler like $intel$'s employ cost models capable of estimating close-to-optimal values for such parameters.
-We leave therefore the backend compiler in charge to select the unroll factor.
+Our compiler chooses the loop along which the number of invariant loads is smaller as outer so that more registers remain available to carry out the computation of the element matrix.
+
+Loop unroll and unroll-and-jam of outer loops are fundamental to expose instruction-level parallelism, so tuning critical parameters, like the unroll factor, is of great importance.
+However, from our experience of inspecting assembly code and comparing with other hand-written implementations, recent versions of a vendor compiler like Intel's employ cost models capable of estimating close-to-optimal values for such parameters.
+This is particularly true for assembly kernels, where the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are unit-stride.
+We therefore leave the backend compiler in charge of selecting the unroll factor.
 This choice also simplifies the COFFEE's cost model.
-The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $intel$ compiler from applying this optimization, even if specific pragmas are added.
+The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code seems to prevent the Intel compiler from applying this optimization, even if specific pragmas are added.
 Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 25 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order \texttt{ijk}).
@@
 
-Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Phi.
+Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Xeon Phi.
 The \texttt{icc 13.1} compiler was used, with optimization level \texttt{-O2} and with auto-vectorization enabled (\texttt{-xAVX} on the Sandy Bridge, and \texttt{TODO} on the Phi).
@@
 In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$.
-Burgers is a time-dependent problem, i.e.
-the assembly is recalculated every time step based on the result of previous iterations.
-It employs differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are accessed.
-
-These problems were studied varying both the shape of mesh elements and the polynomial order $p$ of the method.
+Burgers is a non-linear problem employing differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are accessed.
+
+These problems were studied varying both the shape of mesh elements and the polynomial order $p$ of the method, whereas the element family, Lagrange, is fixed.
 Intuitively, the bigger the element shape and $p$, the larger is the iteration space.
@@
 \label{table:perf-results-licm}
-\caption{Impact of generalized loop-invariant code motion ($licm$ column) on the Helmholtz, Diffusion and Burgers problems, for three kind of elements belonging to the Lagrange family (triangle, tetrahedron, prism), for the range of polynomial orders $p \in \lbrace$1, 4$\rbrace$.
+\caption{Impact of generalized loop-invariant code motion (\emph{licm} column) on the Helmholtz, Diffusion and Burgers problems, for three kind of elements belonging to the Lagrange family (triangle, tetrahedron, prism), for the range of polynomial orders $p \in \lbrace$1, 4$\rbrace$.
 Each entry indicates the range of speed-ups obtained over the non-optimized implementation.
-The column $licm$-$ap$ illustrates the combination of $licm$ with data alignment and padding.
-Results are shown for both the Sandy Bridge and the Phi machine.}
+The column \emph{licm-ap} illustrates the combination of \emph{licm} with data alignment and padding.
+Results are shown for both the Sandy Bridge and the Xeon Phi architectures.}
 \end{table*}
 
-Table~\ref{table:perf-results-licm} illustrates the speed-ups obtained on the Sandy Bridge and the Phi machines when $licm$, data alignment, and padding are used, over non-transformed code.
-We distinguish between $licm$ and $licm$-$ap$; the latter makes use of padding and data alignment.
+Table~\ref{table:perf-results-licm} illustrates the speed-ups obtained on the Sandy Bridge and the Phi machines when \emph{licm}, data alignment, and padding are used, over non-transformed code.
+We distinguish between \emph{licm} and \emph{licm-ap}; the latter makes use of padding and data alignment.
 Inspection of assembly code generated by \texttt{icc} confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant with respect to outer loops are hoisted and, interestingly, not vectorized.
@@
 Padding and data alignment enhance, in general, the quality of SIMD auto-vectorization.
-Sometimes the run-time of $licm$-$ap$ is similar to that of $licm$ because the element matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned.
-Occasionally $licm$-$ap$ is slower than $licm$ (e.g.
+Sometimes the run-time of \emph{licm-ap} is similar to that of \emph{licm} because the element matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned.
+Occasionally \emph{licm-ap} is slower than \emph{licm} (e.g.
 in Burgers 3D p3).
-This is due to the large number of aligned temporaries introduced by $licm$, which probably induces cache associativity conflicts.
+This is due to the large number of aligned temporaries introduced by \emph{licm}, which probably induces cache associativity conflicts.
 
@@
 \label{sec:perf-eval-opvect}
-Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the speed-ups achieved by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmoltz and Diffusion kernels, respectively.
-For each problem instance we report two bars: one indicates the best run-time achieved by auto-tuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved by the COFFEE's cost model.
+Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the speed-ups achieved applying \emph{op-vect} on top of \emph{licm-ap} to the Helmoltz and Diffusion kernels, respectively.
+For each problem instance we report two bars: one indicates the best run-time obtained by auto-tuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved with the COFFEE's cost model.
 In general, there is no substantial difference between the two.
@@
 This is due to the fact that in the considered problems the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short.
-On the other hand, as soon as the iteration space becomes too big, vector-register tiles are essential to minimize pipeline stalls.
-A maximum speed-up of 1.4$\times$ over $licm$-$ap$ was observed in the Diffusion kernel when using a prismatic mesh and $p=4$.
-Application of $op$-$vect$ to the Burgers problem induces meaningful slow downs, because of the large number of temporary arrays that need to be tiled, which exceeds the number of available logical registers on the underlying machine.
-Expression splitting can be used in combination to $op$-$vect$ to alleviate this issue; this is discussed in Section~\ref{sec:perf-results-split}.
-
-We have also used the Intel Architecture Code Analyzer (IACA) tool~\cite{IACA} to demonstrate that the improved run-time is actually due to increased register re-use.
-By applying the tool to the Helmholtz problem on a tetrahedral mesh and $p=4$, it can be noticed that the number of clock cycles to execute one iteration of the $j$ loop decreases by roughly 17$\%$, on the Sandy Bridge, when using $op$-$vect$.
-This is because of less pressure on both data (cache) ports available in the processor.
+On the other hand, as the iteration space becomes bigger, vector-register tiling minimizes pipeline stalls, leading to maximum speed-ups of 1.4$\times$ in Diffusion and of 1.24$\times$ in Helmholtz on the Sandy Bridge architecture.
+This was proved using the Intel Architecture Code Analyzer tool~\cite{IACA}.
+In Helmholtz, for example, the tool showed that when using \emph{op-vect} the number of clock cycles to execute one iteration of the $j$ loop decreases by roughly 17$\%$, and that this is a consequence of increased register re-use, which relieves the pressure on both data (cache) ports available in the processor.
+
+Application of \emph{op-vect} to the Burgers problem induces meaningful slow downs, due to the large number of temporary arrays that need to be tiled, which exceeds the number of available logical registers on the underlying machine.
+Expression splitting can be used in combination to \emph{op-vect} to alleviate this issue; this is discussed in Section~\ref{sec:perf-results-split}.
 
@@
 \includegraphics[scale=0.7]{Pictures/helmholtz-normalized-opvect.pdf}
-\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Helmholtz kernel.
+\caption{Speed-ups obtained by applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel.
 Tri, Tet, and Prism stand, respectively, for triangual, tetrahedral and prismatic mesh.
@@
 \includegraphics[scale=0.7]{Pictures/diffusion-normalized-opvect.pdf}
-\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Diffusion kernel.
+\caption{Speed-ups obtained by applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel.
 Tri, Tet, and Prism stand, respectively, for triangual, tetrahedral and prismatic mesh.
@@
 Expression splitting, or loop fission, contributes to relieve register pressure.
-The transformation is useless if the element matrix can be evaluated by accessing a relatively small number of basis functions (or temporary values introduced by $licm$), for which efficient register allocation can be already guaranteed.
-In the cases of Helmholtz and Diffusion, irrespective of split factor and cut, the only consequence of applying $split$ is to augment the number of increments for evaluating the element matrix.
+The transformation is useless if the element matrix can be evaluated by accessing a relatively small number of basis functions (or temporary values introduced by \emph{licm}), for which efficient register allocation can be already guaranteed.
+In the cases of Helmholtz and Diffusion, irrespective of split factor and cut, the only consequence of applying \emph{split} is to augment the number of increments for evaluating the element matrix.
 On the Sandy Bridge machine, slow downs up to 1.4$\times$ were observed in these two problems.
 The cost model, however, prevents the adoption of this transformation, since the \texttt{while} statement at line 5 in Figure~\ref{algo:applyCostModel} is never entered.
-In the Burgers problem, $split$ allows notable speed-ups to be achieved over the best run-time provided by either $licm$ or $licm$-$ap$.
+In the Burgers problem, \emph{split} allows notable speed-ups to be achieved over the best run-time provided by either \emph{licm} or \emph{licm-ap}.
 Results are shown in Figure~\ref{fig:split-burgers-speedup}.
@@
 
-The $split$ transformation was also tried in combination with $op$-$vect$ ($split$-$op$-$vect$), although the cost model prevents the application of the latter on the Sandy Bridge.
-Despite performance improvements up to 1.22$\times$, $split$-$op$-$vect$ never outperforms $split$.
+The \emph{split} transformation was also tried in combination with \emph{op-vect} (\emph{split-op-vect}), although the cost model prevents the application of the latter on the Sandy Bridge.
+Despite performance improvements up to 1.22$\times$, \emph{split-op-vect} never outperforms \emph{split}.
 This is because of two reasons: for small split factors, like 1 and 2, the data space to be tiled is still too big, and register spilling still affects the run-time; for higher ones, sub-expressions become so small that, as explained in Section~\ref{sec:perf-eval-opvect}, full or partial unrolling of the loop nest allows achieving a certain degree of register re-use.
@@
 \includegraphics[scale=0.7]{Pictures/burgers-normalized-split.pdf}
-\caption{Speed-ups obtained by applying $op$-$vect$ on top of $licm$-$ap$ to the Burgers kernel.
+\caption{Speed-ups obtained by applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel.
 Tri, Tet, and Prism stand, respectively, for triangual, tetrahedral and prismatic mesh.
@@
 This particular kernel can be easily reduced to a sequence of four matrix-matrix multiplies that can be computed via calls to BLAS \texttt{dgemm}.
-In the case of $p=4$, where the element matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only $licm$, data alignment and padding were used, with the slow down being even worse for smaller problem sizes.
+In the case of $p=4$, where the element matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only \emph{licm}, data alignment and padding were used, with the slow down being even worse for smaller problem sizes.
 These experiments justify that there is a set of problems for which turning to MKL is not beneficial in terms of performance.
-It is possible that employing certain BLAS implementations or auto-tuning strategies for particular problem sizes is helpful, although the same problems afflicting $split$ (exposed in Section~\ref{sec:expr-split}) might limit the performance gain.
+It is possible that employing certain BLAS implementations or auto-tuning strategies for particular problem sizes is helpful, although the same problems afflicting \emph{split} (exposed in Section~\ref{sec:expr-split}) might limit the performance gain.
 However, as anticipated, extraction of matrix-matrix multiplies from analysis of the kernel's AST and a strategy for their execution will be comprehensively addressed in further work.

### 2014-04-09 paper/bare_conf.tex
Rephrase vector-register tiling + minor fixes

@@
 Then, two dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$.
-For instance, assume the original size of a basis function array is 3$\times$3 and that the underlying machine possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values.
+For instance, assume the original size of a basis function array is 3$\times$3 and that the underlying architecture possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values.
 In this example, a padded version of the array will have size 3$\times$4.
@@
 \caption{Local assembly code generated by Firedrake when padding, data alignment, and \emph{licm} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}.
-Data alignment and padding are for an AVX machine.
+Data alignment and padding are for an AVX architecture.
 In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$, so they can be precomputed once in a single loop $r$; in general, this might not be the case.}
@@
 \begin{algorithm}[t]
-\label{algo:PyOP2Compiler}
   \textbf{The COFFEE Compiler}\\
@@
   ast.apply\_inter\_kernel\_vectorization(wrapper, isa)
-  \Return{wrapper + ast.from\_ast\_to\_c()}
-}
-\If{isa.backend == gpu}{
-  \If{it\_space}{ast.extract\_iteration\_space(wrapper)}
   \Return{wrapper + ast.from\_ast\_to\_c()}
@@
 \caption{Pseudocode of the COFFEE pipeline.}
+\label{algo:PyOP2Compiler}
 \end{algorithm}
@@
 This feature, currently under development, has been proved to be useful in several finite volume applications~\cite{inter-kernel-vect}.
-The second transformation step is applied if the backend is a manycore architecture, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{pyop2isc}.
 Then, an ordered sequence of optimization steps are executed.
@@
 The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code seems to prevent the Intel compiler from applying this optimization, even if specific pragmas are added.
-Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 25 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order \texttt{ijk}).
+Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 19 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order \texttt{ijk}).
 
@@
 \begin{algorithm}[t]
-\label{algo:applyCostModel}
   \textbf{Cost Model}\\
@@
 \caption{The cost model is employed by the compiler to estimate the most suitable unroll-and-jam (when \emph{op-vect} is used) and split factors, avoiding the overhead of auto-tuning.}
+\label{algo:applyCostModel}
 \end{algorithm}
@@
 
-Experiments were run on two Intel machines, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Xeon Phi.
-The \texttt{icc 13.1} compiler was used, with optimization level \texttt{-O2} and with auto-vectorization enabled (\texttt{-xAVX} on the Sandy Bridge, and \texttt{TODO} on the Phi).
-Other optimization levels performed, in general, slightly worse than \texttt{-O2}.
+Experiments were run on two Intel architectures, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Xeon Phi.
+The \texttt{icc 13.1} compiler was used.
+On the Sandy Bridge, the compilation flags used were \texttt{-O2} (other optimization levels perfomed, in general, slightly worse) and \texttt{-xAVX} for auto-vectorization.
+On the Xeon Phi, the optimization level \texttt{-O3} was used.
 Our code transformations were evaluated in three real-world problems based on the following PDEs:
@@
 
-Table~\ref{table:perf-results-licm} illustrates the speed-ups obtained on the Sandy Bridge and the Phi machines when \emph{licm}, data alignment, and padding are used, over non-transformed code.
+Table~\ref{table:perf-results-licm} illustrates the speed-ups obtained on the Sandy Bridge and the Xeon Phi machines when \emph{licm}, data alignment, and padding are used, over non-transformed code.
 We distinguish between \emph{licm} and \emph{licm-ap}; the latter makes use of padding and data alignment.
@@
 This is chiefly because assembly kernels fit the L1 cache, so estimating the run-time based on the quality of register allocation is relatively simple.
+
 The rationale behind these results is that, for smaller configurations, vector-register tiling is rarely helpful.
 This is due to the fact that in the considered problems the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short.
-On the other hand, as the iteration space becomes bigger, vector-register tiling minimizes pipeline stalls, leading to maximum speed-ups of 1.4$\times$ in Diffusion and of 1.24$\times$ in Helmholtz on the Sandy Bridge architecture.
-This was proved using the Intel Architecture Code Analyzer tool~\cite{IACA}.
-In Helmholtz, for example, the tool showed that when using \emph{op-vect} the number of clock cycles to execute one iteration of the $j$ loop decreases by roughly 17$\%$, and that this is a consequence of increased register re-use, which relieves the pressure on both data (cache) ports available in the processor.
-
-Application of \emph{op-vect} to the Burgers problem induces meaningful slow downs, due to the large number of temporary arrays that need to be tiled, which exceeds the number of available logical registers on the underlying machine.
+On the other hand, as the iteration space becomes bigger, vector-register tiling leads to percentage improvements up to 1.4$\times$ on the Sandy Bridge (Diffusion, prismatic mesh, $p=4$) and up to X$\times$ on the Xeon Phi (X, X, X).
+Using the Intel Architecture Code Analyzer tool~\cite{IACA}, we proved that this is a consequence of increased register re-use.
+In Helmholtz, for example, the tool showed that when using \emph{op-vect} the number of clock cycles to execute one iteration of the $j$ loop decreases by roughly 17$\%$, and that this is a result of the relieved pressure on both of the data (cache) ports available in the core.
+
+On the Sandy Bridge, we have also measured the performance of individual kernels in terms of floating-point operations per second.
+The theoretical peak, with the Intel Turbo Boost technology activated, is 30.4 GFlop/s.
+In the case of Diffusion using a prismatic mesh and $p=4$, we achived a maximum of 21.9 GFlop/s with \emph{op-vect} enabled, whereas 16.4 GFlop/s was obtained when only \emph{licm-ap} is used.
+This result is inline with the expectations: analysis of assembly code showed that, in the \texttt{jk} loop nest, which in this problem represents the bulk of the computation, 73$\%$ of instructions are actually floating-point operations.
+
+Application of \emph{op-vect} to the Burgers problem induces meaningful slow downs, due to the large number of temporary arrays that need to be tiled, which exceeds the number of available logical registers on the underlying architecture.
 Expression splitting can be used in combination to \emph{op-vect} to alleviate this issue; this is discussed in Section~\ref{sec:perf-results-split}.
@@
 \label{table:comparison-to-FFC-opt}
-\caption{Summary of performance comparison between FEniCS and COFFEE on the Sandy Bridge machine.
-The slow down and speed up columns compare the best COFFEE implementation and the FEniCS implementation with optimizations enabled.}
+\caption{Performance comparison between FEniCS (optimizations enabled) and COFFEE on the Sandy Bridge.}
 \end{table}
@@
 These experiments justify that there is a set of problems for which turning to MKL is not beneficial in terms of performance.
-It is possible that employing certain BLAS implementations or auto-tuning strategies for particular problem sizes is helpful, although the same problems afflicting \emph{split} (exposed in Section~\ref{sec:expr-split}) might limit the performance gain.
-However, as anticipated, extraction of matrix-matrix multiplies from analysis of the kernel's AST and a strategy for their execution will be comprehensively addressed in further work.
+It is possible that employing certain BLAS implementations or auto-tuning strategies, particularly for certain problem sizes, is helpful, although the same problems afflicting \emph{split}, exposed in Section~\ref{sec:expr-split}, might limit the performance gain.
+However, as anticipated, the problem of identifying and extracting matrix-matrix multiplies from analysis of the kernel's AST, as well as a strategy for their execution, will be comprehensively addressed in further work.
 

### 2014-04-11 paper/bare_conf.tex
Added new plot. Minor refactor

@@
 On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread.
-Similar programming and execution models are adopted in~\cite{Fenics},~\cite{fluidity_manual_v4},~\cite{lizst}.
+Similar programming and execution models are adopted in~\cite{pyop2isc},~\cite{Fenics},~\cite{fluidity_manual_v4},~\cite{lizst}.
 Kernel optimization is one of the major concerns in unstructured mesh applications.
-In this paper, we tackle this problem by proposing an optimization strategy for a class of kernels used in finite element methods.
+In this paper, we tackle this problem in the context of the finite element method.
 
@@
 BLAS routines could be theoretically employed, although a fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies.
-In addition, as detailed in Section~\ref{sec:perf-results}, in which manual BLAS-based implementations are evaluated, the small dimension of the involved matrices and loss in data locality can limit the performance gain, if any.
+In addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small dimension of the involved matrices and the potential loss in data locality can limit the performance gain, if any.
 
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake for a Helmholtz problem on a 2D triangular mesh with Lagrange $p=1$ elements.}
+\caption{Local assembly code generated by Firedrake for a Helmholtz problem on a 2D triangular mesh using Lagrange $p=1$ elements.}
 \label{code:helmholtz}
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake for a Burgers problem on a 3D tetrahedral mesh with Lagrange $p=1$ elements.}
+\caption{Local assembly code generated by Firedrake for a Burgers problem on a 3D tetrahedral mesh using Lagrange $p=1$ elements.}
 \label{code:burgers}
@@
 The actual complexity of a local assembly kernel depends on the finite element problem being solved.
-In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products.
-An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$.
+In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products involving basis functions.
+An example is provided in Figure~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$.
 In other scenarios, for instance when solving a non-linear problem, like Burgers in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
@@
 One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place.
-In rare, yet important cases, however, this might not be sufficient to achieve notable speed-ups.
+In rare, yet important cases, however, this might not be sufficient to achieve notable performance improvements.
 Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions.
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake when padding, data alignment, and \emph{licm} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}.
-Data alignment and padding are for an AVX architecture.
-In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$, so they can be precomputed once in a single loop $r$; in general, this might not be the case.}
+\caption{Local assembly code generated by Firedrake when padding, data alignment, and \emph{licm} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz} for an AVX architecture.
+In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$, so they can be precomputed once in the $r$ loop.}
 \label{code:helmholtz-licm}
 \end{algorithm}
-
-Figure~\ref{code:helmholtz-licm} shows the Helmholtz assembly code after the application of loop-invariant code motion, padding, and data alignment.
 
@@
 ~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
-~~~~~~~~// Call $Load$ and $set$ intrinsics \\
+~~~~~~~~// $load$ and $set$ intrinsics \\
 ~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
@@
 ~~~~~~$\rbrace$\\
-~~~~// Do Remainder loop (from $j=4$ to $j=6$)\\
+~~~~// Remainder loop (from $j=4$ to $j=6$)\\
 ~~$\rbrace$\\
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake when padding, data alignment, \emph{licm} and \emph{op-vect} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz}.
+\caption{Local assembly code generated by Firedrake for the Helmholtz problem when \emph{op-vect} is applied on top of the optimizations shown in Figure~\ref{code:helmholtz-licm}.
 Here, we assume the polynomial order is $p=2$, since \emph{op-vect} can not be used when an iteration space dimension is smaller than the vector length.
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake when \emph{split} is applied to the optimized Helmholtz problem given in Figure~\ref{code:helmholtz-licm}. In this example, the split factor is 2.}
+\caption{Local assembly code generated by Firedrake for the Helmholtz problem when \emph{split} is applied on top of the optimizations shown in Figure~\ref{code:helmholtz-licm}.
+In this example, the split factor is 2.}
 \label{code:helmholtz-split}
@@
 Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations.
-At run-time, the high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
+At run-time, the high-level specification is translated by the FEniCS Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
 ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}.
@@
 The \texttt{icc 13.1} compiler was used.
-On the Sandy Bridge, the compilation flags used were \texttt{-O2} (other optimization levels perfomed, in general, slightly worse) and \texttt{-xAVX} for auto-vectorization.
+On the Sandy Bridge, the compilation flags used were \texttt{-O2} and \texttt{-xAVX} for auto-vectorization.
 On the Xeon Phi, the optimization level \texttt{-O3} was used.
+Other optimization levels perfomed, in general, slightly worse.
 Our code transformations were evaluated in three real-world problems based on the following PDEs:
@@
 Intuitively, the bigger the element shape and $p$, the larger is the iteration space.
-Triangles (2D), tetrahedron (3D), and prisms (3D) were tested as element shape.
-For instance, in the case of Helmholtz with $p=1$, the size of the $j$ and $k$ loops for the three kind of elements is, respectively, $3$, $4$, and $6$.
-Moving from 2D to 3D has the effect of increasing the number of basis function arrays, since, conceptually, the behaviour of the equation has now to be approximated also along a third axis.
+Triangles, tetrahedron, and prisms were tested as element shape.
+For instance, in the case of Helmholtz with $p=1$, the size of the $j$ and $k$ loops for the three element shapes is, respectively, $3$, $4$, and $6$.
+Moving to bigger shapes has the effect of increasing the number of basis function arrays, since, intuitively, the behaviour of the equation has now to be approximated also along a third axis.
 On the other hand, the polynomial order affects only the problem size (the three loops $i$, $j$, and $k$, and, as a consequence, the size of $X$ and $Y$ arrays).
@@
 \cline{3-6}
-& & \multicolumn{2}{c|}{\texttt{AVX}} & \multicolumn{2}{c|}{\texttt{LRBNi}}\\
+& & \multicolumn{2}{c|}{\texttt{Sandy Bridge}} & \multicolumn{2}{c|}{\texttt{Xeon Phi}}\\
 \cline{3-6}
-& & \multicolumn{1}{c}{\texttt{licm}} & \multicolumn{1}{c|}{\texttt{licm+ap}} & \multicolumn{1}{c}{\texttt{licm}} & \multicolumn{1}{c|}{\texttt{licm+ap}} \\
+& & \multicolumn{1}{c}{\texttt{licm}} & \multicolumn{1}{c|}{\texttt{licm-ap}} & \multicolumn{1}{c}{\texttt{licm}} & \multicolumn{1}{c|}{\texttt{licm-ap}} \\
 \cline{1-6}
 %\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.05$\times$-1.68$\times$ & \multicolumn{1}{c|}{~1.32$\times$-4.14$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.36$\times$-2.64$\times$ & \multicolumn{1}{c|}{~1.35$\times$-3.32$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~2.16$\times$-2.45$\times$ & \multicolumn{1}{c|}{~2.42$\times$-2.63$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.05$\times$-1.68$\times$ & \multicolumn{1}{c|}{~1.32$\times$-4.14$\times$} & ~~1.15$\times$-1.70$\times$ & ~~1.54$\times$-3.27$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.36$\times$-2.64$\times$ & \multicolumn{1}{c|}{~1.35$\times$-3.32$\times$} & ~~1.18$\times$-1.87$\times$ & ~~1.27$\times$-2.92$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~2.16$\times$-2.45$\times$ & \multicolumn{1}{c|}{~2.42$\times$-2.63$\times$} & ~~0.96$\times$-1.64$\times$ & ~~1.84$\times$-1.72$\times$ \\
 \cline{1-6}
-\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.09$\times$-1.64$\times$ & \multicolumn{1}{c|}{~1.37$\times$-4.28$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.30$\times$-3.11$\times$ & \multicolumn{1}{c|}{~1.41$\times$-3.82$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~1.81$\times$-2.70$\times$ & \multicolumn{1}{c|}{~2.55$\times$-3.13$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.09$\times$-1.64$\times$ & \multicolumn{1}{c|}{~1.37$\times$-4.28$\times$} & ~~1.13$\times$-1.22$\times$ & ~~0.92$\times$-2.79$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.30$\times$-3.11$\times$ & \multicolumn{1}{c|}{~1.41$\times$-3.82$\times$} & ~~1.08$\times$-2.11$\times$ & ~~1.10$\times$-5.53$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~1.81$\times$-2.70$\times$ & \multicolumn{1}{c|}{~2.55$\times$-3.13$\times$} & ~~1.04$\times$-2.74$\times$ & ~~1.73$\times$-4.14$\times$ \\
 \cline{1-6}
-\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.53$\times$-2.46$\times$ & \multicolumn{1}{c|}{~1.56$\times$-2.77$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.58$\times$-2.24$\times$ & \multicolumn{1}{c|}{~1.59$\times$-2.10$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~1.32$\times$-2.11$\times$ & \multicolumn{1}{c|}{~1.42$\times$-2.31$\times$} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.53$\times$-2.46$\times$ & \multicolumn{1}{c|}{~1.56$\times$-2.77$\times$} & ~~1.19$\times$-2.97$\times$ & ~~2.70$\times$-4.27$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.58$\times$-2.24$\times$ & \multicolumn{1}{c|}{~1.59$\times$-2.10$\times$} & ~~1.03$\times$-1.21$\times$ & ~~1.31$\times$-1.55$\times$ \\
+\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~1.32$\times$-2.11$\times$ & \multicolumn{1}{c|}{~1.42$\times$-2.31$\times$} & ~~1.04$\times$-1.40$\times$ & ~~1.27$\times$-2.18$\times$ \\
 \cline{1-6}
@@
 
-\begin{table*}[t]\normalsize
+\begin{table*}[t]\small
 \massmatrixtwodresultsnorms
-\label{table:perf-results-licm}
-\caption{Impact of generalized loop-invariant code motion (\emph{licm} column) on the Helmholtz, Diffusion and Burgers problems, for three kind of elements belonging to the Lagrange family (triangle, tetrahedron, prism), for the range of polynomial orders $p \in \lbrace$1, 4$\rbrace$.
-Each entry indicates the range of speed-ups obtained over the non-optimized implementation.
+~
+\caption{Impact of generalized loop-invariant code motion (\emph{licm} column) on the Helmholtz, Diffusion and Burgers problems, varying the element shape (triangle, tetrahedron, prism) and the polynomial order ($p \in [1, 4]$).
+The element family is Lagrange.
+Each entry indicates the minimum and maximum performance improvements obtained over the non-optimized implementation in the range of $p$ values.
 The column \emph{licm-ap} illustrates the combination of \emph{licm} with data alignment and padding.
 Results are shown for both the Sandy Bridge and the Xeon Phi architectures.}
+\label{table:perf-results-licm}
 \end{table*}
 
-Table~\ref{table:perf-results-licm} illustrates the speed-ups obtained on the Sandy Bridge and the Xeon Phi machines when \emph{licm}, data alignment, and padding are used, over non-transformed code.
+Table~\ref{table:perf-results-licm} illustrates the performance improvement obtained on the Sandy Bridge and the Xeon Phi machines when \emph{licm}, data alignment, and padding are used, over non-transformed code.
 We distinguish between \emph{licm} and \emph{licm-ap}; the latter makes use of padding and data alignment.
 Inspection of assembly code generated by \texttt{icc} confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant with respect to outer loops are hoisted and, interestingly, not vectorized.
-This motivates the usually significant gain obtained.
-Padding and data alignment enhance, in general, the quality of SIMD auto-vectorization.
+Padding and data alignment enhance, in general, the quality of auto-vectorization.
 Sometimes the run-time of \emph{licm-ap} is similar to that of \emph{licm} because the element matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned.
@@
 in Burgers 3D p3).
-This is due to the large number of aligned temporaries introduced by \emph{licm}, which probably induces cache associativity conflicts.
+One possible explanation is that the large number of aligned temporaries introduced by \emph{licm} induce cache associativity conflicts.
 
@@
 \label{sec:perf-eval-opvect}
-Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the speed-ups achieved applying \emph{op-vect} on top of \emph{licm-ap} to the Helmoltz and Diffusion kernels, respectively.
+Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the performance improvements achieved applying \emph{op-vect} on top of \emph{licm-ap} to the Helmoltz and Diffusion kernels on the Sandy Bridge, whereas Figure~\ref{fig:opvect-helmholtz-speedup-phi} shows analogous results for Helmholtz on the Xeon Phi.
 For each problem instance we report two bars: one indicates the best run-time obtained by auto-tuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved with the COFFEE's cost model.
@@
 Using the Intel Architecture Code Analyzer tool~\cite{IACA}, we proved that this is a consequence of increased register re-use.
-In Helmholtz, for example, the tool showed that when using \emph{op-vect} the number of clock cycles to execute one iteration of the $j$ loop decreases by roughly 17$\%$, and that this is a result of the relieved pressure on both of the data (cache) ports available in the core.
+In Helmholtz $p=4$, for example, the tool showed that when using \emph{op-vect} the number of clock cycles to execute one iteration of the $j$ loop decreases by roughly 17$\%$, and that this is a result of the relieved pressure on both of the data (cache) ports available in the core.
 
@@
 \includegraphics[scale=0.7]{Pictures/helmholtz-normalized-opvect.pdf}
-\caption{Speed-ups obtained by applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel.
-Tri, Tet, and Prism stand, respectively, for triangual, tetrahedral and prismatic mesh.
-CM stands for cost model.}
+\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel on the Sandy Bridge. CM stands for cost model.}
 \label{fig:opvect-helmholtz-speedup}
@@
 \begin{figure}[h]
+\includegraphics[scale=0.7]{Pictures/helmholtz-normalized-opvect-phi.pdf}
+\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel on the Xeon Phi. CM stands for cost model.}
+\label{fig:opvect-helmholtz-speedup-phi}
+\end{figure}
+
+\begin{figure}[h]
 \includegraphics[scale=0.7]{Pictures/diffusion-normalized-opvect.pdf}
-\caption{Speed-ups obtained by applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel.
-Tri, Tet, and Prism stand, respectively, for triangual, tetrahedral and prismatic mesh.
-CM stands for cost model.}
+\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel on the Sandy Bridge. CM stands for cost model.}
 \label{fig:opvect-diffusion-speedup}
@@
 
+%for which efficient register allocation can be already guaranteed
+
 \subsection{Impact of Expression Splitting}
 \label{sec:perf-results-split}
-Expression splitting, or loop fission, contributes to relieve register pressure.
-The transformation is useless if the element matrix can be evaluated by accessing a relatively small number of basis functions (or temporary values introduced by \emph{licm}), for which efficient register allocation can be already guaranteed.
-In the cases of Helmholtz and Diffusion, irrespective of split factor and cut, the only consequence of applying \emph{split} is to augment the number of increments for evaluating the element matrix.
-On the Sandy Bridge machine, slow downs up to 1.4$\times$ were observed in these two problems.
+Expression splitting relieves the register pressure when the element matrix evaluation reads from a large number of arrays, at the price of increasing accesses to the element matrix itself and of potentially affecting data locality, as detailed in Section~\ref{sec:expr-split}.
+It is not particularly useful, therefore, for Helmholtz and Diffusion, where only between 4 and 8 basis functions or temporaries need to be read at every iteration.
+Slow downs up to 1.4$\times$ were observed in these two kernels on the Sandy Bridge.
 The cost model, however, prevents the adoption of this transformation, since the \texttt{while} statement at line 5 in Figure~\ref{algo:applyCostModel} is never entered.
-In the Burgers problem, \emph{split} allows notable speed-ups to be achieved over the best run-time provided by either \emph{licm} or \emph{licm-ap}.
-Results are shown in Figure~\ref{fig:split-burgers-speedup}.
-In almost all problem configurations, a split factor of 1 induced optimal performance, meaning that the original expression was divided into two parts, with one part performing a slightly larger number of floating point operations than the other.
-Register locality could have been affected by requiring the same array be re-loaded twice as before splitting.
-However, in this experiment, this was only partly true, because most of the arrays appear in one sub-expression only.
+In the case of Burgers, on the contrary, \emph{split} allows notable performance improvement to be achieved.
+Figure~\ref{fig:split-burgers-speedup} shows the normalized run-time over the fastest implementation between \emph{licm} and \emph{licm-ap}.
+In almost all test cases, a split factor of 1, meaning that the original expression was divided into two parts performing a similar number of floating point operations, induced the optimal perforance.
+Register locality was only marginally affected.
 
 The \emph{split} transformation was also tried in combination with \emph{op-vect} (\emph{split-op-vect}), although the cost model prevents the application of the latter on the Sandy Bridge.
-Despite performance improvements up to 1.22$\times$, \emph{split-op-vect} never outperforms \emph{split}.
-This is because of two reasons: for small split factors, like 1 and 2, the data space to be tiled is still too big, and register spilling still affects the run-time; for higher ones, sub-expressions become so small that, as explained in Section~\ref{sec:perf-eval-opvect}, full or partial unrolling of the loop nest allows achieving a certain degree of register re-use.
+Despite improvements up to 1.22$\times$, \emph{split-op-vect} never outperforms \emph{split}.
+This is because of two reasons: for small split factors, like 1 and 2, the data space to be tiled is still too big, and register spilling affects run-time; for higher ones, sub-expressions become so small that, as explained in Section~\ref{sec:perf-eval-opvect}, full or partial unrolling of the loop nest allow achieving a certain degree of register re-use themselves.
 
@@
 \includegraphics[scale=0.7]{Pictures/burgers-normalized-split.pdf}
-\caption{Speed-ups obtained by applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel.
-Tri, Tet, and Prism stand, respectively, for triangual, tetrahedral and prismatic mesh.
-CM stands for cost model.}
+\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel on the Sandy Bridge. CM stands for cost model.}
 \label{fig:split-burgers-speedup}
@@
 
-\subsection{Comparison with Fenics Form Compiler's built-in Optimizations}
-We have modified the Fenics Form Compiler (FFC) to make it return an abstract syntax tree representation of a local assembly kernel, rather than plain C code, so as to enable code transformations in COFFEE.
+\subsection{Comparison with FEniCS Form Compiler's built-in Optimizations}
+We have modified the FEniCS Form Compiler (FFC) to make it return an abstract syntax tree representation of a local assembly kernel, rather than plain C code, so as to enable code transformations in COFFEE.
 Besides Firedrake, FFC is used in the FEniCS project~\cite{Fenics}.
-In FEniCS, FFC can apply its own model-driven optimizations to local assembly kernels~\cite{quadrature1}, which mainly consist of loop-invariant code motion and elimination, at code generation time, of floating point operations involving zero-valued columns in basis function arrays.
-The FEniCS' loop-invariant code motion is, however, definitely different from COFFEE's, since it is based on expansion of arithmetic operations to identify terms that are invariant of the whole loop nest.
-Thus, depending on the way expansion is performed, operation count does not necessarily decrease.
-On the other hand, elimination of zero-valued columns, which are due to certain properties of the input finite element problem, has the effect of introducing indirection arrays in the generated code.
-This kind of optimization is currently under development in COFFEE, although it will differ from FEniCS' by avoiding non-contiguous memory accesses, which would affect vectorization because of the need for gather operations.
-The price to pay is that only some of the zero-columns will be removed.
-
-Table~\ref{table:comparison-to-FFC-opt} summarizes the results of the comparison between FEniCS' (FFC) optimizations and COFFEE on the Sandy Bridge.
-In COFFEE, Burgers' slow downs occur in presence of a small iteration space (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$), although they are generally milder than the maximum one reported in table, obtained with a triangular mesh and $p = 1$.
-This is a result of zero-columns removal in FEniCS: indirection arrays prevent auto-vectorization, which significantly impacts performance as soon as the element matrix becomes bigger than 12$\times$12.
-However, with the forthcoming zero-removal optimization in COFFEE, we expect to outperform FEniCS even in these cases as well.
-In the cases of Helmholtz and Diffusion, the minimum speed-ups are, respectively, 1.10$\times$ and 1.18$\times$ (2D mesh, $p=1$), which keep increasing with polynomial order and element shape up to the values illustrated in the table.
+In FEniCS, FFC can apply its own model-driven optimizations to local assembly kernels~\cite{quadrature1}, which mainly consist of loop-invariant code motion and elimination, at code generation time, of floating point operations involving zero-valued entries in basis function arrays.
+The FEniCS' loop-invariant code motion is, however, totally different from COFFEE's, since it is based on expansion of arithmetic operations to identify terms that are invariant of the whole loop nest.
+In addition, depending on the way expansion is performed, operation count often does not decrease significantly.
+On the other hand, elimination of zero-valued terms, which are due to certain properties of the finite element problem, has the effect of introducing indirection arrays in the generated code.
+This kind of optimization is currently under development in COFFEE, although it will differ from FEniCS' by avoiding non-contiguous memory accesses, which would otherwise affect vectorization, at the price of removing less zero-valued contributions.
+
+Table~\ref{table:comparison-to-FFC-opt} summarizes the performance achived by COFFEE over the best FEniCS (FFC) implementation (i.e.
+the one that solves the problem fastest, regardless of using optimizations) on the Sandy Bridge for the Burgers, Helmholtz and Diffusion kernels.
+Burgers' slow downs occur in presence of a small iteration space (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$), although they are generally smaller than that reported in table, obtained with a triangular mesh and $p = 1$.
+This is a result of removing zero-valued entries in FEniCS' basis functions: some operations are avoided, but indirection arrays prevent auto-vectorization, which significantly impacts performance as soon as the element matrix becomes bigger than 12$\times$12.
+However, with the forthcoming zero-removal optimization in COFFEE, we expect to outperform FEniCS in all problems.
+In the cases of Helmholtz and Diffusion, the minimum improvements are, respectively, 1.10$\times$ and 1.18$\times$ (2D mesh, $p=1$), which tend to increase with polynomial order and element shape up to the values illustrated in the table.
 
@@
 Problem & Max slow down & Max speed up \\\hline\hline
-Helmholtz & / & 11.72$\times$ \\
-Diffusion & / & 14.68$\times$ \\
+Helmholtz & / & 4.14$\times$ \\
+Diffusion & / & 4.28$\times$ \\
 Burgers & 2.24$\times$ & 1.61$\times$ \\\hline
@@
 \end{center}
+\caption{Performance comparison between FEniCS (optimizations enabled) and COFFEE on the Sandy Bridge.}
 \label{table:comparison-to-FFC-opt}
-\caption{Performance comparison between FEniCS (optimizations enabled) and COFFEE on the Sandy Bridge.}
 \end{table}
@@
 \subsection{Comparison with hand-made BLAS-based implementations}
-For the Helmholtz 3D problem, manual implementations based on \texttt{Intel MKL BLAS} were tested on the Sandy Bridge.
+\label{sec:perf-eval-blas}
+For the Helmholtz problem on a tetrahedral mesh, manual implementations based on \texttt{Intel MKL BLAS} were tested on the Sandy Bridge.
 This particular kernel can be easily reduced to a sequence of four matrix-matrix multiplies that can be computed via calls to BLAS \texttt{dgemm}.
-In the case of $p=4$, where the element matrix is of size 35$\times$35, the computation was almost twice slower than the case in which only \emph{licm}, data alignment and padding were used, with the slow down being even worse for smaller problem sizes.
-These experiments justify that there is a set of problems for which turning to MKL is not beneficial in terms of performance.
-It is possible that employing certain BLAS implementations or auto-tuning strategies, particularly for certain problem sizes, is helpful, although the same problems afflicting \emph{split}, exposed in Section~\ref{sec:expr-split}, might limit the performance gain.
-However, as anticipated, the problem of identifying and extracting matrix-matrix multiplies from analysis of the kernel's AST, as well as a strategy for their execution, will be comprehensively addressed in further work.
+In the case of $p=4$, where the element matrix is of size 35$\times$35, the computation was almost twice slower than the case in which \emph{licm-ap} was used, with the slow down being even worse for smaller problem sizes.
+These experiments suggest that the question regarding to what extent linear algebra libraries or, analogously, auto-tuning strategies can improve performance cannot be trivially answered.
+This is due to a combination of issues: the potential loss in data locality as exposed in Section~\ref{sec:expr-split}, the actual effectiveness of external tools when the arrays are relatively small, and the problem inherent to assembly kernels concerning extraction of matrix-matrix multiplies from static analysis of the kernel's code.
+A comprehensive study of these aspects will be addressed in further work.
 
@@
 The finite element method is used in the most disparate contexts to approximate solutions of PDEs.
-Well-known frameworks and applications include nek5000~\cite{nek5000-web-page}, the Fenics project~\cite{Fenics}, Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake; this is not an exhaustive list, though.
+Well-known frameworks and applications include nek5000~\cite{nek5000-web-page}, the FEniCS project~\cite{Fenics}, Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake; this is not an exhaustive list, though.
 Numerical integration is usually employed to implement the local assembly phase.
@@
 In this paper we have presented design, optimizations and systematic performance evaluation of COFFEE, a compiler for finite element local assembly.
-In this context, to the best of our knowledge, this is the first compiler oriented towards the introduction of low-level optimizations to maximize instruction-level parallelism, register locality and SIMD vectorization.
+In this context, to the best of our knowledge, COFFEE is the first compiler oriented towards the introduction of low-level optimizations to maximize instruction-level parallelism, register locality and SIMD vectorization.
 Assembly kernels have peculiar characteristics.
 Their iteration space is usually very small, with the size depending on aspects like the degree of accuracy one wants to reach (polynomial order of the method) and the mesh discretization employed.
-The data space, in terms of number of matrices and scalar, grows proportionally with the complexity of the finite element problem.
+The data space, in terms of number of arrays and scalars required to evaluate the element matrix, grows proportionally with the complexity of the finite element problem.
 COFFEE has been developed taking into account all of these degrees of freedom, based on the idea that reducing the problem of local assembly optimization to a fixed sequence of transformations is way too superficial if close-to-peak performance needs to be reached.
@@
 The exploitation of domain-knowledge allows some of them to be particularly effective, as demonstrated by our experiments on two state-of-the-art Intel platforms.
-Further work include a comprehensive study about feasibility and constraints on transforming the kernel into a sequence of BLAS calls.
+Further work include a comprehensive study about feasibility and constraints on transforming kernels into a sequence of calls to external linear algebra libraries.
 COFFEE supports all of the problems expressible in Firedrake, and it is already integrated with this framework.
@@
 The authors would like to thank Dr.
-Lawrence Mitchell and Dr.
+Carlo Bertolli, Dr.
+Lawrence Mitchell, and Dr.
 Francis Russell for their invaluable suggestions and their contribution to the Firedrake project.

### 2014-04-11 paper/bare_conf.tex
Add new plots

@@
 \label{sec:perf-eval-opvect}
-Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the performance improvements achieved applying \emph{op-vect} on top of \emph{licm-ap} to the Helmoltz and Diffusion kernels on the Sandy Bridge, whereas Figure~\ref{fig:opvect-helmholtz-speedup-phi} shows analogous results for Helmholtz on the Xeon Phi.
+Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the performance improvements achieved applying \emph{op-vect} on top of \emph{licm-ap} to the Helmoltz and Diffusion kernels on the Sandy Bridge, whereas Figures~\ref{fig:opvect-helmholtz-speedup-phi} and~\ref{fig:opvect-diffusion-speedup-phi} illustrate analogous results for the Xeon Phi.
 For each problem instance we report two bars: one indicates the best run-time obtained by auto-tuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved with the COFFEE's cost model.
 In general, there is no substantial difference between the two.
-This is chiefly because assembly kernels fit the L1 cache, so estimating the run-time based on the quality of register allocation is relatively simple.
-
-The rationale behind these results is that, for smaller configurations, vector-register tiling is rarely helpful.
+This is chiefly because these assembly kernels fit the L1 cache, so estimating the run-time based on the quality of register allocation is relatively simple.
+
+The rationale behind these results is that, for smaller configurations, vector-register tiling is marginally helpful.
 This is due to the fact that in the considered problems the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short.
-On the other hand, as the iteration space becomes bigger, vector-register tiling leads to percentage improvements up to 1.4$\times$ on the Sandy Bridge (Diffusion, prismatic mesh, $p=4$) and up to X$\times$ on the Xeon Phi (X, X, X).
+On the other hand, as the iteration space becomes bigger, vector-register tiling leads to percentage improvements up to 1.4$\times$ on the Sandy Bridge (Diffusion, prismatic mesh, $p=4$) and up to 1.4$\times$ on the Xeon Phi (Helmholtz, tetrahedral mesh, $p=3$).
 Using the Intel Architecture Code Analyzer tool~\cite{IACA}, we proved that this is a consequence of increased register re-use.
@@
 \label{fig:opvect-diffusion-speedup}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[scale=0.7]{Pictures/diffusion-normalized-opvect-phi.pdf}
+\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel on the Xeon Phi. CM stands for cost model.}
+\label{fig:opvect-diffusion-speedup-phi}
 \end{figure}

### 2014-04-11 paper/bare_conf.tex
Incorporate Ana's suggestions

@@
 
-We focus on local assembly (``assembly'', in the following), a fundamental step of a finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$.
+We focus on improving the performance of local assembly (``assembly'', in the following), a fundamental step of a finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$.
 During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain.
@@
 The complexity of the mathematical expressions, often characterized by a large number of operations on constants and small matrices, makes it hard to determine a single or specific sequence of transformations that is successfully applicable to all problems.
-The variation in loop trip counts and their typically small value, which prevents the adoption of standard techniques for cache locality, further exacerbates the issue.
+The variation in loop trip counts and their typically small value further exacerbates the issue.
 A compiler-based approach is, therefore, the only reasonable option to obtain close-to-peak performance in a wide range of different local assembly kernels.
-Optimizations like generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but they are not supported by state-of-the-art polyhedral and vendor compilers.
-BLAS routines could be theoretically employed, although a fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies.
+Optimizations like generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but none are supported by state-of-the-art polyhedral and vendor compilers.
+BLAS routines could be theoretically employed, although a fairly complicated control- and data-flow analysis is required to automate identification and extraction of matrix-matrix multiplies.
 In addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small dimension of the involved matrices and the potential loss in data locality can limit the performance gain, if any.
@@
 It supports any problems expressible with this framework.
-This allows us to evaluate our code transformations in a range of real-world problems, varying key parameters that impact both solution accuracy and kernel cost, namely the polynomial order of the method (from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).
-
-Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal.
+This allows us to evaluate our code transformations in a range of real-world problems, varying two key parameters that impact both solution accuracy and kernel cost: the polynomial order of the method (from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).
+
+Our experiments showed that the originally generated code for non-trivial assembly kernels was sub-optimal.
 Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (i.e.
-code where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels.
+code where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.4$\times$ over original kernels.
 The contributions of this paper are threefold
@@
 \item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
-\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
+\item Systematic analysis using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
 \end{itemize}
@@
 Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm.
-It then triggers compilation of such kernels using an available vendor compiler, and eventually manages parallel execution over the mesh.
+It then triggers compilation of such kernels using a vendor compiler, and eventually manages parallel execution over the mesh.
 The subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.
@@
 The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}.
-Input are a pointer to a zero-initialized two dimensional array used to store the element matrix, a pointer to the element's coordinates in the discretized domain, and pointers to data of coefficient fields, for instance indicating value of velocity or pressure in the element.
+Input are a zero-initialized two dimensional array used to store the element matrix, the element's coordinates in the discretized domain, and coefficient fields, for instance indicating value of velocity or pressure in the element.
 The output is the evaluated element matrix.
@@
 All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (i.e.
-using \texttt{static const} in C language).
+using \texttt{static const} in C).
 Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel's objects.
@@
 Object name & Type & Variable name(s) \\\hline\hline
-Determinant of Jacobian matrix & double & det  \\ \hline
-Inverse of Jacobian matrix & double[] & K \\ \hline
+Determinant of the Jacobian matrix & double & det  \\ \hline
+Inverse of the Jacobian matrix & double[] & K \\ \hline
 Coordinates & double** & coords\\ \hline
@@
 Note that in addition to a larger number of operations to compute the element matrix, the Burgers case shows a register pressure higher than that in Helmholtz.
+
 Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.
-
-Some peculiarities of assembly kernels are: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g.
+These include: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g.
 the $F$ values in Burgers) and introduction of temporary arrays; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
@@
 One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place.
-In rare, yet important cases, however, this might not be sufficient to achieve notable performance improvements.
+In rare cases (e.g.
+problems based on Hyperelasticity), however, such optimizations might not be sufficient to achieve notable performance improvements.
 Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions.
@@
 In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure).
-Given its inherent complexity, a comprehensive study of this problem is left as further work.
+A comprehensive study of such cases is left as further work.
 
 As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens iterations, render the optimization process challenging.
-Relatively to machine peak, while some problems achieve good performance when applying a very specific set of transformations, there is also an ample space of cases for which a more elaborated optimization strategy is required.
+It is not always the same set of optimizations that bring performance closest to the machine peak.
 For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces.
@@
 In this context, our goal is to understand the relationship between distinct code transformations, their impact on local assembly kernels, and to what extent their composability is effective in a class of problems and architectures.
+
+% TODO: could add a sentence here to bridge the two paragraphs
 
@@
 Then, two dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$.
-For instance, assume the original size of a basis function array is 3$\times$3 and that the underlying architecture possesses AVX, which implies $vl=4$ since a vector register is 256 bits long and our kernels use 64-bits double-precision floating-point values.
-In this example, a padded version of the array will have size 3$\times$4.
+For instance, assume the original size of a basis function array is 3$\times$3 and $vl=4$ (e.g.
+AVX processor, with 256 bits long vector registers and 64-bits double-precision floating-point values).
+In this case, a padded version of the array will have size 3$\times$4.
 The compiler is explicitly informed about data alignment using a suitable pragma.
@@
 \label{sec:code-transf-licm}
-From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only.
+From the inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only.
 Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion.
@@
 This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest.
-At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop.
+At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, as part of an inner loop.
 Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant.
@@
 Based on the observation that the evaluation of the element matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented.
-If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, we can notice that the computation of the element matrix is abstractly expressible as
+If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, the computation of the element matrix is abstractly expressible as
 \begin{equation}
@@
   y \in B'' \subseteq B}}
-x_j\cdot y_k ~~~~~~ j,k = 0,...,4
+x_j\cdot y_k ~~~~~~ j,k = 0,...,3
 \end{equation}
@@
 The figure shows how 4$\times$4 elements in the top-left block of the element matrix $A$ can be moved to their correct positions.
-Each rotation, represented by a group of three same-colored arrows, is implemented by a single vector shuffle intrinsics.}
+Each rotation, represented by a group of three same-colored arrows, is implemented by a single shuffle intrinsics.}
 \label{fig:restore-layout}
@@
 \label{sec:expr-split}
-In complex kernels, like Burgers' in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
-If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, it is likely to obtain poor register reuse.
+In complex kernels, like Burgers in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
+If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, poor register reuse is likely to be obtained.
 This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large.
-For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the \texttt{ijk} loop order, and compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers.
+For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the \texttt{ijk} loop order, so compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers.
 One potential solution to this problem consists of suitably ``splitting'' the computation of the element matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}.
-Splitting an expression has, in general, several drawbacks.
+
+Splitting an expression has three drawbacks.
 Firstly, it increases the number of accesses to $A$ proportionally to the ``split factor'', which is the number of sub-expressions produced.
@@
 the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}).
-Finally, it might affect register locality, although this is not the case of the Helmholtz example: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
+Finally, it might affect register locality: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
+This is not the case of the Helmholtz example.
 Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.
@@
 The output of COFFEE, C code, is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and execution over the discretized domain take place.
-Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem.
+Because of the large number of parametric transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem.
 Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of ASTs occurs at run-time as soon as problem-specific data are available.
@@
 For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance.
-Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required for every iteration.
-Experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop.
+Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required at every iteration.
+We tested this hyphotesis in manually crafted kernels.
+We found that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop.
 The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension.
@@
 However, from our experience of inspecting assembly code and comparing with other hand-written implementations, recent versions of a vendor compiler like Intel's employ cost models capable of estimating close-to-optimal values for such parameters.
-This is particularly true for assembly kernels, where the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are unit-stride.
+This is particularly true for assembly kernels, where the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are stride-1.
 We therefore leave the backend compiler in charge of selecting the unroll factor.
-This choice also simplifies the COFFEE's cost model.
+This choice also simplifies COFFEE's cost model.
 The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code seems to prevent the Intel compiler from applying this optimization, even if specific pragmas are added.
-Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 19 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order \texttt{ijk}).
+
+%Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 19 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order \texttt{ijk}).
 
@@
 
-Experiments were run on two Intel architectures, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and the Xeon Phi.
+Experiments were run on two Intel architectures, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and a Xeon Phi (5110P, running at 1.05Ghz in native mode, 32KB L1 cache and 512KB L2 cache).
+We have chosen these two architectures because of the differences in the number of logical registers and SIMD lanes, which can impact the effectiveness of the optimization strategy.
 The \texttt{icc 13.1} compiler was used.
@@
 \end{itemize}
-The code was written in UFL and then executed over real unstructured meshes through Firedrake.
+The code was written in UFL (available at~\cite{ufl-code}) and then executed over real unstructured meshes through Firedrake.
 The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}.
-For Advection-Diffusion, the ``Diffusion'' equation, which uses the same differential operators as Helmholtz, is considered.
+For Advection-Diffusion, the Diffusion equation, which uses the same differential operators as Helmholtz, is considered.
 In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$.
-Burgers is a non-linear problem employing differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are accessed.
+Burgers is a non-linear problem employing differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (see Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are needed.
 
 These problems were studied varying both the shape of mesh elements and the polynomial order $p$ of the method, whereas the element family, Lagrange, is fixed.
-Intuitively, the bigger the element shape and $p$, the larger is the iteration space.
+Intuitively, the bigger are the element shape and $p$, the larger is the iteration space.
 Triangles, tetrahedron, and prisms were tested as element shape.
@@
 A range of polynomial orders, from $p=1$ to $p=4$, were tested; higher polynomial orders are excluded from the study because of current Firedrake limitations.
-In such a large space of problems, the size of the element matrix rarely exceeds 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
+In all these cases, the size of the element matrix rarely exceeds 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
 
@@
 The element family is Lagrange.
-Each entry indicates the minimum and maximum performance improvements obtained over the non-optimized implementation in the range of $p$ values.
+Each entry indicates the minimum and maximum performance improvements over the non-optimized code obtained in the range of $p$ values.
 The column \emph{licm-ap} illustrates the combination of \emph{licm} with data alignment and padding.
@@
 
-Table~\ref{table:perf-results-licm} illustrates the performance improvement obtained on the Sandy Bridge and the Xeon Phi machines when \emph{licm}, data alignment, and padding are used, over non-transformed code.
+Table~\ref{table:perf-results-licm} illustrates the performance improvement obtained on the Sandy Bridge and Xeon Phi machines when \emph{licm}, data alignment, and padding are used.
 We distinguish between \emph{licm} and \emph{licm-ap}; the latter makes use of padding and data alignment.
-Inspection of assembly code generated by \texttt{icc} confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant with respect to outer loops are hoisted and, interestingly, not vectorized.
+Inspection of assembly code generated by \texttt{icc} confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant to the inner loop are hoisted and their execution is not vectorized.
 Padding and data alignment enhance, in general, the quality of auto-vectorization.
-Sometimes the run-time of \emph{licm-ap} is similar to that of \emph{licm} because the element matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned.
+Sometimes the run-time of \emph{licm-ap} is similar to that of \emph{licm} because the element matrix size is already a multiple of the vector length, so no scalar reminder loop has to be introduced.
 Occasionally \emph{licm-ap} is slower than \emph{licm} (e.g.
-in Burgers 3D p3).
+in Burgers 3D p3 on the Sandy Bridge).
 One possible explanation is that the large number of aligned temporaries introduced by \emph{licm} induce cache associativity conflicts.
@@
 \label{sec:perf-eval-opvect}
-Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the performance improvements achieved applying \emph{op-vect} on top of \emph{licm-ap} to the Helmoltz and Diffusion kernels on the Sandy Bridge, whereas Figures~\ref{fig:opvect-helmholtz-speedup-phi} and~\ref{fig:opvect-diffusion-speedup-phi} illustrate analogous results for the Xeon Phi.
+Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the performance achieved applying \emph{op-vect} on top of \emph{licm-ap} to the Helmoltz and Diffusion kernels on the Sandy Bridge, whereas Figures~\ref{fig:opvect-helmholtz-speedup-phi} and~\ref{fig:opvect-diffusion-speedup-phi} illustrate analogous results for the Xeon Phi.
 For each problem instance we report two bars: one indicates the best run-time obtained by auto-tuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved with the COFFEE's cost model.
 In general, there is no substantial difference between the two.
-This is chiefly because these assembly kernels fit the L1 cache, so estimating the run-time based on the quality of register allocation is relatively simple.
-
-The rationale behind these results is that, for smaller configurations, vector-register tiling is marginally helpful.
-This is due to the fact that in the considered problems the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short.
+This is chiefly because these kernels fit the L1 cache, so, within a certain degree of confidence, it is possible to predict the fastest implementation based on register allocation.
+
+The rationale behind these results is that, for smaller configurations, vector-register tiling is marginally helpful, while its impact is determinant for larger problems.
+This is due to the fact that in the considered kernels the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short.
 On the other hand, as the iteration space becomes bigger, vector-register tiling leads to percentage improvements up to 1.4$\times$ on the Sandy Bridge (Diffusion, prismatic mesh, $p=4$) and up to 1.4$\times$ on the Xeon Phi (Helmholtz, tetrahedral mesh, $p=3$).
-Using the Intel Architecture Code Analyzer tool~\cite{IACA}, we proved that this is a consequence of increased register re-use.
+Using the Intel Architecture Code Analyzer tool~\cite{IACA} on the Sandy Bridge, we proved that this is a consequence of increased register re-use.
 In Helmholtz $p=4$, for example, the tool showed that when using \emph{op-vect} the number of clock cycles to execute one iteration of the $j$ loop decreases by roughly 17$\%$, and that this is a result of the relieved pressure on both of the data (cache) ports available in the core.
@@
 
-Application of \emph{op-vect} to the Burgers problem induces meaningful slow downs, due to the large number of temporary arrays that need to be tiled, which exceeds the number of available logical registers on the underlying architecture.
+Application of \emph{op-vect} to the Burgers problem induces significant slow downs due to the large number of temporary arrays that need to be tiled, which exceeds the available logical registers on the underlying architecture.
 Expression splitting can be used in combination to \emph{op-vect} to alleviate this issue; this is discussed in Section~\ref{sec:perf-results-split}.
@@
 \label{sec:perf-results-split}
-Expression splitting relieves the register pressure when the element matrix evaluation reads from a large number of arrays, at the price of increasing accesses to the element matrix itself and of potentially affecting data locality, as detailed in Section~\ref{sec:expr-split}.
+Expression splitting relieves the register pressure when the element matrix evaluation reads from a large number of arrays, at the price of increasing accesses to the element matrix itself and potentially affecting data locality, as detailed in Section~\ref{sec:expr-split}.
 It is not particularly useful, therefore, for Helmholtz and Diffusion, where only between 4 and 8 basis functions or temporaries need to be read at every iteration.
-Slow downs up to 1.4$\times$ were observed in these two kernels on the Sandy Bridge.
-The cost model, however, prevents the adoption of this transformation, since the \texttt{while} statement at line 5 in Figure~\ref{algo:applyCostModel} is never entered.
-In the case of Burgers, on the contrary, \emph{split} allows notable performance improvement to be achieved.
-Figure~\ref{fig:split-burgers-speedup} shows the normalized run-time over the fastest implementation between \emph{licm} and \emph{licm-ap}.
-In almost all test cases, a split factor of 1, meaning that the original expression was divided into two parts performing a similar number of floating point operations, induced the optimal perforance.
-Register locality was only marginally affected.
-
-The \emph{split} transformation was also tried in combination with \emph{op-vect} (\emph{split-op-vect}), although the cost model prevents the application of the latter on the Sandy Bridge.
+Slow downs up to 1.4$\times$ and up to 1.6$\times$ were observed, respectively, on the Sandy Bridge and the Xeon Phi.
+The cost model, however, prevents the adoption of the transformation: the \texttt{while} statement at line 5 in Figure~\ref{algo:applyCostModel} is indeed never entered.
+For the Burgers kernel, \emph{split} has a key role on the Sandy Bridge, whereas it has only a minimal impact on the Xeon Phi.
+Figure~\ref{fig:split-burgers-speedup} and~\ref{fig:split-burgers-speedup-phi} show the improvement due to using \emph{split} over the fastest implementation between \emph{licm} and \emph{licm-ap}.
+On the Sandy Bridge, in almost all test cases, a split factor of 1, meaning that the original expression was divided into two parts, ensured close-to-peak perforance.
+The transformation negligbly affected register locality.
+On the contrary, the larger number of vector units present on the Xeon Phi significantly relieve register spilling, making \emph{split} only marginally useful for loop unrolling.
+
+The \emph{split} transformation was also tried in combination with \emph{op-vect} (\emph{split-op-vect}), although the cost model prevents its adoption on both platforms.
 Despite improvements up to 1.22$\times$, \emph{split-op-vect} never outperforms \emph{split}.
@@
 \label{fig:split-burgers-speedup}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[scale=0.7]{Pictures/burgers-normalized-split-phi.pdf}
+\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel on the Xeon Phi. CM stands for cost model.}
+\label{fig:split-burgers-speedup-phi}
 \end{figure}

### 2014-04-11 paper/bare_conf.tex
Re-write related work section

@@
 Their implementation is based on analysis and transformation of the kernel's AST.
-When \emph{op-vect} is selected, the compiler outputs proper AVX or LRBNi intrinsics code.
+When \emph{op-vect} is selected, the compiler outputs proper AVX or AVX-512 intrinsics code.
 Any possible corner cases are handled: for example, if \emph{op-vect} is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.
@@
 \subsection{Comparison with FEniCS Form Compiler's built-in Optimizations}
+\label{sec:perf-results-fenics}
 We have modified the FEniCS Form Compiler (FFC) to make it return an abstract syntax tree representation of a local assembly kernel, rather than plain C code, so as to enable code transformations in COFFEE.
@@
 \label{sec:related-work}
-The finite element method is used in the most disparate contexts to approximate solutions of PDEs.
-Well-known frameworks and applications include nek5000~\cite{nek5000-web-page}, the FEniCS project~\cite{Fenics}, Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake; this is not an exhaustive list, though.
+The finite element method is extensively used to approximate solutions of PDEs.
+Well-known frameworks and applications include nek5000~\cite{nek5000-web-page}, the FEniCS project~\cite{Fenics}, Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake.
 Numerical integration is usually employed to implement the local assembly phase.
 The recent introduction of DSLs to decouple the finite element specification from its underlying implementation facilitated, however, the development of novel approaches.
-Methods based on tensor contraction~\cite{FFC-TC} and symbolic manipulation~\cite{Francis} have been developed, although it has been demonstrated that numerical integration remains the optimal choice for a wide class of problems~\cite{quadrature1}.
-
-Optimization of local assembly using numerical integration for CPU platforms has been tackled in~\cite{quadrature1}.
-However, to the best of our knowledge, ours is the first work targeting low-level optimizations and adopting a real compiler approach.
+Methods based on tensor contraction~\cite{FFC-TC} and symbolic manipulation~\cite{Francis} have been developed.
+We have designed COFFEE to specifically target numerical integration because it has been demonstrated that it remains the optimal choice for a wide class of problems~\cite{quadrature1}.
+
+Optimization of local assembly using numerical integration for CPU platforms has been addressed in FEniCS~\cite{quadrature1}.
+The comparison between COFFEE and this work is presented in Section~\ref{sec:perf-results-fenics}.
 In~\cite{Markall20101815}, and more recently in~\cite{petsc-integration-gpu}, the problem has been studied for GPU architectures.
-In~\cite{assembly-opencl}, variants of the standard numerical integration algorithm have been specialized for the PowerXCell processor and evaluated; the paper lacks, however, an exhaustive study from the compiler viewpoint as we did, and none of the optimizations presented in Section~\ref{sec:code-transf} are mentioned.
-
-Domain-specific languages and automated code generation have been adopted with success in different areas.
-Spiral~\cite{Pueschel:05} automates generation of highly-optimized platform-specific digital signal processing numerical algorithms.
-Similar ideas are adopted in~\cite{Spampinato:14}, where a DSL and a compiler for dense linear algebra kernels are proposed.
-OP2~\cite{op2-main}, and its python implementation~\cite{pyop2isc}, which is used by Firedrake to express iteration over meshes, aim at achieving performance portability for scientific codes based on unstructured meshes.
-Stencil DSLs and compilers, such as Pochoir~\cite{pochoir} and SDSL~\cite{stencil-compiler}, have been developed to support development and high performance in fields like image processing and scientific computations over structured grids.
-
-COFFEE currently uses a simple cost model and heuristics to steer the optimization process.
+In~\cite{assembly-opencl}, variants of the standard numerical integration algorithm have been specialized and evaluated for the PowerXCell processor, but an exhaustive study from the compiler viewpoint - like ours - is missing, and none of the optimizations presented in Section~\ref{sec:code-transf} are mentioned.
+Among these efforts, to the best of our knowledge, COFFEE is the first work targeting low-level optimizations through a real compiler approach.
+
+Our compiler-based optimization approach is made possible by the DSL on top, which enables automated code generation.
+DSLs have been proven successful in auto-generating optimized code for other domains: Spiral~\cite{Pueschel:05} for digital signal processing numerical algorithms, ~\cite{Spampinato:14} for dense linear algebra, or Pochoir~\cite{pochoir} and SDSL~\cite{stencil-compiler} for image processing and scientific computations over structured grids.
+Similarly, PyOP2 is used by Firedrake to express iteration over unstructured meshes in scientific codes.
+COFFEE improves automated code generation in Firedrake.
+
 Many code generators, like those based on the Polyhedral model~\cite{PLUTO} and those driven by domain-knowledge~\cite{modeldriven}, make use of cost models.
-The opposite extreme consists of relying on auto-tuning to select the best implementation for a given problem on a certain platform.
-Notable examples include tuning of small matrix-matrix multiplies in nek5000~\cite{nek5000}, the ATLAS library~\cite{ATLAS}, and FFTW~\cite{FFTW} for fast fourier transforms.
-In both cases, pruning the implementation space, as we did in Section~\ref{sec:pyop2-compiler}, is fundamental to mitigate complexity and overhead.
+The alternative of using auto-tuning to select the best implementation for a given problem on a certain platform has been adopted by nek5000~\cite{nek5000} for small matrix-matrix multiplies, the ATLAS library~\cite{ATLAS}, and FFTW~\cite{FFTW} for fast fourier transforms.
+In both cases, pruning the implementation space is fundamental to mitigate complexity and overhead.
+Likewise, COFFEE uses a cost model and heuristics (Section~\ref{sec:pyop2-compiler}) to steer the optimization process.
 
@@
 
-In this paper we have presented design, optimizations and systematic performance evaluation of COFFEE, a compiler for finite element local assembly.
+In this paper, we have presented design, optimizations and systematic performance evaluation of COFFEE, a compiler for finite element local assembly.
 In this context, to the best of our knowledge, COFFEE is the first compiler oriented towards the introduction of low-level optimizations to maximize instruction-level parallelism, register locality and SIMD vectorization.

### 2014-04-14 paper/bare_conf.tex
Incorporate Paul's suggestions

@@
 It then generates C code including vector intrinsics.
-Experiments using a suite of real-world examples show that significant performance improvement is achieved.
+Experiments using a range of finite-element forms of increasing complexity show that significant performance improvement is achieved.
 \end{abstract}
@@
 To address this problem, domain-specific languages (DSLs) have been developed.
-The successful porting of Hydra, a computational fluid dynamics industrial application devised by Rolls Royce for turbomachinery design (based on Finite Volume Method, roughly 50000 lines of code and mesh sizes that can be over 100 millions edges), to OP2~\cite{op2-main}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}.
+The successful porting of Hydra, an industrial computational fluid dynamics application developed by Rolls Royce for turbomachinery design (based on finite volume method, roughly 50,000 lines of code and with meshes in excess of 100M edges), to OP2~\cite{op2-main}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}.
 
@@
 
-We focus on improving the performance of local assembly (``assembly'', in the following), a fundamental step of a finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$.
+We focus on improving the performance of local assembly (``assembly'', in the following), a fundamental step of the finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$.
 During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain.
@@
 Our experiments showed that the originally generated code for non-trivial assembly kernels was sub-optimal.
-Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (i.e.
-code where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.4$\times$ over original kernels.
-The contributions of this paper are threefold
+Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over code where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment, and up to 4.4$\times$ over original kernels.
+The contributions of this paper are
 \begin{itemize}
@@
 Section~\ref{sec:perf-results} shows performance results.
-Related work are illustrated in Section~\ref{sec:related-work}, while Section~\ref{sec:conclusions} concludes the paper.
+Related work is illustrated in Section~\ref{sec:related-work}, while Section~\ref{sec:conclusions} reviews our contributions in the light of our results, and identifies priorities for future work.
 
@@
 This operation is applied to all cells in the discretized domain.
-In this work we focus on local matrices, or ``element matrices'', which are more costly to compute.
+In this work we focus on local matrices, or ``element matrices'', which are more costly to compute than element vectors.
 
@@
 The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}.
-Input are a zero-initialized two dimensional array used to store the element matrix, the element's coordinates in the discretized domain, and coefficient fields, for instance indicating value of velocity or pressure in the element.
+Input is a zero-initialized two dimensional array used to store the element matrix, the element's coordinates in the discretized domain, and coefficient fields, for instance indicating value of velocity or pressure in the element.
 The output is the evaluated element matrix.
@@
 An example is provided in Figure~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$.
-In other scenarios, for instance when solving a non-linear problem, like Burgers in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
+In other scenarios, for instance when solving a non-linear problem like the Burgers equation, whose assembly code is given in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
 Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse.
-Note that in addition to a larger number of operations to compute the element matrix, the Burgers case shows a register pressure higher than that in Helmholtz.
+In addition to a larger number of operations, more complex cases like the Burgers equation are characterized by high register pressure.
 
 Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.
-These include: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g.
-the $F$ values in Burgers) and introduction of temporary arrays; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
+These include: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas interchanges involving the $i$ loop require pre-computation of values (e.g.
+the $F$ values in Burgers) and introduction of temporary arrays, as explained in Section~\ref{sec:code-transf}; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
 In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.
@@
 One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place.
-In rare cases (e.g.
-problems based on Hyperelasticity), however, such optimizations might not be sufficient to achieve notable performance improvements.
+In rare cases, as in problems involving hyperelasticity, however, such optimizations might not be sufficient to achieve notable performance improvements.
 Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions.
@@
 We want to overcome two deficiencies that both $intel$ and $gnu$ compilers have.
-First, they can identify sub-expressions that are invariant with respect to the innermost loop only.
+First, they identify sub-expressions that are invariant with respect to the innermost loop only.
 This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted in the loop order \texttt{ijk}.
@@
 One notable problem of assembly kernels concerns register allocation and register locality.
-The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high.
+The critical situation occurs when loop trip counts and the variables accessed are such that the vector-registers pressure is high.
 Since the kernel's working set fits the L1 cache, it is remarkably important to optimize register management.
-Canonical optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem.
+Standard optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem.
 In COFFEE, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll).
@@
 ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation.
-Besides a larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.
+In addition to incurring in larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.
 
@@
 \begin{figure}[t]
-\centerline{\includegraphics[scale=0.7]{Pictures/vect-restore.pdf}}
+\centerline{\includegraphics[scale=0.60]{Pictures/vect-restore.pdf}}
 \caption{Restoring the storage layout after \emph{op-vect}.
@@
 Firstly, it increases the number of accesses to $A$ proportionally to the ``split factor'', which is the number of sub-expressions produced.
-Secondly, depending on how the split is executed, it can lead to redundant computation (e.g.
-the product $det*W3[i]$ is performed times number of sub-expressions in the code of Figure~\ref{code:helmholtz-split}).
+Secondly, depending on how the split is executed, it can lead to redundant computation; for example, the times the product $det*W3[i]$ is performed is proportional to the number of sub-expressions, as shown in the code snippet.
 Finally, it might affect register locality: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
@@
 The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}.
-Initially, an AST is inspected, looking for the presence of iteration spaces and domain-specific information provided by the higher layer.
-If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied.
+Initially the AST is inspected, looking for the presence of iteration spaces and domain-specific information provided by the higher layer.
+If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the outer, non-affine loop over mesh elements is vectorized, can be applied.
 This feature, currently under development, has been proved to be useful in several finite volume applications~\cite{inter-kernel-vect}.
@@
 When \emph{op-vect} is selected, the compiler outputs proper AVX or AVX-512 intrinsics code.
-Any possible corner cases are handled: for example, if \emph{op-vect} is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.
+Any possible corner cases are handled: for example, if \emph{op-vect} is to be applied, but the size of the iteration space is not a multiple of the vector length, then a remainder loop, amenable to auto-vectorization, is inserted.
 
@@
 Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required at every iteration.
-We tested this hyphotesis in manually crafted kernels.
+We tested this hypothesis in manually crafted kernels.
 We found that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop.
 The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension.
-Our compiler chooses the loop along which the number of invariant loads is smaller as outer so that more registers remain available to carry out the computation of the element matrix.
+Our compiler chooses as outermost the loop along which the number of invariant loads is smaller so that more registers remain available to carry out the computation of the element matrix.
 
@@
 We have chosen these two architectures because of the differences in the number of logical registers and SIMD lanes, which can impact the effectiveness of the optimization strategy.
+All experiments are sequential.
 The \texttt{icc 13.1} compiler was used.
@@
 These problems were studied varying both the shape of mesh elements and the polynomial order $p$ of the method, whereas the element family, Lagrange, is fixed.
-Intuitively, the bigger are the element shape and $p$, the larger is the iteration space.
+As might be expected, the larger the element shape and $p$, the larger the iteration space.
 Triangles, tetrahedron, and prisms were tested as element shape.
@@
 ~
-\caption{Impact of generalized loop-invariant code motion (\emph{licm} column) on the Helmholtz, Diffusion and Burgers problems, varying the element shape (triangle, tetrahedron, prism) and the polynomial order ($p \in [1, 4]$).
-The element family is Lagrange.
-Each entry indicates the minimum and maximum performance improvements over the non-optimized code obtained in the range of $p$ values.
+\caption{Performance improvement due to generalized loop-invariant code motion (\emph{licm} column) on the Helmholtz, Diffusion and Burgers problems, varying the element shape (triangle, tetrahedron, prism) and the polynomial order ($p \in [1, 4]$), using the Lagrange element family, over the original, non-optimized code.
+Each entry indicates the minimum and maximum speedup obtained in the range of $p$ values.
 The column \emph{licm-ap} illustrates the combination of \emph{licm} with data alignment and padding.
@@
 Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the performance achieved applying \emph{op-vect} on top of \emph{licm-ap} to the Helmoltz and Diffusion kernels on the Sandy Bridge, whereas Figures~\ref{fig:opvect-helmholtz-speedup-phi} and~\ref{fig:opvect-diffusion-speedup-phi} illustrate analogous results for the Xeon Phi.
-For each problem instance we report two bars: one indicates the best run-time obtained by auto-tuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved with the COFFEE's cost model.
+For each problem instance we report two bars: one indicates the best run-time obtained by auto-tuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved with COFFEE's cost model.
 In general, there is no substantial difference between the two.
@@
 
-The rationale behind these results is that, for smaller configurations, vector-register tiling is marginally helpful, while its impact is determinant for larger problems.
-This is due to the fact that in the considered kernels the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short.
+The rationale behind these results is that, for smaller configurations, vector-register tiling is marginally helpful, while its impact is the dominant factor for larger problems.
+This is because in the considered kernels the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short.
 On the other hand, as the iteration space becomes bigger, vector-register tiling leads to percentage improvements up to 1.4$\times$ on the Sandy Bridge (Diffusion, prismatic mesh, $p=4$) and up to 1.4$\times$ on the Xeon Phi (Helmholtz, tetrahedral mesh, $p=3$).
-Using the Intel Architecture Code Analyzer tool~\cite{IACA} on the Sandy Bridge, we proved that this is a consequence of increased register re-use.
+Using the Intel Architecture Code Analyzer tool~\cite{IACA} on the Sandy Bridge, we confirmed that this is a consequence of increased register re-use.
 In Helmholtz $p=4$, for example, the tool showed that when using \emph{op-vect} the number of clock cycles to execute one iteration of the $j$ loop decreases by roughly 17$\%$, and that this is a result of the relieved pressure on both of the data (cache) ports available in the core.
@@
 On the Sandy Bridge, we have also measured the performance of individual kernels in terms of floating-point operations per second.
-The theoretical peak, with the Intel Turbo Boost technology activated, is 30.4 GFlop/s.
-In the case of Diffusion using a prismatic mesh and $p=4$, we achived a maximum of 21.9 GFlop/s with \emph{op-vect} enabled, whereas 16.4 GFlop/s was obtained when only \emph{licm-ap} is used.
-This result is inline with the expectations: analysis of assembly code showed that, in the \texttt{jk} loop nest, which in this problem represents the bulk of the computation, 73$\%$ of instructions are actually floating-point operations.
+The theoretical peak on a single core, with the Intel Turbo Boost technology activated, is 30.4 GFlop/s.
+In the case of Diffusion using a prismatic mesh and $p=4$, we achieved a maximum of 21.9 GFlop/s with \emph{op-vect} enabled, whereas 16.4 GFlop/s was obtained when only \emph{licm-ap} is used.
+This result is in line with the expectations: analysis of assembly code showed that, in the \texttt{jk} loop nest, which in this problem represents the bulk of the computation, 73$\%$ of instructions are actually floating-point operations.
 
 Application of \emph{op-vect} to the Burgers problem induces significant slow downs due to the large number of temporary arrays that need to be tiled, which exceeds the available logical registers on the underlying architecture.
-Expression splitting can be used in combination to \emph{op-vect} to alleviate this issue; this is discussed in Section~\ref{sec:perf-results-split}.
+Expression splitting can be used in combination to \emph{op-vect} to alleviate this issue; this is discussed in the next section.
 
@@
 \includegraphics[scale=0.7]{Pictures/helmholtz-normalized-opvect.pdf}
-\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel on the Sandy Bridge. CM stands for cost model.}
+\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel on the Sandy Bridge.
+CM refers to where the cost model has been used to determine the sequence of code transformations.}
 \label{fig:opvect-helmholtz-speedup}
@@
 \includegraphics[scale=0.7]{Pictures/helmholtz-normalized-opvect-phi.pdf}
-\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel on the Xeon Phi. CM stands for cost model.}
+\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel on the Xeon Phi.
+CM refers to where the cost model has been used to determine the sequence of code transformations.}
 \label{fig:opvect-helmholtz-speedup-phi}
@@
 \includegraphics[scale=0.7]{Pictures/diffusion-normalized-opvect.pdf}
-\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel on the Sandy Bridge. CM stands for cost model.}
+\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel on the Sandy Bridge.
+CM refers to where the cost model has been used to determine the sequence of code transformations.}
 \label{fig:opvect-diffusion-speedup}
@@
 \includegraphics[scale=0.7]{Pictures/diffusion-normalized-opvect-phi.pdf}
-\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel on the Xeon Phi. CM stands for cost model.}
+\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel on the Xeon Phi.
+CM refers to where the cost model has been used to determine the sequence of code transformations.}
 \label{fig:opvect-diffusion-speedup-phi}
@@
 On the Sandy Bridge, in almost all test cases, a split factor of 1, meaning that the original expression was divided into two parts, ensured close-to-peak perforance.
-The transformation negligbly affected register locality.
-On the contrary, the larger number of vector units present on the Xeon Phi significantly relieve register spilling, making \emph{split} only marginally useful for loop unrolling.
+The transformation negligibly affected register locality.
+On the contrary, the larger number of vector units present on the Xeon Phi relieve register spilling, making \emph{split} only marginally useful for loop unrolling.
 
@@
 Despite improvements up to 1.22$\times$, \emph{split-op-vect} never outperforms \emph{split}.
-This is because of two reasons: for small split factors, like 1 and 2, the data space to be tiled is still too big, and register spilling affects run-time; for higher ones, sub-expressions become so small that, as explained in Section~\ref{sec:perf-eval-opvect}, full or partial unrolling of the loop nest allow achieving a certain degree of register re-use themselves.
+This is for two reasons: for small split factors, like 1 and 2, the data space to be tiled is still too big, and register spilling affects run-time; for higher ones, sub-expressions become so small that, as explained in Section~\ref{sec:perf-eval-opvect}, full or partial unrolling of the loop nest allow achieving a certain degree of register re-use themselves.
 
@@
 \includegraphics[scale=0.7]{Pictures/burgers-normalized-split.pdf}
-\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel on the Sandy Bridge. CM stands for cost model.}
+\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel on the Sandy Bridge.
+CM refers to where the cost model has been used to determine the sequence of code transformations.}
 \label{fig:split-burgers-speedup}
@@
 \includegraphics[scale=0.7]{Pictures/burgers-normalized-split-phi.pdf}
-\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel on the Xeon Phi. CM stands for cost model.}
+\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel on the Xeon Phi.
+CM refers to where the cost model has been used to determine the sequence of code transformations.}
 \label{fig:split-burgers-speedup-phi}
@@
 In FEniCS, FFC can apply its own model-driven optimizations to local assembly kernels~\cite{quadrature1}, which mainly consist of loop-invariant code motion and elimination, at code generation time, of floating point operations involving zero-valued entries in basis function arrays.
-The FEniCS' loop-invariant code motion is, however, totally different from COFFEE's, since it is based on expansion of arithmetic operations to identify terms that are invariant of the whole loop nest.
+The FEniCS Form Compiler's loop-invariant code motion is, however, different from COFFEE's, since it is based on expansion of arithmetic operations, for example applying distributivity and associativity to products and sums at code generation time, to identify terms that are invariant of the whole loop nest.
 In addition, depending on the way expansion is performed, operation count often does not decrease significantly.
-On the other hand, elimination of zero-valued terms, which are due to certain properties of the finite element problem, has the effect of introducing indirection arrays in the generated code.
-This kind of optimization is currently under development in COFFEE, although it will differ from FEniCS' by avoiding non-contiguous memory accesses, which would otherwise affect vectorization, at the price of removing less zero-valued contributions.
-
-Table~\ref{table:comparison-to-FFC-opt} summarizes the performance achived by COFFEE over the best FEniCS (FFC) implementation (i.e.
-the one that solves the problem fastest, regardless of using optimizations) on the Sandy Bridge for the Burgers, Helmholtz and Diffusion kernels.
+On the other hand, elimination of zero-valued terms, which are the result of certain properties of the finite element problem, has the effect of introducing indirection arrays in the generated code.
+This kind of optimization is currently under development in COFFEE, although it will differ from FEniCS' by avoiding non-contiguous memory accesses, which would otherwise affect vectorization, at the price of removing fewer zero-valued contributions.
+
+Table~\ref{table:comparison-to-FFC-opt} summarizes the performance achieved by COFFEE over the best FEniCS (FFC) implementation (i.e.
+the one that solves the problem fastest, regardless of the optimizations applied) on the Sandy Bridge for the Burgers, Helmholtz and Diffusion kernels.
 Burgers' slow downs occur in presence of a small iteration space (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$), although they are generally smaller than that reported in table, obtained with a triangular mesh and $p = 1$.
@@
 Our compiler-based optimization approach is made possible by the DSL on top, which enables automated code generation.
-DSLs have been proven successful in auto-generating optimized code for other domains: Spiral~\cite{Pueschel:05} for digital signal processing numerical algorithms, ~\cite{Spampinato:14} for dense linear algebra, or Pochoir~\cite{pochoir} and SDSL~\cite{stencil-compiler} for image processing and scientific computations over structured grids.
+DSLs have been proven successful in auto-generating optimized code for other domains: Spiral~\cite{Pueschel:05} for digital signal processing numerical algorithms, ~\cite{Spampinato:14} for dense linear algebra, or Pochoir~\cite{pochoir} and SDSL~\cite{stencil-compiler} for image processing and finite difference stencils.
 Similarly, PyOP2 is used by Firedrake to express iteration over unstructured meshes in scientific codes.
@@
 In this paper, we have presented design, optimizations and systematic performance evaluation of COFFEE, a compiler for finite element local assembly.
-In this context, to the best of our knowledge, COFFEE is the first compiler oriented towards the introduction of low-level optimizations to maximize instruction-level parallelism, register locality and SIMD vectorization.
+In this context, to the best of our knowledge, COFFEE is the first compiler capable of introducing low-level optimizations to maximize instruction-level parallelism, register locality and SIMD vectorization.
 Assembly kernels have peculiar characteristics.
@@
 The various optimizations overcome limitations of current vendor and research compilers.
-The exploitation of domain-knowledge allows some of them to be particularly effective, as demonstrated by our experiments on two state-of-the-art Intel platforms.
+The exploitation of domain knowledge allows some of them to be particularly effective, as demonstrated by our experiments on two state-of-the-art Intel platforms.
 Further work include a comprehensive study about feasibility and constraints on transforming kernels into a sequence of calls to external linear algebra libraries.
@@
 
-This research is partly funded by the MAPDES project and by the Department of Computing at Imperial College London.
+This research is partly funded by the MAPDES project, by the Department of Computing at Imperial College London, and by EPSRC through grants EP/I00677X/1, EP/I006761/1, and EP/L000407/1.
 The authors would like to thank Dr.

### 2014-04-14 paper/bare_conf.tex
Some copy-editing

@@
 %
-%\usepackage{ifpdf}
-% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
+%\usepackage{ifpdf}% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
 % compilation based on whether the output is pdf or dvi.
@@
 J. Ramanujam\IEEEauthorrefmark{3},
-David A. Ham\IEEEauthorrefmark{4},
+David A. Ham\IEEEauthorrefmark{4}\IEEEauthorrefmark{1},
 and
@@
 \IEEEauthorblockA{\IEEEauthorrefmark{1}Imperial College London, Department of Computing,\\
-\IEEEauthorrefmark{4}Imperial College London, Department of Computing and Department of Mathematics,\\
-\IEEEauthorrefmark{2}TU Delft, Faculty of Engineering, Mathematics and Computer Science (EWI)\\
+\IEEEauthorrefmark{4}Imperial College London, Department of Mathematics,\\
+\IEEEauthorrefmark{2}University of Amsterdam, Department of Computer Sciences,\\
+\IEEEauthorrefmark{3}Louisiana State University, Department of Electrical and Computer Engineering,\\
 \IEEEauthorrefmark{1}\{f.luporini12$|$f.rathgeber10$|$gheorghe-teodor.bercea08$|$p.kelly$|$david.ham\}@imperial.ac.uk,\\
@@
 %\boldmath
-Numerical solution of partial differential equations using the finite element method is one of the key applications of high performance computing.
-The local assembly is its characteristic operation, which executes a problem-specific kernel for each element in the discretized problem domain.
+The numerical solution of partial differential equations using the finite
+element method is one of the key applications of high performance
+computing. Local assembly is its characteristic operation. This entails the
+execution of a problem-specific kernel for each element in the discretized problem domain.
 Since the domain size can be huge, executing efficient kernels is fundamental.
@@
 \section{Introduction}
-In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by means of partial differential equations (PDEs).
-Numerical techniques, like finite volume method and finite element method, are widely-employed to approximate solutions of PDEs.
-Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective.
+In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by partial differential equations (PDEs).
+Numerical techniques, such as the finite volume method and finite element method, are widely-employed to approximate solutions of these PDEs.
+Unstructured meshes are often used to
+discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective.
 The solution is sought in each element of the discretized domain by applying suitable numerical operations.
-As the number of elements can be of the order of millions, a major issue is the time required to execute the computation, which can be hours or days.
+As the number of elements can be of the order of trillions \cite{Rossinelli2013}, a major issue is the time required to execute the computation, which can be hours or days.
 To address this problem, domain-specific languages (DSLs) have been developed.
-The successful porting of Hydra, an industrial computational fluid dynamics application developed by Rolls Royce for turbomachinery design (based on finite volume method, roughly 50,000 lines of code and with meshes in excess of 100M edges), to OP2~\cite{op2-main}, demonstrates the effectiveness of the DSL approach for implementing PDEs solvers~\cite{IstvanHydra}.
+The successful port of Hydra, an industrial computational fluid dynamics application developed by Rolls Royce for turbomachinery design (based on finite volume method, roughly 50,000 lines of code and with meshes in excess of 100M edges), to OP2~\cite{op2-main}, demonstrates the effectiveness of the DSL approach in the implementation of PDE solvers~\cite{IstvanHydra}.
 
@@
 During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain.
-A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution.
+A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed.
 However, we do not consider the latter case.
-An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, where individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for lower order methods.
+An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, in which individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for lower order methods.
 In the innermost loop, a problem-specific expression evaluates a two dimensional array, representing the result of local assembly in an element of the discretized domain.
@@
 BLAS routines could be theoretically employed, although a fairly complicated control- and data-flow analysis is required to automate identification and extraction of matrix-matrix multiplies.
-In addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small dimension of the involved matrices and the potential loss in data locality can limit the performance gain, if any.
-
-Due to the constraints of available compilers and linear-algebra specialized libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for optimizing local assembly kernels.
+In addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small dimension of the  matrices involved and the potential loss in data locality can limit or eliminate the performance gain of this approach.
+
+In order to overcome the constraints of the available compilers and specialized linear-algebra libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for optimizing local assembly kernels.
 COFFEE is integrated with Firedrake~\cite{firedrake-code}, a system for solving PDEs through the finite element method based on the PyOP2 abstraction~\cite{pyop2isc}.
-It supports any problems expressible with this framework.
+It supports all problems expressible with this framework.
 This allows us to evaluate our code transformations in a range of real-world problems, varying two key parameters that impact both solution accuracy and kernel cost: the polynomial order of the method (from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).
@@
 Our experiments showed that the originally generated code for non-trivial assembly kernels was sub-optimal.
-Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over code where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment, and up to 4.4$\times$ over original kernels.
+Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over code in which only basic transformations have been performed, such as generalized loop-invariant code motion, padding, and data alignment, and up to 4.4$\times$ over original kernels.
 The contributions of this paper are
 \begin{itemize}
-\item An optimisation strategy for finite element local assembly. Our approach exploits domain knowledge and goes beyond the limits of both vendor and research compilers.
+\item An optimization strategy for finite element local assembly. Our approach exploits domain knowledge and goes beyond the limits of both vendor and research compilers.
 \item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
@@
 The paper is organized as follows.
-In Section~\ref{sec:background} we provide some background on local assembly, showing code generated by Firedrake and emphasizing critical computational aspects.
+In Section~\ref{sec:background} we provide some background on local assembly, showing code generated by Firedrake and emphasizing the critical computational aspects.
 Section~\ref{sec:code-transf} describes the various code transformations, highlighting when and how domain-knowledge has been exploited.
@@
 Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm.
-It then triggers compilation of such kernels using a vendor compiler, and eventually manages parallel execution over the mesh.
+It then triggers compilation of such kernels using a vendor compiler, and manages parallel execution over the mesh.
 The subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.
@@
 The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}.
-Input is a zero-initialized two dimensional array used to store the element matrix, the element's coordinates in the discretized domain, and coefficient fields, for instance indicating value of velocity or pressure in the element.
+The inputs are a zero-initialized two dimensional array used to store the element matrix, the element's coordinates in the discretized domain, and coefficient fields, for instance indicating value of velocity or pressure in the element.
 The output is the evaluated element matrix.
-A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant two dimensional arrays, called ``basis functions'' (and their derivatives) in finite element terminology, initialized to problem-specific double-precision floating point values; 3) evaluation of the element matrix in an affine loop nest.
+A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of a problem-specific set of constant two dimensional arrays of double precision floats containing the values of the basis functions and their derivatives at the integration points; 3) evaluation of the element matrix in an affine loop nest.
 All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (i.e.
 using \texttt{static const} in C).
-Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel's objects.
+Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel objects.
 
@@
 Fields & double** & w \\ \hline
-Integration points & double[] & W \\ \hline
+Numerical integration weights & double[] & W \\ \hline
 Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ \hline
@@
 In other scenarios, for instance when solving a non-linear problem like the Burgers equation, whose assembly code is given in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
-Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse.
+This may also require the evaluation of constants in outer loops (called $F$ in the code) to act as as scaling factors of arrays; trip counts can be larger (proportionally to the order of the method); and arrays may be block-sparse.
 In addition to a larger number of operations, more complex cases like the Burgers equation are characterized by high register pressure.
 
-Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.
+Despite the infinite variety of problem-specific assembly kernels which Firedrake can generate, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.
 These include: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas interchanges involving the $i$ loop require pre-computation of values (e.g.
@@
 The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake.
-One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place.
+One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, in which the numerical integration takes place.
 In rare cases, as in problems involving hyperelasticity, however, such optimizations might not be sufficient to achieve notable performance improvements.
 Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions.
-It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time.
+It is worth noting there are circumstances in which the number of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so large (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time.
 In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure).
@@
 \label{sec:code-transf-licm}
-From the inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only.
+From the inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions which only depend on two iteration variables.
 Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion.
 Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels.
-We want to overcome two deficiencies that both $intel$ and $gnu$ compilers have.
-First, they identify sub-expressions that are invariant with respect to the innermost loop only.
+We want to overcome two deficiencies that both $intel$ and $gnu$ compilers exhibit.
+First, they only identify sub-expressions that are invariant with respect to the innermost loop.
 This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted in the loop order \texttt{ijk}.
@@
 This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest.
-At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, as part of an inner loop.
+At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized as part of an inner loop.
 Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant.
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake when padding, data alignment, and \emph{licm} are applied to the Helmholtz problem given in Figure~\ref{code:helmholtz} for an AVX architecture.
+\caption{Local assembly code generated by Firedrake in which padding, data alignment, and \emph{licm} have been applied to the Helmholtz problem given in Figure~\ref{code:helmholtz} for an AVX architecture.
 In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$, so they can be precomputed once in the $r$ loop.}
@@
 One notable problem of assembly kernels concerns register allocation and register locality.
-The critical situation occurs when loop trip counts and the variables accessed are such that the vector-registers pressure is high.
+The critical situation occurs when loop trip counts and the variables accessed are such that the vector-register pressure is high.
 Since the kernel's working set fits the L1 cache, it is remarkably important to optimize register management.
@@
 Tiling at the level of vector registers is an additional feature of COFFEE.
-Based on the observation that the evaluation of the element matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented.
+Based on the observation that the evaluation of the element matrix can be reduced to a summation of outer products along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented.
 If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, the computation of the element matrix is abstractly expressible as
@@
 \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets.
-Regardless of the specific input problem, and by abstracting from the presence of all variables independent of both $j$ and $k$, the element matrix computation is always reducible to this form.
+Regardless of the specific input problem, by abstracting from the presence of all variables independent of both $j$ and $k$, the element matrix computation is always reducible to this form.
 Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries ($j,k=0,...,4$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$.
@@
 The figure shows how 4$\times$4 elements in the top-left block of the element matrix $A$ can be moved to their correct positions.
-Each rotation, represented by a group of three same-colored arrows, is implemented by a single shuffle intrinsics.}
+Each rotation, represented by a group of three same-colored arrows, is implemented by a single shuffle intrinsic.}
 \label{fig:restore-layout}
@@
 ~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
-~~~~~~~~// One $permute\_pd$ intrinsics per $k$-loop $load$\\
+~~~~~~~~// One $permute\_pd$ intrinsic per $k$-loop $load$\\
 ~~~~~~~~// Compute A[1,2],A[2,1],A[3,4],A[4,3] \\
-~~~~~~~~// One $permute2f128\_pd$ intrinsics per $k$-loop $load$\\
+~~~~~~~~// One $permute2f128\_pd$ intrinsic per $k$-loop $load$\\
 ~~~~~~~~// ...\\
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake for the Helmholtz problem when \emph{op-vect} is applied on top of the optimizations shown in Figure~\ref{code:helmholtz-licm}.
+\caption{Local assembly code generated by Firedrake for the Helmholtz problem in which \emph{op-vect} has been applied on top of the optimizations shown in Figure~\ref{code:helmholtz-licm}.
 Here, we assume the polynomial order is $p=2$, since \emph{op-vect} can not be used when an iteration space dimension is smaller than the vector length.
@@
 This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large.
-For example, applying loop-invariant code motion to Burgers on a 3D mesh needs 33 temporaries for the \texttt{ijk} loop order, so compiler's hoisting of invariant loads out of the $k$ loop can be inefficient on architectures with a relatively low number of registers.
+For example, applying loop-invariant code motion to Burgers on a 3D mesh requires 33 temporaries for the \texttt{ijk} loop order.
+This can make hoisting of the invariant loads out of the $k$ loop inefficient on architectures with a relatively low number of registers.
 One potential solution to this problem consists of suitably ``splitting'' the computation of the element matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}.
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake for the Helmholtz problem when \emph{split} is applied on top of the optimizations shown in Figure~\ref{code:helmholtz-licm}.
+\caption{Local assembly code generated by Firedrake for the Helmholtz problem in which \emph{split} has been applied on top of the optimizations shown in Figure~\ref{code:helmholtz-licm}.
 In this example, the split factor is 2.}
@@
 
-Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations.
-At run-time, the high-level specification is translated by the FEniCS Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
+Firedrake users employ the Unified Form Language to write problems in a notation resembling mathematical equations.
+At run-time, the high-level specification is translated by a modified version of the FEniCS Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
 ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}.
@@
 
-Loop unroll and unroll-and-jam of outer loops are fundamental to expose instruction-level parallelism, so tuning critical parameters, like the unroll factor, is of great importance.
-However, from our experience of inspecting assembly code and comparing with other hand-written implementations, recent versions of a vendor compiler like Intel's employ cost models capable of estimating close-to-optimal values for such parameters.
-This is particularly true for assembly kernels, where the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are stride-1.
+% "vendor compilers such as Intel's" is awkward and unclear. If Intel is the only one you have looked at, then say "the Intel compiler", otherwise name the ones for which you know this to be true.
+Since loop unroll and unroll-and-jam of outer loops are fundamental to the exposure of instruction-level parallelism, tuning critical parameters such as the unroll factor is of great importance.
+However, inspecting assembly code and comparing with other hand-written implementations indicates that recent versions of vendor compilers such as Intel's are capable of estimating close-to-optimal values for such parameters.
+This is particularly true for assembly kernels, in which the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are stride one.
 We therefore leave the backend compiler in charge of selecting the unroll factor.
@@
 On the Sandy Bridge, the compilation flags used were \texttt{-O2} and \texttt{-xAVX} for auto-vectorization.
-On the Xeon Phi, the optimization level \texttt{-O3} was used.
-Other optimization levels perfomed, in general, slightly worse.
+On the Xeon Phi, optimization level \texttt{-O3} was used.
+Other optimization levels performed, in general, slightly worse.
 Our code transformations were evaluated in three real-world problems based on the following PDEs:
@@
 For instance, in the case of Helmholtz with $p=1$, the size of the $j$ and $k$ loops for the three element shapes is, respectively, $3$, $4$, and $6$.
-Moving to bigger shapes has the effect of increasing the number of basis function arrays, since, intuitively, the behaviour of the equation has now to be approximated also along a third axis.
+Moving to bigger shapes has the effect of increasing the number of basis function arrays, since, intuitively, the behavior of the equation has now to be approximated also along a third axis.
 On the other hand, the polynomial order affects only the problem size (the three loops $i$, $j$, and $k$, and, as a consequence, the size of $X$ and $Y$ arrays).
-A range of polynomial orders, from $p=1$ to $p=4$, were tested; higher polynomial orders are excluded from the study because of current Firedrake limitations.
+A range of polynomial orders from $p=1$ to $p=4$ were tested; higher polynomial orders are excluded from the study because of current Firedrake limitations.
 In all these cases, the size of the element matrix rarely exceeds 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
@@
 Application of \emph{op-vect} to the Burgers problem induces significant slow downs due to the large number of temporary arrays that need to be tiled, which exceeds the available logical registers on the underlying architecture.
-Expression splitting can be used in combination to \emph{op-vect} to alleviate this issue; this is discussed in the next section.
+Expression splitting can be used in combination with \emph{op-vect} to alleviate this issue; this is discussed in the next section.
 
@@
 Despite improvements up to 1.22$\times$, \emph{split-op-vect} never outperforms \emph{split}.
-This is for two reasons: for small split factors, like 1 and 2, the data space to be tiled is still too big, and register spilling affects run-time; for higher ones, sub-expressions become so small that, as explained in Section~\ref{sec:perf-eval-opvect}, full or partial unrolling of the loop nest allow achieving a certain degree of register re-use themselves.
+This is for two reasons: for small split factors, such as 1 and 2, the data space to be tiled is still too big, and register spilling affects run-time; for higher ones, sub-expressions become so small that, as explained in Section~\ref{sec:perf-eval-opvect}, full or partial unrolling of the loop nest achieves a certain degree of register re-use by itself.
 
@@
 \caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel on the Sandy Bridge.
-CM refers to where the cost model has been used to determine the sequence of code transformations.}
+CM refers to the cases in which the cost model has been used to determine the sequence of code transformations.}
 \label{fig:split-burgers-speedup}
@@
 \caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel on the Xeon Phi.
-CM refers to where the cost model has been used to determine the sequence of code transformations.}
+CM refers to the cases in which the cost model has been used to determine the sequence of code transformations.}
 \label{fig:split-burgers-speedup-phi}
@@
 \label{sec:perf-results-fenics}
-We have modified the FEniCS Form Compiler (FFC) to make it return an abstract syntax tree representation of a local assembly kernel, rather than plain C code, so as to enable code transformations in COFFEE.
+We have modified the FEniCS Form Compiler (FFC) to make it return an abstract syntax tree representation of a local assembly kernel, rather than plain C++ code, so as to enable code transformations in COFFEE.
 Besides Firedrake, FFC is used in the FEniCS project~\cite{Fenics}.
@@
 Table~\ref{table:comparison-to-FFC-opt} summarizes the performance achieved by COFFEE over the best FEniCS (FFC) implementation (i.e.
-the one that solves the problem fastest, regardless of the optimizations applied) on the Sandy Bridge for the Burgers, Helmholtz and Diffusion kernels.
-Burgers' slow downs occur in presence of a small iteration space (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$), although they are generally smaller than that reported in table, obtained with a triangular mesh and $p = 1$.
+the combination of optimizations for each problem which results in the fastest code) on the Sandy Bridge for the Burgers, Helmholtz and Diffusion kernels.
+Burgers' slow downs occur in presence of a small iteration space (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$).
+The figure shown is for the worst slow down, which was obtained with a triangular mesh and $p = 1$.
 This is a result of removing zero-valued entries in FEniCS' basis functions: some operations are avoided, but indirection arrays prevent auto-vectorization, which significantly impacts performance as soon as the element matrix becomes bigger than 12$\times$12.
@@
 
-Our compiler-based optimization approach is made possible by the DSL on top, which enables automated code generation.
+Our compiler-based optimization approach is made possible by the top-level DSL, which enables automated code generation.
 DSLs have been proven successful in auto-generating optimized code for other domains: Spiral~\cite{Pueschel:05} for digital signal processing numerical algorithms, ~\cite{Spampinato:14} for dense linear algebra, or Pochoir~\cite{pochoir} and SDSL~\cite{stencil-compiler} for image processing and finite difference stencils.
@@
 The data space, in terms of number of arrays and scalars required to evaluate the element matrix, grows proportionally with the complexity of the finite element problem.
-COFFEE has been developed taking into account all of these degrees of freedom, based on the idea that reducing the problem of local assembly optimization to a fixed sequence of transformations is way too superficial if close-to-peak performance needs to be reached.
+COFFEE has been developed taking into account all of these degrees of freedom, based on the idea that reducing the problem of local assembly optimization to a fixed sequence of transformations is far too superficial if close-to-peak performance needs to be reached.
 The various optimizations overcome limitations of current vendor and research compilers.
@@
 
-This research is partly funded by the MAPDES project, by the Department of Computing at Imperial College London, and by EPSRC through grants EP/I00677X/1, EP/I006761/1, and EP/L000407/1.
+This research is partly funded by the MAPDES project, by the Department of Computing at Imperial College London, by EPSRC through grants EP/I00677X/1, EP/I006761/1, and EP/L000407/1, and by NERC grants NE/K008951/1 and NE/K006789/1.
 The authors would like to thank Dr.

### 2014-04-14 paper/bare_conf.tex
Update plots and table

--- 2014-04-14 paper/bare_conf.tex
@@
 \begin{algorithm}[t]
-\small
+\footnotesize
 \KwSty{void} helmholtz(double A[3][3], double **coords) $\lbrace$\\
@@
 \begin{algorithm}[t]
-\small
+\footnotesize
 \KwSty{void} burgers(double A[12][12], double **c, double **w) $\lbrace$\\
@@
 \begin{algorithm}[t]
-\small
+\footnotesize
 \KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
@@
 \begin{algorithm}[t]
-\small
+\footnotesize
 \KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
@@
 \begin{algorithm}[t]
-\small
+\footnotesize
 \KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
-%~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
 ~~~~// Same code as in Figure~\ref{code:helmholtz-licm} up to line 15 \\
-%~~// Declaration of basis function matrices \\
-%~~\\
-%~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
-%~~~~double LI$\_$0[4] ALIGN;\\
-%~~~~double LI$\_$1[4] ALIGN;\\
-%~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
-%~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
-%~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
-%~~~~$\rbrace$\\
 ~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
@@
 \begin{algorithm}[t]
+\footnotesize
   \textbf{The COFFEE Compiler}\\
@@
 \If{\KwSty{not} it\_space}{
-  ast.apply\_inter\_kernel\_vectorization(wrapper, isa)
+  ast.apply\_inter\_kernel\_vectorization(wrapper, isa)\\
   \Return{wrapper + ast.from\_ast\_to\_c()}
@@
 \subsection{Impact of Generalized Loop-invariant Code Motion}
+%\newcommand{\massmatrixtwodresultsnorms}{
+%\begin{tabularx}{1.0\textwidth}{p{3.5cm}p{3.5cm}|p{2.25cm}p{2.25cm}p{2.25cm}p{2.25cm}|}
+%\cline{3-6}
+%& & \multicolumn{2}{c|}{\texttt{Sandy Bridge}} & \multicolumn{2}{c|}{\texttt{Xeon Phi}}\\
+%\cline{3-6}
+%& & \multicolumn{1}{c}{\texttt{licm}} & \multicolumn{1}{c|}{\texttt{licm-ap}} & \multicolumn{1}{c}{\texttt{licm}} & \multicolumn{1}{c|}{\texttt{licm-ap}} \\
+%\cline{1-6}
+%%\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
+%\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.05$\times$-1.68$\times$ & \multicolumn{1}{c|}{~1.32$\times$-4.14$\times$} & ~~1.15$\times$-1.70$\times$ & ~~1.54$\times$-3.27$\times$ \\
+%\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.36$\times$-2.64$\times$ & \multicolumn{1}{c|}{~1.35$\times$-3.32$\times$} & ~~1.18$\times$-1.87$\times$ & ~~1.27$\times$-2.92$\times$ \\
+%\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~2.16$\times$-2.45$\times$ & \multicolumn{1}{c|}{~2.42$\times$-2.63$\times$} & ~~0.96$\times$-1.64$\times$ & ~~1.84$\times$-1.72$\times$ \\
+%\cline{1-6}
+%\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.09$\times$-1.64$\times$ & \multicolumn{1}{c|}{~1.37$\times$-4.28$\times$} & ~~1.13$\times$-1.22$\times$ & ~~0.92$\times$-2.79$\times$ \\
+%\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.30$\times$-3.11$\times$ & \multicolumn{1}{c|}{~1.41$\times$-3.82$\times$} & ~~1.08$\times$-2.11$\times$ & ~~1.10$\times$-5.53$\times$ \\
+%\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~1.81$\times$-2.70$\times$ & \multicolumn{1}{c|}{~2.55$\times$-3.13$\times$} & ~~1.04$\times$-2.74$\times$ & ~~1.73$\times$-4.14$\times$ \\
+%\cline{1-6}
+%\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.53$\times$-2.46$\times$ & \multicolumn{1}{c|}{~1.56$\times$-2.77$\times$} & ~~1.19$\times$-2.97$\times$ & ~~2.70$\times$-4.27$\times$ \\
+%\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.58$\times$-2.24$\times$ & \multicolumn{1}{c|}{~1.59$\times$-2.10$\times$} & ~~1.03$\times$-1.21$\times$ & ~~1.31$\times$-1.55$\times$ \\
+%\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~1.32$\times$-2.11$\times$ & \multicolumn{1}{c|}{~1.42$\times$-2.31$\times$} & ~~1.04$\times$-1.40$\times$ & ~~1.27$\times$-2.18$\times$ \\
+%\cline{1-6}
+%\end{tabularx}
+%}
+
 \newcommand{\massmatrixtwodresultsnorms}{
-\begin{tabularx}{1.0\textwidth}{p{3.5cm}p{3.5cm}|p{2.25cm}p{2.25cm}p{2.25cm}p{2.25cm}|}
-\cline{3-6}
-& & \multicolumn{2}{c|}{\texttt{Sandy Bridge}} & \multicolumn{2}{c|}{\texttt{Xeon Phi}}\\
-\cline{3-6}
-& & \multicolumn{1}{c}{\texttt{licm}} & \multicolumn{1}{c|}{\texttt{licm-ap}} & \multicolumn{1}{c}{\texttt{licm}} & \multicolumn{1}{c|}{\texttt{licm-ap}} \\
-\cline{1-6}
-%\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.05$\times$-1.68$\times$ & \multicolumn{1}{c|}{~1.32$\times$-4.14$\times$} & ~~1.15$\times$-1.70$\times$ & ~~1.54$\times$-3.27$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.36$\times$-2.64$\times$ & \multicolumn{1}{c|}{~1.35$\times$-3.32$\times$} & ~~1.18$\times$-1.87$\times$ & ~~1.27$\times$-2.92$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~2.16$\times$-2.45$\times$ & \multicolumn{1}{c|}{~2.42$\times$-2.63$\times$} & ~~0.96$\times$-1.64$\times$ & ~~1.84$\times$-1.72$\times$ \\
-\cline{1-6}
-\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.09$\times$-1.64$\times$ & \multicolumn{1}{c|}{~1.37$\times$-4.28$\times$} & ~~1.13$\times$-1.22$\times$ & ~~0.92$\times$-2.79$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.30$\times$-3.11$\times$ & \multicolumn{1}{c|}{~1.41$\times$-3.82$\times$} & ~~1.08$\times$-2.11$\times$ & ~~1.10$\times$-5.53$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~1.81$\times$-2.70$\times$ & \multicolumn{1}{c|}{~2.55$\times$-3.13$\times$} & ~~1.04$\times$-2.74$\times$ & ~~1.73$\times$-4.14$\times$ \\
-\cline{1-6}
-\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.53$\times$-2.46$\times$ & \multicolumn{1}{c|}{~1.56$\times$-2.77$\times$} & ~~1.19$\times$-2.97$\times$ & ~~2.70$\times$-4.27$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.58$\times$-2.24$\times$ & \multicolumn{1}{c|}{~1.59$\times$-2.10$\times$} & ~~1.03$\times$-1.21$\times$ & ~~1.31$\times$-1.55$\times$ \\
-\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~1.32$\times$-2.11$\times$ & \multicolumn{1}{c|}{~1.42$\times$-2.31$\times$} & ~~1.04$\times$-1.40$\times$ & ~~1.27$\times$-2.18$\times$ \\
-\cline{1-6}
+\begin{tabularx}{1.0\textwidth}{cccccc|cccc|cccc|cccc}
+\cline{3-18}
+& & \multicolumn{8}{c}{\texttt{Sandy Bridge}} & \multicolumn{8}{c}{\texttt{Xeon Phi}}\\
+\cline{3-18}
+& & \multicolumn{4}{c}{\texttt{licm}} & \multicolumn{4}{c}{\texttt{licm-ap}} & \multicolumn{4}{c}{\texttt{licm}} & \multicolumn{4}{c}{\texttt{licm-ap}} \\
+\cline{1-18}
+& & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} \\[0.1cm]
+\texttt{Helmholtz} & \texttt{triangle} & 1.05 & 1.46 & 1.68 & 1.67 & 1.32 & 1.88 & 2.87 & 4.13 & 1.05 & 1.46 & 1.68 & 1.67 & 1.32 & 1.88 & 2.87 & 4.13 \\
+\texttt{Helmholtz} & \texttt{tetrahedron} & 1.36 & 2.10 & 2.64 & 2.27 & 1.35 & 3.32 & 2.66 & 3.27 & 1.36 & 2.10 & 2.64 & 2.27 & 1.35 & 3.32 & 2.66 & 3.27 \\
+\texttt{Helmholtz} & \texttt{prism} & 2.16 & 2.28 & 2.45 & 2.06 & 2.63 & 2.74 & 2.43 & 2.75 & 2.16 & 2.28 & 2.45 & 2.06 & 2.63 & 2.74 & 2.43 & 2.75\\[0.1cm]
+%\cline{1-18}
+\texttt{Diffusion} & \texttt{triangle} & 1.09 & 1.68 & 1.97 & 1.64 & 1.38 & 1.99 & 3.07 & 4.28 & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69\\
+\texttt{Diffusion} & \texttt{tetrahedron} & 1.30 & 2.20 & 3.12 & 2.60 & 1.41 & 3.70 & 3.18 & 3.82 & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69\\
+\texttt{Diffusion} & \texttt{prism} & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69 & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69\\[0.1cm]
+%\cline{1-18}
+\texttt{Burgers} & \texttt{triangle} & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 \\
+\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.24 & 1.69 & 1.59 & 1.61 & 2.10 & 1.60 & 1.78 & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 \\
+\texttt{Burgers} & \texttt{prism} & 2.11 & 2.20 & 1.66 & 1.32 & 2.19 & 2.32 & 1.64 & 1.42 & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 \\
+\cline{1-18}
 \end{tabularx}
@@
 
-\begin{table*}[t]\small
+\begin{table*}[t]\footnotesize
 \massmatrixtwodresultsnorms
 ~
-\caption{Performance improvement due to generalized loop-invariant code motion (\emph{licm} column) on the Helmholtz, Diffusion and Burgers problems, varying the element shape (triangle, tetrahedron, prism) and the polynomial order ($p \in [1, 4]$), using the Lagrange element family, over the original, non-optimized code.
-Each entry indicates the minimum and maximum speedup obtained in the range of $p$ values.
-The column \emph{licm-ap} illustrates the combination of \emph{licm} with data alignment and padding.
+\caption{Performance improvement due to generalized loop-invariant code motion (\emph{licm} columns) on the Helmholtz, Diffusion and Burgers problems, varying the element shape (triangle, tetrahedron, prism) and the polynomial order ($p \in [1, 4]$), using the Lagrange element family, over the original, non-optimized code.
+The \emph{licm-ap} columns illustrate the effect of combining \emph{licm} with data alignment and padding.
 Results are shown for both the Sandy Bridge and the Xeon Phi architectures.}

### 2014-04-14 paper/bare_conf.tex
Update licm table
Clean up

@@
 \footnotesize
-\KwSty{void} burgers(double A[12][12], double **c, double **w) $\lbrace$\\
-~~// K, det = Compute Jacobian (c) \\
+\KwSty{void} burgers(double A[12][12], double **coords, double **w) $\lbrace$\\
+~~// K, det = Compute Jacobian (coords) \\
 ~~\\
@@
 
-Experiments were run on two Intel architectures, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and a Xeon Phi (5110P, running at 1.05Ghz in native mode, 32KB L1 cache and 512KB L2 cache).
+Experiments were run on a single core of two Intel architectures, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and a Xeon Phi (5110P, running at 1.05Ghz in native mode, 32KB L1 cache and 512KB L2 cache).
 We have chosen these two architectures because of the differences in the number of logical registers and SIMD lanes, which can impact the effectiveness of the optimization strategy.
-All experiments are sequential.
 The \texttt{icc 13.1} compiler was used.
@@
 \subsection{Impact of Generalized Loop-invariant Code Motion}
-%\newcommand{\massmatrixtwodresultsnorms}{
-%\begin{tabularx}{1.0\textwidth}{p{3.5cm}p{3.5cm}|p{2.25cm}p{2.25cm}p{2.25cm}p{2.25cm}|}
-%\cline{3-6}
-%& & \multicolumn{2}{c|}{\texttt{Sandy Bridge}} & \multicolumn{2}{c|}{\texttt{Xeon Phi}}\\
-%\cline{3-6}
-%& & \multicolumn{1}{c}{\texttt{licm}} & \multicolumn{1}{c|}{\texttt{licm-ap}} & \multicolumn{1}{c}{\texttt{licm}} & \multicolumn{1}{c|}{\texttt{licm-ap}} \\
-%\cline{1-6}
-%%\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
-%\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.05$\times$-1.68$\times$ & \multicolumn{1}{c|}{~1.32$\times$-4.14$\times$} & ~~1.15$\times$-1.70$\times$ & ~~1.54$\times$-3.27$\times$ \\
-%\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.36$\times$-2.64$\times$ & \multicolumn{1}{c|}{~1.35$\times$-3.32$\times$} & ~~1.18$\times$-1.87$\times$ & ~~1.27$\times$-2.92$\times$ \\
-%\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~2.16$\times$-2.45$\times$ & \multicolumn{1}{c|}{~2.42$\times$-2.63$\times$} & ~~0.96$\times$-1.64$\times$ & ~~1.84$\times$-1.72$\times$ \\
-%\cline{1-6}
-%\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.09$\times$-1.64$\times$ & \multicolumn{1}{c|}{~1.37$\times$-4.28$\times$} & ~~1.13$\times$-1.22$\times$ & ~~0.92$\times$-2.79$\times$ \\
-%\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.30$\times$-3.11$\times$ & \multicolumn{1}{c|}{~1.41$\times$-3.82$\times$} & ~~1.08$\times$-2.11$\times$ & ~~1.10$\times$-5.53$\times$ \\
-%\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~1.81$\times$-2.70$\times$ & \multicolumn{1}{c|}{~2.55$\times$-3.13$\times$} & ~~1.04$\times$-2.74$\times$ & ~~1.73$\times$-4.14$\times$ \\
-%\cline{1-6}
-%\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.53$\times$-2.46$\times$ & \multicolumn{1}{c|}{~1.56$\times$-2.77$\times$} & ~~1.19$\times$-2.97$\times$ & ~~2.70$\times$-4.27$\times$ \\
-%\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.58$\times$-2.24$\times$ & \multicolumn{1}{c|}{~1.59$\times$-2.10$\times$} & ~~1.03$\times$-1.21$\times$ & ~~1.31$\times$-1.55$\times$ \\
-%\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~1.32$\times$-2.11$\times$ & \multicolumn{1}{c|}{~1.42$\times$-2.31$\times$} & ~~1.04$\times$-1.40$\times$ & ~~1.27$\times$-2.18$\times$ \\
-%\cline{1-6}
-%\end{tabularx}
-%}
 
@@
 & & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} \\[0.1cm]
-\texttt{Helmholtz} & \texttt{triangle} & 1.05 & 1.46 & 1.68 & 1.67 & 1.32 & 1.88 & 2.87 & 4.13 & 1.05 & 1.46 & 1.68 & 1.67 & 1.32 & 1.88 & 2.87 & 4.13 \\
-\texttt{Helmholtz} & \texttt{tetrahedron} & 1.36 & 2.10 & 2.64 & 2.27 & 1.35 & 3.32 & 2.66 & 3.27 & 1.36 & 2.10 & 2.64 & 2.27 & 1.35 & 3.32 & 2.66 & 3.27 \\
-\texttt{Helmholtz} & \texttt{prism} & 2.16 & 2.28 & 2.45 & 2.06 & 2.63 & 2.74 & 2.43 & 2.75 & 2.16 & 2.28 & 2.45 & 2.06 & 2.63 & 2.74 & 2.43 & 2.75\\[0.1cm]
+\texttt{Helmholtz} & \texttt{triangle} & 1.05 & 1.46 & 1.68 & 1.67 & 1.32 & 1.88 & 2.87 & 4.13 & 1,49 & 1,06 & 1,05 & 1,17 & 1,50 & 2,41 & 1,30 & 1,96 \\
+\texttt{Helmholtz} & \texttt{tetrahedron} & 1.36 & 2.10 & 2.64 & 2.27 & 1.35 & 3.32 & 2.66 & 3.27 & 1,28 & 1,29 & 2,05 & 1,73 & 1,41 & 1,50 & 2,79 & 2,81 \\
+\texttt{Helmholtz} & \texttt{prism} & 2.16 & 2.28 & 2.45 & 2.06 & 2.63 & 2.74 & 2.43 & 2.75 & 1,04 & 2,26 & 1,93 & 1,64 & 2,38 & 2,47 & 2,15 & 1,71\\[0.1cm]
 %\cline{1-18}
-\texttt{Diffusion} & \texttt{triangle} & 1.09 & 1.68 & 1.97 & 1.64 & 1.38 & 1.99 & 3.07 & 4.28 & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69\\
-\texttt{Diffusion} & \texttt{tetrahedron} & 1.30 & 2.20 & 3.12 & 2.60 & 1.41 & 3.70 & 3.18 & 3.82 & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69\\
-\texttt{Diffusion} & \texttt{prism} & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69 & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69\\[0.1cm]
+\texttt{Diffusion} & \texttt{triangle} & 1.09 & 1.68 & 1.97 & 1.64 & 1.38 & 1.99 & 3.07 & 4.28 & 1,07 & 1,06 & 1,18 & 1,16 & 1,08 & 1,88 & 1,20 & 1,97\\
+\texttt{Diffusion} & \texttt{tetrahedron} & 1.30 & 2.20 & 3.12 & 2.60 & 1.41 & 3.70 & 3.18 & 3.82 & 1,00 & 1,38 & 2,02 & 1,74 & 1,05 & 1,51 & 2,76 & 3,00\\
+\texttt{Diffusion} & \texttt{prism} & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69 & 1,11 & 2,16 & 1,85 & 2,83 & 2,41 & 2,52 & 2,05 & 2,48\\[0.1cm]
 %\cline{1-18}
-\texttt{Burgers} & \texttt{triangle} & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 \\
-\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.24 & 1.69 & 1.59 & 1.61 & 2.10 & 1.60 & 1.78 & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 \\
-\texttt{Burgers} & \texttt{prism} & 2.11 & 2.20 & 1.66 & 1.32 & 2.19 & 2.32 & 1.64 & 1.42 & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 \\
+\texttt{Burgers} & \texttt{triangle} & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 & 1,21 & 1,42 & 2,34 & 2,97 & 2,84 & 2,26 & 3,96 & 4,27 \\
+\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.24 & 1.69 & 1.59 & 1.61 & 2.10 & 1.60 & 1.78 & 1,01 & 2,55 & 0,98 & 1,21 & 1,48 & 3,83 & 1,55 & 1,29 \\
+\texttt{Burgers} & \texttt{prism} & 2.11 & 2.20 & 1.66 & 1.32 & 2.19 & 2.32 & 1.64 & 1.42 & 1,39 & 1,56 & 1,18 & 1,04 & 2,18 & 2,82 & 1,24 & 1,25 \\
 \cline{1-18}
@@
 Occasionally \emph{licm-ap} is slower than \emph{licm} (e.g.
-in Burgers 3D p3 on the Sandy Bridge).
+in Burgers $p=3$ on the Sandy Bridge).
 One possible explanation is that the large number of aligned temporaries introduced by \emph{licm} induce cache associativity conflicts.
@@
 The recent introduction of DSLs to decouple the finite element specification from its underlying implementation facilitated, however, the development of novel approaches.
-Methods based on tensor contraction~\cite{FFC-TC} and symbolic manipulation~\cite{Francis} have been developed.
+Methods based on tensor contraction~\cite{FFC-Compiler} and symbolic manipulation~\cite{Francis} have been developed.
 We have designed COFFEE to specifically target numerical integration because it has been demonstrated that it remains the optimal choice for a wide class of problems~\cite{quadrature1}.

### 2014-04-15 paper/bare_conf.tex
Minor fixes + shrink vect pictures

@@
 Numerical techniques, such as the finite volume method and finite element method, are widely-employed to approximate solutions of these PDEs.
-Unstructured meshes are often used to
-discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective.
+Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective.
 The solution is sought in each element of the discretized domain by applying suitable numerical operations.
@@
 \begin{figure}[h]
-\centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect.pdf}}
+\centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect-inline.pdf}}
 \caption{Outer-product vectorization by permuting values in a vector register.}
@@
 \begin{figure}[t]
-\centerline{\includegraphics[scale=0.60]{Pictures/vect-restore.pdf}}
+\centerline{\includegraphics[scale=0.60]{Pictures/vect-restore-inline.pdf}}
 \caption{Restoring the storage layout after \emph{op-vect}.
@@
 Since loop unroll and unroll-and-jam of outer loops are fundamental to the exposure of instruction-level parallelism, tuning critical parameters such as the unroll factor is of great importance.
-However, inspecting assembly code and comparing with other hand-written implementations indicates that recent versions of vendor compilers such as Intel's are capable of estimating close-to-optimal values for such parameters.
+However, inspecting assembly code and comparing with other hand-written implementations indicates that recent versions of the Intel compiler are capable of estimating close-to-optimal values for such parameters.
 This is particularly true for assembly kernels, in which the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are stride one.
@@
 \begin{figure}[h]
-\includegraphics[scale=0.7]{Pictures/helmholtz-normalized-opvect.pdf}
+\includegraphics[scale=0.67]{Pictures/helmholtz-normalized-opvect.pdf}
 \caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel on the Sandy Bridge.
@@
 \begin{figure}[h]
-\includegraphics[scale=0.7]{Pictures/helmholtz-normalized-opvect-phi.pdf}
+\includegraphics[scale=0.67]{Pictures/helmholtz-normalized-opvect-phi.pdf}
 \caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel on the Xeon Phi.
@@
 \begin{figure}[h]
-\includegraphics[scale=0.7]{Pictures/diffusion-normalized-opvect.pdf}
+\includegraphics[scale=0.67]{Pictures/diffusion-normalized-opvect.pdf}
 \caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel on the Sandy Bridge.
@@
 \begin{figure}[h]
-\includegraphics[scale=0.7]{Pictures/diffusion-normalized-opvect-phi.pdf}
+\includegraphics[scale=0.67]{Pictures/diffusion-normalized-opvect-phi.pdf}
 \caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel on the Xeon Phi.
@@
 \begin{figure}[h]
-\includegraphics[scale=0.7]{Pictures/burgers-normalized-split.pdf}
+\includegraphics[scale=0.67]{Pictures/burgers-normalized-split.pdf}
 \caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel on the Sandy Bridge.
@@
 \begin{figure}[h]
-\includegraphics[scale=0.7]{Pictures/burgers-normalized-split-phi.pdf}
+\includegraphics[scale=0.67]{Pictures/burgers-normalized-split-phi.pdf}
 \caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel on the Xeon Phi.
@@
 The recent introduction of DSLs to decouple the finite element specification from its underlying implementation facilitated, however, the development of novel approaches.
-Methods based on tensor contraction~\cite{FFC-Compiler} and symbolic manipulation~\cite{Francis} have been developed.
+Methods based on tensor contraction~\cite{FFC-Compiler} and symbolic manipulation~\cite{Francis} have been implemented.
 We have designed COFFEE to specifically target numerical integration because it has been demonstrated that it remains the optimal choice for a wide class of problems~\cite{quadrature1}.

### 2014-04-16 paper/bare_conf.tex
Minor fixes. SC14-Submission ready.

@@
 In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by partial differential equations (PDEs).
-Numerical techniques, such as the finite volume method and finite element method, are widely-employed to approximate solutions of these PDEs.
+Numerical techniques, such as the finite volume method and finite element method, are widely employed to approximate solutions of these PDEs.
 Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective.
@@
 
-We focus on improving the performance of local assembly (``assembly'', in the following), a fundamental step of the finite element method that covers an important fraction of the overall computation run-time, often in the range 30$\%$-60$\%$.
+We focus on improving the performance of local assembly (``assembly'', in the following), a fundamental step in the finite element method that covers an important fraction of the overall computation run-time, often in the range 30-60$\%$.
 During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain.
@@
 However, we do not consider the latter case.
-An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, in which individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for lower order methods.
+An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, in which individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for low order methods.
 In the innermost loop, a problem-specific expression evaluates a two dimensional array, representing the result of local assembly in an element of the discretized domain.
@@
 Optimizations like generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but none are supported by state-of-the-art polyhedral and vendor compilers.
-BLAS routines could be theoretically employed, although a fairly complicated control- and data-flow analysis is required to automate identification and extraction of matrix-matrix multiplies.
+BLAS routines could theoretically be employed, although a fairly complicated control- and data-flow analysis is required to automate identification and extraction of matrix-matrix multiplies.
 In addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small dimension of the  matrices involved and the potential loss in data locality can limit or eliminate the performance gain of this approach.
@@
 It supports all problems expressible with this framework.
-This allows us to evaluate our code transformations in a range of real-world problems, varying two key parameters that impact both solution accuracy and kernel cost: the polynomial order of the method (from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).
-
-Our experiments showed that the originally generated code for non-trivial assembly kernels was sub-optimal.
+This allows us to evaluate our code transformations in a range of real-world problems, varying two key parameters that impact both solution accuracy and kernel cost: the polynomial order of the method (we investigate from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).
+
+Our experiments show that the original generated code for non-trivial assembly kernels is sub-optimal.
 Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over code in which only basic transformations have been performed, such as generalized loop-invariant code motion, padding, and data alignment, and up to 4.4$\times$ over original kernels.
@@
 \begin{algorithm}[t]
-\small
+\footnotesize
 \KwIn{element matrix (2D array, initialized to 0),\\~~~~~~~~~element coordinates (array),\\~~~~~~~~~coefficient fields (array, e.g. velocity)}
@@
 ~~...\\
-~~\\
 ~~\KwSty{for} (\KwSty{int} i = 0; i$<$5; i++) $\lbrace$\\
@@
 ~~~~~~//10 analogous statements (F1, F2, ...)\\
+~~~~~~...\\
 ~~~~$\rbrace$\\
-~~~~...\\
 ~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$12; j++) \\
@@
 The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}.
-The inputs are a zero-initialized two dimensional array used to store the element matrix, the element's coordinates in the discretized domain, and coefficient fields, for instance indicating value of velocity or pressure in the element.
+The inputs are a zero-initialized two dimensional array used to store the element matrix, the element's coordinates in the discretized domain, and coefficient fields, for instance indicating the values of velocity or pressure in the element.
 The output is the evaluated element matrix.
@@
 \begin{center}
-\begin{tabulary}{1.0\columnwidth}{|C|C|C|}
+\begin{tabulary}{1.0\columnwidth}{C|C|C}
 \hline
-Object name & Type & Variable name(s) \\\hline\hline
-Determinant of the Jacobian matrix & double & det  \\ \hline
-Inverse of the Jacobian matrix & double[] & K \\ \hline
-Coordinates & double** & coords\\ \hline
-Fields & double** & w \\ \hline
-Numerical integration weights & double[] & W \\ \hline
-Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ \hline
+Object name & Type & Variable name(s) \\\hline
+Determinant of the Jacobian matrix & double & det  \\ 
+Inverse of the Jacobian matrix & double[] & K \\ 
+Coordinates & double** & coords\\ 
+Fields & double** & w \\ 
+Numerical integration weights & double[] & W \\ 
+Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ 
 Element matrix & double[][] & A\\ \hline
@@
 An example is provided in Figure~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$.
-In other scenarios, for instance when solving a non-linear problem like the Burgers equation, whose assembly code is given in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
+In other scenarios, for instance when solving the Burgers equation, whose assembly code is given in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
 This may also require the evaluation of constants in outer loops (called $F$ in the code) to act as as scaling factors of arrays; trip counts can be larger (proportionally to the order of the method); and arrays may be block-sparse.
@@
 These include: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas interchanges involving the $i$ loop require pre-computation of values (e.g.
-the $F$ values in Burgers) and introduction of temporary arrays, as explained in Section~\ref{sec:code-transf}; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
+the $F$ values in Burgers) and introduction of temporary arrays, as explained in Section~\ref{sec:code-transf}; 3) depending on the problem being solved, the $j$ and $k$ loops could iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
 In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.
@@
 
-As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens iterations, render the optimization process challenging.
+As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens of iterations, render the optimization process challenging.
 It is not always the same set of optimizations that bring performance closest to the machine peak.
@@
 \subsection{Prerequisites for Effective Auto-vectorization: Padding and Data Alignment}
-Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is smaller than the vector length $vl$.
+Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is not a multiple of the vector length $vl$.
 Data alignment is enforced in two steps.
@@
 For instance, assume the original size of a basis function array is 3$\times$3 and $vl=4$ (e.g.
-AVX processor, with 256 bits long vector registers and 64-bits double-precision floating-point values).
+AVX processor, with 256 bit long vector registers and 64-bit double-precision floating-point values).
 In this case, a padded version of the array will have size 3$\times$4.
@@
 At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized as part of an inner loop.
-Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant.
-It is also worth noting that, in some problems, invariant sub-expressions along $j$ are identical to those along $k$ (e.g.
-in Helmholtz).
+Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could otherwise be significant.
+It is also worth noting that, in some problems (e.g.
+in Helmholtz), invariant sub-expressions along $j$ are identical to those along $k$, and both loops iterate over the same iteration space.
 In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.
+
+Figure~\ref{code:helmholtz-licm} shows the Helmholtz assembly code after the application of loop-invariant code motion, padding, and data alignment.
 
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake in which padding, data alignment, and \emph{licm} have been applied to the Helmholtz problem given in Figure~\ref{code:helmholtz} for an AVX architecture.
-In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$, so they can be precomputed once in the $r$ loop.}
+\caption{Local assembly code for the Helmholtz problem in Figure~\ref{code:helmholtz} after application of padding, data alignment, and \emph{licm}, for an AVX architecture.
+In this example, sub-expressions invariant to $j$ are identical to those invariant to $k$, so they can be precomputed once in the $r$ loop.}
 \label{code:helmholtz-licm}
@@
 
-\begin{figure}[h]
+\begin{figure}[b]
 \centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect-inline.pdf}}
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake for the Helmholtz problem in which \emph{op-vect} has been applied on top of the optimizations shown in Figure~\ref{code:helmholtz-licm}.
+\caption{Local assembly code generated by Firedrake for the Helmholtz problem after application of \emph{op-vect} on top of the optimizations shown in Figure~\ref{code:helmholtz-licm}.
 Here, we assume the polynomial order is $p=2$, since \emph{op-vect} can not be used when an iteration space dimension is smaller than the vector length.
@@
 \begin{center}
-\begin{tabulary}{1.0\columnwidth}{|C|C|}
+\begin{tabulary}{1.0\columnwidth}{C|C}
 \hline
-Name (Abbreviation) & Parameter \\\hline\hline
-Generalized loop-invariant code motion (\emph{licm}) &   \\ \hline
-Padding &  \\ \hline
-Data Alignment & \\ \hline
-Loop interchange      & loops  \\ \hline
-Loop unrolling  & unroll factor \\ \hline
-Outer-product vectorization (\emph{op-vect}) & tile size \\ \hline
+Name (Abbreviation) & Parameter \\\hline
+Generalized loop-invariant code motion (\emph{licm}) &   \\ 
+Padding &  \\ 
+Data Alignment & \\ 
+Loop interchange      & loops  \\ 
+Loop unrolling  & unroll factor \\ 
+Outer-product vectorization (\emph{op-vect}) & tile size \\ 
 Expression splitting (\emph{split}) & split point, split factor \\ \hline
@@
 If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the outer, non-affine loop over mesh elements is vectorized, can be applied.
-This feature, currently under development, has been proved to be useful in several finite volume applications~\cite{inter-kernel-vect}.
-Then, an ordered sequence of optimization steps are executed.
+This feature, currently under development, has been shown to be useful in several finite volume applications~\cite{inter-kernel-vect}.
+Subsequently, an ordered sequence of optimization steps are executed.
 Application of \emph{licm} must precede padding and data alignment, due to the introduction of temporary arrays.
@@
 Their implementation is based on analysis and transformation of the kernel's AST.
-When \emph{op-vect} is selected, the compiler outputs proper AVX or AVX-512 intrinsics code.
+When \emph{op-vect} is selected, the compiler outputs AVX or AVX-512 intrinsics code.
 Any possible corner cases are handled: for example, if \emph{op-vect} is to be applied, but the size of the iteration space is not a multiple of the vector length, then a remainder loop, amenable to auto-vectorization, is inserted.
 
-All loops are interchangeable provided that temporaries are introduced if the nest is not perfect.
+All loops are interchangeable, provided that temporaries are introduced if the nest is not perfect.
 For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance.
@@
 We tested this hypothesis in manually crafted kernels.
-We found that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop.
+We found that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than memory, along the $i$ loop.
 The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension.
@@
 In all these cases, the size of the element matrix rarely exceeds 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
+
+In the following, results for the loop order \texttt{ijk} are shown.
+For the considerations exposed in Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner loop over $i$ caused slow downs; also, interchanging $j$ and $k$ loops while keeping $i$ as outermost loop did not provide any benefits.
 
@@
 This is because in the considered kernels the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short.
-On the other hand, as the iteration space becomes bigger, vector-register tiling leads to percentage improvements up to 1.4$\times$ on the Sandy Bridge (Diffusion, prismatic mesh, $p=4$) and up to 1.4$\times$ on the Xeon Phi (Helmholtz, tetrahedral mesh, $p=3$).
+On the other hand, as the iteration space becomes bigger, vector-register tiling leads to improvements up to 1.4$\times$ on the Sandy Bridge (Diffusion, prismatic mesh, $p=4$) and up to 1.4$\times$ on the Xeon Phi (Helmholtz, tetrahedral mesh, $p=3$).
 Using the Intel Architecture Code Analyzer tool~\cite{IACA} on the Sandy Bridge, we confirmed that this is a consequence of increased register re-use.
@@
 \begin{figure}[h]
-\includegraphics[scale=0.67]{Pictures/helmholtz-normalized-opvect.pdf}
+\begin{center}
+\includegraphics[scale=0.65]{Pictures/helmholtz-normalized-opvect.pdf}
+\end{center}
 \caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel on the Sandy Bridge.
@@
 \begin{figure}[h]
-\includegraphics[scale=0.67]{Pictures/helmholtz-normalized-opvect-phi.pdf}
+\begin{center}
+\includegraphics[scale=0.65]{Pictures/helmholtz-normalized-opvect-phi.pdf}
+\end{center}
 \caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel on the Xeon Phi.
@@
 \begin{figure}[h]
-\includegraphics[scale=0.67]{Pictures/diffusion-normalized-opvect.pdf}
+\begin{center}
+\includegraphics[scale=0.65]{Pictures/diffusion-normalized-opvect.pdf}
+\end{center}
 \caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel on the Sandy Bridge.
@@
 \begin{figure}[h]
-\includegraphics[scale=0.67]{Pictures/diffusion-normalized-opvect-phi.pdf}
+\begin{center}
+\includegraphics[scale=0.65]{Pictures/diffusion-normalized-opvect-phi.pdf}
+\end{center}
 \caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel on the Xeon Phi.
@@
 
+On the Sandy Bridge, the performance of the Burgers kernel on a prismatic mesh was 20.0 GFlop/s from $p=1$ to $p=3$, while it was 21.3 GFlop/s in the case of $p=4$.
+These values are notably close to the peak performance of 30.4 GFlop/s.
+Disabling \emph{split} makes the performance drop to 17.0 GFlop/s for $p=1, 2$, 18.2 GFlop/s for $p=3$, and 14.3 GFlop/s for $p=4$.
+These values are in line with the speedups shown in Figure~\ref{fig:split-burgers-speedup}.
+
 The \emph{split} transformation was also tried in combination with \emph{op-vect} (\emph{split-op-vect}), although the cost model prevents its adoption on both platforms.
@@
 \begin{figure}[h]
-\includegraphics[scale=0.67]{Pictures/burgers-normalized-split.pdf}
+\begin{center}
+\includegraphics[scale=0.65]{Pictures/burgers-normalized-split.pdf}
+\end{center}
 \caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel on the Sandy Bridge.
@@
 \begin{figure}[h]
-\includegraphics[scale=0.67]{Pictures/burgers-normalized-split-phi.pdf}
+\begin{center}
+\includegraphics[scale=0.65]{Pictures/burgers-normalized-split-phi.pdf}
+\end{center}
 \caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel on the Xeon Phi.
@@
 \label{sec:perf-results-fenics}
-We have modified the FEniCS Form Compiler (FFC) to make it return an abstract syntax tree representation of a local assembly kernel, rather than plain C++ code, so as to enable code transformations in COFFEE.
+We have modified the FEniCS Form Compiler (FFC) to return an abstract syntax tree representation of a local assembly kernel, rather than plain C++ code, so as to enable code transformations in COFFEE.
 Besides Firedrake, FFC is used in the FEniCS project~\cite{Fenics}.
@@
 \begin{center}
-\begin{tabulary}{1.0\columnwidth}{|C|C|C|}
+\begin{tabulary}{1.0\columnwidth}{C|C|C}
 \hline
-Problem & Max slow down & Max speed up \\\hline\hline
-Helmholtz & / & 4.14$\times$ \\
-Diffusion & / & 4.28$\times$ \\
+Problem & Max slow down & Max speed up \\\hline
+Helmholtz & - & 4.14$\times$ \\
+Diffusion & - & 4.28$\times$ \\
 Burgers & 2.24$\times$ & 1.61$\times$ \\\hline

### 2014-06-22 SC-paper/bare_conf.tex
Started re-writing section background

--- 2014-04-16 paper/bare_conf.tex
@@
 Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective.
-The solution is sought in each element of the discretized domain by applying suitable numerical operations.
+The solution is sought in each cell, or ``element'', of the discretized domain by applying suitable numerical operations.
 As the number of elements can be of the order of trillions \cite{Rossinelli2013}, a major issue is the time required to execute the computation, which can be hours or days.
@@
 
-\section{Background}
+\section{Background and Motivating Examples}
 \label{sec:background}
@@
 \nlset{1}~~- Compute Jacobian from coordinates \\
-~~- Declare of constant data: basis functions and derivatives\\
-~~- Compute element matrix using numerical quadrature
+~~- Define basis functions \\
+~~- Compute element matrix in an affine loop nest
 \caption{General structure of a local assembly kernel generated by Firedrake.}
@@
 Local assembly is the computation of contributions of a specific cell in the discretized domain to the PDE solution.
-The process consists of numerically evaluating an integral to produce a small dense element matrix or an element vector~\cite{quadrature1, fluidity_manual_v4}.
+The process consists of numerically evaluating problem-specific integrals to produce a matrix and a vector (\cite{quadrature1, fluidity_manual_v4}), whose sizes depend on the order of the method.
 This operation is applied to all cells in the discretized domain.
@@
 
-Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm.
-It then triggers compilation of such kernels using a vendor compiler, and manages parallel execution over the mesh.
-The subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.
-The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used.
-
-The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}.
+Given a finite element description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl} (UFL), Firedrake generates a C-code kernel implementing assembly using a numerical integration algorithm.
+The same kernel can be applied to any elements: this descends from the mathematical formulation of assembly, which states that the evaluation of a local matrix can be reduced to integrating on a fixed ``reference'' element - a special element that does not belong to the domain - after a suitable change of coordinates.
+Firedrake triggers the compilation of an assembly kernel using an available vendor compiler, and manages its (parallel) execution over all elements in the mesh.
+As already stated, the subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.
+
+%The code transformations described in this work are also generalizable to non-Firedrake assembly kernels, provided that the same numerical integration algorithm is employed.
+
+The structure of a local assembly kernel is shown in Figure~\ref{code:general-structure}.
 The inputs are a zero-initialized two dimensional array used to store the element matrix, the element's coordinates in the discretized domain, and coefficient fields, for instance indicating the values of velocity or pressure in the element.
 The output is the evaluated element matrix.
-A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of a problem-specific set of constant two dimensional arrays of double precision floats containing the values of the basis functions and their derivatives at the integration points; 3) evaluation of the element matrix in an affine loop nest.
-All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (i.e.
-using \texttt{static const} in C).
+The kernel body can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse: required for the aforementioned change of coordinates, from the reference element to the one being computed; 2) definition of basis functions: used, intuitively, to interpolate the contribution to the PDE solution over the element.
+The choice of the basis functions is expressed in UFL directly by users.
+In the generated code, they are represented as global read-only two dimensional arrays (i.e.
+using \texttt{static const} in C) of double precision floats; 3) evaluation of the element matrix in an affine loop nest, in which the integration is performed.
 Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel objects.
+
+%JUNK SENTENCES: Since only the values assumed at the integration points are needed to compute the element matrix, a basis function is implemented as a two dimensional array of double precision floats.
+The same basis functions can be used to evaluate all element matrices, provided that a suitable change of coordinates from a ``reference'' element to a real one is performed.
+Therefore, basis functions are declared in the kernel body as global read-only arrays (i.e.
+using \texttt{static const} in C), rather than passed as function parameters.
+Basis functions are chosen at the UFL layer based on the problem being solved, so they are actually an input to the code generator.
 
@@
 Determinant of the Jacobian matrix & double & det  \\ 
-Inverse of the Jacobian matrix & double[] & K \\ 
+Inverse of the Jacobian matrix & double & K1, K2, ... \\ 
 Coordinates & double** & coords\\ 
@@
 In other scenarios, for instance when solving the Burgers equation, whose assembly code is given in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
-This may also require the evaluation of constants in outer loops (called $F$ in the code) to act as as scaling factors of arrays; trip counts can be larger (proportionally to the order of the method); and arrays may be block-sparse.
+This may also require the evaluation of constants in outer loops (called $F$ in the code) to act as scaling factors of arrays; trip counts can be larger (proportionally to the order of the method); and arrays may be block-sparse.
 In addition to a larger number of operations, more complex cases like the Burgers equation are characterized by high register pressure.
+The Helmholtz and Burgers equations, along with the Diffusion problem,
 
@@
 The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake.
-One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, in which the numerical integration takes place.
-In rare cases, as in problems involving hyperelasticity, however, such optimizations might not be sufficient to achieve notable performance improvements.
-Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions.
-It is worth noting there are circumstances in which the number of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so large (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time.
-In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure).
-A comprehensive study of such cases is left as further work.
-
 As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens of iterations, render the optimization process challenging.

### 2014-06-22 TACO-paper/coffee.tex
Add stuff to TACO format

--- 2014-06-22 SC-paper/bare_conf.tex
@@
-\documentclass[conference]{IEEEtran}
-% Add the compsoc option for Computer Society conferences.
+% v2-acmsmall-sample.tex, dated March 6 2012
+% This is a sample file for ACM small trim journals
 %
-% If IEEEtran.cls has not been installed into the LaTeX system files,
-% manually specify the path to it like:
-% \documentclass[conference]{../sty/IEEEtran}
-
-% TODO: ok?
+% Compilation using 'acmsmall.cls' - version 1.3 (March 2012), Aptara Inc.
+% (c) 2010 Association for Computing Machinery (ACM)
+%
+% Questions/Suggestions/Feedback should be addressed to => "acmtexsupport@aptaracorp.com".
+% Users can also go through the FAQs available on the journal's submission webpage.
+%
+% Steps to compile: latex, bibtex, latex latex
+%
+% For tracking purposes => this is v1.3 - March 2012
+
+\documentclass[prodmode,acmtecs]{acmsmall} % Aptara syntax
 
@@
 \usepackage{tabulary}
-
-% Some very useful LaTeX packages include:
-% (uncomment the ones you want to load)
-
-% *** MISC UTILITY PACKAGES ***
-%
-%\usepackage{ifpdf}% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
-% compilation based on whether the output is pdf or dvi.
-% usage:
-% \ifpdf
-%   % pdf code
-% \else
-%   % dvi code
-% \fi
-% The latest version of ifpdf.sty can be obtained from:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
-% Also, note that IEEEtran.cls V1.7 and later provides a builtin
-% \ifCLASSINFOpdf conditional that works the same way.
-% When switching from latex to pdflatex and vice-versa, the compiler may
-% have to be run twice to clear warning/error messages.
-
-% *** CITATION PACKAGES ***
-%
-%\usepackage{cite}
-% cite.sty was written by Donald Arseneau
-% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
-% \cite{} output to follow that of IEEE. Loading the cite package will
-% result in citation numbers being automatically sorted and properly
-% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
-% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
-% \cite will automatically add leading space, if needed. Use cite.sty's
-% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
-% cite.sty is already installed on most LaTeX systems. Be sure and use
-% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
-% not currently provide for hyperlinked citations.
-% The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
-% The documentation is contained in the cite.sty file itself.
-
-% *** GRAPHICS RELATED PACKAGES ***
-%
-\ifCLASSINFOpdf
-  \usepackage[pdftex]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../pdf/}{../jpeg/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
-\else
-  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
-  % will default to the driver specified in the system graphics.cfg if no
-  % driver is specified.
-  % \usepackage[dvips]{graphicx}
-  % declare the path(s) where your graphic files are
-  % \graphicspath{{../eps/}}
-  % and their extensions so you won't have to specify these with
-  % every instance of \includegraphics
-  % \DeclareGraphicsExtensions{.eps}
-\fi
-% graphicx was written by David Carlisle and Sebastian Rahtz. It is
-% required if you want graphics, photos, etc. graphicx.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
-% Another good source of documentation is "Using Imported Graphics in
-% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
-% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
-%
-% latex, and pdflatex in dvi mode, support graphics in encapsulated
-% postscript (.eps) format. pdflatex in pdf mode supports graphics
-% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
-% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
-% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
-% which can result in "jaggedy"/blurry rendering of lines and letters as
-% well as large increases in file sizes.
-%
-% You can find documentation about the pdfTeX application at:
-% http://www.tug.org/applications/pdftex
-
-% *** MATH PACKAGES ***
-%
 \usepackage[cmex10]{amsmath}
-% A popular package from the American Mathematical Society that provides
-% many useful and powerful commands for dealing with mathematics. If using
-% it, be sure to load this package with the cmex10 option to ensure that
-% only type 1 fonts will utilized at all point sizes. Without this option,
-% it is possible that some math symbols, particularly those within
-% footnotes, will be rendered in bitmap form which will result in a
-% document that can not be IEEE Xplore compliant!
-%
-% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
-% thus preventing page breaks from occurring within multiline equations. Use:
-%\interdisplaylinepenalty=2500
-% after loading amsmath to restore such page breaks as IEEEtran.cls normally
-% does. amsmath.sty is already installed on most LaTeX systems. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
-
-% *** SPECIALIZED LIST PACKAGES ***
-%
-\usepackage[longend,linesnumbered,figure]{algorithm2e}
-\usepackage{algorithmic}
-
-\newcommand{\code}[1]{\lstset{basicstyle=\small\tt}\lstinline£#1£\lstset
-{basicstyle=\footnotesize\tt}}
-
-% algorithmic.sty was written by Peter Williams and Rogerio Brito.
-% This package provides an algorithmic environment fo describing algorithms.
-% You can use the algorithmic environment in-text or within a figure
-% environment to provide for a floating algorithm. Do NOT use the algorithm
-% floating environment provided by algorithm.sty (by the same authors) or
-% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
-% algorithm float types and packages that provide these will not provide
-% correct IEEE style captions. The latest version and documentation of
-% algorithmic.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
-% There is also a support site at:
-% http://algorithms.berlios.de/index.html
-% Also of interest may be the (relatively newer and more customizable)
-% algorithmicx.sty package by Szasz Janos:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
-
-% *** ALIGNMENT PACKAGES ***
-%
-%\usepackage{array}
-% Frank Mittelbach's and David Carlisle's array.sty patches and improves
-% the standard LaTeX2e array and tabular environments to provide better
-% appearance and additional user controls. As the default LaTeX2e table
-% generation code is lacking to the point of almost being broken with
-% respect to the quality of the end results, all users are strongly
-% advised to use an enhanced (at the very least that provided by array.sty)
-% set of table tools. array.sty is already installed on most systems. The
-% latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/required/tools/
-
-%\usepackage{mdwmath}
-%\usepackage{mdwtab}
-% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
-% especially mdwmath.sty and mdwtab.sty which are used to format equations
-% and tables, respectively. The MDWtools set is already installed on most
-% LaTeX systems. The lastest version and documentation is available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
-
-% IEEEtran contains the IEEEeqnarray family of commands that can be used to
-% generate multiline equations as well as matrices, tables, etc., of high
-% quality.
-
-%\usepackage{eqparbox}
-% Also of notable interest is Scott Pakin's eqparbox package for creating
-% (automatically sized) equal width boxes - aka "natural width parboxes".
-% Available at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
-
-% *** SUBFIGURE PACKAGES ***
-%\usepackage[tight,footnotesize]{subfigure}
-% subfigure.sty was written by Steven Douglas Cochran. This package makes it
-% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
-% work, it is a good idea to load it with the tight package option to reduce
-% the amount of white space around the subfigures. subfigure.sty is already
-% installed on most LaTeX systems. The latest version and documentation can
-% be obtained at:
-% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
-% subfigure.sty has been superceeded by subfig.sty.
-
-%\usepackage[caption=false]{caption}
-%\usepackage[font=footnotesize]{subfig}
-% subfig.sty, also written by Steven Douglas Cochran, is the modern
-% replacement for subfigure.sty. However, subfig.sty requires and
-% automatically loads Axel Sommerfeldt's caption.sty which will override
-% IEEEtran.cls handling of captions and this will result in nonIEEE style
-% figure/table captions. To prevent this problem, be sure and preload
-% caption.sty with its "caption=false" package option. This is will preserve
-% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later
-% (recommended due to many improvements over 1.2) of subfig.sty supports
-% the caption=false option directly:
-%\usepackage[caption=false,font=footnotesize]{subfig}
-%
-% The latest version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
-% The latest version and documentation of caption.sty can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
-
-% *** FLOAT PACKAGES ***
-%
-%\usepackage{fixltx2e}
-% fixltx2e, the successor to the earlier fix2col.sty, was written by
-% Frank Mittelbach and David Carlisle. This package corrects a few problems
-% in the LaTeX2e kernel, the most notable of which is that in current
-% LaTeX2e releases, the ordering of single and double column floats is not
-% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
-% single column figure to be placed prior to an earlier double column
-% figure. The latest version and documentation can be found at:
-% http://www.ctan.org/tex-archive/macros/latex/base/
-
-%\usepackage{stfloats}
-% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
-% the ability to do double column floats at the bottom of the page as well
-% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
-% LaTeX2e). It also provides a command:
-%\fnbelowfloat
-% to enable the placement of footnotes below bottom floats (the standard
-% LaTeX2e kernel puts them above bottom floats). This is an invasive package
-% which rewrites many portions of the LaTeX2e float routines. It may not work
-% with other packages that modify the LaTeX2e float routines. The latest
-% version and documentation can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
-% Documentation is contained in the stfloats.sty comments as well as in the
-% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
-% does not allow \baselineskip to stretch. Authors submitting work to the
-% IEEE should note that IEEE rarely uses double column equations and
-% that authors should try to avoid such use. Do not be tempted to use the
-% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
-% not format its papers in such ways.
-
-% *** PDF, URL AND HYPERLINK PACKAGES ***
-%
 \usepackage{url}
-% url.sty was written by Donald Arseneau. It provides better support for
-% handling and breaking URLs. url.sty is already installed on most LaTeX
-% systems. The latest version can be obtained at:
-% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
-% Read the url.sty source comments for usage information. Basically,
-% \url{my_url_here}.
-
-% *** Do not adjust lengths that control margins, column widths, etc. ***
-% *** Do not use packages that alter fonts (such as pslatex).         ***
-% There should be no need to do such things with IEEEtran.cls V1.6 and later.
-% (Unless specifically asked to do so by the journal or conference you plan
-% to submit to, of course. )
-
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
-
+
+% Package to generate and customize Algorithm as per ACM style
+\usepackage[ruled]{algorithm2e}
+\renewcommand{\algorithmcfname}{ALGORITHM}
+\SetAlFnt{\small}
+\SetAlCapFnt{\small}
+\SetAlCapNameFnt{\small}
+\SetAlCapHSkip{0pt}
+\IncMargin{-\parindent}
+
+% Metadata Information
+\acmVolume{9}
+\acmNumber{4}
+\acmArticle{39}
+\acmYear{2010}
+\acmMonth{3}
+
+% Document starts
 \begin{document}
-%
-% paper title
-% can use linebreaks \\ within to get better formatting as desired
-\title{COFFEE: an Optimizing Compiler\\for Finite Element Local Assembly}
-
-\author{\IEEEauthorblockN{Fabio Luporini\IEEEauthorrefmark{1},
-Ana Lucia Varbanescu\IEEEauthorrefmark{2},
-Florian Rathgeber\IEEEauthorrefmark{1},
-Gheorghe-Teodor Bercea\IEEEauthorrefmark{1}, \\
-J. Ramanujam\IEEEauthorrefmark{3},
-David A. Ham\IEEEauthorrefmark{4}\IEEEauthorrefmark{1},
-and
-Paul H. J. Kelly\IEEEauthorrefmark{1}}
-\IEEEauthorblockA{\IEEEauthorrefmark{1}Imperial College London, Department of Computing,\\
-\IEEEauthorrefmark{4}Imperial College London, Department of Mathematics,\\
-\IEEEauthorrefmark{2}University of Amsterdam, Department of Computer Sciences,\\
-\IEEEauthorrefmark{3}Louisiana State University, Department of Electrical and Computer Engineering,\\
-\IEEEauthorrefmark{1}\{f.luporini12$|$f.rathgeber10$|$gheorghe-teodor.bercea08$|$p.kelly$|$david.ham\}@imperial.ac.uk,\\
-\IEEEauthorrefmark{2}a.l.varbanescu@uva.nl,
-\IEEEauthorrefmark{3}jxr@ece.lsu.edu
-}
-}
-
-% conference papers do not typically use \thanks and this command
-% is locked out in conference mode. If really needed, such as for
-% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
-% after \documentclass
-
-% for over three affiliations, or if they all won't fit within the width
-% of the page, use this alternative format:
-%
-%\author{\IEEEauthorblockN{Michael Shell\IEEEauthorrefmark{1},
-%Homer Simpson\IEEEauthorrefmark{2},
-%James Kirk\IEEEauthorrefmark{3},
-%Montgomery Scott\IEEEauthorrefmark{3} and
-%Eldon Tyrell\IEEEauthorrefmark{4}}
-%\IEEEauthorblockA{\IEEEauthorrefmark{1}School of Electrical and Computer Engineering\\
-%Georgia Institute of Technology,
-%Atlanta, Georgia 30332--0250\\ Email: see http://www.michaelshell.org/contact.html}
-%\IEEEauthorblockA{\IEEEauthorrefmark{2}Twentieth Century Fox, Springfield, USA\\
-%Email: homer@thesimpsons.com}
-%\IEEEauthorblockA{\IEEEauthorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
-%Telephone: (800) 555--1212, Fax: (888) 555--1212}
-%\IEEEauthorblockA{\IEEEauthorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
-\maketitle
-%OLD 1
-%The finite element method is widely-employed to approximate solutions of partial differential equations.
-Local assembly is one of its steps, in which a problem-specific kernel is applied to all elements in the discretized domain of the equation.
-Since the domain size can be huge, executing efficient kernels is fundamental.
-Automating their optimization is, however, a challenging issue.
-Even if an affine loop nest can be generally distinguished, small problem size and complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations.
-In this context, we present COFFEE, a domain-specific compiler for local assembly kernels.
-COFFEE manipulates kernels’ abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality.
-It then generates C code including vector intrinsics.
-Systematic performance evaluation using a suite of real-world examples shows that speed-ups over non-optimized kernels up to 4.44$\times$ are achievable.
-
-%OLD 2 before feedback
-%The finite element method is widely-employed to approximate solutions of partial differential equations.
-Local assembly is one of its steps, in which a problem-specific kernel is applied to all elements in the discretized domain of the equation.
-Since the domain size can be huge, executing efficient kernels is fundamental.
-Automating their optimization is, however, a challenging issue.
-Even if an affine loop nest can be generally distinguished, small problem size and complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations.
-In this context, we present design and systematic evaluation of COFFEE, a domain-specific compiler for local assembly kernels.
-COFFEE manipulates abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality.
-It then generates C code including vector intrinsics.
-Performance evaluation using a suite of real-world examples shows that signifcant speed-ups up are achievable.
+
+% Page heads
+\markboth{F. Luporini et al.}{COFFEE: an Optimizing Compiler for Finite Element Local Assembly}
+
+% Title portion
+\title{COFFEE: an Optimizing Compiler for Finite Element Local Assembly}
+\author{GANG ZHOU
+\affil{College of William and Mary}
+YAFENG WU
+\affil{University of Virginia}
+TING YAN
+\affil{Eaton Innovation Center}
+TIAN HE
+\affil{University of Minnesota}
+CHENGDU HUANG
+\affil{Google}
+JOHN A. STANKOVIC
+\affil{University of Virginia}
+TAREK F. ABDELZAHER
+\affil{University of Illinois at Urbana-Champaign}}
+% NOTE! Affiliations placed here should be for the institution where the
+%       BULK of the research was done. If the author has gone to a new
+%       institution, before publication, the (above) affiliation should NOT be changed.
+%       The authors 'current' address may be given in the "Author's addresses:" block (below).
+%       So for example, Mr. Abdelzaher, the bulk of the research was done at UIUC, and he is
+%       currently affiliated with NASA.
 
 \begin{abstract}
-%\boldmath
 The numerical solution of partial differential equations using the finite
@@
 
-% IEEEtran.cls defaults to using nonbold math in the Abstract.
-% This preserves the distinction between vectors and scalars. However,
-% if the conference you are submitting to favors bold math in the abstract,
-% then you can use LaTeX's standard command \boldmath at the very start
-% of the abstract to achieve this. Many IEEE journals/conferences frown on
-% math in the abstract anyway.
-
-\begin{IEEEkeywords}
-Finite element integration, local assembly, compilers, optimizations, simd vectorization
-\end{IEEEkeywords}
-
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-\IEEEpeerreviewmaketitle
+\category{C.2.2}{Computer-Communication Networks}{Network Protocols}
+
+\terms{Design, Algorithms, Performance}
+
+\keywords{Finite element integration, local assembly, compilers, optimizations, simd vectorization}
+
+\acmformat{Gang Zhou, Yafeng Wu, Ting Yan, Tian He, Chengdu Huang, John A. Stankovic,
+and Tarek F. Abdelzaher, 2010. A multifrequency MAC specially
+designed for  wireless sensor network applications.}
+% At a minimum you need to supply the author names, year and a title.
+% IMPORTANT:
+% Full first names whenever they are known, surname last, followed by a period.
+% In the case of two authors, 'and' is placed between them.
+% In the case of three or more authors, the serial comma is used, that is, all author names
+% except the last one but including the penultimate author's name are followed by a comma,
+% and then 'and' is placed before the final author's name.
+% If only first and middle initials are known, then each initial
+% is followed by a period and they are separated by a space.
+% The remaining information (journal title, volume, article number, date, etc.) is 'auto-generated'.
+
+\begin{bottomstuff}
+This research is partly funded by the MAPDES project, by the Department of Computing at Imperial College London, by EPSRC through grants EP/I00677X/1, EP/I006761/1, and EP/L000407/1, and by NERC grants NE/K008951/1 and NE/K006789/1.
+The authors would like to thank Dr.
+Carlo Bertolli, Dr.
+Lawrence Mitchell, and Dr.
+Francis Russell for their invaluable suggestions and their contribution to the Firedrake project.
+
+Author's addresses: G. Zhou, Computer Science Department,
+College of William and Mary; Y. Wu  {and} J. A. Stankovic,
+Computer Science Department, University of Virginia; T. Yan,
+Eaton Innovation Center; T. He, Computer Science Department,
+University of Minnesota; C. Huang, Google; T. F. Abdelzaher,
+(Current address) NASA Ames Research Center, Moffett Field, California 94035.
+\end{bottomstuff}
+
+\maketitle
 
@@
 The contributions of this paper are
-\begin{itemize}
+\begin{enumerate}
 \item An optimization strategy for finite element local assembly. Our approach exploits domain knowledge and goes beyond the limits of both vendor and research compilers.
@@
 \item Systematic analysis using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
-\end{itemize}
+\end{enumerate}
 
@@
 \KwOut{element matrix (2D array)}
-\nlset{1}~~- Compute Jacobian from coordinates \\
+~~- Compute Jacobian from coordinates \\
 ~~- Define basis functions \\
@@
 ~~\\
-~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
+~~\KwSty{static const double} W[3] = $\lbrace$...$\rbrace$\\
 ~~\KwSty{static const double} X$\_$D10[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
@@
 ~~~~~~~~A[j][k] += ((Y[i][k]*Y[i][j]+\\
-~~~~~~~~~~~+((K1*X$\_$D10[i][k]+K3*X$\_$D01[i][k])*\\
-~~~~~~~~~~~*(K1*X$\_$D10[i][j]+K3*X$\_$D01[i][j]))+\\
-~~~~~~~~~~~+((K0*X$\_$D10[i][k]+K2*X$\_$D01[i][k])*\\
-~~~~~~~~~~~*(K0*X$\_$D10[i][j]+K2*X$\_$D01[i][j])))*\\
-~~~~~~~~~~~*det*W3[i]);\\
+~~~~~~~~~~~+((K1*X$\_$D10[i][k]+K3*X$\_$D01[i][k])*(K1*X$\_$D10[i][j]+K3*X$\_$D01[i][j]))+\\
+~~~~~~~~~~~+((K0*X$\_$D10[i][k]+K2*X$\_$D01[i][k])*(K0*X$\_$D10[i][j]+K2*X$\_$D01[i][j])))*\\
+~~~~~~~~~~~*det*W[i]);\\
 $\rbrace$
@@
 ~~\\
-~~\KwSty{static const double} W5[5] = $\lbrace$...$\rbrace$\\
+~~\KwSty{static const double} W[5] = $\lbrace$...$\rbrace$\\
 ~~\KwSty{static const double} X1$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
@@
 ~~~~~~~~A[j][k] += (..(K5*F9)+(K8*F10))*Y1[i][j])+\\
-~~~~~~~~~+(((K0*X1$\_$D100[i][k])+(K3*X1$\_$D010[i][k])+\\
-~~~~~~~~~+(K6*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
-~~~~~~~~~+(..((K2*X2$\_$D100[i][k])+...+(K8*X2$\_$D001[i][k])*\\
-~~~~~~~~~*(K2*X2$\_$D100[i][j])+...+(K8*X2$\_$D001[i][j]))..)+\\
+~~~~~~~~~+(((K0*X1$\_$D100[i][k])+(K3*X1$\_$D010[i][k])+(K6*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
+~~~~~~~~~+(..((K2*X2$\_$D100[i][k])+...+(K8*X2$\_$D001[i][k]))*((K2*X2$\_$D100[i][j])+...+(K8*X2$\_$D001[i][j]))..)+\\
 ~~~~~~~~~+ $<$roughly a hundred sum/muls go here$>$)..)*\\
-~~~~~~~~~*det*W5[i]);\\
+~~~~~~~~~*det*W[i]);\\
 ~~$\rbrace$ \\
@@
 \end{center}
+\label{table:map-name-letters}
 \caption{Type and variable names used in the various code snippets to identify local assembly objects.}
-\label{table:map-name-letters}
 \end{table}
@@
 ~~\\
-~~\KwSty{static const double} W3[3] ALIGN = $\lbrace$...$\rbrace$\\
+~~\KwSty{static const double} W[3] ALIGN = $\lbrace$...$\rbrace$\\
 ~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
@@
 ~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
-~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+\\
-~~~~~~~~~~~~~~~~~~~~+LI$\_$1[k]*LI$\_$1[j])*det*W3[i]);\\
+~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+LI$\_$1[k]*LI$\_$1[j])*det*W[i]);\\
 ~~$\rbrace$\\
@@
 Our code transformations were evaluated in three real-world problems based on the following PDEs:
-\begin{itemize}
+\begin{enumerate}
 \item Helmholtz
@@
 \item Burgers
-\end{itemize}
+\end{enumerate}
 The code was written in UFL (available at~\cite{ufl-code}) and then executed over real unstructured meshes through Firedrake.
@@
 
-% use section* for acknowledgement
-\section*{Acknowledgment}
-
-This research is partly funded by the MAPDES project, by the Department of Computing at Imperial College London, by EPSRC through grants EP/I00677X/1, EP/I006761/1, and EP/L000407/1, and by NERC grants NE/K008951/1 and NE/K006789/1.
-The authors would like to thank Dr.
-Carlo Bertolli, Dr.
-Lawrence Mitchell, and Dr.
-Francis Russell for their invaluable suggestions and their contribution to the Firedrake project.
-
-\bibliographystyle{plain}	% (uses file "plain.bst")
-\bibliography{biblio}		% expects file "myrefs.bib"
-
-% that's all folks
+% Bibliography
+\bibliographystyle{ACM-Reference-Format-Journals}
+\bibliography{biblio}
+                             % Sample .bib file with references that match those in
+                             % the 'Specifications Document (V1.5)' as well containing
+                             % 'legacy' bibs and bibs with 'alternate codings'.
+                             % Gerry Murray - March 2012
+
+% Electronic Appendix
+\elecappendix
+
+\medskip
+
 \end{document}

### 2014-06-23 TACO-paper/coffee.tex
went or adjusting stuff

@@
 
-\usepackage{tabularx}
 \usepackage{tabulary}
@@
 \title{COFFEE: an Optimizing Compiler for Finite Element Local Assembly}
-\author{GANG ZHOU
-\affil{College of William and Mary}
-YAFENG WU
-\affil{University of Virginia}
-TING YAN
-\affil{Eaton Innovation Center}
-TIAN HE
-\affil{University of Minnesota}
-CHENGDU HUANG
-\affil{Google}
-JOHN A. STANKOVIC
-\affil{University of Virginia}
-TAREK F. ABDELZAHER
-\affil{University of Illinois at Urbana-Champaign}}
+\author{Fabio Luporini
+\affil{Imperial College London}
+Ana Lucia Varbanescu
+\affil{University of Amsterdam}
+Florian Rathgeber
+\affil{Imperial College London}
+Gheorghe-Teodor Bercea
+\affil{Imperial College London}
+J. Ramanujam
+\affil{Louisiana State University}
+David A. Ham
+\affil{Imperial College London}
+Paul H.J. Kelly
+\affil{Imperial College London}}
 % NOTE! Affiliations placed here should be for the institution where the
@@
 \begin{abstract}
-The numerical solution of partial differential equations using the finite
-element method is one of the key applications of high performance
-computing. Local assembly is its characteristic operation. This entails the
-execution of a problem-specific kernel for each element in the discretized problem domain.
+The numerical solution of partial differential equations using the finite element method is one of the key applications of high performance computing.
+Local assembly is its characteristic operation.
+This entails the execution of a problem-specific kernel for each element in the discretized problem domain.
 Since the domain size can be huge, executing efficient kernels is fundamental.
@@
 
-\category{C.2.2}{Computer-Communication Networks}{Network Protocols}
+\category{G.1.8}{Numerical Analysis}{Partial Differential Equations - Finite element methods}
 
-\terms{Design, Algorithms, Performance}
+\terms{Design, Performance}
 
@@
 
-\acmformat{Gang Zhou, Yafeng Wu, Ting Yan, Tian He, Chengdu Huang, John A. Stankovic,
-and Tarek F. Abdelzaher, 2010. A multifrequency MAC specially
-designed for  wireless sensor network applications.}
+\acmformat{Fabio Luporini, Ana Lucia Varbanescu, Florian Rathgeber, Gheorghe-Teodor Bercea, J.
+Ramanujam, David A.
+Ham, and Paul H.
+J.
+Kelly, 2014.
+COFFEE: an Optimizing Compilerfor Finite Element Local Assembly.}
 % At a minimum you need to supply the author names, year and a title.
@@
 
-Author's addresses: G. Zhou, Computer Science Department,
-College of William and Mary; Y. Wu  {and} J. A. Stankovic,
-Computer Science Department, University of Virginia; T. Yan,
-Eaton Innovation Center; T. He, Computer Science Department,
-University of Minnesota; C. Huang, Google; T. F. Abdelzaher,
-(Current address) NASA Ames Research Center, Moffett Field, California 94035.
+Author's addresses: Fabio Luporini $\&$ Florian Rathgeber $\&$ Gheorghe-Teodor Bercea $\&$ Paul H.
+J.
+Kelly, Department of Computing, Imperial College London; Ana Lucia Varbanescu, Department of Computer Sciences, University of Amsterdam; David A.
+Ham, Department of Computing and Department of Mathematics, Imperial College London; J.
+Ramanujam, Department of Electrical and Computer Engineering, Louisiana State University.
 \end{bottomstuff}
@@
 Local assembly is the computation of contributions of a specific cell in the discretized domain to the PDE solution.
-The process consists of numerically evaluating problem-specific integrals to produce a matrix and a vector (\cite{quadrature1, fluidity_manual_v4}), whose sizes depend on the order of the method.
+The process consists of numerically evaluating problem-specific integrals to produce a matrix and a vector (\cite{quadrature1}, \cite{fluidity_manual_v4}), whose sizes depend on the order of the method.
 This operation is applied to all cells in the discretized domain.
@@
 \begin{center}
+\tbl{Type and variable names used in the various code snippets to identify local assembly objects.}{
 \begin{tabulary}{1.0\columnwidth}{C|C|C}
@@
 \end{tabulary}
+}
 \end{center}
 \label{table:map-name-letters}
-\caption{Type and variable names used in the various code snippets to identify local assembly objects.}
 \end{table}
@@
 \begin{center}
+\tbl{Overview of code transformations for Firedrake-generated assembly kernels.}{
 \begin{tabulary}{1.0\columnwidth}{C|C}
@@
 \end{tabulary}
+}
 \end{center}
-\caption{Overview of code transformations for Firedrake-generated assembly kernels.}
 \label{table:code-transformations}
@@
 \newcommand{\massmatrixtwodresultsnorms}{
-\begin{tabularx}{1.0\textwidth}{cccccc|cccc|cccc|cccc}
+\begin{tabulary}{1.0\textwidth}{cccccc|cccc|cccc|cccc}
 \cline{3-18}
@@
 \cline{1-18}
-\end{tabularx}
+\end{tabulary}
 }
 
-\begin{table*}[t]\footnotesize
+\begin{table*}[t]
+\tbl{Performance improvement due to generalized loop-invariant code motion (\emph{licm} columns) on the Helmholtz, Diffusion and Burgers problems, varying the element shape (triangle, tetrahedron, prism) and the polynomial order ($p \in [1, 4]$), using the Lagrange element family, over the original, non-optimized code.
+The \emph{licm-ap} columns illustrate the effect of combining \emph{licm} with data alignment and padding.
+Results are shown for both the Sandy Bridge and the Xeon Phi architectures.}{
+\scriptsize
 \massmatrixtwodresultsnorms
-~
-\caption{Performance improvement due to generalized loop-invariant code motion (\emph{licm} columns) on the Helmholtz, Diffusion and Burgers problems, varying the element shape (triangle, tetrahedron, prism) and the polynomial order ($p \in [1, 4]$), using the Lagrange element family, over the original, non-optimized code.
-The \emph{licm-ap} columns illustrate the effect of combining \emph{licm} with data alignment and padding.
-Results are shown for both the Sandy Bridge and the Xeon Phi architectures.}
+}
 \label{table:perf-results-licm}
@@
 \begin{center}
+\tbl{Performance comparison between FEniCS (optimizations enabled) and COFFEE on the Sandy Bridge.}{
 \begin{tabulary}{1.0\columnwidth}{C|C|C}
@@
 \end{tabulary}
+}
 \end{center}
-\caption{Performance comparison between FEniCS (optimizations enabled) and COFFEE on the Sandy Bridge.}
 \label{table:comparison-to-FFC-opt}
@@
 
-% Electronic Appendix
-\elecappendix
-
 \medskip

### 2014-06-23 TACO-paper/coffee.tex
Clarify why we do not consider high order methods

@@
 
-We focus on improving the performance of local assembly (``assembly'', in the following), a fundamental step in the finite element method that covers an important fraction of the overall computation run-time, often in the range 30-60$\%$.
+We focus on improving the performance of local assembly (or ``assembly'', in the following), a fundamental step in the finite element method that covers an important fraction of the overall computation run-time, often in the range 30-60$\%$.
 During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain.
-A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed.
-However, we do not consider the latter case.
+In this work, we focus on relatively low order finite element methods, in which an assembly kernel's working set is usually small enough to fit the L1 cache.
+Low order methods are by no means exotic: they are employed in a wide variety of fields, including climate and ocean modeling, computational fluid dynamics, and structural mechanics.
+High order methods are often based upon special formulations of the finite element method, which result in the adoption of different sets of algorithms to solve PDEs~\cite{SPENCER}, and therefore are excluded from our study.
+
 An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, in which individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for low order methods.

### 2014-06-23 TACO-paper/coffee.tex
Explain the three kernels

@@
 In addition to a larger number of operations, more complex cases like the Burgers equation are characterized by high register pressure.
-The Helmholtz and Burgers equations, along with the Diffusion problem,
+
+The Helmholtz and Burgers equations are exponents of a large class of problems and, as such, will constitute our benchmark, along with the Diffusion equation.
+We carefully motivate this choice in Section~\ref{sec:real-examples}.
 
@@
 Other optimization levels performed, in general, slightly worse.
-Our code transformations were evaluated in three real-world problems based on the following PDEs:
+
+\subsection{The Helmholtz, Diffusion, and Burgers equations}
+\label{sec:real-examples}
+Our code transformations were evaluated in three real-world problems based on the following PDEs
 \begin{enumerate}
@@
 \end{enumerate}
+The three chosen benchmarks are completely real examples and comprise the core operators in some of the most frequently encountered finite element problems in scientific computing.
+The Helmholtz and Diffusion kernels are archetypal second order elliptic operators.
+They are complete and unsimplified examples of the operators which occur for diffusion and viscosity in fluids, and for imposing pressure in compressible fluids.
+For instance, they are both extensively used in climate modeling.
+Very similar operators, for which the same optimisations are expected to be equally effective, apply to elasticity problems, which are at the core of computational structural mechanics.
+The Burgers kernel is a typical example of a first order hyperbolic conservation law, which occurs in real applications whenever a quantity is transported by a fluid (the momentum itself, in our case).
+We chose this particular kernel since it applies to a vector-valued quantity, while the elliptic operators apply to scalar quantities.
+In this way, the operators we have selected as examples are characteristic of both the second and first order operators which dominate in fluids and solids simulations.
+
 The code was written in UFL (available at~\cite{ufl-code}) and then executed over real unstructured meshes through Firedrake.
@@
 For Advection-Diffusion, the Diffusion equation, which uses the same differential operators as Helmholtz, is considered.
-In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$.
-Burgers is a non-linear problem employing differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (see Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are needed.
+In the Diffusion kernel code, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$.
+Burgers is a non-linear problem employing differential operators different from those of Helmholtz and relying on vector-valued quantities, which has a major impact on the generated assembly code (see Figure~\ref{code:burgers}), where a larger number of basis function arrays ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are generated.
 

### 2014-06-23 TACO-paper/coffee.tex
Add a few comments to splitting section

@@
 
+\usepackage{graphicx}
+\usepackage{caption}
+\usepackage{subfig}
 \usepackage{tabulary}
@@
 % Padding and Data Alignment
-\subsection{Prerequisites for Effective Auto-vectorization: Padding and Data Alignment}
+\subsection{Padding and Data Alignment}
 Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is not a multiple of the vector length $vl$.
@@
 
-\begin{figure}[b]
+\begin{figure}[h]
 \centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect-inline.pdf}}
@@
 
-\begin{figure}[t]
+\begin{figure}[h]
 \centerline{\includegraphics[scale=0.60]{Pictures/vect-restore-inline.pdf}}
@@
 One potential solution to this problem consists of suitably ``splitting'' the computation of the element matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}.
+The transformation can be regarded as a special case of classic loop fission, in which sum's associativity is exploited to distribute the expression across multiple loops.
+To the best of our knowledge, expression splitting is not supported by available compilers.
 
-Splitting an expression has three drawbacks.
+Splitting an expression has, however, several drawbacks.
 Firstly, it increases the number of accesses to $A$ proportionally to the ``split factor'', which is the number of sub-expressions produced.
-Secondly, depending on how the split is executed, it can lead to redundant computation; for example, the times the product $det*W3[i]$ is performed is proportional to the number of sub-expressions, as shown in the code snippet.
+Also, depending on how the split is executed, it can lead to redundant computation; for example, the times the product $det*W3[i]$ is performed is proportional to the number of sub-expressions, as shown in the code snippet.
+Further, it increases loop overhead (e.g.
+branch instructions).
 Finally, it might affect register locality: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
@@
 \label{sec:perf-eval-opvect}
-Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the performance achieved applying \emph{op-vect} on top of \emph{licm-ap} to the Helmoltz and Diffusion kernels on the Sandy Bridge, whereas Figures~\ref{fig:opvect-helmholtz-speedup-phi} and~\ref{fig:opvect-diffusion-speedup-phi} illustrate analogous results for the Xeon Phi.
+Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the performance achieved applying \emph{op-vect} on top of \emph{licm-ap} to the Helmoltz and Diffusion kernels on the Sandy Bridge and Xeon Phi architectures, respectively.
 For each problem instance we report two bars: one indicates the best run-time obtained by auto-tuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved with COFFEE's cost model.
@@
 \begin{figure}[h]
-\begin{center}
-\includegraphics[scale=0.65]{Pictures/helmholtz-normalized-opvect.pdf}
-\end{center}
-\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel on the Sandy Bridge.
+	\centering
+	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.52]{Pictures/helmholtz-normalized-opvect.pdf}}
+	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/helmholtz-normalized-opvect-phi.pdf}}
+	\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel.
 CM refers to where the cost model has been used to determine the sequence of code transformations.}
@@
 \begin{figure}[h]
-\begin{center}
-\includegraphics[scale=0.65]{Pictures/helmholtz-normalized-opvect-phi.pdf}
-\end{center}
-\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel on the Xeon Phi.
-CM refers to where the cost model has been used to determine the sequence of code transformations.}
-\label{fig:opvect-helmholtz-speedup-phi}
-\end{figure}
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[scale=0.65]{Pictures/diffusion-normalized-opvect.pdf}
-\end{center}
-\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel on the Sandy Bridge.
+	\centering
+	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.52]{Pictures/diffusion-normalized-opvect.pdf}}
+	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/diffusion-normalized-opvect-phi.pdf}}
+	\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel.
 CM refers to where the cost model has been used to determine the sequence of code transformations.}
 	\label{fig:opvect-diffusion-speedup}
-\end{figure}
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[scale=0.65]{Pictures/diffusion-normalized-opvect-phi.pdf}
-\end{center}
-\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel on the Xeon Phi.
-CM refers to where the cost model has been used to determine the sequence of code transformations.}
-\label{fig:opvect-diffusion-speedup-phi}
 \end{figure}
@@
 For the Burgers kernel, \emph{split} has a key role on the Sandy Bridge, whereas it has only a minimal impact on the Xeon Phi.
-Figure~\ref{fig:split-burgers-speedup} and~\ref{fig:split-burgers-speedup-phi} show the improvement due to using \emph{split} over the fastest implementation between \emph{licm} and \emph{licm-ap}.
+Figure~\ref{fig:split-burgers-speedup} shows the improvement due to using \emph{split} over the fastest implementation between \emph{licm} and \emph{licm-ap}.
 On the Sandy Bridge, in almost all test cases, a split factor of 1, meaning that the original expression was divided into two parts, ensured close-to-peak perforance.
@@
 \begin{figure}[h]
-\begin{center}
-\includegraphics[scale=0.65]{Pictures/burgers-normalized-split.pdf}
-\end{center}
-\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel on the Sandy Bridge.
-CM refers to the cases in which the cost model has been used to determine the sequence of code transformations.}
+	\centering
+	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.52]{Pictures/burgers-normalized-split.pdf}}
+	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/burgers-normalized-split-phi.pdf}}
+	\caption{Performance improvement obtained applying \emph{split} on top of \emph{licm-ap} to the Burgers kernel.
+CM refers to where the cost model has been used to determine the sequence of code transformations.}
 	\label{fig:split-burgers-speedup}
-\end{figure}
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[scale=0.65]{Pictures/burgers-normalized-split-phi.pdf}
-\end{center}
-\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Burgers kernel on the Xeon Phi.
-CM refers to the cases in which the cost model has been used to determine the sequence of code transformations.}
-\label{fig:split-burgers-speedup-phi}
 \end{figure}
@@
 \begin{center}
-\tbl{Performance comparison between FEniCS (optimizations enabled) and COFFEE on the Sandy Bridge.}{
+\tbl{Performance comparison between FEniCS (optimizations enabled) and COFFEE on the Sandy Bridge}{
+\label{table:comparison-to-FFC-opt}
 \begin{tabulary}{1.0\columnwidth}{C|C|C}
@@
 \end{center}
-\label{table:comparison-to-FFC-opt}
 \end{table}

### 2014-06-23 TACO-paper/coffee.tex
Elaborate padding. Minor fixes

@@
 
-% TODO: could add a sentence here to bridge the two paragraphs
-
 % Padding and Data Alignment
 \subsection{Padding and Data Alignment}
-Auto-vectorization of assembly code computing the element matrix can be less effective if data are not aligned and if the length of the innermost loop is not a multiple of the vector length $vl$.
+The absence of stencils renders the element matrix computation easily auto-vectorizable by a vendor compiler.
+Nevertheless, auto-vectorization is not efficient if data are not aligned and if the length of the innermost loop is not a multiple of the vector length $vl$, especially when the loops are particularly small as in the case of assembly.
+
 Data alignment is enforced in two steps.
-Initially, all arrays are allocated to addresses that are multiples of $vl$.
+Firstly, all arrays are allocated to addresses that are multiples of $vl$.
 Then, two dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$.
 For instance, assume the original size of a basis function array is 3$\times$3 and $vl=4$ (e.g.
-AVX processor, with 256 bit long vector registers and 64-bit double-precision floating-point values).
+AVX processor, with 256 bit long vector registers and 64-bit double-precision floats).
 In this case, a padded version of the array will have size 3$\times$4.
 The compiler is explicitly informed about data alignment using a suitable pragma.
-Padding of all two dimensional arrays involved in the evaluation of the element matrix also allows us to safely round the loop trip count to the nearest multiple of $vl$.
+
+Padding of all two dimensional arrays involved in the evaluation of the element matrix also allows to safely round the loop trip count to the nearest multiple of $vl$.
 This avoids the introduction of a remainder (scalar) loop from the compiler, which would render vectorization less efficient.
+These extra iterations only write to the padded region of the element matrix, and therefore have no side effects on the final result.
 
@@
 it is not subjected to auto-vectorization.
+
 We work around these limitations with source-level loop-invariant code motion.
@@
 Regardless of the specific input problem, by abstracting from the presence of all variables independent of both $j$ and $k$, the element matrix computation is always reducible to this form.
-Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries ($j,k=0,...,4$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$.
+Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries ($j,k=0,...,3$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$.
 Values in a register are shuffled each time a product is performed.
@@
 \begin{figure}[h]
-\centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect-inline.pdf}}
+\centerline{\includegraphics[scale=0.8]{Pictures/vect-by-vect-inline.pdf}}
 \caption{Outer-product vectorization by permuting values in a vector register.}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[scale=0.60]{Pictures/vect-restore-inline.pdf}}
+\centerline{\includegraphics[scale=0.8]{Pictures/vect-restore-inline.pdf}}
 \caption{Restoring the storage layout after \emph{op-vect}.
@@
 \Return{wrapper + ast.from\_ast\_to\_c()}
-\caption{Pseudocode of the COFFEE pipeline.}
+\caption{Pseudocode of the COFFEE architecture.}
 \label{algo:PyOP2Compiler}

### 2014-06-23 TACO-paper/coffee.tex
Minor fixes. About to break licm-licmap table

@@
 
+%\NoCaptionOfAlgo
+%\RestoreCaptionOfAlgo
+
 \begin{algorithm}[t]
@@
 Despite the infinite variety of problem-specific assembly kernels which Firedrake can generate, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.
-These include: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas interchanges involving the $i$ loop require pre-computation of values (e.g.
-the $F$ values in Burgers) and introduction of temporary arrays, as explained in Section~\ref{sec:code-transf}; 3) depending on the problem being solved, the $j$ and $k$ loops could iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$).
+These include: 1) memory accesses along the three loop dimensions are always stride-1; 2) the \texttt{j} and \texttt{k} loops are interchangeable, whereas interchanges involving the $i$ loop require pre-computation of values (e.g.
+the $F$ values in Burgers) and introduction of temporary arrays, as explained in Section~\ref{sec:code-transf}; 3) depending on the problem being solved, the \texttt{j} and \texttt{k} loops could iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either \texttt{i}-\texttt{j} or \texttt{i}-\texttt{k}).
 In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.
@@
 First, they only identify sub-expressions that are invariant with respect to the innermost loop.
-This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted in the loop order \texttt{ijk}.
+This is an issue for sub-expressions depending on \texttt{i}-\texttt{k}, which are not automatically lifted in the loop order \texttt{ijk}.
 Second, the hoisted code is scalar, i.e.
@@
 It is also worth noting that, in some problems (e.g.
-in Helmholtz), invariant sub-expressions along $j$ are identical to those along $k$, and both loops iterate over the same iteration space.
-In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.
+in Helmholtz), invariant sub-expressions along \texttt{j} are identical to those along \texttt{k}, and both loops iterate over the same iteration space.
+In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that \texttt{j} and \texttt{k} loops share the same iteration space.
 
@@
 \caption{Local assembly code for the Helmholtz problem in Figure~\ref{code:helmholtz} after application of padding, data alignment, and \emph{licm}, for an AVX architecture.
-In this example, sub-expressions invariant to $j$ are identical to those invariant to $k$, so they can be precomputed once in the $r$ loop.}
+In this example, sub-expressions invariant to \texttt{j} are identical to those invariant to \texttt{k}, so they can be precomputed once in the $r$ loop.}
 \label{code:helmholtz-licm}
@@
 Tiling at the level of vector registers is an additional feature of COFFEE.
-Based on the observation that the evaluation of the element matrix can be reduced to a summation of outer products along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented.
+Based on the observation that the evaluation of the element matrix can be reduced to a summation of outer products along the \texttt{j} and \texttt{k} dimensions, a model-driven vector-register tiling strategy can be implemented.
 If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, the computation of the element matrix is abstractly expressible as
@@
 \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets.
-Regardless of the specific input problem, by abstracting from the presence of all variables independent of both $j$ and $k$, the element matrix computation is always reducible to this form.
+Regardless of the specific input problem, by abstracting from the presence of all variables independent of both \texttt{j} and \texttt{k}, the element matrix computation is always reducible to this form.
 Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries ($j,k=0,...,3$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$.
@@
 
-%are sufficiently large to prevent register reuse along the innermost dimension $k$.
+%are sufficiently large to prevent register reuse along the innermost dimension \texttt{k}.
 A combination of the two scenarios is also possible.
@@
 ~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
-~~~~~~~~// One $permute\_pd$ intrinsic per $k$-loop $load$\\
+~~~~~~~~// One $permute\_pd$ intrinsic per \texttt{k}-loop $load$\\
 ~~~~~~~~// Compute A[1,2],A[2,1],A[3,4],A[4,3] \\
-~~~~~~~~// One $permute2f128\_pd$ intrinsic per $k$-loop $load$\\
+~~~~~~~~// One $permute2f128\_pd$ intrinsic per \texttt{k}-loop $load$\\
 ~~~~~~~~// ...\\
@@
 Here, we assume the polynomial order is $p=2$, since \emph{op-vect} can not be used when an iteration space dimension is smaller than the vector length.
-The original size of the $j$-$k$ iteration space (i.e.
+The original size of the \texttt{j}-\texttt{k} iteration space (i.e.
 before padding was applied) was 6$\times$6.
@@
 For example, applying loop-invariant code motion to Burgers on a 3D mesh requires 33 temporaries for the \texttt{ijk} loop order.
-This can make hoisting of the invariant loads out of the $k$ loop inefficient on architectures with a relatively low number of registers.
+This can make hoisting of the invariant loads out of the \texttt{k} loop inefficient on architectures with a relatively low number of registers.
 One potential solution to this problem consists of suitably ``splitting'' the computation of the element matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}.
@@
 For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance.
-Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required at every iteration.
+Conceptually, this is motivated by the fact that if the \texttt{i} loop were in an inner position, then a significantly higher number of load instructions would be required at every iteration.
 We tested this hypothesis in manually crafted kernels.
-We found that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than memory, along the $i$ loop.
+We found that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than memory, along the \texttt{i} loop.
 The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension.
@@
 The cost model is shown in Figure~\ref{algo:applyCostModel}.
-It takes into account the number of available logical vector registers, $n\_regs$, and the number of unique variables accessed: $n\_consts$ counts variables independent of both $j$ and $k$ loops and temporary registers, $n\_outer\_arrays$ counts $j$-dependent variables, and $n\_inner\_arrays$ counts $k$-dependent variables, assuming the \texttt{ijk} loop order.
+It takes into account the number of available logical vector registers, \emph{n\_regs}, and the number of unique variables accessed: \emph{n\_consts} counts variables independent of both \texttt{j} and \texttt{k} loops and temporary registers, \emph{n\_outer\_arrays} counts \texttt{j}-dependent variables, and \emph{n\_inner\_arrays} counts \texttt{k}-dependent variables, assuming the \texttt{ijk} loop order.
 These values are used to estimate unroll-and-jam and split factors for \emph{op-vect} and \emph{split}.
@@
 Triangles, tetrahedron, and prisms were tested as element shape.
-For instance, in the case of Helmholtz with $p=1$, the size of the $j$ and $k$ loops for the three element shapes is, respectively, $3$, $4$, and $6$.
+For instance, in the case of Helmholtz with $p=1$, the size of the \texttt{j} and \texttt{k} loops for the three element shapes is, respectively, $3$, $4$, and $6$.
 Moving to bigger shapes has the effect of increasing the number of basis function arrays, since, intuitively, the behavior of the equation has now to be approximated also along a third axis.
-On the other hand, the polynomial order affects only the problem size (the three loops $i$, $j$, and $k$, and, as a consequence, the size of $X$ and $Y$ arrays).
+On the other hand, the polynomial order affects only the problem size (the three loops \texttt{i}, \texttt{j}, and \texttt{k}, and, as a consequence, the size of $X$ and $Y$ arrays).
 A range of polynomial orders from $p=1$ to $p=4$ were tested; higher polynomial orders are excluded from the study because of current Firedrake limitations.
@@
 In the following, results for the loop order \texttt{ijk} are shown.
-For the considerations exposed in Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner loop over $i$ caused slow downs; also, interchanging $j$ and $k$ loops while keeping $i$ as outermost loop did not provide any benefits.
+For the considerations exposed in Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner loop over \texttt{i} caused slow downs; also, interchanging \texttt{j} and \texttt{k} loops while keeping \texttt{i} as outermost loop did not provide any benefits.
 
@@
 Using the Intel Architecture Code Analyzer tool~\cite{IACA} on the Sandy Bridge, we confirmed that this is a consequence of increased register re-use.
-In Helmholtz $p=4$, for example, the tool showed that when using \emph{op-vect} the number of clock cycles to execute one iteration of the $j$ loop decreases by roughly 17$\%$, and that this is a result of the relieved pressure on both of the data (cache) ports available in the core.
+In Helmholtz $p=4$, for example, the tool showed that when using \emph{op-vect} the number of clock cycles to execute one iteration of the \texttt{j} loop decreases by roughly 17$\%$, and that this is a result of the relieved pressure on both of the data (cache) ports available in the core.
 

### 2014-06-23 TACO-paper/coffee.tex
Still in the process of writing licm-licmap results

@@
 \end{table*}
+Table~\ref{table:perf-results-licm} illustrates the performance improvement obtained when \emph{licm} is used.
+The gain grows with the computational cost of the kernels: bigger loop nests usually benefit from the reduction in redundant computation, even though extra memory for the temporary arrays is required.
+Inspection of assembly code generated by \texttt{icc} confirmed all limitations described in Section~\ref{sec:code-transf-licm}: in the original version of the code, only sub-expressions invariant to the inner loop were hoisted, and their execution was not vectorized, which motivates the generally good speed ups achieved by \emph{licm}.
 
-Table~\ref{table:perf-results-licm} illustrates the performance improvement obtained on the Sandy Bridge and Xeon Phi machines when \emph{licm}, data alignment, and padding are used.
-We distinguish between \emph{licm} and \emph{licm-ap}; the latter makes use of padding and data alignment.
-Inspection of assembly code generated by \texttt{icc} confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant to the inner loop are hoisted and their execution is not vectorized.
+\subsection{Impact of Padding and Data Alignment}
+Table~\ref{table:perf-results-licmap} shows the cumulative impact of \emph{licm}, data alignment, and padding over the original code.
+In the following, this version of the code is referred to as \emph{licm-ap}.
 Padding and data alignment enhance, in general, the quality of auto-vectorization.
-Sometimes the run-time of \emph{licm-ap} is similar to that of \emph{licm} because the element matrix size is already a multiple of the vector length, so no scalar reminder loop has to be introduced.
-Occasionally \emph{licm-ap} is slower than \emph{licm} (e.g.
+Occasionally the run-time of \emph{licm-ap} is close to that of \emph{licm}, since the element matrix size is already a multiple of the vector length, and no penalty for a scalar reminder loop has to be paid.
+Rarely \emph{licm-ap} is slower than \emph{licm} (e.g.
 in Burgers $p=3$ on the Sandy Bridge).
-One possible explanation is that the large number of aligned temporaries introduced by \emph{licm} induce cache associativity conflicts.
+One possible explanation is that the number of aligned temporaries introduced by \emph{licm} is so large to induce cache associativity conflicts.
 

### 2014-06-23 TACO-paper/coffee.tex
Fix layout and style and names of algos/listings

@@
 \usepackage{tabulary}
+
+\usepackage{alltt}
+\renewcommand{\ttdefault}{txtt}
+
+\usepackage{listings}
+\lstset{language=C}
+
 \usepackage[cmex10]{amsmath}
@@
 
-\begin{algorithm}[t]
-\footnotesize
-\KwIn{element matrix (2D array, initialized to 0),\\~~~~~~~~~element coordinates (array),\\~~~~~~~~~coefficient fields (array, e.g. velocity)}
-\KwOut{element matrix (2D array)}
-~~- Compute Jacobian from coordinates \\
-~~- Define basis functions \\
-~~- Compute element matrix in an affine loop nest
-\caption{General structure of a local assembly kernel generated by Firedrake.}
-\label{code:general-structure}
-\end{algorithm}
-
 %\NoCaptionOfAlgo
@@
 
+\begin{figure}
+\begin{alltt}
+\footnotesize
+\textbf{Input:} element matrix (2D array, initialized to 0), 
+       element coordinates (array), 
+       coefficient fields (array, e.g. velocity) 
+\textbf{Output:} element matrix (2D array)
+- Compute Jacobian from coordinates 
+- Define basis functions 
+- Compute element matrix in an affine loop nest
+\end{alltt}
+\caption{Structure of a local assembly kernel}
+\label{code:general-structure}
+\end{figure}
+
 \begin{algorithm}[t]
+\SetAlgorithmName{LISTING}{}
 \footnotesize
@@
 \begin{algorithm}[t]
+\SetAlgorithmName{LISTING}{}
 \footnotesize
@@
 In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products involving basis functions.
-An example is provided in Figure~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$.
-In other scenarios, for instance when solving the Burgers equation, whose assembly code is given in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
+An example is provided in Listing~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$.
+In other scenarios, for instance when solving the Burgers equation, whose assembly code is given in Listing~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
 This may also require the evaluation of constants in outer loops (called $F$ in the code) to act as scaling factors of arrays; trip counts can be larger (proportionally to the order of the method); and arrays may be block-sparse.
@@
 It is not always the same set of optimizations that bring performance closest to the machine peak.
-For example, the Burgers problem in Figure~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Figure~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces.
+For example, the Burgers problem in Listing~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Listing~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces.
 Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger.
@@
 \label{sec:code-transf-licm}
-From the inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions which only depend on two iteration variables.
+From the inspection of the codes in Listings~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions which only depend on two iteration variables.
 Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion.
@@
 
-Figure~\ref{code:helmholtz-licm} shows the Helmholtz assembly code after the application of loop-invariant code motion, padding, and data alignment.
+Listing~\ref{code:helmholtz-licm} shows the Helmholtz assembly code after the application of loop-invariant code motion, padding, and data alignment.
 
 \begin{algorithm}[t]
+\SetAlgorithmName{LISTING}{}
 \footnotesize
@@
 $\rbrace$
-\caption{Local assembly code for the Helmholtz problem in Figure~\ref{code:helmholtz} after application of padding, data alignment, and \emph{licm}, for an AVX architecture.
+\caption{Local assembly code for the Helmholtz problem in Listing~\ref{code:helmholtz} after application of padding, data alignment, and \emph{licm}, for an AVX architecture.
 In this example, sub-expressions invariant to \texttt{j} are identical to those invariant to \texttt{k}, so they can be precomputed once in the $r$ loop.}
@@
 Based on the observation that the evaluation of the element matrix can be reduced to a summation of outer products along the \texttt{j} and \texttt{k} dimensions, a model-driven vector-register tiling strategy can be implemented.
-If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, the computation of the element matrix is abstractly expressible as
+If we consider the code snippet in Listing~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, the computation of the element matrix is abstractly expressible as
 \begin{equation}
@@
 ~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
-~~~~// Do loop-invariant code motion \\
+~~~~// Do generalized loop-invariant code motion \\
 ~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
@@
 ~~~~~~~~// $load$ and $set$ intrinsics \\
-~~~~~~~~// Compute A[1,1],A[2,2],A[3,3],A[4,4] \\
+~~~~~~~~// Compute A[0,0], A[1,1], A[2,2], A[3,3] \\
 ~~~~~~~~// One $permute\_pd$ intrinsic per \texttt{k}-loop $load$\\
-~~~~~~~~// Compute A[1,2],A[2,1],A[3,4],A[4,3] \\
+~~~~~~~~// Compute A[0,1], A[1,0], A[2,3], A[3,2] \\
 ~~~~~~~~// One $permute2f128\_pd$ intrinsic per \texttt{k}-loop $load$\\
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake for the Helmholtz problem after application of \emph{op-vect} on top of the optimizations shown in Figure~\ref{code:helmholtz-licm}.
+\caption{Local assembly code generated by Firedrake for the Helmholtz problem after application of \emph{op-vect} on top of the optimizations shown in Listing~\ref{code:helmholtz-licm}.
 Here, we assume the polynomial order is $p=2$, since \emph{op-vect} can not be used when an iteration space dimension is smaller than the vector length.
@@
 \label{sec:expr-split}
-In complex kernels, like Burgers in Figure~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
+In complex kernels, like Burgers in Listing~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
 If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, poor register reuse is likely to be obtained.
@@
 This can make hoisting of the invariant loads out of the \texttt{k} loop inefficient on architectures with a relatively low number of registers.
-One potential solution to this problem consists of suitably ``splitting'' the computation of the element matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Figure~\ref{code:helmholtz-split}.
+One potential solution to this problem consists of suitably ``splitting'' the computation of the element matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Listing~\ref{code:helmholtz-split}.
 The transformation can be regarded as a special case of classic loop fission, in which sum's associativity is exploited to distribute the expression across multiple loops.
@@
 
-\begin{algorithm}[t]
+\begin{algorithm}
+\SetAlgorithmName{LISTING}{}
 \footnotesize
 \KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
-~~~~// Same code as in Figure~\ref{code:helmholtz-licm} up to line 15 \\
+~~~~// Same code as in Listing~\ref{code:helmholtz-licm} up to line 15 \\
 ~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake for the Helmholtz problem in which \emph{split} has been applied on top of the optimizations shown in Figure~\ref{code:helmholtz-licm}.
+\caption{Local assembly code generated by Firedrake for the Helmholtz problem in which \emph{split} has been applied on top of the optimizations shown in Listing~\ref{code:helmholtz-licm}.
 In this example, the split factor is 2.}
@@
 
-\begin{table}[h]
+\begin{table}[t]
 \begin{center}
@@
 
-\begin{algorithm}[t]
+%\rule[0.2ex]{16cm}{0.5pt}
+
+\begin{figure}[t]
+\begin{alltt}
 \footnotesize
-  \textbf{The COFFEE Compiler}\\
-  \KwIn{ast, wrapper, isa}
-  \KwOut{code}
-// Analyze ast and build optimization plan \\
-it\_space = analyze(ast) \\
-\If{\KwSty{not} it\_space}{
-  ast.apply\_inter\_kernel\_vectorization(wrapper, isa)\\
-  \Return{wrapper + ast.from\_ast\_to\_c()}
-}
-plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs) \\
-// Optimize ast based on plan \\
-ast.licm() \\
-ast.padding() \\
-ast.data\_align() \\
-\If{plan.permute}{
+\textbf{Input:} ast: abstract syntax tree of a local assembly kernel produced by FFC
+       wrapper: the function within which the kernel is invoked
+	       isa: represents the SIMD Instruction Set Architecture
+\textbf{Output:} assembly code (C string)
+\textbf{// Analyze ast and build optimization plan }
+it\_space = analyze(ast) 
+\textbf{if not} it\_space
+  ast.apply\_inter\_kernel\_vectorization(wrapper, isa)
+  \textbf{return} wrapper + ast.from\_ast\_to\_c()
+plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs) 
+\textbf{// Optimize ast based on plan}
+ast.licm()
+ast.padding() 
+ast.align\_data()
+\textbf{if} plan.permute
   ast.permute\_assembly\_loops()
-}
-\If{plan.sz\_split}{
-  ast.split(plan.sz\_split)
-}
-\If{plan.uaj\_factor}{
-  uaj = MIN(plan.uaj\_factor, $\lfloor$it\_space.j.size/isa.vf$\rfloor$) \\
+\textbf{if} plan.split\_factor
+  ast.split(plan.split\_factor)
+\textbf{if} plan.uaj\_factor
+  uaj = MIN(plan.uaj\_factor, it\_space.j.size/isa.vf) 
   ast.op\_vect(uaj)
-}
-\Return{wrapper + ast.from\_ast\_to\_c()}
-\caption{Pseudocode of the COFFEE architecture.}
+\textbf{return} wrapper + ast.from\_ast\_to\_c()
+\end{alltt}
+
+\caption{Outline of the COFFEE architecture.}
 \label{algo:PyOP2Compiler}
-\end{algorithm}
+\end{figure}
 
 Firedrake users employ the Unified Form Language to write problems in a notation resembling mathematical equations.
-At run-time, the high-level specification is translated by a modified version of the FEniCS Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
+At run-time, the high-level specification is translated by a modified version of the FEniCS Form Compiler~\cite{FFC-Compiler} (FFC) into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
 ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}.
@@
 Because of the large number of parametric transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem.
-Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of ASTs occurs at run-time as soon as problem-specific data are available.
+Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of ASTs occurs at run-time as soon as problem-secific data are available.
 Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.
@@
 
+\begin{figure}
+\begin{alltt}
+\footnotesize
+\textbf{Input:} n\_outer\_arrays, n\_inner\_arrays, n\_consts, n\_regs
+\textbf{Output:} uaj\_factor, split\_factor
+n\_outer\_regs = n\_regs / 2 
+split\_factor = 0 
+\textbf{// Compute spltting factor }
+\textbf{while} n\_outer\_arrays > n\_outer\_regs
+  n\_outer\_arrays = n\_outer\_arrays / 2 
+  split\_factor = split\_factor + 1
+\textbf{// Compute unroll-and-jam factor for \emph{op-vect}}
+n\_regs\_avail = n\_regs - (n\_outer\_arrays + n\_consts)
+uaj\_factor = n\_reg\_avail / n\_inner\_arrays
+\textbf{// Estimate the benefit of permuting loops}
+permute = n\_outer\_arrays > n\_inner\_arrays
+\textbf{return} <permute, split\_factor, uaj\_factor>
+\end{alltt}
+\caption{The cost model is employed by the compiler to estimate the most suitable unroll-and-jam (when \emph{op-vect} is used) and split factors, avoiding the overhead of auto-tuning.}
+\label{algo:applyCostModel}
+\end{figure}
+
 The cost model is shown in Figure~\ref{algo:applyCostModel}.
@@
 Finally, the profitability of loop interchange is evaluated (line 12-16).
-
-\begin{algorithm}[t]
-  \textbf{Cost Model}\\
-  \KwIn{n\_outer\_arrays, n\_inner\_arrays, n\_consts, n\_regs}
-  \KwOut{uaj\_factor, split\_factor}
-
-n\_outer\_regs = n\_regs / 2 \\
-split\_factor = 0 \\
-// Compute spltting factor \\
-\While{n\_outer\_arrays $>$ n\_outer\_regs}{
-  n\_outer\_arrays = n\_outer\_arrays / 2 \\
-  split\_factor = split\_factor + 1
-}
-
-// Compute unroll-and-jam factor for \emph{op-vect} \\
-n\_regs\_avail = n\_regs - (n\_outer\_arrays + n\_consts)\\
-uaj\_factor = $\lceil$n\_reg\_avail / n\_inner\_arrays$\rceil$ \\
-\eIf{n\_outer\_arrays $>$ n\_inner\_arrays}{
-  permute = \KwSty{True}
-}{
-  permute = \KwSty{False}
-}
-\Return{$<$permute, split\_factor, uaj\_factor$>$}
-\caption{The cost model is employed by the compiler to estimate the most suitable unroll-and-jam (when \emph{op-vect} is used) and split factors, avoiding the overhead of auto-tuning.}
-\label{algo:applyCostModel}
-\end{algorithm}
 
@@
 The code was written in UFL (available at~\cite{ufl-code}) and then executed over real unstructured meshes through Firedrake.
-The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}.
+The Helmholtz code has already been shown in Listing~\ref{code:helmholtz}.
 For Advection-Diffusion, the Diffusion equation, which uses the same differential operators as Helmholtz, is considered.
 In the Diffusion kernel code, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$.
-Burgers is a non-linear problem employing differential operators different from those of Helmholtz and relying on vector-valued quantities, which has a major impact on the generated assembly code (see Figure~\ref{code:burgers}), where a larger number of basis function arrays ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are generated.
+Burgers is a non-linear problem employing differential operators different from those of Helmholtz and relying on vector-valued quantities, which has a major impact on the generated assembly code (see Listing~\ref{code:burgers}), where a larger number of basis function arrays ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are generated.
 
@@
 Burgers' slow downs occur in presence of a small iteration space (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$).
-The figure shown is for the worst slow down, which was obtained with a triangular mesh and $p = 1$.
-This is a result of removing zero-valued entries in FEniCS' basis functions: some operations are avoided, but indirection arrays prevent auto-vectorization, which significantly impacts performance as soon as the element matrix becomes bigger than 12$\times$12.
+The result shown represents the worst slow down, which was obtained with a triangular mesh and $p = 1$.
+This is a result of removing zero-valued entries in FEniCS' basis function arrays: some operations are avoided, but indirection arrays prevent auto-vectorization, which significantly impacts performance as soon as the element matrix becomes bigger than 12$\times$12.
 However, with the forthcoming zero-removal optimization in COFFEE, we expect to outperform FEniCS in all problems.

### 2014-06-24 TACO-paper/coffee.tex
Now really about to break licm-licmap tables
Broke tables; re-wrote impact of licm, ap, op-vect

@@
 
+\begin{table}[b]
+\tbl{Type and variable names used in the various listings to identify local assembly objects.}{
+\begin{tabulary}{1.0\columnwidth}{C|C|C}
+\hline
+Object name & Type & Variable name(s) \\\hline
+Determinant of the Jacobian matrix & double & det  \\ 
+Inverse of the Jacobian matrix & double & K1, K2, ... \\ 
+Coordinates & double** & coords\\ 
+Fields & double** & w \\ 
+Numerical integration weights & double[] & W \\ 
+Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ 
+Element matrix & double[][] & A\\ \hline
+\end{tabulary}
+}
+\label{table:map-name-letters}
+\end{table}
+
+Local assembly is the computation of contributions of a specific cell in the discretized domain to the PDE solution.
+The process consists of numerically evaluating problem-specific integrals to produce a matrix and a vector (\cite{quadrature1}, \cite{fluidity_manual_v4}), whose sizes depend on the order of the method.
+This operation is applied to all cells in the discretized domain.
+In this work we focus on local matrices, or ``element matrices'', which are more costly to compute than element vectors.
+
+Given a finite element description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl} (UFL), Firedrake generates a C-code kernel implementing assembly using a numerical integration algorithm.
+The same kernel can be applied to any elements: this descends from the mathematical formulation of assembly, which states that the evaluation of a local matrix can be reduced to integrating on a fixed ``reference'' element - a special element that does not belong to the domain - after a suitable change of coordinates.
+Firedrake triggers the compilation of an assembly kernel using an available vendor compiler, and manages its (parallel) execution over all elements in the mesh.
+As already stated, the subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.
+
+%The code transformations described in this work are also generalizable to non-Firedrake assembly kernels, provided that the same numerical integration algorithm is employed.
+
+The structure of a local assembly kernel is shown in Figure~\ref{code:general-structure}.
+The inputs are a zero-initialized two dimensional array used to store the element matrix, the element's coordinates in the discretized domain, and coefficient fields, for instance indicating the values of velocity or pressure in the element.
+The output is the evaluated element matrix.
+The kernel body can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse: required for the aforementioned change of coordinates, from the reference element to the one being computed; 2) definition of basis functions: used, intuitively, to interpolate the contribution to the PDE solution over the element.
+The choice of the basis functions is expressed in UFL directly by users.
+In the generated code, they are represented as global read-only two dimensional arrays (i.e.
+using \texttt{static const} in C) of double precision floats; 3) evaluation of the element matrix in an affine loop nest, in which the integration is performed.
+Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel objects.
+
+%JUNK SENTENCES: Since only the values assumed at the integration points are needed to compute the element matrix, a basis function is implemented as a two dimensional array of double precision floats.
+The same basis functions can be used to evaluate all element matrices, provided that a suitable change of coordinates from a ``reference'' element to a real one is performed.
+Therefore, basis functions are declared in the kernel body as global read-only arrays (i.e.
+using \texttt{static const} in C), rather than passed as function parameters.
+Basis functions are chosen at the UFL layer based on the problem being solved, so they are actually an input to the code generator.
+
 \begin{algorithm}[t]
@@
 \end{algorithm}
-
-Local assembly is the computation of contributions of a specific cell in the discretized domain to the PDE solution.
-The process consists of numerically evaluating problem-specific integrals to produce a matrix and a vector (\cite{quadrature1}, \cite{fluidity_manual_v4}), whose sizes depend on the order of the method.
-This operation is applied to all cells in the discretized domain.
-In this work we focus on local matrices, or ``element matrices'', which are more costly to compute than element vectors.
-
-Given a finite element description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl} (UFL), Firedrake generates a C-code kernel implementing assembly using a numerical integration algorithm.
-The same kernel can be applied to any elements: this descends from the mathematical formulation of assembly, which states that the evaluation of a local matrix can be reduced to integrating on a fixed ``reference'' element - a special element that does not belong to the domain - after a suitable change of coordinates.
-Firedrake triggers the compilation of an assembly kernel using an available vendor compiler, and manages its (parallel) execution over all elements in the mesh.
-As already stated, the subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.
-
-%The code transformations described in this work are also generalizable to non-Firedrake assembly kernels, provided that the same numerical integration algorithm is employed.
-
-The structure of a local assembly kernel is shown in Figure~\ref{code:general-structure}.
-The inputs are a zero-initialized two dimensional array used to store the element matrix, the element's coordinates in the discretized domain, and coefficient fields, for instance indicating the values of velocity or pressure in the element.
-The output is the evaluated element matrix.
-The kernel body can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse: required for the aforementioned change of coordinates, from the reference element to the one being computed; 2) definition of basis functions: used, intuitively, to interpolate the contribution to the PDE solution over the element.
-The choice of the basis functions is expressed in UFL directly by users.
-In the generated code, they are represented as global read-only two dimensional arrays (i.e.
-using \texttt{static const} in C) of double precision floats; 3) evaluation of the element matrix in an affine loop nest, in which the integration is performed.
-Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel objects.
-
-%JUNK SENTENCES: Since only the values assumed at the integration points are needed to compute the element matrix, a basis function is implemented as a two dimensional array of double precision floats.
-The same basis functions can be used to evaluate all element matrices, provided that a suitable change of coordinates from a ``reference'' element to a real one is performed.
-Therefore, basis functions are declared in the kernel body as global read-only arrays (i.e.
-using \texttt{static const} in C), rather than passed as function parameters.
-Basis functions are chosen at the UFL layer based on the problem being solved, so they are actually an input to the code generator.
-
-\begin{table}[h]
-\begin{center}
-\tbl{Type and variable names used in the various code snippets to identify local assembly objects.}{
-\begin{tabulary}{1.0\columnwidth}{C|C|C}
-\hline
-Object name & Type & Variable name(s) \\\hline
-Determinant of the Jacobian matrix & double & det  \\ 
-Inverse of the Jacobian matrix & double & K1, K2, ... \\ 
-Coordinates & double** & coords\\ 
-Fields & double** & w \\ 
-Numerical integration weights & double[] & W \\ 
-Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ 
-Element matrix & double[][] & A\\ \hline
-\end{tabulary}
-}
-\end{center}
-\label{table:map-name-letters}
-\end{table}
 
@@
 \begin{table}[t]
-\begin{center}
 \tbl{Overview of code transformations for Firedrake-generated assembly kernels.}{
@@
 }
-\end{center}
 \label{table:code-transformations}
@@
 We therefore leave the backend compiler in charge of selecting the unroll factor.
-This choice also simplifies COFFEE's cost model.
+This choice also simplifies the COFFEE's cost model.
 The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code seems to prevent the Intel compiler from applying this optimization, even if specific pragmas are added.
@@
 
-In the following, results for the loop order \texttt{ijk} are shown.
-For the considerations exposed in Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner loop over \texttt{i} caused slow downs; also, interchanging \texttt{j} and \texttt{k} loops while keeping \texttt{i} as outermost loop did not provide any benefits.
+\subsection{Loop permutation}
+In the following, only results for the loop order \texttt{ijk} are shown.
+For the considerations exposed in Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner loop along \texttt{i} caused slow downs; also, interchanging \texttt{j} and \texttt{k} loops while keeping \texttt{i} as outermost loop did not provide any benefits.
 
 \subsection{Impact of Generalized Loop-invariant Code Motion}
+\label{licm-impact}
 
-\newcommand{\massmatrixtwodresultsnorms}{
-\begin{tabulary}{1.0\textwidth}{cccccc|cccc|cccc|cccc}
-\cline{3-18}
-& & \multicolumn{8}{c}{\texttt{Sandy Bridge}} & \multicolumn{8}{c}{\texttt{Xeon Phi}}\\
-\cline{3-18}
-& & \multicolumn{4}{c}{\texttt{licm}} & \multicolumn{4}{c}{\texttt{licm-ap}} & \multicolumn{4}{c}{\texttt{licm}} & \multicolumn{4}{c}{\texttt{licm-ap}} \\
-\cline{1-18}
-& & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} \\[0.1cm]
-\texttt{Helmholtz} & \texttt{triangle} & 1.05 & 1.46 & 1.68 & 1.67 & 1.32 & 1.88 & 2.87 & 4.13 & 1,49 & 1,06 & 1,05 & 1,17 & 1,50 & 2,41 & 1,30 & 1,96 \\
-\texttt{Helmholtz} & \texttt{tetrahedron} & 1.36 & 2.10 & 2.64 & 2.27 & 1.35 & 3.32 & 2.66 & 3.27 & 1,28 & 1,29 & 2,05 & 1,73 & 1,41 & 1,50 & 2,79 & 2,81 \\
-\texttt{Helmholtz} & \texttt{prism} & 2.16 & 2.28 & 2.45 & 2.06 & 2.63 & 2.74 & 2.43 & 2.75 & 1,04 & 2,26 & 1,93 & 1,64 & 2,38 & 2,47 & 2,15 & 1,71\\[0.1cm]
+\newcommand{\licmresultsnorms}{
+\begin{tabulary}{1.0\textwidth}{cccccc|cccc}
+\cline{3-10}
+& & \multicolumn{4}{c}{\texttt{Sandy Bridge}} & \multicolumn{4}{c}{\texttt{Xeon Phi}} \\
+\cline{1-10}
+\texttt{problem} & \texttt{shape} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} \\
+\texttt{Helmholtz} & \texttt{triangle} & 1.05 & 1.46 & 1.68 & 1.67 & 1,49 & 1,06 & 1,05 & 1,17 \\
+\texttt{Helmholtz} & \texttt{tetrahedron} & 1.36 & 2.10 & 2.64 & 2.27 & 1,28 & 1,29 & 2,05 & 1,73 \\
+\texttt{Helmholtz} & \texttt{prism} & 2.16 & 2.28 & 2.45 & 2.06 & 1,04 & 2,26 & 1,93 & 1,64 \\[0.1cm]
 %\cline{1-18}
-\texttt{Diffusion} & \texttt{triangle} & 1.09 & 1.68 & 1.97 & 1.64 & 1.38 & 1.99 & 3.07 & 4.28 & 1,07 & 1,06 & 1,18 & 1,16 & 1,08 & 1,88 & 1,20 & 1,97\\
-\texttt{Diffusion} & \texttt{tetrahedron} & 1.30 & 2.20 & 3.12 & 2.60 & 1.41 & 3.70 & 3.18 & 3.82 & 1,00 & 1,38 & 2,02 & 1,74 & 1,05 & 1,51 & 2,76 & 3,00\\
-\texttt{Diffusion} & \texttt{prism} & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69 & 1,11 & 2,16 & 1,85 & 2,83 & 2,41 & 2,52 & 2,05 & 2,48\\[0.1cm]
+\texttt{Diffusion} & \texttt{triangle} & 1.09 & 1.68 & 1.97 & 1.64 & 1,07 & 1,06 & 1,18 & 1,16 \\
+\texttt{Diffusion} & \texttt{tetrahedron} & 1.30 & 2.20 & 3.12 & 2.60 & 1,00 & 1,38 & 2,02 & 1,74\\
+\texttt{Diffusion} & \texttt{prism} & 2.15 & 1.82 & 2.71 & 2.32 & 1,11 & 2,16 & 1,85 & 2,83\\[0.1cm]
 %\cline{1-18}
-\texttt{Burgers} & \texttt{triangle} & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 & 1,21 & 1,42 & 2,34 & 2,97 & 2,84 & 2,26 & 3,96 & 4,27 \\
-\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.24 & 1.69 & 1.59 & 1.61 & 2.10 & 1.60 & 1.78 & 1,01 & 2,55 & 0,98 & 1,21 & 1,48 & 3,83 & 1,55 & 1,29 \\
-\texttt{Burgers} & \texttt{prism} & 2.11 & 2.20 & 1.66 & 1.32 & 2.19 & 2.32 & 1.64 & 1.42 & 1,39 & 1,56 & 1,18 & 1,04 & 2,18 & 2,82 & 1,24 & 1,25 \\
-\cline{1-18}
+\texttt{Burgers} & \texttt{triangle} & 1.53 & 1.81 & 2.68 & 2.46 & 1,21 & 1,42 & 2,34 & 2,97  \\
+\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.24 & 1.69 & 1.59 & 1,01 & 2,55 & 0,98 & 1,21  \\
+\texttt{Burgers} & \texttt{prism} & 2.11 & 2.20 & 1.66 & 1.32 & 1,39 & 1,56 & 1,18 & 1,04 \\
+\cline{1-10}
 \end{tabulary}
@@
 \begin{table*}[t]
-\tbl{Performance improvement due to generalized loop-invariant code motion (\emph{licm} columns) on the Helmholtz, Diffusion and Burgers problems, varying the element shape (triangle, tetrahedron, prism) and the polynomial order ($p \in [1, 4]$), using the Lagrange element family, over the original, non-optimized code.
-The \emph{licm-ap} columns illustrate the effect of combining \emph{licm} with data alignment and padding.
-Results are shown for both the Sandy Bridge and the Xeon Phi architectures.}{
+\tbl{Performance improvement due to generalized loop-invariant code motion, for different element shapes (triangle, tetrahedron, prism) and polynomial orders ($p \in [1, 4]$), over the original non-optimized code, for the Helmholtz, Diffusion and Burgers problems.}{
 \scriptsize
-\massmatrixtwodresultsnorms
+\licmresultsnorms
 }
@@
 \end{table*}
-Table~\ref{table:perf-results-licm} illustrates the performance improvement obtained when \emph{licm} is used.
-The gain grows with the computational cost of the kernels: bigger loop nests usually benefit from the reduction in redundant computation, even though extra memory for the temporary arrays is required.
-Inspection of assembly code generated by \texttt{icc} confirmed all limitations described in Section~\ref{sec:code-transf-licm}: in the original version of the code, only sub-expressions invariant to the inner loop were hoisted, and their execution was not vectorized, which motivates the generally good speed ups achieved by \emph{licm}.
+Table~\ref{table:perf-results-licm} illustrates the performance improvement obtained when \emph{licm} is applied.
+In general, the speed ups are notable.
+The main reasons were anticipated in Section~\ref{sec:code-transf-licm}: in the original code, 1) sub-expressions invariant to outer loops are not automatically hoisted, while 2) sub-expressions invariant to the innermost loop are hoisted, but their execution is not auto-vectorized.
+These observations come from inspection of assembly code generated by the compiler.
+
+The gain tends to grow with the computational cost of the kernels: bigger loop nests (i.e.
+larger element shapes and polynomial orders) usually benefit from the reduction in redundant computation, even though extra memory for the temporary arrays is required.
+Some discrepancies to this trend are due to a less effective auto-vectorization.
+For instance, on the Sandy Bridge, the improvement at $p=3$ is larger than that at $p=4$ because, in the latter case, the size of the innermost loop is not a multiple of the vector length, and a reminder scalar loop is introduced at compile time.
+Since the loop nest is small, the cost of executing the extra scalar iterations can have a significant impact.
+The reminder loop overhead is more pronounced on the Xeon Phi, where the vector length is twice as big, which leads to proportionally larger scalar reminder loops.
 
 \subsection{Impact of Padding and Data Alignment}
+
+\newcommand{\licmapresultsnorms}{
+\begin{tabulary}{1.0\textwidth}{cccccc|cccc}
+\cline{3-10}
+& & \multicolumn{4}{c}{\texttt{Sandy Bridge}} & \multicolumn{4}{c}{\texttt{Xeon Phi}} \\
+\cline{1-10}
+\texttt{problem} & \texttt{shape} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} \\[0.1cm]
+\texttt{Helmholtz} & \texttt{triangle} & 1.32 & 1.88 & 2.87 & 4.13 & 1,50 & 2,41 & 1,30 & 1,96 \\
+\texttt{Helmholtz} & \texttt{tetrahedron} & 1.35 & 3.32 & 2.66 & 3.27 & 1,41 & 1,50 & 2,79 & 2,81 \\
+\texttt{Helmholtz} & \texttt{prism} & 2.63 & 2.74 & 2.43 & 2.75 & 2,38 & 2,47 & 2,15 & 1,71\\[0.1cm]
+%\cline{1-18}
+\texttt{Diffusion} & \texttt{triangle} & 1.38 & 1.99 & 3.07 & 4.28 & 1,08 & 1,88 & 1,20 & 1,97\\
+\texttt{Diffusion} & \texttt{tetrahedron} & 1.41 & 3.70 & 3.18 & 3.82 & 1,05 & 1,51 & 2,76 & 3,00\\
+\texttt{Diffusion} & \texttt{prism} & 2.55 & 3.13 & 2.73 & 2.69 & 2,41 & 2,52 & 2,05 & 2,48\\[0.1cm]
+%\cline{1-18}
+\texttt{Burgers} & \texttt{triangle} & 1.56 & 2.28 & 2.61 & 2.77 & 2,84 & 2,26 & 3,96 & 4,27 \\
+\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.10 & 1.60 & 1.78 & 1,48 & 3,83 & 1,55 & 1,29 \\
+\texttt{Burgers} & \texttt{prism} & 2.19 & 2.32 & 1.64 & 1.42 & 2,18 & 2,82 & 1,24 & 1,25 \\
+\cline{1-10}
+\end{tabulary}
+}
+
+\begin{table*}[t]
+\tbl{Performance improvement due to generalized loop-invariant code motion, data alignment, and padding, for different element shapes (triangle, tetrahedron, prism) and polynomial orders ($p \in [1, 4]$), over the original non-optimized code, for the Helmholtz, Diffusion and Burgers problems.}{
+\scriptsize
+\licmapresultsnorms
+}
+\label{table:perf-results-licmap}
+\end{table*}
+
 Table~\ref{table:perf-results-licmap} shows the cumulative impact of \emph{licm}, data alignment, and padding over the original code.
 In the following, this version of the code is referred to as \emph{licm-ap}.
-Padding and data alignment enhance, in general, the quality of auto-vectorization.
-Occasionally the run-time of \emph{licm-ap} is close to that of \emph{licm}, since the element matrix size is already a multiple of the vector length, and no penalty for a scalar reminder loop has to be paid.
+Padding, which avoids the introduction of a reminder loop as described in Section~\ref{licm-impact}, as well as data alignment, enhance the quality of auto-vectorization.
+Occasionally the run-time of \emph{licm-ap} is close to that of \emph{licm}, since the non-padded element matrix size is already a multiple of the vector length.
 Rarely \emph{licm-ap} is slower than \emph{licm} (e.g.
@@
 \label{sec:perf-eval-opvect}
-Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} show the performance achieved applying \emph{op-vect} on top of \emph{licm-ap} to the Helmoltz and Diffusion kernels on the Sandy Bridge and Xeon Phi architectures, respectively.
-For each problem instance we report two bars: one indicates the best run-time obtained by auto-tuning the unroll/unroll-and-jam factors, whereas the other shows the result retrieved with COFFEE's cost model.
-In general, there is no substantial difference between the two.
-This is chiefly because these kernels fit the L1 cache, so, within a certain degree of confidence, it is possible to predict the fastest implementation based on register allocation.
+In this section, we evaluate the impact of vector-register tiling.
+We compare two versions: the baseline, \emph{licm-ap}; and vector-register tiling on top of \emph{licm-ap}, which in the following is referred to simply as \emph{op-vect}.
 
-The rationale behind these results is that, for smaller configurations, vector-register tiling is marginally helpful, while its impact is the dominant factor for larger problems.
-This is because in the considered kernels the number of accessed arrays along the innermost loop dimension is rather small (between 2 and 4), so extensive unrolling is often enough to maximize register re-use when the loops are relatively short.
-On the other hand, as the iteration space becomes bigger, vector-register tiling leads to improvements up to 1.4$\times$ on the Sandy Bridge (Diffusion, prismatic mesh, $p=4$) and up to 1.4$\times$ on the Xeon Phi (Helmholtz, tetrahedral mesh, $p=3$).
-Using the Intel Architecture Code Analyzer tool~\cite{IACA} on the Sandy Bridge, we confirmed that this is a consequence of increased register re-use.
+Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} illustrate the speed up achieved by \emph{op-vect} over \emph{licm-ap} in the Helmoltz and Diffusion kernels, respectively.
+As explained in Section~\ref{sec:pyop2-compiler}, the \emph{op-vect} version requires the unroll-and-jam factor to be explicitly set.
+To distinguish between the two ways this parameter was determined, for each problem instance (equation, element shape, polynomial order) we report two bars: one shows the best speed-up obtained after all feasible unroll-and-jam factors were tried; the other shows the speed up when the unroll-and-jam factor was retrieved via the cost model.
+In a plot legend, cost model bars are suffixed with ``CM''.
+
+It is worth noticing that, in most cases, the cost model successfully determines how to transform a kernel to maximize its performance.
+This is chiefly because assembly kernels fit the L1 cache, so, within a certain degree of confidence, it is possible to predict how to obtain a fast implementation by simply reasoning on the register pressure.
+For each problem, the cost model stated to use the default loop permutation, to apply a particular unroll-and-jam factor, and not to perform expression splitting, which, as explained in Section~\ref{sec:perf-results-split}, only deteriorates performance in Helmholtz and Diffusion.
+
+The rationale behind these results is that the effect of \emph{op-vect} is significant in problems in which the assembly loop nest is relatively big.
+When the loops are short, since the number of arrays accessed at every loop iteration is rather small (between 5 and 9), there is no need for vector-register tiling; extensive unrolling is sufficient to improve register re-use and, therefore, to maximize the performance.
+However, as the iteration space becomes larger, \emph{op-vect} leads to improvements up to 1.4$\times$ on the Sandy Bridge (Diffusion, prismatic mesh, $p=4$ - increasing the overall speed up from 2.69$\times$ to 3.87$\times$), and up to 1.4$\times$ on the Xeon Phi (Helmholtz, tetrahedral mesh, $p=3$ - bringing the overall speed up from 1.71$\times$ to 2.42$\times$).
+
+Using the Intel Architecture Code Analyzer tool~\cite{IACA} on the Sandy Bridge, we confirmed that speed ups are a consequence of increased register re-use.
 In Helmholtz $p=4$, for example, the tool showed that when using \emph{op-vect} the number of clock cycles to execute one iteration of the \texttt{j} loop decreases by roughly 17$\%$, and that this is a result of the relieved pressure on both of the data (cache) ports available in the core.
@@
 	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/helmholtz-normalized-opvect-phi.pdf}}
-	\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Helmholtz kernel.
-CM refers to where the cost model has been used to determine the sequence of code transformations.}
+	\caption{Performance improvement over \emph{licm-ap} obtained by \emph{op-vect} in the Helmholtz kernel. Bars suffixed with ``CM'' indicate that the cost model was used to transform the kernel.}
 	\label{fig:opvect-helmholtz-speedup}
@@
 	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/diffusion-normalized-opvect-phi.pdf}}
-	\caption{Performance improvement obtained applying \emph{op-vect} on top of \emph{licm-ap} to the Diffusion kernel.
-CM refers to where the cost model has been used to determine the sequence of code transformations.}
+	\caption{Performance improvement over \emph{licm-ap} obtained by \emph{op-vect} in the Diffusion kernel. Bars suffixed with ``CM'' indicate that the cost model was used to transform the kernel.}
 	\label{fig:opvect-diffusion-speedup}
@@
 	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/burgers-normalized-split-phi.pdf}}
-	\caption{Performance improvement obtained applying \emph{split} on top of \emph{licm-ap} to the Burgers kernel.
-CM refers to where the cost model has been used to determine the sequence of code transformations.}
+	\caption{Performance improvement over \emph{licm-ap} obtained by \emph{split} in the Burgers kernel. Bars suffixed with ``CM'' indicate that the cost model was used to transform the kernel.}
 	\label{fig:split-burgers-speedup}
@@
 \begin{table}[h]
-\begin{center}
 \tbl{Performance comparison between FEniCS (optimizations enabled) and COFFEE on the Sandy Bridge}{
-\label{table:comparison-to-FFC-opt}
 \begin{tabulary}{1.0\columnwidth}{C|C|C}
@@
 }
-\end{center}
+\label{table:comparison-to-FFC-opt}
 \end{table}

### 2014-06-24 TACO-paper/coffee.tex
Partly re-written impact of expression splitting

@@
 This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large.
-For example, applying loop-invariant code motion to Burgers on a 3D mesh requires 33 temporaries for the \texttt{ijk} loop order.
+For example, applying loop-invariant code motion to Burgers on a 3D mesh requires 24 temporaries for the \texttt{ijk} loop order.
 This can make hoisting of the invariant loads out of the \texttt{k} loop inefficient on architectures with a relatively low number of registers.
@@
 The rationale behind these results is that the effect of \emph{op-vect} is significant in problems in which the assembly loop nest is relatively big.
-When the loops are short, since the number of arrays accessed at every loop iteration is rather small (between 5 and 9), there is no need for vector-register tiling; extensive unrolling is sufficient to improve register re-use and, therefore, to maximize the performance.
+When the loops are short, since the number of arrays accessed at every loop iteration is rather small (between 4 and 8 temporaries, plus the element matrix itself), there is no need for vector-register tiling; extensive unrolling is sufficient to improve register re-use and, therefore, to maximize the performance.
 However, as the iteration space becomes larger, \emph{op-vect} leads to improvements up to 1.4$\times$ on the Sandy Bridge (Diffusion, prismatic mesh, $p=4$ - increasing the overall speed up from 2.69$\times$ to 3.87$\times$), and up to 1.4$\times$ on the Xeon Phi (Helmholtz, tetrahedral mesh, $p=3$ - bringing the overall speed up from 1.71$\times$ to 2.42$\times$).
@@
 \label{sec:perf-results-split}
-Expression splitting relieves the register pressure when the element matrix evaluation reads from a large number of arrays, at the price of increasing accesses to the element matrix itself and potentially affecting data locality, as detailed in Section~\ref{sec:expr-split}.
-It is not particularly useful, therefore, for Helmholtz and Diffusion, where only between 4 and 8 basis functions or temporaries need to be read at every iteration.
+Expression splitting relieves the register pressure when the element matrix evaluation needs to read from a large number of basis function arrays.
+As detailed in Section~\ref{sec:expr-split}, in the price to pay for this optimazion, there are an increased number of accesses to the element matrix and, potentially, redundant computation.
+Similarly to the analysis of vector-register tiling, we compare two versions: the baseline, \emph{licm-ap}; and expression splitting on top of \emph{licm-ap}, which, for simplicity, in the following is referred to as \emph{split}.
+
+For the Helmholtz and Diffusion kernels, in which only between 4 and 8 temporaries are read at every loop iteration, \texttt{split} tends to slow down the computation, because of the aforementioned drawbacks.
 Slow downs up to 1.4$\times$ and up to 1.6$\times$ were observed, respectively, on the Sandy Bridge and the Xeon Phi.
-The cost model, however, prevents the adoption of the transformation: the \texttt{while} statement at line 5 in Figure~\ref{algo:applyCostModel} is indeed never entered.
-For the Burgers kernel, \emph{split} has a key role on the Sandy Bridge, whereas it has only a minimal impact on the Xeon Phi.
-Figure~\ref{fig:split-burgers-speedup} shows the improvement due to using \emph{split} over the fastest implementation between \emph{licm} and \emph{licm-ap}.
-On the Sandy Bridge, in almost all test cases, a split factor of 1, meaning that the original expression was divided into two parts, ensured close-to-peak perforance.
+Note that the cost model prevents the adoption of the transformation: the \texttt{while} statement in Figure~\ref{algo:applyCostModel} is indeed never entered.
+
+In the Burgers kernels, between 12 and 24 temporaries are accessed at every loop iteration, so \emph{split} plays a key role on the Sandy Bridge, where the number of available logical registers is only 16.
+Figure~\ref{fig:split-burgers-speedup} shows the performance improvement achieved by \emph{split} over \emph{licm-ap}.
+In almost all cases, a split factor of 1, meaning that the original expression was divided into two parts, ensured close-to-peak perforance.
 The transformation negligibly affected register locality.
-On the contrary, the larger number of vector units present on the Xeon Phi relieve register spilling, making \emph{split} only marginally useful for loop unrolling.
+On the Xeon Phi, the impact of \emph{split} is only marginal, since register spilling is limited by the presence of 32 logical vector units.
 
@@
 Despite improvements up to 1.22$\times$, \emph{split-op-vect} never outperforms \emph{split}.
-This is for two reasons: for small split factors, such as 1 and 2, the data space to be tiled is still too big, and register spilling affects run-time; for higher ones, sub-expressions become so small that, as explained in Section~\ref{sec:perf-eval-opvect}, full or partial unrolling of the loop nest achieves a certain degree of register re-use by itself.
+This is motivated by two factors: for small split factors, such as 1 and 2, the data space to be tiled is still too big, and register spilling affects run-time; for higher ones, sub-expressions become so small that, as explained in Section~\ref{sec:perf-eval-opvect}, extensive unrolling already allows to achieve a certain degree of register re-use.
 

### 2014-06-24 TACO-paper/coffee.tex
Minor adjustments

@@
 \usepackage{tabulary}
+\usepackage{lineno}
 
@@
 \category{G.1.8}{Numerical Analysis}{Partial Differential Equations - Finite element methods}
+\category{G.4}{Mathematical Software}{Parallel and vector implementations}
 
@@
 \footnotesize
-\KwSty{void} helmholtz(double A[8][8], double **coords) $\lbrace$\\
+\KwSty{void} helmholtz(double A[4][4], double **coords) $\lbrace$\\
 ~~// K, det = Compute Jacobian (coords) \\
@@
 ~~\\
-~~\KwSty{for} (\KwSty{int} i = 0; i$<$6; i++) $\lbrace$ \\
+~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
 ~~~~// Do generalized loop-invariant code motion \\
 ~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) \\
-~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
+~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k+=4) $\lbrace$\\
 ~~~~~~~~// $load$ and $set$ intrinsics \\
@@
 ~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
-~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$8; k+=4) $\lbrace$\\
+~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k+=4) $\lbrace$\\
 ~~~~~~r0 = $\_$mm256$\_$load$\_$pd ($\&$A[j+0][k]);\\
@@
 \caption{Local assembly code generated by Firedrake for the Helmholtz problem after application of \emph{op-vect} on top of the optimizations shown in Listing~\ref{code:helmholtz-licm}.
-Here, we assume the polynomial order is $p=2$, since \emph{op-vect} can not be used when an iteration space dimension is smaller than the vector length.
-The original size of the \texttt{j}-\texttt{k} iteration space (i.e.
-before padding was applied) was 6$\times$6.
 In this example, the unroll-and-jam factor is 1.}
@@
 \KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
-~~~~// Same code as in Listing~\ref{code:helmholtz-licm} up to line 15 \\
+~~~~// Same code as in Listing~\ref{code:helmholtz-licm} up to the j loop\\
 ~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
@@
 \end{alltt}
-
 \caption{Outline of the COFFEE architecture.}
@@
 % "vendor compilers such as Intel's" is awkward and unclear. If Intel is the only one you have looked at, then say "the Intel compiler", otherwise name the ones for which you know this to be true.
-Since loop unroll and unroll-and-jam of outer loops are fundamental to the exposure of instruction-level parallelism, tuning critical parameters such as the unroll factor is of great importance.
-However, inspecting assembly code and comparing with other hand-written implementations indicates that recent versions of the Intel compiler are capable of estimating close-to-optimal values for such parameters.
-This is particularly true for assembly kernels, in which the loop nest is affine, bounds are usually very small and known at compile-time, and memory accesses are stride one.
-We therefore leave the backend compiler in charge of selecting the unroll factor.
+Loop unroll  (or unroll-and-jam of outer loops) is fundamental to the exposure of instruction-level parallelism, and tuning the unroll factor is particularly important.
+However, we noticed, by comparison with implementations having manually-unrolled loops, that recent versions of the Intel compiler estimate close-to-optimal unroll factors when the loops are affine and their bounds are small and known at compile-time, which is the case of our kernels.
+We therefore leave the backend compiler in charge of selecting unroll factors.
 This choice also simplifies the COFFEE's cost model.
-The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code seems to prevent the Intel compiler from applying this optimization, even if specific pragmas are added.
+The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code prevents the Intel compiler from applying this optimization, even if specific pragmas are added.
 
@@
 \begin{figure}
-\begin{alltt}
-\footnotesize
+\begin{alltt}\footnotesize\internallinenumbers
 \textbf{Input:} n\_outer\_arrays, n\_inner\_arrays, n\_consts, n\_regs
@@
 If a factor is 0, then the corresponding transformation is not applied.
-The \emph{split} transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 2).
+The \emph{split} transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 3).
 A split factor of $n$ means that the assembly expression should be ``cut'' into $n$ sub-expressions.
@@
 those not used for storing hoisted terms (line 9-11).
-Finally, the profitability of loop interchange is evaluated (line 12-16).
+Finally, the profitability of loop interchange is evaluated (line 13).
 
@@
 \end{enumerate}
-The three chosen benchmarks are completely real examples and comprise the core operators in some of the most frequently encountered finite element problems in scientific computing.
+The three chosen benchmarks are completely real examples and comprise the core differential operators in some of the most frequently encountered finite element problems in scientific computing.
 The Helmholtz and Diffusion kernels are archetypal second order elliptic operators.
-They are complete and unsimplified examples of the operators which occur for diffusion and viscosity in fluids, and for imposing pressure in compressible fluids.
-For instance, they are both extensively used in climate modeling.
-Very similar operators, for which the same optimisations are expected to be equally effective, apply to elasticity problems, which are at the core of computational structural mechanics.
+They are complete and unsimplified examples of the operators occurring for diffusion and viscosity in fluids, and for imposing pressure in compressible fluids.
+As such, they are both extensively used in climate and ocean modeling.
+Very similar operators, for which the same optimisations are expected to be equally effective, apply to elasticity problems, which are at the base of computational structural mechanics.
 The Burgers kernel is a typical example of a first order hyperbolic conservation law, which occurs in real applications whenever a quantity is transported by a fluid (the momentum itself, in our case).
-We chose this particular kernel since it applies to a vector-valued quantity, while the elliptic operators apply to scalar quantities.
-In this way, the operators we have selected as examples are characteristic of both the second and first order operators which dominate in fluids and solids simulations.
+We chose this particular kernel since it applies to a vector-valued quantity, while the elliptic operators apply to scalar quantities; this impacts the generated code, as explained next.
+In this way, the operators we have selected are characteristic of both the second and first order operators that dominate in fluids and solids simulations, and a wide variety on local assembly codes can be tested.
 
-The code was written in UFL (available at~\cite{ufl-code}) and then executed over real unstructured meshes through Firedrake.
+The benchmarks were written in UFL (code available at~\cite{ufl-code}) and executed over real unstructured meshes through Firedrake.
 The Helmholtz code has already been shown in Listing~\ref{code:helmholtz}.
 For Advection-Diffusion, the Diffusion equation, which uses the same differential operators as Helmholtz, is considered.
-In the Diffusion kernel code, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$.
+In the Diffusion kernel code, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix.
 Burgers is a non-linear problem employing differential operators different from those of Helmholtz and relying on vector-valued quantities, which has a major impact on the generated assembly code (see Listing~\ref{code:burgers}), where a larger number of basis function arrays ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are generated.

### 2014-06-25 TACO-paper/coffee.tex
Introduction revised

@@
 
+%Computational cost is a critical limitation in scientific computing, especially
+%for finite element simulations. To provide one particular example we are
+%particularly concerned about, it has been well established that mesh resolution
+ Consequently, our aggressive optimization of
+%local assembly, which may even take up to 60% of the overall FEM's execution
+%time, directly impacts the performance of large-scale scientific simulations
+%running on supercomputers.
+
 \section{Introduction}
 In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by partial differential equations (PDEs).
-Numerical techniques, such as the finite volume method and finite element method, are widely employed to approximate solutions of these PDEs.
+Numerical techniques, like the finite volume method and the finite element method, are widely employed to approximate solutions of these PDEs.
 Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective.
-The solution is sought in each cell, or ``element'', of the discretized domain by applying suitable numerical operations.
-As the number of elements can be of the order of trillions \cite{Rossinelli2013}, a major issue is the time required to execute the computation, which can be hours or days.
-To address this problem, domain-specific languages (DSLs) have been developed.
-The successful port of Hydra, an industrial computational fluid dynamics application developed by Rolls Royce for turbomachinery design (based on finite volume method, roughly 50,000 lines of code and with meshes in excess of 100M edges), to OP2~\cite{op2-main}, demonstrates the effectiveness of the DSL approach in the implementation of PDE solvers~\cite{IstvanHydra}.
+The solution is sought by applying suitable numerical operations, or kernels, to the components of a mesh, such as edges, vertices, or cells.
+On standard clusters of multi-cores, typically, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a different node or thread.
+Such execution model, with minor variations, is adopted, for example, in \cite{pyop2isc}, \cite{Fenics}, \cite{fluidity_manual_v4}, \cite{lizst}.
 
-OP2 adopts a programming model in which computations are expressed through self-contained functions, referred to as ``kernels''.
-A kernel is applied to all elements in a set of mesh components, such as edges, vertices, or cells, with an implicit synchronization between the application of two consecutive kernels.
-On commodity multi-cores, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a thread.
-Similar programming and execution models are adopted in~\cite{pyop2isc},~\cite{Fenics},~\cite{fluidity_manual_v4},~\cite{lizst}.
-Kernel optimization is one of the major concerns in unstructured mesh applications.
-In this paper, we tackle this problem in the context of the finite element method.
+%JUNK: , with synchronization handled at the level of the run-time support
 
-We focus on improving the performance of local assembly (or ``assembly'', in the following), a fundamental step in the finite element method that covers an important fraction of the overall computation run-time, often in the range 30-60$\%$.
-During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain.
+The time required to apply the numerical kernels is a major issue, since the equation domain needs to be discretized into an extremely large number of elements, possibly order of trillions \cite{Rossinelli2013}, in order to obtain a satisfactory approximation of the PDE.
+For example, it has been well established that mesh resolution is critical in the accuracy of numerical weather forecasts.
+However, operational forecast centres have a strict time limit in which to produce a forecast (60 minutes in the case of the UK Met Office, who are partially funding this work), so producing efficient kernels has a direct scientific payoff in higher resolution, and therefore more accurate, forecasts.
+In finite elements, numerical kernels are executed during the local assembly phase (or ``assembly'', in the following), which can cover a significant fraction of the overall computation run-time, often in the range 30-60$\%$.
+In this paper, we aim at improving the performance of local assembly kernels in a wide variety of notably important PDEs though aggressive code transformation.
+
+During the assembly phase, the solution of the PDE is approximated by executing a problem-specific kernel over all cells, or elements, in the discretized domain.
 In this work, we focus on relatively low order finite element methods, in which an assembly kernel's working set is usually small enough to fit the L1 cache.
@@
 A compiler-based approach is, therefore, the only reasonable option to obtain close-to-peak performance in a wide range of different local assembly kernels.
-Optimizations like generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but none are supported by state-of-the-art polyhedral and vendor compilers.
-BLAS routines could theoretically be employed, although a fairly complicated control- and data-flow analysis is required to automate identification and extraction of matrix-matrix multiplies.
+Optimizations like padding, generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but none are supported by state-of-the-art polyhedral and vendor compilers.
+BLAS routines could theoretically be employed, although a fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies.
 In addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small dimension of the  matrices involved and the potential loss in data locality can limit or eliminate the performance gain of this approach.
@@
 COFFEE is integrated with Firedrake~\cite{firedrake-code}, a system for solving PDEs through the finite element method based on the PyOP2 abstraction~\cite{pyop2isc}.
-It supports all problems expressible with this framework.
-This allows us to evaluate our code transformations in a range of real-world problems, varying two key parameters that impact both solution accuracy and kernel cost: the polynomial order of the method (we investigate from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).
+It supports all problems expressible with this framework, including equations that can be found at the core of real -world simulations, such as those used in our performance evaluation~\ref{sec:real-examples}.
+In evaluating our code transformations in a range of notable problems, we also vary two key parameters that impact solution accuracy and kernel cost: the polynomial order of the method (we investigate from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).
 
+%JUNK: 1.5$\times$ over code in which only basic transformations have been performed, such as generalized loop-invariant code motion, padding, and data alignment, and up to
 Our experiments show that the original generated code for non-trivial assembly kernels is sub-optimal.
-Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over code in which only basic transformations have been performed, such as generalized loop-invariant code motion, padding, and data alignment, and up to 4.4$\times$ over original kernels.
+Our cost-model-driven sequence of code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 4.4$\times$ over original kernels.
 The contributions of this paper are
@@
 
-Despite the infinite variety of problem-specific assembly kernels which Firedrake can generate, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.
+Despite the infinite variety of assembly kernels which Firedrake can generate, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.
 These include: 1) memory accesses along the three loop dimensions are always stride-1; 2) the \texttt{j} and \texttt{k} loops are interchangeable, whereas interchanges involving the $i$ loop require pre-computation of values (e.g.
@@
 The absence of stencils renders the element matrix computation easily auto-vectorizable by a vendor compiler.
-Nevertheless, auto-vectorization is not efficient if data are not aligned and if the length of the innermost loop is not a multiple of the vector length $vl$, especially when the loops are particularly small as in the case of assembly.
+Nevertheless, auto-vectorization is not efficient if data are not aligned and if the length of the innermost loop is not a multiple of the vector length $vl$, especially when the loops are small as in local assembly.
 
@@
 It is also worth noting that, in some problems (e.g.
-in Helmholtz), invariant sub-expressions along \texttt{j} are identical to those along \texttt{k}, and both loops iterate over the same iteration space.
-In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that \texttt{j} and \texttt{k} loops share the same iteration space.
+Helmholtz), invariant sub-expressions along \texttt{j} are identical to those along \texttt{k}, and both loops iterate over the same iteration space, as anticipated in Section~\ref{sec:background}.
+In these cases, we safely avoid redundant pre-computation.
 
@@
 Numerical integration is usually employed to implement the local assembly phase.
-The recent introduction of DSLs to decouple the finite element specification from its underlying implementation facilitated, however, the development of novel approaches.
+The recent introduction of domain specific languages (DSLs) to decouple the finite element specification from its underlying implementation facilitated, however, the development of novel approaches.
 Methods based on tensor contraction~\cite{FFC-Compiler} and symbolic manipulation~\cite{Francis} have been implemented.

### 2014-06-25 TACO-paper/coffee.tex
minor fix
Introduction further re-written

@@
 %particularly concerned about, it has been well established that mesh resolution
- Consequently, our aggressive optimization of
+% Consequently, our aggressive optimization of
 %local assembly, which may even take up to 60% of the overall FEM's execution
@@
 %JUNK: , with synchronization handled at the level of the run-time support
+%Computational cost is a critical limitation in finite element simulations.
+High resolution in time and space is critical to the level of error in the simulations, but computational cost rises superlinearly with resolution.
+The result of this is that simulations are conducted at the resolution that the user can afford (wall time, allocated supercomputing units or the direct financial cost of cloud time)
 
-The time required to apply the numerical kernels is a major issue, since the equation domain needs to be discretized into an extremely large number of elements, possibly order of trillions \cite{Rossinelli2013}, in order to obtain a satisfactory approximation of the PDE.
+The time required to apply the numerical kernels is a major issue, since the equation domain needs to be discretized into an extremely large number of cells, possibly order of trillions \cite{Rossinelli2013}, in order to obtain a satisfactory approximation of the PDE.
 For example, it has been well established that mesh resolution is critical in the accuracy of numerical weather forecasts.
 However, operational forecast centres have a strict time limit in which to produce a forecast (60 minutes in the case of the UK Met Office, who are partially funding this work), so producing efficient kernels has a direct scientific payoff in higher resolution, and therefore more accurate, forecasts.
-In finite elements, numerical kernels are executed during the local assembly phase (or ``assembly'', in the following), which can cover a significant fraction of the overall computation run-time, often in the range 30-60$\%$.
-In this paper, we aim at improving the performance of local assembly kernels in a wide variety of notably important PDEs though aggressive code transformation.
+Hence, the computational cost is a dominant problem in computational science simulations, especially for those based on finite elements, which are the subject of this paper.
+We address, in particular, the well-known problem of optimizing the local assembly phase of the finite element method~\cite{Francis},~\cite{quadrature1},~\cite{petsc-integration-gpu},~\cite{Kirby-FEM-opt}, which can cover a significant fraction of the overall computation run-time, often in the range 30-60$\%$.
+With respect to these studies, we propose a novel set of composable code transformations targeting, for the first time, instruction-level parallelism - with emphasis on SIMD vectorization - and register locality.
 
@@
 Given a finite element description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl} (UFL), Firedrake generates a C-code kernel implementing assembly using a numerical integration algorithm.
-The same kernel can be applied to any elements: this descends from the mathematical formulation of assembly, which states that the evaluation of a local matrix can be reduced to integrating on a fixed ``reference'' element - a special element that does not belong to the domain - after a suitable change of coordinates.
+This kernel can be applied to any element in the mesh: this descends from the mathematical formulation of assembly, which states that the evaluation of a local matrix can be reduced to integrating on a fixed ``reference'' element - a special element that does not belong to the domain - after a suitable change of coordinates.
 Firedrake triggers the compilation of an assembly kernel using an available vendor compiler, and manages its (parallel) execution over all elements in the mesh.
-As already stated, the subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.
+As already explained, the subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.
 

### 2014-06-26 TACO-paper/coffee.tex
Some minor fixes - add a few key sentences/words

@@
 \begin{bottomstuff}
-This research is partly funded by the MAPDES project, by the Department of Computing at Imperial College London, by EPSRC through grants EP/I00677X/1, EP/I006761/1, and EP/L000407/1, and by NERC grants NE/K008951/1 and NE/K006789/1.
+This research is partly funded by the MAPDES project, by the Department of Computing at Imperial College London, by EPSRC through grants EP/I00677X/1, EP/I006761/1, and EP/L000407/1, by NERC grants NE/K008951/1 and NE/K006789/1, and by a HiPEAC collaboration grant.
 The authors would like to thank Dr.
@@
 J.
-Kelly, Department of Computing, Imperial College London; Ana Lucia Varbanescu, Department of Computer Sciences, University of Amsterdam; David A.
+Kelly, Department of Computing, Imperial College London; Ana Lucia Varbanescu, Informatics Institute, University of Amsterdam; David A.
 Ham, Department of Computing and Department of Mathematics, Imperial College London; J.
@@
 For example, it has been well established that mesh resolution is critical in the accuracy of numerical weather forecasts.
-However, operational forecast centres have a strict time limit in which to produce a forecast (60 minutes in the case of the UK Met Office, who are partially funding this work), so producing efficient kernels has a direct scientific payoff in higher resolution, and therefore more accurate, forecasts.
+However, operational forecast centres have a strict time limit in which to produce a forecast - 60 minutes in the case of the UK Met Office - so producing efficient kernels has a direct scientific payoff in higher resolution, and therefore more accurate, forecasts.
 Hence, the computational cost is a dominant problem in computational science simulations, especially for those based on finite elements, which are the subject of this paper.
@@
 An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, in which individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for low order methods.
-In the innermost loop, a problem-specific expression evaluates a two dimensional array, representing the result of local assembly in an element of the discretized domain.
+In the innermost loop, a problem-specific, flops-intensive expression evaluates a two dimensional array, representing the result of local assembly in an element of the discretized domain.
 With such a kernel structure, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.
@@
 COFFEE is integrated with Firedrake~\cite{firedrake-code}, a system for solving PDEs through the finite element method based on the PyOP2 abstraction~\cite{pyop2isc}.
-It supports all problems expressible with this framework, including equations that can be found at the core of real -world simulations, such as those used in our performance evaluation~\ref{sec:real-examples}.
+It supports all problems expressible with this framework, including equations that can be found at the core of real-world simulations, such as those used in our performance evaluation.
 In evaluating our code transformations in a range of notable problems, we also vary two key parameters that impact solution accuracy and kernel cost: the polynomial order of the method (we investigate from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).
+From the point of view of the generated code, these two parameters directly impact the size of both loop nests and mathematical expressions, as elaborated in Section~\ref{sec:real-examples}.
 
 %JUNK: 1.5$\times$ over code in which only basic transformations have been performed, such as generalized loop-invariant code motion, padding, and data alignment, and up to
-Our experiments show that the original generated code for non-trivial assembly kernels is sub-optimal.
-Our cost-model-driven sequence of code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 4.4$\times$ over original kernels.
-The contributions of this paper are
+
+Our experiments show that the original generated code for non-trivial assembly kernels, despite following state-of-the-art techniques, remains suboptimal in the context of modern multi-core architectures.
+Our cost-model-driven sequence of code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 4.4$\times$ over original kernels, and around 70$\%$ of the test cases obtain a speed up greater than 2$\times$.
+The contributions of this paper, therefore, are
 \begin{enumerate}
@@
 \end{enumerate}
-The three chosen benchmarks are completely real examples and comprise the core differential operators in some of the most frequently encountered finite element problems in scientific computing.
+The three chosen benchmarks are \emph{real-life kernels} and comprise the core differential operators in some of the most frequently encountered finite element problems in scientific computing.
 The Helmholtz and Diffusion kernels are archetypal second order elliptic operators.

### 2014-06-28 TACO-paper/coffee.tex
Minor fixes. Add coffee overview picture

@@
 \label{sec:expr-split}
-In complex kernels, like Burgers in Listing~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
-If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, poor register reuse is likely to be obtained.
-This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large.
-For example, applying loop-invariant code motion to Burgers on a 3D mesh requires 24 temporaries for the \texttt{ijk} loop order.
-This can make hoisting of the invariant loads out of the \texttt{k} loop inefficient on architectures with a relatively low number of registers.
-One potential solution to this problem consists of suitably ``splitting'' the computation of the element matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Listing~\ref{code:helmholtz-split}.
-The transformation can be regarded as a special case of classic loop fission, in which sum's associativity is exploited to distribute the expression across multiple loops.
-To the best of our knowledge, expression splitting is not supported by available compilers.
-
-Splitting an expression has, however, several drawbacks.
-Firstly, it increases the number of accesses to $A$ proportionally to the ``split factor'', which is the number of sub-expressions produced.
-Also, depending on how the split is executed, it can lead to redundant computation; for example, the times the product $det*W3[i]$ is performed is proportional to the number of sub-expressions, as shown in the code snippet.
-Further, it increases loop overhead (e.g.
-branch instructions).
-Finally, it might affect register locality: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
-This is not the case of the Helmholtz example.
-Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.
 
@@
 
+In complex kernels, like Burgers in Listing~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
+If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, poor register reuse is likely to be obtained.
+This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large.
+For example, applying loop-invariant code motion to Burgers on a 3D mesh requires 24 temporaries for the \texttt{ijk} loop order.
+This can make hoisting of the invariant loads out of the \texttt{k} loop inefficient on architectures with a relatively low number of registers.
+One potential solution to this problem consists of suitably ``splitting'' the computation of the element matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Listing~\ref{code:helmholtz-split}.
+The transformation can be regarded as a special case of classic loop fission, in which sum's associativity is exploited to distribute the expression across multiple loops.
+To the best of our knowledge, expression splitting is not supported by available compilers.
+
+Splitting an expression has, however, several drawbacks.
+Firstly, it increases the number of accesses to $A$ proportionally to the ``split factor'', which is the number of sub-expressions produced.
+Also, depending on how the split is executed, it can lead to redundant computation; for example, the times the product $det*W3[i]$ is performed is proportional to the number of sub-expressions, as shown in the code snippet.
+Further, it increases loop overhead (e.g.
+branch instructions).
+Finally, it might affect register locality: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
+This is not the case of the Helmholtz example.
+Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.
+
+\section*{}
 Table~\ref{table:code-transformations} summarizes the code transformations described so far.
@@
 The output of COFFEE, C code, is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and execution over the discretized domain take place.
+The flow is summarized in Figure~\ref{fig:coffee-pipeline}.
+
 Because of the large number of parametric transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem.
-Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of ASTs occurs at run-time as soon as problem-secific data are available.
+Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of ASTs occurs at run-time as soon as problem-specific data are available.
 Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[scale=0.75]{Pictures/coffee-pipeline.pdf}
+\caption{High-level view of Firedrake. COFFEE is at the core, receiving ASTs from a modified version of the FEniCS Form compiler and producing optimized C code kernels.}
+\label{fig:coffee-pipeline}
+\end{center}
+\end{figure}
 
@@
 In almost all cases, a split factor of 1, meaning that the original expression was divided into two parts, ensured close-to-peak perforance.
-The transformation negligibly affected register locality.
+The transformation negligibly affected register locality, so speed ups up to 1.5$\times$ were observed.
+For instance, on the Sandy Bridge, when $p=4$ and a prismatic mesh is employed, the overall performance improvement (i.e.
+the one over the original code) increases from 1.44$\times$ to 2.11$\times$.
 On the Xeon Phi, the impact of \emph{split} is only marginal, since register spilling is limited by the presence of 32 logical vector units.
@@
 In FEniCS, FFC can apply its own model-driven optimizations to local assembly kernels~\cite{quadrature1}, which mainly consist of loop-invariant code motion and elimination, at code generation time, of floating point operations involving zero-valued entries in basis function arrays.
-The FEniCS Form Compiler's loop-invariant code motion is, however, different from COFFEE's, since it is based on expansion of arithmetic operations, for example applying distributivity and associativity to products and sums at code generation time, to identify terms that are invariant of the whole loop nest.
-In addition, depending on the way expansion is performed, operation count often does not decrease significantly.
-On the other hand, elimination of zero-valued terms, which are the result of certain properties of the finite element problem, has the effect of introducing indirection arrays in the generated code.
+
+The FEniCS Form Compiler's loop-invariant code motion is different from COFFEE's.
+It is based on expansion of arithmetic operations, for example applying distributivity and associativity to products and sums at code generation time, to identify terms that are invariant of the whole loop nest.
+Depending on the way expansion is performed, operation count may not decrease significantly.
+
+Elimination of zero-valued terms, which are the result of using vector-valued quantities in the finite element problem, has the effect of introducing indirection arrays in the generated code.
 This kind of optimization is currently under development in COFFEE, although it will differ from FEniCS' by avoiding non-contiguous memory accesses, which would otherwise affect vectorization, at the price of removing fewer zero-valued contributions.
 
-Table~\ref{table:comparison-to-FFC-opt} summarizes the performance achieved by COFFEE over the best FEniCS (FFC) implementation (i.e.
-the combination of optimizations for each problem which results in the fastest code) on the Sandy Bridge for the Burgers, Helmholtz and Diffusion kernels.
+Table~\ref{table:comparison-to-FFC-opt} summarizes the performance achieved by COFFEE over the \emph{fastest} FEniCS (FFC) implementation on the Sandy Bridge for the Burgers, Helmholtz and Diffusion kernels.
 Burgers' slow downs occur in presence of a small iteration space (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$).
@@
 \begin{table}[h]
-\tbl{Performance comparison between FEniCS (optimizations enabled) and COFFEE on the Sandy Bridge}{
+\tbl{Performance comparison between FEniCS and COFFEE on the Sandy Bridge.}{
 \begin{tabulary}{1.0\columnwidth}{C|C|C}
@@
 In the case of $p=4$, where the element matrix is of size 35$\times$35, the computation was almost twice slower than the case in which \emph{licm-ap} was used, with the slow down being even worse for smaller problem sizes.
-These experiments suggest that the question regarding to what extent linear algebra libraries or, analogously, auto-tuning strategies can improve performance cannot be trivially answered.
-This is due to a combination of issues: the potential loss in data locality as exposed in Section~\ref{sec:expr-split}, the actual effectiveness of external tools when the arrays are relatively small, and the problem inherent to assembly kernels concerning extraction of matrix-matrix multiplies from static analysis of the kernel's code.
+These experiments suggest that the question regarding to what extent linear algebra libraries can improve performance cannot be trivially answered.
+This is due to a combination of issues: the potential loss in data locality, as exposed in Section~\ref{sec:expr-split}, the actual effectiveness of these libraries when the arrays are relatively small, and the problem inherent to assembly kernels concerning extraction of matrix-matrix multiplies from static analysis of the kernel's code.
 A comprehensive study of these aspects will be addressed in further work.

### 2014-06-28 TACO-paper/coffee.tex
Layout adjustment
edit

@@
 \begin{bottomstuff}
-This research is partly funded by the MAPDES project, by the Department of Computing at Imperial College London, by EPSRC through grants EP/I00677X/1, EP/I006761/1, and EP/L000407/1, by NERC grants NE/K008951/1 and NE/K006789/1, and by a HiPEAC collaboration grant.
-The authors would like to thank Dr.
-Carlo Bertolli, Dr.
-Lawrence Mitchell, and Dr.
-Francis Russell for their invaluable suggestions and their contribution to the Firedrake project.
-
-Author's addresses: Fabio Luporini $\&$ Florian Rathgeber $\&$ Gheorghe-Teodor Bercea $\&$ Paul H.
-J.
-Kelly, Department of Computing, Imperial College London; Ana Lucia Varbanescu, Informatics Institute, University of Amsterdam; David A.
-Ham, Department of Computing and Department of Mathematics, Imperial College London; J.
-Ramanujam, Department of Electrical and Computer Engineering, Louisiana State University.
+
+This research is partly funded by the MAPDES project, by the
+Department of Computing at Imperial College London, by EPSRC through
+grants EP/I00677X/1, EP/I006761/1, and EP/L000407/1, by NERC grants
+NE/K008951/1 and NE/K006789/1, by the U.S.  National Science
+Foundation through grants 0811457 and 0926687, by the U.S. Army
+through contract W911NF-10-1-000, and by a HiPEAC collaboration
+grant. The authors would like to thank Dr. Carlo Bertolli,
+Dr. Lawrence Mitchell, and Dr. Francis Russell for their invaluable
+suggestions and their contribution to the Firedrake project.
+
+Author's addresses: Fabio Luporini $\&$ Florian Rathgeber $\&$
+Gheorghe-Teodor Bercea $\&$ Paul H. J. Kelly, Department of Computing,
+Imperial College London; Ana Lucia Varbanescu, Informatics Institute,
+University of Amsterdam; David A. Ham, Department of Computing and
+Department of Mathematics, Imperial College London; J. Ramanujam,
+iCenter for Computation and Technology and the School of Elec.~Engr.\&
+Comp.~Sci., Louisiana State University.
 \end{bottomstuff}
@@
 \section*{}
-Table~\ref{table:code-transformations} summarizes the code transformations described so far.
-Given that many of these transformations depend on some parameters (e.g.
-tile size), we need a mechanism to prune such a large space of optimization.
-This aspect is treated in Section~\ref{sec:pyop2-compiler}.
-
-\begin{table}[t]
+
+\begin{table}[h]
 \tbl{Overview of code transformations for Firedrake-generated assembly kernels.}{
@@
 
-% mention (obvious) composability
+Table~\ref{table:code-transformations} summarizes the code transformations described so far.
+Given that many of these transformations depend on some parameters (e.g.
+tile size), we need a mechanism to prune such a large space of optimization.
+This aspect is treated in Section~\ref{sec:pyop2-compiler}.
 
@@
 %\rule[0.2ex]{16cm}{0.5pt}
+\begin{figure}[b]
+\begin{center}
+\includegraphics[scale=0.75]{Pictures/coffee-pipeline.pdf}
+\caption{High-level view of Firedrake. COFFEE is at the core, receiving ASTs from a modified version of the FEniCS Form compiler and producing optimized C code kernels.}
+\label{fig:coffee-pipeline}
+\end{center}
+\end{figure}
+
+Firedrake users employ the Unified Form Language to write problems in a notation resembling mathematical equations.
+At run-time, the high-level specification is translated by a modified version of the FEniCS Form Compiler~\cite{FFC-Compiler} (FFC) into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
+ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}.
+The output of COFFEE, C code, is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and execution over the discretized domain take place.
+The flow is summarized in Figure~\ref{fig:coffee-pipeline}.
+
+Because of the large number of parametric transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem.
+Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of ASTs occurs at run-time as soon as problem-specific data are available.
+Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.
+
+The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}.
+Initially the AST is inspected, looking for the presence of iteration spaces and domain-specific information provided by the higher layer.
+If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the outer, non-affine loop over mesh elements is vectorized, can be applied.
+This feature, currently under development, has been shown to be useful in several finite volume applications~\cite{inter-kernel-vect}.
+Subsequently, an ordered sequence of optimization steps are executed.
+Application of \emph{licm} must precede padding and data alignment, due to the introduction of temporary arrays.
+Based on a cost model, loop interchange, \emph{split}, and \emph{op-vect} may be introduced.
+Their implementation is based on analysis and transformation of the kernel's AST.
+When \emph{op-vect} is selected, the compiler outputs AVX or AVX-512 intrinsics code.
+Any possible corner cases are handled: for example, if \emph{op-vect} is to be applied, but the size of the iteration space is not a multiple of the vector length, then a remainder loop, amenable to auto-vectorization, is inserted.
 
@@
 
-Firedrake users employ the Unified Form Language to write problems in a notation resembling mathematical equations.
-At run-time, the high-level specification is translated by a modified version of the FEniCS Form Compiler~\cite{FFC-Compiler} (FFC) into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
-ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}.
-The output of COFFEE, C code, is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and execution over the discretized domain take place.
-The flow is summarized in Figure~\ref{fig:coffee-pipeline}.
-
-Because of the large number of parametric transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem.
-Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of ASTs occurs at run-time as soon as problem-specific data are available.
-Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.
-
-\begin{figure}[t]
-\begin{center}
-\includegraphics[scale=0.75]{Pictures/coffee-pipeline.pdf}
-\caption{High-level view of Firedrake. COFFEE is at the core, receiving ASTs from a modified version of the FEniCS Form compiler and producing optimized C code kernels.}
-\label{fig:coffee-pipeline}
-\end{center}
-\end{figure}
-
-The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}.
-Initially the AST is inspected, looking for the presence of iteration spaces and domain-specific information provided by the higher layer.
-If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the outer, non-affine loop over mesh elements is vectorized, can be applied.
-This feature, currently under development, has been shown to be useful in several finite volume applications~\cite{inter-kernel-vect}.
-Subsequently, an ordered sequence of optimization steps are executed.
-Application of \emph{licm} must precede padding and data alignment, due to the introduction of temporary arrays.
-Based on a cost model, loop interchange, \emph{split}, and \emph{op-vect} may be introduced.
-Their implementation is based on analysis and transformation of the kernel's AST.
-When \emph{op-vect} is selected, the compiler outputs AVX or AVX-512 intrinsics code.
-Any possible corner cases are handled: for example, if \emph{op-vect} is to be applied, but the size of the iteration space is not a multiple of the vector length, then a remainder loop, amenable to auto-vectorization, is inserted.
-
 All loops are interchangeable, provided that temporaries are introduced if the nest is not perfect.

### 2014-06-28 TACO-paper/coffee.tex
edits

@@
 \begin{abstract}
-The numerical solution of partial differential equations using the finite element method is one of the key applications of high performance computing.
-Local assembly is its characteristic operation.
-This entails the execution of a problem-specific kernel for each element in the discretized problem domain.
-Since the domain size can be huge, executing efficient kernels is fundamental.
-Their optimization is, however, a challenging issue.
-Even though affine loop nests are generally present, the short trip counts and the complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations.
-We present, therefore, the design and systematic evaluation of COFFEE, a domain-specific compiler for local assembly kernels.
-COFFEE manipulates abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality.
-It then generates C code including vector intrinsics.
-Experiments using a range of finite-element forms of increasing complexity show that significant performance improvement is achieved.
+The numerical solution of partial differential equations using the
+finite element method is one of the key applications of high
+performance computing. Local assembly is its characteristic
+operation. This entails the execution of a problem-specific kernel for
+each element in the discretized problem domain. Since the domain size
+can be huge, executing efficient kernels is fundamental. Their
+optimization is, however, a challenging issue. Even though affine loop
+nests are generally present, the short trip counts and the complexity
+of mathematical expressions make it hard to determine a single or
+unique sequence of successful transformations. Therefore, we present
+the design and systematic evaluation of COFFEE, a domain-specific
+compiler for local assembly kernels. COFFEE manipulates abstract
+syntax trees by introducing domain-aware composable optimizations aimed at
+improving instruction-level parallelism, especially SIMD
+vectorization, and register locality. It then generates C code
+including vector intrinsics. Experiments using a range of
+finite-element forms of increasing complexity show that significant
+performance improvement is achieved.
 \end{abstract}
 
-\category{G.1.8}{Numerical Analysis}{Partial Differential Equations - Finite element methods}
-\category{G.4}{Mathematical Software}{Parallel and vector implementations}
+\category{G.1.8}{Numerical Analysis}{Partial Differential Equations -
+  Finite element methods}
+
+\category{G.4}{Mathematical Software}{Parallel and vector
+  implementations}
 
@@
 
-\keywords{Finite element integration, local assembly, compilers, optimizations, simd vectorization}
-
-\acmformat{Fabio Luporini, Ana Lucia Varbanescu, Florian Rathgeber, Gheorghe-Teodor Bercea, J.
-Ramanujam, David A.
-Ham, and Paul H.
-J.
-Kelly, 2014.
-COFFEE: an Optimizing Compilerfor Finite Element Local Assembly.}
+\keywords{Finite element integration, local assembly, compilers,
+  optimizations, SIMD vectorization}
+
+\acmformat{Fabio Luporini, Ana Lucia Varbanescu, Florian Rathgeber,
+  Gheorghe-Teodor Bercea, J. Ramanujam, David A. Ham, and Paul
+  H. J. Kelly, 2014. COFFEE: an Optimizing Compilerfor Finite Element
+  Local Assembly.}
+
 % At a minimum you need to supply the author names, year and a title.
@@
 \section{Introduction}
-In many fields, such as computational fluid dynamics, computational electromagnetics, and structural mechanics, phenomena are modelled by partial differential equations (PDEs).
-Numerical techniques, like the finite volume method and the finite element method, are widely employed to approximate solutions of these PDEs.
-Unstructured meshes are often used to discretize the equation domain, since their geometric flexibility allows solvers to be extremely effective.
-The solution is sought by applying suitable numerical operations, or kernels, to the components of a mesh, such as edges, vertices, or cells.
-On standard clusters of multi-cores, typically, a kernel is executed sequentially by a thread, while parallelism is achieved partitioning the mesh and assigning each partition to a different node or thread.
-Such execution model, with minor variations, is adopted, for example, in \cite{pyop2isc}, \cite{Fenics}, \cite{fluidity_manual_v4}, \cite{lizst}.
+In many fields, such as computational fluid dynamics, computational
+electromagnetics, and structural mechanics, phenomena are modelled by
+partial differential equations (PDEs). Numerical techniques, like the
+finite volume method and the finite element method, are widely
+employed to approximate solutions of these PDEs. Unstructured meshes
+are often used to discretize the equation domain, since their
+geometric flexibility allows solvers to be extremely effective. The
+solution is sought by applying suitable numerical operations, or
+kernels, to the components of a mesh, such as edges, vertices, or
+cells. On standard clusters of multi-cores, typically, a kernel is
+executed sequentially by a thread, while parallelism is achieved
+partitioning the mesh and assigning each partition to a different node
+or thread. Such execution model, with minor variations, is adopted,
+for example, in \cite{pyop2isc}, \cite{Fenics},
+\cite{fluidity_manual_v4}, \cite{lizst}.
 
@@
 
-The time required to apply the numerical kernels is a major issue, since the equation domain needs to be discretized into an extremely large number of cells, possibly order of trillions \cite{Rossinelli2013}, in order to obtain a satisfactory approximation of the PDE.
-For example, it has been well established that mesh resolution is critical in the accuracy of numerical weather forecasts.
-However, operational forecast centres have a strict time limit in which to produce a forecast - 60 minutes in the case of the UK Met Office - so producing efficient kernels has a direct scientific payoff in higher resolution, and therefore more accurate, forecasts.
-Hence, the computational cost is a dominant problem in computational science simulations, especially for those based on finite elements, which are the subject of this paper.
-We address, in particular, the well-known problem of optimizing the local assembly phase of the finite element method~\cite{Francis},~\cite{quadrature1},~\cite{petsc-integration-gpu},~\cite{Kirby-FEM-opt}, which can cover a significant fraction of the overall computation run-time, often in the range 30-60$\%$.
-With respect to these studies, we propose a novel set of composable code transformations targeting, for the first time, instruction-level parallelism - with emphasis on SIMD vectorization - and register locality.
-
-During the assembly phase, the solution of the PDE is approximated by executing a problem-specific kernel over all cells, or elements, in the discretized domain.
-In this work, we focus on relatively low order finite element methods, in which an assembly kernel's working set is usually small enough to fit the L1 cache.
-Low order methods are by no means exotic: they are employed in a wide variety of fields, including climate and ocean modeling, computational fluid dynamics, and structural mechanics.
-High order methods are often based upon special formulations of the finite element method, which result in the adoption of different sets of algorithms to solve PDEs~\cite{SPENCER}, and therefore are excluded from our study.
-
-An assembly kernel is characterized by the presence of an affine, often non-perfect loop nest, in which individual loops are rather small: their trip count rarely exceeds 30, and may be as low as 3 for low order methods.
-In the innermost loop, a problem-specific, flops-intensive expression evaluates a two dimensional array, representing the result of local assembly in an element of the discretized domain.
-With such a kernel structure, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization.
-
-Achieving high-performance is non-trivial.
-The complexity of the mathematical expressions, often characterized by a large number of operations on constants and small matrices, makes it hard to determine a single or specific sequence of transformations that is successfully applicable to all problems.
-The variation in loop trip counts and their typically small value further exacerbates the issue.
-A compiler-based approach is, therefore, the only reasonable option to obtain close-to-peak performance in a wide range of different local assembly kernels.
-Optimizations like padding, generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but none are supported by state-of-the-art polyhedral and vendor compilers.
-BLAS routines could theoretically be employed, although a fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies.
-In addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small dimension of the  matrices involved and the potential loss in data locality can limit or eliminate the performance gain of this approach.
-
-In order to overcome the constraints of the available compilers and specialized linear-algebra libraries, we have automated a set of generic and model-driven code transformations in COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for optimizing local assembly kernels.
-COFFEE is integrated with Firedrake~\cite{firedrake-code}, a system for solving PDEs through the finite element method based on the PyOP2 abstraction~\cite{pyop2isc}.
-It supports all problems expressible with this framework, including equations that can be found at the core of real-world simulations, such as those used in our performance evaluation.
-In evaluating our code transformations in a range of notable problems, we also vary two key parameters that impact solution accuracy and kernel cost: the polynomial order of the method (we investigate from $p=1$ to $p=4$) and the geometry of elements in the discretized domain (2D triangle, 3D tetrahedron, 3D prism).
-From the point of view of the generated code, these two parameters directly impact the size of both loop nests and mathematical expressions, as elaborated in Section~\ref{sec:real-examples}.
+The time required to apply the numerical kernels is a major issue,
+since the equation domain needs to be discretized into an extremely
+large number of cells, possibly order of trillions
+\cite{Rossinelli2013}, in order to obtain a satisfactory approximation
+of the PDE. For example, it has been well established that mesh
+resolution is critical in the accuracy of numerical weather
+forecasts. However, operational forecast centres have a strict time
+limit in which to produce a forecast - 60 minutes in the case of the
+UK Met Office - so producing efficient kernels has a direct scientific
+payoff in higher resolution, and therefore more accurate,
+forecasts. Hence, the computational cost is a dominant problem in
+computational science simulations, especially for those based on
+finite elements, which are the subject of this paper. We address, in
+particular, the well-known problem of optimizing the local assembly
+phase of the finite element
+method~\cite{Francis},~\cite{quadrature1},~\cite{petsc-integration-gpu},~\cite{Kirby-FEM-opt},
+which can cover a significant fraction of the overall computation
+run-time, often in the range 30-60$\%$. With respect to these studies,
+we propose a novel set of composable code transformations targeting,
+for the first time, instruction-level parallelism - with emphasis on
+SIMD vectorization - and register locality.
+
+During the assembly phase, the solution of the PDE is approximated by
+executing a problem-specific kernel over all cells, or elements, in
+the discretized domain. In this work, we focus on relatively low order
+finite element methods, in which an assembly kernel's working set is
+usually small enough to fit the L1 cache. Low order methods are by no
+means exotic: they are employed in a wide variety of fields, including
+climate and ocean modeling, computational fluid dynamics, and
+structural mechanics. High order methods are often based upon special
+formulations of the finite element method, which result in the
+adoption of different sets of algorithms to solve PDEs~\cite{SPENCER},
+and therefore are excluded from our study.
+
+An assembly kernel is characterized by the presence of an affine,
+often non-perfect loop nest, in which individual loops are rather
+small: their trip count rarely exceeds 30, and may be as low as 3 for
+low order methods. In the innermost loop, a problem-specific,
+flops-intensive expression evaluates a two dimensional array,
+representing the result of local assembly in an element of the
+discretized domain. With such a kernel structure, we focus on aspects
+like minimization of floating-point operations, register allocation
+and instruction-level parallelism, especially in the form of SIMD
+vectorization.
+
+Achieving high-performance is non-trivial. The complexity of the
+mathematical expressions, often characterized by a large number of
+operations on constants and small matrices, makes it hard to determine
+a single or specific sequence of transformations that is successfully
+applicable to all problems. The variation in loop trip counts and
+their typically small value further exacerbates the issue. A
+compiler-based approach is, therefore, the only reasonable option to
+obtain close-to-peak performance in a wide range of different local
+assembly kernels. Optimizations like padding, generalized
+loop-invariant code motion, vector-register tiling, and expression
+splitting, as well as their composition, are essential, but none are
+supported by state-of-the-art polyhedral and vendor compilers. BLAS
+routines could theoretically be employed, although a fairly
+complicated control- and data-flow analysis would be required to
+automate identification and extraction of matrix-matrix multiplies. In
+addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small
+dimension of the matrices involved and the potential loss in data
+locality can limit or eliminate the performance gain of this approach.
+
+In order to overcome the constraints of the available compilers and
+specialized linear-algebra libraries, we have automated a set of
+generic and model-driven code transformations in
+COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local
+  assembly.}, a compiler for optimizing local assembly kernels. COFFEE
+is integrated with Firedrake~\cite{firedrake-code}, a system for
+solving PDEs through the finite element method based on the PyOP2
+abstraction~\cite{pyop2isc}. It supports all problems expressible with
+this framework, including equations that can be found at the core of
+real-world simulations, such as those used in our performance
+evaluation. In evaluating our code transformations in a range of
+notable problems, we also vary two key parameters that impact solution
+accuracy and kernel cost: the polynomial order of the method (we
+investigate from $p=1$ to $p=4$) and the geometry of elements in the
+discretized domain (2D triangle, 3D tetrahedron, 3D prism). From the
+point of view of the generated code, these two parameters directly
+impact the size of both loop nests and mathematical expressions, as
+elaborated in Section~\ref{sec:real-examples}.
 
@@
 
-Our experiments show that the original generated code for non-trivial assembly kernels, despite following state-of-the-art techniques, remains suboptimal in the context of modern multi-core architectures.
-Our cost-model-driven sequence of code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 4.4$\times$ over original kernels, and around 70$\%$ of the test cases obtain a speed up greater than 2$\times$.
-The contributions of this paper, therefore, are
+Our experiments show that the original generated code for non-trivial
+assembly kernels, despite following state-of-the-art techniques,
+remains suboptimal in the context of modern multi-core
+architectures. Our cost-model-driven sequence of code transformations,
+aimed at improving SIMD vectorization and register data locality, can
+result in performance improvements up to 4.4$\times$ over original
+kernels, and around 70$\%$ of the test cases obtain a speed up greater
+than 2$\times$. The contributions of this paper, therefore, are
 \begin{enumerate}
-\item An optimization strategy for finite element local assembly. Our approach exploits domain knowledge and goes beyond the limits of both vendor and research compilers.
-\item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
-\item Systematic analysis using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
+\item An optimization strategy for finite element local assembly. Our
+  approach exploits domain knowledge and goes beyond the limits of
+  both vendor and research compilers.
+\item Design and implementation of a compiler that automates the
+  proposed code transformations for any problems expressible in
+  Firedrake.
+\item Systematic analysis using a suite of examples of real-world
+  importance, and evidence of significant performance improvements on
+  two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
 \end{enumerate}
 
-The paper is organized as follows.
-In Section~\ref{sec:background} we provide some background on local assembly, showing code generated by Firedrake and emphasizing the critical computational aspects.
-Section~\ref{sec:code-transf} describes the various code transformations, highlighting when and how domain-knowledge has been exploited.
-The design and implementation of our compiler is discussed in Section~\ref{sec:pyop2-compiler}.
-Section~\ref{sec:perf-results} shows performance results.
-Related work is illustrated in Section~\ref{sec:related-work}, while Section~\ref{sec:conclusions} reviews our contributions in the light of our results, and identifies priorities for future work.
+The paper is organized as follows. In Section~\ref{sec:background} we
+provide some background on local assembly, showing code generated by
+Firedrake and emphasizing the critical computational
+aspects. Section~\ref{sec:code-transf} describes the various code
+transformations, highlighting when and how domain-knowledge has been
+exploited. The design and implementation of our compiler is discussed
+in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results}
+shows performance results. Related work is illustrated in
+Section~\ref{sec:related-work}, while Section~\ref{sec:conclusions}
+reviews our contributions in the light of our results, and identifies
+priorities for future work.
 
@@
 
-Local assembly is the computation of contributions of a specific cell in the discretized domain to the PDE solution.
-The process consists of numerically evaluating problem-specific integrals to produce a matrix and a vector (\cite{quadrature1}, \cite{fluidity_manual_v4}), whose sizes depend on the order of the method.
-This operation is applied to all cells in the discretized domain.
-In this work we focus on local matrices, or ``element matrices'', which are more costly to compute than element vectors.
-
-Given a finite element description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl} (UFL), Firedrake generates a C-code kernel implementing assembly using a numerical integration algorithm.
-This kernel can be applied to any element in the mesh: this descends from the mathematical formulation of assembly, which states that the evaluation of a local matrix can be reduced to integrating on a fixed ``reference'' element - a special element that does not belong to the domain - after a suitable change of coordinates.
-Firedrake triggers the compilation of an assembly kernel using an available vendor compiler, and manages its (parallel) execution over all elements in the mesh.
-As already explained, the subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code.
-
-%The code transformations described in this work are also generalizable to non-Firedrake assembly kernels, provided that the same numerical integration algorithm is employed.
-
-The structure of a local assembly kernel is shown in Figure~\ref{code:general-structure}.
-The inputs are a zero-initialized two dimensional array used to store the element matrix, the element's coordinates in the discretized domain, and coefficient fields, for instance indicating the values of velocity or pressure in the element.
-The output is the evaluated element matrix.
-The kernel body can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse: required for the aforementioned change of coordinates, from the reference element to the one being computed; 2) definition of basis functions: used, intuitively, to interpolate the contribution to the PDE solution over the element.
-The choice of the basis functions is expressed in UFL directly by users.
-In the generated code, they are represented as global read-only two dimensional arrays (i.e.
-using \texttt{static const} in C) of double precision floats; 3) evaluation of the element matrix in an affine loop nest, in which the integration is performed.
-Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel objects.
-
-%JUNK SENTENCES: Since only the values assumed at the integration points are needed to compute the element matrix, a basis function is implemented as a two dimensional array of double precision floats.
-The same basis functions can be used to evaluate all element matrices, provided that a suitable change of coordinates from a ``reference'' element to a real one is performed.
-Therefore, basis functions are declared in the kernel body as global read-only arrays (i.e.
-using \texttt{static const} in C), rather than passed as function parameters.
-Basis functions are chosen at the UFL layer based on the problem being solved, so they are actually an input to the code generator.
+Local assembly is the computation of contributions of a specific cell
+in the discretized domain to the PDE solution. The process consists of
+numerically evaluating problem-specific integrals to produce a matrix
+and a vector (\cite{quadrature1}, \cite{fluidity_manual_v4}), whose
+sizes depend on the order of the method. This operation is applied to
+all cells in the discretized domain. In this work we focus on local
+matrices, or ``element matrices,'' which are more costly to compute
+than element vectors.
+
+Given a finite element description of the input problem, expressed
+through the domain-specific Unified Form Language~\cite{ufl} (UFL),
+Firedrake generates a C-code kernel implementing assembly using a
+numerical integration algorithm. This kernel can be applied to any
+element in the mesh: this descends from the mathematical formulation
+of assembly, which states that the evaluation of a local matrix can be
+reduced to integrating on a fixed ``reference'' element---a special
+element that does not belong to the domain---after a suitable change
+of coordinates. Firedrake triggers the compilation of an assembly
+kernel using an available vendor compiler, and manages its (parallel)
+execution over all elements in the mesh. As already explained, the
+subject of this paper is to enhance this execution model by adding an
+optimization stage prior to the generation of C code.
+
+%The code transformations described in this work are also
+%generalizable to non-Firedrake assembly kernels, provided that the
+%same numerical integration algorithm is employed.
+
+The structure of a local assembly kernel is shown in
+Figure~\ref{code:general-structure}. The inputs are a zero-initialized
+two dimensional array used to store the element matrix, the element's
+coordinates in the discretized domain, and coefficient fields, for
+instance indicating the values of velocity or pressure in the
+element. The output is the evaluated element matrix. The kernel body
+can be logically split into three parts: 1) calculation of the
+Jacobian matrix, its determinant and its inverse: required for the
+aforementioned change of coordinates, from the reference element to
+the one being computed; 2) definition of basis functions: used,
+intuitively, to interpolate the contribution to the PDE solution over
+the element. The choice of the basis functions is expressed in UFL
+directly by users. In the generated code, they are represented as
+global read-only two dimensional arrays (i.e., using \texttt{static
+  const} in C) of double precision floats; 3) evaluation of the
+element matrix in an affine loop nest, in which the integration is
+performed. Table~\ref{table:map-name-letters} shows the variable names
+we will use in the upcoming code snippets to refer to the various
+kernel objects.
+
+%JUNK SENTENCES: Since only the values assumed at the integration
+%points are needed to compute the element matrix, a basis function is
+%implemented as a two dimensional array of double precision
+%floats. The same basis functions can be used to evaluate all element
+%matrices, provided that a suitable change of coordinates from a
+%``reference'' element to a real one is performed. Therefore, basis
+%functions are declared in the kernel body as global read-only arrays
+%(i.e., using \texttt{static const} in C), rather than passed as
+%function parameters. Basis functions are chosen at the UFL layer
+%based on the problem being solved, so they are actually an input to
+%the code generator.
 
@@
 
-The actual complexity of a local assembly kernel depends on the finite element problem being solved.
-In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products involving basis functions.
-An example is provided in Listing~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$.
-In other scenarios, for instance when solving the Burgers equation, whose assembly code is given in Listing~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression.
-This may also require the evaluation of constants in outer loops (called $F$ in the code) to act as scaling factors of arrays; trip counts can be larger (proportionally to the order of the method); and arrays may be block-sparse.
-In addition to a larger number of operations, more complex cases like the Burgers equation are characterized by high register pressure.
-
-The Helmholtz and Burgers equations are exponents of a large class of problems and, as such, will constitute our benchmark, along with the Diffusion equation.
-We carefully motivate this choice in Section~\ref{sec:real-examples}.
-
-Despite the infinite variety of assembly kernels which Firedrake can generate, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.
-These include: 1) memory accesses along the three loop dimensions are always stride-1; 2) the \texttt{j} and \texttt{k} loops are interchangeable, whereas interchanges involving the $i$ loop require pre-computation of values (e.g.
-the $F$ values in Burgers) and introduction of temporary arrays, as explained in Section~\ref{sec:code-transf}; 3) depending on the problem being solved, the \texttt{j} and \texttt{k} loops could iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either \texttt{i}-\texttt{j} or \texttt{i}-\texttt{k}).
-In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.
+The actual complexity of a local assembly kernel depends on the finite
+element problem being solved. In simpler cases, the loop nest is
+perfect, it has short trip counts (in the range 3--15), and the
+computation reduces to a summation of a few products involving basis
+functions. An example is provided in Listing~\ref{code:helmholtz},
+which shows the assembly kernel for a Helmholtz problem using Lagrange
+basis functions on 2D elements with polynomial order $p=1$. In other
+scenarios, for instance when solving the Burgers equation, whose
+assembly code is given in Listing~\ref{code:burgers}, the number of
+arrays involved in the computation of the element matrix can be much
+larger: in this case, 14 unique arrays are accessed, and the same
+array can be referenced multiple times within the expression. This may
+also require the evaluation of constants in outer loops (called $F$ in
+the code) to act as scaling factors of arrays; trip counts can be
+larger (proportionally to the order of the method); and arrays may be
+block-sparse. In addition to a larger number of operations, more
+complex cases like the Burgers equation are characterized by high
+register pressure.
+
+The Helmholtz and Burgers equations are exponents of a large class of
+problems and, as such, will constitute our benchmark, along with the
+Diffusion equation. We carefully motivate this choice in
+Section~\ref{sec:real-examples}.
+
+Despite the infinite variety of assembly kernels which Firedrake can
+generate, it is still possible to identify common domain-specific
+traits that can be exploited for effective code transformations and
+SIMD vectorization. These include: 1) memory accesses along the three
+loop dimensions are always stride-1; 2) the \texttt{j} and \texttt{k}
+loops are interchangeable, whereas interchanges involving the $i$ loop
+require pre-computation of values (e.g. the $F$ values in Burgers) and
+introduction of temporary arrays, as explained in
+Section~\ref{sec:code-transf}; 3) depending on the problem being
+solved, the \texttt{j} and \texttt{k} loops could iterate over the
+same iteration space; 4) most of the sub-expressions on the right hand
+side of the element matrix computation depend on just two loops
+(either \texttt{i}-\texttt{j} or \texttt{i}-\texttt{k}). In
+Section~\ref{sec:code-transf} we show how to exploit these
+observations to define a set of systematic, composable optimizations.
 
@@
 \label{sec:code-transf}
-The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake.
-As already emphasized, the structure of mathematical expressions evaluating the element matrix and the variation in loop trip counts, although typically limited to the order of tens of iterations, render the optimization process challenging.
-It is not always the same set of optimizations that bring performance closest to the machine peak.
-For example, the Burgers problem in Listing~\ref{code:burgers}, given the large number of arrays accessed, suffers from high register pressure, whereas the Helmholtz problem in Listing~\ref{code:helmholtz} does not; this intuitively suggests that the two problems require a different treatment, based on an in-depth analysis of both data and iteration spaces.
-Furthermore, domain-knowledge enables transformations that a general-purpose compiler could not apply, making the optimization space even larger.
-In this context, our goal is to understand the relationship between distinct code transformations, their impact on local assembly kernels, and to what extent their composability is effective in a class of problems and architectures.
+The code transformations presented in this section are applicable to
+all finite element problems that can be formulated in Firedrake. As
+already emphasized, the structure of mathematical expressions
+evaluating the element matrix and the variation in loop trip counts,
+although typically limited to the order of tens of iterations, render
+the optimization process challenging. It is not always the same set of
+optimizations that bring performance closest to the machine peak. For
+example, the Burgers problem in Listing~\ref{code:burgers}, given the
+large number of arrays accessed, suffers from high register pressure,
+whereas the Helmholtz problem in Listing~\ref{code:helmholtz} does
+not; this intuitively suggests that the two problems require a
+different treatment, based on an in-depth analysis of both data and
+iteration spaces. Furthermore, domain-knowledge enables
+transformations that a general-purpose compiler could not apply,
+making the optimization space even larger. In this context, our goal
+is to understand the relationship between distinct code
+transformations, their impact on local assembly kernels, and to what
+extent their composability is effective in a class of problems and
+architectures.
 
@@
 \subsection{Padding and Data Alignment}
-The absence of stencils renders the element matrix computation easily auto-vectorizable by a vendor compiler.
-Nevertheless, auto-vectorization is not efficient if data are not aligned and if the length of the innermost loop is not a multiple of the vector length $vl$, especially when the loops are small as in local assembly.
-
-Data alignment is enforced in two steps.
-Firstly, all arrays are allocated to addresses that are multiples of $vl$.
-Then, two dimensional arrays are padded by rounding the number of columns to the nearest multiple of $vl$.
-For instance, assume the original size of a basis function array is 3$\times$3 and $vl=4$ (e.g.
-AVX processor, with 256 bit long vector registers and 64-bit double-precision floats).
-In this case, a padded version of the array will have size 3$\times$4.
-The compiler is explicitly informed about data alignment using a suitable pragma.
-
-Padding of all two dimensional arrays involved in the evaluation of the element matrix also allows to safely round the loop trip count to the nearest multiple of $vl$.
-This avoids the introduction of a remainder (scalar) loop from the compiler, which would render vectorization less efficient.
-These extra iterations only write to the padded region of the element matrix, and therefore have no side effects on the final result.
+The absence of stencils renders the element matrix computation easily
+auto-vectorizable by a vendor compiler. Nevertheless,
+auto-vectorization is not efficient if data are not aligned and if the
+length of the innermost loop is not a multiple of the vector length
+$vl$, especially when the loops are small as in local assembly.
+
+Data alignment is enforced in two steps. Firstly, all arrays are
+allocated to addresses that are multiples of $vl$. Then, two
+dimensional arrays are padded by rounding the number of columns to the
+nearest multiple of $vl$. For instance, assume the original size of a
+basis function array is 3$\times$3 and $vl=4$ (e.g. AVX processor,
+with 256 bit long vector registers and 64-bit double-precision
+floats). In this case, a padded version of the array will have size
+3$\times$4. The compiler is explicitly informed about data alignment
+using a suitable pragma.
+
+Padding of all two dimensional arrays involved in the evaluation of
+the element matrix also allows to safely round the loop trip count to
+the nearest multiple of $vl$. This avoids the introduction of a
+remainder (scalar) loop from the compiler, which would render
+vectorization less efficient. These extra iterations only write to the
+padded region of the element matrix, and therefore have no side
+effects on the final result.
 
@@
 \label{sec:code-transf-licm}
-From the inspection of the codes in Listings~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions which only depend on two iteration variables.
-Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion.
-Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels.
-We want to overcome two deficiencies that both $intel$ and $gnu$ compilers exhibit.
-First, they only identify sub-expressions that are invariant with respect to the innermost loop.
-This is an issue for sub-expressions depending on \texttt{i}-\texttt{k}, which are not automatically lifted in the loop order \texttt{ijk}.
-Second, the hoisted code is scalar, i.e.
-it is not subjected to auto-vectorization.
-
-We work around these limitations with source-level loop-invariant code motion.
-In particular, we pre-compute all values that an invariant sub-expression assumes along its fastest varying dimension.
-This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest.
-At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized as part of an inner loop.
-Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could otherwise be significant.
-It is also worth noting that, in some problems (e.g.
-Helmholtz), invariant sub-expressions along \texttt{j} are identical to those along \texttt{k}, and both loops iterate over the same iteration space, as anticipated in Section~\ref{sec:background}.
-In these cases, we safely avoid redundant pre-computation.
-
-Listing~\ref{code:helmholtz-licm} shows the Helmholtz assembly code after the application of loop-invariant code motion, padding, and data alignment.
+From the inspection of the codes in Listings~\ref{code:helmholtz}
+and~\ref{code:burgers}, it can be noticed that the computation of $A$
+involves evaluating many sub-expressions which only depend on two
+iteration variables. Since symbols in most of these sub-expressions
+are read-only variables, there is ample space for loop-invariant code
+motion. Vendor compilers apply this technique, although not in the
+systematic way we need for our assembly kernels. We want to overcome
+two deficiencies that both $intel$ and $gnu$ compilers exhibit. First,
+they only identify sub-expressions that are invariant with respect to
+the innermost loop. This is an issue for sub-expressions depending on
+\texttt{i}-\texttt{k}, which are not automatically lifted in the loop
+order \texttt{ijk}. Second, the hoisted code is scalar, i.e., it is not
+subjected to auto-vectorization.
+
+We work around these limitations with source-level loop-invariant code
+motion. In particular, we pre-compute all values that an invariant
+sub-expression assumes along its fastest varying dimension. This is
+implemented by introducing a temporary array per invariant
+sub-expression and by adding a new loop to the nest. At the price of
+extra memory for storing temporaries, the gain is that lifted terms
+can be auto-vectorized as part of an inner loop. Given the short trip
+counts of our loops, it is important to achieve auto-vectorization of
+hoisted terms in order to minimize the percentage of scalar
+instructions, which could otherwise be significant. It is also worth
+noting that, in some problems (e.g. Helmholtz), invariant
+sub-expressions along \texttt{j} are identical to those along
+\texttt{k}, and both loops iterate over the same iteration space, as
+anticipated in Section~\ref{sec:background}. In these cases, we safely
+avoid redundant pre-computation.
+
+Listing~\ref{code:helmholtz-licm} shows the Helmholtz assembly code
+after the application of loop-invariant code motion, padding, and data
+alignment.
 
@@
 $\rbrace$
-\caption{Local assembly code for the Helmholtz problem in Listing~\ref{code:helmholtz} after application of padding, data alignment, and \emph{licm}, for an AVX architecture.
-In this example, sub-expressions invariant to \texttt{j} are identical to those invariant to \texttt{k}, so they can be precomputed once in the $r$ loop.}
+\caption{Local assembly code for the Helmholtz problem in
+  Listing~\ref{code:helmholtz} after application of padding, data
+  alignment, and \emph{licm}, for an AVX architecture. In this
+  example, sub-expressions invariant to \texttt{j} are identical to
+  those invariant to \texttt{k}, so they can be precomputed once in
+  the $r$ loop.}
 \label{code:helmholtz-licm}
@@
 \subsection{Model-driven Vector-register Tiling}
-One notable problem of assembly kernels concerns register allocation and register locality.
-The critical situation occurs when loop trip counts and the variables accessed are such that the vector-register pressure is high.
-Since the kernel's working set fits the L1 cache, it is remarkably important to optimize register management.
-Standard optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem.
-In COFFEE, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll).
-Tiling at the level of vector registers is an additional feature of COFFEE.
-Based on the observation that the evaluation of the element matrix can be reduced to a summation of outer products along the \texttt{j} and \texttt{k} dimensions, a model-driven vector-register tiling strategy can be implemented.
-If we consider the code snippet in Listing~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, the computation of the element matrix is abstractly expressible as
+One notable problem of assembly kernels concerns register allocation
+and register locality. The critical situation occurs when loop trip
+counts and the variables accessed are such that the vector-register
+pressure is high. Since the kernel's working set fits the L1 cache, it
+is remarkably important to optimize register management. Standard
+optimizations, such as loop interchange, unroll, and unroll-and-jam,
+can be employed to deal with this problem. In COFFEE, these
+optimizations are supported either by means of explicit code
+transformations (interchange, unroll-and-jam) or indirectly by
+delegation to the compiler through standard pragmas (unroll). Tiling
+at the level of vector registers is an additional feature of
+COFFEE. Based on the observation that the evaluation of the element
+matrix can be reduced to a summation of outer products along the
+\texttt{j} and \texttt{k} dimensions, a model-driven vector-register
+tiling strategy can be implemented. If we consider the code snippet in
+Listing~\ref{code:helmholtz-licm} and we ignore the presence of the
+operation \texttt{det*W3[i]}, the computation of the element matrix is
+abstractly expressible as
 \begin{equation}
@@
 \end{equation}
-where $B$ is the set of all basis functions (or temporary variables, e.g.
-\texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets.
-Regardless of the specific input problem, by abstracting from the presence of all variables independent of both \texttt{j} and \texttt{k}, the element matrix computation is always reducible to this form.
-Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries ($j,k=0,...,3$) of the element matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$.
-Values in a register are shuffled each time a product is performed.
-Standard compiler auto-vectorization ($gnu$ and $intel$), instead, executes 4 broadcast operations (i.e.
-``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation.
-In addition to incurring in larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.
-
-%More importantly, we avoid using $vl$-1 registers per each outer-loop-dependent variable (i.e.
-all $x$ terms in Equation~\ref{outer-product}), with $vl$ being the vector length.
-This is a considerable gain, which allows us to slice the iteration space into bigger tiles, implemented directly by vector registers.
+where $B$ is the set of all basis functions (or temporary variables,
+e.g., \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$
+are generic problem-dependent subsets. Regardless of the specific
+input problem, by abstracting from the presence of all variables
+independent of both \texttt{j} and \texttt{k}, the element matrix
+computation is always reducible to this
+form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16
+entries ($j,k=0,...,3$) of the element matrix using just 2 vector
+registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert
+= \vert B'' \vert = 1$. Values in a register are shuffled each time a
+product is performed. Standard compiler auto-vectorization ($gnu$ and
+$intel$), instead, executes 4 broadcast operations (i.e., ``splat'' of
+a value over all of the register locations) along the outer dimension
+to perform the calculation. In addition to incurring in larger number
+of cache accesses, it needs to keep between $f=1$ and $f=3$ extra
+registers to perform the same 16 evaluations when unroll-and-jam is
+used, with $f$ being the unroll-and-jam factor.
+
+%More importantly, we avoid using $vl$-1 registers per each
+%outer-loop-dependent variable (i.e., all $x$ terms in
+%Equation~\ref{outer-product}), with $vl$ being the vector
+%length. This is a considerable gain, which allows us to slice the
+%iteration space into bigger tiles, implemented directly by vector
+%registers.
 
@@
 \centerline{\includegraphics[scale=0.8]{Pictures/vect-by-vect-inline.pdf}}
-\caption{Outer-product vectorization by permuting values in a vector register.}
+\caption{Outer-product vectorization by permuting values in a vector
+  register.}
 \label{fig:vect-by-vect}
@@
 
-The storage layout of $A$, however, is incorrect after the application of this outer-product-based vectorization (\emph{op-vect}, in the following).
-It can be efficiently restored with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the \texttt{ijk} loop nest.
-The generated pseudo-code for the simple Helmholtz problem when using \emph{op-vect} is shown in Figure~\ref{code:helmholtz-opvect}.
+The storage layout of $A$, however, is incorrect after the application
+of this outer-product-based vectorization (\emph{op-vect}, in the
+following). It can be efficiently restored with a sequence of vector
+shuffles following the pattern highlighted in
+Figure~\ref{fig:restore-layout}, executed once outside of the
+\texttt{ijk} loop nest. The generated pseudo-code for the simple
+Helmholtz problem when using \emph{op-vect} is shown in
+Figure~\ref{code:helmholtz-opvect}.
 
@@
 \centerline{\includegraphics[scale=0.8]{Pictures/vect-restore-inline.pdf}}
-\caption{Restoring the storage layout after \emph{op-vect}.
-The figure shows how 4$\times$4 elements in the top-left block of the element matrix $A$ can be moved to their correct positions.
-Each rotation, represented by a group of three same-colored arrows, is implemented by a single shuffle intrinsic.}
+\caption{Restoring the storage layout after \emph{op-vect}. The figure
+  shows how 4$\times$4 elements in the top-left block of the element
+  matrix $A$ can be moved to their correct positions. Each rotation,
+  represented by a group of three same-colored arrows, is implemented
+  by a single shuffle intrinsic.}
 \label{fig:restore-layout}
@@
 
-%are sufficiently large to prevent register reuse along the innermost dimension \texttt{k}.
-A combination of the two scenarios is also possible.
-We want to exploit the small size of the iteration space and the structure of the computation to optimize register management.
+%are sufficiently large to prevent register reuse along the innermost
+%dimension \texttt{k}. A combination of the two scenarios is also
+%possible. We want to exploit the small size of the iteration space
+%and the structure of the computation to optimize register management.
 
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake for the Helmholtz problem after application of \emph{op-vect} on top of the optimizations shown in Listing~\ref{code:helmholtz-licm}.
-In this example, the unroll-and-jam factor is 1.}
+\caption{Local assembly code generated by Firedrake for the Helmholtz
+  problem after application of \emph{op-vect} on top of the
+  optimizations shown in Listing~\ref{code:helmholtz-licm}. In this
+  example, the unroll-and-jam factor is 1.}
 \label{code:helmholtz-opvect}
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake for the Helmholtz problem in which \emph{split} has been applied on top of the optimizations shown in Listing~\ref{code:helmholtz-licm}.
-In this example, the split factor is 2.}
+\caption{Local assembly code generated by Firedrake for the Helmholtz
+  problem in which \emph{split} has been applied on top of the
+  optimizations shown in Listing~\ref{code:helmholtz-licm}. In this
+  example, the split factor is 2.}
 \label{code:helmholtz-split}
@@
 
-In complex kernels, like Burgers in Listing~\ref{code:burgers}, and on certain architectures, achieving effective register allocation can be challenging.
-If the number of variables independent of the innermost-loop dimension is close to or greater than the number of available CPU registers, poor register reuse is likely to be obtained.
-This usually happens when the number of basis function arrays, temporaries introduced by generalized loop-invariant code motion, and problem constants is large.
-For example, applying loop-invariant code motion to Burgers on a 3D mesh requires 24 temporaries for the \texttt{ijk} loop order.
-This can make hoisting of the invariant loads out of the \texttt{k} loop inefficient on architectures with a relatively low number of registers.
-One potential solution to this problem consists of suitably ``splitting'' the computation of the element matrix $A$ into multiple sub-expressions; an example, for the Helmholtz problem, is given in Listing~\ref{code:helmholtz-split}.
-The transformation can be regarded as a special case of classic loop fission, in which sum's associativity is exploited to distribute the expression across multiple loops.
-To the best of our knowledge, expression splitting is not supported by available compilers.
-
-Splitting an expression has, however, several drawbacks.
-Firstly, it increases the number of accesses to $A$ proportionally to the ``split factor'', which is the number of sub-expressions produced.
-Also, depending on how the split is executed, it can lead to redundant computation; for example, the times the product $det*W3[i]$ is performed is proportional to the number of sub-expressions, as shown in the code snippet.
-Further, it increases loop overhead (e.g.
-branch instructions).
-Finally, it might affect register locality: for instance, the same array could be accessed in different sub-expressions, requiring a proportional number of loads be performed.
-This is not the case of the Helmholtz example.
-Nevertheless, as shown in Section~\ref{sec:perf-results}, the performance gain from improved register reuse along inner dimensions can still be greater, especially if the split factor and the splitting itself use heuristics to minimize the aforementioned issues.
+In complex kernels, like Burgers in Listing~\ref{code:burgers}, and on
+certain architectures, achieving effective register allocation can be
+challenging. If the number of variables independent of the
+innermost-loop dimension is close to or greater than the number of
+available CPU registers, poor register reuse is likely to be
+obtained. This usually happens when the number of basis function
+arrays, temporaries introduced by generalized loop-invariant code
+motion, and problem constants is large. For example, applying
+loop-invariant code motion to Burgers on a 3D mesh requires 24
+temporaries for the \texttt{ijk} loop order. This can make hoisting of
+the invariant loads out of the \texttt{k} loop inefficient on
+architectures with a relatively low number of registers. One potential
+solution to this problem consists of suitably ``splitting'' the
+computation of the element matrix $A$ into multiple sub-expressions;
+an example, for the Helmholtz problem, is given in
+Listing~\ref{code:helmholtz-split}. The transformation can be regarded
+as a special case of classic loop fission, in which sum's
+associativity is exploited to distribute the expression across
+multiple loops. To the best of our knowledge, expression splitting is
+not supported by available compilers.
+
+Splitting an expression has, however, several drawbacks. Firstly, it
+increases the number of accesses to $A$ proportionally to the ``split
+factor'', which is the number of sub-expressions produced. Also,
+depending on how the split is executed, it can lead to redundant
+computation; for example, the times the product $det*W3[i]$ is
+performed is proportional to the number of sub-expressions, as shown
+in the code snippet. Further, it increases loop overhead (e.g. branch
+instructions). Finally, it might affect register locality: for
+instance, the same array could be accessed in different
+sub-expressions, requiring a proportional number of loads be
+performed. This is not the case of the Helmholtz
+example. Nevertheless, as shown in Section~\ref{sec:perf-results}, the
+performance gain from improved register reuse along inner dimensions
+can still be greater, especially if the split factor and the splitting
+itself use heuristics to minimize the aforementioned issues.
 
@@
 
-Table~\ref{table:code-transformations} summarizes the code transformations described so far.
-Given that many of these transformations depend on some parameters (e.g.
-tile size), we need a mechanism to prune such a large space of optimization.
-This aspect is treated in Section~\ref{sec:pyop2-compiler}.
+Table~\ref{table:code-transformations} summarizes the code
+transformations described so far. Given that many of these
+transformations depend on some parameters (e.g. tile size), we need a
+mechanism to prune such a large space of optimization. This aspect is
+treated in Section~\ref{sec:pyop2-compiler}.
 
@@
 \label{sec:pyop2-compiler}
-%Design and structure of the PyOP2 Compiler.
-Show the steps through which the IR is transformed (need to cite Firedrake here).
-Briefly describe a simple cost model that allows us to prune the space of code transformations.
+%Design and structure of the PyOP2 Compiler. Show the steps through
+%which the IR is transformed (need to cite Firedrake here). Briefly
+%describe a simple cost model that allows us to prune the space of
+%code transformations.
 
@@
 \includegraphics[scale=0.75]{Pictures/coffee-pipeline.pdf}
-\caption{High-level view of Firedrake. COFFEE is at the core, receiving ASTs from a modified version of the FEniCS Form compiler and producing optimized C code kernels.}
+\caption{High-level view of Firedrake. COFFEE is at the core,
+  receiving ASTs from a modified version of the FEniCS Form compiler
+  and producing optimized C code kernels.}
 \label{fig:coffee-pipeline}
@@
 
-Firedrake users employ the Unified Form Language to write problems in a notation resembling mathematical equations.
-At run-time, the high-level specification is translated by a modified version of the FEniCS Form Compiler~\cite{FFC-Compiler} (FFC) into an abstract syntax tree (AST) representation of one or more finite element assembly kernels.
-ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}.
-The output of COFFEE, C code, is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and execution over the discretized domain take place.
-The flow is summarized in Figure~\ref{fig:coffee-pipeline}.
-
-Because of the large number of parametric transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem.
-Auto-tuning might be used, although it would significantly increase the run-time overhead, since the generation of ASTs occurs at run-time as soon as problem-specific data are available.
-Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.
-
-The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}.
-Initially the AST is inspected, looking for the presence of iteration spaces and domain-specific information provided by the higher layer.
-If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the outer, non-affine loop over mesh elements is vectorized, can be applied.
-This feature, currently under development, has been shown to be useful in several finite volume applications~\cite{inter-kernel-vect}.
-Subsequently, an ordered sequence of optimization steps are executed.
-Application of \emph{licm} must precede padding and data alignment, due to the introduction of temporary arrays.
-Based on a cost model, loop interchange, \emph{split}, and \emph{op-vect} may be introduced.
-Their implementation is based on analysis and transformation of the kernel's AST.
-When \emph{op-vect} is selected, the compiler outputs AVX or AVX-512 intrinsics code.
-Any possible corner cases are handled: for example, if \emph{op-vect} is to be applied, but the size of the iteration space is not a multiple of the vector length, then a remainder loop, amenable to auto-vectorization, is inserted.
+Firedrake users employ the Unified Form Language to write problems in
+a notation resembling mathematical equations. At run-time, the
+high-level specification is translated by a modified version of the
+FEniCS Form Compiler~\cite{FFC-Compiler} (FFC) into an abstract syntax
+tree (AST) representation of one or more finite element assembly
+kernels. ASTs are then passed to COFFEE, which is capable of applying
+the transformations described in Section~\ref{sec:code-transf}. The
+output of COFFEE, C code, is eventually provided to
+PyOP2~\cite{pyop2isc}, where just-in-time compilation and execution
+over the discretized domain take place. The flow is summarized in
+Figure~\ref{fig:coffee-pipeline}.
+
+Because of the large number of parametric transformations, COFFEE
+needs a mechanism to select the most suitable optimization strategy
+for a given problem. Auto-tuning might be used, although it would
+significantly increase the run-time overhead, since the generation of
+ASTs occurs at run-time as soon as problem-specific data are
+available. Our optimization strategy, based on heuristics and a simple
+cost model, is described in the following, along with an overview of
+the compiler.
+
+The compiler structure is outlined in
+Figure~\ref{algo:PyOP2Compiler}. Initially the AST is inspected,
+looking for the presence of iteration spaces and domain-specific
+information provided by the higher layer. If the kernel lacks an
+iteration space, then so-called inter-kernel vectorization, in which
+the outer, non-affine loop over mesh elements is vectorized, can be
+applied. This feature, currently under development, has been shown to
+be useful in several finite volume
+applications~\cite{inter-kernel-vect}. Subsequently, an ordered
+sequence of optimization steps are executed. Application of
+\emph{licm} must precede padding and data alignment, due to the
+introduction of temporary arrays. Based on a cost model, loop
+interchange, \emph{split}, and \emph{op-vect} may be introduced. Their
+implementation is based on analysis and transformation of the kernel's
+AST. When \emph{op-vect} is selected, the compiler outputs AVX or
+AVX-512 intrinsics code. Any possible corner cases are handled: for
+example, if \emph{op-vect} is to be applied, but the size of the
+iteration space is not a multiple of the vector length, then a
+remainder loop, amenable to auto-vectorization, is inserted.
 
@@
 \footnotesize
-\textbf{Input:} ast: abstract syntax tree of a local assembly kernel produced by FFC
-       wrapper: the function within which the kernel is invoked
-	       isa: represents the SIMD Instruction Set Architecture
+\textbf{Input:} ast: abstract syntax tree of a local assembly kernel
+produced by FFC wrapper: the function within which the kernel is
+invoked isa: represents the SIMD Instruction Set Architecture
 \textbf{Output:} assembly code (C string)
@@
 
-All loops are interchangeable, provided that temporaries are introduced if the nest is not perfect.
-For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance.
-Conceptually, this is motivated by the fact that if the \texttt{i} loop were in an inner position, then a significantly higher number of load instructions would be required at every iteration.
-We tested this hypothesis in manually crafted kernels.
-We found that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than memory, along the \texttt{i} loop.
-The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension.
-Our compiler chooses as outermost the loop along which the number of invariant loads is smaller so that more registers remain available to carry out the computation of the element matrix.
-
-% "vendor compilers such as Intel's" is awkward and unclear. If Intel is the only one you have looked at, then say "the Intel compiler", otherwise name the ones for which you know this to be true.
-Loop unroll  (or unroll-and-jam of outer loops) is fundamental to the exposure of instruction-level parallelism, and tuning the unroll factor is particularly important.
-However, we noticed, by comparison with implementations having manually-unrolled loops, that recent versions of the Intel compiler estimate close-to-optimal unroll factors when the loops are affine and their bounds are small and known at compile-time, which is the case of our kernels.
-We therefore leave the backend compiler in charge of selecting unroll factors.
-This choice also simplifies the COFFEE's cost model.
-The only situation in which we explicitly unroll-and-jam a loop is when \emph{op-vect} is used, since the transformed code prevents the Intel compiler from applying this optimization, even if specific pragmas are added.
-
-%Note that, regardless of the output of the cost model, the unroll-and-jam factor never exceeds the actual size of the outer loop (line 19 in Figure~\ref{algo:PyOP2Compiler}, assuming the default loop order \texttt{ijk}).
+All loops are interchangeable, provided that temporaries are
+introduced if the nest is not perfect. For the employed storage
+layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely
+to maximize performance. Conceptually, this is motivated by the fact
+that if the \texttt{i} loop were in an inner position, then a
+significantly higher number of load instructions would be required at
+every iteration. We tested this hypothesis in manually crafted
+kernels. We found that the performance loss is greater than the gain
+due to the possibility of accumulating increments in a register,
+rather than memory, along the \texttt{i} loop. The choice between
+\texttt{ijk} and \texttt{ikj} depends on the number of load
+instructions that can be hoisted out of the innermost dimension. Our
+compiler chooses as outermost the loop along which the number of
+invariant loads is smaller so that more registers remain available to
+carry out the computation of the element matrix.
+
+% "vendor compilers such as Intel's" is awkward and unclear. If Intel
+% is the only one you have looked at, then say "the Intel compiler",
+% otherwise name the ones for which you know this to be true.
+Loop unroll (or unroll-and-jam of outer loops) is fundamental to the
+exposure of instruction-level parallelism, and tuning the unroll
+factor is particularly important. However, we noticed, by comparison
+with implementations having manually-unrolled loops, that recent
+versions of the Intel compiler estimate close-to-optimal unroll
+factors when the loops are affine and their bounds are small and known
+at compile-time, which is the case of our kernels. We therefore leave
+the backend compiler in charge of selecting unroll factors. This
+choice also simplifies the COFFEE's cost model. The only situation in
+which we explicitly unroll-and-jam a loop is when \emph{op-vect} is
+used, since the transformed code prevents the Intel compiler from
+applying this optimization, even if specific pragmas are added.
+
+%Note that, regardless of the output of the cost model, the
+%unroll-and-jam factor never exceeds the actual size of the outer loop
+%(line 19 in Figure~\ref{algo:PyOP2Compiler}, assuming the default
+%loop order \texttt{ijk}).
 
@@
 \end{alltt}
-\caption{The cost model is employed by the compiler to estimate the most suitable unroll-and-jam (when \emph{op-vect} is used) and split factors, avoiding the overhead of auto-tuning.}
+\caption{The cost model is employed by the compiler to estimate the
+  most suitable unroll-and-jam (when \emph{op-vect} is used) and split
+  factors, avoiding the overhead of auto-tuning.}
 \label{algo:applyCostModel}
@@
 
-The cost model is shown in Figure~\ref{algo:applyCostModel}.
-It takes into account the number of available logical vector registers, \emph{n\_regs}, and the number of unique variables accessed: \emph{n\_consts} counts variables independent of both \texttt{j} and \texttt{k} loops and temporary registers, \emph{n\_outer\_arrays} counts \texttt{j}-dependent variables, and \emph{n\_inner\_arrays} counts \texttt{k}-dependent variables, assuming the \texttt{ijk} loop order.
-These values are used to estimate unroll-and-jam and split factors for \emph{op-vect} and \emph{split}.
-If a factor is 0, then the corresponding transformation is not applied.
-The \emph{split} transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 3).
-A split factor of $n$ means that the assembly expression should be ``cut'' into $n$ sub-expressions.
-Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size.
-The unroll-and-jam factor for the \emph{op-vect} transformation is determined as a function of the available logical registers, i.e.
-those not used for storing hoisted terms (line 9-11).
-Finally, the profitability of loop interchange is evaluated (line 13).
+The cost model is shown in Figure~\ref{algo:applyCostModel}. It takes
+into account the number of available logical vector registers,
+\emph{n\_regs}, and the number of unique variables accessed:
+\emph{n\_consts} counts variables independent of both \texttt{j} and
+\texttt{k} loops and temporary registers, \emph{n\_outer\_arrays}
+counts \texttt{j}-dependent variables, and \emph{n\_inner\_arrays}
+counts \texttt{k}-dependent variables, assuming the \texttt{ijk} loop
+order. These values are used to estimate unroll-and-jam and split
+factors for \emph{op-vect} and \emph{split}. If a factor is 0, then
+the corresponding transformation is not applied. The \emph{split}
+transformation is triggered whenever the number of hoistable terms is
+larger than the available registers along the outer dimension (lines
+3-8), which is approximated as half of the total (line 3). A split
+factor of $n$ means that the assembly expression should be ``cut''
+into $n$ sub-expressions. Depending on the structure of the assembly
+expression, each sub-expression might end up accessing a different
+number of arrays; the cost model is simplified by assuming that all
+sub-expressions are of the same size. The unroll-and-jam factor for
+the \emph{op-vect} transformation is determined as a function of the
+available logical registers, i.e., those not used for storing hoisted
+terms (line 9-11). Finally, the profitability of loop interchange is
+evaluated (line 13).
 
@@
 
-Experiments were run on a single core of two Intel architectures, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and a Xeon Phi (5110P, running at 1.05Ghz in native mode, 32KB L1 cache and 512KB L2 cache).
-We have chosen these two architectures because of the differences in the number of logical registers and SIMD lanes, which can impact the effectiveness of the optimization strategy.
-The \texttt{icc 13.1} compiler was used.
-On the Sandy Bridge, the compilation flags used were \texttt{-O2} and \texttt{-xAVX} for auto-vectorization.
-On the Xeon Phi, optimization level \texttt{-O3} was used.
-Other optimization levels performed, in general, slightly worse.
+Experiments were run on a single core of two Intel architectures, a
+Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB
+L2 cache) and a Xeon Phi (5110P, running at 1.05Ghz in native mode,
+32KB L1 cache and 512KB L2 cache). We have chosen these two
+architectures because of the differences in the number of logical
+registers and SIMD lanes, which can impact the effectiveness of the
+optimization strategy. The \texttt{icc 13.1} compiler was used. On the
+Sandy Bridge, the compilation flags used were \texttt{-O2} and
+\texttt{-xAVX} for auto-vectorization. On the Xeon Phi, optimization
+level \texttt{-O3} was used. Other optimization levels performed, in
+general, slightly worse.
 
@@
 \label{sec:real-examples}
-Our code transformations were evaluated in three real-world problems based on the following PDEs
+Our code transformations were evaluated in three real-world problems
+based on the following PDEs
 \begin{enumerate}
@@
 \end{enumerate}
-The three chosen benchmarks are \emph{real-life kernels} and comprise the core differential operators in some of the most frequently encountered finite element problems in scientific computing.
-The Helmholtz and Diffusion kernels are archetypal second order elliptic operators.
-They are complete and unsimplified examples of the operators occurring for diffusion and viscosity in fluids, and for imposing pressure in compressible fluids.
-As such, they are both extensively used in climate and ocean modeling.
-Very similar operators, for which the same optimisations are expected to be equally effective, apply to elasticity problems, which are at the base of computational structural mechanics.
-The Burgers kernel is a typical example of a first order hyperbolic conservation law, which occurs in real applications whenever a quantity is transported by a fluid (the momentum itself, in our case).
-We chose this particular kernel since it applies to a vector-valued quantity, while the elliptic operators apply to scalar quantities; this impacts the generated code, as explained next.
-In this way, the operators we have selected are characteristic of both the second and first order operators that dominate in fluids and solids simulations, and a wide variety on local assembly codes can be tested.
-
-The benchmarks were written in UFL (code available at~\cite{ufl-code}) and executed over real unstructured meshes through Firedrake.
-The Helmholtz code has already been shown in Listing~\ref{code:helmholtz}.
-For Advection-Diffusion, the Diffusion equation, which uses the same differential operators as Helmholtz, is considered.
-In the Diffusion kernel code, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix.
-Burgers is a non-linear problem employing differential operators different from those of Helmholtz and relying on vector-valued quantities, which has a major impact on the generated assembly code (see Listing~\ref{code:burgers}), where a larger number of basis function arrays ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are generated.
-
-These problems were studied varying both the shape of mesh elements and the polynomial order $p$ of the method, whereas the element family, Lagrange, is fixed.
-As might be expected, the larger the element shape and $p$, the larger the iteration space.
-Triangles, tetrahedron, and prisms were tested as element shape.
-For instance, in the case of Helmholtz with $p=1$, the size of the \texttt{j} and \texttt{k} loops for the three element shapes is, respectively, $3$, $4$, and $6$.
-Moving to bigger shapes has the effect of increasing the number of basis function arrays, since, intuitively, the behavior of the equation has now to be approximated also along a third axis.
-On the other hand, the polynomial order affects only the problem size (the three loops \texttt{i}, \texttt{j}, and \texttt{k}, and, as a consequence, the size of $X$ and $Y$ arrays).
-A range of polynomial orders from $p=1$ to $p=4$ were tested; higher polynomial orders are excluded from the study because of current Firedrake limitations.
-In all these cases, the size of the element matrix rarely exceeds 30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and $p=4$.
+The three chosen benchmarks are \emph{real-life kernels} and comprise
+the core differential operators in some of the most frequently
+encountered finite element problems in scientific computing. The
+Helmholtz and Diffusion kernels are archetypal second order elliptic
+operators. They are complete and unsimplified examples of the
+operators occurring for diffusion and viscosity in fluids, and for
+imposing pressure in compressible fluids. As such, they are both
+extensively used in climate and ocean modeling. Very similar
+operators, for which the same optimisations are expected to be equally
+effective, apply to elasticity problems, which are at the base of
+computational structural mechanics. The Burgers kernel is a typical
+example of a first order hyperbolic conservation law, which occurs in
+real applications whenever a quantity is transported by a fluid (the
+momentum itself, in our case). We chose this particular kernel since
+it applies to a vector-valued quantity, while the elliptic operators
+apply to scalar quantities; this impacts the generated code, as
+explained next. In this way, the operators we have selected are
+characteristic of both the second and first order operators that
+dominate in fluids and solids simulations, and a wide variety on local
+assembly codes can be tested.
+
+The benchmarks were written in UFL (code available at~\cite{ufl-code})
+and executed over real unstructured meshes through Firedrake. The
+Helmholtz code has already been shown in
+Listing~\ref{code:helmholtz}. For Advection-Diffusion, the Diffusion
+equation, which uses the same differential operators as Helmholtz, is
+considered. In the Diffusion kernel code, the main differences with
+respect to Helmholtz are the absence of the $Y$ array and the presence
+of a few more constants for computing the element matrix. Burgers is a
+non-linear problem employing differential operators different from
+those of Helmholtz and relying on vector-valued quantities, which has
+a major impact on the generated assembly code (see
+Listing~\ref{code:burgers}), where a larger number of basis function
+arrays ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$,
+...) are generated.
+
+These problems were studied varying both the shape of mesh elements
+and the polynomial order $p$ of the method, whereas the element
+family, Lagrange, is fixed. As might be expected, the larger the
+element shape and $p$, the larger the iteration space. Triangles,
+tetrahedron, and prisms were tested as element shape. For instance, in
+the case of Helmholtz with $p=1$, the size of the \texttt{j} and
+\texttt{k} loops for the three element shapes is, respectively, $3$,
+$4$, and $6$. Moving to bigger shapes has the effect of increasing the
+number of basis function arrays, since, intuitively, the behavior of
+the equation has now to be approximated also along a third axis. On
+the other hand, the polynomial order affects only the problem size
+(the three loops \texttt{i}, \texttt{j}, and \texttt{k}, and, as a
+consequence, the size of $X$ and $Y$ arrays). A range of polynomial
+orders from $p=1$ to $p=4$ were tested; higher polynomial orders are
+excluded from the study because of current Firedrake limitations. In
+all these cases, the size of the element matrix rarely exceeds
+30$\times$30, with a peak of 105$\times$105 in Burgers with prisms and
+$p=4$.
 
 \subsection{Loop permutation}
-In the following, only results for the loop order \texttt{ijk} are shown.
-For the considerations exposed in Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner loop along \texttt{i} caused slow downs; also, interchanging \texttt{j} and \texttt{k} loops while keeping \texttt{i} as outermost loop did not provide any benefits.
+In the following, only results for the loop order \texttt{ijk} are
+shown. For the considerations exposed in
+Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner
+loop along \texttt{i} caused slow downs; also, interchanging
+\texttt{j} and \texttt{k} loops while keeping \texttt{i} as outermost
+loop did not provide any benefits.
 
@@
 \begin{table*}[t]
-\tbl{Performance improvement due to generalized loop-invariant code motion, for different element shapes (triangle, tetrahedron, prism) and polynomial orders ($p \in [1, 4]$), over the original non-optimized code, for the Helmholtz, Diffusion and Burgers problems.}{
-\scriptsize
-\licmresultsnorms
-}
+\tbl{Performance improvement due to generalized loop-invariant code
+  motion, for different element shapes (triangle, tetrahedron, prism)
+  and polynomial orders ($p \in [1, 4]$), over the original
+  non-optimized code, for the Helmholtz, Diffusion and Burgers
+  problems.}{ \scriptsize \licmresultsnorms }
 \label{table:perf-results-licm}
 \end{table*}
-Table~\ref{table:perf-results-licm} illustrates the performance improvement obtained when \emph{licm} is applied.
-In general, the speed ups are notable.
-The main reasons were anticipated in Section~\ref{sec:code-transf-licm}: in the original code, 1) sub-expressions invariant to outer loops are not automatically hoisted, while 2) sub-expressions invariant to the innermost loop are hoisted, but their execution is not auto-vectorized.
-These observations come from inspection of assembly code generated by the compiler.
-
-The gain tends to grow with the computational cost of the kernels: bigger loop nests (i.e.
-larger element shapes and polynomial orders) usually benefit from the reduction in redundant computation, even though extra memory for the temporary arrays is required.
-Some discrepancies to this trend are due to a less effective auto-vectorization.
-For instance, on the Sandy Bridge, the improvement at $p=3$ is larger than that at $p=4$ because, in the latter case, the size of the innermost loop is not a multiple of the vector length, and a reminder scalar loop is introduced at compile time.
-Since the loop nest is small, the cost of executing the extra scalar iterations can have a significant impact.
-The reminder loop overhead is more pronounced on the Xeon Phi, where the vector length is twice as big, which leads to proportionally larger scalar reminder loops.
+Table~\ref{table:perf-results-licm} illustrates the performance
+improvement obtained when \emph{licm} is applied. In general, the
+speed ups are notable. The main reasons were anticipated in
+Section~\ref{sec:code-transf-licm}: in the original code, 1)
+sub-expressions invariant to outer loops are not automatically
+hoisted, while 2) sub-expressions invariant to the innermost loop are
+hoisted, but their execution is not auto-vectorized. These
+observations come from inspection of assembly code generated by the
+compiler.
+
+The gain tends to grow with the computational cost of the kernels:
+bigger loop nests (i.e., larger element shapes and polynomial orders)
+usually benefit from the reduction in redundant computation, even
+though extra memory for the temporary arrays is required. Some
+discrepancies to this trend are due to a less effective
+auto-vectorization. For instance, on the Sandy Bridge, the improvement
+at $p=3$ is larger than that at $p=4$ because, in the latter case, the
+size of the innermost loop is not a multiple of the vector length, and
+a reminder scalar loop is introduced at compile time. Since the loop
+nest is small, the cost of executing the extra scalar iterations can
+have a significant impact. The reminder loop overhead is more
+pronounced on the Xeon Phi, where the vector length is twice as big,
+which leads to proportionally larger scalar reminder loops.
 
@@
 \begin{table*}[t]
-\tbl{Performance improvement due to generalized loop-invariant code motion, data alignment, and padding, for different element shapes (triangle, tetrahedron, prism) and polynomial orders ($p \in [1, 4]$), over the original non-optimized code, for the Helmholtz, Diffusion and Burgers problems.}{
-\scriptsize
-\licmapresultsnorms
-}
+\tbl{Performance improvement due to generalized loop-invariant code
+  motion, data alignment, and padding, for different element shapes
+  (triangle, tetrahedron, prism) and polynomial orders ($p \in [1,
+    4]$), over the original non-optimized code, for the Helmholtz,
+  Diffusion and Burgers problems.}{ \scriptsize \licmapresultsnorms }
 \label{table:perf-results-licmap}
@@
 
-Table~\ref{table:perf-results-licmap} shows the cumulative impact of \emph{licm}, data alignment, and padding over the original code.
-In the following, this version of the code is referred to as \emph{licm-ap}.
-Padding, which avoids the introduction of a reminder loop as described in Section~\ref{licm-impact}, as well as data alignment, enhance the quality of auto-vectorization.
-Occasionally the run-time of \emph{licm-ap} is close to that of \emph{licm}, since the non-padded element matrix size is already a multiple of the vector length.
-Rarely \emph{licm-ap} is slower than \emph{licm} (e.g.
-in Burgers $p=3$ on the Sandy Bridge).
-One possible explanation is that the number of aligned temporaries introduced by \emph{licm} is so large to induce cache associativity conflicts.
+Table~\ref{table:perf-results-licmap} shows the cumulative impact of
+\emph{licm}, data alignment, and padding over the original code. In
+the following, this version of the code is referred to as
+\emph{licm-ap}. Padding, which avoids the introduction of a reminder
+loop as described in Section~\ref{licm-impact}, as well as data
+alignment, enhance the quality of auto-vectorization. Occasionally the
+run-time of \emph{licm-ap} is close to that of \emph{licm}, since the
+non-padded element matrix size is already a multiple of the vector
+length. Rarely \emph{licm-ap} is slower than \emph{licm} (e.g. in
+Burgers $p=3$ on the Sandy Bridge). One possible explanation is that
+the number of aligned temporaries introduced by \emph{licm} is so
+large to induce cache associativity conflicts.
 
@@
 \label{sec:perf-eval-opvect}
-In this section, we evaluate the impact of vector-register tiling.
-We compare two versions: the baseline, \emph{licm-ap}; and vector-register tiling on top of \emph{licm-ap}, which in the following is referred to simply as \emph{op-vect}.
-
-Figures~\ref{fig:opvect-helmholtz-speedup} and~\ref{fig:opvect-diffusion-speedup} illustrate the speed up achieved by \emph{op-vect} over \emph{licm-ap} in the Helmoltz and Diffusion kernels, respectively.
-As explained in Section~\ref{sec:pyop2-compiler}, the \emph{op-vect} version requires the unroll-and-jam factor to be explicitly set.
-To distinguish between the two ways this parameter was determined, for each problem instance (equation, element shape, polynomial order) we report two bars: one shows the best speed-up obtained after all feasible unroll-and-jam factors were tried; the other shows the speed up when the unroll-and-jam factor was retrieved via the cost model.
-In a plot legend, cost model bars are suffixed with ``CM''.
-
-It is worth noticing that, in most cases, the cost model successfully determines how to transform a kernel to maximize its performance.
-This is chiefly because assembly kernels fit the L1 cache, so, within a certain degree of confidence, it is possible to predict how to obtain a fast implementation by simply reasoning on the register pressure.
-For each problem, the cost model stated to use the default loop permutation, to apply a particular unroll-and-jam factor, and not to perform expression splitting, which, as explained in Section~\ref{sec:perf-results-split}, only deteriorates performance in Helmholtz and Diffusion.
-
-The rationale behind these results is that the effect of \emph{op-vect} is significant in problems in which the assembly loop nest is relatively big.
-When the loops are short, since the number of arrays accessed at every loop iteration is rather small (between 4 and 8 temporaries, plus the element matrix itself), there is no need for vector-register tiling; extensive unrolling is sufficient to improve register re-use and, therefore, to maximize the performance.
-However, as the iteration space becomes larger, \emph{op-vect} leads to improvements up to 1.4$\times$ on the Sandy Bridge (Diffusion, prismatic mesh, $p=4$ - increasing the overall speed up from 2.69$\times$ to 3.87$\times$), and up to 1.4$\times$ on the Xeon Phi (Helmholtz, tetrahedral mesh, $p=3$ - bringing the overall speed up from 1.71$\times$ to 2.42$\times$).
-
-Using the Intel Architecture Code Analyzer tool~\cite{IACA} on the Sandy Bridge, we confirmed that speed ups are a consequence of increased register re-use.
-In Helmholtz $p=4$, for example, the tool showed that when using \emph{op-vect} the number of clock cycles to execute one iteration of the \texttt{j} loop decreases by roughly 17$\%$, and that this is a result of the relieved pressure on both of the data (cache) ports available in the core.
-
-On the Sandy Bridge, we have also measured the performance of individual kernels in terms of floating-point operations per second.
-The theoretical peak on a single core, with the Intel Turbo Boost technology activated, is 30.4 GFlop/s.
-In the case of Diffusion using a prismatic mesh and $p=4$, we achieved a maximum of 21.9 GFlop/s with \emph{op-vect} enabled, whereas 16.4 GFlop/s was obtained when only \emph{licm-ap} is used.
-This result is in line with the expectations: analysis of assembly code showed that, in the \texttt{jk} loop nest, which in this problem represents the bulk of the computation, 73$\%$ of instructions are actually floating-point operations.
-
-Application of \emph{op-vect} to the Burgers problem induces significant slow downs due to the large number of temporary arrays that need to be tiled, which exceeds the available logical registers on the underlying architecture.
-Expression splitting can be used in combination with \emph{op-vect} to alleviate this issue; this is discussed in the next section.
+In this section, we evaluate the impact of vector-register tiling. We
+compare two versions: the baseline, \emph{licm-ap}; and
+vector-register tiling on top of \emph{licm-ap}, which in the
+following is referred to simply as \emph{op-vect}.
+
+Figures~\ref{fig:opvect-helmholtz-speedup}
+and~\ref{fig:opvect-diffusion-speedup} illustrate the speed up
+achieved by \emph{op-vect} over \emph{licm-ap} in the Helmoltz and
+Diffusion kernels, respectively. As explained in
+Section~\ref{sec:pyop2-compiler}, the \emph{op-vect} version requires
+the unroll-and-jam factor to be explicitly set. To distinguish between
+the two ways this parameter was determined, for each problem instance
+(equation, element shape, polynomial order) we report two bars: one
+shows the best speed-up obtained after all feasible unroll-and-jam
+factors were tried; the other shows the speed up when the
+unroll-and-jam factor was retrieved via the cost model. In a plot
+legend, cost model bars are suffixed with ``CM''.
+
+It is worth noticing that, in most cases, the cost model successfully
+determines how to transform a kernel to maximize its performance. This
+is chiefly because assembly kernels fit the L1 cache, so, within a
+certain degree of confidence, it is possible to predict how to obtain
+a fast implementation by simply reasoning on the register
+pressure. For each problem, the cost model stated to use the default
+loop permutation, to apply a particular unroll-and-jam factor, and not
+to perform expression splitting, which, as explained in
+Section~\ref{sec:perf-results-split}, only deteriorates performance in
+Helmholtz and Diffusion.
+
+The rationale behind these results is that the effect of
+\emph{op-vect} is significant in problems in which the assembly loop
+nest is relatively big. When the loops are short, since the number of
+arrays accessed at every loop iteration is rather small (between 4 and
+8 temporaries, plus the element matrix itself), there is no need for
+vector-register tiling; extensive unrolling is sufficient to improve
+register re-use and, therefore, to maximize the performance. However,
+as the iteration space becomes larger, \emph{op-vect} leads to
+improvements up to 1.4$\times$ on the Sandy Bridge (Diffusion,
+prismatic mesh, $p=4$ - increasing the overall speed up from
+2.69$\times$ to 3.87$\times$), and up to 1.4$\times$ on the Xeon Phi
+(Helmholtz, tetrahedral mesh, $p=3$ - bringing the overall speed up
+from 1.71$\times$ to 2.42$\times$).
+
+Using the Intel Architecture Code Analyzer tool~\cite{IACA} on the
+Sandy Bridge, we confirmed that speed ups are a consequence of
+increased register re-use. In Helmholtz $p=4$, for example, the tool
+showed that when using \emph{op-vect} the number of clock cycles to
+execute one iteration of the \texttt{j} loop decreases by roughly
+17$\%$, and that this is a result of the relieved pressure on both of
+the data (cache) ports available in the core.
+
+On the Sandy Bridge, we have also measured the performance of
+individual kernels in terms of floating-point operations per
+second. The theoretical peak on a single core, with the Intel Turbo
+Boost technology activated, is 30.4 GFlop/s. In the case of Diffusion
+using a prismatic mesh and $p=4$, we achieved a maximum of 21.9
+GFlop/s with \emph{op-vect} enabled, whereas 16.4 GFlop/s was obtained
+when only \emph{licm-ap} is used. This result is in line with the
+expectations: analysis of assembly code showed that, in the
+\texttt{jk} loop nest, which in this problem represents the bulk of
+the computation, 73$\%$ of instructions are actually floating-point
+operations.
+
+Application of \emph{op-vect} to the Burgers problem induces
+significant slow downs due to the large number of temporary arrays
+that need to be tiled, which exceeds the available logical registers
+on the underlying architecture. Expression splitting can be used in
+combination with \emph{op-vect} to alleviate this issue; this is
+discussed in the next section.
 
@@
 	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/helmholtz-normalized-opvect-phi.pdf}}
-	\caption{Performance improvement over \emph{licm-ap} obtained by \emph{op-vect} in the Helmholtz kernel. Bars suffixed with ``CM'' indicate that the cost model was used to transform the kernel.}
+	\caption{Performance improvement over \emph{licm-ap} obtained
+          by \emph{op-vect} in the Helmholtz kernel. Bars suffixed
+          with ``CM'' indicate that the cost model was used to
+          transform the kernel.}
 	\label{fig:opvect-helmholtz-speedup}
@@
 	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/diffusion-normalized-opvect-phi.pdf}}
-	\caption{Performance improvement over \emph{licm-ap} obtained by \emph{op-vect} in the Diffusion kernel. Bars suffixed with ``CM'' indicate that the cost model was used to transform the kernel.}
+	\caption{Performance improvement over \emph{licm-ap} obtained
+          by \emph{op-vect} in the Diffusion kernel. Bars suffixed
+          with ``CM'' indicate that the cost model was used to
+          transform the kernel.}
 	\label{fig:opvect-diffusion-speedup}
@@
 \label{sec:perf-results-split}
-Expression splitting relieves the register pressure when the element matrix evaluation needs to read from a large number of basis function arrays.
-As detailed in Section~\ref{sec:expr-split}, in the price to pay for this optimazion, there are an increased number of accesses to the element matrix and, potentially, redundant computation.
-Similarly to the analysis of vector-register tiling, we compare two versions: the baseline, \emph{licm-ap}; and expression splitting on top of \emph{licm-ap}, which, for simplicity, in the following is referred to as \emph{split}.
-
-For the Helmholtz and Diffusion kernels, in which only between 4 and 8 temporaries are read at every loop iteration, \texttt{split} tends to slow down the computation, because of the aforementioned drawbacks.
-Slow downs up to 1.4$\times$ and up to 1.6$\times$ were observed, respectively, on the Sandy Bridge and the Xeon Phi.
-Note that the cost model prevents the adoption of the transformation: the \texttt{while} statement in Figure~\ref{algo:applyCostModel} is indeed never entered.
-
-In the Burgers kernels, between 12 and 24 temporaries are accessed at every loop iteration, so \emph{split} plays a key role on the Sandy Bridge, where the number of available logical registers is only 16.
-Figure~\ref{fig:split-burgers-speedup} shows the performance improvement achieved by \emph{split} over \emph{licm-ap}.
-In almost all cases, a split factor of 1, meaning that the original expression was divided into two parts, ensured close-to-peak perforance.
-The transformation negligibly affected register locality, so speed ups up to 1.5$\times$ were observed.
-For instance, on the Sandy Bridge, when $p=4$ and a prismatic mesh is employed, the overall performance improvement (i.e.
-the one over the original code) increases from 1.44$\times$ to 2.11$\times$.
-On the Xeon Phi, the impact of \emph{split} is only marginal, since register spilling is limited by the presence of 32 logical vector units.
-
-On the Sandy Bridge, the performance of the Burgers kernel on a prismatic mesh was 20.0 GFlop/s from $p=1$ to $p=3$, while it was 21.3 GFlop/s in the case of $p=4$.
-These values are notably close to the peak performance of 30.4 GFlop/s.
-Disabling \emph{split} makes the performance drop to 17.0 GFlop/s for $p=1, 2$, 18.2 GFlop/s for $p=3$, and 14.3 GFlop/s for $p=4$.
-These values are in line with the speedups shown in Figure~\ref{fig:split-burgers-speedup}.
-
-The \emph{split} transformation was also tried in combination with \emph{op-vect} (\emph{split-op-vect}), although the cost model prevents its adoption on both platforms.
-Despite improvements up to 1.22$\times$, \emph{split-op-vect} never outperforms \emph{split}.
-This is motivated by two factors: for small split factors, such as 1 and 2, the data space to be tiled is still too big, and register spilling affects run-time; for higher ones, sub-expressions become so small that, as explained in Section~\ref{sec:perf-eval-opvect}, extensive unrolling already allows to achieve a certain degree of register re-use.
+Expression splitting relieves the register pressure when the element
+matrix evaluation needs to read from a large number of basis function
+arrays. As detailed in Section~\ref{sec:expr-split}, in the price to
+pay for this optimazion, there are an increased number of accesses to
+the element matrix and, potentially, redundant computation. Similarly
+to the analysis of vector-register tiling, we compare two versions:
+the baseline, \emph{licm-ap}; and expression splitting on top of
+\emph{licm-ap}, which, for simplicity, in the following is referred to
+as \emph{split}.
+
+For the Helmholtz and Diffusion kernels, in which only between 4 and 8
+temporaries are read at every loop iteration, \texttt{split} tends to
+slow down the computation, because of the aforementioned
+drawbacks. Slow downs up to 1.4$\times$ and up to 1.6$\times$ were
+observed, respectively, on the Sandy Bridge and the Xeon Phi. Note
+that the cost model prevents the adoption of the transformation: the
+\texttt{while} statement in Figure~\ref{algo:applyCostModel} is indeed
+never entered.
+
+In the Burgers kernels, between 12 and 24 temporaries are accessed at
+every loop iteration, so \emph{split} plays a key role on the Sandy
+Bridge, where the number of available logical registers is only
+16. Figure~\ref{fig:split-burgers-speedup} shows the performance
+improvement achieved by \emph{split} over \emph{licm-ap}. In almost
+all cases, a split factor of 1, meaning that the original expression
+was divided into two parts, ensured close-to-peak perforance. The
+transformation negligibly affected register locality, so speed ups up
+to 1.5$\times$ were observed. For instance, on the Sandy Bridge, when
+$p=4$ and a prismatic mesh is employed, the overall performance
+improvement (i.e., the one over the original code) increases from
+1.44$\times$ to 2.11$\times$. On the Xeon Phi, the impact of
+\emph{split} is only marginal, since register spilling is limited by
+the presence of 32 logical vector units.
+
+On the Sandy Bridge, the performance of the Burgers kernel on a
+prismatic mesh was 20.0 GFlop/s from $p=1$ to $p=3$, while it was 21.3
+GFlop/s in the case of $p=4$. These values are notably close to the
+peak performance of 30.4 GFlop/s. Disabling \emph{split} makes the
+performance drop to 17.0 GFlop/s for $p=1, 2$, 18.2 GFlop/s for $p=3$,
+and 14.3 GFlop/s for $p=4$. These values are in line with the speedups
+shown in Figure~\ref{fig:split-burgers-speedup}.
+
+The \emph{split} transformation was also tried in combination with
+\emph{op-vect} (\emph{split-op-vect}), although the cost model
+prevents its adoption on both platforms. Despite improvements up to
+1.22$\times$, \emph{split-op-vect} never outperforms
+\emph{split}. This is motivated by two factors: for small split
+factors, such as 1 and 2, the data space to be tiled is still too big,
+and register spilling affects run-time; for higher ones,
+sub-expressions become so small that, as explained in
+Section~\ref{sec:perf-eval-opvect}, extensive unrolling already allows
+to achieve a certain degree of register re-use.
 
@@
 	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/burgers-normalized-split-phi.pdf}}
-	\caption{Performance improvement over \emph{licm-ap} obtained by \emph{split} in the Burgers kernel. Bars suffixed with ``CM'' indicate that the cost model was used to transform the kernel.}
+	\caption{Performance improvement over \emph{licm-ap} obtained
+          by \emph{split} in the Burgers kernel. Bars suffixed with
+          ``CM'' indicate that the cost model was used to transform
+          the kernel.}
 	\label{fig:split-burgers-speedup}
@@
 \label{sec:perf-results-fenics}
-We have modified the FEniCS Form Compiler (FFC) to return an abstract syntax tree representation of a local assembly kernel, rather than plain C++ code, so as to enable code transformations in COFFEE.
-Besides Firedrake, FFC is used in the FEniCS project~\cite{Fenics}.
-In FEniCS, FFC can apply its own model-driven optimizations to local assembly kernels~\cite{quadrature1}, which mainly consist of loop-invariant code motion and elimination, at code generation time, of floating point operations involving zero-valued entries in basis function arrays.
-
-The FEniCS Form Compiler's loop-invariant code motion is different from COFFEE's.
-It is based on expansion of arithmetic operations, for example applying distributivity and associativity to products and sums at code generation time, to identify terms that are invariant of the whole loop nest.
-Depending on the way expansion is performed, operation count may not decrease significantly.
-
-Elimination of zero-valued terms, which are the result of using vector-valued quantities in the finite element problem, has the effect of introducing indirection arrays in the generated code.
-This kind of optimization is currently under development in COFFEE, although it will differ from FEniCS' by avoiding non-contiguous memory accesses, which would otherwise affect vectorization, at the price of removing fewer zero-valued contributions.
-
-Table~\ref{table:comparison-to-FFC-opt} summarizes the performance achieved by COFFEE over the \emph{fastest} FEniCS (FFC) implementation on the Sandy Bridge for the Burgers, Helmholtz and Diffusion kernels.
-Burgers' slow downs occur in presence of a small iteration space (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$).
-The result shown represents the worst slow down, which was obtained with a triangular mesh and $p = 1$.
-This is a result of removing zero-valued entries in FEniCS' basis function arrays: some operations are avoided, but indirection arrays prevent auto-vectorization, which significantly impacts performance as soon as the element matrix becomes bigger than 12$\times$12.
-However, with the forthcoming zero-removal optimization in COFFEE, we expect to outperform FEniCS in all problems.
-In the cases of Helmholtz and Diffusion, the minimum improvements are, respectively, 1.10$\times$ and 1.18$\times$ (2D mesh, $p=1$), which tend to increase with polynomial order and element shape up to the values illustrated in the table.
+We have modified the FEniCS Form Compiler (FFC) to return an abstract
+syntax tree representation of a local assembly kernel, rather than
+plain C++ code, so as to enable code transformations in
+COFFEE. Besides Firedrake, FFC is used in the FEniCS
+project~\cite{Fenics}. In FEniCS, FFC can apply its own model-driven
+optimizations to local assembly kernels~\cite{quadrature1}, which
+mainly consist of loop-invariant code motion and elimination, at code
+generation time, of floating point operations involving zero-valued
+entries in basis function arrays.
+
+The FEniCS Form Compiler's loop-invariant code motion is different
+from COFFEE's. It is based on expansion of arithmetic operations, for
+example applying distributivity and associativity to products and sums
+at code generation time, to identify terms that are invariant of the
+whole loop nest. Depending on the way expansion is performed,
+operation count may not decrease significantly.
+
+Elimination of zero-valued terms, which are the result of using
+vector-valued quantities in the finite element problem, has the effect
+of introducing indirection arrays in the generated code. This kind of
+optimization is currently under development in COFFEE, although it
+will differ from FEniCS' by avoiding non-contiguous memory accesses,
+which would otherwise affect vectorization, at the price of removing
+fewer zero-valued contributions.
+
+Table~\ref{table:comparison-to-FFC-opt} summarizes the performance
+achieved by COFFEE over the \emph{fastest} FEniCS (FFC) implementation
+on the Sandy Bridge for the Burgers, Helmholtz and Diffusion
+kernels. Burgers' slow downs occur in presence of a small iteration
+space (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1,
+  2]$; prismatic mesh, $p = 1$). The result shown represents the worst
+slow down, which was obtained with a triangular mesh and $p = 1$. This
+is a result of removing zero-valued entries in FEniCS' basis function
+arrays: some operations are avoided, but indirection arrays prevent
+auto-vectorization, which significantly impacts performance as soon as
+the element matrix becomes bigger than 12$\times$12. However, with the
+forthcoming zero-removal optimization in COFFEE, we expect to
+outperform FEniCS in all problems. In the cases of Helmholtz and
+Diffusion, the minimum improvements are, respectively, 1.10$\times$
+and 1.18$\times$ (2D mesh, $p=1$), which tend to increase with
+polynomial order and element shape up to the values illustrated in the
+table.
 
@@
 \label{sec:perf-eval-blas}
-For the Helmholtz problem on a tetrahedral mesh, manual implementations based on \texttt{Intel MKL BLAS} were tested on the Sandy Bridge.
-This particular kernel can be easily reduced to a sequence of four matrix-matrix multiplies that can be computed via calls to BLAS \texttt{dgemm}.
-In the case of $p=4$, where the element matrix is of size 35$\times$35, the computation was almost twice slower than the case in which \emph{licm-ap} was used, with the slow down being even worse for smaller problem sizes.
-These experiments suggest that the question regarding to what extent linear algebra libraries can improve performance cannot be trivially answered.
-This is due to a combination of issues: the potential loss in data locality, as exposed in Section~\ref{sec:expr-split}, the actual effectiveness of these libraries when the arrays are relatively small, and the problem inherent to assembly kernels concerning extraction of matrix-matrix multiplies from static analysis of the kernel's code.
-A comprehensive study of these aspects will be addressed in further work.
+For the Helmholtz problem on a tetrahedral mesh, manual
+implementations based on \texttt{Intel MKL BLAS} were tested on the
+Sandy Bridge. This particular kernel can be easily reduced to a
+sequence of four matrix-matrix multiplies that can be computed via
+calls to BLAS \texttt{dgemm}. In the case of $p=4$, where the element
+matrix is of size 35$\times$35, the computation was almost twice
+slower than the case in which \emph{licm-ap} was used, with the slow
+down being even worse for smaller problem sizes. These experiments
+suggest that the question regarding to what extent linear algebra
+libraries can improve performance cannot be trivially answered. This
+is due to a combination of issues: the potential loss in data
+locality, as exposed in Section~\ref{sec:expr-split}, the actual
+effectiveness of these libraries when the arrays are relatively small,
+and the problem inherent to assembly kernels concerning extraction of
+matrix-matrix multiplies from static analysis of the kernel's code. A
+comprehensive study of these aspects will be addressed in further
+work.
 
@@
 \label{sec:related-work}
-The finite element method is extensively used to approximate solutions of PDEs.
-Well-known frameworks and applications include nek5000~\cite{nek5000-web-page}, the FEniCS project~\cite{Fenics}, Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake.
-Numerical integration is usually employed to implement the local assembly phase.
-The recent introduction of domain specific languages (DSLs) to decouple the finite element specification from its underlying implementation facilitated, however, the development of novel approaches.
-Methods based on tensor contraction~\cite{FFC-Compiler} and symbolic manipulation~\cite{Francis} have been implemented.
-We have designed COFFEE to specifically target numerical integration because it has been demonstrated that it remains the optimal choice for a wide class of problems~\cite{quadrature1}.
-
-Optimization of local assembly using numerical integration for CPU platforms has been addressed in FEniCS~\cite{quadrature1}.
-The comparison between COFFEE and this work is presented in Section~\ref{sec:perf-results-fenics}.
-In~\cite{Markall20101815}, and more recently in~\cite{petsc-integration-gpu}, the problem has been studied for GPU architectures.
-In~\cite{assembly-opencl}, variants of the standard numerical integration algorithm have been specialized and evaluated for the PowerXCell processor, but an exhaustive study from the compiler viewpoint - like ours - is missing, and none of the optimizations presented in Section~\ref{sec:code-transf} are mentioned.
-Among these efforts, to the best of our knowledge, COFFEE is the first work targeting low-level optimizations through a real compiler approach.
-
-Our compiler-based optimization approach is made possible by the top-level DSL, which enables automated code generation.
-DSLs have been proven successful in auto-generating optimized code for other domains: Spiral~\cite{Pueschel:05} for digital signal processing numerical algorithms, ~\cite{Spampinato:14} for dense linear algebra, or Pochoir~\cite{pochoir} and SDSL~\cite{stencil-compiler} for image processing and finite difference stencils.
-Similarly, PyOP2 is used by Firedrake to express iteration over unstructured meshes in scientific codes.
-COFFEE improves automated code generation in Firedrake.
-
-Many code generators, like those based on the Polyhedral model~\cite{PLUTO} and those driven by domain-knowledge~\cite{modeldriven}, make use of cost models.
-The alternative of using auto-tuning to select the best implementation for a given problem on a certain platform has been adopted by nek5000~\cite{nek5000} for small matrix-matrix multiplies, the ATLAS library~\cite{ATLAS}, and FFTW~\cite{FFTW} for fast fourier transforms.
-In both cases, pruning the implementation space is fundamental to mitigate complexity and overhead.
-Likewise, COFFEE uses a cost model and heuristics (Section~\ref{sec:pyop2-compiler}) to steer the optimization process.
-
-%(i.e. given a mesh, those that lead to ``less accurate'' solutions than when using high-order polynomials, at a lower computational cost)
+The finite element method is extensively used to approximate solutions
+of PDEs. Well-known frameworks and applications include
+nek5000~\cite{nek5000-web-page}, the FEniCS project~\cite{Fenics},
+Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake. Numerical
+integration is usually employed to implement the local assembly
+phase. The recent introduction of domain specific languages (DSLs) to
+decouple the finite element specification from its underlying
+implementation facilitated, however, the development of novel
+approaches. Methods based on tensor contraction~\cite{FFC-Compiler}
+and symbolic manipulation~\cite{Francis} have been implemented. We
+have designed COFFEE to specifically target numerical integration
+because it has been demonstrated that it remains the optimal choice
+for a wide class of problems~\cite{quadrature1}.
+
+Optimization of local assembly using numerical integration for CPU
+platforms has been addressed in FEniCS~\cite{quadrature1}. The
+comparison between COFFEE and this work is presented in
+Section~\ref{sec:perf-results-fenics}. In~\cite{Markall20101815}, and
+more recently in~\cite{petsc-integration-gpu}, the problem has been
+studied for GPU architectures. In~\cite{assembly-opencl}, variants of
+the standard numerical integration algorithm have been specialized and
+evaluated for the PowerXCell processor, but an exhaustive study from
+the compiler viewpoint - like ours - is missing, and none of the
+optimizations presented in Section~\ref{sec:code-transf} are
+mentioned. Among these efforts, to the best of our knowledge, COFFEE
+is the first work targeting low-level optimizations through a real
+compiler approach.
+
+Our compiler-based optimization approach is made possible by the
+top-level DSL, which enables automated code generation. DSLs have been
+proven successful in auto-generating optimized code for other domains:
+Spiral~\cite{Pueschel:05} for digital signal processing numerical
+algorithms, ~\cite{Spampinato:14} for dense linear algebra, or
+Pochoir~\cite{pochoir} and SDSL~\cite{stencil-compiler} for image
+processing and finite difference stencils. Similarly, PyOP2 is used by
+Firedrake to express iteration over unstructured meshes in scientific
+codes. COFFEE improves automated code generation in Firedrake.
+
+Many code generators, like those based on the Polyhedral
+model~\cite{PLUTO} and those driven by
+domain-knowledge~\cite{modeldriven}, make use of cost models. The
+alternative of using auto-tuning to select the best implementation for
+a given problem on a certain platform has been adopted by
+nek5000~\cite{nek5000} for small matrix-matrix multiplies, the ATLAS
+library~\cite{ATLAS}, and FFTW~\cite{FFTW} for fast fourier
+transforms. In both cases, pruning the implementation space is
+fundamental to mitigate complexity and overhead. Likewise, COFFEE uses
+a cost model and heuristics (Section~\ref{sec:pyop2-compiler}) to
+steer the optimization process.
+
+%(i.e., given a mesh, those that lead to ``less accurate'' solutions than when using high-order polynomials, at a lower computational cost)
 
@@
 
-In this paper, we have presented design, optimizations and systematic performance evaluation of COFFEE, a compiler for finite element local assembly.
-In this context, to the best of our knowledge, COFFEE is the first compiler capable of introducing low-level optimizations to maximize instruction-level parallelism, register locality and SIMD vectorization.
-Assembly kernels have peculiar characteristics.
-Their iteration space is usually very small, with the size depending on aspects like the degree of accuracy one wants to reach (polynomial order of the method) and the mesh discretization employed.
-The data space, in terms of number of arrays and scalars required to evaluate the element matrix, grows proportionally with the complexity of the finite element problem.
-COFFEE has been developed taking into account all of these degrees of freedom, based on the idea that reducing the problem of local assembly optimization to a fixed sequence of transformations is far too superficial if close-to-peak performance needs to be reached.
-The various optimizations overcome limitations of current vendor and research compilers.
-The exploitation of domain knowledge allows some of them to be particularly effective, as demonstrated by our experiments on two state-of-the-art Intel platforms.
-Further work include a comprehensive study about feasibility and constraints on transforming kernels into a sequence of calls to external linear algebra libraries.
-COFFEE supports all of the problems expressible in Firedrake, and it is already integrated with this framework.
+In this paper, we have presented design, optimizations and systematic
+performance evaluation of COFFEE, a compiler for finite element local
+assembly. In this context, to the best of our knowledge, COFFEE is the
+first compiler capable of introducing low-level optimizations to
+maximize instruction-level parallelism, register locality and SIMD
+vectorization. Assembly kernels have peculiar characteristics. Their
+iteration space is usually very small, with the size depending on
+aspects like the degree of accuracy one wants to reach (polynomial
+order of the method) and the mesh discretization employed. The data
+space, in terms of number of arrays and scalars required to evaluate
+the element matrix, grows proportionally with the complexity of the
+finite element problem. COFFEE has been developed taking into account
+all of these degrees of freedom, based on the idea that reducing the
+problem of local assembly optimization to a fixed sequence of
+transformations is far too superficial if close-to-peak performance
+needs to be reached. The various optimizations overcome limitations of
+current vendor and research compilers. The exploitation of domain
+knowledge allows some of them to be particularly effective, as
+demonstrated by our experiments on two state-of-the-art Intel
+platforms. Further work include a comprehensive study about
+feasibility and constraints on transforming kernels into a sequence of
+calls to external linear algebra libraries. COFFEE supports all of the
+problems expressible in Firedrake, and it is already integrated with
+this framework.
 

### 2014-06-28 TACO-paper/coffee.tex
edits

@@
 compiler for local assembly kernels. COFFEE manipulates abstract
-syntax trees by introducing domain-aware composable optimizations aimed at
+syntax trees generated from a high-level domain-specific language for
+PDEs by introducing domain-aware composable optimizations aimed at
 improving instruction-level parallelism, especially SIMD
@@
 performance improvement is achieved.
+%
+%The code is generated from a high level language for PDEs. An
+%invocation of the kernel typically does tensor-like calculators over
+%relatively small matrices; e.g., triply nested loops operating on
+%about a dozen iterations. Because of the DSL context, COFFEE has a lot
+%of flexibility on how it organizes/orders the calculations to tailor
+%execution to a target architecture.
+%
 \end{abstract}
@@
 partitioning the mesh and assigning each partition to a different node
-or thread. Such execution model, with minor variations, is adopted,
+or thread. Such an execution model, with minor variations, is adopted,
 for example, in \cite{pyop2isc}, \cite{Fenics},
@@
 since the equation domain needs to be discretized into an extremely
-large number of cells, possibly order of trillions
+large number of cells, possibly of the order of trillions
 \cite{Rossinelli2013}, in order to obtain a satisfactory approximation
@@
 particular, the well-known problem of optimizing the local assembly
-phase of the finite element
-method~\cite{Francis},~\cite{quadrature1},~\cite{petsc-integration-gpu},~\cite{Kirby-FEM-opt},
-which can cover a significant fraction of the overall computation
-run-time, often in the range 30-60$\%$. With respect to these studies,
-we propose a novel set of composable code transformations targeting,
-for the first time, instruction-level parallelism - with emphasis on
-SIMD vectorization - and register locality.
+phase of the finite element method
+\cite{Francis,quadrature1,petsc-integration-gpu,Kirby-FEM-opt}, which
+can cover a significant fraction of the overall computation run-time,
+often in the range 30-60$\%$. With respect to these studies, we
+propose a novel set of composable code transformations targeting, for
+the first time, instruction-level parallelism - with emphasis on SIMD
+vectorization - and register locality.
 

### 2014-06-29 TACO-paper/coffee.tex
edits

@@
 
-%Computational cost is a critical limitation in scientific computing, especially
-%for finite element simulations. To provide one particular example we are
-%particularly concerned about, it has been well established that mesh resolution
-% Consequently, our aggressive optimization of
-%local assembly, which may even take up to 60% of the overall FEM's execution
-%time, directly impacts the performance of large-scale scientific simulations
-%running on supercomputers.
+%Computational cost is a critical limitation in scientific computing,
+%especially for finite element simulations. To provide one particular
+%example we are particularly concerned about, it has been well
+%established that mesh resolution Consequently, our aggressive
+%optimization of local assembly, which may even take up to 60% of the
+%overall FEM's execution time, directly impacts the performance of
+%large-scale scientific simulations running on supercomputers.
 
@@
 
-%JUNK: , with synchronization handled at the level of the run-time support
-%Computational cost is a critical limitation in finite element simulations.
-High resolution in time and space is critical to the level of error in the simulations, but computational cost rises superlinearly with resolution.
-The result of this is that simulations are conducted at the resolution that the user can afford (wall time, allocated supercomputing units or the direct financial cost of cloud time)
+%JUNK: , with synchronization handled at the level of the run-time
+%support Computational cost is a critical limitation in finite element
+%simulations. High resolution in time and space is critical to the
+%level of error in the simulations, but computational cost rises
+%superlinearly with resolution. The result of this is that simulations
+%are conducted at the resolution that the user can afford (wall time,
+%allocated supercomputing units or the direct financial cost of cloud
+%time)
 
@@
 
-%JUNK: 1.5$\times$ over code in which only basic transformations have been performed, such as generalized loop-invariant code motion, padding, and data alignment, and up to
+%JUNK: 1.5$\times$ over code in which only basic transformations have
+%been performed, such as generalized loop-invariant code motion,
+%padding, and data alignment, and up to
 
@@
 
-% systematic optimization strategy -> discuss the importance of domain knowledge
-
-% A python implementation of OP2, namely PyOP2, is used in the Firedrake project as an intermediate representation for FEM computations.
-In Firedrake all computational kernels are generated by a compiler, (the Fenics Form Compiler, FFC), whereas in OP2 applications kernels are provided by users.
+% systematic optimization strategy -> discuss the importance of domain
+% knowledge
+
+% A python implementation of OP2, namely PyOP2, is used in the
+% Firedrake project as an intermediate representation for FEM
+% computations. In Firedrake all computational kernels are generated
+% by a compiler, (the Fenics Form Compiler, FFC), whereas in OP2
+% applications kernels are provided by users.
 
@@
 \label{sec:background}
-%Here we discuss about the computational characteristics of computational science kernels.
-Briefly cite op2 kernels.
-Emphasis on Finite Element Assembly.
-Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.
+%Here we discuss about the computational characteristics of
+%computational science kernels. Briefly cite op2 kernels. Emphasis on
+%Finite Element Assembly. Generalization and formalization of a Finite
+%Element Assembly kernel using quadrature representation.
 
@@
 \begin{alltt}
-\footnotesize
-\textbf{Input:} element matrix (2D array, initialized to 0), 
-       element coordinates (array), 
-       coefficient fields (array, e.g. velocity) 
-\textbf{Output:} element matrix (2D array)
-- Compute Jacobian from coordinates 
-- Define basis functions 
-- Compute element matrix in an affine loop nest
+\footnotesize \textbf{Input:} element matrix (2D array, initialized to
+0), element coordinates (array), coefficient fields (array,
+e.g. velocity) \textbf{Output:} element matrix (2D array) - Compute
+Jacobian from coordinates - Define basis functions - Compute element
+matrix in an affine loop nest
 \end{alltt}
@@
 length of the innermost loop is not a multiple of the vector length
-$vl$, especially when the loops are small as in local assembly.
+$\mbox{\texttt{VL}}$, especially when the loops are small as in local assembly.
 
 Data alignment is enforced in two steps. Firstly, all arrays are
-allocated to addresses that are multiples of $vl$. Then, two
+allocated to addresses that are multiples of $\mbox{\texttt{VL}}$. Then, two
 dimensional arrays are padded by rounding the number of columns to the
-nearest multiple of $vl$. For instance, assume the original size of a
-basis function array is 3$\times$3 and $vl=4$ (e.g. AVX processor,
+nearest multiple of $\mbox{\texttt{VL}}$. For instance, assume the original size of a
+basis function array is 3$\times$3 and $\mbox{\texttt{VL}}=4$ (e.g. AVX processor,
 with 256 bit long vector registers and 64-bit double-precision
@@
 the element matrix also allows to safely round the loop trip count to
-the nearest multiple of $vl$. This avoids the introduction of a
+the nearest multiple of $\mbox{\texttt{VL}}$. This avoids the introduction of a
 remainder (scalar) loop from the compiler, which would render
@@
 
-%More importantly, we avoid using $vl$-1 registers per each
+%More importantly, we avoid using $\mbox{\texttt{VL}}$-1 registers per each
 %outer-loop-dependent variable (i.e., all $x$ terms in
-%Equation~\ref{outer-product}), with $vl$ being the vector
+%Equation~\ref{outer-product}), with $\mbox{\texttt{VL}}$ being the vector
 %length. This is a considerable gain, which allows us to slice the

### 2014-06-29 TACO-paper/coffee.tex
edits

@@
 % At a minimum you need to supply the author names, year and a title.
-% IMPORTANT:
-% Full first names whenever they are known, surname last, followed by a period.
-% In the case of two authors, 'and' is placed between them.
-% In the case of three or more authors, the serial comma is used, that is, all author names
-% except the last one but including the penultimate author's name are followed by a comma,
-% and then 'and' is placed before the final author's name.
-% If only first and middle initials are known, then each initial
-% is followed by a period and they are separated by a space.
-% The remaining information (journal title, volume, article number, date, etc.) is 'auto-generated'.
+% IMPORTANT: Full first names whenever they are known, surname last,
+% followed by a period.  In the case of two authors, 'and' is placed
+% between them.  In the case of three or more authors, the serial
+% comma is used, that is, all author names except the last one but
+% including the penultimate author's name are followed by a comma, and
+% then 'and' is placed before the final author's name.  If only first
+% and middle initials are known, then each initial is followed by a
+% period and they are separated by a space.  The remaining information
+% (journal title, volume, article number, date, etc.) is
+% 'auto-generated'.
 
@@
 \section{Introduction}
-In many fields, such as computational fluid dynamics, computational
-electromagnetics, and structural mechanics, phenomena are modelled by
+In many fields such as computational fluid dynamics, computational
+electromagnetics and structural mechanics, phenomena are modelled by
 partial differential equations (PDEs). Numerical techniques, like the
@@
 kernels, to the components of a mesh, such as edges, vertices, or
-cells. On standard clusters of multi-cores, typically, a kernel is
+cells. On standard clusters of multicores, typically, a kernel is
 executed sequentially by a thread, while parallelism is achieved
@@
 assembly kernels, despite following state-of-the-art techniques,
-remains suboptimal in the context of modern multi-core
-architectures. Our cost-model-driven sequence of code transformations,
-aimed at improving SIMD vectorization and register data locality, can
-result in performance improvements up to 4.4$\times$ over original
-kernels, and around 70$\%$ of the test cases obtain a speed up greater
-than 2$\times$. The contributions of this paper, therefore, are
+remains suboptimal in the context of modern multicore
+architectures. Our domian-aware cost-model-driven sequence of code
+transformations, aimed at improving SIMD vectorization and register
+data locality, can result in performance improvements up to
+4.4$\times$ over original kernels, and around 70$\%$ of the test cases
+obtain a speed up greater than 2$\times$. The contributions of this
+paper, therefore, are
 \begin{enumerate}

### 2014-06-30 TACO-paper/coffee.tex
Copy editing of introduction

@@
 \section{Introduction}
+
 In many fields such as computational fluid dynamics, computational
@@
 employed to approximate solutions of these PDEs. Unstructured meshes
-are often used to discretize the equation domain, since their
-geometric flexibility allows solvers to be extremely effective. The
+are often used to discretize the computational domain, since they
+allow an accurate representation of complex geometries. The
 solution is sought by applying suitable numerical operations, or
-kernels, to the components of a mesh, such as edges, vertices, or
+kernels, to the entities of a mesh, such as edges, vertices, or
 cells. On standard clusters of multicores, typically, a kernel is
-executed sequentially by a thread, while parallelism is achieved
+executed sequentially by a thread, while parallelism is achieved by
 partitioning the mesh and assigning each partition to a different node
@@
 
+% FIXME: citation for Met Office?
+
 The time required to apply the numerical kernels is a major issue,
 since the equation domain needs to be discretized into an extremely
-large number of cells, possibly of the order of trillions
-\cite{Rossinelli2013}, in order to obtain a satisfactory approximation
-of the PDE. For example, it has been well established that mesh
+large number of cells to obtain a satisfactory approximation
+of the PDE, possibly of the order of trillions
+\cite{Rossinelli2013}. For example, it has been well established that mesh
 resolution is critical in the accuracy of numerical weather
-forecasts. However, operational forecast centres have a strict time
+forecasts. However, operational forecast centers have a strict time
 limit in which to produce a forecast - 60 minutes in the case of the
-UK Met Office - so producing efficient kernels has a direct scientific
+UK Met Office. Producing efficient kernels has a direct scientific
 payoff in higher resolution, and therefore more accurate,
-forecasts. Hence, the computational cost is a dominant problem in
+forecasts. Computational cost is a dominant problem in
 computational science simulations, especially for those based on
@@
 \cite{Francis,quadrature1,petsc-integration-gpu,Kirby-FEM-opt}, which
-can cover a significant fraction of the overall computation run-time,
-often in the range 30-60$\%$. With respect to these studies, we
-propose a novel set of composable code transformations targeting, for
+can be responsible for a significant fraction of the overall computation
+run-time, often in the range 30-60$\%$. With respect to these studies,
+we propose a novel set of composable code transformations targeting, for
 the first time, instruction-level parallelism - with emphasis on SIMD
@@
 During the assembly phase, the solution of the PDE is approximated by
-executing a problem-specific kernel over all cells, or elements, in
-the discretized domain. In this work, we focus on relatively low order
+executing a problem-specific kernel over all cells, or elements, in the
+discretized domain. In this work, we focus on relatively low order
 finite element methods, in which an assembly kernel's working set is
@@
 means exotic: they are employed in a wide variety of fields, including
-climate and ocean modeling, computational fluid dynamics, and
-structural mechanics. High order methods are often based upon special
-formulations of the finite element method, which result in the
-adoption of different sets of algorithms to solve PDEs~\cite{SPENCER},
-and therefore are excluded from our study.
+climate and ocean modeling, computational fluid dynamics, and structural
+mechanics. High order methods such as the spectral element method
+\cite{SPENCER} commonly require different algorithms to solve PDEs and
+therefore are excluded from our study.
 
@@
 low order methods. In the innermost loop, a problem-specific,
-flops-intensive expression evaluates a two dimensional array,
+compute intensive expression evaluates a two dimensional array,
 representing the result of local assembly in an element of the
@@
 a single or specific sequence of transformations that is successfully
-applicable to all problems. The variation in loop trip counts and
-their typically small value further exacerbates the issue. A
+applicable to all problems. Loop trip counts are typically small and can
+vary significantly, which further exacerbates the issue. A
 compiler-based approach is, therefore, the only reasonable option to
@@
 loop-invariant code motion, vector-register tiling, and expression
-splitting, as well as their composition, are essential, but none are
+splitting, as well as their composition, are essential. but none are
 supported by state-of-the-art polyhedral and vendor compilers. BLAS
 routines could theoretically be employed, although a fairly
-complicated control- and data-flow analysis would be required to
+complicated control and data flow analysis would be required to
 automate identification and extraction of matrix-matrix multiplies. In
@@
 In order to overcome the constraints of the available compilers and
-specialized linear-algebra libraries, we have automated a set of
-generic and model-driven code transformations in
-COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local
-  assembly.}, a compiler for optimizing local assembly kernels. COFFEE
-is integrated with Firedrake~\cite{firedrake-code}, a system for
-solving PDEs through the finite element method based on the PyOP2
-abstraction~\cite{pyop2isc}. It supports all problems expressible with
-this framework, including equations that can be found at the core of
-real-world simulations, such as those used in our performance
-evaluation. In evaluating our code transformations in a range of
-notable problems, we also vary two key parameters that impact solution
-accuracy and kernel cost: the polynomial order of the method (we
-investigate from $p=1$ to $p=4$) and the geometry of elements in the
+specialized linear algebra libraries, we have automated a set of generic
+and model-driven code transformations in COFFEE\footnote{COFFEE stands
+for COmpiler For FinitE Element local assembly.}, a compiler for
+optimizing local assembly kernels. COFFEE is integrated with
+Firedrake~\cite{firedrake-code}, a system for solving PDEs through the
+finite element method based on the PyOP2
+abstraction~\cite{pyop2ws,pyop2isc}. All problems expressible with this
+framework are supported by COFFEE, including equations that can be found
+at the core of real-world simulations, such as those used in our
+performance evaluation. In evaluating our code transformations for a
+range of relevant problems, we vary two key parameters that impact
+solution accuracy and kernel cost: the polynomial order of the method
+(we investigate from $p=1$ to $p=4$) and the geometry of elements in the
 discretized domain (2D triangle, 3D tetrahedron, 3D prism). From the
@@
 assembly kernels, despite following state-of-the-art techniques,
+% FIXME: what are these state-of-the-art techniques?
 remains suboptimal in the context of modern multicore
-architectures. Our domian-aware cost-model-driven sequence of code
+architectures. Our domain-aware cost-model-driven sequence of code
 transformations, aimed at improving SIMD vectorization and register
@@
 obtain a speed up greater than 2$\times$. The contributions of this
-paper, therefore, are
+paper are
 \begin{enumerate}
-\item An optimization strategy for finite element local assembly. Our
-  approach exploits domain knowledge and goes beyond the limits of
-  both vendor and research compilers.
-\item Design and implementation of a compiler that automates the
+\item An optimization strategy for finite element local assembly that
+  exploits domain knowledge and goes beyond the limits of both vendor
+  and research compilers.
+\item The design and implementation of a compiler that automates the
   proposed code transformations for any problems expressible in
   Firedrake.
-\item Systematic analysis using a suite of examples of real-world
-  importance, and evidence of significant performance improvements on
+\item A systematic analysis using a suite of examples of real-world
+  importance that is evidence of significant performance improvements on
   two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
@@
 The paper is organized as follows. In Section~\ref{sec:background} we
-provide some background on local assembly, showing code generated by
-Firedrake and emphasizing the critical computational
+provide some background on local assembly, show code generated by
+Firedrake and emphasize the critical computational
 aspects. Section~\ref{sec:code-transf} describes the various code
-transformations, highlighting when and how domain-knowledge has been
+transformations, highlighting when and how domain knowledge has been
 exploited. The design and implementation of our compiler is discussed
 in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results}
-shows performance results. Related work is illustrated in
+shows performance results. Related work is described in
 Section~\ref{sec:related-work}, while Section~\ref{sec:conclusions}

### 2014-06-30 TACO-paper/coffee.tex
Copy editing of introduction
Mention numerical integration in abstract
Copy editing of background

@@
 The numerical solution of partial differential equations using the
-finite element method is one of the key applications of high
-performance computing. Local assembly is its characteristic
-operation. This entails the execution of a problem-specific kernel for
-each element in the discretized problem domain. Since the domain size
-can be huge, executing efficient kernels is fundamental. Their
-optimization is, however, a challenging issue. Even though affine loop
-nests are generally present, the short trip counts and the complexity
-of mathematical expressions make it hard to determine a single or
-unique sequence of successful transformations. Therefore, we present
-the design and systematic evaluation of COFFEE, a domain-specific
-compiler for local assembly kernels. COFFEE manipulates abstract
-syntax trees generated from a high-level domain-specific language for
-PDEs by introducing domain-aware composable optimizations aimed at
-improving instruction-level parallelism, especially SIMD
-vectorization, and register locality. It then generates C code
-including vector intrinsics. Experiments using a range of
-finite-element forms of increasing complexity show that significant
-performance improvement is achieved.
+finite element method is one of the key applications of high performance
+computing. Local assembly is its characteristic operation. This entails
+the execution of a problem-specific kernel to numerically evaluate an
+integral for each element in the discretized problem domain. Since the
+domain size can be huge, executing efficient kernels is fundamental.
+Their optimization is, however, a challenging issue. Even though affine
+loop nests are generally present, the short trip counts and the
+complexity of mathematical expressions make it hard to determine a
+single or unique sequence of successful transformations. Therefore, we
+present the design and systematic evaluation of COFFEE, a
+domain-specific compiler for local assembly kernels. COFFEE manipulates
+abstract syntax trees generated from a high-level domain-specific
+language for PDEs by introducing domain-aware composable optimizations
+aimed at improving instruction-level parallelism, especially SIMD
+vectorization, and register locality. It then generates C code including
+vector intrinsics. Experiments using a range of finite-element forms of
+increasing complexity show that significant performance improvement is
+achieved.
 %
@@
 \begin{alltt}
-\footnotesize \textbf{Input:} element matrix (2D array, initialized to
+\footnotesize
+\textbf{Input:} element matrix (2D array, initialized to
 0), element coordinates (array), coefficient fields (array,
-e.g. velocity) \textbf{Output:} element matrix (2D array) - Compute
-Jacobian from coordinates - Define basis functions - Compute element
-matrix in an affine loop nest
+e.g. velocity)
+\textbf{Output:} element matrix (2D array)
+- Compute Jacobian from coordinates
+- Define basis functions
+- Compute element matrix in an affine loop nest
 \end{alltt}
@@
 
-Local assembly is the computation of contributions of a specific cell
-in the discretized domain to the PDE solution. The process consists of
-numerically evaluating problem-specific integrals to produce a matrix
-and a vector (\cite{quadrature1}, \cite{fluidity_manual_v4}), whose
-sizes depend on the order of the method. This operation is applied to
-all cells in the discretized domain. In this work we focus on local
-matrices, or ``element matrices,'' which are more costly to compute
-than element vectors.
+Local assembly is the computation of contributions of a specific cell in
+the discretized domain to the linear system which yields the PDE
+solution. The process consists of numerically evaluating
+problem-specific integrals to produce a matrix and a vector
+\cite{quadrature1,fluidity_manual_v4}, whose sizes depend on the order
+of the method. This operation is applied to all cells in the discretized
+domain. In this work we focus on local matrices, or ``element
+matrices'', which are more costly to compute than element vectors.
 
 Given a finite element description of the input problem, expressed
-through the domain-specific Unified Form Language~\cite{ufl} (UFL),
-Firedrake generates a C-code kernel implementing assembly using a
-numerical integration algorithm. This kernel can be applied to any
-element in the mesh: this descends from the mathematical formulation
-of assembly, which states that the evaluation of a local matrix can be
-reduced to integrating on a fixed ``reference'' element---a special
-element that does not belong to the domain---after a suitable change
+through the domain-specific Unified Form Language (UFL)~\cite{ufl},
+Firedrake employs the FEniCS form compiler (FFC)~\cite{FFC-Compiler} to
+generate a C-code kernel implementing assembly using a
+numerical quadrature. This kernel can be applied to any
+element in the mesh, which follows from a mathematical property of the
+finite element method. The evaluation of a local matrix can be
+reduced to integration on a fixed ``reference'' element --- a special
+element that does not belong to the domain --- after a suitable change
 of coordinates. Firedrake triggers the compilation of an assembly
@@
 element. The output is the evaluated element matrix. The kernel body
-can be logically split into three parts: 1) calculation of the
-Jacobian matrix, its determinant and its inverse: required for the
-aforementioned change of coordinates, from the reference element to
-the one being computed; 2) definition of basis functions: used,
-intuitively, to interpolate the contribution to the PDE solution over
-the element. The choice of the basis functions is expressed in UFL
-directly by users. In the generated code, they are represented as
-global read-only two dimensional arrays (i.e., using \texttt{static
-  const} in C) of double precision floats; 3) evaluation of the
-element matrix in an affine loop nest, in which the integration is
-performed. Table~\ref{table:map-name-letters} shows the variable names
+can be logically split into three parts:
+\begin{enumerate}
+  \item Calculation of the Jacobian matrix, its determinant and its
+    inverse required for the aforementioned change of coordinates from
+    the reference element to the one being computed.
+  % FIXME: this rather hand-wavy and not technically correct
+  \item Definition of basis functions used, intuitively, to interpolate
+    the contribution to the PDE solution over the element. The choice of
+    basis functions is expressed in UFL directly by users. In the
+    generated code, they are represented as global read-only two
+    dimensional arrays (i.e., using \texttt{static const} in C) of
+    double precision floats.
+  \item Evaluation of the element matrix in an affine loop nest, in which
+    the integration is performed.
+\end{enumerate}
+Table~\ref{table:map-name-letters} shows the variable names
 we will use in the upcoming code snippets to refer to the various
@@
 $\rbrace$
-\caption{Local assembly code generated by Firedrake for a Burgers problem on a 3D tetrahedral mesh using Lagrange $p=1$ elements.}
+\caption{Local assembly code generated by Firedrake for a Burgers
+problem on a 3D tetrahedral mesh using Lagrange $p=1$ elements.}
 \label{code:burgers}
@@
 element problem being solved. In simpler cases, the loop nest is
-perfect, it has short trip counts (in the range 3--15), and the
-computation reduces to a summation of a few products involving basis
-functions. An example is provided in Listing~\ref{code:helmholtz},
-which shows the assembly kernel for a Helmholtz problem using Lagrange
-basis functions on 2D elements with polynomial order $p=1$. In other
-scenarios, for instance when solving the Burgers equation, whose
-assembly code is given in Listing~\ref{code:burgers}, the number of
-arrays involved in the computation of the element matrix can be much
-larger: in this case, 14 unique arrays are accessed, and the same
-array can be referenced multiple times within the expression. This may
-also require the evaluation of constants in outer loops (called $F$ in
-the code) to act as scaling factors of arrays; trip counts can be
-larger (proportionally to the order of the method); and arrays may be
-block-sparse. In addition to a larger number of operations, more
-complex cases like the Burgers equation are characterized by high
-register pressure.
-
-The Helmholtz and Burgers equations are exponents of a large class of
-problems and, as such, will constitute our benchmark, along with the
+perfect, has short trip counts (in the range 3--15), and the computation
+reduces to a summation of a few products involving basis functions. An
+example is provided in Listing~\ref{code:helmholtz}, which shows the
+assembly kernel for a Helmholtz problem using Lagrange basis functions
+on 2D elements with polynomial order $p=1$. In other scenarios, for
+instance when solving the Burgers equation, the number of arrays
+involved in the computation of the element matrix can be much larger.
+The assembly code is given in Listing~\ref{code:burgers} and contains 14
+unique arrays that are accessed, where the same array can be referenced
+multiple times within the same expression. This may also require the
+evaluation of constants in outer loops (called $F$ in the code) to act
+as scaling factors of arrays. Trip counts grow proportionally to the
+order of the method and arrays may be block-sparse. In addition to a
+larger number of operations, more complex cases like the Burgers
+equation are characterized by high register pressure.
+
+The Helmholtz and Burgers equations exemplify a large class of problems
+and, as such, will constitute our benchmark problems, along with the
 Diffusion equation. We carefully motivate this choice in
@@
 SIMD vectorization. These include: 1) memory accesses along the three
-loop dimensions are always stride-1; 2) the \texttt{j} and \texttt{k}
+loop dimensions are always unit stride; 2) the \texttt{j} and \texttt{k}
 loops are interchangeable, whereas interchanges involving the $i$ loop

### 2014-06-30 TACO-paper/coffee.tex
Copy editing of background
Copy editing of code transformations

@@
 The numerical solution of partial differential equations using the
-finite element method is one of the key applications of high performance
-computing. Local assembly is its characteristic operation. This entails
-the execution of a problem-specific kernel to numerically evaluate an
-integral for each element in the discretized problem domain. Since the
-domain size can be huge, executing efficient kernels is fundamental.
-Their optimization is, however, a challenging issue. Even though affine
-loop nests are generally present, the short trip counts and the
-complexity of mathematical expressions make it hard to determine a
-single or unique sequence of successful transformations. Therefore, we
-present the design and systematic evaluation of COFFEE, a
-domain-specific compiler for local assembly kernels. COFFEE manipulates
-abstract syntax trees generated from a high-level domain-specific
-language for PDEs by introducing domain-aware composable optimizations
-aimed at improving instruction-level parallelism, especially SIMD
-vectorization, and register locality. It then generates C code including
-vector intrinsics. Experiments using a range of finite-element forms of
-increasing complexity show that significant performance improvement is
-achieved.
+finite element method is one of the key applications of high
+performance computing. Local assembly is its characteristic
+operation. This entails the execution of a problem-specific kernel for
+each element in the discretized problem domain. Since the domain size
+can be huge, executing efficient kernels is fundamental. Their
+optimization is, however, a challenging issue. Even though affine loop
+nests are generally present, the short trip counts and the complexity
+of mathematical expressions make it hard to determine a single or
+unique sequence of successful transformations. Therefore, we present
+the design and systematic evaluation of COFFEE, a domain-specific
+compiler for local assembly kernels. COFFEE manipulates abstract
+syntax trees generated from a high-level domain-specific language for
+PDEs by introducing domain-aware composable optimizations aimed at
+improving instruction-level parallelism, especially SIMD
+vectorization, and register locality. It then generates C code
+including vector intrinsics. Experiments using a range of
+finite-element forms of increasing complexity show that significant
+performance improvement is achieved.
 %
@@
 systematic way we need for our assembly kernels. We want to overcome
-two deficiencies that both $intel$ and $gnu$ compilers exhibit. First,
+two deficiencies that both Intel and GNU compilers exhibit. First,
 they only identify sub-expressions that are invariant with respect to
@@
 \texttt{i}-\texttt{k}, which are not automatically lifted in the loop
-order \texttt{ijk}. Second, the hoisted code is scalar, i.e., it is not
+order \texttt{ijk}. Second, the hoisted code is scalar and therefore not
 subjected to auto-vectorization.
@@
 pressure is high. Since the kernel's working set fits the L1 cache, it
-is remarkably important to optimize register management. Standard
+is particularly important to optimize register management. Standard
 optimizations, such as loop interchange, unroll, and unroll-and-jam,
@@
   y \in B'' \subseteq B}}
-x_j\cdot y_k ~~~~~~ j,k = 0,...,3
+x_j\cdot y_k ~~~~~~ j,k = 0,...,2
 \end{equation}
@@
 independent of both \texttt{j} and \texttt{k}, the element matrix
-computation is always reducible to this
-form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16
-entries ($j,k=0,...,3$) of the element matrix using just 2 vector
-registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert
-= \vert B'' \vert = 1$. Values in a register are shuffled each time a
-product is performed. Standard compiler auto-vectorization ($gnu$ and
-$intel$), instead, executes 4 broadcast operations (i.e., ``splat'' of
-a value over all of the register locations) along the outer dimension
-to perform the calculation. In addition to incurring in larger number
-of cache accesses, it needs to keep between $f=1$ and $f=3$ extra
-registers to perform the same 16 evaluations when unroll-and-jam is
-used, with $f$ being the unroll-and-jam factor.
+computation is always reducible to this form.
+Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries
+($j,k=0,...,3$) of the element matrix using just 2 vector registers,
+which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B''
+\vert = 1$. Values in a register are shuffled each time a product is
+performed. Standard compiler auto-vectorization for both GNU and Intel
+compilers, instead, executes 4 broadcast operations (i.e., ``splat'' of
+a value over all of the register locations) along the outer dimension to
+perform the calculation. In addition to incurring a larger number of
+cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers
+to perform the same 16 evaluations when unroll-and-jam is used, with $f$
+being the unroll-and-jam factor.
 
@@
 innermost-loop dimension is close to or greater than the number of
-available CPU registers, poor register reuse is likely to be
-obtained. This usually happens when the number of basis function
-arrays, temporaries introduced by generalized loop-invariant code
-motion, and problem constants is large. For example, applying
-loop-invariant code motion to Burgers on a 3D mesh requires 24
-temporaries for the \texttt{ijk} loop order. This can make hoisting of
-the invariant loads out of the \texttt{k} loop inefficient on
-architectures with a relatively low number of registers. One potential
-solution to this problem consists of suitably ``splitting'' the
-computation of the element matrix $A$ into multiple sub-expressions;
-an example, for the Helmholtz problem, is given in
+available CPU registers, poor register reuse is likely. This usually
+happens when the number of basis function arrays, temporaries introduced
+by generalized loop-invariant code motion, and problem constants is
+large. For example, applying loop-invariant code motion to Burgers on a
+3D mesh requires 24 temporaries for the \texttt{ijk} loop order. This
+can make hoisting of the invariant loads out of the \texttt{k} loop
+inefficient on architectures with a relatively low number of registers.
+One potential solution to this problem consists of suitably
+``splitting'' the computation of the element matrix $A$ into multiple
+sub-expressions; an example, for the Helmholtz problem, is given in
 Listing~\ref{code:helmholtz-split}. The transformation can be regarded
-as a special case of classic loop fission, in which sum's
-associativity is exploited to distribute the expression across
-multiple loops. To the best of our knowledge, expression splitting is
-not supported by available compilers.
+as a special case of classic loop fission, in which associativity of the
+sum is exploited to distribute the expression across multiple loops. To
+the best of our knowledge, expression splitting is not supported by
+available compilers.
 
@@
 depending on how the split is executed, it can lead to redundant
-computation; for example, the times the product $det*W3[i]$ is
-performed is proportional to the number of sub-expressions, as shown
-in the code snippet. Further, it increases loop overhead (e.g. branch
-instructions). Finally, it might affect register locality: for
-instance, the same array could be accessed in different
-sub-expressions, requiring a proportional number of loads be
-performed. This is not the case of the Helmholtz
-example. Nevertheless, as shown in Section~\ref{sec:perf-results}, the
-performance gain from improved register reuse along inner dimensions
-can still be greater, especially if the split factor and the splitting
-itself use heuristics to minimize the aforementioned issues.
-
-\section*{}
+computation. For example, the number of times the product $det*W3[i]$ is
+performed is proportional to the number of sub-expressions, as shown in
+the code snippet. Further, it increases loop overhead, for example
+through additional branch instructions. Finally, it might affect
+register locality: for instance, the same array could be accessed in
+different sub-expressions, requiring a proportional number of loads be
+performed. This is not the case for the Helmholtz example. Nevertheless,
+as shown in Section~\ref{sec:perf-results}, the performance gain from
+improved register reuse along inner dimensions can still be greater,
+especially if the split factor and the splitting itself use heuristics
+to minimize the aforementioned issues.
 
@@
 transformations depend on some parameters (e.g. tile size), we need a
-mechanism to prune such a large space of optimization. This aspect is
+mechanism to prune such a large optimization space. This aspect is
 treated in Section~\ref{sec:pyop2-compiler}.

### 2014-06-30 TACO-paper/coffee.tex
Copy editing of code transformations
Copy editing of coffee overview

@@
 
+% FIXME: ``mesh used to discretize''
 %\rule[0.2ex]{16cm}{0.5pt}
@@
 
-Firedrake users employ the Unified Form Language to write problems in
+Firedrake users employ the Unified Form Language to express problems in
 a notation resembling mathematical equations. At run-time, the
 high-level specification is translated by a modified version of the
-FEniCS Form Compiler~\cite{FFC-Compiler} (FFC) into an abstract syntax
+FEniCS Form Compiler (FFC)~\cite{FFC-Compiler} into an abstract syntax
 tree (AST) representation of one or more finite element assembly
-kernels. ASTs are then passed to COFFEE, which is capable of applying
-the transformations described in Section~\ref{sec:code-transf}. The
-output of COFFEE, C code, is eventually provided to
-PyOP2~\cite{pyop2isc}, where just-in-time compilation and execution
-over the discretized domain take place. The flow is summarized in
-Figure~\ref{fig:coffee-pipeline}.
+kernels. ASTs are then passed to COFFEE to apply the transformations
+described in Section~\ref{sec:code-transf}. The output of COFFEE, C
+code, is eventually provided to PyOP2~\cite{pyop2ws,pyop2isc}, where
+just-in-time compilation and execution over the discretized domain take
+place. The flow is summarized in Figure~\ref{fig:coffee-pipeline}.
 
@@
 \footnotesize
-\textbf{Input:} ast: abstract syntax tree of a local assembly kernel
-produced by FFC wrapper: the function within which the kernel is
-invoked isa: represents the SIMD Instruction Set Architecture
+\textbf{Input:}
+ast: abstract syntax tree of a local assembly kernel produced by FFC
+wrapper: the function within which the kernel is invoked
+isa: represents the SIMD Instruction Set Architecture
 \textbf{Output:} assembly code (C string)
@@
 the backend compiler in charge of selecting unroll factors. This
-choice also simplifies the COFFEE's cost model. The only situation in
+choice also simplifies COFFEE's cost model. The only situation in
 which we explicitly unroll-and-jam a loop is when \emph{op-vect} is
@@
 into account the number of available logical vector registers,
-\emph{n\_regs}, and the number of unique variables accessed:
-\emph{n\_consts} counts variables independent of both \texttt{j} and
-\texttt{k} loops and temporary registers, \emph{n\_outer\_arrays}
-counts \texttt{j}-dependent variables, and \emph{n\_inner\_arrays}
+\texttt{n\_regs}, and the number of unique variables accessed:
+\texttt{n\_consts} counts variables independent of both \texttt{j} and
+\texttt{k} loops and temporary registers, \texttt{n\_outer\_arrays}
+counts \texttt{j}-dependent variables, and \texttt{n\_inner\_arrays}
 counts \texttt{k}-dependent variables, assuming the \texttt{ijk} loop

### 2014-06-30 TACO-paper/coffee.tex
Copy editing of coffee overview
Copy editing of results

@@
 \end{enumerate}
-The three chosen benchmarks are \emph{real-life kernels} and comprise
-the core differential operators in some of the most frequently
-encountered finite element problems in scientific computing. The
-Helmholtz and Diffusion kernels are archetypal second order elliptic
-operators. They are complete and unsimplified examples of the
-operators occurring for diffusion and viscosity in fluids, and for
-imposing pressure in compressible fluids. As such, they are both
-extensively used in climate and ocean modeling. Very similar
+The three chosen benchmarks comprise the core differential operators in
+some of the most frequently encountered finite element problems in
+scientific computing. The Helmholtz and Diffusion kernels are archetypal
+second order elliptic operators. They are complete and unsimplified
+examples of the operators used to model diffusion and viscosity in
+fluids, and for imposing pressure in compressible fluids. As such, they
+are both extensively used in climate and ocean modeling. Very similar
 operators, for which the same optimisations are expected to be equally
-effective, apply to elasticity problems, which are at the base of
+effective, apply to elasticity problems, which are a foundation of
 computational structural mechanics. The Burgers kernel is a typical
@@
 real applications whenever a quantity is transported by a fluid (the
-momentum itself, in our case). We chose this particular kernel since
-it applies to a vector-valued quantity, while the elliptic operators
-apply to scalar quantities; this impacts the generated code, as
-explained next. In this way, the operators we have selected are
-characteristic of both the second and first order operators that
-dominate in fluids and solids simulations, and a wide variety on local
-assembly codes can be tested.
-
-The benchmarks were written in UFL (code available at~\cite{ufl-code})
+momentum itself, in our case). We chose this particular kernel since it
+applies to a vector-valued quantity, while the elliptic operators apply
+to scalar quantities. This impacts the generated code, as explained
+next. The operators we have selected are characteristic of both the
+second and first order operators that dominate simulations in fluids and
+solids, and a wide variety of local assembly codes can be tested.
+
+The benchmarks, available at~\cite{ufl-code}, were written in UFL
 and executed over real unstructured meshes through Firedrake. The
 Helmholtz code has already been shown in
-Listing~\ref{code:helmholtz}. For Advection-Diffusion, the Diffusion
+Listing~\ref{code:helmholtz}. For advection-diffusion, the diffusion
 equation, which uses the same differential operators as Helmholtz, is
-considered. In the Diffusion kernel code, the main differences with
+considered. In the diffusion kernel code, the main differences with
 respect to Helmholtz are the absence of the $Y$ array and the presence
-of a few more constants for computing the element matrix. Burgers is a
-non-linear problem employing differential operators different from
+of additional constants for computing the element matrix. Burgers is a
+nonlinear problem employing differential operators different from
 those of Helmholtz and relying on vector-valued quantities, which has
@@
 and the polynomial order $p$ of the method, whereas the element
-family, Lagrange, is fixed. As might be expected, the larger the
+family, Lagrange, is fixed. As expected, the larger the
 element shape and $p$, the larger the iteration space. Triangles,
-tetrahedron, and prisms were tested as element shape. For instance, in
+tetrahedra, and prisms were tested as element shape. For instance, in
 the case of Helmholtz with $p=1$, the size of the \texttt{j} and
@@
 (the three loops \texttt{i}, \texttt{j}, and \texttt{k}, and, as a
-consequence, the size of $X$ and $Y$ arrays). A range of polynomial
-orders from $p=1$ to $p=4$ were tested; higher polynomial orders are
+consequence, the size of the $X$ and $Y$ arrays). A range of polynomial
+orders from $p=1$ to $p=4$ were tested. Higher polynomial orders are
 excluded from the study because of current Firedrake limitations. In
@@
 
+% FIXME: weren't those FIAT limitations?
+
 \subsection{Loop permutation}
+
 In the following, only results for the loop order \texttt{ijk} are
 shown. For the considerations exposed in
-Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner
-loop along \texttt{i} caused slow downs; also, interchanging
-\texttt{j} and \texttt{k} loops while keeping \texttt{i} as outermost
-loop did not provide any benefits.
+Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner loop
+along \texttt{i} caused slow downs; also, interchanging \texttt{j} and
+\texttt{k} loops while keeping \texttt{i} as outermost loop did not
+provide any benefits.
 
@@
 improvement obtained when \emph{licm} is applied. In general, the
-speed ups are notable. The main reasons were anticipated in
+speedups are notable. The main reasons were anticipated in
 Section~\ref{sec:code-transf-licm}: in the original code, 1)
@@
 bigger loop nests (i.e., larger element shapes and polynomial orders)
-usually benefit from the reduction in redundant computation, even
-though extra memory for the temporary arrays is required. Some
-discrepancies to this trend are due to a less effective
-auto-vectorization. For instance, on the Sandy Bridge, the improvement
-at $p=3$ is larger than that at $p=4$ because, in the latter case, the
-size of the innermost loop is not a multiple of the vector length, and
-a reminder scalar loop is introduced at compile time. Since the loop
-nest is small, the cost of executing the extra scalar iterations can
-have a significant impact. The reminder loop overhead is more
-pronounced on the Xeon Phi, where the vector length is twice as big,
-which leads to proportionally larger scalar reminder loops.
+usually benefit from the reduction in redundant computation, even though
+extra memory for the temporary arrays is required.  Exceptions to this
+trend are due to a less effective auto-vectorization. For instance, on
+the Sandy Bridge, the improvement at $p=3$ is larger than that at $p=4$
+because, in the latter case, the size of the innermost loop is not a
+multiple of the vector length, and a remainder scalar loop is introduced
+at compile time. Since the loop nest is small, the cost of executing the
+extra scalar iterations can have a significant impact. The remainder
+loop overhead is more pronounced on the Xeon Phi, where the vector
+length is twice as big, which leads to proportionally larger scalar
+remainder loops.
 
@@
 Table~\ref{table:perf-results-licmap} shows the cumulative impact of
-\emph{licm}, data alignment, and padding over the original code. In
-the following, this version of the code is referred to as
-\emph{licm-ap}. Padding, which avoids the introduction of a reminder
-loop as described in Section~\ref{licm-impact}, as well as data
-alignment, enhance the quality of auto-vectorization. Occasionally the
-run-time of \emph{licm-ap} is close to that of \emph{licm}, since the
-non-padded element matrix size is already a multiple of the vector
-length. Rarely \emph{licm-ap} is slower than \emph{licm} (e.g. in
-Burgers $p=3$ on the Sandy Bridge). One possible explanation is that
-the number of aligned temporaries introduced by \emph{licm} is so
-large to induce cache associativity conflicts.
+\emph{licm}, data alignment, and padding over the original code. In the
+following, this version of the code is referred to as \emph{licm-ap}.
+Padding, which avoids the introduction of a remainder loop as described
+in Section~\ref{licm-impact}, as well as data alignment, enhance the
+quality of auto-vectorization. Occasionally the run-time of
+\emph{licm-ap} is close to that of \emph{licm}, since the non-padded
+element matrix size is already a multiple of the vector length. In rare
+cases \emph{licm-ap} is slower than \emph{licm} (e.g. in Burgers $p=3$
+on the Sandy Bridge). One possible explanation is that the number of
+aligned temporaries introduced by \emph{licm} is so large to induce
+cache associativity conflicts.
 
@@
 \label{sec:perf-eval-opvect}
+
 In this section, we evaluate the impact of vector-register tiling. We
@@
 
-It is worth noticing that, in most cases, the cost model successfully
+It is worth noting that, in most cases, the cost model successfully
 determines how to transform a kernel to maximize its performance. This
-is chiefly because assembly kernels fit the L1 cache, so, within a
+is chiefly because assembly kernels fit the L1 cache, such that, with a
 certain degree of confidence, it is possible to predict how to obtain
@@
 Sandy Bridge, we confirmed that speed ups are a consequence of
-increased register re-use. In Helmholtz $p=4$, for example, the tool
+improved register re-use. In Helmholtz $p=4$, for example, the tool
 showed that when using \emph{op-vect} the number of clock cycles to
@@
 \label{sec:perf-results-split}
+
 Expression splitting relieves the register pressure when the element
-matrix evaluation needs to read from a large number of basis function
-arrays. As detailed in Section~\ref{sec:expr-split}, in the price to
-pay for this optimazion, there are an increased number of accesses to
-the element matrix and, potentially, redundant computation. Similarly
-to the analysis of vector-register tiling, we compare two versions:
-the baseline, \emph{licm-ap}; and expression splitting on top of
-\emph{licm-ap}, which, for simplicity, in the following is referred to
-as \emph{split}.
-
-For the Helmholtz and Diffusion kernels, in which only between 4 and 8
+matrix evaluation needs to be read from a large number of basis function
+arrays. As detailed in Section~\ref{sec:expr-split}, the price to pay
+for this optimazion is an increased number of accesses to the element
+matrix and, potentially, redundant computation. Similarly to the
+analysis of vector-register tiling, we compare two versions: the
+baseline, \emph{licm-ap}; and expression splitting on top of
+\emph{licm-ap}, which, for simplicity, is referred to as \emph{split} in
+the following.
+
+For the Helmholtz and diffusion kernels, in which only between 4 and 8
 temporaries are read at every loop iteration, \texttt{split} tends to
-slow down the computation, because of the aforementioned
-drawbacks. Slow downs up to 1.4$\times$ and up to 1.6$\times$ were
-observed, respectively, on the Sandy Bridge and the Xeon Phi. Note
-that the cost model prevents the adoption of the transformation: the
-\texttt{while} statement in Figure~\ref{algo:applyCostModel} is indeed
-never entered.
+slow down the computation, because of the aforementioned drawbacks. Slow
+downs up to 1.4$\times$ and up to 1.6$\times$ respectively were observed
+on the Sandy Bridge and the Xeon Phi. Note that the cost model prevents
+the adoption of the transformation: the \texttt{while} statement in
+Figure~\ref{algo:applyCostModel} is indeed never entered.
 
@@
 $p=4$ and a prismatic mesh is employed, the overall performance
-improvement (i.e., the one over the original code) increases from
-1.44$\times$ to 2.11$\times$. On the Xeon Phi, the impact of
-\emph{split} is only marginal, since register spilling is limited by
-the presence of 32 logical vector units.
+improvement over the original code increases from 1.44$\times$ to
+2.11$\times$. On the Xeon Phi, the impact of \emph{split} is only
+marginal, since register spilling is limited by the presence of 32
+logical vector units.
 
@@
 1.22$\times$, \emph{split-op-vect} never outperforms
-\emph{split}. This is motivated by two factors: for small split
+\emph{split}. This is explained by two factors: for small split
 factors, such as 1 and 2, the data space to be tiled is still too big,
-and register spilling affects run-time; for higher ones,
+and register spilling affects run-time. For higher factors,
 sub-expressions become so small that, as explained in
@@
 \label{sec:perf-results-fenics}
+
 We have modified the FEniCS Form Compiler (FFC) to return an abstract
-syntax tree representation of a local assembly kernel, rather than
-plain C++ code, so as to enable code transformations in
-COFFEE. Besides Firedrake, FFC is used in the FEniCS
-project~\cite{Fenics}. In FEniCS, FFC can apply its own model-driven
-optimizations to local assembly kernels~\cite{quadrature1}, which
-mainly consist of loop-invariant code motion and elimination, at code
+syntax tree representation of a local assembly kernel, rather than plain
+C++ code to enable code transformations in COFFEE. In the FEniCS
+project~\cite{Fenics}, FFC's role is generating an optimized C++ code
+string confirming to the UFC interface. The model-driven optimizations
+to local assembly kernels~\cite{quadrature1} supported by FFC mainly
+consist of loop-invariant code motion and elimination, at code
 generation time, of floating point operations involving zero-valued
@@
 example applying distributivity and associativity to products and sums
-at code generation time, to identify terms that are invariant of the
+at code generation time to identify terms that are invariant of the
 whole loop nest. Depending on the way expansion is performed,
-operation count may not decrease significantly.
+operation counts may not decrease significantly.
 
@@
 on the Sandy Bridge for the Burgers, Helmholtz and Diffusion
-kernels. Burgers' slow downs occur in presence of a small iteration
-space (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1,
-  2]$; prismatic mesh, $p = 1$). The result shown represents the worst
-slow down, which was obtained with a triangular mesh and $p = 1$. This
-is a result of removing zero-valued entries in FEniCS' basis function
-arrays: some operations are avoided, but indirection arrays prevent
-auto-vectorization, which significantly impacts performance as soon as
-the element matrix becomes bigger than 12$\times$12. However, with the
-forthcoming zero-removal optimization in COFFEE, we expect to
-outperform FEniCS in all problems. In the cases of Helmholtz and
-Diffusion, the minimum improvements are, respectively, 1.10$\times$
-and 1.18$\times$ (2D mesh, $p=1$), which tend to increase with
-polynomial order and element shape up to the values illustrated in the
-table.
+kernels. Slow downs in the Burgers benchmark occur in presence of a
+small iteration space (triangular mesh, $p \in [1, 2]$; tetrahedral
+mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$). The result shown
+represents the worst slow down, which was obtained with a triangular
+mesh and $p = 1$. This is a result of removing zero-valued entries in
+basis function arrays: some operations are avoided, but indirection
+arrays prevent auto-vectorization, which significantly impacts
+performance as soon as the element matrix becomes bigger than
+12$\times$12. However, with the forthcoming zero-removal optimization in
+COFFEE, we expect to outperform FEniCS in all problems. In the cases of
+Helmholtz and Diffusion, the minimum improvements are, respectively,
+1.10$\times$ and 1.18$\times$ (2D mesh, $p=1$), which tend to increase
+with polynomial order and element shape up to the values illustrated in
+the table.
 
@@
 \label{sec:perf-eval-blas}
+
 For the Helmholtz problem on a tetrahedral mesh, manual

### 2014-06-30 TACO-paper/coffee.tex
Copy editing of related work

@@
 \label{sec:related-work}
+
 The finite element method is extensively used to approximate solutions
@@
 nek5000~\cite{nek5000-web-page}, the FEniCS project~\cite{Fenics},
-Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake. Numerical
-integration is usually employed to implement the local assembly
-phase. The recent introduction of domain specific languages (DSLs) to
-decouple the finite element specification from its underlying
-implementation facilitated, however, the development of novel
-approaches. Methods based on tensor contraction~\cite{FFC-Compiler}
-and symbolic manipulation~\cite{Francis} have been implemented. We
-have designed COFFEE to specifically target numerical integration
-because it has been demonstrated that it remains the optimal choice
-for a wide class of problems~\cite{quadrature1}.
+Fluidity~\cite{fluidity_manual_v4}. Numerical integration is usually
+employed to implement the local assembly phase. The recent introduction
+of domain-specific languages (DSLs) to decouple the finite element
+specification from its underlying implementation facilitated the
+development of novel approaches. Methods based on tensor
+contraction~\cite{FFC-Compiler} and symbolic manipulation~\cite{Francis}
+have been implemented. We have designed COFFEE to specifically target
+numerical integration because it has been demonstrated that it remains
+the optimal choice for a wide class of problems~\cite{quadrature1}.
 
 Optimization of local assembly using numerical integration for CPU
-platforms has been addressed in FEniCS~\cite{quadrature1}. The
-comparison between COFFEE and this work is presented in
+platforms has been addressed in FEniCS~\cite{quadrature1} and a
+comparison to COFFEE is presented in
 Section~\ref{sec:perf-results-fenics}. In~\cite{Markall20101815}, and
-more recently in~\cite{petsc-integration-gpu}, the problem has been
-studied for GPU architectures. In~\cite{assembly-opencl}, variants of
-the standard numerical integration algorithm have been specialized and
-evaluated for the PowerXCell processor, but an exhaustive study from
-the compiler viewpoint - like ours - is missing, and none of the
-optimizations presented in Section~\ref{sec:code-transf} are
-mentioned. Among these efforts, to the best of our knowledge, COFFEE
-is the first work targeting low-level optimizations through a real
-compiler approach.
+more recently in~\cite{petsc-integration-gpu}, numerical integration has
+been studied for GPU architectures. In~\cite{assembly-opencl}, variants
+of the standard numerical integration algorithm have been specialized
+and evaluated for the PowerXCell processor, but an exhaustive study from
+the compiler viewpoint --- like ours --- is missing, and none of the
+optimizations presented in Section~\ref{sec:code-transf} are mentioned.
+Among these efforts, to the best of our knowledge, COFFEE is the first
+work targeting low-level optimizations through a real compiler approach.
 
@@
 proven successful in auto-generating optimized code for other domains:
-Spiral~\cite{Pueschel:05} for digital signal processing numerical
-algorithms, ~\cite{Spampinato:14} for dense linear algebra, or
+Spiral~\cite{Pueschel:05} for digital signal processing algorithms,
+~\cite{Spampinato:14} for dense linear algebra, or
 Pochoir~\cite{pochoir} and SDSL~\cite{stencil-compiler} for image
@@
 Firedrake to express iteration over unstructured meshes in scientific
-codes. COFFEE improves automated code generation in Firedrake.
+codes. COFFEE improves the efficiency of automatically generated kernels
+in Firedrake.
 

### 2014-07-01 TACO-paper/coffee.tex
Copy editing of conclusions

@@
 
-In this paper, we have presented design, optimizations and systematic
+In this paper, we have presented the design, optimizations and systematic
 performance evaluation of COFFEE, a compiler for finite element local
@@
 maximize instruction-level parallelism, register locality and SIMD
-vectorization. Assembly kernels have peculiar characteristics. Their
+vectorization. Assembly kernels have particular characteristics. Their
 iteration space is usually very small, with the size depending on
@@
 demonstrated by our experiments on two state-of-the-art Intel
-platforms. Further work include a comprehensive study about
+platforms. Further work includes a comprehensive study about
 feasibility and constraints on transforming kernels into a sequence of
 calls to external linear algebra libraries. COFFEE supports all of the
-problems expressible in Firedrake, and it is already integrated with
+problems expressible in Firedrake, and is already integrated with
 this framework.

### 2014-07-01 TACO-paper/coffee.tex
Copy editing of conclusions

--- 2014-06-30 TACO-paper/coffee.tex
@@
 \end{enumerate}
-The three chosen benchmarks comprise the core differential operators in
-some of the most frequently encountered finite element problems in
-scientific computing. The Helmholtz and Diffusion kernels are archetypal
-second order elliptic operators. They are complete and unsimplified
-examples of the operators used to model diffusion and viscosity in
-fluids, and for imposing pressure in compressible fluids. As such, they
-are both extensively used in climate and ocean modeling. Very similar
+The three chosen benchmarks are \emph{real-life kernels} and comprise
+the core differential operators in some of the most frequently
+encountered finite element problems in scientific computing. The
+Helmholtz and Diffusion kernels are archetypal second order elliptic
+operators. They are complete and unsimplified examples of the
+operators occurring for diffusion and viscosity in fluids, and for
+imposing pressure in compressible fluids. As such, they are both
+extensively used in climate and ocean modeling. Very similar
 operators, for which the same optimisations are expected to be equally
-effective, apply to elasticity problems, which are a foundation of
+effective, apply to elasticity problems, which are at the base of
 computational structural mechanics. The Burgers kernel is a typical
@@
 real applications whenever a quantity is transported by a fluid (the
-momentum itself, in our case). We chose this particular kernel since it
-applies to a vector-valued quantity, while the elliptic operators apply
-to scalar quantities. This impacts the generated code, as explained
-next. The operators we have selected are characteristic of both the
-second and first order operators that dominate simulations in fluids and
-solids, and a wide variety of local assembly codes can be tested.
-
-The benchmarks, available at~\cite{ufl-code}, were written in UFL
+momentum itself, in our case). We chose this particular kernel since
+it applies to a vector-valued quantity, while the elliptic operators
+apply to scalar quantities; this impacts the generated code, as
+explained next. In this way, the operators we have selected are
+characteristic of both the second and first order operators that
+dominate in fluids and solids simulations, and a wide variety on local
+assembly codes can be tested.
+
+The benchmarks were written in UFL (code available at~\cite{ufl-code})
 and executed over real unstructured meshes through Firedrake. The
 Helmholtz code has already been shown in
-Listing~\ref{code:helmholtz}. For advection-diffusion, the diffusion
+Listing~\ref{code:helmholtz}. For Advection-Diffusion, the Diffusion
 equation, which uses the same differential operators as Helmholtz, is
-considered. In the diffusion kernel code, the main differences with
+considered. In the Diffusion kernel code, the main differences with
 respect to Helmholtz are the absence of the $Y$ array and the presence
-of additional constants for computing the element matrix. Burgers is a
-nonlinear problem employing differential operators different from
+of a few more constants for computing the element matrix. Burgers is a
+non-linear problem employing differential operators different from
 those of Helmholtz and relying on vector-valued quantities, which has
@@
 and the polynomial order $p$ of the method, whereas the element
-family, Lagrange, is fixed. As expected, the larger the
+family, Lagrange, is fixed. As might be expected, the larger the
 element shape and $p$, the larger the iteration space. Triangles,
-tetrahedra, and prisms were tested as element shape. For instance, in
+tetrahedron, and prisms were tested as element shape. For instance, in
 the case of Helmholtz with $p=1$, the size of the \texttt{j} and
@@
 (the three loops \texttt{i}, \texttt{j}, and \texttt{k}, and, as a
-consequence, the size of the $X$ and $Y$ arrays). A range of polynomial
-orders from $p=1$ to $p=4$ were tested. Higher polynomial orders are
+consequence, the size of $X$ and $Y$ arrays). A range of polynomial
+orders from $p=1$ to $p=4$ were tested; higher polynomial orders are
 excluded from the study because of current Firedrake limitations. In
@@
 
-% FIXME: weren't those FIAT limitations?
-
 \subsection{Loop permutation}
-
 In the following, only results for the loop order \texttt{ijk} are
 shown. For the considerations exposed in
-Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner loop
-along \texttt{i} caused slow downs; also, interchanging \texttt{j} and
-\texttt{k} loops while keeping \texttt{i} as outermost loop did not
-provide any benefits.
+Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner
+loop along \texttt{i} caused slow downs; also, interchanging
+\texttt{j} and \texttt{k} loops while keeping \texttt{i} as outermost
+loop did not provide any benefits.
 
@@
 improvement obtained when \emph{licm} is applied. In general, the
-speedups are notable. The main reasons were anticipated in
+speed ups are notable. The main reasons were anticipated in
 Section~\ref{sec:code-transf-licm}: in the original code, 1)
@@
 bigger loop nests (i.e., larger element shapes and polynomial orders)
-usually benefit from the reduction in redundant computation, even though
-extra memory for the temporary arrays is required.  Exceptions to this
-trend are due to a less effective auto-vectorization. For instance, on
-the Sandy Bridge, the improvement at $p=3$ is larger than that at $p=4$
-because, in the latter case, the size of the innermost loop is not a
-multiple of the vector length, and a remainder scalar loop is introduced
-at compile time. Since the loop nest is small, the cost of executing the
-extra scalar iterations can have a significant impact. The remainder
-loop overhead is more pronounced on the Xeon Phi, where the vector
-length is twice as big, which leads to proportionally larger scalar
-remainder loops.
+usually benefit from the reduction in redundant computation, even
+though extra memory for the temporary arrays is required. Some
+discrepancies to this trend are due to a less effective
+auto-vectorization. For instance, on the Sandy Bridge, the improvement
+at $p=3$ is larger than that at $p=4$ because, in the latter case, the
+size of the innermost loop is not a multiple of the vector length, and
+a reminder scalar loop is introduced at compile time. Since the loop
+nest is small, the cost of executing the extra scalar iterations can
+have a significant impact. The reminder loop overhead is more
+pronounced on the Xeon Phi, where the vector length is twice as big,
+which leads to proportionally larger scalar reminder loops.
 
@@
 Table~\ref{table:perf-results-licmap} shows the cumulative impact of
-\emph{licm}, data alignment, and padding over the original code. In the
-following, this version of the code is referred to as \emph{licm-ap}.
-Padding, which avoids the introduction of a remainder loop as described
-in Section~\ref{licm-impact}, as well as data alignment, enhance the
-quality of auto-vectorization. Occasionally the run-time of
-\emph{licm-ap} is close to that of \emph{licm}, since the non-padded
-element matrix size is already a multiple of the vector length. In rare
-cases \emph{licm-ap} is slower than \emph{licm} (e.g. in Burgers $p=3$
-on the Sandy Bridge). One possible explanation is that the number of
-aligned temporaries introduced by \emph{licm} is so large to induce
-cache associativity conflicts.
+\emph{licm}, data alignment, and padding over the original code. In
+the following, this version of the code is referred to as
+\emph{licm-ap}. Padding, which avoids the introduction of a reminder
+loop as described in Section~\ref{licm-impact}, as well as data
+alignment, enhance the quality of auto-vectorization. Occasionally the
+run-time of \emph{licm-ap} is close to that of \emph{licm}, since the
+non-padded element matrix size is already a multiple of the vector
+length. Rarely \emph{licm-ap} is slower than \emph{licm} (e.g. in
+Burgers $p=3$ on the Sandy Bridge). One possible explanation is that
+the number of aligned temporaries introduced by \emph{licm} is so
+large to induce cache associativity conflicts.
 
@@
 \label{sec:perf-eval-opvect}
-
 In this section, we evaluate the impact of vector-register tiling. We
@@
 
-It is worth noting that, in most cases, the cost model successfully
+It is worth noticing that, in most cases, the cost model successfully
 determines how to transform a kernel to maximize its performance. This
-is chiefly because assembly kernels fit the L1 cache, such that, with a
+is chiefly because assembly kernels fit the L1 cache, so, within a
 certain degree of confidence, it is possible to predict how to obtain
@@
 Sandy Bridge, we confirmed that speed ups are a consequence of
-improved register re-use. In Helmholtz $p=4$, for example, the tool
+increased register re-use. In Helmholtz $p=4$, for example, the tool
 showed that when using \emph{op-vect} the number of clock cycles to
@@
 \label{sec:perf-results-split}
-
 Expression splitting relieves the register pressure when the element
-matrix evaluation needs to be read from a large number of basis function
-arrays. As detailed in Section~\ref{sec:expr-split}, the price to pay
-for this optimazion is an increased number of accesses to the element
-matrix and, potentially, redundant computation. Similarly to the
-analysis of vector-register tiling, we compare two versions: the
-baseline, \emph{licm-ap}; and expression splitting on top of
-\emph{licm-ap}, which, for simplicity, is referred to as \emph{split} in
-the following.
-
-For the Helmholtz and diffusion kernels, in which only between 4 and 8
+matrix evaluation needs to read from a large number of basis function
+arrays. As detailed in Section~\ref{sec:expr-split}, in the price to
+pay for this optimazion, there are an increased number of accesses to
+the element matrix and, potentially, redundant computation. Similarly
+to the analysis of vector-register tiling, we compare two versions:
+the baseline, \emph{licm-ap}; and expression splitting on top of
+\emph{licm-ap}, which, for simplicity, in the following is referred to
+as \emph{split}.
+
+For the Helmholtz and Diffusion kernels, in which only between 4 and 8
 temporaries are read at every loop iteration, \texttt{split} tends to
-slow down the computation, because of the aforementioned drawbacks. Slow
-downs up to 1.4$\times$ and up to 1.6$\times$ respectively were observed
-on the Sandy Bridge and the Xeon Phi. Note that the cost model prevents
-the adoption of the transformation: the \texttt{while} statement in
-Figure~\ref{algo:applyCostModel} is indeed never entered.
+slow down the computation, because of the aforementioned
+drawbacks. Slow downs up to 1.4$\times$ and up to 1.6$\times$ were
+observed, respectively, on the Sandy Bridge and the Xeon Phi. Note
+that the cost model prevents the adoption of the transformation: the
+\texttt{while} statement in Figure~\ref{algo:applyCostModel} is indeed
+never entered.
 
@@
 $p=4$ and a prismatic mesh is employed, the overall performance
-improvement over the original code increases from 1.44$\times$ to
-2.11$\times$. On the Xeon Phi, the impact of \emph{split} is only
-marginal, since register spilling is limited by the presence of 32
-logical vector units.
+improvement (i.e., the one over the original code) increases from
+1.44$\times$ to 2.11$\times$. On the Xeon Phi, the impact of
+\emph{split} is only marginal, since register spilling is limited by
+the presence of 32 logical vector units.
 
@@
 1.22$\times$, \emph{split-op-vect} never outperforms
-\emph{split}. This is explained by two factors: for small split
+\emph{split}. This is motivated by two factors: for small split
 factors, such as 1 and 2, the data space to be tiled is still too big,
-and register spilling affects run-time. For higher factors,
+and register spilling affects run-time; for higher ones,
 sub-expressions become so small that, as explained in
@@
 \label{sec:perf-results-fenics}
-
 We have modified the FEniCS Form Compiler (FFC) to return an abstract
-syntax tree representation of a local assembly kernel, rather than plain
-C++ code to enable code transformations in COFFEE. In the FEniCS
-project~\cite{Fenics}, FFC's role is generating an optimized C++ code
-string confirming to the UFC interface. The model-driven optimizations
-to local assembly kernels~\cite{quadrature1} supported by FFC mainly
-consist of loop-invariant code motion and elimination, at code
+syntax tree representation of a local assembly kernel, rather than
+plain C++ code, so as to enable code transformations in
+COFFEE. Besides Firedrake, FFC is used in the FEniCS
+project~\cite{Fenics}. In FEniCS, FFC can apply its own model-driven
+optimizations to local assembly kernels~\cite{quadrature1}, which
+mainly consist of loop-invariant code motion and elimination, at code
 generation time, of floating point operations involving zero-valued
@@
 example applying distributivity and associativity to products and sums
-at code generation time to identify terms that are invariant of the
+at code generation time, to identify terms that are invariant of the
 whole loop nest. Depending on the way expansion is performed,
-operation counts may not decrease significantly.
+operation count may not decrease significantly.
 
@@
 on the Sandy Bridge for the Burgers, Helmholtz and Diffusion
-kernels. Slow downs in the Burgers benchmark occur in presence of a
-small iteration space (triangular mesh, $p \in [1, 2]$; tetrahedral
-mesh, $p \in [1, 2]$; prismatic mesh, $p = 1$). The result shown
-represents the worst slow down, which was obtained with a triangular
-mesh and $p = 1$. This is a result of removing zero-valued entries in
-basis function arrays: some operations are avoided, but indirection
-arrays prevent auto-vectorization, which significantly impacts
-performance as soon as the element matrix becomes bigger than
-12$\times$12. However, with the forthcoming zero-removal optimization in
-COFFEE, we expect to outperform FEniCS in all problems. In the cases of
-Helmholtz and Diffusion, the minimum improvements are, respectively,
-1.10$\times$ and 1.18$\times$ (2D mesh, $p=1$), which tend to increase
-with polynomial order and element shape up to the values illustrated in
-the table.
+kernels. Burgers' slow downs occur in presence of a small iteration
+space (triangular mesh, $p \in [1, 2]$; tetrahedral mesh, $p \in [1,
+  2]$; prismatic mesh, $p = 1$). The result shown represents the worst
+slow down, which was obtained with a triangular mesh and $p = 1$. This
+is a result of removing zero-valued entries in FEniCS' basis function
+arrays: some operations are avoided, but indirection arrays prevent
+auto-vectorization, which significantly impacts performance as soon as
+the element matrix becomes bigger than 12$\times$12. However, with the
+forthcoming zero-removal optimization in COFFEE, we expect to
+outperform FEniCS in all problems. In the cases of Helmholtz and
+Diffusion, the minimum improvements are, respectively, 1.10$\times$
+and 1.18$\times$ (2D mesh, $p=1$), which tend to increase with
+polynomial order and element shape up to the values illustrated in the
+table.
 
@@
 \label{sec:perf-eval-blas}
-
 For the Helmholtz problem on a tetrahedral mesh, manual

### 2014-08-20 TACO-paper-v2/coffee.tex
Add TACO-paper-v2, which is the revised version

--- 2014-07-01 TACO-paper/coffee.tex
@@
 % Metadata Information
-\acmVolume{9}
-\acmNumber{4}
-\acmArticle{39}
-\acmYear{2010}
-\acmMonth{3}
+%\acmVolume{9}
+%\acmNumber{4}
+%\acmArticle{39}
+%\acmYear{2010}
+%\acmMonth{3}
 
@@
 large number of cells to obtain a satisfactory approximation
-of the PDE, possibly of the order of trillions
+of the PDE, possibly of the order of trillions, as in
 \cite{Rossinelli2013}. For example, it has been well established that mesh

### 2014-08-21 TACO-paper-v2/coffee.tex
Start addressing the points of the reviewers

@@
 loop-invariant code motion, vector-register tiling, and expression
-splitting, as well as their composition, are essential. but none are
-supported by state-of-the-art polyhedral and vendor compilers. BLAS
+splitting, as well as their composition, are essential, but very limited is
+the support of state-of-the-art polyhedral and vendor compilers, if any. BLAS
 routines could theoretically be employed, although a fairly
@@
 auto-vectorizable by a vendor compiler. Nevertheless,
-auto-vectorization is not efficient if data are not aligned and if the
+auto-vectorization is not efficient if data are not aligned to cache-line boundaries and if the
 length of the innermost loop is not a multiple of the vector length
@@
 basis function array is 3$\times$3 and $\mbox{\texttt{VL}}=4$ (e.g. AVX processor,
-with 256 bit long vector registers and 64-bit double-precision
+with 256-bit long vector registers and 64-bit double-precision
 floats). In this case, a padded version of the array will have size
-3$\times$4. The compiler is explicitly informed about data alignment
-using a suitable pragma.
+3$\times$4. The compiler is explicitly told about data alignment
+using suitable pragmas; for example, in the case of the Intel compiler, the annotation \texttt{$\#$pragma vector aligned} is added over a loop
+(as shown in later figures) to inform that all of the memory accesses in the loop body will be properly aligned.
+This allows the compiler to issue aligned load and store instructions, which are notably faster than unaligned ones.
 
@@
 
+\paragraph{Loop interchange}
 All loops are interchangeable, provided that temporaries are
@@
 % otherwise name the ones for which you know this to be true.
+\paragraph{Loop unrolling}
 Loop unroll (or unroll-and-jam of outer loops) is fundamental to the
-exposure of instruction-level parallelism, and tuning the unroll
-factor is particularly important. However, we noticed, by comparison
-with implementations having manually-unrolled loops, that recent
-versions of the Intel compiler estimate close-to-optimal unroll
-factors when the loops are affine and their bounds are small and known
-at compile-time, which is the case of our kernels. We therefore leave
+exposure of instruction-level parallelism, and tuning unroll
+factors is particularly important.
+
+We first observe that manual full (or extensive) unrolling is unlikely to be effective.
+The \texttt{ijk} loop nest would need to be small enough such that the unrolled instructions do not exceed the instruction cache, which is rarely the case: it is true that in a local assembly kernel the minimum size of the \texttt{ijk} loop nest is 3$\times$3$\times$3 (triangular mesh and polynomial order 1), but this increases rapidly with the polynomial order of the method and the discretization employed (e.g.
+tetrahedral meshes imply larger loop nests than triangular ones), so sizes greater than 10$\times$10$\times$10, for which extensive unrolling would already be harmful, are in practice very common.
+In addition, manual unrolling is dangerous because it may compromise compiler auto-vectorization by either removing loops (most compilers search for vectorizable loops) or losing spatial locality within a vector register.
+
+By comparison with implementations characterized by manually-unrolled loops, we noticed that recent
+versions of compilers like GNU's and Intel's estimate close-to-optimal unroll
+factors when the loops are affine and their bounds are relatively small and known
+at compile-time, which is the case of our kernels. Our choice, therefore, is to leave
 the backend compiler in charge of selecting unroll factors. This
-choice also simplifies COFFEE's cost model. The only situation in
+also simplifies the COFFEE's cost model, described next. The only situation in
 which we explicitly unroll-and-jam a loop is when \emph{op-vect} is
@@
 
+\paragraph{Cost model}
 The cost model is shown in Figure~\ref{algo:applyCostModel}. It takes
@@
 architectures because of the differences in the number of logical
-registers and SIMD lanes, which can impact the effectiveness of the
+registers and SIMD lanes (16 256-bit registers in the Sandy Bridge, 
+and 32 512-bit registers in the Xeon Phi), which can impact the 
 optimization strategy. The \texttt{icc 13.1} compiler was used. On the

### 2014-08-21 TACO-paper-v2/coffee.tex
Start writing the full application study

@@
 
-\subsection{The Helmholtz, Diffusion, and Burgers equations}
+We present two studies. Firstly, in Section~\ref{sec:analysis-code-transf}, we
+analyze the impact of the proposed code transformations, and their interplay,
+in three real-world equations. We quantify the speed ups achievable at
+the level of the local assembly kernel, i.e. by excluding from the measurements
+the other sections of the computation. Then, in Section~\ref{sec:full-app}, we
+provide a full-application investigation. Here, we use another equation (so as to 
+bring the total number of examined problems to four) to demonstrate that COFFEE's
+optimizations actually allow a significant reduction in the overall computation run-time 
+to be achieved, which practically motivates our research.
+
+\subsection{Impact of Code Transformations}
+\label{sec:analysis-code-transf}
+
+\subsubsection{The Helmholtz, Diffusion, and Burgers equations}
 \label{sec:real-examples}
@@
 $4$, and $6$. Moving to bigger shapes has the effect of increasing the
-number of basis function arrays, since, intuitively, the behavior of
+number of basis function arrays, since, intuitively, the behaviour of
 the equation has now to be approximated also along a third axis. On
@@
 
-\subsection{Loop permutation}
+\subsubsection{Loop interchange}
 In the following, only results for the loop order \texttt{ijk} are
@@
 
-\subsection{Impact of Generalized Loop-invariant Code Motion}
+\subsubsection{Impact of Generalized Loop-invariant Code Motion}
 \label{licm-impact}
@@
 
-\subsection{Impact of Padding and Data Alignment}
+\subsubsection{Impact of Padding and Data Alignment}
 
@@
 
-\subsection{Impact of Vector-register Tiling}
+\subsubsection{Impact of Vector-register Tiling}
 \label{sec:perf-eval-opvect}
@@
 
-\subsection{Impact of Expression Splitting}
+\subsubsection{Impact of Expression Splitting}
 \label{sec:perf-results-split}
@@
 
-\subsection{Comparison with FEniCS Form Compiler's built-in Optimizations}
+\subsubsection{Comparison with FEniCS Form Compiler's built-in Optimizations}
 \label{sec:perf-results-fenics}
@@
 
-\subsection{Comparison with hand-made BLAS-based implementations}
+\subsubsection{Comparison with hand-made BLAS-based implementations}
 \label{sec:perf-eval-blas}
@@
 work.
+
+\subsection{Full-application Study}
+\label{sec:full-app}
+In this section, we investigate the performance gain for an entire finite element computation
+developed in Firedrake, a static linear elasticity problem. The source code is available at~\cite{elasticity-code}. The application simulates ... .
+
+The execution time of a (non-explicit) finite element method is dominated by two factors: assembly and solve.
+Once computed, the local element matrices and vectors are respectively inserted in a ``global'' sparse matrix and a global vector; these form a linear system, which is usually solved with an iterative method (the ``solve'' phase).
+The amount of time spent in assembly and solve varies from problem to problem.
+As reiterated in the available literature, for example in~\cite{quadrature1}, the computational cost of local assembly grows with the complexity of the partial differential equation (e.g.
+because of larger loops, more articulated expressions), while the solve, at worst, exhibits a minimal increase (what impacts the solve are, for instance, polynomial order and mesh discretization).
+The complexity of an equation depends on several factors, including the number of derivatives and coefficient functions (i.e.
+known functions that are prescribed with the finite element specification).
+We study two versions of our static elasticity problem: 1) with one coefficient function ($f=1$) and 2) with two coefficient functions ($f=2$).
+More coefficient functions are also plausible; however, as we will see, the assembly process starts being particularly expensive already in case 1).
+
+The application runs on a tetrahedral mesh composed of 196608 cells.
+In the two versions of the elasticity problem, we examine the cases of polynomial order $p=1$ and $p=2$.
+To run our experiments, we use a single core of the Sandy Bridge architecture described in Section~\ref{sec:perf-results}.
+
+In Figure~\ref{fig:full-app-res}, we show the execution times for the four test cases, without and with optimizations enabled.
+There are four groups of two bars, each group corresponding to a specific test case; for example, the name $f_1p_1$ below the first group of bars refers to the case in which the elasticity problem has 1 coefficient function and has polynomial order 1.
+A stacked bar captures the time spent in the assembly phase (\texttt{assembly}), to solve the linear system (\texttt{solve}), and in the other various parts of the program (\texttt{other} - for example, setup of the problem and initialization of the coefficient functions).
+The cost of the insertion of the computed local element matrices (and vectors) in the global matrix (vector) is incorporated in \texttt{assembly}.
+
+We first notice that, in the scenario $f_1p_1$, the assembly is dominated by matrix insertion: despite the application of generalized loop-invariant code motion, data alignment and padding, there is only a minimal performance gain.
+This changes instantly moving to $p=2$ and $f=2$.
+In these scenarios, not only the cost of \texttt{assembly} becomes larger with respect to \texttt{solve}, but also does it make insertion cost negligible.
+The transformations automatically applied by COFFEE successfully decrease the impact of \texttt{assembly} over \texttt{solve}: this was achieved by a combined application of generalized loop-invariant code motion, alignment and padding, and expression splitting.
+Interesting are two facts.
+First, generalized loop-invariant code motion was particularly intensive: ...
+.
+Second, ...
+.
+
+%We conclude noting that in explicit finite element methods you would have more local assembly , more often, no solver
 

### 2014-08-22 TACO-paper-v2/coffee.tex
Add elasticity plots
Rewrite partially the elasticity section

@@
 As reiterated in the available literature, for example in~\cite{quadrature1}, the computational cost of local assembly grows with the complexity of the partial differential equation (e.g.
-because of larger loops, more articulated expressions), while the solve, at worst, exhibits a minimal increase (what impacts the solve are, for instance, polynomial order and mesh discretization).
+because of larger loops, more articulated expressions), while the solve, at worst, exhibits a minimal increase (what instead impact the solve are, for instance, polynomial order and mesh discretization).
 The complexity of an equation depends on several factors, including the number of derivatives and coefficient functions (i.e.
-known functions that are prescribed with the finite element specification).
+additional known functions that characterize the equation).
 We study two versions of our static elasticity problem: 1) with one coefficient function ($f=1$) and 2) with two coefficient functions ($f=2$).
@@
 
-The application runs on a tetrahedral mesh composed of 196608 cells.
+The application runs on a tetrahedral mesh composed of 196608 elements.
 In the two versions of the elasticity problem, we examine the cases of polynomial order $p=1$ and $p=2$.
@@
 
-In Figure~\ref{fig:full-app-res}, we show the execution times for the four test cases, without and with optimizations enabled.
-There are four groups of two bars, each group corresponding to a specific test case; for example, the name $f_1p_1$ below the first group of bars refers to the case in which the elasticity problem has 1 coefficient function and has polynomial order 1.
-A stacked bar captures the time spent in the assembly phase (\texttt{assembly}), to solve the linear system (\texttt{solve}), and in the other various parts of the program (\texttt{other} - for example, setup of the problem and initialization of the coefficient functions).
+\begin{figure}[t]
+	\centering
+	\subfloat[]{\includegraphics[scale=0.52]{Pictures/elasticity-p1.pdf}}
+	\subfloat[]{\includegraphics[scale=0.52]{Pictures/elasticity-p2.pdf}}
+	\caption{Performance improvement over non-optimized code for the static linear elasticity equation.
+The tetrahedral mesh is composed of 196608 elements.
+Experiments were executed on a single core of a Sandy Bridge architecture.}
+	\label{fig:full-app-res}
+\end{figure}
+
+In Figure~\ref{fig:full-app-res}, we show the execution times for the four test cases, split over two figures (one for each polynomial order), without and with optimizations enabled.
+A stacked bar captures the time spent in the assembly phase (\texttt{assembly}), in solving the linear system (\texttt{solve}), and in the other various parts of the program (\texttt{other} - for example, setup of the problem and initialization of the coefficient functions).
+The \emph{non-opt} and \emph{opt} bars correspond, respectively, to the cases in which no optimizations and a combination of optimizations have been applied.
+The optimizations applied are generalized loop-invariant code motion, alignment and padding, and expression splitting.
 The cost of the insertion of the computed local element matrices (and vectors) in the global matrix (vector) is incorporated in \texttt{assembly}.
 
-We first notice that, in the scenario $f_1p_1$, the assembly is dominated by matrix insertion: despite the application of generalized loop-invariant code motion, data alignment and padding, there is only a minimal performance gain.
+We first notice that, in the scenario $[f=1, p=1]$, the assembly is dominated by matrix insertion: despite the application of several transformations, there is only a minimal performance gain.
 This changes instantly moving to $p=2$ and $f=2$.
 In these scenarios, not only the cost of \texttt{assembly} becomes larger with respect to \texttt{solve}, but also does it make insertion cost negligible.
-The transformations automatically applied by COFFEE successfully decrease the impact of \texttt{assembly} over \texttt{solve}: this was achieved by a combined application of generalized loop-invariant code motion, alignment and padding, and expression splitting.
-Interesting are two facts.
-First, generalized loop-invariant code motion was particularly intensive: ...
-.
-Second, ...
-.
-
-%We conclude noting that in explicit finite element methods you would have more local assembly , more often, no solver
+In such cases, the transformations automatically applied by COFFEE successfully decrease the impact of \texttt{assembly} over \texttt{solve}.
+Interesting is the fact that generalized loop-invariant code motion was particularly invasive, with 23 temporaries generated and several redundancies discovered (see Section~\ref{sec:code-transf-licm}).
+
+In these experiments, we observe a maximum performance improvement of 1.47$\times$ over the non-optimized local assembly code, obtained in the case $[f=2, p=2]$.
+However, it is notably important noting that, as we anticipated, full-application speed ups rise proportionally with the time spent in assembly and, therefore, with the complexity of the equation.
+Further, by increasing polynomial order and number of coefficient functions, or by simply studying a different, more complex equation, it is our experience that performance gains become increasingly more relevant.
 

### 2014-08-22 TACO-paper-v2/coffee.tex
Change title, abstract, list of contributions, bits of intro

@@
 % Page heads
-\markboth{F. Luporini et al.}{COFFEE: an Optimizing Compiler for Finite Element Local Assembly}
+\markboth{F. Luporini et al.}{Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly}
+% TODO: "...Intensity for Finite..." OR rather "...Intensity in Finite..."
 
 % Title portion
-\title{COFFEE: an Optimizing Compiler for Finite Element Local Assembly}
+\title{Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly}
 \author{Fabio Luporini
@@
 \begin{abstract}
-The numerical solution of partial differential equations using the
-finite element method is one of the key applications of high performance
-computing. Local assembly is its characteristic operation. This entails
-the execution of a problem-specific kernel to numerically evaluate an
-integral for each element in the discretized problem domain. Since the
-domain size can be huge, executing efficient kernels is fundamental.
-Their optimization is, however, a challenging issue. Even though affine
-loop nests are generally present, the short trip counts and the
-complexity of mathematical expressions make it hard to determine a
-single or unique sequence of successful transformations. Therefore, we
-present the design and systematic evaluation of COFFEE, a
-domain-specific compiler for local assembly kernels. COFFEE manipulates
-abstract syntax trees generated from a high-level domain-specific
-language for PDEs by introducing domain-aware composable optimizations
-aimed at improving instruction-level parallelism, especially SIMD
-vectorization, and register locality. It then generates C code including
-vector intrinsics. Experiments using a range of finite-element forms of
-increasing complexity show that significant performance improvement is
-achieved.
-%
-%The code is generated from a high level language for PDEs. An
-%invocation of the kernel typically does tensor-like calculators over
-%relatively small matrices; e.g., triply nested loops operating on
-%about a dozen iterations. Because of the DSL context, COFFEE has a lot
-%of flexibility on how it organizes/orders the calculations to tailor
-%execution to a target architecture.
-%
+We study and systematically evaluate a class of composable code 
+transformations that improve arithmetic intensity in local assembly 
+operations, which represent a significant fraction of the execution
+time in finite element methods. Their performance optimization is indeed a 
+challenging issue. Even though affine loop nests are generally present, 
+the short trip counts and the complexity of mathematical expressions, which 
+vary among different problems, make it hard to determine a single or unique 
+sequence of successful transformations. Our investigation has resulted in the 
+implementation of a compiler for local assembly kernels, fully
+integrated with a real framework for developing finite element methods. 
+The compiler manipulates abstract syntax trees generated from a domain-specific
+language for partial differential equations by introducing domain-aware optimizations,
+tailored to instruction-level parallelism and register locality. Eventually, C code including 
+vector SIMD intrinsics is produced. Experiments using a range of finite-element 
+problems of increasing complexity show that significant performance improvement 
+is achieved. The generality of the approach and the applicability of the proposed 
+code transformations to other domains is also discussed.
+
 \end{abstract}
@@
   Gheorghe-Teodor Bercea, J. Ramanujam, David A. Ham, and Paul
-  H. J. Kelly, 2014. COFFEE: an Optimizing Compilerfor Finite Element
-  Local Assembly.}
+  H. J. Kelly, 2014. Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly.}
 
@@
 we propose a novel set of composable code transformations targeting, for
-the first time, instruction-level parallelism - with emphasis on SIMD
-vectorization - and register locality.
+the first time, cross-loop arithmetic intensity, with emphasis on instruction-level 
+parallelism, redundant computation, and register locality. How such transformations generalize
+to domains other than local assembly will also be discussed.
 
@@
   two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
+ \item A full-application evaluation to demonstrate that the optimization
+ strategy greatly impacts the whole finite element method.
 \end{enumerate}
-
+Our contributions also include a theoretical study involving
+\begin{enumerate}
+ \item An analysis of the generality of the proposed code transformations.
+ \item An investigation of their applicability to different computational
+ domains.  
+\end{enumerate}
 The paper is organized as follows. In Section~\ref{sec:background} we
@@
 shows performance results. Related work is described in
-Section~\ref{sec:related-work}, while Section~\ref{sec:conclusions}
-reviews our contributions in the light of our results, and identifies
-priorities for future work.
+Section~\ref{sec:related-work}. The generality of our optimization strategy
+and the applicability to other domains is discussed in Section~\ref{sec:generality}. 
+Finally, Section~\ref{sec:conclusions} reviews our contributions in the light 
+of our results, and identifies priorities for future work.
 
@@
 
+\section{Generality of the Approach and Applicability to Other Domains}
+\label{sec:generality}
+aa
+
 \section{Related Work}

### 2014-08-22 TACO-paper-v2/coffee.tex
Minor fixes

@@
 
-%Computational cost is a critical limitation in scientific computing,
-%especially for finite element simulations. To provide one particular
-%example we are particularly concerned about, it has been well
-%established that mesh resolution Consequently, our aggressive
-%optimization of local assembly, which may even take up to 60% of the
-%overall FEM's execution time, directly impacts the performance of
-%large-scale scientific simulations running on supercomputers.
-
 \section{Introduction}
@@
 
-%JUNK: , with synchronization handled at the level of the run-time
-%support Computational cost is a critical limitation in finite element
-%simulations. High resolution in time and space is critical to the
-%level of error in the simulations, but computational cost rises
-%superlinearly with resolution. The result of this is that simulations
-%are conducted at the resolution that the user can afford (wall time,
-%allocated supercomputing units or the direct financial cost of cloud
-%time)
-
-% FIXME: citation for Met Office?
-
 The time required to apply the numerical kernels is a major issue,
@@
 computational science simulations, especially for those based on
-finite elements, which are the subject of this paper. We address, in
-particular, the well-known problem of optimizing the local assembly
-phase of the finite element method
-\cite{Francis,quadrature1,petsc-integration-gpu,Kirby-FEM-opt}, which
-can be responsible for a significant fraction of the overall computation
+finite elements, which are the subject of this paper.
+
+We address, in particular, the well-known problem of optimizing the local assembly
+phase of the finite element method \cite{Francis,quadrature1, petsc-integration-gpu, Kirby-FEM-opt}, 
+which can be responsible for a significant fraction of the overall computation
 run-time, often in the range 30-60$\%$. With respect to these studies,
@@
 the first time, cross-loop arithmetic intensity, with emphasis on instruction-level 
-parallelism, redundant computation, and register locality. How such transformations generalize
-to domains other than local assembly will also be discussed.
+parallelism, redundant computation, and register locality. How such transformations 
+generalize to domains other than finite element local assembly will also be discussed.
 
@@
  strategy greatly impacts the whole finite element method.
+ \newcounter{enumTemp}
+ \setcounter{enumTemp}{\theenumi}
 \end{enumerate}
@@
 \begin{enumerate}
- \item An analysis of the generality of the proposed code transformations.
- \item An investigation of their applicability to different computational
+ \setcounter{enumi}{\theenumTemp}
+ \item An analysis of the generality of the proposed code transformations
+ and an investigation of their applicability to different computational
  domains.  
@@
 in Section~\ref{sec:pyop2-compiler}. Section~\ref{sec:perf-results}
-shows performance results. Related work is described in
-Section~\ref{sec:related-work}. The generality of our optimization strategy
+shows performance results. The generality of our optimization strategy
 and the applicability to other domains is discussed in Section~\ref{sec:generality}. 
-Finally, Section~\ref{sec:conclusions} reviews our contributions in the light 
+Related work is described in Section~\ref{sec:related-work}. Finally, 
+Section~\ref{sec:conclusions} reviews our contributions in the light 
 of our results, and identifies priorities for future work.

### 2014-08-22 TACO-paper-v2/coffee.tex
Add key sentence about choice of the three benchmarks

@@
 We address, in particular, the well-known problem of optimizing the local assembly
-phase of the finite element method \cite{Francis,quadrature1, petsc-integration-gpu, Kirby-FEM-opt}, 
+phase of the finite element method \cite{Francis,quadrature1,petsc-integration-gpu,Kirby-FEM-opt}, 
 which can be responsible for a significant fraction of the overall computation
@@
 analyze the impact of the proposed code transformations, and their interplay,
-in three real-world equations. We quantify the speed ups achievable at
+in three real-world representative equations. We quantify the speed ups achievable at
 the level of the local assembly kernel, i.e. by excluding from the measurements
@@
 
-\subsection{Impact of Code Transformations}
+\subsection{In-depth Performance Analysis of the Code Transformations and their Interplay}
 \label{sec:analysis-code-transf}
 
-\subsubsection{The Helmholtz, Diffusion, and Burgers equations}
+\subsubsection{Choice and Properties of the Benchmarks}
 \label{sec:real-examples}
@@
 the core differential operators in some of the most frequently
-encountered finite element problems in scientific computing. The
-Helmholtz and Diffusion kernels are archetypal second order elliptic
+encountered finite element problems in scientific computing. This is
+of crucial importance because distinct problems, possibly arising  in
+completely different fields, may employ (subsets of) the same differential
+operators of our benchmarks, which implies similarities
+and redundant patterns in the generated code. Consequently, the proposed
+code transformations have a domain of applicability that goes far beyond
+that of the three analyzed equations.
+
+The Helmholtz and Diffusion kernels are archetypal second order elliptic
 operators. They are complete and unsimplified examples of the

### 2014-08-22 TACO-paper-v2/coffee.tex
Draft (not finished) of the approach generality

@@
 \label{sec:generality}
-aa
+We have demonstrated that our optimizations for cross-loop arithmetic intensity are actually effective and that in a field such as automated code generation for finite element methods they can dramatically impact the full-application performance.
+The next, logical question is to ask whether applications in other computational domains would benefit from the same transformations that COFFEE is capable of applying; or, more in general, to what extent a generic vendor compiler should (could) integrate some of them.
+
+We begin observing that, strategically, our choice was to develop COFFEE as a separate, self-contained software module, with clear input/output interfaces, rather then incorporating it within PyOP2.
+This is because there are neither conceptual restrictions nor technical reasons for which the compiler could not be integrated in frameworks other than Firedrake.
+It would be quite easy, for example, to achieve that with the popular FEniCS framework, the pioneer of automated code generation for finite element methods.
+However, more challenging is to assess the generality of the approach; that is, how COFFEE and its optimizations could be transferred to other computational domains, and whether that is really significant.
+To answer this question we need to jump back to the origins of the compiler.
+
+In Firedrake, COFFEE has the fundamental role of decoupling the mathematical specification of a finite element method, captured by the Unified Form Language and the FEniCS Form Compiler, from platform-specific code optimization.
+Although at the COFFEE layer there is no notion of finite element method, it is clear that the construction of the compiler and the code transformations were dictated and inspired by our domain of interest.
+The starting point of our work was a mathematical formulation of a local assembly operation, which is expressible as follows
+
+\begin{equation*}
+\scriptsize
+\forall_{i, j} (A_{ij}^K = \sum_{q=1}^{n_1} \sum_{k=1}^{n_2} \alpha_{k, q}(a, b, c, ...) \beta_{q, i, j}(a, c, d, e, ...) \gamma_{q, K}(f, g) 
+\end{equation*}
+
+The expression represents the numerical evaluation of an integral at $n_1$ points in the mesh element $K$, computing the local element matrix $A$.
+Functions $\alpha$, $\beta$ and $\gamma$ depend on the specific problem being solved and can be intricately complex, involving for example the evaluation of derivatives.
+Translating into code such a formulation of local assembly is a challenging process for complicated equations; for this reasons, research in automated code generation has been performed over the last few years.
+Premised that, and abstracting from the inherent structure of $\alpha$, $\beta$ and $\gamma$, a number of aspects can be highlighted
+\begin{enumerate}
+\item All quantities are known values (apart obviously from the unknown $A$).
+This means that ``good'' code can be generated, for example with explicit loop bounds (i.e.
+no need to leave them as parameters), statically-initialized constant n-dimensional arrays, and no pointer aliasing.
+All of these factors contribute to the effectiveness of source-to-source code transformations: by knowing iteration spaces in advance and having a clean memory access pattern, transformations such as padding, data alignment, and vector-register tiling can be automated by a compiler like COFFEE.
+\item In the expression above, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel.
+The various sub-expressions, such as $a$, $b$, and $c$, in general only depend on (i.e.
+iterate along) a subset of the enclosing loops.
+In addition, some of these sub-expressions might reduce to the same values as iterating along certain iteration spaces.
+This code structure lies at the core of generalized loop-invariant code motion.
+Whenever sub-expressions invariant with respect to different sets of loops can be found, the question of whether, where and how to hoist them arises.
+Any codes presenting these characteristics is amenable to this kind of transformation.
+\end{enumerate}
+
+%In general, there can be a number of loops, with
 

### 2014-08-28 TACO-paper-v2/coffee.tex
Rewrite full application study

@@
 The execution time of a (non-explicit) finite element method is dominated by two factors: assembly and solve.
-Once computed, the local element matrices and vectors are respectively inserted in a ``global'' sparse matrix and a global vector; these form a linear system, which is usually solved with an iterative method (the ``solve'' phase).
-The amount of time spent in assembly and solve varies from problem to problem.
-As reiterated in the available literature, for example in~\cite{quadrature1}, the computational cost of local assembly grows with the complexity of the partial differential equation (e.g.
-because of larger loops, more articulated expressions), while the solve, at worst, exhibits a minimal increase (what instead impact the solve are, for instance, polynomial order and mesh discretization).
+The evaluation of all local element matrices and vectors, and their insertion, respectively, in a ``global'' sparse matrix and a global vector, compose the ``assembly'' phase.
+The global matrix and vector form a linear system, usually solved by an iterative method; this is the ``solve'' phase.
+The percentage of time spent in assembly and solve varies from problem to problem.
+As reiterated in the available literature, for example in~\cite{quadrature1}, the computational cost of local assembly grows with the \textit{complexity} of the partial differential equation (e.g.
+because of larger loops and more articulated expressions), while the solve time increases with polynomial order and mesh discretization.
 The complexity of an equation depends on several factors, including the number of derivatives and coefficient functions (i.e.
 additional known functions that characterize the equation).
-We study two versions of our static elasticity problem: 1) with one coefficient function ($f=1$) and 2) with two coefficient functions ($f=2$).
-More coefficient functions are also plausible; however, as we will see, the assembly process starts being particularly expensive already in case 1).
-
+
+We study two versions of the static linear elasticity problem: 1) with one coefficient function, $f=1$, and 2) with two coefficient functions, $f=2$.
+More coefficient functions are also plausible; however, as we will see, the assembly process starts being particularly expensive already with $f=1$.
+For each of these two versions, we examine the cases of polynomial order $p=1$ and $p=2$.
+To run our experiments, we use a single core of the Sandy Bridge architecture described in Section~\ref{sec:perf-results}.
 The application runs on a tetrahedral mesh composed of 196608 elements.
-In the two versions of the elasticity problem, we examine the cases of polynomial order $p=1$ and $p=2$.
-To run our experiments, we use a single core of the Sandy Bridge architecture described in Section~\ref{sec:perf-results}.
 
@@
 The optimizations applied are generalized loop-invariant code motion, alignment and padding, and expression splitting.
-The cost of the insertion of the computed local element matrices (and vectors) in the global matrix (vector) is incorporated in \texttt{assembly}.
-
-We first notice that, in the scenario $[f=1, p=1]$, the assembly is dominated by matrix insertion: despite the application of several transformations, there is only a minimal performance gain.
+We recall that the cost of the insertion of the computed local element matrices (and vectors) in the global matrix (vector) is incorporated in \texttt{assembly}.
+
+We first notice that, in the scenario $[f=1, p=1]$, the assembly is dominated by matrix insertion: despite the application of several transformations, only a minimal performance gain is achieved.
 This changes instantly moving to $p=2$ and $f=2$.
@@
 In these experiments, we observe a maximum performance improvement of 1.47$\times$ over the non-optimized local assembly code, obtained in the case $[f=2, p=2]$.
-However, it is notably important noting that, as we anticipated, full-application speed ups rise proportionally with the time spent in assembly and, therefore, with the complexity of the equation.
-Further, by increasing polynomial order and number of coefficient functions, or by simply studying a different, more complex equation, it is our experience that performance gains become increasingly more relevant.
+However, we reiterate the fact that full-application speed ups rise proportionally with the amount of time spent in assembly and, therefore, with the complexity of the equation.
+By increasing polynomial order and number of coefficient functions, or by simply studying a different, more complex equation, it is our experience that performance gains become increasingly more relevant.
+The choice of studying the static linear elasticity equation was to show that even relatively simple problems can be characterized by a large proportion of execution time spent in assembly.
 
@@
 Although at the COFFEE layer there is no notion of finite element method, it is clear that the construction of the compiler and the code transformations were dictated and inspired by our domain of interest.
-The starting point of our work was a mathematical formulation of a local assembly operation, which is expressible as follows
+The starting point of our work was the mathematical formulation of a local assembly operation, which is expressible as follows
 
@@
 \scriptsize
-\forall_{i, j} (A_{ij}^K = \sum_{q=1}^{n_1} \sum_{k=1}^{n_2} \alpha_{k, q}(a, b, c, ...) \beta_{q, i, j}(a, c, d, e, ...) \gamma_{q, K}(f, g) 
+\forall_{i, j} ~~ (~ A_{ij}^K = \sum_{q=1}^{n_1} \sum_{k=1}^{n_2} \alpha_{k, q}(a, b, c, ...) \beta_{q, i, j}(a, c, d, e, ...) \gamma_{q, K}(f, g) ~)
 \end{equation*}
@@
 \begin{enumerate}
-\item All quantities are known values (apart obviously from the unknown $A$).
+\item All quantities are known values (apart, obviously, from the unknown $A$).
 This means that ``good'' code can be generated, for example with explicit loop bounds (i.e.
-no need to leave them as parameters), statically-initialized constant n-dimensional arrays, and no pointer aliasing.
+no parameters), statically-initialized constant n-dimensional arrays, and no pointer aliasing.
 All of these factors contribute to the effectiveness of source-to-source code transformations: by knowing iteration spaces in advance and having a clean memory access pattern, transformations such as padding, data alignment, and vector-register tiling can be automated by a compiler like COFFEE.
@@
 This code structure lies at the core of generalized loop-invariant code motion.
-Whenever sub-expressions invariant with respect to different sets of loops can be found, the question of whether, where and how to hoist them arises.
+Whenever sub-expressions invariant with respect to different sets of loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises.
 Any codes presenting these characteristics is amenable to this kind of transformation.
+Note that, in general, for applying our technique the loop nest could be arbitrarily complex (e.g.
+more loops), provided that it is still affine.
+\item 
 \end{enumerate}
@@
 %In general, there can be a number of loops, with
+
+%TODO: from the expression tree to a decorated AST.
+% Diderot, tensor contraction, image processing?
 

### 2014-08-28 TACO-paper-v2/coffee.tex
Trying to rewrite the generality section...

@@
 4.4$\times$ over original kernels, and around 70$\%$ of the test cases
-obtain a speed up greater than 2$\times$. The contributions of this
-paper are
+obtain a speed up greater than 2$\times$.
+
+Summarizing, the contributions of this paper are
 \begin{enumerate}
@@
 The code transformations presented in this section are applicable to
-all finite element problems that can be formulated in Firedrake. As
-already emphasized, the structure of mathematical expressions
-evaluating the element matrix and the variation in loop trip counts,
-although typically limited to the order of tens of iterations, render
-the optimization process challenging. It is not always the same set of
-optimizations that bring performance closest to the machine peak. For
-example, the Burgers problem in Listing~\ref{code:burgers}, given the
-large number of arrays accessed, suffers from high register pressure,
+all finite element problems that can be formulated in Firedrake. 
+Their generality and the potential for applicability in other domains is also discussed in
+Section~\ref{sec:generality}.
+
+As already emphasized, the variations in the structure of 
+mathematical expressions and in loop trip counts, although typically 
+limited to the order of tens of iterations, render the optimization process 
+challenging, requiring distinct sets of transformations to bring performance 
+closest to the machine peak in different problems. For example, the Burgers problem in Listing~\ref{code:burgers}, 
+given the large number of arrays accessed, suffers from high register pressure,
 whereas the Helmholtz problem in Listing~\ref{code:helmholtz} does
@@
 is to understand the relationship between distinct code
-transformations, their impact on local assembly kernels, and to what
+transformations, their impact on cross-loop arithmetic intensity, and to what
 extent their composability is effective in a class of problems and
@@
 \begin{alltt}
-\footnotesize
-\textbf{Input:}
-ast: abstract syntax tree of a local assembly kernel produced by FFC
-wrapper: the function within which the kernel is invoked
-isa: represents the SIMD Instruction Set Architecture
+\scriptsize
+\textbf{Input:} ast: kernel abstract syntax tree
+       wrapper: function invoking the kernel
+       isa: SIMD Instruction Set Architecture
 \textbf{Output:} assembly code (C string)
@@
 \begin{figure}
-\begin{alltt}\footnotesize\internallinenumbers
+\begin{alltt}\scriptsize\internallinenumbers
 \textbf{Input:} n\_outer\_arrays, n\_inner\_arrays, n\_consts, n\_regs
@@
 
-\begin{figure}[t]
+\begin{figure}
 	\centering
@@
 \label{sec:generality}
-We have demonstrated that our optimizations for cross-loop arithmetic intensity are actually effective and that in a field such as automated code generation for finite element methods they can dramatically impact the full-application performance.
-The next, logical question is to ask whether applications in other computational domains would benefit from the same transformations that COFFEE is capable of applying; or, more in general, to what extent a generic vendor compiler should (could) integrate some of them.
+We have demonstrated that our optimizations for cross-loop arithmetic intensity are actually effective and that in a field such as automated code generation for finite element methods they can significantly impact full-application performance.
+The next, logical question is about whether applications in other computational domains would benefit from the same transformations that COFFEE is capable of applying; or, more in general, to what extent a general-purpose compiler should (could) integrate some of them.
 
@@
 It would be quite easy, for example, to achieve that with the popular FEniCS framework, the pioneer of automated code generation for finite element methods.
-However, more challenging is to assess the generality of the approach; that is, how COFFEE and its optimizations could be transferred to other computational domains, and whether that is really significant.
-To answer this question we need to jump back to the origins of the compiler.
+More challenging is to assess the generality of the approach; that is, if and how COFFEE and its optimizations could be transferred to other computational domains, and to what extent that is really useful.
+To answer this question, we need to jump back to the origins of the compiler.
 
 In Firedrake, COFFEE has the fundamental role of decoupling the mathematical specification of a finite element method, captured by the Unified Form Language and the FEniCS Form Compiler, from platform-specific code optimization.
-Although at the COFFEE layer there is no notion of finite element method, it is clear that the construction of the compiler and the code transformations were dictated and inspired by our domain of interest.
-The starting point of our work was the mathematical formulation of a local assembly operation, which is expressible as follows
+The starting point of our work was the mathematical formulation of a local assembly operation, expressible as follows
 
@@
 \scriptsize
-\forall_{i, j} ~~ (~ A_{ij}^K = \sum_{q=1}^{n_1} \sum_{k=1}^{n_2} \alpha_{k, q}(a, b, c, ...) \beta_{q, i, j}(a, c, d, e, ...) \gamma_{q, K}(f, g) ~)
+\forall_{i, j} ~~~ A_{ij}^K = \sum_{q=1}^{n_1} \sum_{k=1}^{n_2} \alpha_{k, q}(a, b, c, ...) \beta_{q, i, j}(a, c, d, e, ...) \gamma_{q}(f_K, g_K) 
 \end{equation*}
 
-The expression represents the numerical evaluation of an integral at $n_1$ points in the mesh element $K$, computing the local element matrix $A$.
-Functions $\alpha$, $\beta$ and $\gamma$ depend on the specific problem being solved and can be intricately complex, involving for example the evaluation of derivatives.
-Translating into code such a formulation of local assembly is a challenging process for complicated equations; for this reasons, research in automated code generation has been performed over the last few years.
+The expression represents the numerical evaluation of an integral at $n_1$ points in the mesh element $K$ computing the local element matrix $A$.
+Functions $\alpha$, $\beta$ and $\gamma$ are problem-specific and can be intricately complex, involving for example the evaluation of derivatives.
+Translating into efficient code such a formulation of local assembly is a challenging process, especially for complicated equations; this was actually one of the problems motivating research in DSLs and automated code generation.
 Premised that, and abstracting from the inherent structure of $\alpha$, $\beta$ and $\gamma$, a number of aspects can be highlighted
 \begin{enumerate}
-\item All quantities are known values (apart, obviously, from the unknown $A$).
-This means that ``good'' code can be generated, for example with explicit loop bounds (i.e.
-no parameters), statically-initialized constant n-dimensional arrays, and no pointer aliasing.
-All of these factors contribute to the effectiveness of source-to-source code transformations: by knowing iteration spaces in advance and having a clean memory access pattern, transformations such as padding, data alignment, and vector-register tiling can be automated by a compiler like COFFEE.
+\item Expression manipulation at the mathematical layer (e.g.
+simplification, decomposition into sub-expressions) opens multiple semantically-equivalent code generation opportunities, characterized by different trade-offs in parallelism, redundant computation, and data locality.
+The basic idea is to exploit properties of arithmetic operators, like associativity and commutativity, to re-schedule the computation suitably for the underlying architecture.
+Loop-invariant code motion and expression splitting follow this principle, so they can theoretically be re-adapted or extended to domains sharing some of the computational characteristics of local assembly, like those involving tensor contractions~\cite{modeldriven}.
 \item In the expression above, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel.
@@
 Note that, in general, for applying our technique the loop nest could be arbitrarily complex (e.g.
-more loops), provided that it is still affine.
-\item 
+more loops), provided that it is affine.
+\item TODO expression splitting -> key to loop/kernel fission
+\item The quantities in the local assembly expression are known at code generation time.
+This means that ``good'' code can be generated, for example with explicit loop bounds (i.e.
+no parameters), statically-initialized constant n-dimensional arrays, and no pointer aliasing.
+All of these factors contribute to the effectiveness of source-to-source code transformations: by knowing iteration spaces in advance and having a clean memory access pattern, transformations such as padding, data alignment, and vector-register tiling can be automated by a compiler like COFFEE.
+\item TODO lack of stencils -> simplifies most of the proposed code transformations; the interplay with a ployhedral/general-purpose compiler would need to be studied to understand how the transformations for arithmetic intensity would impact things like auto-vectorization.
+\item TODO: cross-loop vectorization ... relationship between indices/memory access patterns and vectorization techniques.
 \end{enumerate}
+
+%In this paper, we have shown how to optimize cross-loop arithmetic intensity for standard CPUs, although different optimization strategies are probably needed for other architectures (e.g.
+GPUs may benefit from distribution of fully-parallel loops among different threads to optimize shared memory usage).
 

### 2014-08-28 TACO-paper-v2/coffee.tex
Some more re-writing...now fuck off, I go out

@@
 \begin{alltt}
-\footnotesize
-\textbf{Input:} element matrix (2D array, initialized to
-0), element coordinates (array), coefficient fields (array,
-e.g. velocity)
+\scriptsize
+\textbf{Input:} element matrix (2D array, initialized to 0)
+       element coordinates (array)
+       coefficient fields (array, e.g. velocity)
 \textbf{Output:} element matrix (2D array)
@@
 \begin{algorithm}[t]
+\footnotesize
 \SetAlgorithmName{LISTING}{}
-\footnotesize
 \KwSty{void} helmholtz(double A[3][3], double **coords) $\lbrace$\\
@@
 \begin{algorithm}[t]
+\footnotesize
 \SetAlgorithmName{LISTING}{}
-\footnotesize
 \KwSty{void} burgers(double A[12][12], double **coords, double **w) $\lbrace$\\
@@
 \begin{algorithm}[t]
+\footnotesize
 \SetAlgorithmName{LISTING}{}
-\footnotesize
 \KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
@@
 \begin{algorithm}[t]
+\footnotesize
 \SetAlgorithmName{LISTING}{}
-\footnotesize
 \KwSty{void} helmholtz(double A[4][4], double **coords) $\lbrace$\\
@@
 \begin{algorithm}
+\footnotesize
 \SetAlgorithmName{LISTING}{}
-\footnotesize
 \KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
@@
 The basic idea is to exploit properties of arithmetic operators, like associativity and commutativity, to re-schedule the computation suitably for the underlying architecture.
-Loop-invariant code motion and expression splitting follow this principle, so they can theoretically be re-adapted or extended to domains sharing some of the computational characteristics of local assembly, like those involving tensor contractions~\cite{modeldriven}.
+Loop-invariant code motion and expression splitting follow this principle, so they can be re-adapted or extended to any domain computationally similar to local assembly, like those involving tensor contractions~\cite{TCE}.
 \item In the expression above, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel.
@@
 In addition, some of these sub-expressions might reduce to the same values as iterating along certain iteration spaces.
-This code structure lies at the core of generalized loop-invariant code motion.
-Whenever sub-expressions invariant with respect to different sets of loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises.
-Any codes presenting these characteristics is amenable to this kind of transformation.
-Note that, in general, for applying our technique the loop nest could be arbitrarily complex (e.g.
-more loops), provided that it is affine.
-\item TODO expression splitting -> key to loop/kernel fission
+This code structure motivated the generalized loop-invariant code motion technique.
+Whenever sub-expressions invariant with respect to different sets of affine loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises.
+\item Associative arithmetic operators are the prerequisite for expression splitting.
+In essence, this transformation is about resource-aware execution.
+It is indeed true that, in the context of COFFEE, it has been successfully applied to improve register pressure; however, the general idea is to re-schedule (re-associate) operations so as to maximize some parameters.
+Expression splitting could be used, for example, as a starting point to perform kernel fission; that is, dividing a kernel into multiple parts, each part characterized by memory requirements less stringent than the original kernel.
+An example of this, specifically about fissioning associative non-affine loops in unstructured mesh applications, is~\cite{carlo-lcpc}.
 \item The quantities in the local assembly expression are known at code generation time.
@@
 \end{enumerate}
+
+From the perspective of the domain-specific language and compiler designers, our technique represents an incentive to produce an extremely simple representation of the code (e.g.
+inlined mathematical expressions and perfect loop nests, if possible, in the form of an abstract syntax tree in the case of COFFEE) and then leave the (possibly domain-aware) code optimizer responsible for choosing and applying the best set of transformations for the underlying architecture.
+This aspect is of crucial importance since it allows achieving \textit{separation of concerns}, with domain specialists that can completely abstract from the performance optimization problem.
 

### 2014-08-30 TACO-paper-v2/coffee.tex
Add few lines, but almost time to go out

@@
 \label{sec:generality}
-We have demonstrated that our optimizations for cross-loop arithmetic intensity are actually effective and that in a field such as automated code generation for finite element methods they can significantly impact full-application performance.
-The next, logical question is about whether applications in other computational domains would benefit from the same transformations that COFFEE is capable of applying; or, more in general, to what extent a general-purpose compiler should (could) integrate some of them.
-
-We begin observing that, strategically, our choice was to develop COFFEE as a separate, self-contained software module, with clear input/output interfaces, rather then incorporating it within PyOP2.
-This is because there are neither conceptual restrictions nor technical reasons for which the compiler could not be integrated in frameworks other than Firedrake.
-It would be quite easy, for example, to achieve that with the popular FEniCS framework, the pioneer of automated code generation for finite element methods.
-More challenging is to assess the generality of the approach; that is, if and how COFFEE and its optimizations could be transferred to other computational domains, and to what extent that is really useful.
-To answer this question, we need to jump back to the origins of the compiler.
+We have demonstrated that our optimizations for cross-loop arithmetic intensity are effective in the context of automated code generation for finite element local assembly.
+In this section, we discuss about the potential for their applicability in other computational domains and, more in general, to what extent a general-purpose compiler could integrate some of them.
+
+We begin observing that our strategic choice was to develop COFFEE as a separate, self-contained software module, with clear input/output interfaces, rather then incorporating it within PyOP2.
+This is because there are neither conceptual nor technical reasons for which the compiler could not be used to optimize codes different than those produced by Firedrake.
+It would be quite easy, for example, to integrate COFFEE with the popular FEniCS framework, the pioneer of automated code generation for finite element local assembly.
+More challenging is to assess the generality of our optimization strategy; that is, if and how COFFEE and its transformations are transferable to other computational domains, perhaps other DSLs, and to what extent it would be helpful for improving full-application performance.
+To answer these questions, we need to go back to the origins of our compiler.
 
@@
 
-The expression represents the numerical evaluation of an integral at $n_1$ points in the mesh element $K$ computing the local element matrix $A$.
+The expression represents the numerical evaluation of an integral at $n_1$ points in the mesh element $K$ to compute the local element matrix $A$.
 Functions $\alpha$, $\beta$ and $\gamma$ are problem-specific and can be intricately complex, involving for example the evaluation of derivatives.
-Translating into efficient code such a formulation of local assembly is a challenging process, especially for complicated equations; this was actually one of the problems motivating research in DSLs and automated code generation.
-Premised that, and abstracting from the inherent structure of $\alpha$, $\beta$ and $\gamma$, a number of aspects can be highlighted
+Abstracting from the inherent structure of $\alpha$, $\beta$ and $\gamma$, a number of aspects can be highlighted
 \begin{enumerate}
@@
 Expression splitting could be used, for example, as a starting point to perform kernel fission; that is, dividing a kernel into multiple parts, each part characterized by memory requirements less stringent than the original kernel.
-An example of this, specifically about fissioning associative non-affine loops in unstructured mesh applications, is~\cite{carlo-lcpc}.
+An example of this, specifically about fissioning associative non-affine loops in unstructured mesh applications, is~\cite{carlolcpc}.
 \item The quantities in the local assembly expression are known at code generation time.
@@
 All of these factors contribute to the effectiveness of source-to-source code transformations: by knowing iteration spaces in advance and having a clean memory access pattern, transformations such as padding, data alignment, and vector-register tiling can be automated by a compiler like COFFEE.
-\item TODO lack of stencils -> simplifies most of the proposed code transformations; the interplay with a ployhedral/general-purpose compiler would need to be studied to understand how the transformations for arithmetic intensity would impact things like auto-vectorization.
-\item TODO: cross-loop vectorization ... relationship between indices/memory access patterns and vectorization techniques.
+\item Compiler auto-vectorization has become effective for a variety of codes.
+Nevertheless, multi-loop vectorization, which involves loading and storing data along a subset of the loops characterizing the iteration space (rather than just along the innermost loop), is not supported by available general-purpose and polyhedral compilers to the best of our knowledge.
+However, the outer-product vectorization technique presented in this paper shows that two-loop vectorization can outperform standard auto-vectorization; more importantly, we expect the performance gain to scale with the number of vectorized loops and the vector length (as proved by the Xeon Phi results).
+Despite being not straightforward to automate multi-loop vectorization in a general-purpose compiler, especially if stencils are present, we believe that in specific domains this could be more easily achieved.
+The intuition is to map the memory access pattern on vector registers, and then relying on in-register shuffling to minimize the traffic between memory and processor.
 \end{enumerate}
 
-From the perspective of the domain-specific language and compiler designers, our technique represents an incentive to produce an extremely simple representation of the code (e.g.
-inlined mathematical expressions and perfect loop nests, if possible, in the form of an abstract syntax tree in the case of COFFEE) and then leave the (possibly domain-aware) code optimizer responsible for choosing and applying the best set of transformations for the underlying architecture.
+We conclude highlighting the fact that, from the perspective of the domain-specific language and compiler designers, our optimization strategy represents an incentive to produce extremely simple representations of the code (e.g.
+fully-inlined mathematical expressions in the form of an abstract syntax tree) so as to make the architecture-aware code optimizer completely responsible for choosing and applying the most suitable set of transformations.
 This aspect is of crucial importance since it allows achieving \textit{separation of concerns}, with domain specialists that can completely abstract from the performance optimization problem.

### 2014-08-30 TACO-paper-v2/coffee.tex
Add a few other lines, bye bye

@@
 \scriptsize
-\textbf{Input:} element matrix (2D array, initialized to 0)
-       element coordinates (array)
-       coefficient fields (array, e.g. velocity)
+\textbf{Input:} element matrix (2D array, initialized to 0), coordinates (array), coefficients (array, e.g. velocity)
 \textbf{Output:} element matrix (2D array)
@@
 \scriptsize
-\textbf{Input:} ast: kernel abstract syntax tree
-       wrapper: function invoking the kernel
-       isa: SIMD Instruction Set Architecture
+\textbf{Input:} ast (kernel abstract syntax tree), wrapper (function invoking the kernel), 
+       isa (SIMD Instruction Set Architecture)
 \textbf{Output:} assembly code (C string)
@@
 
-\begin{equation*}
+\begin{equation}
+\label{eq:assembly-model}
 \scriptsize
 \forall_{i, j} ~~~ A_{ij}^K = \sum_{q=1}^{n_1} \sum_{k=1}^{n_2} \alpha_{k, q}(a, b, c, ...) \beta_{q, i, j}(a, c, d, e, ...) \gamma_{q}(f_K, g_K) 
-\end{equation*}
+\end{equation}
 
@@
 Loop-invariant code motion and expression splitting follow this principle, so they can be re-adapted or extended to any domain computationally similar to local assembly, like those involving tensor contractions~\cite{TCE}.
-\item In the expression above, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel.
-The various sub-expressions, such as $a$, $b$, and $c$, in general only depend on (i.e.
+\item In Equation~\ref{eq:assembly-model}, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel.
+The various sub-expressions in $\beta$ only depend on (i.e.
 iterate along) a subset of the enclosing loops.
@@
 This code structure motivated the generalized loop-invariant code motion technique.
-Whenever sub-expressions invariant with respect to different sets of affine loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises.
+The intuition is that whenever sub-expressions invariant with respect to different sets of affine loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises.
 \item Associative arithmetic operators are the prerequisite for expression splitting.
-In essence, this transformation is about resource-aware execution.
-It is indeed true that, in the context of COFFEE, it has been successfully applied to improve register pressure; however, the general idea is to re-schedule (re-associate) operations so as to maximize some parameters.
-Expression splitting could be used, for example, as a starting point to perform kernel fission; that is, dividing a kernel into multiple parts, each part characterized by memory requirements less stringent than the original kernel.
-An example of this, specifically about fissioning associative non-affine loops in unstructured mesh applications, is~\cite{carlolcpc}.
+In essence, this transformation concerns resource-aware execution.
+It is indeed true that, in the context of COFFEE, expression splitting has been successfully applied to improve register pressure; however, the basic idea is to re-schedule (re-associate) operations so as to maximize some generic parameters, so it is far more general.
+It could be used, for example, as a starting point to perform kernel fission; that is, splitting a kernel into multiple parts, each part characterized by memory requirements less stringent than the original code (a variant of this idea for non-affine loops in unstructured mesh applications has been adopted in~\cite{carlolcpc}).
+In Equation~\ref{eq:assembly-model}, for instance, not only any of the functions $\alpha$, $\beta$ and $\gamma$ could be split (assuming they include associative operators), but also the evaluation of $\alpha$ in the summation over \texttt{k} could be completely extracted from the expression and evaluated in a separate kernel, allowing to save space for the allocation of $b$.
 \item The quantities in the local assembly expression are known at code generation time.
-This means that ``good'' code can be generated, for example with explicit loop bounds (i.e.
-no parameters), statically-initialized constant n-dimensional arrays, and no pointer aliasing.
+This means that ``good'' code can be generated, for example with explicit loop bounds, statically-initialized constant n-dimensional arrays, and no pointer aliasing.
 All of these factors contribute to the effectiveness of source-to-source code transformations: by knowing iteration spaces in advance and having a clean memory access pattern, transformations such as padding, data alignment, and vector-register tiling can be automated by a compiler like COFFEE.

### 2014-09-01 TACO-paper-v2/coffee.tex
Add a few other bits, now really stop, see ya
Fix commas in tables
Rewrite the generality section

@@
 \subsection{Model-driven Vector-register Tiling}
+\label{sec:op-vect}
 One notable problem of assembly kernels concerns register allocation
@@
 \texttt{problem} & \texttt{shape} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} \\
-\texttt{Helmholtz} & \texttt{triangle} & 1.05 & 1.46 & 1.68 & 1.67 & 1,49 & 1,06 & 1,05 & 1,17 \\
-\texttt{Helmholtz} & \texttt{tetrahedron} & 1.36 & 2.10 & 2.64 & 2.27 & 1,28 & 1,29 & 2,05 & 1,73 \\
-\texttt{Helmholtz} & \texttt{prism} & 2.16 & 2.28 & 2.45 & 2.06 & 1,04 & 2,26 & 1,93 & 1,64 \\[0.1cm]
+\texttt{Helmholtz} & \texttt{triangle} & 1.05 & 1.46 & 1.68 & 1.67 & 1.49 & 1.06 & 1.05 & 1.17 \\
+\texttt{Helmholtz} & \texttt{tetrahedron} & 1.36 & 2.10 & 2.64 & 2.27 & 1.28 & 1.29 & 2.05 & 1.73 \\
+\texttt{Helmholtz} & \texttt{prism} & 2.16 & 2.28 & 2.45 & 2.06 & 1.04 & 2.26 & 1.93 & 1.64 \\[0.1cm]
 %\cline{1-18}
-\texttt{Diffusion} & \texttt{triangle} & 1.09 & 1.68 & 1.97 & 1.64 & 1,07 & 1,06 & 1,18 & 1,16 \\
-\texttt{Diffusion} & \texttt{tetrahedron} & 1.30 & 2.20 & 3.12 & 2.60 & 1,00 & 1,38 & 2,02 & 1,74\\
-\texttt{Diffusion} & \texttt{prism} & 2.15 & 1.82 & 2.71 & 2.32 & 1,11 & 2,16 & 1,85 & 2,83\\[0.1cm]
+\texttt{Diffusion} & \texttt{triangle} & 1.09 & 1.68 & 1.97 & 1.64 & 1.07 & 1.06 & 1.18 & 1.16 \\
+\texttt{Diffusion} & \texttt{tetrahedron} & 1.30 & 2.20 & 3.12 & 2.60 & 1.00 & 1.38 & 2.02 & 1.74\\
+\texttt{Diffusion} & \texttt{prism} & 2.15 & 1.82 & 2.71 & 2.32 & 1.11 & 2.16 & 1.85 & 2.83\\[0.1cm]
 %\cline{1-18}
-\texttt{Burgers} & \texttt{triangle} & 1.53 & 1.81 & 2.68 & 2.46 & 1,21 & 1,42 & 2,34 & 2,97  \\
-\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.24 & 1.69 & 1.59 & 1,01 & 2,55 & 0,98 & 1,21  \\
-\texttt{Burgers} & \texttt{prism} & 2.11 & 2.20 & 1.66 & 1.32 & 1,39 & 1,56 & 1,18 & 1,04 \\
+\texttt{Burgers} & \texttt{triangle} & 1.53 & 1.81 & 2.68 & 2.46 & 1.21 & 1.42 & 2.34 & 2.97  \\
+\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.24 & 1.69 & 1.59 & 1.01 & 2.55 & 0.98 & 1.21  \\
+\texttt{Burgers} & \texttt{prism} & 2.11 & 2.20 & 1.66 & 1.32 & 1.39 & 1.56 & 1.18 & 1.04 \\
 \cline{1-10}
@@
 \texttt{problem} & \texttt{shape} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} \\[0.1cm]
-\texttt{Helmholtz} & \texttt{triangle} & 1.32 & 1.88 & 2.87 & 4.13 & 1,50 & 2,41 & 1,30 & 1,96 \\
-\texttt{Helmholtz} & \texttt{tetrahedron} & 1.35 & 3.32 & 2.66 & 3.27 & 1,41 & 1,50 & 2,79 & 2,81 \\
-\texttt{Helmholtz} & \texttt{prism} & 2.63 & 2.74 & 2.43 & 2.75 & 2,38 & 2,47 & 2,15 & 1,71\\[0.1cm]
+\texttt{Helmholtz} & \texttt{triangle} & 1.32 & 1.88 & 2.87 & 4.13 & 1.50 & 2.41 & 1.30 & 1.96 \\
+\texttt{Helmholtz} & \texttt{tetrahedron} & 1.35 & 3.32 & 2.66 & 3.27 & 1.41 & 1.50 & 2.79 & 2.81 \\
+\texttt{Helmholtz} & \texttt{prism} & 2.63 & 2.74 & 2.43 & 2.75 & 2.38 & 2.47 & 2.15 & 1.71\\[0.1cm]
 %\cline{1-18}
-\texttt{Diffusion} & \texttt{triangle} & 1.38 & 1.99 & 3.07 & 4.28 & 1,08 & 1,88 & 1,20 & 1,97\\
-\texttt{Diffusion} & \texttt{tetrahedron} & 1.41 & 3.70 & 3.18 & 3.82 & 1,05 & 1,51 & 2,76 & 3,00\\
-\texttt{Diffusion} & \texttt{prism} & 2.55 & 3.13 & 2.73 & 2.69 & 2,41 & 2,52 & 2,05 & 2,48\\[0.1cm]
+\texttt{Diffusion} & \texttt{triangle} & 1.38 & 1.99 & 3.07 & 4.28 & 1.08 & 1.88 & 1.20 & 1.97\\
+\texttt{Diffusion} & \texttt{tetrahedron} & 1.41 & 3.70 & 3.18 & 3.82 & 1.05 & 1.51 & 2.76 & 3.00\\
+\texttt{Diffusion} & \texttt{prism} & 2.55 & 3.13 & 2.73 & 2.69 & 2.41 & 2.52 & 2.05 & 2.48\\[0.1cm]
 %\cline{1-18}
-\texttt{Burgers} & \texttt{triangle} & 1.56 & 2.28 & 2.61 & 2.77 & 2,84 & 2,26 & 3,96 & 4,27 \\
-\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.10 & 1.60 & 1.78 & 1,48 & 3,83 & 1,55 & 1,29 \\
-\texttt{Burgers} & \texttt{prism} & 2.19 & 2.32 & 1.64 & 1.42 & 2,18 & 2,82 & 1,24 & 1,25 \\
+\texttt{Burgers} & \texttt{triangle} & 1.56 & 2.28 & 2.61 & 2.77 & 2.84 & 2.26 & 3.96 & 4.27 \\
+\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.10 & 1.60 & 1.78 & 1.48 & 3.83 & 1.55 & 1.29 \\
+\texttt{Burgers} & \texttt{prism} & 2.19 & 2.32 & 1.64 & 1.42 & 2.18 & 2.82 & 1.24 & 1.25 \\
 \cline{1-10}
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
 	\centering
@@
 
-\begin{figure}[h]
+\begin{figure}
 	\centering
@@
 
-We begin observing that our strategic choice was to develop COFFEE as a separate, self-contained software module, with clear input/output interfaces, rather then incorporating it within PyOP2.
-This is because there are neither conceptual nor technical reasons for which the compiler could not be used to optimize codes different than those produced by Firedrake.
-It would be quite easy, for example, to integrate COFFEE with the popular FEniCS framework, the pioneer of automated code generation for finite element local assembly.
-More challenging is to assess the generality of our optimization strategy; that is, if and how COFFEE and its transformations are transferable to other computational domains, perhaps other DSLs, and to what extent it would be helpful for improving full-application performance.
-To answer these questions, we need to go back to the origins of our compiler.
-
-In Firedrake, COFFEE has the fundamental role of decoupling the mathematical specification of a finite element method, captured by the Unified Form Language and the FEniCS Form Compiler, from platform-specific code optimization.
+We observe that our choice was to develop COFFEE as a separate, self-contained software module, with clear input/output interfaces, rather then incorporating it within PyOP2.
+This was motivated by two critical aspects.
+
+\paragraph{Separation of concerns} We believe that in domain-specific frameworks there must be a clear, logical separation of roles reflecting the various levels of abstractions, with domain-specialists that can completely abstract from performance optimization.
+In Firedrake, for instance, COFFEE decouples the mathematical specification of a finite element method, captured by the Unified Form Language and the FEniCS Form Compiler, from code optimization.
+This is of fundamental importance to maximize productivity by allowing scientists to focus only on their area of expertise.
+Practically speaking, from the perspective of the domain-specific language and compiler designers, our optimization strategy represents an incentive to produce extremely simple representations of the code (e.g.
+fully-inlined mathematical expressions in the form of an abstract syntax tree, in the case of Firedrake) so as to make the architecture-aware code optimizer completely responsible for choosing and applying the most suitable set of transformations.
+
+\paragraph{Generalization to other domains} There are neither conceptual nor technical reasons for which our compiler could not be used outside of the Firedrake system.
+For example, integration with the popular FEniCS framework, the pioneer of automated code generation for finite element local assembly, would be relatively easy to achieve.
+More challenging is to assess the generality of our optimization strategy; that is, if and how COFFEE and its transformations are transferable to other computational domains, perhaps other DSLs, and to what extent this would be helpful for improving full-application performance.
+To answer these questions, we first need to go back to the origins of our compiler.
 The starting point of our work was the mathematical formulation of a local assembly operation, expressible as follows
@@
 \scriptsize
-\forall_{i, j} ~~~ A_{ij}^K = \sum_{q=1}^{n_1} \sum_{k=1}^{n_2} \alpha_{k, q}(a, b, c, ...) \beta_{q, i, j}(a, c, d, e, ...) \gamma_{q}(f_K, g_K) 
+\forall_{i, j} ~~~ A_{ij}^K = \sum_{q=1}^{n_1} \sum_{k=1}^{n_2} \alpha_{k, q}(a', b', c', ...) \beta_{q, i, j}(a, b, c, d, ...) \gamma_{q}(w_K, z_K) 
 \end{equation}
@@
 Abstracting from the inherent structure of $\alpha$, $\beta$ and $\gamma$, a number of aspects can be highlighted
-\begin{enumerate}
-\item Expression manipulation at the mathematical layer (e.g.
+
+\paragraph{Optimizing mathematical expressions} 
+Expression manipulation at the mathematical layer (e.g.
 simplification, decomposition into sub-expressions) opens multiple semantically-equivalent code generation opportunities, characterized by different trade-offs in parallelism, redundant computation, and data locality.
@@
 Loop-invariant code motion and expression splitting follow this principle, so they can be re-adapted or extended to any domain computationally similar to local assembly, like those involving tensor contractions~\cite{TCE}.
+We highlight three notable points.
+\begin{enumerate}
 \item In Equation~\ref{eq:assembly-model}, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel.
-The various sub-expressions in $\beta$ only depend on (i.e.
+Throughout the paper we assumed to have a single thread per kernel, although it is worth noting that for certain architectures (e.g.
+GPUs) this could be prohibitive due to memory requirements.
+Intra-kernel parallelization is one possible solution: a domain-specific compiler like COFFEE could map mathematical quantifiers and operators to different parallelization schemes, and transform both the kernel and the function calling it for multi-threaded execution.
+This approach is fundamental if performance portability needs to be achieved.
+We believe that analogous considerations apply to different computational domains.
+\item The various sub-expressions in $\beta$ only depend on (i.e.
 iterate along) a subset of the enclosing loops.
@@
 In essence, this transformation concerns resource-aware execution.
-It is indeed true that, in the context of COFFEE, expression splitting has been successfully applied to improve register pressure; however, the basic idea is to re-schedule (re-associate) operations so as to maximize some generic parameters, so it is far more general.
-It could be used, for example, as a starting point to perform kernel fission; that is, splitting a kernel into multiple parts, each part characterized by memory requirements less stringent than the original code (a variant of this idea for non-affine loops in unstructured mesh applications has been adopted in~\cite{carlolcpc}).
-In Equation~\ref{eq:assembly-model}, for instance, not only any of the functions $\alpha$, $\beta$ and $\gamma$ could be split (assuming they include associative operators), but also the evaluation of $\alpha$ in the summation over \texttt{k} could be completely extracted from the expression and evaluated in a separate kernel, allowing to save space for the allocation of $b$.
-\item The quantities in the local assembly expression are known at code generation time.
-This means that ``good'' code can be generated, for example with explicit loop bounds, statically-initialized constant n-dimensional arrays, and no pointer aliasing.
-All of these factors contribute to the effectiveness of source-to-source code transformations: by knowing iteration spaces in advance and having a clean memory access pattern, transformations such as padding, data alignment, and vector-register tiling can be automated by a compiler like COFFEE.
-\item Compiler auto-vectorization has become effective for a variety of codes.
+It is indeed true that, in the context of COFFEE, expression splitting has been successfully applied to improve register pressure; nevertheless, the basic idea is far more general, which is re-scheduling (re-associating) operations to optimize for some generic parameters.
+It could be used, for example, as a starting point to perform kernel fission; that is, splitting a kernel into multiple parts, each part characterized by less stringent memory requirements (a variant of this idea for non-affine loops in unstructured mesh applications has been adopted in~\cite{carlolcpc}).
+In Equation~\ref{eq:assembly-model}, for instance, not only any of the functions $\alpha$, $\beta$ and $\gamma$ could be split (assuming they include associative operators), but also the evaluation of $\alpha$ could be completely extracted and evaluated in a separate kernel: this would avoid allocating space for $[a', b', c', ...]$ in a second kernel computing the rest of the expression.
+This may be particularly indicated for many-core architectures in which the available per-core memory is much smaller than that in traditional CPUs.
+\end{enumerate}
+\paragraph{Code generation and applicability of the transformations} 
+All array sizes and loop bounds, like $n1$ and $n2$ in Equation~\ref{eq:assembly-model}, are known at code generation time.
+This means that ``good'' code can be generated.
+For example, loop bounds can be made explicit, arrays can be statically initialized, and pointer aliasing is easily avoidable.
+Further, all of these factors contribute to the applicability and the effectiveness of all of our code transformations.
+For instance, knowing loop bounds allows both generating correct code when applying vector-register tiling (see Section~\ref{sec:op-vect}) and discovering redundant computation opportunities (see Section~\ref{sec:code-transf-licm}).
+On the other hand, the padding and data alignment transformation could be performed directly at run-time; note that, theoretically, it could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.
+\paragraph{Multi-loop vectorization} 
+Compiler auto-vectorization has become effective for a variety of codes.
 Nevertheless, multi-loop vectorization, which involves loading and storing data along a subset of the loops characterizing the iteration space (rather than just along the innermost loop), is not supported by available general-purpose and polyhedral compilers to the best of our knowledge.
@@
 The intuition is to map the memory access pattern on vector registers, and then relying on in-register shuffling to minimize the traffic between memory and processor.
-\end{enumerate}
-
-We conclude highlighting the fact that, from the perspective of the domain-specific language and compiler designers, our optimization strategy represents an incentive to produce extremely simple representations of the code (e.g.
-fully-inlined mathematical expressions in the form of an abstract syntax tree) so as to make the architecture-aware code optimizer completely responsible for choosing and applying the most suitable set of transformations.
-This aspect is of crucial importance since it allows achieving \textit{separation of concerns}, with domain specialists that can completely abstract from the performance optimization problem.
 

### 2014-09-02 TACO-paper-v2/coffee.tex
Add a bit to the conclusions
Change compiler picture

@@
 \caption{High-level view of Firedrake. COFFEE is at the core,
-  receiving ASTs from a modified version of the FEniCS Form compiler
+  receiving ASTs from a modified version of the FEniCS Form Compiler
   and producing optimized C code kernels.}
@@
 just-in-time compilation and execution over the discretized domain take
-place. The flow is summarized in Figure~\ref{fig:coffee-pipeline}.
+place. The flow and the compiler structure are outlined in Figure~\ref{fig:coffee-pipeline}.
 
@@
 ASTs occurs at run-time as soon as problem-specific data are
-available. Our optimization strategy, based on heuristics and a simple
-cost model, is described in the following, along with an overview of
-the compiler.
-
-The compiler structure is outlined in
-Figure~\ref{algo:PyOP2Compiler}. Initially the AST is inspected,
-looking for the presence of iteration spaces and domain-specific
-information provided by the higher layer. If the kernel lacks an
-iteration space, then so-called inter-kernel vectorization, in which
-the outer, non-affine loop over mesh elements is vectorized, can be
-applied. This feature, currently under development, has been shown to
-be useful in several finite volume
-applications~\cite{inter-kernel-vect}. Subsequently, an ordered
-sequence of optimization steps are executed. Application of
+available. Our optimization strategy is rather based on heuristics and a simple
+cost model, as described later.
+
+Once received an AST from FFC, an ordered sequence of optimization steps are executed. Application of
 \emph{licm} must precede padding and data alignment, due to the
@@
 remainder loop, amenable to auto-vectorization, is inserted.
-
-\begin{figure}[t]
-\begin{alltt}
-\scriptsize
-\textbf{Input:} ast (kernel abstract syntax tree), wrapper (function invoking the kernel), 
-       isa (SIMD Instruction Set Architecture)
-\textbf{Output:} assembly code (C string)
-\textbf{// Analyze ast and build optimization plan }
-it\_space = analyze(ast)
-\textbf{if not} it\_space
-  ast.apply\_inter\_kernel\_vectorization(wrapper, isa)
-  \textbf{return} wrapper + ast.from\_ast\_to\_c()
-plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs)
-\textbf{// Optimize ast based on plan}
-ast.licm()
-ast.padding()
-ast.align\_data()
-\textbf{if} plan.permute
-  ast.permute\_assembly\_loops()
-\textbf{if} plan.split\_factor
-  ast.split(plan.split\_factor)
-\textbf{if} plan.uaj\_factor
-  uaj = MIN(plan.uaj\_factor, it\_space.j.size/isa.vf)
-  ast.op\_vect(uaj)
-\textbf{return} wrapper + ast.from\_ast\_to\_c()
-\end{alltt}
-\caption{Outline of the COFFEE architecture.}
-\label{algo:PyOP2Compiler}
-\end{figure}
 
@@
 problems expressible in Firedrake, and is already integrated with
-this framework.
+this framework. The generality and the applicability of the proposed 
+code transformations to other domains has also been discussed.
 

### 2014-09-02 TACO-paper-v2/coffee.tex
Rewrite bits of the compiler section

@@
 
-Splitting an expression has, however, several drawbacks. Firstly, it
+Splitting an expression (henceforth \emph{split}) has, however, several drawbacks. Firstly, it
 increases the number of accesses to $A$ proportionally to the ``split
@@
 
-Because of the large number of parametric transformations, COFFEE
-needs a mechanism to select the most suitable optimization strategy
-for a given problem. Auto-tuning might be used, although it would
-significantly increase the run-time overhead, since the generation of
-ASTs occurs at run-time as soon as problem-specific data are
-available. Our optimization strategy is rather based on heuristics and a simple
-cost model, as described later.
-
-Once received an AST from FFC, an ordered sequence of optimization steps are executed. Application of
-\emph{licm} must precede padding and data alignment, due to the
-introduction of temporary arrays. Based on a cost model, loop
-interchange, \emph{split}, and \emph{op-vect} may be introduced. Their
-implementation is based on analysis and transformation of the kernel's
-AST. When \emph{op-vect} is selected, the compiler outputs AVX or
-AVX-512 intrinsics code. Any possible corner cases are handled: for
-example, if \emph{op-vect} is to be applied, but the size of the
-iteration space is not a multiple of the vector length, then a
-remainder loop, amenable to auto-vectorization, is inserted.
+COFFEE uses heuristics and a simple cost model to select the most suitable 
+optimization strategy for a given problem. Auto-tuning might be used, although
+it would significantly increase the  run-time overhead, since the generation of 
+ASTs occurs at run-time as soon as problem-specific data are available.
+
+The compiler executes an ordered sequence of optimization steps to the 
+ASTs received from FFC. Application of \emph{licm} must precede padding 
+and data alignment, due to the introduction of temporary arrays. These
+transformations are always performed because they are likely to improve
+the run-time performance, as demonstrated by our results in Section~\ref{sec:analysis-code-transf}.
+Based on a cost model (described later), loop interchange, \emph{split}, 
+and \emph{op-vect} may be introduced. Their implementation is centered 
+on analysis and manipulation of the kernel's AST. When \emph{op-vect}
+is selected, the compiler outputs AVX or AVX-512 intrinsics code. Any 
+possible corner cases are handled: for example, if \emph{op-vect} is to 
+be applied, but the size of the iteration space is not a multiple of the vector 
+length, then a remainder loop, amenable to auto-vectorization, is inserted.
 

### 2014-09-02 TACO-paper-v2/coffee.tex
minor fix
small rewriting of generality section

@@
 We first notice that, in the scenario $[f=1, p=1]$, the assembly is dominated by matrix insertion: despite the application of several transformations, only a minimal performance gain is achieved.
-This changes instantly moving to $p=2$ and $f=2$.
+This changes instantly increasing $p$ or $f$.
 In these scenarios, not only the cost of \texttt{assembly} becomes larger with respect to \texttt{solve}, but also does it make insertion cost negligible.
@@
 \label{sec:generality}
-We have demonstrated that our optimizations for cross-loop arithmetic intensity are effective in the context of automated code generation for finite element local assembly.
-In this section, we discuss about the potential for their applicability in other computational domains and, more in general, to what extent a general-purpose compiler could integrate some of them.
+We have demonstrated that our cross-loop optimizations for arithmetic intensity are effective in the context of automated code generation for finite element local assembly.
+In this section, we discuss about their applicability in other computational domains and, more in general, their integrability within a general-purpose compiler.
 
 We observe that our choice was to develop COFFEE as a separate, self-contained software module, with clear input/output interfaces, rather then incorporating it within PyOP2.
-This was motivated by two critical aspects.
+This was motivated by two critical aspects that characterize the generality of our research.
 
@@
 
-\paragraph{Generalization to other domains} There are neither conceptual nor technical reasons for which our compiler could not be used outside of the Firedrake system.
+\paragraph{Generalizability to other domains} There are neither conceptual nor technical reasons for which our compiler could not be used outside of Firedrake.
 For example, integration with the popular FEniCS framework, the pioneer of automated code generation for finite element local assembly, would be relatively easy to achieve.
-More challenging is to assess the generality of our optimization strategy; that is, if and how COFFEE and its transformations are transferable to other computational domains, perhaps other DSLs, and to what extent this would be helpful for improving full-application performance.
+More challenging is to assess the generality of the optimization strategy; that is, if and how COFFEE and its transformations are transferable to other computational domains, perhaps other DSLs, and to what extent this would be helpful for improving full-application performance.
 To answer these questions, we first need to go back to the origins of our compiler.
@@
 
-The expression represents the numerical evaluation of an integral at $n_1$ points in the mesh element $K$ to compute the local element matrix $A$.
+The expression represents the numerical evaluation of an integral at $n_1$ points in the mesh element $K$ computing the local element matrix $A$.
 Functions $\alpha$, $\beta$ and $\gamma$ are problem-specific and can be intricately complex, involving for example the evaluation of derivatives.
-Abstracting from the inherent structure of $\alpha$, $\beta$ and $\gamma$, a number of aspects can be highlighted
-
-\paragraph{Optimizing mathematical expressions} 
-Expression manipulation at the mathematical layer (e.g.
+We can however abstract from the inherent structure of $\alpha$, $\beta$ and $\gamma$ to highlight a number of aspects
+
+\begin{itemize}
+\item \textbf{Optimizing mathematical expressions.} Expression manipulation (e.g.
 simplification, decomposition into sub-expressions) opens multiple semantically-equivalent code generation opportunities, characterized by different trade-offs in parallelism, redundant computation, and data locality.
 The basic idea is to exploit properties of arithmetic operators, like associativity and commutativity, to re-schedule the computation suitably for the underlying architecture.
-Loop-invariant code motion and expression splitting follow this principle, so they can be re-adapted or extended to any domain computationally similar to local assembly, like those involving tensor contractions~\cite{TCE}.
-We highlight three notable points.
+Loop-invariant code motion and expression splitting follow this principle, so they can be re-adapted or extended to any domains involving numerical evaluation of complex mathematical expressions (e.g.
+electronic structure calculations in physics and quantum chemistry relying on tensor contractions~\cite{TCE}).
+In this context, we highlight three notable points.
 \begin{enumerate}
 \item In Equation~\ref{eq:assembly-model}, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel.
-Throughout the paper we assumed to have a single thread per kernel, although it is worth noting that for certain architectures (e.g.
+Throughout the paper we assumed to assign the execution of a kernel to a single thread, which is likely to be the best strategy for a standard multi-core CPU.
+On the other hand, we note that for certain architectures (e.g.
 GPUs) this could be prohibitive due to memory requirements.
-Intra-kernel parallelization is one possible solution: a domain-specific compiler like COFFEE could map mathematical quantifiers and operators to different parallelization schemes, and transform both the kernel and the function calling it for multi-threaded execution.
-This approach is fundamental if performance portability needs to be achieved.
-We believe that analogous considerations apply to different computational domains.
+Intra-kernel parallelization is one possible solution: a domain-specific compiler like COFFEE could map mathematical quantifiers and operators to different parallelization schemes and generate distinct variants of multi-threaded kernel code.
+Based on our experience, we believe this is the right approach to achieve performance portability.
 \item The various sub-expressions in $\beta$ only depend on (i.e.
@@
 \end{enumerate}
-\paragraph{Code generation and applicability of the transformations} 
-All array sizes and loop bounds, like $n1$ and $n2$ in Equation~\ref{eq:assembly-model}, are known at code generation time.
+\item \textbf{Code generation and applicability of the transformations.} All array sizes and loop bounds, like $n1$ and $n2$ in Equation~\ref{eq:assembly-model}, are known at code generation time.
 This means that ``good'' code can be generated.
@@
 On the other hand, the padding and data alignment transformation could be performed directly at run-time; note that, theoretically, it could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.
-\paragraph{Multi-loop vectorization} 
-Compiler auto-vectorization has become effective for a variety of codes.
+\item \textbf{Multi-loop vectorization.} Compiler auto-vectorization has become effective for a variety of codes.
 Nevertheless, multi-loop vectorization, which involves loading and storing data along a subset of the loops characterizing the iteration space (rather than just along the innermost loop), is not supported by available general-purpose and polyhedral compilers to the best of our knowledge.
@@
 The intuition is to map the memory access pattern on vector registers, and then relying on in-register shuffling to minimize the traffic between memory and processor.
+\end{itemize}
 

### 2014-09-02 TACO-paper-v2/coffee.tex
minor fixes, add TODOs

@@
 The intuition is that whenever sub-expressions invariant with respect to different sets of affine loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises.
+TODO: hoisting requires extra memory, so what should one do?
 \item Associative arithmetic operators are the prerequisite for expression splitting.
@@
 Nevertheless, multi-loop vectorization, which involves loading and storing data along a subset of the loops characterizing the iteration space (rather than just along the innermost loop), is not supported by available general-purpose and polyhedral compilers to the best of our knowledge.
-However, the outer-product vectorization technique presented in this paper shows that two-loop vectorization can outperform standard auto-vectorization; more importantly, we expect the performance gain to scale with the number of vectorized loops and the vector length (as proved by the Xeon Phi results).
+The outer-product vectorization technique presented in this paper shows that two-loop vectorization can outperform standard auto-vectorization; more importantly, we expect the performance gain to scale with the number of vectorized loops and the vector length (as proved by the Xeon Phi results).
 Despite being not straightforward to automate multi-loop vectorization in a general-purpose compiler, especially if stencils are present, we believe that in specific domains this could be more easily achieved.

### 2014-09-03 TACO-paper-v2/coffee.tex
Add explanation of linear elasticity
Some other minor adjustments

--- 2014-09-02 TACO-paper-v2/coffee.tex
@@
 implementation of a compiler for local assembly kernels, fully
-integrated with a real framework for developing finite element methods. 
+integrated with a framework for developing finite element methods. 
 The compiler manipulates abstract syntax trees generated from a domain-specific
-language for partial differential equations by introducing domain-aware optimizations,
-tailored to instruction-level parallelism and register locality. Eventually, C code including 
-vector SIMD intrinsics is produced. Experiments using a range of finite-element 
+language by introducing domain-aware optimizations and, eventually,
+produces C code including vector SIMD intrinsics. Experiments using a range of finite-element 
 problems of increasing complexity show that significant performance improvement 
@@
 transformations described so far. Given that many of these
-transformations depend on some parameters (e.g. tile size), we need a
-mechanism to prune such a large optimization space. This aspect is
-treated in Section~\ref{sec:pyop2-compiler}.
+transformations depend on some parameters, we need a
+mechanism to prune such a large optimization space, which can be as large as order of thousands variants for medium-size loops and moderately complex mathematical expressions (all parametric transformations can indeed be combined).
+This aspect is treated in Section~\ref{sec:pyop2-compiler}.
 
@@
 
-The compiler executes an ordered sequence of optimization steps to the 
+The compiler applies an ordered sequence of optimization steps to the 
 ASTs received from FFC. Application of \emph{licm} must precede padding 
@@
 
-We first observe that manual full (or extensive) unrolling is unlikely to be effective.
-The \texttt{ijk} loop nest would need to be small enough such that the unrolled instructions do not exceed the instruction cache, which is rarely the case: it is true that in a local assembly kernel the minimum size of the \texttt{ijk} loop nest is 3$\times$3$\times$3 (triangular mesh and polynomial order 1), but this increases rapidly with the polynomial order of the method and the discretization employed (e.g.
+We first observe that manual full (or extensive) unrolling is unlikely to be effective for two reasons.
+Firstly, the \texttt{ijk} loop nest would need to be small enough such that the unrolled instructions do not exceed the instruction cache, which is rarely the case: it is true that in a local assembly kernel the minimum size of the \texttt{ijk} loop nest is 3$\times$3$\times$3 (triangular mesh and polynomial order 1), but this increases rapidly with the polynomial order of the method and the discretization employed (e.g.
 tetrahedral meshes imply larger loop nests than triangular ones), so sizes greater than 10$\times$10$\times$10, for which extensive unrolling would already be harmful, are in practice very common.
-In addition, manual unrolling is dangerous because it may compromise compiler auto-vectorization by either removing loops (most compilers search for vectorizable loops) or losing spatial locality within a vector register.
+Secondly, manual unrolling is dangerous because it may compromise compiler auto-vectorization by either removing loops (most compilers search for vectorizable loops) or losing spatial locality within a vector register.
 
@@
 In this section, we investigate the performance gain for an entire finite element computation
-developed in Firedrake, a static linear elasticity problem. The source code is available at~\cite{elasticity-code}. The application simulates ... .
+developed in Firedrake, a static linear elasticity problem.
+The source code is available at~\cite{elasticity-code}.
+The equation is used to simulate deformation of an object caused by pre-established loading conditions.
 
@@
 \item In Equation~\ref{eq:assembly-model}, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel.
-Throughout the paper we assumed to assign the execution of a kernel to a single thread, which is likely to be the best strategy for a standard multi-core CPU.
+Throughout the paper we assumed a kernel be executed by a single thread, which is likely to be the best strategy for standard multi-core CPUs.
 On the other hand, we note that for certain architectures (e.g.
 GPUs) this could be prohibitive due to memory requirements.
-Intra-kernel parallelization is one possible solution: a domain-specific compiler like COFFEE could map mathematical quantifiers and operators to different parallelization schemes and generate distinct variants of multi-threaded kernel code.
+Intra-kernel parallelization is one possible solution: a domain-specific compiler such as COFFEE could map mathematical quantifiers and operators to different parallelization schemes and generate distinct variants of multi-threaded kernel code.
 Based on our experience, we believe this is the right approach to achieve performance portability.
@@
 The intuition is that whenever sub-expressions invariant with respect to different sets of affine loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises.
+Pre-computation of invariant terms also increases memory requirements due to the need for temporary arrays, so it is possible that for certain architectures the transformation could actually cause slow downs (e.g.
+whenever the available per-core memory is small).
 \item Associative arithmetic operators are the prerequisite for expression splitting.
@@
 It could be used, for example, as a starting point to perform kernel fission; that is, splitting a kernel into multiple parts, each part characterized by less stringent memory requirements (a variant of this idea for non-affine loops in unstructured mesh applications has been adopted in~\cite{carlolcpc}).
-In Equation~\ref{eq:assembly-model}, for instance, not only any of the functions $\alpha$, $\beta$ and $\gamma$ could be split (assuming they include associative operators), but also the evaluation of $\alpha$ could be completely extracted and evaluated in a separate kernel: this would avoid allocating space for $[a', b', c', ...]$ in a second kernel computing the rest of the expression.
+In Equation~\ref{eq:assembly-model}, for instance, not only any of the functions $\alpha$, $\beta$ and $\gamma$ could be split (assuming they include associative operators), but also $\alpha$ could be completely extracted and evaluated in a separate kernel: this would avoid allocating space for $[a', b', c', ...]$ in a second kernel computing the rest of the expression.
 This may be particularly indicated for many-core architectures in which the available per-core memory is much smaller than that in traditional CPUs.
@@
 For example, loop bounds can be made explicit, arrays can be statically initialized, and pointer aliasing is easily avoidable.
-Further, all of these factors contribute to the applicability and the effectiveness of all of our code transformations.
+Further, all of these factors contribute to the applicability and the effectiveness of some of our code transformations.
 For instance, knowing loop bounds allows both generating correct code when applying vector-register tiling (see Section~\ref{sec:op-vect}) and discovering redundant computation opportunities (see Section~\ref{sec:code-transf-licm}).
-On the other hand, the padding and data alignment transformation could be performed directly at run-time; note that, theoretically, it could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.
-\item \textbf{Multi-loop vectorization.} Compiler auto-vectorization has become effective for a variety of codes.
-Nevertheless, multi-loop vectorization, which involves loading and storing data along a subset of the loops characterizing the iteration space (rather than just along the innermost loop), is not supported by available general-purpose and polyhedral compilers to the best of our knowledge.
-However, the outer-product vectorization technique presented in this paper shows that two-loop vectorization can outperform standard auto-vectorization; more importantly, we expect the performance gain to scale with the number of vectorized loops and the vector length (as proved by the Xeon Phi results).
-Despite being not straightforward to automate multi-loop vectorization in a general-purpose compiler, especially if stencils are present, we believe that in specific domains this could be more easily achieved.
-The intuition is to map the memory access pattern on vector registers, and then relying on in-register shuffling to minimize the traffic between memory and processor.
+A special case are padding and data alignment since they could be performed even at run-time if some values were not known at code generation time; theoretically, they could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.
+\item \textbf{Multi-loop vectorization.} Compiler auto-vectorization has become increasingly effective in a variety of codes.
+Nevertheless, multi-loop vectorization, which involves loading and storing data along a subset of the loops characterizing the iteration space (rather than just along the innermost loop), is not supported by available general-purpose and polyhedral compilers, to the best of our knowledge.
+The outer-product vectorization technique presented in this paper shows that two-loop vectorization can outperform standard auto-vectorization.
+In addition, we expect the performance gain to scale with the number of vectorized loops and the vector length (as proved by the Xeon Phi experiments).
+Despite being not straightforward to automate multi-loop vectorization in a general-purpose compiler, especially if stencils are present, we believe that this could be more easily achieved in specific domains.
+The intuition is to map the memory access pattern on vector registers, and then exploiting in-register shuffling to minimize the traffic between memory and processor.
+By demonstrating the effectiveness of multi-loop vectorization in a real scenario, our research represents an incentive for studying this technique in a broader and systematic way.
 \end{itemize}
+%We conclude noting that despite experiments were performed on Intel platforms only, performance gains are expected even on completely different CPU architectures, since the techniques proposed in this paper involves re-scheduling
 

### 2014-09-03 TACO-paper-v2/coffee.tex
Adjust conclusions

@@
 
-In this paper, we have presented the design, optimizations and systematic
-performance evaluation of COFFEE, a compiler for finite element local
-assembly. In this context, to the best of our knowledge, COFFEE is the
-first compiler capable of introducing low-level optimizations to
-maximize instruction-level parallelism, register locality and SIMD
-vectorization. Assembly kernels have particular characteristics. Their
-iteration space is usually very small, with the size depending on
-aspects like the degree of accuracy one wants to reach (polynomial
-order of the method) and the mesh discretization employed. The data
-space, in terms of number of arrays and scalars required to evaluate
+In this paper, we have presented the study and systematic performance 
+evaluation of a class of composable cross-loop optimizations for improving 
+arithmetic intensity in finite element local assembly kernels, and their 
+integration in a novel compiler, COFFEE. In the context of automated code 
+generation for finite element local assembly, COFFEE is the first compiler 
+capable of introducing low-level optimizations to maximize instruction-level 
+parallelism, register locality and SIMD vectorization. Assembly kernels have 
+particular characteristics. Their iteration space is usually very small, with 
+the size depending on aspects like the degree of accuracy one wants to 
+reach (polynomial order of the method) and the mesh discretization employed. 
+The data space, in terms of number of arrays and scalars required to evaluate
 the element matrix, grows proportionally with the complexity of the

### 2014-09-03 TACO-paper-v2/TACO-Rebuttal.txt
Add rebuttal file

We thank our reviewers (R1-R4). We appreciate the positive feedback and the
critical comments on our paper. Below we answers the various points arose by the
reviewers.

### 2014-09-03 TACO-paper-v2/coffee.tex

@@
 executing a problem-specific kernel over all cells, or elements, in the
-discretized domain. In this work, we focus on relatively low order
-finite element methods, in which an assembly kernel's working set is
-usually small enough to fit the L1 cache. Low order methods are by no
-means exotic: they are employed in a wide variety of fields, including
-climate and ocean modeling, computational fluid dynamics, and structural
-mechanics. High order methods such as the spectral element method
-\cite{SPENCER} commonly require different algorithms to solve PDEs and
-therefore are excluded from our study.
+discretized domain. In this work, we focus on relatively low order finite
+element methods, in which an assembly kernel's working set is usually small
+enough to fit the L1 cache. Low order methods are by no means exotic: they
+are employed in a wide variety of fields, including climate and ocean
+modeling, computational fluid dynamics, and structural mechanics. The
+efficient assembly of high order methods such as the spectral element
+method \cite{SPENCER} requires a significantly different loop nest structure.
+High order methods  therefore are excluded from our study.
 
@@
   % FIXME: this rather hand-wavy and not technically correct
-  \item Definition of basis functions used, intuitively, to interpolate
-    the contribution to the PDE solution over the element. The choice of
-    basis functions is expressed in UFL directly by users. In the
-    generated code, they are represented as global read-only two
-    dimensional arrays (i.e., using \texttt{static const} in C) of
-    double precision floats.
+  \item Definition of basis functions used to interpolate fields at the
+    quadrature points in the element. The choice of basis functions is
+    expressed in UFL directly by users. In the generated code, they are
+    represented as global read-only two dimensional arrays (i.e., using
+    \texttt{static const} in C) of double precision floats.
   \item Evaluation of the element matrix in an affine loop nest, in which
@@
 
-We first notice that, in the scenario $[f=1, p=1]$, the assembly is dominated by matrix insertion: despite the application of several transformations, only a minimal performance gain is achieved.
-This changes instantly increasing $p$ or $f$.
-In these scenarios, not only the cost of \texttt{assembly} becomes larger with respect to \texttt{solve}, but also does it make insertion cost negligible.
-In such cases, the transformations automatically applied by COFFEE successfully decrease the impact of \texttt{assembly} over \texttt{solve}.
-Interesting is the fact that generalized loop-invariant code motion was particularly invasive, with 23 temporaries generated and several redundancies discovered (see Section~\ref{sec:code-transf-licm}).
-
-In these experiments, we observe a maximum performance improvement of 1.47$\times$ over the non-optimized local assembly code, obtained in the case $[f=2, p=2]$.
+We first notice that, in the scenario ($f=1, p=1$), the assembly is
+dominated by matrix insertion: despite the application of several
+transformations, only a minimal performance gain is achieved. However if
+either $p$ or $f$ is increased then the cost of \texttt{assembly} becomes
+larger with respect to \texttt{solve}, and the matrix insertion cost becomes
+negligible. In such cases, the transformations automatically applied by
+COFFEE dramatically decrease the cost of \texttt{assembly}.  This results in
+a significant overall speedup and a simulation which is now dominated by
+\texttt{solve} time. It is interesting to note that
+generalized loop-invariant code motion was particularly invasive in this
+case, with 23 temporaries generated and several redundancies discovered (see
+Section~\ref{sec:code-transf-licm}).
+
+In these experiments, we observe a maximum performance improvement of 1.47$\times$ over the non-optimized local assembly code, obtained in the case ($f=2, p=2$).
 However, we reiterate the fact that full-application speed ups rise proportionally with the amount of time spent in assembly and, therefore, with the complexity of the equation.
@@
 
-\paragraph{Generalizability to other domains} There are neither conceptual nor technical reasons for which our compiler could not be used outside of Firedrake.
-For example, integration with the popular FEniCS framework, the pioneer of automated code generation for finite element local assembly, would be relatively easy to achieve.
-More challenging is to assess the generality of the optimization strategy; that is, if and how COFFEE and its transformations are transferable to other computational domains, perhaps other DSLs, and to what extent this would be helpful for improving full-application performance.
+\paragraph{Generalizability to other domains} There are neither conceptual
+nor technical reasons which prevent our compiler being used in applications
+other than Firedrake. For example, integration with the popular FEniCS
+framework, the pioneer of automated code generation for finite element local
+assembly, would be relatively easy to achieve. It is more challenging to
+assess the generality of the optimization strategy: the extent to which COFFEE and its transformations are transferable to other computational domains, perhaps other DSLs, and to what extent this would be helpful for improving full-application performance.
 To answer these questions, we first need to go back to the origins of our compiler.
@@
 \begin{itemize}
-\item \textbf{Optimizing mathematical expressions.} Expression manipulation (e.g.
-simplification, decomposition into sub-expressions) opens multiple semantically-equivalent code generation opportunities, characterized by different trade-offs in parallelism, redundant computation, and data locality.
-The basic idea is to exploit properties of arithmetic operators, like associativity and commutativity, to re-schedule the computation suitably for the underlying architecture.
+\item \textbf{Optimizing mathematical expressions.} Expression manipulation
+  (e.g. simplification, decomposition into sub-expressions) opens multiple
+  semantically equivalent code generation opportunities, characterized by
+  different trade-offs in parallelism, redundant computation, and data
+  locality. The basic idea is to exploit properties of arithmetic operators,
+  such as associativity and commutativity, to re-schedule the computation suitably for the underlying architecture.
 Loop-invariant code motion and expression splitting follow this principle, so they can be re-adapted or extended to any domains involving numerical evaluation of complex mathematical expressions (e.g.
@@
 whenever the available per-core memory is small).
-\item Associative arithmetic operators are the prerequisite for expression splitting.
-In essence, this transformation concerns resource-aware execution.
-It is indeed true that, in the context of COFFEE, expression splitting has been successfully applied to improve register pressure; nevertheless, the basic idea is far more general, which is re-scheduling (re-associating) operations to optimize for some generic parameters.
-It could be used, for example, as a starting point to perform kernel fission; that is, splitting a kernel into multiple parts, each part characterized by less stringent memory requirements (a variant of this idea for non-affine loops in unstructured mesh applications has been adopted in~\cite{carlolcpc}).
-In Equation~\ref{eq:assembly-model}, for instance, not only any of the functions $\alpha$, $\beta$ and $\gamma$ could be split (assuming they include associative operators), but also $\alpha$ could be completely extracted and evaluated in a separate kernel: this would avoid allocating space for $[a', b', c', ...]$ in a second kernel computing the rest of the expression.
-This may be particularly indicated for many-core architectures in which the available per-core memory is much smaller than that in traditional CPUs.
+\item Associative arithmetic operators are the prerequisite for expression
+  splitting. In essence, this transformation concerns resource-aware
+  execution. In the context of COFFEE, expression splitting has been
+  successfully applied to improve register pressure. However, underlying
+  idea of re-scheduling (re-associating) operations to optimize for some
+  generic parameters is far more general. It could be used, for example, as
+  a starting point to perform kernel fission; that is, splitting a kernel
+  into multiple parts, each part characterized by less stringent memory
+  requirements (a variant of this idea for non-affine loops in unstructured
+  mesh applications has been adopted in~\cite{carlolcpc}). In
+  Equation~\ref{eq:assembly-model}, for instance, not only can  any of the
+  functions $\alpha$, $\beta$ and $\gamma$ be split (assuming they include
+  associative operators), but $\alpha$ could be completely extracted and
+  evaluated in a separate kernel. This would reduce the working set size of
+  each of the kernel functions, an option which is particularly attractive for many-core architectures in which the available per-core memory is much smaller than that in traditional CPUs.
 \end{enumerate}
-\item \textbf{Code generation and applicability of the transformations.} All array sizes and loop bounds, like $n1$ and $n2$ in Equation~\ref{eq:assembly-model}, are known at code generation time.
-This means that ``good'' code can be generated.
-For example, loop bounds can be made explicit, arrays can be statically initialized, and pointer aliasing is easily avoidable.
-Further, all of these factors contribute to the applicability and the effectiveness of some of our code transformations.
-For instance, knowing loop bounds allows both generating correct code when applying vector-register tiling (see Section~\ref{sec:op-vect}) and discovering redundant computation opportunities (see Section~\ref{sec:code-transf-licm}).
-A special case are padding and data alignment since they could be performed even at run-time if some values were not known at code generation time; theoretically, they could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.
-\item \textbf{Multi-loop vectorization.} Compiler auto-vectorization has become increasingly effective in a variety of codes.
-Nevertheless, multi-loop vectorization, which involves loading and storing data along a subset of the loops characterizing the iteration space (rather than just along the innermost loop), is not supported by available general-purpose and polyhedral compilers, to the best of our knowledge.
-The outer-product vectorization technique presented in this paper shows that two-loop vectorization can outperform standard auto-vectorization.
-In addition, we expect the performance gain to scale with the number of vectorized loops and the vector length (as proved by the Xeon Phi experiments).
-Despite being not straightforward to automate multi-loop vectorization in a general-purpose compiler, especially if stencils are present, we believe that this could be more easily achieved in specific domains.
-The intuition is to map the memory access pattern on vector registers, and then exploiting in-register shuffling to minimize the traffic between memory and processor.
+\item \textbf{Code generation and applicability of the transformations.} All
+  array sizes and loop bounds, like $n1$ and $n2$ in
+  Equation~\ref{eq:assembly-model}, are known at code generation time. This
+  means that ``good'' code can be generated. For example, loop bounds can be
+  made explicit, arrays can be statically initialized, and pointer aliasing
+  is easily avoidable. Further, all of these factors contribute to the
+  applicability and the effectiveness of some of our code
+  transformations. For instance, knowing loop bounds allows both generating
+  correct code when applying vector-register tiling (see
+  Section~\ref{sec:op-vect}) and discovering redundant computation
+  opportunities (see Section~\ref{sec:code-transf-licm}). Padding and data
+  alignment are spectial cases, since they could be performed at run-time if some values were not known at code generation time.
+Theoretically, they could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.
+\item \textbf{Multi-loop vectorization.} Compiler auto-vectorization has
+  become increasingly effective in a variety of codes. However, to the best
+  of our knowledge, multi-loop vectorization involving the loading and
+  storing of data along a subset of the loops characterizing the iteration
+  space (rather than just along the innermost loop), is not supported by
+  available general-purpose and polyhedral compilers. The outer-product
+  vectorization technique presented in this paper shows that two-loop
+  vectorization can outperform standard auto-vectorization. In addition, we
+  expect the performance gain to scale with the number of vectorized loops
+  and the vector length (as demonstrated in  the Xeon Phi
+  experiments). Although the automation of multi-loop vectorization in a
+  general-purpose compiler is far from straightforward, especially if stencils are present, we believe that this could be more easily achieved in specific domains.
+The intuition is to map the memory access pattern on vector registers, and then exploit in-register shuffling to minimize the traffic between memory and processor.
 By demonstrating the effectiveness of multi-loop vectorization in a real scenario, our research represents an incentive for studying this technique in a broader and systematic way.

### 2014-09-04 TACO-paper-v2/TACO-Rebuttal.txt
Start writing TACO rebuttal

@@
 We thank our reviewers (R1-R4). We appreciate the positive feedback and the
-critical comments on our paper. Below we answers the various points arose by the
+critical comments on our paper. Below we answer the various points arose by the
 reviewers.
+
+R1:
+We agree the big missing part was a discussion on the generality of our
+approach and the applicability of the code transformations to other
+computational domains (or, perhaps, their integration within a general-purpose
+compilers). We added therefore a complete new section (Section 6) that explains
+these aspects in details. We believe this addition was necessary to let readers
+understand how to transfer our work to a different domain or compiler (or, at
+least, to let them gain something from our experience of designing/developing 
+the code transformations and COFFEE).
+
+We also addressed all of the other useful comments (1-8) provided:
+1) The goal is to develop a class of optimizations for an infinite variety of 
+local assembly code. In that paragraph we say "(...) makes it hard to determine 
+a single or specific sequence of transformations that is successfully 
+applicable to all problems (...)", which implies that different finite element 
+problems may require different code optimizations due to the inherent diversity 
+in mathematical expressions and loop trip counts (this aspect is reiterated 
+throughout the paper). Therefore, a compiler-based approach, in which the input 
+and some parameters are analyzed to decide how to optimally transform the code 
+is needed to get as close as possible to peak performance. Another possibility 
+would be a library-based approach, although it would be quite pointless since 
+we can directly generate the most suitable code for each problem.
+2) The variation in loop trip counts impacts loop transformation. If the loop 
+is relatively large, for example, vector-register tiling is likely to be useful 
+to minimize register spilling; if it's small, then full or partial unrolling is 
+an option. Understanding the border between "small" and "large" is challenging.  
+The cost model helps in this direction. We tried to clarify the point you make 
+in various sections in the paper: the sentence at page 6 has been re-written to 
+anticipate the fact that "different loop sizes require different 
+transformations". Then we discuss on the importance and the impact of loop 
+unrolling in Section 4 (there is a complete new paragraph about that).  
+Finally, in Section 5.2.5, we demonstrate our claim as discussing the impact of 
+vector-register tiling (paragraphs 4, 5 and 6 of that section) in our 
+benchmarks.
+3) Clarified in the paper by adding/re-writing a few sentences. The short 
+answer is: yes.
+4) Explained in the paper.
+5) We agree. We removed the code listing and expanded Figure 4.
+6) Explained in the paper.
+7) This was already explained by the sentence "Other optimization levels 
+performed, in general, slightly worse."
+8) We believe each of those tables can be represented by 6 different charts (1 
+for each <Architecture, Equation> examined>). We have not done that for lack of 
+space (the submission for acceptance is limited to 25 pages) but we can easily 
+do that, if really needed, in the final version of the paper (which can be 
+30-page long)
+
+******************************************************************************
+
+R2:
+Two main critiques are moved to our paper.
+
+a) We do not think that "the techniques of the paper are mostly interesting to 
+a subset of the people in the intersection of code optimization and finite 
+element methods". However, we do agree that the paper was missing an in-depth 
+explanation about how to re-adapt/re-use our transformations in other contexts, 
+perhaps other DSLs or even general-purpose compilers. Also, we believe that our 
+experience of developing COFFEE and integrating it in a real-world framework 
+based on a domain-specific language can be of inspiration for other research 
+and/or engineering work. We therefore introduced a complete new section, 
+Section 6, about the generality of our approach and the applicability of the 
+code transformations to other computational domains. We hope you can read this 
+section carefully, because it represents novel contribution of the paper and 
+aims to answer exactly the concern you pointed out.
+
+b) The results of the paper are *really* applicable to other kernels beyond 
+those analyzed and presented. First of all, we want to emphasize a number of 
+aspects. What is important to clarify is that kernels are not "mere small 
+pieces of code" but rather numerical routines that evaluate an integral in each 
+element of the discretized equation domain, produced by means of a real (used 
+by many scientists around the world) domain-specific language (and compiler).  
+Therefore, there is an infinite variety of kernels that can be generated in 
+Firedrake, and their code structure (loop sizes and mathematical expression) 
+depends on the partial differential equation that is being solved. This is 
+explained and re-iterated in the paper. Also, when we vary the polynomial order 
+of the method or the domain discretization (triangles, tetrahedrons, prisms) we 
+actually change the code generation by increasing loops and arrays sizes and by 
+making mathematical expressions more articulated (this happens, for example, 
+when moving from 2D domains to 3D ones). So we are already testing many 
+different code variants. This is carefully explained in Section 5.2.1. Now, 
+premised that, back to your very point, which is "applicability and 
+effectiveness of our transformations in other local assembly kernels". Our 
+answer is twofold:
+
+- Section 5.2.1 explains in-depth the motivations behind the choice of the 
+  three problems (helmholtz, Diffusion, Burgers), explaining (in short) that 
+"(...) distinct problems, possibly arising in completely different fields, may 
+employ (subsets of) the same differential operators of our benchmarks, which 
+implies similarities and redundant patterns in the generated code. 
+Consequently, the proposed code transformations have a domain of applicability 
+that goes far beyond that of the three analyzed equations. (...)". Then we 
+re-inforce this later on in the section and in the paper.
+
+- In Section 5.3 we provide a full-application study in which we analyze a 
+  different equation (Linear Elasticity), which brings the total number of 
+studied equations to 4 (Helmholtz, Diffusion, Burgers, Elasticity)
+Adding more kernels to the performance evaluation section would make the paper 
+only more chaotic: we selected these three problems to highlight all pros and 
+cons of our code transformations. This is the reason for which we comment on 
+the code transformations and their interplay in specific sub-sections, 
+referring to, describing, and carefully analyzing all of the provided plots.
+
+Other kernels arising from completely different equations will still have the 
+same high-level structure of the kernels we examined (e.g. same memory access 
+pattern, same ranges of loop sizes, same problems concerning the presence of 
+loop-invariant code and register pressure) so there is no reason for which our 
+code optimizations would not be effective. The critical issue in our research 
+was to understand what transformations our local assembly kernels required, and 
+how to instantiate parametric transformations (vector-register tiling, loop 
+interchange, expression splitting) optimally in each possible problem (this led 
+to the development of the cost model).

### 2014-09-04 TACO-paper-v2/TACO-Rebuttal.txt
Answer R2

@@
 that goes far beyond that of the three analyzed equations. (...)". Then we 
-re-inforce this later on in the section and in the paper.
+re-inforce this later on in the section and in the paper. Other kernels arising 
+from completely different equations will still have the same high-level 
+structure of the kernels we examined (e.g. same memory access pattern, same 
+ranges of loop sizes, same problems concerning the presence of loop-invariant 
+code and register pressure) so there is no reason for which our code 
+optimizations would not be effective. The critical issue in our research was to 
+understand what transformations our local assembly kernels required, and how to 
+instantiate parametric transformations (vector-register tiling, loop 
+interchange, expression splitting) optimally in each possible problem (this led 
+to the development of the cost model).
 
@@
 
-Other kernels arising from completely different equations will still have the 
-same high-level structure of the kernels we examined (e.g. same memory access 
-pattern, same ranges of loop sizes, same problems concerning the presence of 
-loop-invariant code and register pressure) so there is no reason for which our 
-code optimizations would not be effective. The critical issue in our research 
-was to understand what transformations our local assembly kernels required, and 
-how to instantiate parametric transformations (vector-register tiling, loop 
-interchange, expression splitting) optimally in each possible problem (this led 
-to the development of the cost model).
+Other concerns:
+- It is actually not true that we do not consider other optimization 
+  strategies: 1) a comparison with using linear algebra libriries (the 
+comparison uses hand-crafted kernels) and 2) a comparison with the FEniCS Form 
+Compiler's built-in optimization system (which is the result of a cited paper) 
+are provided.
+- We agree it is useful to report speed ups and execution times for the whole finite element 
+  calculation. We have added a new section (Section 5.3) about a 
+full-application study, showing exactly these values.
+- At some points in the paper (introduction and related work) we cite studies 
+  concerning local assembly and GPUs. These works however deeply differ 
+from ours for a number of reasons, including the fact that they are outside of 
+the context of automated code generation, they are limited to only very 
+low-order methods (typically only polynomial order 1), and they do not try to 
+optimize for cross-loop arithmetic intensity as we do. The goal in these works 
+is rather to understand how to effectively parallelize the local assembly code (so as 
+to have multiple threads executing a single kernel, as opposed to our execution 
+model in which we assume a single thread per kernel) in specific equations (not 
+in a possibly infinite range as we do). A critical issue when executing local 
+assembly kernels on GPUs is represented by memory requirements. We give here 
+two examples, although more problems could be discussed. First, the transfer of 
+the mesh from host to GPU may be a notable obstacle: if too big, the mesh
+will not fit the available GPU memory, and distributed execution, which is 
+marginally supported in available finite element frameworks, would be required. 
+Secondly, optimizations like generalized loop-invariant code motion further increase 
+the memory requirements due to the introduction of temporary arrays. This may 
+actually lead to poor per-thread shared memory and registers allocation. These 
+aspects suggest that the problem of executing generic, automatically-generated local 
+assembly kernels on GPUs require a separate, in-depth study. However, in the 
+new Section 6, we provide some intuitions of how we could re-adapt COFFEE's 
+transformations for execution on GPUs.

### 2014-09-04 TACO-paper-v2/TACO-Rebuttal.txt
Start answering R3

@@
 are provided.
-- We agree it is useful to report speed ups and execution times for the whole finite element 
-  calculation. We have added a new section (Section 5.3) about a 
+- We agree it is useful to report speed ups and execution times for the whole 
+  finite element calculation. We have added a new section (Section 5.3) about a 
 full-application study, showing exactly these values.
 - At some points in the paper (introduction and related work) we cite studies 
-  concerning local assembly and GPUs. These works however deeply differ 
-from ours for a number of reasons, including the fact that they are outside of 
-the context of automated code generation, they are limited to only very 
-low-order methods (typically only polynomial order 1), and they do not try to 
-optimize for cross-loop arithmetic intensity as we do. The goal in these works 
-is rather to understand how to effectively parallelize the local assembly code (so as 
-to have multiple threads executing a single kernel, as opposed to our execution 
+  concerning local assembly and GPUs. These works however deeply differ from 
+ours for a number of reasons, including the fact that they are outside of the 
+context of automated code generation, they are limited to only very low-order 
+methods (typically only polynomial order 1), and they do not try to optimize 
+for cross-loop arithmetic intensity as we do. The goal in these works is rather 
+to understand how to effectively parallelize the local assembly code (so as to 
+have multiple threads executing a single kernel, as opposed to our execution 
 model in which we assume a single thread per kernel) in specific equations (not 
@@
 marginally supported in available finite element frameworks, would be required.  
-Secondly, optimizations like generalized loop-invariant code motion further increase 
-the memory requirements due to the introduction of temporary arrays. This may 
-actually lead to poor per-thread shared memory and registers allocation. These 
-aspects suggest that the problem of executing generic, automatically-generated local 
-assembly kernels on GPUs require a separate, in-depth study. However, in the 
-new Section 6, we provide some intuitions of how we could re-adapt COFFEE's 
-transformations for execution on GPUs.
+Secondly, optimizations like generalized loop-invariant code motion further 
+increase the memory requirements due to the introduction of temporary arrays.  
+This may actually lead to poor per-thread shared memory and registers 
+allocation. These aspects suggest that the problem of executing generic, 
+automatically-generated local assembly kernels on GPUs require a separate, 
+in-depth study. However, in the new Section 6, we provide some intuitions of 
+how we could re-adapt COFFEE's transformations for execution on GPUs.
+
+******************************************************************************
+
+R3:
+We answer here the 5 points of the review.
+1) For what concerns execution on a GPGPU, we replicate here the same answer to 
+R2.
+2) We agree that profile-guided optimization could be used to improve the 
+performance of kernels, although this aspect is orthogonal to most of our code
+transformations, because in no way profile information could be used, for 
+example, to trigger generalized loop-invariant code motion and the avoidance of 
+redundant computation (as explained in Section 3.2). Profile-guided 
+optimization could theoretically be used for padding and data-alignment of 
+arrays, provided that the compiler has sufficient machinery in place to perform 
+(we quote the revised paper, Section 6) "some sort of data-flow analysis (...) 
+to ensure that the extra loop iterations over the padded region do not affect 
+the numerical results.". We discuss on the generality of our code 
+transformation and on their integrability in a general-purpose compiler (and in 
+its profile-guided optimization system) in the new Section 6 of the paper, 
+which also represents novel contribution of our work.
+3) TODO
+4) Firedrake is *not* specifically designed for Intel architectures and 
+compilers. The decision of using Intel platforms and tools is twofold: from one 
+hand, we believe that Intel's is one of the most powerful vectorizing and 
+optimizing compilers, so for a research like ours it was important to compare 
+to the best available on the market; from the other hand, their wide 
+availability represents a strong incentive for studying the problems object 
+of the paper on these machines.
+It is also worth noting that all general-purpose compilers are, to different 
+extents, limited in optimizing for register locality and code motion (for 
+instance, this is true for the Intel and GNU compilers, the latter only used 
+for "local" experimentation), so our work is expected to be equally/similarly 
+effective on different "standard" CPU architectures.
+From the practical viewpoint, we observe that COFFEE introduces optimizations 
+aimed at improving cross-loop arithmetic intensity by performing source-level 
+transformations, so there are absolutely no technical issues for 
+straightforwardly using our domain-specific compiler on architectures different 
+than Intel. Actually, this fact is a strength of our work, because it allows 
+COFFEE to be almost completely platform-independent.  The only exception 
+regards intrinsics code: COFFEE is currently only capable of outputing AVX 
+intrinsics, so at the moment we cannot apply vector-register tiling (the also 
+called outer-product vectorization technique) to architectures different than 
+x86+AVX. However, COFFEE's internal algorithms use an intermediate 
+representation of intrinsics code, so extending our compiler to other 
+instruction sets would be just a matter of adding a new dictionary of 
+compiler-specific intrinsics.
+5) We agree that expression split has many issues involved. Our cost model 
+offers a simple alternative for driving expression splitting. Simplicity is the 
+key to effectiveness, portability, and robustness. Probably using a graph 
+partitioning approach as you rightly point out would allow us to obtain better 
+performance gains, but three points are worth to be highlighted:
+- The split factor is a parameter of the transformation. In our 
+  experiments we have tested several split factors, which means testing as many 
+code variants (i.e. as many ways of splitting an expression based on sum's 
+associativity) and, as shown by the various graphs in the paper, our cost model 
+seems to be capable of properly approximating such a parameter. This is of 
+course encouraging.
+- In the perspective of integrating expression splitting in a general-purpose 
+  compiler, or generalizing this technique to other problems (as elaborated in 
+the new Section 6), we feel that a simple model is preferable to keep the 
+implementation complexity as low as possible.
+- TODO
+
+******************************************************************************
+
+R4:

### 2014-09-05 TACO-paper-v2/TACO-Rebuttal.txt
Add bits to rebuttal

@@
 transformations, because in no way profile information could be used, for 
-example, to trigger generalized loop-invariant code motion and the avoidance of 
+example, to trigger generalized loop-invariant code motion and avoidance of 
 redundant computation (as explained in Section 3.2). Profile-guided 
@@
 to the best available on the market; from the other hand, their wide 
-availability represents a strong incentive for studying the problems object 
-of the paper on these machines.
+availability represents a strong incentive for studying the problems object of 
+the paper on these machines.
 It is also worth noting that all general-purpose compilers are, to different 
@@
 performance gains, but three points are worth to be highlighted:
-- The split factor is a parameter of the transformation. In our 
-  experiments we have tested several split factors, which means testing as many 
-code variants (i.e. as many ways of splitting an expression based on sum's 
-associativity) and, as shown by the various graphs in the paper, our cost model 
-seems to be capable of properly approximating such a parameter. This is of 
-course encouraging.
+- The split factor is a parameter of the transformation. In our experiments we 
+  have tested several split factors, which means testing as many code variants 
+(i.e. as many ways of splitting an expression based on sum's associativity) 
+and, as shown by the various graphs in the paper, our cost model seems to be 
+capable of properly approximating such a parameter. This is of course 
+encouraging.
 - In the perspective of integrating expression splitting in a general-purpose 
@@
 
+Other comments:
+a) We elaborate on the use of loop unwinding (or unrolling) in Section 4, in 
+which we have added a new paragraph (named "Loop unrolling"). This paragraph 
+elaborates on "complete/partial loop unwinding" as well as other general 
+aspects concerning pros and cons of applying this transformation.
+b) It was copy-paste leftover, now fixed
+c) Right, fixed by making the statement less strong
+d) This is all clarified and explained in details in the new Section 6. The 
+short answer is that we hope COFFEE can be re-used in frameworks other than 
+Firedrake (e.g. the FEniCS framework that we compare to in the paper).
+
 ******************************************************************************
@@
 R4:
+We addressed all of the observations in your review:
+1) Fixed, thanks
+2) The polyhedral model is well-known among compilers people. Basically, it is 
+a mathematical model for representing affine loop nests and determining in a 
+systemtic way all of its possible legal transformations. Cost models are often 
+used by polyhedral compilers to estimate the effectiveness of a certain 
+transformation. Polyhedral optimizations often aim at improving cache locality.  
+In the Related Work section, we cite one of the most relevant researches on
+polyhedral compilers; note that one of the co-authors of our paper is also a 
+co-author of that paper (i.e. he is an expert of polyhedral compilers).
+3) The speed ups only refer to the computation of the local matrix. This is now 
+clarified in Section 5.1, and also re-iterated later on in the paper. 
+We used different meshes for the various experiments. There is a link to 
+the code of the experiments as well as the mesh sizes employed, for each 
+experiment. This is now explained in the paper, Section 5.2.1.
+4)

### 2014-09-05 TACO-paper-v2/TACO-Rebuttal.txt
Answer R4

@@
 R2.
+
 2) We agree that profile-guided optimization could be used to improve the 
@@
 which also represents novel contribution of our work.
+
 3) TODO
+
 4) Firedrake is *not* specifically designed for Intel architectures and 
@@
 compiler-specific intrinsics.
+
 5) We agree that expression split has many issues involved. Our cost model 
@@
 aspects concerning pros and cons of applying this transformation.
+
 b) It was copy-paste leftover, now fixed
+
 c) Right, fixed by making the statement less strong
+
 d) This is all clarified and explained in details in the new Section 6. The 
@@
 1) Fixed, thanks
+
 2) The polyhedral model is well-known among compilers people. Basically, it is 
@@
 co-author of that paper (i.e. he is an expert of polyhedral compilers).
-3) The speed ups only refer to the computation of the local matrix. This is now 
-clarified in Section 5.1, and also re-iterated later on in the paper. 
-We used different meshes for the various experiments. There is a link to 
-the code of the experiments as well as the mesh sizes employed, for each 
-experiment. This is now explained in the paper, Section 5.2.1.
-4)
+
+3 and 4) The speed ups only refer to the computation of the local element 
+matrix.  This is now clarified in Section 5.1, and also re-iterated later on in 
+the paper. We used different meshes for the various experiments. In Section 
+5.2.1 we provide a link to a repository in which the code of the experiments as 
+well as the mesh sizes employed can be found. We note that for these 
+experiments the mesh size was set such that the working set of the computation 
+could at least fit the last level cache (L3 on the Sandy Bridge architecture). 
+The goal of these experiments (Section 5.2 in the revised paper) was only to 
+measure speed ups achieved by optimizing local assembly kernels, so there was 
+no point in running over big meshes (i.e. meshes that would not fit the last 
+level cache), which would have only increased the completion time of our test 
+suite (which is already order of hours).
+We however completely agree that reporting execution times and speed ups of the 
+overall finite element calculation is of fundamental importance. Therefore, 
+following your suggestion, we have added a new section to the paper, Section 
+5.3, which is about a full-application study concerning a linear elasticity 
+equation.  This means that the performance evaluation section is now structured 
+in two parts: Section 5.2, about local assembly speed ups; Section 5.3, about a 
+real application study. We think this is the best structuring to let readers 
+understand the effectiveness of our transformations "at the level of loop 
+nests" as well as their impact on the whole application. In the new section, we 
+provide the numbers you suggested: we show execution times of assembly and 
+solve, we comment on the achieved speed ups, and we highlight the cases in 
+which matrix insertion is negligible because of the high computational cost of 
+local assembly. For this application, we use a realistic mesh composed of 
+196608 cells (as reported in the paper). We also provide a link to the code of 
+the application.

### 2014-09-06 TACO-paper-v2/coffee.tex
minor fix
Try to gain some space in the paper

@@
 \begin{algorithm}[t]
-\footnotesize
+\scriptsize
 \SetAlgorithmName{LISTING}{}
-\KwSty{void} helmholtz(double A[3][3], double **coords) $\lbrace$\\
+
+\KwSty{void} helmholtz(\KwSty{double} A[3][3], \KwSty{double} **coords) $\lbrace$\\
 ~~// K, det = Compute Jacobian (coords) \\
@@
 \begin{algorithm}[t]
-\footnotesize
+\scriptsize
 \SetAlgorithmName{LISTING}{}
-\KwSty{void} burgers(double A[12][12], double **coords, double **w) $\lbrace$\\
+
+\KwSty{void} burgers(\KwSty{double} A[12][12], \KwSty{double} **coords, \KwSty{double} **w) $\lbrace$\\
 ~~// K, det = Compute Jacobian (coords) \\
@@
 \begin{algorithm}[t]
-\footnotesize
+\scriptsize
 \SetAlgorithmName{LISTING}{}
-\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
+
+\KwSty{void} helmholtz(\KwSty{double} A[3][4], \KwSty{double} **coords) $\lbrace$\\
 ~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
@@
 ~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
-~~~~double LI$\_$0[4] ALIGN;\\
-~~~~double LI$\_$1[4] ALIGN;\\
+~~~~\KwSty{double} LI$\_$0[4] ALIGN;\\
+~~~~\KwSty{double} LI$\_$1[4] ALIGN;\\
 ~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
@@
 \begin{algorithm}[t]
-\footnotesize
+\scriptsize
 \SetAlgorithmName{LISTING}{}
-\KwSty{void} helmholtz(double A[4][4], double **coords) $\lbrace$\\
+
+\KwSty{void} helmholtz(\KwSty{double} A[4][4], \KwSty{double} **coords) $\lbrace$\\
 ~~// K, det = Compute Jacobian (coords) \\
@@
 ~~\KwSty{for} (\KwSty{int} j = 0; j$<$4; j+=4) $\lbrace$\\
-~~~~$\_\_$m256d r0, r1, r2, r3, r4, r5, r6, r7;\\
+~~~~\KwSty{$\_\_$m256d} r0, r1, r2, r3, r4, r5, r6, r7;\\
 ~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k+=4) $\lbrace$\\
@@
 \begin{algorithm}
-\footnotesize
+\scriptsize
 \SetAlgorithmName{LISTING}{}
-\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
+
+\KwSty{void} helmholtz(\KwSty{double} A[3][4], \KwSty{double} **coords) $\lbrace$\\
 ~~~~// Same code as in Listing~\ref{code:helmholtz-licm} up to the j loop\\
@@
   opportunities (see Section~\ref{sec:code-transf-licm}). Padding and data
-  alignment are spectial cases, since they could be performed at run-time if some values were not known at code generation time.
+  alignment are special cases, since they could be performed at run-time if some values were not known at code generation time.
 Theoretically, they could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.

### 2014-09-08 TACO-paper-v2/TACO-Rebuttal.txt
Rewrite bits of the rebuttal document

@@
 1) The goal is to develop a class of optimizations for an infinite variety of 
-local assembly code. In that paragraph we say "(...) makes it hard to determine 
-a single or specific sequence of transformations that is successfully 
+local assembly codes. In that paragraph we say "(...) makes it hard to 
+determine a single or specific sequence of transformations that is successfully 
 applicable to all problems (...)", which implies that different finite element 
@@
 and some parameters are analyzed to decide how to optimally transform the code 
-is needed to get as close as possible to peak performance. Another possibility 
-would be a library-based approach, although it would be quite pointless since 
-we can directly generate the most suitable code for each problem.
-2) The variation in loop trip counts impacts loop transformation. If the loop 
+is needed to maximize the performance. Another possibility would be a 
+library-based approach, although it would be quite pointless since we can 
+generate the most suitable code for each problem.
+2) The variation in loop trip counts impacts loops transformation. If the loop 
 is relatively large, for example, vector-register tiling is likely to be useful 
-to minimize register spilling; if it's small, then full or partial unrolling is 
-an option. Understanding the border between "small" and "large" is challenging.  
-The cost model helps in this direction. We tried to clarify the point you make 
-in various sections in the paper: the sentence at page 6 has been re-written to 
-anticipate the fact that "different loop sizes require different 
-transformations". Then we discuss on the importance and the impact of loop 
-unrolling in Section 4 (there is a complete new paragraph about that).  
-Finally, in Section 5.2.5, we demonstrate our claim as discussing the impact of 
-vector-register tiling (paragraphs 4, 5 and 6 of that section) in our 
+to minimize register spilling; if it is small, then full or partial unrolling 
+is an option. Understanding the border between "small" and "large" is 
+challenging.  The cost model helps in this direction. We tried to clarify the 
+point you make in various sections in the paper: the sentence at page 6 has 
+been re-written to anticipate the fact that "different loop sizes require 
+different transformations". Then we elaborate on the importance and the impact 
+of loop unrolling in Section 4 (there is a complete new paragraph about that).  
+Finally, in Section 5.2.5, we demonstrate our claim when discussing the impact 
+of vector-register tiling (paragraphs 4, 5 and 6 of that section) in our 
 benchmarks.
@@
 aspects. What is important to clarify is that kernels are not "mere small 
-pieces of code" but rather numerical routines that evaluate an integral in each 
-element of the discretized equation domain, produced by means of a real (used 
-by many scientists around the world) domain-specific language (and compiler).  
-Therefore, there is an infinite variety of kernels that can be generated in 
-Firedrake, and their code structure (loop sizes and mathematical expression) 
-depends on the partial differential equation that is being solved. This is 
-explained and re-iterated in the paper. Also, when we vary the polynomial order 
-of the method or the domain discretization (triangles, tetrahedrons, prisms) we 
-actually change the code generation by increasing loops and arrays sizes and by 
-making mathematical expressions more articulated (this happens, for example, 
-when moving from 2D domains to 3D ones). So we are already testing many 
-different code variants. This is carefully explained in Section 5.2.1. Now, 
-premised that, back to your very point, which is "applicability and 
-effectiveness of our transformations in other local assembly kernels". Our 
-answer is twofold:
+pieces of code", but rather, in general, quite complex numerical routines 
+evaluating integrals in an
+element of the discretized equation domain, produced by means of real (used by 
+many scientists around the world) domain-specific language and compiler.  There 
+is an infinite variety of kernels that can be generated in Firedrake; the 
+generated code (e.g. loop sizes and mathematical expression) depends on the 
+partial differential equation being solved. This is explained and re-iterated 
+in the paper. Also, varying the polynomial order of the method or the element 
+shape (e.g. triangles, tetrahedrons, prisms), as we do and explain in the 
+performance evaluation section, changes the generated code by increasing loops 
+and arrays sizes and by making mathematical expressions more articulated (this 
+happens, for example, when moving from 2D to 3D domains). So we are already 
+testing many different code variants (three different kernels, and then for 
+each kernel we try multiple polynomial orders and element shapes). This is 
+carefully explained in Section 5.2.1. Now, premised this, back to your point, 
+which is "applicability and effectiveness of our transformations in other local 
+assembly kernels". Our answer is twofold:
 
 - Section 5.2.1 explains in-depth the motivations behind the choice of the 
-  three problems (helmholtz, Diffusion, Burgers), explaining (in short) that 
-"(...) distinct problems, possibly arising in completely different fields, may 
-employ (subsets of) the same differential operators of our benchmarks, which 
-implies similarities and redundant patterns in the generated code.  
-Consequently, the proposed code transformations have a domain of applicability 
-that goes far beyond that of the three analyzed equations. (...)". Then we 
-re-inforce this later on in the section and in the paper. Other kernels arising 
-from completely different equations will still have the same high-level 
-structure of the kernels we examined (e.g. same memory access pattern, same 
-ranges of loop sizes, same problems concerning the presence of loop-invariant 
-code and register pressure) so there is no reason for which our code 
-optimizations would not be effective. The critical issue in our research was to 
-understand what transformations our local assembly kernels required, and how to 
-instantiate parametric transformations (vector-register tiling, loop 
-interchange, expression splitting) optimally in each possible problem (this led 
-to the development of the cost model).
+  three problems (Helmholtz, Diffusion, Burgers), explaining that "(...) 
+distinct problems, possibly arising in completely different fields, may employ 
+(subsets of) the same differential operators of our benchmarks, which implies 
+similarities and redundant patterns in the generated code.  Consequently, the 
+proposed code transformations have a domain of applicability that goes far 
+beyond that of the three analyzed equations. (...)". Then we re-inforce this 
+later on in the section and in the paper. Other kernels arising from completely 
+different equations will still have the same structure of the kernels we 
+examined (e.g. same memory access pattern, same ranges of loop sizes, same 
+problems concerning the presence of loop-invariant code and register pressure) 
+as well as similar patterns in the mathematical expressions, so there is no 
+evident reason for which our code optimizations would not be effective. The 
+critical issue in our research was to understand what transformations our local 
+assembly kernels required, and how to instantiate parametric transformations 
+(vector-register tiling, loop interchange, expression splitting) optimally in 
+each possible problem (this led to the development of the cost model).
 
@@
 cons of our code transformations. This is the reason for which we comment on 
-the code transformations and their interplay in specific sub-sections, 
-referring to, describing, and carefully analyzing all of the provided plots.
+the performance of the code transformations (and their interplay) in specific 
+sub-sections, referring to, describing, and carefully analyzing the provided 
+plots.
 
@@
 full-application study, showing exactly these values.
-- At some points in the paper (introduction and related work) we cite studies 
+- In some points in the paper (introduction and related work) we cite studies 
   concerning local assembly and GPUs. These works however deeply differ from 
-ours for a number of reasons, including the fact that they are outside of the 
-context of automated code generation, they are limited to only very low-order 
-methods (typically only polynomial order 1), and they do not try to optimize 
-for cross-loop arithmetic intensity as we do. The goal in these works is rather 
-to understand how to effectively parallelize the local assembly code (so as to 
+ours for a number of reasons, including the facts that they do not rely on 
+automated code generation, they are limited to only very low-order methods 
+(typically only polynomial order 1), and they do not try to optimize for 
+cross-loop arithmetic intensity as we do. The goal in these works is rather to 
+understand how to effectively parallelize the local assembly code (so as to 
 have multiple threads executing a single kernel, as opposed to our execution 
@@
 assembly kernels on GPUs is represented by memory requirements. We give here 
-two examples, although more problems could be discussed. First, the transfer of 
-the mesh from host to GPU may be a notable obstacle: if too big, the mesh
+two examples of this problem, although more issues could be found. First, the 
+transfer of the mesh from host to GPU may be a notable obstacle: if too big, 
+the mesh
 will not fit the available GPU memory, and distributed execution, which is 
 marginally supported in available finite element frameworks, would be required.  
-Secondly, optimizations like generalized loop-invariant code motion further 
-increase the memory requirements due to the introduction of temporary arrays.  
-This may actually lead to poor per-thread shared memory and registers 
-allocation. These aspects suggest that the problem of executing generic, 
-automatically-generated local assembly kernels on GPUs require a separate, 
-in-depth study. However, in the new Section 6, we provide some intuitions of 
-how we could re-adapt COFFEE's transformations for execution on GPUs.
+Secondly, optimizations such as generalized loop-invariant code motion increase 
+the memory requirements due to the introduction of temporary arrays.  This may 
+actually lead to poor per-thread shared memory and registers allocation. These 
+aspects suggest that the problem of executing generic, automatically-generated 
+local assembly kernels on GPUs require a separate, in-depth study. However, in 
+the new Section 6, we provide some intuitions of how we could re-adapt COFFEE's 
+transformations for execution on GPUs.  Therefore, we invite once more to go 
+through the new Section 6 carefully.
 
@@
 
+In some points in the paper (introduction and related work) we cite studies 
+concerning local assembly and GPUs. These works however deeply differ from ours 
+for a number of reasons, including the facts that they do not rely on automated 
+code generation, they are limited to only very low-order methods (typically 
+only polynomial order 1), and they do not try to optimize for cross-loop 
+arithmetic intensity as we do. The goal in these works is rather to understand 
+how to effectively parallelize the local assembly code (so as to have multiple 
+threads executing a single kernel, as opposed to our execution model in which 
+we assume a single thread per kernel) in specific equations (not in a possibly 
+infinite range as we do). A critical issue when executing local assembly 
+kernels on GPUs is represented by memory requirements. We give here two 
+examples of this problem, although many more issues could be found. First, the 
+transfer of the mesh from host to device may be a great obstacle: if too big, 
+the mesh
+will not fit the available GPU memory, and distributed execution, which is 
+marginally supported in available finite element frameworks, would be required.  
+Secondly, optimizations such as generalized loop-invariant code motion increase 
+the memory requirements due to the introduction of temporary arrays.  This may 
+actually lead to poor per-thread shared memory and registers allocation. These 
+aspects suggest that the problem of executing generic, automatically-generated 
+local assembly kernels on GPUs require a separate, in-depth study. However, in 
+the new Section 6 of the revised paper, we provide some intuitions of how we 
+could re-adapt COFFEE's transformations for execution on GPUs.
+
 2) We agree that profile-guided optimization could be used to improve the 
 performance of kernels, although this aspect is orthogonal to most of our code
-transformations, because in no way profile information could be used, for 
+transformations, because in no way profile information could suggest, for 
 example, to trigger generalized loop-invariant code motion and avoidance of 
-redundant computation (as explained in Section 3.2). Profile-guided 
-optimization could theoretically be used for padding and data-alignment of 
-arrays, provided that the compiler has sufficient machinery in place to perform 
-(we quote the revised paper, Section 6) "some sort of data-flow analysis (...) 
-to ensure that the extra loop iterations over the padded region do not affect 
-the numerical results.". We discuss on the generality of our code 
-transformation and on their integrability in a general-purpose compiler (and in 
-its profile-guided optimization system) in the new Section 6 of the paper, 
-which also represents novel contribution of our work.
+redundant computation (described in Section 3.2). Profile-guided optimization 
+could theoretically be used for padding and data-alignment of arrays, provided 
+that the compiler has sufficient machinery in place to perform (we quote the 
+revised paper, Section 6) "some sort of data-flow analysis (...) to ensure that 
+the extra loop iterations over the padded region do not affect the numerical 
+results.". We discuss on the generality of our code transformation and on their 
+integrability in a general-purpose compiler (and in its profile-guided 
+optimization system) in the new Section 6 of the paper, which also represents 
+novel contribution of our work.
 
@@
 to the best available on the market; from the other hand, their wide 
-availability represents a strong incentive for studying the problems object of 
-the paper on these machines.
+availability represents a strong incentive because users will be likely to run 
+their experiments on these machines.
 It is also worth noting that all general-purpose compilers are, to different 
 extents, limited in optimizing for register locality and code motion (for 
-instance, this is true for the Intel and GNU compilers, the latter only used 
-for "local" experimentation), so our work is expected to be equally/similarly 
-effective on different "standard" CPU architectures.
+instance, this is true for the Intel and GNU compilers, although we only used 
+the latter for "local" experimentation), so our work is expected to be 
+equally/similarly effective on different yet "standard" CPU architectures.
 From the practical viewpoint, we observe that COFFEE introduces optimizations 
 aimed at improving cross-loop arithmetic intensity by performing source-level 
-transformations, so there are absolutely no technical issues for 
-straightforwardly using our domain-specific compiler on architectures different 
-than Intel. Actually, this fact is a strength of our work, because it allows 
-COFFEE to be almost completely platform-independent.  The only exception 
-regards intrinsics code: COFFEE is currently only capable of outputing AVX 
-intrinsics, so at the moment we cannot apply vector-register tiling (the also 
-called outer-product vectorization technique) to architectures different than 
-x86+AVX. However, COFFEE's internal algorithms use an intermediate 
-representation of intrinsics code, so extending our compiler to other 
-instruction sets would be just a matter of adding a new dictionary of 
-compiler-specific intrinsics.
+transformations, so there are absolutely no technical issues for using our 
+domain-specific compiler on architectures different than Intel. Actually, this 
+fact is a strength of our work, because it allows COFFEE to be almost 
+completely platform-independent.  The only exception regards intrinsics code: 
+COFFEE is currently only capable of outputing AVX intrinsics, so at the moment 
+we cannot apply vector-register tiling (the also called outer-product 
+vectorization technique) to architectures different than x86+AVX and x86+AVX2. 
+However, COFFEE's internal algorithms use an intermediate representation of 
+intrinsics code, so extending our compiler to other instruction sets would be 
+just a matter of adding a new dictionary of intrinsics functions.
 
@@
   have tested several split factors, which means testing as many code variants 
-(i.e. as many ways of splitting an expression based on sum's associativity) 
+(i.e. as many as the ways an expression can be split based on sum's associativity) 
 and, as shown by the various graphs in the paper, our cost model seems to be 
@@
 implementation complexity as low as possible.
-- TODO
+- Graph partitioning for expression splitting seems to us a complete new 
+  research problem. On the other hand, this transformation is only one of the 
+many applicable by COFFEE, and we just do not want the focus of the paper being 
+on "how to optimally and rapidly split a mathematical expression". Rather, we 
+prefer to show that, for our problems, there is clearly a problem of register pressure 
+and that by rescheduling the computation in different loops (and 
+sub-expressions) we can actually obtain a performance gain. We have a strategy 
+for expression splitting; we are not claiming it is the best strategy, but at 
+least we demonstrate it is effective and easily predictable.
 

### 2014-09-08 TACO-paper-v2/coffee.tex
Introduction: copy editing

@@
 partial differential equations (PDEs). Numerical techniques, like the
-finite volume method and the finite element method, are widely
-employed to approximate solutions of these PDEs. Unstructured meshes
-are often used to discretize the computational domain, since they
-allow an accurate representation of complex geometries. The
-solution is sought by applying suitable numerical operations, or
-kernels, to the entities of a mesh, such as edges, vertices, or
-cells. On standard clusters of multicores, typically, a kernel is
-executed sequentially by a thread, while parallelism is achieved by
-partitioning the mesh and assigning each partition to a different node
-or thread. Such an execution model, with minor variations, is adopted,
-for example, in \cite{pyop2isc}, \cite{Fenics},
-\cite{fluidity_manual_v4}, \cite{lizst}.
+finite volume method and the finite element method, are widely employed
+to approximate solutions of these PDEs. Unstructured meshes are often
+used to discretize the computational domain, since they allow an
+accurate representation of complex geometries. The solution is sought by
+applying suitable numerical operations, described by computational
+kernels, to all entities of a mesh, such as edges, vertices, or cells.
+On standard clusters of multicores, typically, a kernel is executed
+sequentially by a thread, while parallelism is achieved by partitioning
+the mesh and assigning each partition to a different node or thread.
+Such an execution model, with minor variations, is adopted, for example,
+by \citeN{pyop2isc}, \citeN{Fenics}, \citeN{fluidity_manual_v4},
+\citeN{lizst}.
 
@@
 of the PDE, possibly of the order of trillions, as in
-\cite{Rossinelli2013}. For example, it has been well established that mesh
+\citeN{Rossinelli2013}. For example, it has been well established that mesh
 resolution is critical in the accuracy of numerical weather
@@
 finite elements, which are the subject of this paper.
+
+% FIXME: citation for the 30-60% figure
 
@@
 method \cite{SPENCER} requires a significantly different loop nest structure.
-High order methods  therefore are excluded from our study.
+High order methods are therefore excluded from our study.
 
@@
 discretized domain. With such a kernel structure, we focus on aspects
-like minimization of floating-point operations, register allocation
+like the minimization of floating-point operations, register allocation
 and instruction-level parallelism, especially in the form of SIMD
@@
 
-Achieving high-performance is non-trivial. The complexity of the
+Achieving high performance is non-trivial. The complexity of the
 mathematical expressions, often characterized by a large number of
-operations on constants and small matrices, makes it hard to determine
-a single or specific sequence of transformations that is successfully
+operations on constants and small matrices, makes it hard to determine a
+single or specific sequence of transformations that is successfully
 applicable to all problems. Loop trip counts are typically small and can
@@
 obtain close-to-peak performance in a wide range of different local
-assembly kernels. Optimizations like padding, generalized
-loop-invariant code motion, vector-register tiling, and expression
-splitting, as well as their composition, are essential, but very limited is
-the support of state-of-the-art polyhedral and vendor compilers, if any. BLAS
-routines could theoretically be employed, although a fairly
-complicated control and data flow analysis would be required to
+assembly kernels. Optimizations like padding, generalized loop-invariant
+code motion, vector-register tiling, and expression splitting, as well
+as their composition, are essential, but their support in
+state-of-the-art polyhedral and vendor compilers, if any, is very
+limited. BLAS routines could theoretically be employed, although a
+fairly complicated control and data flow analysis would be required to
 automate identification and extraction of matrix-matrix multiplies. In
@@
 data locality, can result in performance improvements up to
-4.4$\times$ over original kernels, and around 70$\%$ of the test cases
+4.4$\times$ over original kernels. Around 70$\%$ of the test cases
 obtain a speed up greater than 2$\times$.
 
-Summarizing, the contributions of this paper are
+Summarizing, the contributions of this paper are:
 \begin{enumerate}
@@
  \item A full-application evaluation to demonstrate that the optimization
- strategy greatly impacts the whole finite element method.
+ strategy greatly impacts the whole finite element problem.
  \newcounter{enumTemp}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[scale=0.8]{Pictures/vect-by-vect-inline.pdf}}
+\centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect-inline.pdf}}
 \caption{Outer-product vectorization by permuting values in a vector
@@
 \begin{figure}[h]
-\centerline{\includegraphics[scale=0.8]{Pictures/vect-restore-inline.pdf}}
+\centerline{\includegraphics[scale=0.6]{Pictures/vect-restore-inline.pdf}}
 \caption{Restoring the storage layout after \emph{op-vect}. The figure
@@
 to minimize the aforementioned issues.
-
-\begin{table}[h]
-\tbl{Overview of code transformations for Firedrake-generated assembly kernels.}{
-\begin{tabulary}{1.0\columnwidth}{C|C}
-\hline
-Name (Abbreviation) & Parameter \\\hline
-Generalized loop-invariant code motion (\emph{licm}) &   \\
-Padding &  \\
-Data Alignment & \\
-Loop interchange      & loops  \\
-Loop unrolling  & unroll factor \\
-Outer-product vectorization (\emph{op-vect}) & tile size \\
-Expression splitting (\emph{split}) & split point, split factor \\ \hline
-\end{tabulary}
-}
-\label{table:code-transformations}
-\end{table}
-
-Table~\ref{table:code-transformations} summarizes the code
-transformations described so far. Given that many of these
-transformations depend on some parameters, we need a
-mechanism to prune such a large optimization space, which can be as large as order of thousands variants for medium-size loops and moderately complex mathematical expressions (all parametric transformations can indeed be combined).
-This aspect is treated in Section~\ref{sec:pyop2-compiler}.
+~\\
+
+Some of the transformations presented in this section, as well as other typical compiler optimizations applicable by COFFEE, like loop interchange and unrolling, are parametric; for instance, expression splitting and loop unrolling depend on the split and the unroll factors, respectively.
+Since all transformations are composable, the optimization space can be as large as order of thousands variants for local assembly kernels in which mathematical expressions are moderately complex and the iteration space is not particularly small, so a strategy to prune it is required.
+This problem is treated in the following section.
 
@@
 work.
+
+\subsubsection{On the compilation time}
+\label{sec:compile-time}
+Firedrake and its various software components (see Figure~\ref{fig:coffee-pipeline}) are implemented in Python and Cython, whereas the generated code is pure C.
+COFFEE has been written in Python to be naturally integrated within Firedrake.
+Among the various Firedrake modules, COFFEE is where, in general, the least amount of time is spent.
+The process of transforming the abstract syntax tree usually takes order of milliseconds.
+For instance, for the Helmholtz equation with $p=4$ on a tiny tetrahedral mesh, it took 0.0009s to transform the abstract syntax tree, while the whole assembly process needed 0.384s.
+Being a domain-specific compiler, COFFEE expects C code with the structure outlined in Figure~\ref{code:general-structure}, so typical compilers functionalities such as dependency tracking are simplified with respect to a general-purpose compiler, which implies a clear advantage in compilation time and implementation complexity.
 
@@
 Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake. Numerical
-integration is usually employed to implement the local assembly
+integration based on quadrature, as in Firedrake, is usually employed to implement the local assembly
 phase. The recent introduction of domain specific languages (DSLs) to
@@
 approaches. Methods based on tensor contraction~\cite{FFC-Compiler}
-and symbolic manipulation~\cite{Francis} have been implemented. We
-have designed COFFEE to specifically target numerical integration
-because it has been demonstrated that it remains the optimal choice
-for a wide class of problems~\cite{quadrature1}.
-
-Optimization of local assembly using numerical integration for CPU
-platforms has been addressed in FEniCS~\cite{quadrature1}. The
-comparison between COFFEE and this work is presented in
+and symbolic manipulation~\cite{Francis} have been implemented.
+Nevertheless, it has been demonstrated that quadrature-based integration remains the most efficient choice for a wide class of problems~\cite{quadrature1}, which motivates our work in COFFEE.
+
+Optimization of quadrature-based local assembly for CPU
+architectures has been addressed in FEniCS~\cite{quadrature1}. The
+comparison between COFFEE and this work has been presented in
 Section~\ref{sec:perf-results-fenics}. In~\cite{Markall20101815}, and
-more recently in~\cite{petsc-integration-gpu}, the problem has been
+more recently in~\cite{petsc-integration-gpu}, the same problem has been
 studied for GPU architectures. In~\cite{assembly-opencl}, variants of

### 2014-09-08 TACO-paper-v2/coffee.tex
Background: copy editing

@@
 Coordinates & double** & coords\\
-Fields & double** & w \\
+Fields (coefficients) & double** & w \\
 Numerical integration weights & double[] & W \\
@@
 of the method. This operation is applied to all cells in the discretized
-domain. In this work we focus on local matrices, or ``element
-matrices'', which are more costly to compute than element vectors.
+domain. In this work we focus on local matrices, or \emph{element
+matrices}, which are more costly to compute than element vectors.
 
@@
 Firedrake employs the FEniCS form compiler (FFC)~\cite{FFC-Compiler} to
-generate a C-code kernel implementing assembly using a
+generate an abstract syntax tree of a kernel implementing assembly using
 numerical quadrature. This kernel can be applied to any
@@
 finite element method. The evaluation of a local matrix can be
-reduced to integration on a fixed ``reference'' element --- a special
+reduced to integration on a fixed \emph{reference} element --- a special
 element that does not belong to the domain --- after a suitable change

### 2014-09-08 TACO-paper-v2/coffee.tex
Section 3: copy editing

@@
 different treatment, based on an in-depth analysis of both data and
-iteration spaces. Furthermore, domain-knowledge enables
+iteration spaces. Furthermore, domain knowledge enables
 transformations that a general-purpose compiler could not apply,
@@
 Data alignment is enforced in two steps. Firstly, all arrays are
-allocated to addresses that are multiples of $\mbox{\texttt{VL}}$. Then, two
-dimensional arrays are padded by rounding the number of columns to the
-nearest multiple of $\mbox{\texttt{VL}}$. For instance, assume the original size of a
-basis function array is 3$\times$3 and $\mbox{\texttt{VL}}=4$ (e.g. AVX processor,
-with 256-bit long vector registers and 64-bit double-precision
-floats). In this case, a padded version of the array will have size
-3$\times$4. The compiler is explicitly told about data alignment
-using suitable pragmas; for example, in the case of the Intel compiler, the annotation \texttt{$\#$pragma vector aligned} is added over a loop
-(as shown in later figures) to inform that all of the memory accesses in the loop body will be properly aligned.
-This allows the compiler to issue aligned load and store instructions, which are notably faster than unaligned ones.
+allocated to addresses that are multiples of $\mbox{\texttt{VL}}$. Then,
+two dimensional arrays are padded by rounding the number of columns to
+the nearest multiple of $\mbox{\texttt{VL}}$. For instance, assume the
+original size of a basis function array is 3$\times$3 and
+$\mbox{\texttt{VL}}=4$ (e.g. AVX processor, with 256-bit long vector
+registers and 64-bit double-precision floats). In this case, a padded
+version of the array will have size 3$\times$4. The compiler is
+explicitly told about data alignment using suitable pragmas; for
+example, in the case of the Intel compiler, the annotation
+\texttt{$\#$pragma vector aligned} is added before the loop (as shown in
+later figures) to inform that all of the memory accesses in the loop
+body will be properly aligned. This allows the compiler to issue aligned
+load and store instructions, which are notably faster than unaligned
+ones.
 
@@
 instructions, which could otherwise be significant. It is also worth
-noting that, in some problems (e.g. Helmholtz), invariant
+noting that, in some problems, for instance Helmholtz, invariant
 sub-expressions along \texttt{j} are identical to those along
@@
 ~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k+=4) $\lbrace$\\
-~~~~~~~~// $load$ and $set$ intrinsics \\
+~~~~~~~~// \emph{load} and \emph{set} intrinsics \\
 ~~~~~~~~// Compute A[0,0], A[1,1], A[2,2], A[3,3] \\
-~~~~~~~~// One $permute\_pd$ intrinsic per \texttt{k}-loop $load$\\
+~~~~~~~~// One \emph{permute\_pd} intrinsic per \texttt{k}-loop \emph{load}\\
 ~~~~~~~~// Compute A[0,1], A[1,0], A[2,3], A[3,2] \\
-~~~~~~~~// One $permute2f128\_pd$ intrinsic per \texttt{k}-loop $load$\\
+~~~~~~~~// One \emph{permute2f128\_pd} intrinsic per \texttt{k}-loop \emph{load}\\
 ~~~~~~~~// ...\\

### 2014-09-08 TACO-paper-v2/TACO-Rebuttal.txt
various additions

@@
 library-based approach, although it would be quite pointless since we can 
-generate the most suitable code for each problem.
+already generate the most suitable code for each problem.
 2) The variation in loop trip counts impacts loops transformation. If the loop 
@@
 section carefully, because it represents novel contribution of the paper and 
-aims to answer exactly the concern you pointed out.
+aims to answer exactly the concerns you pointed out.
 
@@
 pieces of code", but rather, in general, quite complex numerical routines 
-evaluating integrals in an
-element of the discretized equation domain, produced by means of real (used by 
-many scientists around the world) domain-specific language and compiler.  There 
-is an infinite variety of kernels that can be generated in Firedrake; the 
-generated code (e.g. loop sizes and mathematical expression) depends on the 
-partial differential equation being solved. This is explained and re-iterated 
-in the paper. Also, varying the polynomial order of the method or the element 
-shape (e.g. triangles, tetrahedrons, prisms), as we do and explain in the 
-performance evaluation section, changes the generated code by increasing loops 
-and arrays sizes and by making mathematical expressions more articulated (this 
-happens, for example, when moving from 2D to 3D domains). So we are already 
-testing many different code variants (three different kernels, and then for 
-each kernel we try multiple polynomial orders and element shapes). This is 
-carefully explained in Section 5.2.1. Now, premised this, back to your point, 
-which is "applicability and effectiveness of our transformations in other local 
-assembly kernels". Our answer is twofold:
+evaluating integrals in an element of the discretized equation domain, produced 
+by means of real (used by many scientists around the world) domain-specific 
+language and compiler.  There is an infinite variety of kernels that can be 
+generated in Firedrake; the generated code (e.g. loop sizes and mathematical 
+expression) depends on the partial differential equation being solved. This is 
+explained and re-iterated in the paper. Also, varying the polynomial order of 
+the method or the element shape (e.g. triangles, tetrahedrons, prisms), as we 
+do and explain in the performance evaluation section, changes the generated 
+code by increasing loops and arrays sizes and by making mathematical 
+expressions more articulated (this happens, for example, when moving from 2D to 
+3D domains). So we are already testing many different code variants (three 
+different kernels, and then for each kernel we try multiple polynomial orders 
+and element shapes). This is carefully explained in Section 5.2.1. Premised 
+this, we now go back to your point about "applicability and effectiveness of 
+our transformations in other local assembly kernels". Our answer is twofold:
 
 - Section 5.2.1 explains in-depth the motivations behind the choice of the 
-  three problems (Helmholtz, Diffusion, Burgers), explaining that "(...) 
+  three benchmarks (Helmholtz, Diffusion, Burgers), explaining that "(...) 
 distinct problems, possibly arising in completely different fields, may employ 
@@
 later on in the section and in the paper. Other kernels arising from completely 
-different equations will still have the same structure of the kernels we 
-examined (e.g. same memory access pattern, same ranges of loop sizes, same 
-problems concerning the presence of loop-invariant code and register pressure) 
-as well as similar patterns in the mathematical expressions, so there is no 
-evident reason for which our code optimizations would not be effective. The 
-critical issue in our research was to understand what transformations our local 
-assembly kernels required, and how to instantiate parametric transformations 
+different equations will still have the same structure of the ones we examined 
+(e.g. same memory access pattern, same ranges of loop sizes, same problems 
+concerning the presence of loop-invariant code and register pressure) as well 
+as similar patterns in the mathematical expressions, so there is no evident 
+reason for which our code optimizations would not be effective. The critical 
+issue in our research was to understand what transformations our local assembly 
+kernels required, and how to instantiate parametric transformations 
 (vector-register tiling, loop interchange, expression splitting) optimally in 
@@
 
-- In Section 5.3 we provide a full-application study in which we analyze a 
-  different equation (Linear Elasticity), which brings the total number of 
-studied equations to 4 (Helmholtz, Diffusion, Burgers, Elasticity)
+- In Section 5.3 we provide a full-application study in which we analyze a new 
+  equation (Linear Elasticity), which brings the total number of studied 
+equations to 4 (Helmholtz, Diffusion, Burgers, Elasticity)
 Adding more kernels to the performance evaluation section would make the paper 
-only more chaotic: we selected these three problems to highlight all pros and 
-cons of our code transformations. This is the reason for which we comment on 
-the performance of the code transformations (and their interplay) in specific 
-sub-sections, referring to, describing, and carefully analyzing the provided 
-plots.
+only more chaotic: we specifically selected these three problems because they 
+allow us to highlight all pros and cons of our code transformations. This is 
+also the reason we comment on the performance of the code transformations (and 
+their interplay) in specific sub-sections (referring to, describing, and 
+carefully analyzing the provided plots), rather than in a single, centralized 
+section.
 
@@
 assembly kernels on GPUs is represented by memory requirements. We give here 
-two examples of this problem, although more issues could be found. First, the 
-transfer of the mesh from host to GPU may be a notable obstacle: if too big, 
-the mesh
-will not fit the available GPU memory, and distributed execution, which is 
-marginally supported in available finite element frameworks, would be required.  
-Secondly, optimizations such as generalized loop-invariant code motion increase 
-the memory requirements due to the introduction of temporary arrays.  This may 
-actually lead to poor per-thread shared memory and registers allocation. These 
-aspects suggest that the problem of executing generic, automatically-generated 
-local assembly kernels on GPUs require a separate, in-depth study. However, in 
-the new Section 6, we provide some intuitions of how we could re-adapt COFFEE's 
-transformations for execution on GPUs.  Therefore, we invite once more to go 
-through the new Section 6 carefully.
+two examples of this problem, although more issues could be found and 
+discussed. Firstly, the transfer of the mesh from host to device may be a great 
+obstacle: if too big, the mesh will not fit the available GPU memory, and 
+distributed execution, which is marginally supported in available finite 
+element frameworks, would be required. Secondly, optimizations such as 
+generalized loop-invariant code motion increase the memory requirements due to 
+the introduction of temporary arrays.  This may actually lead to poor 
+per-thread shared memory and registers allocation. These aspects suggest that 
+the problem of executing generic, automatically-generated local assembly 
+kernels on GPUs require a separate, in-depth study. However, in the new Section 
+6, we provide some intuitions of how we could re-adapt COFFEE's transformations 
+for execution on GPUs.  Therefore, we invite once more to go through the new 
+Section 6 carefully.
 
@@
 kernels on GPUs is represented by memory requirements. We give here two 
-examples of this problem, although many more issues could be found. First, the 
+examples of this problem, although many more issues could be found. Firstly, the 
 transfer of the mesh from host to device may be a great obstacle: if too big, 
-the mesh
-will not fit the available GPU memory, and distributed execution, which is 
-marginally supported in available finite element frameworks, would be required.  
-Secondly, optimizations such as generalized loop-invariant code motion increase 
-the memory requirements due to the introduction of temporary arrays.  This may 
-actually lead to poor per-thread shared memory and registers allocation. These 
-aspects suggest that the problem of executing generic, automatically-generated 
-local assembly kernels on GPUs require a separate, in-depth study. However, in 
-the new Section 6 of the revised paper, we provide some intuitions of how we 
-could re-adapt COFFEE's transformations for execution on GPUs.
+the mesh will not fit the available GPU memory, and distributed execution, 
+which is marginally supported in available finite element frameworks, would be 
+required.  Secondly, optimizations such as generalized loop-invariant code 
+motion increase the memory requirements due to the introduction of temporary 
+arrays.  This may actually lead to poor per-thread shared memory and registers 
+allocation. These aspects suggest that the problem of executing generic, 
+automatically-generated local assembly kernels on GPUs require a separate, 
+in-depth study. However, in the new Section 6 of the revised paper, we provide 
+some intuitions of how we could re-adapt COFFEE's transformations for execution 
+on GPUs.
 
@@
   have tested several split factors, which means testing as many code variants 
-(i.e. as many as the ways an expression can be split based on sum's associativity) 
-and, as shown by the various graphs in the paper, our cost model seems to be 
-capable of properly approximating such a parameter. This is of course 
-encouraging.
+(i.e. as many as the ways an expression can be split based on sum's 
+associativity) and, as shown by the various graphs in the paper, our cost model 
+seems to be capable of properly approximating such a parameter. This is of 
+course encouraging.
 - In the perspective of integrating expression splitting in a general-purpose 
@@
 on "how to optimally and rapidly split a mathematical expression". Rather, we 
-prefer to show that, for our problems, there is clearly a problem of register pressure 
-and that by rescheduling the computation in different loops (and 
+prefer to show that, for our problems, there is clearly a problem of register 
+pressure and that by rescheduling the computation in different loops (and 
 sub-expressions) we can actually obtain a performance gain. We have a strategy 
@@
 least we demonstrate it is effective and easily predictable. 
+We spent some time thinking about the problem: two things that we may want to 
+try in the future are 1) a model based on graph partitioning and 2) a model 
+based on a set of rewrite rules that exploit arithmetic properties such as 
+commutativity, associativity, and distributivity/factorization to 
+systematically optimize the arithmetic intensity of the expression.
 

### 2014-09-08 TACO-paper-v2/coffee.tex
various additions

--- 2014-09-08 TACO-paper-v2/coffee.tex
@@
 partial differential equations (PDEs). Numerical techniques, like the
-finite volume method and the finite element method, are widely employed
-to approximate solutions of these PDEs. Unstructured meshes are often
-used to discretize the computational domain, since they allow an
-accurate representation of complex geometries. The solution is sought by
-applying suitable numerical operations, described by computational
-kernels, to all entities of a mesh, such as edges, vertices, or cells.
-On standard clusters of multicores, typically, a kernel is executed
-sequentially by a thread, while parallelism is achieved by partitioning
-the mesh and assigning each partition to a different node or thread.
-Such an execution model, with minor variations, is adopted, for example,
-by \citeN{pyop2isc}, \citeN{Fenics}, \citeN{fluidity_manual_v4},
-\citeN{lizst}.
+finite volume method and the finite element method, are widely
+employed to approximate solutions of these PDEs. Unstructured meshes
+are often used to discretize the computational domain, since they
+allow an accurate representation of complex geometries. The
+solution is sought by applying suitable numerical operations, or
+kernels, to the entities of a mesh, such as edges, vertices, or
+cells. On standard clusters of multicores, typically, a kernel is
+executed sequentially by a thread, while parallelism is achieved by
+partitioning the mesh and assigning each partition to a different node
+or thread. Such an execution model, with minor variations, is adopted,
+for example, in \cite{pyop2isc}, \cite{Fenics},
+\cite{fluidity_manual_v4}, \cite{lizst}.
 
@@
 of the PDE, possibly of the order of trillions, as in
-\citeN{Rossinelli2013}. For example, it has been well established that mesh
+\cite{Rossinelli2013}. For example, it has been well established that mesh
 resolution is critical in the accuracy of numerical weather
@@
 finite elements, which are the subject of this paper.
-
-% FIXME: citation for the 30-60% figure
 
@@
 method \cite{SPENCER} requires a significantly different loop nest structure.
-High order methods are therefore excluded from our study.
+High order methods  therefore are excluded from our study.
 
@@
 discretized domain. With such a kernel structure, we focus on aspects
-like the minimization of floating-point operations, register allocation
+like minimization of floating-point operations, register allocation
 and instruction-level parallelism, especially in the form of SIMD
@@
 
-Achieving high performance is non-trivial. The complexity of the
+Achieving high-performance is non-trivial. The complexity of the
 mathematical expressions, often characterized by a large number of
-operations on constants and small matrices, makes it hard to determine a
-single or specific sequence of transformations that is successfully
+operations on constants and small matrices, makes it hard to determine
+a single or specific sequence of transformations that is successfully
 applicable to all problems. Loop trip counts are typically small and can
@@
 obtain close-to-peak performance in a wide range of different local
-assembly kernels. Optimizations like padding, generalized loop-invariant
-code motion, vector-register tiling, and expression splitting, as well
-as their composition, are essential, but their support in
-state-of-the-art polyhedral and vendor compilers, if any, is very
-limited. BLAS routines could theoretically be employed, although a
-fairly complicated control and data flow analysis would be required to
+assembly kernels. Optimizations like padding, generalized
+loop-invariant code motion, vector-register tiling, and expression
+splitting, as well as their composition, are essential, but very limited is
+the support of state-of-the-art polyhedral and vendor compilers, if any. BLAS
+routines could theoretically be employed, although a fairly
+complicated control and data flow analysis would be required to
 automate identification and extraction of matrix-matrix multiplies. In
@@
 data locality, can result in performance improvements up to
-4.4$\times$ over original kernels. Around 70$\%$ of the test cases
+4.4$\times$ over original kernels, and around 70$\%$ of the test cases
 obtain a speed up greater than 2$\times$.
 
-Summarizing, the contributions of this paper are:
+Summarizing, the contributions of this paper are
 \begin{enumerate}
@@
  \item A full-application evaluation to demonstrate that the optimization
- strategy greatly impacts the whole finite element problem.
+ strategy greatly impacts the whole finite element method.
  \newcounter{enumTemp}
@@
 \begin{figure}[h]
-\centerline{\includegraphics[scale=0.6]{Pictures/vect-by-vect-inline.pdf}}
+\centerline{\includegraphics[scale=0.8]{Pictures/vect-by-vect-inline.pdf}}
 \caption{Outer-product vectorization by permuting values in a vector
@@
 \begin{figure}[h]
-\centerline{\includegraphics[scale=0.6]{Pictures/vect-restore-inline.pdf}}
+\centerline{\includegraphics[scale=0.8]{Pictures/vect-restore-inline.pdf}}
 \caption{Restoring the storage layout after \emph{op-vect}. The figure
@@
 to minimize the aforementioned issues.
-~\\
-
-Some of the transformations presented in this section, as well as other typical compiler optimizations applicable by COFFEE, like loop interchange and unrolling, are parametric; for instance, expression splitting and loop unrolling depend on the split and the unroll factors, respectively.
-Since all transformations are composable, the optimization space can be as large as order of thousands variants for local assembly kernels in which mathematical expressions are moderately complex and the iteration space is not particularly small, so a strategy to prune it is required.
-This problem is treated in the following section.
+
+\begin{table}[h]
+\tbl{Overview of code transformations for Firedrake-generated assembly kernels.}{
+\begin{tabulary}{1.0\columnwidth}{C|C}
+\hline
+Name (Abbreviation) & Parameter \\\hline
+Generalized loop-invariant code motion (\emph{licm}) &   \\
+Padding &  \\
+Data Alignment & \\
+Loop interchange      & loops  \\
+Loop unrolling  & unroll factor \\
+Outer-product vectorization (\emph{op-vect}) & tile size \\
+Expression splitting (\emph{split}) & split point, split factor \\ \hline
+\end{tabulary}
+}
+\label{table:code-transformations}
+\end{table}
+
+Table~\ref{table:code-transformations} summarizes the code
+transformations described so far. Given that many of these
+transformations depend on some parameters, we need a
+mechanism to prune such a large optimization space, which can be as large as order of thousands variants for medium-size loops and moderately complex mathematical expressions (all parametric transformations can indeed be combined).
+This aspect is treated in Section~\ref{sec:pyop2-compiler}.
 
@@
 Among the various Firedrake modules, COFFEE is where, in general, the least amount of time is spent.
-The process of transforming the abstract syntax tree usually takes order of milliseconds.
-For instance, for the Helmholtz equation with $p=4$ on a tiny tetrahedral mesh, it took 0.0009s to transform the abstract syntax tree, while the whole assembly process needed 0.384s.
-Being a domain-specific compiler, COFFEE expects C code with the structure outlined in Figure~\ref{code:general-structure}, so typical compilers functionalities such as dependency tracking are simplified with respect to a general-purpose compiler, which implies a clear advantage in compilation time and implementation complexity.
+The process of transforming the abstract syntax tree never exceeds ...
+For example, in X, it takes .., while FFC needed Y secs to translate UFL code into the abstract syntax tree.
+Being a domain-specific compiler, COFFEE expects C code with a certain structure (e.g.
+it know there is no control flow), so data-flow analysis for dependency tracking can be kept to a minimum.
+This is also a clear advantage over a general-purpose compiler.
 

### 2014-09-09 TACO-paper-v2/coffee.tex
Add paragraph on compilation time

@@
 Among the various Firedrake modules, COFFEE is where, in general, the least amount of time is spent.
-The process of transforming the abstract syntax tree never exceeds ...
-For example, in X, it takes .., while FFC needed Y secs to translate UFL code into the abstract syntax tree.
-Being a domain-specific compiler, COFFEE expects C code with a certain structure (e.g.
-it know there is no control flow), so data-flow analysis for dependency tracking can be kept to a minimum.
-This is also a clear advantage over a general-purpose compiler.
+The process of transforming the abstract syntax tree usually takes order of milliseconds.
+For instance, for the Helmholtz equation with $p=4$ on a tiny tetrahedral mesh, it took 0.0009s to transform the abstract syntax tree, while the whole assembly process needed 0.384s.
+Being a domain-specific compiler, COFFEE expects C code with the structure outlined in Figure~\ref{code:general-structure}, so typical compilers functionalities such as dependency tracking are simplified with respect to a general-purpose compiler, which implies a clear advantage in compilation time and implementation complexity.
 

### 2014-09-09 TACO-paper-v2/coffee.tex
Section 4: copy editing

@@
 and \emph{op-vect} may be introduced. Their implementation is centered
-on analysis and manipulation of the kernel's AST. When \emph{op-vect}
+on analysis and manipulation of the kernel AST. When \emph{op-vect}
 is selected, the compiler outputs AVX or AVX-512 intrinsics code. Any
@@
 the backend compiler in charge of selecting unroll factors. This
-also simplifies the COFFEE's cost model, described next. The only situation in
+also simplifies COFFEE's cost model, described next. The only situation in
 which we explicitly unroll-and-jam a loop is when \emph{op-vect} is

### 2014-09-09 TACO-paper-v2/coffee.tex
UTF8 input encoding
Section 5: copy editing

@@
 
+\usepackage[utf8]{inputenc}
 \usepackage{graphicx}
@@
 
-We present two studies. Firstly, in Section~\ref{sec:analysis-code-transf}, we
-analyze the impact of the proposed code transformations, and their interplay,
-in three real-world representative equations. We quantify the speed ups achievable at
-the level of the local assembly kernel, i.e. by excluding from the measurements
-the other sections of the computation. Then, in Section~\ref{sec:full-app}, we
-provide a full-application investigation. Here, we use another equation (so as to
-bring the total number of examined problems to four) to demonstrate that COFFEE's
-optimizations actually allow a significant reduction in the overall computation run-time
-to be achieved, which practically motivates our research.
+We present two studies. Firstly, in
+Section~\ref{sec:analysis-code-transf}, we analyze the impact of the
+proposed code transformations, and their interplay, in three real-world
+representative equations. We quantify the speedups achievable at the
+level of the local assembly kernel, by excluding the other stages of the
+computation from the measurements. Then, in Section~\ref{sec:full-app},
+we provide a full-application investigation. Here, we use another
+equation, bringing the total number of examined problems to four,
+to demonstrate that COFFEE's optimizations actually allow a significant
+reduction in the overall computation run-time to be achieved, which
+practically motivates our research.
 
@@
 Our code transformations were evaluated in three real-world problems
-based on the following PDEs
+based on the following PDEs:
 \begin{enumerate}
@@
 operators. They are complete and unsimplified examples of the
-operators occurring for diffusion and viscosity in fluids, and for
+operators used to model diffusion and viscosity in fluids, and for
 imposing pressure in compressible fluids. As such, they are both
@@
 apply to scalar quantities; this impacts the generated code, as
-explained next. In this way, the operators we have selected are
-characteristic of both the second and first order operators that
-dominate in fluids and solids simulations, and a wide variety on local
-assembly codes can be tested.
+explained next. The operators we have selected are characteristic of
+both the second and first order operators that dominate fluids and
+solids simulations.
+%and a wide variety on local assembly codes can be tested.
 
@@
 respect to Helmholtz are the absence of the $Y$ array and the presence
-of a few more constants for computing the element matrix. Burgers is a
+of additional constants for computing the element matrix. Burgers is a
 non-linear problem employing differential operators different from
@@
 element shape and $p$, the larger the iteration space. Triangles,
-tetrahedron, and prisms were tested as element shape. For instance, in
+tetrahedra, and prisms were tested as element shape. For instance, in
 the case of Helmholtz with $p=1$, the size of the \texttt{j} and
@@
 number of basis function arrays, since, intuitively, the behaviour of
-the equation has now to be approximated also along a third axis. On
+the equation has to be approximated also along a third axis. On
 the other hand, the polynomial order affects only the problem size
@@
 Section~\ref{sec:pyop2-compiler}, loop interchanges having an inner
-loop along \texttt{i} caused slow downs; also, interchanging
+loop along \texttt{i} caused slowdowns; also, interchanging
 \texttt{j} and \texttt{k} loops while keeping \texttt{i} as outermost
@@
 size of the innermost loop is not a multiple of the vector length, and
-a reminder scalar loop is introduced at compile time. Since the loop
+a remainder scalar loop is introduced at compile time. Since the loop
 nest is small, the cost of executing the extra scalar iterations can
-have a significant impact. The reminder loop overhead is more
-pronounced on the Xeon Phi, where the vector length is twice as big,
-which leads to proportionally larger scalar reminder loops.
+have a significant impact. The remainder loop overhead is more
+pronounced on the Xeon Phi, where the vector length is twice as long,
+which leads to proportionally larger scalar remainder loops.
 
@@
 the following, this version of the code is referred to as
-\emph{licm-ap}. Padding, which avoids the introduction of a reminder
+\emph{licm-ap}. Padding, which avoids the introduction of a remainder
 loop as described in Section~\ref{licm-impact}, as well as data
@@
 Application of \emph{op-vect} to the Burgers problem induces
-significant slow downs due to the large number of temporary arrays
+significant slowdowns due to the large number of temporary arrays
 that need to be tiled, which exceeds the available logical registers
@@
 \subsubsection{Impact of Expression Splitting}
-\label{sec:perf-results-split}
-Expression splitting relieves the register pressure when the element
-matrix evaluation needs to read from a large number of basis function
-arrays. As detailed in Section~\ref{sec:expr-split}, in the price to
-pay for this optimazion, there are an increased number of accesses to
-the element matrix and, potentially, redundant computation. Similarly
-to the analysis of vector-register tiling, we compare two versions:
-the baseline, \emph{licm-ap}; and expression splitting on top of
-\emph{licm-ap}, which, for simplicity, in the following is referred to
-as \emph{split}.
+\label{sec:perf-results-split} Expression splitting relieves the
+register pressure when the element matrix evaluation needs to read from
+a large number of basis function arrays. As detailed in
+Section~\ref{sec:expr-split}, the price to pay for this optimazion is an
+increased number of accesses to the element matrix and, potentially,
+redundant computation. Similarly to the analysis of vector-register
+tiling, we compare two versions: the baseline, \emph{licm-ap}, and
+expression splitting on top of \emph{licm-ap}, which, for simplicity, in
+the following is referred to as \emph{split}.
 
@@
 that the cost model prevents the adoption of the transformation: the
-\texttt{while} statement in Figure~\ref{algo:applyCostModel} is indeed
-never entered.
+\texttt{while} statement in Figure~\ref{algo:applyCostModel} is never
+entered.
 
@@
 fewer zero-valued contributions.
+
+% FIXME: fastest quadrature or does that include tensor representation?
 
@@
 
-In these experiments, we observe a maximum performance improvement of 1.47$\times$ over the non-optimized local assembly code, obtained in the case ($f=2, p=2$).
-However, we reiterate the fact that full-application speed ups rise proportionally with the amount of time spent in assembly and, therefore, with the complexity of the equation.
-By increasing polynomial order and number of coefficient functions, or by simply studying a different, more complex equation, it is our experience that performance gains become increasingly more relevant.
-The choice of studying the static linear elasticity equation was to show that even relatively simple problems can be characterized by a large proportion of execution time spent in assembly.
+In these experiments, we observe a maximum performance improvement of
+1.47$\times$ over the non-optimized local assembly code, obtained in the
+case ($f=2, p=2$). However, we reiterate the fact that full-application
+speed ups increase proportionally with the amount of time spent in
+assembly and, therefore, with the complexity of the equation. By
+increasing polynomial order and number of coefficient functions, or by
+simply studying a different, more complex equation, it is our experience
+that performance gains become increasingly more relevant. The choice of
+studying the static linear elasticity equation was to show that even
+relatively simple problems can be characterized by a large proportion of
+execution time spent in assembly.
 

### 2014-09-09 TACO-paper-v2/coffee.tex
Section 6: copy editing

@@
 
-We observe that our choice was to develop COFFEE as a separate, self-contained software module, with clear input/output interfaces, rather then incorporating it within PyOP2.
-This was motivated by two critical aspects that characterize the generality of our research.
-
-\paragraph{Separation of concerns} We believe that in domain-specific frameworks there must be a clear, logical separation of roles reflecting the various levels of abstractions, with domain-specialists that can completely abstract from performance optimization.
-In Firedrake, for instance, COFFEE decouples the mathematical specification of a finite element method, captured by the Unified Form Language and the FEniCS Form Compiler, from code optimization.
-This is of fundamental importance to maximize productivity by allowing scientists to focus only on their area of expertise.
-Practically speaking, from the perspective of the domain-specific language and compiler designers, our optimization strategy represents an incentive to produce extremely simple representations of the code (e.g.
-fully-inlined mathematical expressions in the form of an abstract syntax tree, in the case of Firedrake) so as to make the architecture-aware code optimizer completely responsible for choosing and applying the most suitable set of transformations.
-
-\paragraph{Generalizability to other domains} There are neither conceptual
-nor technical reasons which prevent our compiler being used in applications
-other than Firedrake. For example, integration with the popular FEniCS
-framework, the pioneer of automated code generation for finite element local
-assembly, would be relatively easy to achieve. It is more challenging to
-assess the generality of the optimization strategy: the extent to which COFFEE and its transformations are transferable to other computational domains, perhaps other DSLs, and to what extent this would be helpful for improving full-application performance.
-To answer these questions, we first need to go back to the origins of our compiler.
-The starting point of our work was the mathematical formulation of a local assembly operation, expressible as follows
+COFFEE was developed as a separate, self-contained software module, with
+clear input/output interfaces, rather then incorporating it within
+PyOP2. This choice was motivated by two critical aspects that
+characterize the generality of our research.
+
+\paragraph{Separation of concerns} We believe that in domain-specific
+frameworks there must be a clear, logical separation of roles reflecting
+the various levels of abstraction, where domain specialists are
+completely separated from performance optimization. In Firedrake, for
+instance, COFFEE decouples the mathematical specification of a finite
+element method, captured by the Unified Form Language and the FEniCS
+Form Compiler, from code optimization. This is of fundamental importance
+to maximize productivity by allowing scientists to focus only on their
+area of expertise. Practically speaking, from the perspective of the
+domain-specific language and compiler designers, our optimization
+strategy represents an incentive to produce extremely simple
+representations of the code (e.g. fully-inlined mathematical expressions
+in the form of an abstract syntax tree, in the case of Firedrake) so as
+to make the architecture-aware code optimizer completely responsible for
+choosing and applying the most suitable set of transformations.
+
+\paragraph{Generalizability to other domains} There are neither
+conceptual nor technical reasons which prevent our compiler from being
+used in applications other than Firedrake. For example, integration with
+the popular FEniCS framework, the pioneer of automated code generation
+for finite element local assembly, would be relatively easy to achieve.
+It is more challenging to assess the generality of the optimization
+strategy: the extent to which COFFEE and its transformations are
+transferable to other computational domains, perhaps other DSLs, and to
+what extent this would be helpful for improving full-application
+performance. To answer these questions, we first need to go back to the
+origins of our compiler. The starting point of our work was the
+mathematical formulation of a local assembly operation, expressible as
+follows
 
@@
 \begin{enumerate}
-\item In Equation~\ref{eq:assembly-model}, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel.
-Throughout the paper we assumed a kernel be executed by a single thread, which is likely to be the best strategy for standard multi-core CPUs.
+\item In Equation~\eqref{eq:assembly-model}, the summations correspond
+  to reduction loops, whereas loops over indices $i$ and $j$ are fully
+  parallel.
+Throughout the paper we assumed a kernel to be executed by a single thread, which is likely to be the best strategy for standard multi-core CPUs.
 On the other hand, we note that for certain architectures (e.g.
@@
   execution. In the context of COFFEE, expression splitting has been
-  successfully applied to improve register pressure. However, underlying
+  successfully applied to improve register pressure. However, the underlying
   idea of re-scheduling (re-associating) operations to optimize for some
@@
   requirements (a variant of this idea for non-affine loops in unstructured
-  mesh applications has been adopted in~\cite{carlolcpc}). In
-  Equation~\ref{eq:assembly-model}, for instance, not only can  any of the
+  mesh applications has been adopted in~\citeN{carlolcpc}). In
+  Equation~\eqref{eq:assembly-model}, for instance, not only can  any of the
   functions $\alpha$, $\beta$ and $\gamma$ be split (assuming they include
@@
   opportunities (see Section~\ref{sec:code-transf-licm}). Padding and data
-  alignment are special cases, since they could be performed at run-time if some values were not known at code generation time.
-Theoretically, they could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.
+  alignment are special cases, since they could be performed at run-time
+  if some values were not known at code generation time. Theoretically,
+  they could also be automated by a general-purpose compiler through profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.
 \item \textbf{Multi-loop vectorization.} Compiler auto-vectorization has

### 2014-09-09 TACO-paper-v2/coffee.tex
Section 7: copy editing

@@
 comparison between COFFEE and this work has been presented in
-Section~\ref{sec:perf-results-fenics}. In~\cite{Markall20101815}, and
-more recently in~\cite{petsc-integration-gpu}, the same problem has been
-studied for GPU architectures. In~\cite{assembly-opencl}, variants of
+Section~\ref{sec:perf-results-fenics}. In~\citeN{Markall20101815}, and
+more recently in~\citeN{petsc-integration-gpu}, the same problem has been
+studied for GPU architectures. In~\citeN{assembly-opencl}, variants of
 the standard numerical integration algorithm have been specialized and
@@
 Spiral~\cite{Pueschel:05} for digital signal processing numerical
-algorithms, ~\cite{Spampinato:14} for dense linear algebra, or
+algorithms, ~\citeN{Spampinato:14} for dense linear algebra, or
 Pochoir~\cite{pochoir} and SDSL~\cite{stencil-compiler} for image

### 2014-09-09 TACO-paper-v2/coffee.tex
Add paragrah on compilation techniques in related work section

@@
 \tbl{Type and variable names used in the various listings to identify local assembly objects.}{
+\scriptsize
 \begin{tabulary}{1.0\columnwidth}{C|C|C}
@@
 
+The code transformations presented are inspired by standard compilers optimizations and exploit domain properties.
+Our loop-invariant code motion technique individuates invariant sub-expressions and redundant computation by analyzing all loops in an iteration space, which is a generalization of the algorithms often implemented by general-purpose compilers.
+Expression splitting is an abstract variant of loop fission based on properties of arithmetic operators.
+The outer-product vectorization is an implementation of tiling at the level of vector registers; tiling, or ``loop blocking'', is commonly used to improve data locality (especially for caches).
+Padding has been used to achieve data alignment and to improve the effectiveness of vectorization.
+A standard reference for the compilation techniques re-adapted in this work is~\citeN{dragonbook}.
+
 Our compiler-based optimization approach is made possible by the

### 2014-09-09 TACO-paper-v2/coffee.tex
2-column layout for listing 4
minor fix

@@
 \IncMargin{-\parindent}
+\usepackage{multicol}
 
@@
 element methods, in which an assembly kernel's working set is usually small
-enough to fit the L1 cache. Low order methods are by no means exotic: they
+enough to fit the L1 cache of traditional CPU architectures. Low order methods are by no means exotic: they
 are employed in a wide variety of fields, including climate and ocean
@@
 
+\begin{multicols}{2}
 \KwSty{void} helmholtz(\KwSty{double} A[4][4], \KwSty{double} **coords) $\lbrace$\\
@@
 $\rbrace$
+\end{multicols}
 \caption{Local assembly code generated by Firedrake for the Helmholtz

### 2014-09-10 TACO-paper-v2/coffee.tex
Incorporate gpu paragraph. Several other fixes

@@
 language by introducing domain-aware optimizations and, eventually,
-produces C code including vector SIMD intrinsics. Experiments using a range of finite-element
+produces C code including vector SIMD intrinsics. Experiments using a range of real-world finite-element
 problems of increasing complexity show that significant performance improvement
@@
 
+We aim to maximize our impact on the platforms that are realistically used
+for finite element applications, so we target conventional CPU architectures. 
+The key limiting factor to the execution on GPUs is the stringent memory requirements. 
+Only relatively small meshes can fit a GPU memory, and distributed execution is minimally
+supported by available frameworks (if any). There has been some research on
+adapting local assembly to GPUs (which we mention in Section~\ref{sec:related-work}),
+although it differs from ours in several ways, including: (i) not relying on
+automated code generation from a domain-specific language (explained next),
+(ii) testing only very low order methods, (iii) not optimizing for
+cross-loop arithmetic intensity (the goal is rather effective multi-thread
+parallelization). In addition, our code transformations would drastically impact the 
+GPU parallelization strategy, for example by increasing a thread's working set. 
+For all these reasons, a study on extending our work to GPU architectures is 
+beyond the scope of this paper. In Section~\ref{sec:generality}, however, we provide 
+some intuitions about this research direction.
+
 An assembly kernel is characterized by the presence of an affine,
@@
 
-Achieving high performance is non-trivial. The complexity of the
+Achieving high performance on CPUs is non-trivial. The complexity of the
 mathematical expressions, often characterized by a large number of
@@
 single or specific sequence of transformations that is successfully
-applicable to all problems. Loop trip counts are typically small and can
+applicable to different problems (equations). Loop trip counts are typically small and can
 vary significantly, which further exacerbates the issue. A
@@
 
-\begin{figure}
+\begin{figure}[b]
 \begin{alltt}
@@
 \begin{equation}
+\scriptsize
 \label{outer-product}
@@
 \begin{center}
-\includegraphics[scale=0.75]{Pictures/coffee-pipeline.pdf}
+\includegraphics[scale=0.70]{Pictures/coffee-pipeline.pdf}
 \caption{High-level view of Firedrake. COFFEE is at the core,
@@
 Our code transformations were evaluated in three real-world problems
-based on the following PDEs:
-\begin{enumerate}
-\item Helmholtz
-\item Advection-Diffusion
-\item Burgers
-\end{enumerate}
+based on the following PDEs: (i) Helmholtz, (ii) Advection-Diffusion, 
+and (iii) Burgers.
+
 The three chosen benchmarks are \emph{real-life kernels} and comprise
@@
 \tbl{Performance improvement due to generalized loop-invariant code
-  motion, for different element shapes (triangle, tetrahedron, prism)
-  and polynomial orders ($p \in [1, 4]$), over the original
-  non-optimized code, for the Helmholtz, Diffusion and Burgers
-  problems.}{ \scriptsize \licmresultsnorms }
+  motion over the original non-optimized code.}{ \scriptsize \licmresultsnorms }
 \label{table:perf-results-licm}
@@
 \tbl{Performance improvement due to generalized loop-invariant code
-  motion, data alignment, and padding, for different element shapes
-  (triangle, tetrahedron, prism) and polynomial orders ($p \in [1,
-    4]$), over the original non-optimized code, for the Helmholtz,
-  Diffusion and Burgers problems.}{ \scriptsize \licmapresultsnorms }
+  motion, data alignment, and padding over the original non-optimized code.}{ \scriptsize \licmapresultsnorms }
 \label{table:perf-results-licmap}
@@
 	\centering
-	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.52]{Pictures/helmholtz-normalized-opvect.pdf}}
-	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/helmholtz-normalized-opvect-phi.pdf}}
+	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.48]{Pictures/helmholtz-normalized-opvect.pdf}}
+	\subfloat[Xeon Phi]{	\includegraphics[scale=0.48]{Pictures/helmholtz-normalized-opvect-phi.pdf}}
 	\caption{Performance improvement over \emph{licm-ap} obtained
@@
 	\centering
-	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.52]{Pictures/diffusion-normalized-opvect.pdf}}
-	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/diffusion-normalized-opvect-phi.pdf}}
+	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.48]{Pictures/diffusion-normalized-opvect.pdf}}
+	\subfloat[Xeon Phi]{	\includegraphics[scale=0.48]{Pictures/diffusion-normalized-opvect-phi.pdf}}
 	\caption{Performance improvement over \emph{licm-ap} obtained
@@
 	\centering
-	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.52]{Pictures/burgers-normalized-split.pdf}}
-	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/burgers-normalized-split-phi.pdf}}
+	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.48]{Pictures/burgers-normalized-split.pdf}}
+	\subfloat[Xeon Phi]{	\includegraphics[scale=0.48]{Pictures/burgers-normalized-split-phi.pdf}}
 	\caption{Performance improvement over \emph{licm-ap} obtained
@@
 \begin{table}[h]
-\tbl{Performance comparison between FEniCS and COFFEE on the Sandy Bridge.}{
+\tbl{Performance comparison between FEniCS and COFFEE on Sandy Bridge.}{
+\scriptsize
 \begin{tabulary}{1.0\columnwidth}{C|C|C}
@@
 	\centering
-	\subfloat[]{\includegraphics[scale=0.52]{Pictures/elasticity-p1.pdf}}
-	\subfloat[]{\includegraphics[scale=0.52]{Pictures/elasticity-p2.pdf}}
-	\caption{Performance improvement over non-optimized code for the static linear elasticity equation.
-The tetrahedral mesh is composed of 196608 elements.
-Experiments were executed on a single core of a Sandy Bridge architecture.}
+	\subfloat[]{\includegraphics[scale=0.48]{Pictures/elasticity-p1.pdf}}
+	\subfloat[]{\includegraphics[scale=0.48]{Pictures/elasticity-p2.pdf}}
+	\caption{Performance improvement over non-optimized code for the static linear elasticity equation on a single core of the Sandy Bridge architecture.}
 	\label{fig:full-app-res}
@@
 We have demonstrated that our cross-loop optimizations for arithmetic intensity are effective in the context of automated code generation for finite element local assembly.
-In this section, we discuss about their applicability in other computational domains and, more in general, their integrability within a general-purpose compiler.
+In this section, we discuss their applicability in other computational domains and, in general, their integrability within a general-purpose compiler.
 
@@
 PyOP2. This choice was motivated by two critical aspects that
-characterize the generality of our research.
+characterize the generality of our research:
 
@@
 The intuition is that whenever sub-expressions invariant with respect to different sets of affine loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises.
-Pre-computation of invariant terms also increases memory requirements due to the need for temporary arrays, so it is possible that for certain architectures the transformation could actually cause slow downs (e.g.
+Pre-computation of invariant terms also increases memory requirements due to the need for temporary arrays, so it is possible that for certain architectures the transformation could actually cause slowdowns (e.g.
 whenever the available per-core memory is small).
@@
   applicability and the effectiveness of some of our code
-  transformations. For instance, knowing loop bounds allows both generating
+  transformations. For instance, knowing loop bounds allows both generation of
   correct code when applying vector-register tiling (see
-  Section~\ref{sec:op-vect}) and discovering redundant computation
+  Section~\ref{sec:op-vect}) and discovery of redundant computation
   opportunities (see Section~\ref{sec:code-transf-licm}). Padding and data
@@
   general-purpose compiler is far from straightforward, especially if stencils are present, we believe that this could be more easily achieved in specific domains.
-The intuition is to map the memory access pattern on vector registers, and then exploit in-register shuffling to minimize the traffic between memory and processor.
+The intuition is to map the memory access pattern onto vector registers, and then to exploit in-register shuffling to minimize the traffic between memory and processor.
 By demonstrating the effectiveness of multi-loop vectorization in a real scenario, our research represents an incentive for studying this technique in a broader and systematic way.

### 2014-09-10 TACO-paper-v2/TACO-Rebuttal.txt
Update rebuttal (from shared google docs)

@@
 We thank our reviewers (R1-R4). We appreciate the positive feedback and the
-critical comments on our paper. Below we answer the various points arose by the
-reviewers.
+critical comments on our paper. Below we answer the various points raised by
+the reviewers.  We have addressed many minor issues, and we have added two
+substantial new sections - an application case study (§5.3) and an extended
+discussion of the generality of the work (§6).
 
-R1:
-We agree the big missing part was a discussion on the generality of our
+R1: We agree on the need for more specific discussion of the generality of our
 approach and the applicability of the code transformations to other
 computational domains (or, perhaps, their integration within a general-purpose
-compilers). We added therefore a complete new section (Section 6) that explains
-these aspects in details. We believe this addition was necessary to let readers
-understand how to transfer our work to a different domain or compiler (or, at
-least, to let them gain something from our experience of designing/developing 
-the code transformations and COFFEE).
+compilers). We therefore added a complete new section (Section 6) that explains
+these aspects in detail. This should help readers understand how to transfer
+our work to a different domain or compiler.
 
-We also addressed all of the other useful comments (1-8) provided:
-1) The goal is to develop a class of optimizations for an infinite variety of 
-local assembly codes. In that paragraph we say "(...) makes it hard to 
-determine a single or specific sequence of transformations that is successfully 
-applicable to all problems (...)", which implies that different finite element 
-problems may require different code optimizations due to the inherent diversity 
-in mathematical expressions and loop trip counts (this aspect is reiterated 
-throughout the paper). Therefore, a compiler-based approach, in which the input 
-and some parameters are analyzed to decide how to optimally transform the code 
-is needed to maximize the performance. Another possibility would be a 
-library-based approach, although it would be quite pointless since we can 
-already generate the most suitable code for each problem.
-2) The variation in loop trip counts impacts loops transformation. If the loop 
-is relatively large, for example, vector-register tiling is likely to be useful 
-to minimize register spilling; if it is small, then full or partial unrolling 
-is an option. Understanding the border between "small" and "large" is 
-challenging.  The cost model helps in this direction. We tried to clarify the 
-point you make in various sections in the paper: the sentence at page 6 has 
-been re-written to anticipate the fact that "different loop sizes require 
-different transformations". Then we elaborate on the importance and the impact 
-of loop unrolling in Section 4 (there is a complete new paragraph about that).  
+We have also addressed all of the other very comments (1-8) provided: 
+(1) “a compiler is the only reasonable option for obtaining close-to-peak performance
+for these loops, but you don't say what other options were considered nor what
+their downsides are”: 
+The goal is to develop a class of optimizations for an
+infinite variety of local assembly codes. In that paragraph we say "(...) makes
+it hard to determine a single or specific sequence of transformations that is
+successfully applicable to all problems (...)", which implies that different
+finite element problems may require different code optimizations due to the
+inherent diversity in mathematical expressions and loop trip counts (this
+aspect is reiterated throughout the paper). Therefore, a compiler-based
+approach, in which the input and some parameters are analyzed to decide how to
+optimally transform the code is needed to maximize the performance. Another
+possibility would be a library-based approach, although it would be quite
+pointless since we can already generate the most suitable code for each
+problem.
+
+2) On page 6, why do varying iteration counts for your loops make
+optimization challenging?  If your loop trip count is small enough, why not
+just unroll a number of times to remove this loop?” 
+The variation in loop trip counts impacts loops transformation. If the loop is relatively large, for
+example, vector-register tiling is likely to be useful to minimize register
+spilling; if it is small, then full or partial unrolling is an option.
+Understanding the border between "small" and "large" is challenging.  The cost
+model helps in this direction. We tried to address the point you make in
+various sections in the paper: the sentence at page 6 has been re-written to
+anticipate the fact that "different loop sizes require different
+transformations". Then we elaborate on the importance and the impact of loop
+unrolling in Section 4 (there is a complete new paragraph about that).
 Finally, in Section 5.2.5, we demonstrate our claim when discussing the impact
@@
 benchmarks.
-3) Clarified in the paper by adding/re-writing a few sentences. The short 
-answer is: yes.
-4) Explained in the paper.
-5) We agree. We removed the code listing and expanded Figure 4.
-6) Explained in the paper.
-7) This was already explained by the sentence "Other optimization levels 
-performed, in general, slightly worse."
-8) We believe each of those tables can be represented by 6 different charts (1 
-for each <Architecture, Equation> examined>). We have not done that for lack of 
-space (the submission for acceptance is limited to 25 pages) but we can easily 
-do that, if really needed, in the final version of the paper (which can be 
-30-page long)
+
+3) “Do you mean that COFFEE produces this pragma for the C
+iompiler?“: 
+Clarified in the paper by adding/re-writing a few sentences. The short answer is: yes.
+
+4) “a large optimization space, but never actually give an indication of its size”: 
+Explained in the paper.
+
+5) “picture rather than the code listing in figure 5“: 
+We agree. We removed the code listing and expanded Figure 4.
+
+6) “two architectures...differences”: 
+Explained in the paper.
+
+7) “different compilation flags”: 
+This was already explained by the sentence "Other optimization levels performed, in general, slightly worse."
+
+8) “Tables III and IV would definitely be more readable as charts”: 
+We believe each of those tables can be represented by 6 different charts (1 for each
+<Architecture, Equation> examined>). We have not done that for lack of space
+(the submission for acceptance is limited to 25 pages) but we can easily do
+that, if really needed, in the final version of the paper (which can be 30-page
+long)
 
@@
 
-R2:
-Two main critiques are moved to our paper.
+R2: Two main critiques are offered on our paper.
 
-a) We do not think that "the techniques of the paper are mostly interesting to 
-a subset of the people in the intersection of code optimization and finite 
+a) “the techniques of the paper are mostly interesting to a subset of the
+people in the intersection of code optimization and finite elements methods.”:
+
+We do not think that "the techniques of the paper are mostly interesting to a
+subset of the people in the intersection of code optimization and finite
 element methods". However, we do agree that the paper was missing an in-depth
@@
 Section 6, about the generality of our approach and the applicability of the
-code transformations to other computational domains. We hope you can read this 
-section carefully, because it represents novel contribution of the paper and 
-aims to answer exactly the concerns you pointed out.
+code transformations to other computational domains. Your comments have
+prompted us to think pretty hard about this, and we think this adds
+significantly to the paper’s contributions.
 
-b) The results of the paper are *really* applicable to other kernels beyond 
-those analyzed and presented. First of all, we want to emphasize a number of 
-aspects. What is important to clarify is that kernels are not "mere small 
-pieces of code", but rather, in general, quite complex numerical routines 
-evaluating integrals in an element of the discretized equation domain, produced 
-by means of real (used by many scientists around the world) domain-specific 
-language and compiler.  There is an infinite variety of kernels that can be 
-generated in Firedrake; the generated code (e.g. loop sizes and mathematical 
-expression) depends on the partial differential equation being solved. This is 
-explained and re-iterated in the paper. Also, varying the polynomial order of 
-the method or the element shape (e.g. triangles, tetrahedrons, prisms), as we 
-do and explain in the performance evaluation section, changes the generated 
-code by increasing loops and arrays sizes and by making mathematical 
-expressions more articulated (this happens, for example, when moving from 2D to 
-3D domains). So we are already testing many different code variants (three 
-different kernels, and then for each kernel we try multiple polynomial orders 
-and element shapes). This is carefully explained in Section 5.2.1. Premised 
-this, we now go back to your point about "applicability and effectiveness of 
-our transformations in other local assembly kernels". Our answer is twofold:
+b) “it is not clear if the results of the paper are applicable to other kernels
+beyond the ones analysed and presented” and “The paper appears to be a
+collection of optimizations that work for these 3 kernels but is not clear how
+broad these techniques are and if they will be effective for other kernels. It
+would make sense to at least include a number of kernels in the study from real
+applications and use some of them as a black box in the evaluation section.
+Kernels are small pieces of code, so presumably there will be many of them
+available. If on the other hand there is only a limited number of kernels
+available, then why not optimize them directly by hand.”
+
+The results of the paper are *really* applicable to other kernels beyond those
+analyzed and presented. First of all, we want to emphasize a number of aspects.
+What is important to clarify is that kernels are not "mere small pieces of
+code", but rather, in general, quite complex numerical routines evaluating
+integrals in an element of the discretized equation domain, produced by means
+of real (used by many scientists around the world) domain-specific language and
+compiler.  There is an infinite variety of kernels that can be generated in
+Firedrake; the generated code (e.g. loop sizes and mathematical expression)
+depends on the partial differential equation being solved. This is explained
+and re-iterated in the paper. Also, varying the polynomial order of the method
+or the element shape (e.g. triangles, tetrahedrons, prisms), as we do and
+explain in the performance evaluation section, changes the generated code by
+increasing loops and arrays sizes and by making mathematical expressions more
+articulated (this happens, for example, when moving from 2D to 3D domains). So
+we are already testing many different code variants (three different kernels,
+and then for each kernel we try multiple polynomial orders and element shapes).
+This is carefully explained in Section 5.2.1. Premised this, we now go back to
+your point about "applicability and effectiveness of our transformations in
+other local assembly kernels". Our answer is twofold:
 
@@
 proposed code transformations have a domain of applicability that goes far
-beyond that of the three analyzed equations. (...)". Then we re-inforce this 
+beyond that of the three analyzed equations. (...)". Then we reinforce this
 later on in the section and in the paper. Other kernels arising from completely
@@
 equation (Linear Elasticity), which brings the total number of studied
-equations to 4 (Helmholtz, Diffusion, Burgers, Elasticity)
-Adding more kernels to the performance evaluation section would make the paper 
-only more chaotic: we specifically selected these three problems because they 
-allow us to highlight all pros and cons of our code transformations. This is 
-also the reason we comment on the performance of the code transformations (and 
-their interplay) in specific sub-sections (referring to, describing, and 
-carefully analyzing the provided plots), rather than in a single, centralized 
-section.
+equations to 4 (Helmholtz, Diffusion, Burgers, Elasticity). Adding more kernels
+to the performance evaluation section would make the paper only more chaotic:
+we specifically selected these three problems because they allow us to
+highlight all pros and cons of our code transformations. This is also the
+reason we comment on the performance of the code transformations (and their
+interplay) in specific sub-sections (referring to, describing, and carefully
+analyzing the provided plots), rather than in a single, centralized section.
 
-Other concerns:
-- It is actually not true that we do not consider other optimization 
-  strategies: 1) a comparison with using linear algebra libriries (the 
-comparison uses hand-crafted kernels) and 2) a comparison with the FEniCS Form 
-Compiler's built-in optimization system (which is the result of a cited paper) 
-are provided.
-- We agree it is useful to report speed ups and execution times for the whole 
+R2’s recommendations: * “It does not discuss and/or compare to other approaches
+to improving the performance of finite element methods: (a) The authors briefly
+comment on using linear algebra libs but then dismiss this approach as
+requiring more work”
+
+- Section 5.2.7 and Table V present a head-to-head performance comparison with
+the FEniCS Form Compiler's built-in optimization system, which is very widely
+used and, as indicated in the cited article, the state of the art.  Section
+5.2.8 presents our experience with using MKL BLAS.
+
+(b) It seems that some type of hardware support, e.g. graphics card, would be
+more appropriate for floating point intensive applications. How does the
+performance presented compare to a graphics card (or other) approach? Does it
+make sense for someone to go down the path presented in this paper or just buy
+and use one (or more) graphics card(s)?
+
+This is now clarified in the introduction of the paper, paragrah 5. Here, we
+give a more detailed explanation.
+
+At various points in the paper (introduction and related work) we cite studies
+concerning local assembly and GPUs. These works however deeply differ from ours
+in a few ways: (i) that they do not rely on automated code generation, (ii)
+they are limited to only very low-order methods (typically only polynomial
+order 1), (iii) they do not try to optimize for cross-loop arithmetic intensity
+as we do, and (iv) their performance comparisons with CPU implementations are
+based on code which has not benefited from the optimizations in this paper. The
+goal in these works is rather to understand how to effectively parallelize the
+local assembly code (so as to have multiple threads executing a single kernel,
+as opposed to our execution model in which we assume a single thread per
+kernel) in specific equations (not in a possibly infinite range as we do).  A
+critical issue when executing local assembly kernels on GPUs is represented by
+memory requirements. We give here two examples of this problem, although more
+issues could be found and discussed. Firstly, the transfer of the mesh from
+host to device may be a great obstacle: if too big, the mesh will not fit the
+available GPU memory, and distributed execution, which is marginally supported
+in available finite element frameworks, would be required. Secondly,
+optimizations such as generalized loop-invariant code motion increase the
+memory requirements due to the introduction of temporary arrays.  This may
+actually lead to poor per-thread shared memory and registers allocation. These
+aspects suggest that the problem of executing generic, automatically-generated
+local assembly kernels on GPUs require a separate, in-depth study. However, in
+the new Section 6, we provide some intuitions of how we could re-adapt COFFEE's
+transformations for execution on GPUs.  Therefore, we invite once more to go
+through the new Section 6 carefully.
+
+* “The speedup results presented refer to the kernels only. It would be better
+to show the improvement on the execution time of the whole finite element
+calculation, end-to-end”:
+
+We agree it is useful to report speedups and execution times for the whole
 finite element calculation. We have added a new section (Section 5.3) about a
 full-application study, showing exactly these values.
-- In some points in the paper (introduction and related work) we cite studies 
-  concerning local assembly and GPUs. These works however deeply differ from 
-ours for a number of reasons, including the facts that they do not rely on 
-automated code generation, they are limited to only very low-order methods 
-(typically only polynomial order 1), and they do not try to optimize for 
-cross-loop arithmetic intensity as we do. The goal in these works is rather to 
-understand how to effectively parallelize the local assembly code (so as to 
-have multiple threads executing a single kernel, as opposed to our execution 
-model in which we assume a single thread per kernel) in specific equations (not 
-in a possibly infinite range as we do). A critical issue when executing local 
-assembly kernels on GPUs is represented by memory requirements. We give here 
-two examples of this problem, although more issues could be found and 
-discussed. Firstly, the transfer of the mesh from host to device may be a great 
-obstacle: if too big, the mesh will not fit the available GPU memory, and 
-distributed execution, which is marginally supported in available finite 
-element frameworks, would be required. Secondly, optimizations such as 
-generalized loop-invariant code motion increase the memory requirements due to 
-the introduction of temporary arrays.  This may actually lead to poor 
-per-thread shared memory and registers allocation. These aspects suggest that 
-the problem of executing generic, automatically-generated local assembly 
-kernels on GPUs require a separate, in-depth study. However, in the new Section 
-6, we provide some intuitions of how we could re-adapt COFFEE's transformations 
-for execution on GPUs.  Therefore, we invite once more to go through the new 
-Section 6 carefully.
+
+* “Related work should cover the individual compilation techniques discussed
+and how they are adapted in this work”:
+
+A new paragraph has been added to the related work section addressing this
+issue (Section 7, paragraph 3).
 
@@
 
-R3:
-We answer here the 5 points of the review.
-1) For what concerns execution on a GPGPU, we replicate here the same answer to 
-R2.
+R3: We answer here the 5 points of the review.  (1) “The targeted computational
+problems seem ideal for GPGPU computing. So why not target at optimizations of
+the kernel functions for GPGPU architectures rather than Intel CPUs. In fact,
+Sandy Bridge also contains a GPGPU.”
 
-In some points in the paper (introduction and related work) we cite studies 
+We replicate here the same answer to R2.
+
+This is now clarified in the introduction of the paper, paragrah 5. Here, we
+give a more detailed explanation.
+
+At various points in the paper (introduction and related work) we cite studies
 concerning local assembly and GPUs. These works however deeply differ from ours
-for a number of reasons, including the facts that they do not rely on automated 
-code generation, they are limited to only very low-order methods (typically 
-only polynomial order 1), and they do not try to optimize for cross-loop 
-arithmetic intensity as we do. The goal in these works is rather to understand 
-how to effectively parallelize the local assembly code (so as to have multiple 
-threads executing a single kernel, as opposed to our execution model in which 
-we assume a single thread per kernel) in specific equations (not in a possibly 
-infinite range as we do). A critical issue when executing local assembly 
-kernels on GPUs is represented by memory requirements. We give here two 
-examples of this problem, although many more issues could be found. Firstly, the 
-transfer of the mesh from host to device may be a great obstacle: if too big, 
-the mesh will not fit the available GPU memory, and distributed execution, 
-which is marginally supported in available finite element frameworks, would be 
-required.  Secondly, optimizations such as generalized loop-invariant code 
-motion increase the memory requirements due to the introduction of temporary 
-arrays.  This may actually lead to poor per-thread shared memory and registers 
-allocation. These aspects suggest that the problem of executing generic, 
-automatically-generated local assembly kernels on GPUs require a separate, 
-in-depth study. However, in the new Section 6 of the revised paper, we provide 
-some intuitions of how we could re-adapt COFFEE's transformations for execution 
-on GPUs.
+in a few ways: (i) that they do not rely on automated code generation, (ii)
+they are limited to only very low-order methods (typically only polynomial
+order 1), (iii) they do not try to optimize for cross-loop arithmetic intensity
+as we do, and (iv) their performance comparisons with CPU implementations are
+based on code which has not benefited from the optimizations in this paper. The
+goal in these works is rather to understand how to effectively parallelize the
+local assembly code (so as to have multiple threads executing a single kernel,
+as opposed to our execution model in which we assume a single thread per
+kernel) in specific equations (not in a possibly infinite range as we do).  A
+critical issue when executing local assembly kernels on GPUs is represented by
+memory requirements. We give here two examples of this problem, although more
+issues could be found and discussed. Firstly, the transfer of the mesh from
+host to device may be a great obstacle: if too big, the mesh will not fit the
+available GPU memory, and distributed execution, which is marginally supported
+in available finite element frameworks, would be required. Secondly,
+optimizations such as generalized loop-invariant code motion increase the
+memory requirements due to the introduction of temporary arrays.  This may
+actually lead to poor per-thread shared memory and registers allocation. These
+aspects suggest that the problem of executing generic, automatically-generated
+local assembly kernels on GPUs require a separate, in-depth study. However, in
+the new Section 6, we provide some intuitions of how we could re-adapt COFFEE's
+transformations for execution on GPUs.  Therefore, we invite once more to go
+through the new Section 6 carefully.
 
-2) We agree that profile-guided optimization could be used to improve the 
+2) “Using profile guided optimization has been very commonplace, even for some
+commercial compilers. Some of the domain specific knowledge listed in this work
+could be discovered by profile runs. For example, it is rather easy for a
+profile run to tell that the trip counts are small, the kernel working set can
+fit the L1 data cache, whether data alignment and padding are needed. This work
+should discuss what domain knowledge would not be available from profiles and
+thus call for domain specific transformations.”
+
+We agree that profile-guided optimization could be used to improve the
 performance of kernels, although this aspect is orthogonal to most of our code
@@
 
-3) TODO
+3) “Compile time has always been an important issue. A domain specific compiler
+should have clear advantages over general compilers. This issue has not been
+discussed in the paper.”
 
-4) Firedrake is *not* specifically designed for Intel architectures and 
-compilers. The decision of using Intel platforms and tools is twofold: from one 
-hand, we believe that Intel's is one of the most powerful vectorizing and 
-optimizing compilers, so for a research like ours it was important to compare 
-to the best available on the market; from the other hand, their wide 
-availability represents a strong incentive because users will be likely to run 
-their experiments on these machines.
-It is also worth noting that all general-purpose compilers are, to different 
-extents, limited in optimizing for register locality and code motion (for 
-instance, this is true for the Intel and GNU compilers, although we only used 
-the latter for "local" experimentation), so our work is expected to be 
-equally/similarly effective on different yet "standard" CPU architectures.
-From the practical viewpoint, we observe that COFFEE introduces optimizations 
-aimed at improving cross-loop arithmetic intensity by performing source-level 
-transformations, so there are absolutely no technical issues for using our 
-domain-specific compiler on architectures different than Intel. Actually, this 
-fact is a strength of our work, because it allows COFFEE to be almost 
-completely platform-independent.  The only exception regards intrinsics code: 
-COFFEE is currently only capable of outputing AVX intrinsics, so at the moment 
-we cannot apply vector-register tiling (the also called outer-product 
-vectorization technique) to architectures different than x86+AVX and x86+AVX2.  
+We have added a new subsection (5.2.9) about the compilation time. In short,
+compilation time is never a problem for COFFEE, since it usually takes order of
+milliseconds, while the overall assembly process can take order of hours or
+realistic meshes.
+
+4) “This goal of this paper is to understand the relationship between distinct
+code transformations, impact on local assembly kernels, and to what extent
+their composibility is effective in a class of problems and architectures.
+However, it is quite clear that the code transformations discussed in this work
+is tightly coupled with Intel architectures with SSE/AVX SIMD extensions. It is
+not clear whether such transformations are effective on other architectures
+such as PPC AltiVec, MIPS, SPARC, ARM with Neon, and Vector computers such as
+Cray. COFFEE generates C code, so in theory, the performance of the code will
+be very dependent on the target architecture. Is Firedrake designed only for
+Intel architectures? If yes, this has to be clarified in the paper.”
+
+Firedrake is *not* specifically designed for Intel architectures and compilers.
+The decision of using Intel platforms and tools is twofold: from one hand, we
+believe that Intel's is one of the most powerful vectorizing and optimizing
+compilers, so for research like ours it was important to compare to the best
+available on the market; from the other hand, their wide availability
+represents a strong incentive because users will be likely to run their
+experiments on these machines.  It is also worth noting that all
+general-purpose compilers are, to different extents, limited in optimizing for
+register locality and code motion (for instance, this is true for the Intel and
+GNU compilers, although we only used the latter for "local" experimentation),
+so our work is expected to be equally/similarly effective on different yet
+"standard" CPU architectures.  From the practical viewpoint, we observe that
+COFFEE introduces optimizations aimed at improving cross-loop arithmetic
+intensity by performing source-level transformations, so there are absolutely
+no technical issues for using our domain-specific compiler on architectures
+different than Intel. Actually, this fact is a strength of our work, because it
+allows COFFEE to be almost completely platform-independent.  The only exception
+regards intrinsics code: COFFEE is currently only capable of outputting AVX
+intrinsics, so at the moment we cannot currently apply vector-register tiling
+(the so-called outer-product vectorization technique) to other architectures.
 However, COFFEE's internal algorithms use an intermediate representation of
@@
 
-5) We agree that expression split has many issues involved. Our cost model 
-offers a simple alternative for driving expression splitting. Simplicity is the 
-key to effectiveness, portability, and robustness. Probably using a graph 
-partitioning approach as you rightly point out would allow us to obtain better 
-performance gains, but three points are worth to be highlighted:
-- The split factor is a parameter of the transformation. In our experiments we 
-  have tested several split factors, which means testing as many code variants 
-(i.e. as many as the ways an expression can be split based on sum's 
-associativity) and, as shown by the various graphs in the paper, our cost model 
-seems to be capable of properly approximating such a parameter. This is of 
-course encouraging.
-- In the perspective of integrating expression splitting in a general-purpose 
-  compiler, or generalizing this technique to other problems (as elaborated in 
-the new Section 6), we feel that a simple model is preferable to keep the 
-implementation complexity as low as possible.
-- Graph partitioning for expression splitting seems to us a complete new 
-  research problem. On the other hand, this transformation is only one of the 
-many applicable by COFFEE, and we just do not want the focus of the paper being 
-on "how to optimally and rapidly split a mathematical expression". Rather, we 
-prefer to show that, for our problems, there is clearly a problem of register 
-pressure and that by rescheduling the computation in different loops (and 
-sub-expressions) we can actually obtain a performance gain. We have a strategy 
-for expression splitting; we are not claiming it is the best strategy, but at 
-least we demonstrate it is effective and easily predictable. 
-We spent some time thinking about the problem: two things that we may want to 
-try in the future are 1) a model based on graph partitioning and 2) a model 
-based on a set of rewrite rules that exploit arithmetic properties such as 
-commutativity, associativity, and distributivity/factorization to 
-systematically optimize the arithmetic intensity of the expression.
+5) “Expression split has many issues involved: how to limit the number of
+registers used, how to avoid incurring too many redundant operations, how to
+minimize loop overhead, how to minimize the impact from reduced cache and
+register locality. In this work, expression split is handled with a simple
+heuristic. This is insufficient. Ideally, this problem should be modeled as a
+graph partition problem with a cost model to select the best partitioning.
+Perhaps in Firedrake, this is not a big issue that the performance of splitting
+is insensitive. If this is the case, it should be addressed so that this work
+could be qualified for a journal article.”
 
-Other comments:
-a) We elaborate on the use of loop unwinding (or unrolling) in Section 4, in 
-which we have added a new paragraph (named "Loop unrolling"). This paragraph 
+We agree that there are many issues involved. Our cost model offers a simple
+alternative for driving expression splitting. Simplicity is the key to
+effectiveness, portability, and robustness. Probably using a graph partitioning
+approach as you rightly point out would allow us to obtain better performance
+gains, but three points are worth highlighting: - The split factor is a
+parameter of the transformation. In our experiments we have tested several
+split factors, which means testing as many code variants (i.e. as many as the
+ways an expression can be split based on sum's associativity) and, as shown by
+the various graphs in the paper, our cost model seems to be capable of properly
+approximating such a parameter. This is of course encouraging.  - In the
+perspective of integrating expression splitting in a general-purpose compiler,
+or generalizing this technique to other problems (as elaborated in the new
+Section 6), we feel that a simple model is preferable to keep the
+implementation complexity as low as possible.  - Graph partitioning for
+expression splitting seems to us a complete new research problem. On the other
+hand, this transformation is only one of the many applicable by COFFEE, and we
+just do not want the focus of the paper being on "how to optimally and rapidly
+split a mathematical expression". Rather, we prefer to show that, for our
+problems, there is clearly a problem of register pressure and that by
+rescheduling the computation in different loops (and sub-expressions) we can
+actually obtain a performance gain. We have a strategy for expression
+splitting; we are not claiming it is the best strategy, but at least we
+demonstrate it is effective and easily predictable.  We spent some time
+thinking about the problem: two things that we may want to try in the future
+are 1) a model based on graph partitioning and 2) a model based on a set of
+rewrite rules that exploit arithmetic properties such as commutativity,
+associativity, and distributivity/factorization to systematically optimize the
+arithmetic intensity of the expression.
+
+Other comments: a) “Complete loop unwinding: in all the code examples, such as
+Listing 1,2,3 and 4, the trip counts in the kernels are constants (such as 3,
+5, 12, and 4). In ILP compilers, complete loop unwinding is often used to
+increase ILP and enable more effective optimizations. Is it also used in
+COFFEE?”
+
+We elaborate on the use of loop unwinding (or unrolling) in Section 4, in which
+we have added a new paragraph (named "Loop unrolling"). This paragraph
 elaborates on "complete/partial loop unwinding" as well as other general
@@
 
-b) It was copy-paste leftover, now fixed
+b) “On page 14 and 15, in Table III and IV (...) any reason why such
+inconsistency exist?”
 
-c) Right, fixed by making the statement less strong
+It was copy-paste leftover, now fixed
 
-d) This is all clarified and explained in details in the new Section 6. The 
-short answer is that we hope COFFEE can be re-used in frameworks other than 
-Firedrake (e.g. the FEniCS framework that we compare to in the paper).
+c) “Padding, LICM (not as general, though) are supported in vendor compilers.
+Some forms of expression splitting are also supported to control register
+pressure and compile time. The quoted statement (...) is too strong.”
+
+Right, fixed by making the statement less strong
+
+d) “COFFEE is currently implemented as a stage between FFC and PyOP2. Could you
+explain in the paper why COFFEE and PyOP2 is not integrated. Current
+transformations in COFFEE do not have direct control on the final code to be
+executed on the target machine.”
+
+This is all clarified and explained in details in the new Section 6. The short
+answer is that we hope COFFEE can be re-used in frameworks other than Firedrake
+(e.g. the FEniCS framework that we compare to in the paper).
 
@@
 
-R4:
-We addressed all of the observations in your review:
-1) Fixed, thanks
+R4: We addressed all of the observations in your review:
 
-2) The polyhedral model is well-known among compilers people. Basically, it is 
-a mathematical model for representing affine loop nests and determining in a 
-systemtic way all of its possible legal transformations. Cost models are often 
+1) TYPO page 3
+
+Fixed, thanks
+
+2) “What is a "polyhedral compiler"?”
+
+The polyhedral model is well-known among compilers people. Basically, it is a
+mathematical model for representing affine loop nests and determining in a
+systematic way all of its possible legal transformations. Cost models are often
 used by polyhedral compilers to estimate the effectiveness of a certain
@@
 polyhedral compilers; note that one of the co-authors of our paper is also a
-co-author of that paper (i.e. he is an expert of polyhedral compilers).
+co-author of that paper (i.e. he is an expert in polyhedral compilers).
 
-3 and 4) The speed ups only refer to the computation of the local element 
-matrix.  This is now clarified in Section 5.1, and also re-iterated later on in 
-the paper. We used different meshes for the various experiments. In Section 
-5.2.1 we provide a link to a repository in which the code of the experiments as 
-well as the mesh sizes employed can be found. We note that for these 
-experiments the mesh size was set such that the working set of the computation 
-could at least fit the last level cache (L3 on the Sandy Bridge architecture).  
-The goal of these experiments (Section 5.2 in the revised paper) was only to 
-measure speed ups achieved by optimizing local assembly kernels, so there was 
-no point in running over big meshes (i.e. meshes that would not fit the last 
-level cache), which would have only increased the completion time of our test 
-suite (which is already order of hours).
-We however completely agree that reporting execution times and speed ups of the 
+3 and 4) “Which exact timings are used for measuring the reported speedups in
+Section 5? Is it computing the local matrix? Assembling the full matrix? Or
+solving the PDE? The speedups seem to be unreasonably good if they are not for
+computing only the local matrix (if they also include sparse matrix insertion
+and linear solve). Please explain and comment. I would also like to see some
+numbers on how large a mesh was used (number of cells) and execution time in
+seconds (not only speedups).” and “Related to the previous remark: if the
+speedups are for calling the kernel, what is the corresponding (smaller - if
+that is the case) speedup for the full assembly (including sparse matrix
+insertion) and for the full solution process (including sparse matrix insertion
+and linear solve)?”
+
+The speed ups only refer to the computation of the local element matrix.  This
+is now clarified in Section 5.1, and also re-iterated later on in the paper. We
+used different meshes for the various experiments. In Section 5.2.1 we provide
+a link to a repository in which the code of the experiments as well as the mesh
+sizes employed can be found. We note that for these experiments the mesh size
+was set such that the working set of the computation could at least fit the
+last level cache (L3 on the Sandy Bridge architecture).  The goal of these
+experiments (Section 5.2 in the revised paper) was only to measure speed ups
+achieved by optimizing local assembly kernels, so there was no point in running
+over big meshes (i.e. meshes that would not fit the last level cache), which
+would have only increased the completion time of our test suite (which is
+already order of hours).
+
+We however completely agree that reporting execution times and speedups of the
 overall finite element calculation is of fundamental importance. Therefore,

### 2014-09-10 TACO-paper-v2/coffee.tex
It makes no sense to talk about explicit or implicit for steady problems

@@
   two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
- \item A full-application evaluation to demonstrate that the optimization
+ \item A full application evaluation to demonstrate that the optimization
  strategy greatly impacts the whole finite element problem.
@@
 computation from the measurements. Then, in Section~\ref{sec:full-app},
-we provide a full-application investigation. Here, we use another
+we provide a full application investigation. Here, we use another
 equation, bringing the total number of examined problems to four,
@@
 
-\subsection{Full-application Study}
+\subsection{Full application study}
 \label{sec:full-app}
@@
 
-The execution time of a (non-explicit) finite element method is dominated by two factors: assembly and solve.
+The execution time of a steady finite element problem is dominated by two factors: assembly and solve.
 The evaluation of all local element matrices and vectors, and their insertion, respectively, in a ``global'' sparse matrix and a global vector, compose the ``assembly'' phase.
@@
 1.47$\times$ over the non-optimized local assembly code, obtained in the
-case ($f=2, p=2$). However, we reiterate the fact that full-application
+case ($f=2, p=2$). However, we reiterate the fact that full application
 speed ups increase proportionally with the amount of time spent in
@@
 transferable to other computational domains, perhaps other DSLs, and to
-what extent this would be helpful for improving full-application
+what extent this would be helpful for improving full application
 performance. To answer these questions, we first need to go back to the

### 2014-09-10 TACO-paper-v2/coffee.tex
Some minor English fixes

@@
 electromagnetics and structural mechanics, phenomena are modelled by
-partial differential equations (PDEs). Numerical techniques, like the
+partial differential equations (PDEs). Numerical techniques, such as the
 finite volume method and the finite element method, are widely employed
@@
 We aim to maximize our impact on the platforms that are realistically used
-for finite element applications, so we target conventional CPU architectures. 
+for finite element applications, so we target conventional CPU architectures
+rather than GPUs. 
 The key limiting factor to the execution on GPUs is the stringent memory requirements. 
-Only relatively small meshes can fit a GPU memory, and distributed execution is minimally
-supported by available frameworks (if any). There has been some research on
+Only relatively small problems fit in a GPU memory, and support for
+distributed GPU execution in general purpose finite element frameworks is minimal. There has been some research on
 adapting local assembly to GPUs (which we mention in Section~\ref{sec:related-work}),
@@
   to reduction loops, whereas loops over indices $i$ and $j$ are fully
-  parallel.
-Throughout the paper we assumed a kernel to be executed by a single thread, which is likely to be the best strategy for standard multi-core CPUs.
-On the other hand, we note that for certain architectures (e.g.
-GPUs) this could be prohibitive due to memory requirements.
+  parallel. Throughout the paper we assumed that a kernel will be executed
+  by a single thread, which is likely to be the best strategy for standard
+  multi-core CPUs. On the other hand, we note that for certain architectures
+  (for example GPUs) this could be prohibitive due to memory requirements.
 Intra-kernel parallelization is one possible solution: a domain-specific compiler such as COFFEE could map mathematical quantifiers and operators to different parallelization schemes and generate distinct variants of multi-threaded kernel code.
@@
 \item \textbf{Code generation and applicability of the transformations.} All
-  array sizes and loop bounds, like $n1$ and $n2$ in
+  array sizes and loop bounds, for example $n1$ and $n2$ in
   Equation~\ref{eq:assembly-model}, are known at code generation time. This

### 2014-10-20 TACO-paper-v2/coffee.tex
Address minor issues in the second round of revisions

@@
 distributed GPU execution in general purpose finite element frameworks is minimal. There has been some research on
-adapting local assembly to GPUs (which we mention in Section~\ref{sec:related-work}),
+adapting local assembly to GPUs, which we mention in Section~\ref{sec:related-work},
 although it differs from ours in several ways, including: (i) not relying on
 automated code generation from a domain-specific language (explained next),
-(ii) testing only very low order methods, (iii) not optimizing for
+(ii) testing only order 1 methods (while we focus on the range 1-4), (iii) not optimizing for
 cross-loop arithmetic intensity (the goal is rather effective multi-thread
@@
 and 32 512-bit registers in the Xeon Phi), which can impact the
-optimization strategy. The \texttt{icc 13.1} compiler was used. On the
-Sandy Bridge, the compilation flags used were \texttt{-O2} and
-\texttt{-xAVX} for auto-vectorization. On the Xeon Phi, optimization
-level \texttt{-O3} was used. Other optimization levels performed, in
-general, slightly worse.
+optimization strategy.
+The \texttt{icc 13.1} compiler was used.
+For what concerns the compilation flags, we selected the best optimization levels on each platform: \texttt{-O2}, and \texttt{-xAVX} for auto-vectorization, were used on the Sandy Bridge; \texttt{-O3} was adopted on the Xeon Phi.
 
@@
 COFFEE was developed as a separate, self-contained software module, with
-clear input/output interfaces, rather then incorporating it within
+clear input/output interfaces, rather than incorporating it within
 PyOP2. This choice was motivated by two critical aspects that

### 2014-10-20 TACO-paper-v2/coffee.tex
edited files

@@
 suggestions and their contribution to the Firedrake project.
+We thank the ACM TACO reviewers for their suggestions that led
+to a significantly improved version.
 
@@
 and 32 512-bit registers in the Xeon Phi), which can impact the
-optimization strategy.
-The \texttt{icc 13.1} compiler was used.
-For what concerns the compilation flags, we selected the best optimization levels on each platform: \texttt{-O2}, and \texttt{-xAVX} for auto-vectorization, were used on the Sandy Bridge; \texttt{-O3} was adopted on the Xeon Phi.
+optimization strategy. The \texttt{icc 13.1} compiler was used.  We
+selected the best optimization levels for each platform using the
+following compilation flags: \texttt{-xAVX} for auto-vectorization and
+\texttt{-O2} were used on the Sandy Bridge; \texttt{-O3} was adopted
+on the Xeon Phi.
 

### 2014-10-20 TACO-paper-v2/coffee.tex
Reapply Ram's edits
edits

@@
 rather than GPUs. 
-The key limiting factor to the execution on GPUs is the stringent memory requirements. 
-Only relatively small problems fit in a GPU memory, and support for
-distributed GPU execution in general purpose finite element frameworks is minimal. There has been some research on
-adapting local assembly to GPUs, which we mention in Section~\ref{sec:related-work},
-although it differs from ours in several ways, including: (i) not relying on
-automated code generation from a domain-specific language (explained next),
-(ii) testing only order 1 methods (while we focus on the range 1-4), (iii) not optimizing for
-cross-loop arithmetic intensity (the goal is rather effective multi-thread
-parallelization). In addition, our code transformations would drastically impact the 
-GPU parallelization strategy, for example by increasing a thread's working set. 
-For all these reasons, a study on extending our work to GPU architectures is 
-beyond the scope of this paper. In Section~\ref{sec:generality}, however, we provide 
-some intuitions about this research direction.
+The key limiting factor to the execution on GPUs is the stringent
+memory requirements.  Only relatively small problems fit in a GPU
+memory, and support for distributed GPU execution in general purpose
+finite element frameworks is minimal. There has been some research on
+adapting local assembly to GPUs, which we mention in
+Section~\ref{sec:related-work}, although it differs from ours in
+several ways, including: (i) not optimizing for cross-loop arithmetic
+intensity (the goal is rather effective multi-thread parallelization),
+and (ii) not relying on automated code generation from a
+domain-specific language (explained next); moreover, they have done
+testing only on order 1 methods, while we focus on the order range
+1--4. In addition, our code transformations would drastically impact
+the GPU parallelization strategy, for example by increasing a thread's
+working set.  For all these reasons, a study on extending our work to
+GPU architectures is beyond the scope of this paper. In
+Section~\ref{sec:generality}, however, we provide some intuitions
+about this research direction.
 
@@
 increases the number of accesses to $A$ proportionally to the ``split
-factor'', which is the number of sub-expressions produced. Also,
+factor,'' which is the number of sub-expressions produced. Also,
 depending on how the split is executed, it can lead to redundant
@@
 and 32 512-bit registers in the Xeon Phi), which can impact the
-optimization strategy. The \texttt{icc 13.1} compiler was used.  We
-selected the best optimization levels for each platform using the
+optimization strategy. The \texttt{icc 13.1} compiler was used. 
+We selected the best optimization levels for each platform using the
 following compilation flags: \texttt{-xAVX} for auto-vectorization and

### 2014-10-20 TACO-paper-v2/coffee.tex
edits

@@
 and model-driven code transformations in COFFEE\footnote{COFFEE stands
-for COmpiler For FinitE Element local assembly.}, a compiler for
+for COmpiler For Fast Expression Evaluation.}, a compiler for
 optimizing local assembly kernels. COFFEE is integrated with
@@
 evaluation of constants in outer loops (called $F$ in the code) to act
-as scaling factors of arrays. Trip counts grow proportionally to the
+as scaling factors of arrays. Trip counts grow proportionately to the
 order of the method and arrays may be block-sparse. In addition to a
@@
 
-Splitting an expression (henceforth \emph{split}) has, however, several drawbacks. Firstly, it
-increases the number of accesses to $A$ proportionally to the ``split
-factor,'' which is the number of sub-expressions produced. Also,
-depending on how the split is executed, it can lead to redundant
-computation. For example, the number of times the product $det*W3[i]$ is
-performed is proportional to the number of sub-expressions, as shown in
-the code snippet. Further, it increases loop overhead, for example
-through additional branch instructions. Finally, it might affect
-register locality: for instance, the same array could be accessed in
-different sub-expressions, requiring a proportional number of loads be
-performed. This is not the case for the Helmholtz example. Nevertheless,
-as shown in Section~\ref{sec:perf-results}, the performance gain from
-improved register reuse along inner dimensions can still be greater,
-especially if the split factor and the splitting itself use heuristics
-to minimize the aforementioned issues.
-~\\
-
-Some of the transformations presented in this section, as well as other typical compiler optimizations applicable by COFFEE, like loop interchange and unrolling, are parametric; for instance, expression splitting and loop unrolling depend on the split and the unroll factors, respectively.
-Since all transformations are composable, the optimization space can be as large as order of thousands variants for local assembly kernels in which mathematical expressions are moderately complex and the iteration space is not particularly small, so a strategy to prune it is required.
-This problem is treated in the following section.
+Splitting an expression (henceforth \emph{split}) has, however,
+several drawbacks. Firstly, it increases the number of accesses to $A$
+in proportion to the ``split factor,'' which is the number of
+sub-expressions produced. Also, depending on how splitting is done, it
+can lead to redundant computation. For example, the number of times
+the product $det*W3[i]$ is performed is proportional to the number of
+sub-expressions, as shown in the code snippet. Further, it increases
+loop overhead, for example through additional branch
+instructions. Finally, it might affect register locality: for
+instance, the same array could be accessed in different
+sub-expressions, requiring a proportional number of loads be
+performed. This is not the case for the Helmholtz
+example. Nevertheless, as shown in Section~\ref{sec:perf-results}, the
+performance gain from improved register reuse along inner dimensions
+can still be greater, especially if the split factor and the splitting
+itself use heuristics to minimize the aforementioned issues.
 
@@
 
-COFFEE uses heuristics and a simple cost model to select the most suitable
-optimization strategy for a given problem. Auto-tuning might be used, although
-it would significantly increase the  run-time overhead, since the generation of
-ASTs occurs at run-time as soon as problem-specific data are available.
+Some of the transformations presented in the previous section, as well
+as other typical compiler optimizations used in COFFEE, like loop
+interchange and unrolling, are parametric; for instance, expression
+splitting and loop unrolling depend on the split and the unroll
+factors, respectively. Since all transformations are composable, the
+optimization space can be as large as order of thousands variants for
+local assembly kernels in which mathematical expressions are
+moderately complex and the iteration space is not particularly small,
+so a strategy to prune it is required.  For this, COFFEE uses
+heuristics and a simple cost model to select the most suitable
+optimization strategy for a given problem. Auto-tuning can also be
+used, although it would significantly increase the run-time overhead,
+since the generation of ASTs occurs at run-time as soon as
+problem-specific data are available.
 
@@
 and 32 512-bit registers in the Xeon Phi), which can impact the
-optimization strategy. The \texttt{icc 13.1} compiler was used. 
-We selected the best optimization levels for each platform using the
-following compilation flags: \texttt{-xAVX} for auto-vectorization and
-\texttt{-O2} were used on the Sandy Bridge; \texttt{-O3} was adopted
-on the Xeon Phi.
+optimization strategy. The \texttt{icc 13.1} compiler was used.  We
+selected the best optimization levels for each platform; we used
+\texttt{-xAVX} for auto-vectorization and \texttt{-O2} on the Sandy
+Bridge, and \texttt{-O3} on the Xeon Phi.
 
@@
 pronounced on the Xeon Phi, where the vector length is twice as long,
-which leads to proportionally larger scalar remainder loops.
+which leads to proportionately larger scalar remainder loops.
 
@@
 case ($f=2, p=2$). However, we reiterate the fact that full application
-speed ups increase proportionally with the amount of time spent in
+speed ups increase proportionately with the amount of time spent in
 assembly and, therefore, with the complexity of the equation. By
@@
 The data space, in terms of number of arrays and scalars required to evaluate
-the element matrix, grows proportionally with the complexity of the
+the element matrix, grows proportionately with the complexity of the
 finite element problem. COFFEE has been developed taking into account

### 2014-10-20 TACO-paper-v2/coffee.tex
Minor edits, including the splitting stuff

@@
 sub-expressions, requiring a proportional number of loads be
-performed. This is not the case for the Helmholtz
-example. Nevertheless, as shown in Section~\ref{sec:perf-results}, the
-performance gain from improved register reuse along inner dimensions
-can still be greater, especially if the split factor and the splitting
-itself use heuristics to minimize the aforementioned issues.
+performed; this is not the case for the Helmholtz
+example, though. Nevertheless the performance gain from improved register reuse 
+can still be greater if suitable heuristics are used. Our approach 
+consists of traversing the expression tree and recursively splitting it into 
+multiple sub-expressions if the number of innermost-loop independent 
+variables is larger than a certain threshold. This is elaborated in the next section, 
+and validated against empirical search in Section~\ref{sec:perf-results-split}.
+
+% Nevertheless, as shown in Section~\ref{sec:perf-results}, the
+%performance gain from improved register reuse along inner dimensions
+%can still be greater, especially if the split factor and the splitting
+%itself use heuristics to minimize the aforementioned issues.
 
@@
 \subsubsection{Impact of Expression Splitting}
-\label{sec:perf-results-split} Expression splitting relieves the
-register pressure when the element matrix evaluation needs to read from
-a large number of basis function arrays. As detailed in
+\label{sec:perf-results-split} 
+Expression splitting relieves the register pressure when the element matrix 
+evaluation needs to read from a large number of basis function arrays. As detailed in
 Section~\ref{sec:expr-split}, the price to pay for this optimazion is an
@@
 feasibility and constraints on transforming kernels into a sequence of
-calls to external linear algebra libraries. COFFEE supports all of the
+calls to external linear algebra libraries.
+As suggested by the performance evaluation, an analysis of more sophisticated approaches for expression splitting also looks promising.
+COFFEE supports all of the
 problems expressible in Firedrake, and is already integrated with

### 2014-10-20 TACO-paper-v2/TACO-Rebuttal-final.txt
Minor fixes to paper. Add a draft of covering letter.

We thank our reviewers (R1-R4). We appreciate the feedback and the critical
comments on our paper obtained in the revision process, which really led
to a significantly improved version. Below we detail how we answered the minor
points raised by R1 and R4.

R1:
1) "I'm not hung up over tables II and III, but putting them in small graphs
would, I think, still be good for the final version, space permitting."

We agree that it would probably be easier to read plots than tables; however,
we also feel that the best way of representing a table such as Tables II and III
is by means of 6 plots (3 plots per architecture, each plot corresponding to a
specific partial differential equation). Due to lack of space, we prefer to stick
to the tables rather than adding a total of 12 tiny plots, which would really
hinder readability.

2) "Also, your comment about different compilation flags (page 13, line 18) is
still not clear, in my opinion.  It reads like other optimization levels were
worse only for the Xeon Phi. If you could just clarify that you picked the best
optimization levels for each platform, that make your methodology more clear."

By re-thinking about this, we actually agree with you: the sentence was not that
clear. So we rephrased it this way, which hopefully avoid any ambiguity:

<< (...). The icc 13.1 compiler was used. We selected the best optimization levels
for each platform; we used -xAVX for auto-vectorization and -O2 on the Sandy Bridge,
and -O3 on the Xeon Phi. >>

R4:
1) "Ideally, we would like COFFEE to be platform independent. However, in this work,
COFFEE is tightly coupled with the target platform. It generates vector intrinsic and
works closely with the PyOP2 runtime. It would be better if the authors could show
that the approach would also work well on a different platform, such as ARM, SPARC,
PowerPC or MIPS. Nevertheless, this is merely a suggestion, not a requirement,
since we understand how much work is involved in tuning for a different platform."

We emphasize that the only way in which COFFEE is tied with the architecture is by means
of vector intrinsics, which are related to the x86 instruction set, while all of the other
transformations occur at the source code level. Using some abstractions on top of vector
intrinsics would probably fix this; this is something we planned for the near future,
since Firedrake is going to be ported to BLUE GENE architectures. COFFEE is
not tied with the PyOP2 runtime either: it is true that a suitable interface is used in
PyOP2 to interact with COFFEE; however, COFFEE can be used in scenarios other than Firedrake/PyOP2,
as demonstrated by the fact that the code lives in a completely different (public)
repository.

2) "The new title of the paper is now changed to focus on cross-loop optimization.
However, for modern microprocessors, cross-loop optimization usually includes locality
improvement transformations to increase the performance of cache hierarchies. COFFEE
may not need to do that since it focuses on lower order methods where the working set
is already fit in the L1 cache. However, the paper should point out why locality
improvement is not in their scope."

We note that locality improvement really is in the scope of the paper: we optimize for
data locality at the level of registers, which is the level of the memory hierarchy
immediately below the L1 cache. In particular, the outer-product vectorization and
expression splitting aim to optimize for register data locality. However, we agree with
you that it is useful to remark this aspect. Therefore, following your adivse, we have
rewritten a few key sentences. Firstly, in the abstract, we changed the following sentence:

<< The compiler manipulates abstract syntax trees generated from a domain-specific language
by introducing domain-aware optimizations and, eventually, produces C code including vector
SIMD intrinsics. >>

into:

<< The compiler manipulates abstract syntax trees generated from a domain-specific
language by introducing domain-aware optimizations for instruction-level parallelism
and register locality. Eventually, it produces C code including vector SIMD
intrinsics. >>

This should aware the reader from the very beginning that locality at the level of
registers is the target of our optimization strategy. Note that the term "(register)
locality" is also reiterated throughout the paper. 
Secondly, we have modified another key sentence in the Introduction (page 3):

<< (...). With such a kernel structure, we focus on aspects like the minimization of
floating-point operations, register allocation and instruction-level parallelism,
especially in the form of SIMD vectorization.>>

into:

<< (...). With such kernel and working set structures, traditional transformations for
cache locality (e.g. blocking) are not helpful; instead we focus on aspects like the
minimization of floating-point operations, register locality and instruction-level
parallelism, especially vectorization.>>

which we think explicitly targets your concern.

3) "The current trend for GPGPU is to have the GPU sharing the memory with the CPU
(so called SVM – Shared Virtual Memory). This SVM feature has already shown up in
OpenCL 2.0 and Cuda 6.0 standards. With this trend, such claims about the stringent
memory requirements of GPU will no longer to be true."

Even though SVM is going to be part of the new GPU programming models (OpenCL, CUDA),
that does not mean that "the stringent memory requirements will no longer be true". SVM
is just an abstraction that allows programmers to avoid explictly handling memory management;
however, behind the scenes, the continuous transfers between host and device will
still impair the performance. We therefore prefer to be really coutious about this matter.

4) "<<There has been some research on adapting local assembly to GPUs (which we mention
in Section 7), although it differs from ours in several ways, including: (i) not relying
on automated code generation from a domain-specific language (explained next),
(ii) testing only very low order methods, (iii) not optimizing for cross-loop arithmetic
intensity (the goal is rather effective multi-thread parallelization). >>

Your work focuses on low order methods, yet you claim that similar work on GPU is different
from yours since they work on “very low” order methods. What are the differences between “low”
and “very low”? From the description of this paper, it seems that they are very close. Some
of the low order methods have trip count s as low as 3, so how much lower can “very low” go?"

We definitely agree our description was unclear. Therefore, we rewrote that part of
the paper as follows:

<<There has been some research on adapting local assembly to GPUs, which we men-
tion in Section 7, although it differs from ours in several ways, including: (i) not op-
timizing for cross-loop arithmetic intensity (the goal is rather effective multi-thread
parallelization), and (ii) not relying on automated code generation from a domain-specific
language (explained next); moreover, they have done testing only on order 1 methods,
while we focus on the order range 1–4. >>

This way not only we clarify the previous ambiguity (by explicitly stating they only test
order 1 methods), also do we emphasize the importance of points (i) and (ii).

5) "Section 6, page 20, line 41
rather then -->  rather than"

Fixed

6) "The treatment of expression split is still too simple. Although real implementations
favor simplicity, a technical paper does not."

For this aspect, we have added a few sentences to Section 3.4, where expression
splitting is described:

<< (...) Nevertheless the performance gain from improved register reuse can still be greater
if suitable heuristics are used. Our approach consists of traversing the expression tree
and recursively splitting it into multiple sub-expressions if the number of innermost-
loop independent variables is larger than a certain threshold. This is elaborated in the
next section, and validated against empirical search in Section 5.2.6. >>

This does not make the treatment less simplistic, but helps understanding the heuristics
we have used (by also adding a forward pointer to the upcoming section, in which COFFEE's
cost model is detailed) and by stating that this heuristics was "validated against empirical
search". The performance evaluation shows that this is indeed a good heuristics, but as
you correctly point out, there is space for further and systematic improvement. That is why
we also added a sentence to the Conclusion section, in which we mention planned further work:

<< (...). As suggested by the performance evaluation, an analysis of more sophisticated
approaches for expression splitting also looks promising.>>

With the expression splitting transformation, our intention was to show that by generalizing
the well-known loop fission transformation, and by using simple intuitions, register locality
can notably be improved. Probably, a formal study would require a complete new paper, because
the complexity of the problem grows considerably if we consider generic expressions embedded
in generic loop nests. We believe that the ideas provided in this paper represent a starting
point for an in-depth, sufficiently abstract analysis of the topic.

### 2014-10-20 TACO-paper-v2/coffee.tex
Minor fixes to paper. Add a draft of covering letter.

@@
 The compiler manipulates abstract syntax trees generated from a domain-specific
-language by introducing domain-aware optimizations and, eventually,
-produces C code including vector SIMD intrinsics. Experiments using a range of real-world finite-element
+language by introducing domain-aware optimizations for instruction-level parallelism and register locality. Eventually,
+it produces C code including vector SIMD intrinsics. Experiments using a range of real-world finite-element
 problems of increasing complexity show that significant performance improvement
@@
 representing the result of local assembly in an element of the
-discretized domain. With such a kernel structure, we focus on aspects
-like the minimization of floating-point operations, register allocation
-and instruction-level parallelism, especially in the form of SIMD
-vectorization.
+discretized domain.
+With such kernel and working set structures, traditional transformations for cache locality (e.g.
+blocking) are not helpful; instead we focus on aspects like the minimization of floating-point operations, register locality
+and instruction-level parallelism, especially vectorization.
 

### 2014-10-22 TACO-paper-v2/TACO-Rebuttal-final.txt
edits

@@
-We thank our reviewers (R1-R4). We appreciate the feedback and the critical
-comments on our paper obtained in the revision process, which really led
-to a significantly improved version. Below we detail how we answered the minor
-points raised by R1 and R4.
+We thank our reviewers (R1-R4). We appreciate the feedback and the
+critical comments on our paper obtained in the revision process, which
+really led to a significantly improved version. Below we detail how we
+answered the minor points raised by R1 and R4.
 
@@
 
-We agree that it would probably be easier to read plots than tables; however,
-we also feel that the best way of representing a table such as Tables II and III
-is by means of 6 plots (3 plots per architecture, each plot corresponding to a
-specific partial differential equation). Due to lack of space, we prefer to stick
-to the tables rather than adding a total of 12 tiny plots, which would really
-hinder readability.
+  We agree that it would probably be easier to read plots than
+  tables. We will exceed the 25-page limit, if we do so. In addition,
+  we feel that the best way of representing a table such as Tables II
+  and III is by means of 6 plots (3 plots per architecture, each plot
+  corresponding to a specific partial differential equation). Due to
+  lack of space, we prefer to stick to the tables rather than adding a
+  total of 12 tiny plots, which would really hinder readability.
 
-2) "Also, your comment about different compilation flags (page 13, line 18) is
-still not clear, in my opinion.  It reads like other optimization levels were
-worse only for the Xeon Phi. If you could just clarify that you picked the best
-optimization levels for each platform, that make your methodology more clear."
+2) "Also, your comment about different compilation flags (page 13,
+line 18) is still not clear, in my opinion.  It reads like other
+optimization levels were worse only for the Xeon Phi. If you could
+just clarify that you picked the best optimization levels for each
+platform, that make your methodology more clear."
 
-By re-thinking about this, we actually agree with you: the sentence was not that
-clear. So we rephrased it this way, which hopefully avoid any ambiguity:
+  Thanks for the suggestion. The sentence was not that clear. So we
+  rephrased it as follows, which hopefully avoids any ambiguity:
 
-<< (...). The icc 13.1 compiler was used. We selected the best optimization levels
-for each platform; we used -xAVX for auto-vectorization and -O2 on the Sandy Bridge,
-and -O3 on the Xeon Phi. >>
+    "... The icc 13.1 compiler was used. We selected the best
+     optimization levels for each platform; we used -xAVX for
+     auto-vectorization and -O2 on the Sandy Bridge, and -O3 on the
+     Xeon Phi."
 
 R4:
-1) "Ideally, we would like COFFEE to be platform independent. However, in this work,
-COFFEE is tightly coupled with the target platform. It generates vector intrinsic and
-works closely with the PyOP2 runtime. It would be better if the authors could show
-that the approach would also work well on a different platform, such as ARM, SPARC,
-PowerPC or MIPS. Nevertheless, this is merely a suggestion, not a requirement,
-since we understand how much work is involved in tuning for a different platform."
 
-We emphasize that the only way in which COFFEE is tied with the architecture is by means
-of vector intrinsics, which are related to the x86 instruction set, while all of the other
-transformations occur at the source code level. Using some abstractions on top of vector
-intrinsics would probably fix this; this is something we planned for the near future,
-since Firedrake is going to be ported to BLUE GENE architectures. COFFEE is
-not tied with the PyOP2 runtime either: it is true that a suitable interface is used in
-PyOP2 to interact with COFFEE; however, COFFEE can be used in scenarios other than Firedrake/PyOP2,
-as demonstrated by the fact that the code lives in a completely different (public)
-repository.
+1) "Ideally, we would like COFFEE to be platform independent. However,
+in this work, COFFEE is tightly coupled with the target platform. It
+generates vector intrinsic and works closely with the PyOP2
+runtime. It would be better if the authors could show that the
+approach would also work well on a different platform, such as ARM,
+SPARC, PowerPC or MIPS. Nevertheless, this is merely a suggestion, not
+a requirement, since we understand how much work is involved in tuning
+for a different platform."
 
-2) "The new title of the paper is now changed to focus on cross-loop optimization.
-However, for modern microprocessors, cross-loop optimization usually includes locality
-improvement transformations to increase the performance of cache hierarchies. COFFEE
-may not need to do that since it focuses on lower order methods where the working set
-is already fit in the L1 cache. However, the paper should point out why locality
-improvement is not in their scope."
+  We emphasize that the only way in which COFFEE is tied with the
+  architecture is by means of vector intrinsics (in this case, related
+  to the x86 instruction set), while all of the other transformations
+  occur at the source code level. Using some abstractions on top of
+  vector intrinsics would probably fix this; this is something we
+  planned for the near future, since Firedrake is going to be ported
+  to BLUE GENE architectures. COFFEE is not tied with the PyOP2
+  runtime either: it is true that a suitable interface is used in
+  PyOP2 to interact with COFFEE; however, COFFEE easily can be used in
+  scenarios other than Firedrake/PyOP2.
 
-We note that locality improvement really is in the scope of the paper: we optimize for
-data locality at the level of registers, which is the level of the memory hierarchy
-immediately below the L1 cache. In particular, the outer-product vectorization and
-expression splitting aim to optimize for register data locality. However, we agree with
-you that it is useful to remark this aspect. Therefore, following your adivse, we have
-rewritten a few key sentences. Firstly, in the abstract, we changed the following sentence:
+2) "The new title of the paper is now changed to focus on cross-loop
+optimization.  However, for modern microprocessors, cross-loop
+optimization usually includes locality improvement transformations to
+increase the performance of cache hierarchies. COFFEE may not need to
+do that since it focuses on lower order methods where the working set
+is already fit in the L1 cache. However, the paper should point out
+why locality improvement is not in their scope."
 
-<< The compiler manipulates abstract syntax trees generated from a domain-specific language
-by introducing domain-aware optimizations and, eventually, produces C code including vector
+  We note that locality improvement is already in the scope of the
+  paper: we optimize for data locality at the level of registers,
+  which is the level of the memory hierarchy immediately below the L1
+  cache. In particular, the outer-product vectorization and expression
+  splitting aim to optimize for register data locality. However, we
+  agree with you that it is useful to remark this aspect. Therefore,
+  following your suggestion, we have rewritten a few key
+  sentences. Firstly, in the abstract, we changed the following
+  sentence:
+
+    "The compiler manipulates abstract syntax trees generated from a
+     domain-specific language by introducing domain-aware
+     optimizations and, eventually, produces C code including vector
      SIMD intrinsics. >>
@@
 
-<< The compiler manipulates abstract syntax trees generated from a domain-specific
-language by introducing domain-aware optimizations for instruction-level parallelism
-and register locality. Eventually, it produces C code including vector SIMD
-intrinsics. >>
+    "The compiler manipulates abstract syntax trees generated from a
+     domain-specific language by introducing domain-aware
+     optimizations for instruction-level parallelism and register
+     locality. Eventually, it produces C code including vector SIMD
+     intrinsics."
 
-This should aware the reader from the very beginning that locality at the level of
-registers is the target of our optimization strategy. Note that the term "(register)
-locality" is also reiterated throughout the paper. 
-Secondly, we have modified another key sentence in the Introduction (page 3):
+  This would alert the reader from the very beginning that locality at
+  the level of registers is the target of our optimization
+  strategy. Note that the term "(register) locality" is also
+  mentioned throughout the paper.  Secondly, we have modified another
+  key sentence in the Introduction (page 3):
 
-<< (...). With such a kernel structure, we focus on aspects like the minimization of
-floating-point operations, register allocation and instruction-level parallelism,
-especially in the form of SIMD vectorization.>>
+    "... With such a kernel structure, we focus on aspects like the
+     minimization of floating-point operations, register allocation
+     and instruction-level parallelism, especially in the form of SIMD
+     vectorization."
 
-into:
+  into one that we think explicitly targets your concern:
 
-<< (...). With such kernel and working set structures, traditional transformations for
-cache locality (e.g. blocking) are not helpful; instead we focus on aspects like the
-minimization of floating-point operations, register locality and instruction-level
-parallelism, especially vectorization.>>
+    "... With such kernel and working set structures, traditional
+     transformations for cache locality (e.g., blocking) are not
+     helpful; instead we focus on aspects like the minimization of
+     floating-point operations, register locality and
+     instruction-level parallelism, especially vectorization.>>
 
-which we think explicitly targets your concern.
+3) "The current trend for GPGPU is to have the GPU sharing the memory
+with the CPU (so called SVM – Shared Virtual Memory). This SVM feature
+has already shown up in OpenCL 2.0 and Cuda 6.0 standards. With this
+trend, such claims about the stringent memory requirements of GPU will
+no longer to be true."
 
-3) "The current trend for GPGPU is to have the GPU sharing the memory with the CPU
-(so called SVM – Shared Virtual Memory). This SVM feature has already shown up in
-OpenCL 2.0 and Cuda 6.0 standards. With this trend, such claims about the stringent
-memory requirements of GPU will no longer to be true."
+  Even though SVM is going to be part of the new GPU programming
+  models (OpenCL, CUDA), that does not mean that "the stringent memory
+  requirements will no longer be true." SVM is just an abstraction
+  that allows programmers to avoid explictly handling memory
+  management; however, behind the scenes, the continuous transfers
+  between host and device will still impair the performance. We
+  therefore prefer to be really cautious about this matter.
 
-Even though SVM is going to be part of the new GPU programming models (OpenCL, CUDA),
-that does not mean that "the stringent memory requirements will no longer be true". SVM
-is just an abstraction that allows programmers to avoid explictly handling memory management;
-however, behind the scenes, the continuous transfers between host and device will
-still impair the performance. We therefore prefer to be really coutious about this matter.
+4) "<<There has been some research on adapting local assembly to GPUs
+(which we mention in Section 7), although it differs from ours in
+several ways, including: (i) not relying on automated code generation
+from a domain-specific language (explained next), (ii) testing only
+very low order methods, (iii) not optimizing for cross-loop arithmetic
+intensity (the goal is rather effective multi-thread
+parallelization). >>
 
-4) "<<There has been some research on adapting local assembly to GPUs (which we mention
-in Section 7), although it differs from ours in several ways, including: (i) not relying
-on automated code generation from a domain-specific language (explained next),
-(ii) testing only very low order methods, (iii) not optimizing for cross-loop arithmetic
-intensity (the goal is rather effective multi-thread parallelization). >>
+Your work focuses on low order methods, yet you claim that similar
+work on GPU is different from yours since they work on “very low”
+order methods. What are the differences between “low” and “very low”?
+From the description of this paper, it seems that they are very
+close. Some of the low order methods have trip count s as low as 3, so
+how much lower can “very low” go?"
 
-Your work focuses on low order methods, yet you claim that similar work on GPU is different
-from yours since they work on “very low” order methods. What are the differences between “low”
-and “very low”? From the description of this paper, it seems that they are very close. Some
-of the low order methods have trip count s as low as 3, so how much lower can “very low” go?"
+  We definitely agree our description was unclear. Therefore, we
+  rewrote that part of the paper as follows:
 
-We definitely agree our description was unclear. Therefore, we rewrote that part of
-the paper as follows:
+    "There has been some research on adapting local assembly to GPUs,
+     which we mention in Section 7, although it differs from ours in
+     several ways, including: (i) not optimizing for cross-loop
+     arithmetic intensity (the goal is rather effective multi-thread
+     parallelization), and (ii) not relying on automated code
+     generation from a domain-specific language (explained next);
+     moreover, they have done testing only on order 1 methods, while
+     we focus on the order range 1–4."
 
-<<There has been some research on adapting local assembly to GPUs, which we men-
-tion in Section 7, although it differs from ours in several ways, including: (i) not op-
-timizing for cross-loop arithmetic intensity (the goal is rather effective multi-thread
-parallelization), and (ii) not relying on automated code generation from a domain-specific
-language (explained next); moreover, they have done testing only on order 1 methods,
-while we focus on the order range 1–4. >>
-
-This way not only we clarify the previous ambiguity (by explicitly stating they only test
-order 1 methods), also do we emphasize the importance of points (i) and (ii).
+  This way, we not only remove the previous ambiguity (by explicitly
+  stating they only test order 1 methods), but also emphasize the
+  importance of points (i) and (ii).
 
@@
 
-6) "The treatment of expression split is still too simple. Although real implementations
-favor simplicity, a technical paper does not."
+6) "The treatment of expression split is still too simple. Although
+real implementations favor simplicity, a technical paper does not."
 
-For this aspect, we have added a few sentences to Section 3.4, where expression
-splitting is described:
+  For this aspect, we have added a few sentences to Section 3.4, where
+  expression splitting is described:
 
-<< (...) Nevertheless the performance gain from improved register reuse can still be greater
-if suitable heuristics are used. Our approach consists of traversing the expression tree
-and recursively splitting it into multiple sub-expressions if the number of innermost-
-loop independent variables is larger than a certain threshold. This is elaborated in the
-next section, and validated against empirical search in Section 5.2.6. >>
+    "... Nevertheless the performance gain from improved register
+     reuse can still be greater if suitable heuristics are used. Our
+     approach consists of traversing the expression tree and
+     recursively splitting it into multiple sub-expressions as long as
+     the number of variables independent of the innermost loop exceeds
+     a certain threshold. This is elaborated in the next section, and
+     validated against empirical search in Section 5.2.6. >>
 
-This does not make the treatment less simplistic, but helps understanding the heuristics
-we have used (by also adding a forward pointer to the upcoming section, in which COFFEE's
-cost model is detailed) and by stating that this heuristics was "validated against empirical
-search". The performance evaluation shows that this is indeed a good heuristics, but as
-you correctly point out, there is space for further and systematic improvement. That is why
-we also added a sentence to the Conclusion section, in which we mention planned further work:
+  This does not make the treatment less simplistic, but helps
+  understanding the heuristics we have used (by also adding a forward
+  pointer to the upcoming section, in which COFFEE's cost model is
+  detailed) and by stating that this heuristics was "validated against
+  empirical search." Our performance evaluation shows that this is
+  indeed a good heuristic, but as you correctly point out, there is
+  space for further systematic exploration. That is why we also
+  added a sentence to the Conclusion section, in which we mention
+  planned further work:
 
-<< (...). As suggested by the performance evaluation, an analysis of more sophisticated
-approaches for expression splitting also looks promising.>>
+    "... As suggested by the performance evaluation, an exploration of
+     more sophisticated approaches for expression splitting also looks
+     promising."
 
-With the expression splitting transformation, our intention was to show that by generalizing
-the well-known loop fission transformation, and by using simple intuitions, register locality
-can notably be improved. Probably, a formal study would require a complete new paper, because
-the complexity of the problem grows considerably if we consider generic expressions embedded
-in generic loop nests. We believe that the ideas provided in this paper represent a starting
-point for an in-depth, sufficiently abstract analysis of the topic.
+  With the expression splitting transformation, our intention was to
+  show that by generalizing the well-known loop fission
+  transformation, and through simple intuition, register locality can
+  notably be improved. It is clear that a formal study would require
+  another complete paper, because the complexity of the problem grows
+  considerably if we consider generic expressions embedded in generic
+  loop nests. We believe that the ideas provided in this paper
+  represent a starting point for an in-depth analysis of the topic.

### 2014-10-22 TACO-paper-v2/coffee.tex
edits

@@
 small: their trip count rarely exceeds 30, and may be as low as 3 for
-low order methods. In the innermost loop, a problem-specific,
-compute intensive expression evaluates a two dimensional array,
-representing the result of local assembly in an element of the
-discretized domain.
-With such kernel and working set structures, traditional transformations for cache locality (e.g.
-blocking) are not helpful; instead we focus on aspects like the minimization of floating-point operations, register locality
-and instruction-level parallelism, especially vectorization.
+low order methods. In the innermost loop, a problem-specific, compute
+intensive expression evaluates a two dimensional array, representing
+the result of local assembly in an element of the discretized
+domain. With such kernel and working set structures, traditional
+transformations for cache locality (e.g., blocking) are not helpful;
+instead we focus on aspects like the minimization of floating-point
+operations, register locality and instruction-level parallelism,
+especially vectorization.
 
@@
 sub-expressions, requiring a proportional number of loads be
-performed; this is not the case for the Helmholtz
-example, though. Nevertheless the performance gain from improved register reuse 
+performed; this is not the case for the Helmholtz example,
+though. Nevertheless the performance gain from improved register reuse
 can still be greater if suitable heuristics are used. Our approach
-consists of traversing the expression tree and recursively splitting it into 
-multiple sub-expressions if the number of innermost-loop independent 
-variables is larger than a certain threshold. This is elaborated in the next section, 
-and validated against empirical search in Section~\ref{sec:perf-results-split}.
+consists of traversing the expression tree and recursively splitting
+it into multiple sub-expressions as long as the number of variables
+independent of the innermost loop exceeds a certain threshold. This is
+elaborated in the next section, and validated against empirical search
+in Section~\ref{sec:perf-results-split}.
 
@@
 In this paper, we have presented the study and systematic performance
-evaluation of a class of composable cross-loop optimizations for improving
-arithmetic intensity in finite element local assembly kernels, and their
-integration in a novel compiler, COFFEE. In the context of automated code
-generation for finite element local assembly, COFFEE is the first compiler
-capable of introducing low-level optimizations to maximize instruction-level
-parallelism, register locality and SIMD vectorization. Assembly kernels have
-particular characteristics. Their iteration space is usually very small, with
-the size depending on aspects like the degree of accuracy one wants to
-reach (polynomial order of the method) and the mesh discretization employed.
-The data space, in terms of number of arrays and scalars required to evaluate
-the element matrix, grows proportionately with the complexity of the
-finite element problem. COFFEE has been developed taking into account
-all of these degrees of freedom, based on the idea that reducing the
-problem of local assembly optimization to a fixed sequence of
-transformations is far too superficial if close-to-peak performance
-needs to be reached. The various optimizations overcome limitations of
-current vendor and research compilers. The exploitation of domain
-knowledge allows some of them to be particularly effective, as
-demonstrated by our experiments on two state-of-the-art Intel
-platforms. Further work includes a comprehensive study about
-feasibility and constraints on transforming kernels into a sequence of
-calls to external linear algebra libraries.
-As suggested by the performance evaluation, an analysis of more sophisticated approaches for expression splitting also looks promising.
-COFFEE supports all of the
-problems expressible in Firedrake, and is already integrated with
-this framework. The generality and the applicability of the proposed
-code transformations to other domains has also been discussed.
+evaluation of a class of composable cross-loop optimizations for
+improving arithmetic intensity in finite element local assembly
+kernels, and their integration in a novel compiler, COFFEE. In the
+context of automated code generation for finite element local
+assembly, COFFEE is the first compiler capable of introducing
+low-level optimizations to maximize instruction-level parallelism,
+register locality and SIMD vectorization. Assembly kernels have
+particular characteristics. Their iteration space is usually very
+small, with the size depending on aspects like the degree of accuracy
+one wants to reach (polynomial order of the method) and the mesh
+discretization employed.  The data space, in terms of number of arrays
+and scalars required to evaluate the element matrix, grows
+proportionately with the complexity of the finite element
+problem. COFFEE has been developed taking into account all of these
+degrees of freedom, based on the idea that reducing the problem of
+local assembly optimization to a fixed sequence of transformations is
+far too superficial if close-to-peak performance needs to be
+reached. The various optimizations overcome limitations of current
+vendor and research compilers. The exploitation of domain knowledge
+allows some of them to be particularly effective, as demonstrated by
+our experiments on two state-of-the-art Intel platforms. Further work
+includes a comprehensive study about feasibility and constraints on
+transforming kernels into a sequence of calls to external linear
+algebra libraries. As suggested by the performance evaluation, an
+exploration of more sophisticated approaches for expression splitting
+also looks promising. COFFEE supports all of the problems expressible
+in Firedrake, and is already integrated with this framework. The
+generality and the applicability of the proposed code transformations
+to other domains has also been discussed.
 

### 2014-10-23 TACO-paper-v2/coffee.tex
changes suggested by Paul on Oct 22 via email

@@
 operations, which represent a significant fraction of the execution
-time in finite element methods. Their performance optimization is indeed a
-challenging issue. Even though affine loop nests are generally present,
-the short trip counts and the complexity of mathematical expressions, which
-vary among different problems, make it hard to determine a single or unique
-sequence of successful transformations. Our investigation has resulted in the
-implementation of a compiler for local assembly kernels, fully
-integrated with a framework for developing finite element methods.
-The compiler manipulates abstract syntax trees generated from a domain-specific
-language by introducing domain-aware optimizations for instruction-level parallelism and register locality. Eventually,
-it produces C code including vector SIMD intrinsics. Experiments using a range of real-world finite-element
-problems of increasing complexity show that significant performance improvement
-is achieved. The generality of the approach and the applicability of the proposed
-code transformations to other domains is also discussed.
+time in finite element methods. Their performance optimization is
+indeed a challenging issue. Even though affine loop nests are
+generally present, the short trip counts and the complexity of
+mathematical expressions, which vary among different problems, make it
+hard to determine an optimal sequence of successful
+transformations. Our investigation has resulted in the implementation
+of a compiler for local assembly kernels, fully integrated with a
+framework for developing finite element methods.  The compiler
+manipulates abstract syntax trees generated from a domain-specific
+language by introducing domain-aware optimizations for
+instruction-level parallelism and register locality. Eventually, it
+produces C code including vector SIMD intrinsics. Experiments using a
+range of real-world finite-element problems of increasing complexity
+show that significant performance improvement is achieved. The
+generality of the approach and the applicability of the proposed code
+transformations to other domains is also discussed.
 
@@
 
-Achieving high performance on CPUs is non-trivial. The complexity of the
-mathematical expressions, often characterized by a large number of
-operations on constants and small matrices, makes it hard to determine a
-single or specific sequence of transformations that is successfully
-applicable to different problems (equations). Loop trip counts are typically small and can
-vary significantly, which further exacerbates the issue. A
-compiler-based approach is, therefore, the only reasonable option to
-obtain close-to-peak performance in a wide range of different local
-assembly kernels. Optimizations like padding, generalized loop-invariant
-code motion, vector-register tiling, and expression splitting, as well
-as their composition, are essential, but their support in
-state-of-the-art polyhedral and vendor compilers, if any, is very
-limited. BLAS routines could theoretically be employed, although a
-fairly complicated control and data flow analysis would be required to
-automate identification and extraction of matrix-matrix multiplies. In
-addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small
-dimension of the matrices involved and the potential loss in data
-locality can limit or eliminate the performance gain of this approach.
+Achieving high performance on CPUs is non-trivial. The complexity of
+the mathematical expressions, often characterized by a large number of
+operations on constants and small matrices, makes it hard to determine
+an optimal sequence of transformations that is successfully applicable
+to different problems (equations). Loop trip counts are typically
+small and can vary significantly, which further exacerbates the
+issue. A compiler-based approach is, therefore, the only reasonable
+option to obtain close-to-peak performance in a wide range of
+different local assembly kernels. Optimizations like padding,
+generalized loop-invariant code motion, vector-register tiling, and
+expression splitting, as well as their composition, are essential, but
+their support in state-of-the-art polyhedral and vendor compilers, if
+any, is very limited. BLAS routines could theoretically be employed,
+although a fairly complicated control and data flow analysis would be
+required to automate identification and extraction of matrix-matrix
+multiplies. In addition, as detailed in
+Section~\ref{sec:perf-eval-blas}, the small dimension of the matrices
+involved and the potential loss in data locality can limit or
+eliminate the performance gain of this approach.
 
@@
 The assembly code is given in Listing~\ref{code:burgers} and contains 14
-unique arrays that are accessed, where the same array can be referenced
+different arrays that are accessed, where the same array can be referenced
 multiple times within the same expression. This may also require the
@@
 local assembly optimization to a fixed sequence of transformations is
-far too superficial if close-to-peak performance needs to be
+far too simple-minded if close-to-peak performance needs to be
 reached. The various optimizations overcome limitations of current

### 2014-10-29 TACO-paper-v2/coffee.tex
fixes minor

@@
 Foundation through grants 0811457 and 0926687, by the U.S. Army
-through contract W911NF-10-1-000, and by a HiPEAC collaboration
-grant. The authors would like to thank Dr. Carlo Bertolli,
-Dr. Lawrence Mitchell, and Dr. Francis Russell for their invaluable
-suggestions and their contribution to the Firedrake project.
-We thank the ACM TACO reviewers for their suggestions that led
-to a significantly improved version.
+through contract W911NF-10-1-000, by Louisiana State University and by
+a HiPEAC collaboration grant. The authors would like to thank Carlo
+Bertolli, Lawrence Mitchell and Francis Russell for their invaluable
+suggestions and their contribution to the Firedrake project.  We thank
+the ACM TACO reviewers for their suggestions that led to a
+significantly improved version.
 
@@
 Department of Mathematics, Imperial College London; J. Ramanujam,
-iCenter for Computation and Technology and the School of Elec.~Engr.\&
+Center for Computation and Technology and the School of Elec.~Engr.\&
 Comp.~Sci., Louisiana State University.
@@
 
-The time required to apply the numerical kernels is a major issue,
+The time required to execute the numerical kernels is a major issue,
 since the equation domain needs to be discretized into an extremely
@@
 forecasts. However, operational forecast centers have a strict time
-limit in which to produce a forecast - 60 minutes in the case of the
+limit in which to produce a forecast---60 minutes in the case of the
 UK Met Office. Producing efficient kernels has a direct scientific

### 2014-10-29 TACO-paper-v2/coffee.tex
minor edits

@@
 
-We address, in particular, the well-known problem of optimizing the local assembly
-phase of the finite element method \cite{Francis,quadrature1,petsc-integration-gpu,Kirby-FEM-opt},
-which can be responsible for a significant fraction of the overall computation
-run-time, often in the range 30-60$\%$. With respect to these studies,
-we propose a novel set of composable code transformations targeting, for
-the first time, cross-loop arithmetic intensity, with emphasis on instruction-level
-parallelism, redundant computation, and register locality. How such transformations
-generalize to domains other than finite element local assembly will also be discussed.
+In particular, we address the well-known problem of optimizing the
+local assembly phase of the finite element method
+\cite{Francis,quadrature1,petsc-integration-gpu,Kirby-FEM-opt}, which
+can be responsible for a significant fraction of the overall
+computation run-time, often in the range 30-60$\%$. With respect to
+these studies, we propose a novel set of composable code
+transformations targeting, for the first time, cross-loop arithmetic
+intensity, with emphasis on instruction-level parallelism, redundant
+computation and register locality. How such transformations generalize
+to domains other than finite element local assembly will also be
+discussed.
 

### 2014-11-02 TACO-paper-v2/coffee.tex
some edits

@@
 \renewcommand{\ttdefault}{txtt}
+
+\long\def\comment#1{}
 
@@
 \begin{bottomstuff}
-
-This research is partly funded by the MAPDES project, by the
-Department of Computing at Imperial College London, by EPSRC through
-grants EP/I00677X/1, EP/I006761/1, and EP/L000407/1, by NERC grants
-NE/K008951/1 and NE/K006789/1, by the U.S.  National Science
-Foundation through grants 0811457 and 0926687, by the U.S. Army
-through contract W911NF-10-1-000, by Louisiana State University and by
-a HiPEAC collaboration grant. The authors would like to thank Carlo
-Bertolli, Lawrence Mitchell and Francis Russell for their invaluable
-suggestions and their contribution to the Firedrake project.  We thank
-the ACM TACO reviewers for their suggestions that led to a
-significantly improved version.
-
 Author's addresses: Fabio Luporini $\&$ Florian Rathgeber $\&$
@@
 
+\section*{Acknowledgements}
+This research is partly funded by the MAPDES project, by the
+Department of Computing at Imperial College London, by EPSRC through
+grants EP/I00677X/1, EP/I006761/1, and EP/L000407/1, by NERC grants
+NE/K008951/1 and NE/K006789/1, by the U.S.  National Science
+Foundation through grants 0811457, 0926687 and 1059417, by the
+U.S. Army through contract W911NF-10-1-000, by 
+Louisiana State University and by a
+HiPEAC collaboration grant. The authors would like to thank Carlo
+Bertolli, Lawrence Mitchell and Francis Russell for their invaluable
+suggestions and their contribution to the Firedrake project.  We thank
+the ACM TACO and HiPEAC'15 reviewers for their suggestions that led to
+a significantly improved version.
+
 % Bibliography

### 2014-11-02 TACO-paper-v2/coffee.tex
edits

@@
 since the equation domain needs to be discretized into an extremely
-large number of cells to obtain a satisfactory approximation
-of the PDE, possibly of the order of trillions, as in
-\citeN{Rossinelli2013}. For example, it has been well established that mesh
-resolution is critical in the accuracy of numerical weather
+large number of cells to obtain a satisfactory approximation of the
+PDE, possibly of the order of trillions, as in
+\citeN{Rossinelli2013}. For example, it has been well established that
+mesh resolution is critical in the accuracy of numerical weather
 forecasts. However, operational forecast centers have a strict time
@@
 payoff in higher resolution, and therefore more accurate,
-forecasts. Computational cost is a dominant problem in
-computational science simulations, especially for those based on
-finite elements, which are the subject of this paper.
+forecasts. Computational cost is a dominant problem in computational
+science simulations, especially for those based on finite elements,
+that are the subject of this paper.
 
@@
 intensity, with emphasis on instruction-level parallelism, redundant
-computation and register locality. How such transformations generalize
-to domains other than finite element local assembly will also be
-discussed.
+computation and register locality. Also, we discuss 
+how such transformations generalize
+to domains other than finite element local assembly.
 
 During the assembly phase, the solution of the PDE is approximated by
-executing a problem-specific kernel over all cells, or elements, in the
-discretized domain. In this work, we focus on relatively low order finite
-element methods, in which an assembly kernel's working set is usually small
-enough to fit the L1 cache of traditional CPU architectures. Low order methods are by no means exotic: they
-are employed in a wide variety of fields, including climate and ocean
-modeling, computational fluid dynamics, and structural mechanics. The
+executing a problem-specific kernel over all cells, or elements, in
+the discretized domain. In this work, we focus on relatively low order
+finite element methods, in which an assembly kernel's working set is
+usually small enough to fit in the L1 cache of traditional CPU
+architectures. Low order methods are by no means exotic: they are
+employed in a wide variety of fields, including climate and ocean
+modeling, computational fluid dynamics and structural mechanics. The
 efficient assembly of high order methods such as the spectral element
-method \cite{SPENCER} requires a significantly different loop nest structure.
-High order methods are therefore excluded from our study.
-
-We aim to maximize our impact on the platforms that are realistically used
-for finite element applications, so we target conventional CPU architectures
-rather than GPUs. 
-The key limiting factor to the execution on GPUs is the stringent
-memory requirements.  Only relatively small problems fit in a GPU
-memory, and support for distributed GPU execution in general purpose
-finite element frameworks is minimal. There has been some research on
-adapting local assembly to GPUs, which we mention in
-Section~\ref{sec:related-work}, although it differs from ours in
-several ways, including: (i) not optimizing for cross-loop arithmetic
-intensity (the goal is rather effective multi-thread parallelization),
-and (ii) not relying on automated code generation from a
-domain-specific language (explained next); moreover, they have done
-testing only on order 1 methods, while we focus on the order range
-1--4. In addition, our code transformations would drastically impact
-the GPU parallelization strategy, for example by increasing a thread's
-working set.  For all these reasons, a study on extending our work to
-GPU architectures is beyond the scope of this paper. In
-Section~\ref{sec:generality}, however, we provide some intuitions
-about this research direction.
+method \cite{SPENCER} requires a significantly different loop nest
+structure.  High order methods are therefore excluded from our study.
+
+We aim to maximize our impact on the platforms that are realistically
+used for finite element applications, so we target conventional CPU
+architectures rather than GPUs.  The key limiting factor for the
+execution on GPUs is the stringent memory requirements.  Only
+relatively small problems fit in GPU memory, and support for
+distributed GPU execution in general-purpose finite element frameworks
+is minimal. There has been some research on adapting local assembly to
+GPUs, which we mention in Section~\ref{sec:related-work}; these works 
+differ from ours in several ways, including: (i) not optimizing for
+cross-loop arithmetic intensity (the goal is rather effective
+multi-thread parallelization), and (ii) not relying on automated code
+generation from a domain-specific language (explained next); moreover,
+they have done testing only on order 1 methods, while we focus on the
+order range 1--4. In addition, our code transformations would
+drastically impact the GPU parallelization strategy, for example by
+increasing a thread's working set.  For all these reasons, a study on
+extending our work to GPU architectures is beyond the scope of this
+paper. In Section~\ref{sec:generality}, however, we provide some
+intuitions about this research direction.
 
@@
 small: their trip count rarely exceeds 30, and may be as low as 3 for
-low order methods. In the innermost loop, a problem-specific, compute
-intensive expression evaluates a two dimensional array, representing
-the result of local assembly in an element of the discretized
-domain. With such kernel and working set structures, traditional
-transformations for cache locality (e.g., blocking) are not helpful;
-instead we focus on aspects like the minimization of floating-point
-operations, register locality and instruction-level parallelism,
-especially vectorization.
+low order methods. In the innermost loop, a problem-specific
+compute-intensive expression evaluates a two dimensional array,
+representing the result of local assembly in an element of the
+discretized domain. With such kernel and working set structures,
+traditional transformations for cache locality (e.g., blocking) are
+not helpful; instead, we focus on aspects such as the optimization of
+floating-point operations, register locality and instruction-level
+parallelism, especially vectorization.
 
@@
 to different problems (equations). Loop trip counts are typically
-small and can vary significantly, which further exacerbates the
+small and can vary significantly and this further exacerbates the
 issue. A compiler-based approach is, therefore, the only reasonable
@@
 expression splitting, as well as their composition, are essential, but
-their support in state-of-the-art polyhedral and vendor compilers, if
+their support in state-of-the-art vendor and polyhedral compilers, if
 any, is very limited. BLAS routines could theoretically be employed,
@@
 required to automate identification and extraction of matrix-matrix
-multiplies. In addition, as detailed in
+multiplications. In addition, as detailed in
 Section~\ref{sec:perf-eval-blas}, the small dimension of the matrices
@@
 \end{enumerate}
-The paper is organized as follows. In Section~\ref{sec:background} we
+The paper is organized as follows. In Section~\ref{sec:background}, we
 provide some background on local assembly, show code generated by
@@
 Section~\ref{sec:conclusions} reviews our contributions in the light
-of our results, and identifies priorities for future work.
+of our results and identifies priorities for future work.
 
@@
 
-Local assembly is the computation of contributions of a specific cell in
-the discretized domain to the linear system which yields the PDE
+Local assembly is the computation of contributions of a specific cell
+in the discretized domain to the linear system that yields the PDE
 solution. The process consists of numerically evaluating
@@
 \cite{quadrature1,fluidity_manual_v4}, whose sizes depend on the order
-of the method. This operation is applied to all cells in the discretized
-domain. In this work we focus on local matrices, or \emph{element
-matrices}, which are more costly to compute than element vectors.
-
-Given a finite element description of the input problem, expressed
+of the method. This operation is applied to all cells in the
+discretized domain. In this work we focus on local matrices or
+\emph{element matrices}, which are more costly to compute than element
+vectors.
+
+Given a finite element description of the input problem expressed
 through the domain-specific Unified Form Language (UFL)~\cite{ufl},
@@
 finite element method. The evaluation of a local matrix can be
-reduced to integration on a fixed \emph{reference} element --- a special
-element that does not belong to the domain --- after a suitable change
+reduced to integration on a fixed \emph{reference} element---a special
+element that does not belong to the domain---after a suitable change
 of coordinates. Firedrake triggers the compilation of an assembly
-kernel using an available vendor compiler, and manages its (parallel)
+kernel using an available vendor compiler, and manages its parallel
 execution over all elements in the mesh. As already explained, the
@@
 two dimensional array used to store the element matrix, the element's
-coordinates in the discretized domain, and coefficient fields, for
+coordinates in the discretized domain and coefficient fields, for
 instance indicating the values of velocity or pressure in the
 element. The output is the evaluated element matrix. The kernel body
-can be logically split into three parts:
+can be logically split into three parts.
 \begin{enumerate}
@@
     \texttt{static const} in C) of double precision floats.
-  \item Evaluation of the element matrix in an affine loop nest, in which
+  \item Evaluation of the element matrix in an affine loop nest in which
     the integration is performed.
@@
 reduces to a summation of a few products involving basis functions. An
-example is provided in Listing~\ref{code:helmholtz}, which shows the
+example is provided in Listing~\ref{code:helmholtz} that shows the
 assembly kernel for a Helmholtz problem using Lagrange basis functions
@@
 
-Despite the infinite variety of assembly kernels which Firedrake can
+Despite the infinite variety of assembly kernels that Firedrake can
 generate, it is still possible to identify common domain-specific
@@
 all finite element problems that can be formulated in Firedrake.
-Their generality and the potential for applicability in other domains is also discussed in
-Section~\ref{sec:generality}.
-
-As already emphasized, the variations in the structure of
-mathematical expressions and in loop trip counts, although typically
-limited to the order of tens of iterations, render the optimization process
-challenging, requiring distinct sets of transformations to bring performance
-closest to the machine peak in different problems. For example, the Burgers problem in Listing~\ref{code:burgers},
-given the large number of arrays accessed, suffers from high register pressure,
-whereas the Helmholtz problem in Listing~\ref{code:helmholtz} does
-not; this intuitively suggests that the two problems require a
-different treatment, based on an in-depth analysis of both data and
-iteration spaces. Furthermore, domain knowledge enables
-transformations that a general-purpose compiler could not apply,
-making the optimization space even larger. In this context, our goal
-is to understand the relationship between distinct code
-transformations, their impact on cross-loop arithmetic intensity, and to what
-extent their composability is effective in a class of problems and
-architectures.
+Their generality and the potential for applicability in other domains
+is discussed in Section~\ref{sec:generality}.
+
+As already emphasized, the variations in the structure of mathematical
+expressions and in loop trip counts, although typically limited to
+tens of iterations, render the optimization process challenging,
+requiring distinct sets of transformations to bring performance
+closest to the machine peak in different problems. For example, the
+Burgers problem in Listing~\ref{code:burgers}, given the large number
+of arrays accessed, suffers from high register pressure, whereas the
+Helmholtz problem in Listing~\ref{code:helmholtz} does not; this
+intuitively suggests that the two problems require a different
+treatment, based on an in-depth analysis of both data and iteration
+spaces. Furthermore, domain knowledge enables transformations that a
+general-purpose compiler could not apply, making the optimization
+space even larger. In this context, our goal is to understand the
+relationship between distinct code transformations, their impact on
+cross-loop arithmetic intensity, and to what extent their
+composability is effective in a class of problems and architectures.
 
@@
 auto-vectorizable by a vendor compiler. Nevertheless,
-auto-vectorization is not efficient if data are not aligned to cache-line boundaries and if the
-length of the innermost loop is not a multiple of the vector length
-$\mbox{\texttt{VL}}$, especially when the loops are small as in local assembly.
+auto-vectorization is not efficient if data are not aligned to
+cache-line boundaries and if the length of the innermost loop is not a
+multiple of the vector length $\mbox{\texttt{VL}}$, especially when
+the loops are small as in local assembly.
 
@@
 body will be properly aligned. This allows the compiler to issue aligned
-load and store instructions, which are notably faster than unaligned
+load and store instructions, that are notably faster than unaligned
 ones.
@@
 the element matrix also allows to safely round the loop trip count to
-the nearest multiple of $\mbox{\texttt{VL}}$. This avoids the introduction of a
-remainder (scalar) loop from the compiler, which would render
-vectorization less efficient. These extra iterations only write to the
-padded region of the element matrix, and therefore have no side
-effects on the final result.
+the nearest multiple of $\mbox{\texttt{VL}}$. This avoids the
+introduction of a remainder (scalar) loop from the compiler, which
+would render vectorization less efficient. These extra iterations only
+write to the padded region of the element matrix, and therefore have
+no side effects on the final result.
 
 % Autovectorizable LICM -> show code snippet?
-\subsection{Generalized Loop-invariant Code Motion}
+\subsection{Generalized Loop-Invariant Code Motion}
 \label{sec:code-transf-licm}
-From the inspection of the codes in Listings~\ref{code:helmholtz}
-and~\ref{code:burgers}, it can be noticed that the computation of $A$
-involves evaluating many sub-expressions which only depend on two
+From an inspection of the codes in Listings~\ref{code:helmholtz}
+and~\ref{code:burgers}, it can be noted that the computation of $A$
+involves evaluating many sub-expressions that only depend on two
 iteration variables. Since symbols in most of these sub-expressions
-are read-only variables, there is ample space for loop-invariant code
+are read-only variables, there is ample room for loop-invariant code
 motion. Vendor compilers apply this technique, although not in the
@@
   Listing~\ref{code:helmholtz} after application of padding, data
-  alignment, and \emph{licm}, for an AVX architecture. In this
+  alignment, and loop-invariant code motion, for an AVX architecture. In this
   example, sub-expressions invariant to \texttt{j} are identical to
@@
 ($j,k=0,...,3$) of the element matrix using just 2 vector registers,
-which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B''
+that represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B''
 \vert = 1$. Values in a register are shuffled each time a product is
@@
 
-Firedrake users employ the Unified Form Language to express problems in
-a notation resembling mathematical equations. At run-time, the
+Firedrake users employ the Unified Form Language to express problems
+in a notation resembling mathematical equations. At run-time, the
 high-level specification is translated by a modified version of the
@@
 code, is eventually provided to PyOP2~\cite{pyop2ws,pyop2isc}, where
-just-in-time compilation and execution over the discretized domain take
-place. The flow and the compiler structure are outlined in Figure~\ref{fig:coffee-pipeline}.
+just-in-time compilation and execution over the discretized domain
+take place. The flow and the compiler structure are outlined in
+Figure~\ref{fig:coffee-pipeline}.
 
@@
 The compiler applies an ordered sequence of optimization steps to the
-ASTs received from FFC. Application of \emph{licm} must precede padding
+ASTs received from FFC. Application of loop-invariant code motion 
+(or \emph{licm}) must precede padding
 and data alignment, due to the introduction of temporary arrays. These
@@
 
-By comparison with implementations characterized by manually-unrolled loops, we noticed that recent
+By comparison with implementations characterized by manually-unrolled loops, we note that recent
 versions of compilers like GNU's and Intel's estimate close-to-optimal unroll
@@
 extensively used in climate and ocean modeling. Very similar
-operators, for which the same optimisations are expected to be equally
+operators, for which the same optimizations are expected to be equally
 effective, apply to elasticity problems, which are at the base of
@@
 
-\subsubsection{Impact of Generalized Loop-invariant Code Motion}
+\subsubsection{Impact of Generalized Loop-Invariant Code Motion}
 \label{licm-impact}
@@
 Table~\ref{table:perf-results-licm} illustrates the performance
-improvement obtained when \emph{licm} is applied. In general, the
+improvement obtained when loop-invariant code motion (or \emph{licm})
+is applied. In general, the
 speedups are notable. The main reasons were anticipated in
@@
 Table~\ref{table:perf-results-licmap} shows the cumulative impact of
-\emph{licm}, data alignment, and padding over the original code. In
+loop-invariant code motion (or \emph{licm}), data alignment, and padding over the original code. In
 the following, this version of the code is referred to as
@@
 alignment, enhance the quality of auto-vectorization. Occasionally the
-run-time of \emph{licm-ap} is close to that of \emph{licm}, since the
+execution time of \emph{licm-ap} is close to that of \emph{licm}, since the
 non-padded element matrix size is already a multiple of the vector
@@
 of PDEs. Well-known frameworks and applications include
-nek5000~\cite{nek5000-web-page}, the FEniCS project~\cite{Fenics},
+Nek5000~\cite{nek5000-web-page}, the FEniCS project~\cite{Fenics},
 Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake. Numerical
-integration based on quadrature, as in Firedrake, is usually employed to implement the local assembly
-phase. The recent introduction of domain specific languages (DSLs) to
-decouple the finite element specification from its underlying
-implementation facilitated, however, the development of novel
-approaches. Methods based on tensor contraction~\cite{FFC-Compiler}
-and symbolic manipulation~\cite{Francis} have been implemented.
-Nevertheless, it has been demonstrated that quadrature-based integration remains the most efficient choice for a wide class of problems~\cite{quadrature1}, which motivates our work in COFFEE.
-
-Optimization of quadrature-based local assembly for CPU
-architectures has been addressed in FEniCS~\cite{quadrature1}. The
-comparison between COFFEE and this work has been presented in
+integration based on quadrature, as in Firedrake, is usually employed
+to implement the local assembly phase. The recent introduction of
+domain specific languages (DSLs) to decouple the finite element
+specification from its underlying implementation has facilitated the
+development of novel approaches. Methods based on tensor
+contraction~\cite{FFC-Compiler} and symbolic
+manipulation~\cite{Francis} have been implemented. Nevertheless, it
+has been demonstrated that quadrature-based integration remains the
+most efficient choice for a wide class of problems~\cite{quadrature1},
+which motivates our work in COFFEE.
+
+Optimization of quadrature-based local assembly for CPU architectures
+has been addressed in FEniCS~\cite{quadrature1}. The comparison
+between COFFEE and this work has been presented in
 Section~\ref{sec:perf-results-fenics}. In~\citeN{Markall20101815}, and
-more recently in~\citeN{petsc-integration-gpu}, the same problem has been
-studied for GPU architectures. In~\citeN{assembly-opencl}, variants of
-the standard numerical integration algorithm have been specialized and
-evaluated for the PowerXCell processor, but an exhaustive study from
-the compiler viewpoint - like ours - is missing, and none of the
-optimizations presented in Section~\ref{sec:code-transf} are
-mentioned. Among these efforts, to the best of our knowledge, COFFEE
-is the first work targeting low-level optimizations through a real
-compiler approach.
-
-The code transformations presented are inspired by standard compilers optimizations and exploit domain properties.
-Our loop-invariant code motion technique individuates invariant sub-expressions and redundant computation by analyzing all loops in an iteration space, which is a generalization of the algorithms often implemented by general-purpose compilers.
-Expression splitting is an abstract variant of loop fission based on properties of arithmetic operators.
-The outer-product vectorization is an implementation of tiling at the level of vector registers; tiling, or ``loop blocking'', is commonly used to improve data locality (especially for caches).
-Padding has been used to achieve data alignment and to improve the effectiveness of vectorization.
-A standard reference for the compilation techniques re-adapted in this work is~\citeN{dragonbook}.
+more recently in~\citeN{petsc-integration-gpu}, the same problem has
+been studied for GPU architectures. In~\citeN{assembly-opencl},
+variants of the standard numerical integration algorithm have been
+specialized and evaluated for the PowerXCell processor, but an
+exhaustive study from the compiler viewpoint---like ours---is missing,
+and none of the optimizations presented in
+Section~\ref{sec:code-transf} are mentioned. Among these efforts, to
+the best of our knowledge, COFFEE is the first work targeting
+low-level optimizations through a real compiler approach.
+
+The code transformations presented are inspired by standard compilers
+optimizations and exploit domain properties. Our loop-invariant code
+motion technique individuates invariant sub-expressions and redundant
+computation by analyzing all loops in an iteration space, which is a
+generalization of the algorithms often implemented by general-purpose
+compilers. Expression splitting is an abstract variant of loop fission
+based on properties of arithmetic operators. The outer-product
+vectorization is an implementation of tiling at the level of vector
+registers; tiling or ``loop blocking,'' is commonly used to improve
+data locality especially for caches. Padding has been used to
+achieve data alignment and to improve the effectiveness of
+vectorization. A standard reference for the compilation techniques
+re-adapted in this work is~\citeN{dragonbook}.
 
@@
 Spiral~\cite{Pueschel:05} for digital signal processing numerical
-algorithms, ~\citeN{Spampinato:14} for dense linear algebra, or
+algorithms, LGen~\citeN{Spampinato:14} for dense linear algebra, and
 Pochoir~\cite{pochoir} and SDSL~\cite{stencil-compiler} for image
@@
 a given problem on a certain platform has been adopted by
-nek5000~\cite{nek5000} for small matrix-matrix multiplies, the ATLAS
+Nek5000~\cite{nek5000} for small matrix-matrix multiplies, the ATLAS
 library~\cite{ATLAS}, and FFTW~\cite{FFTW} for fast fourier
 transforms. In both cases, pruning the implementation space is
-fundamental to mitigate complexity and overhead. Likewise, COFFEE uses
-a cost model and heuristics (Section~\ref{sec:pyop2-compiler}) to
-steer the optimization process.
-
-%(i.e., given a mesh, those that lead to ``less accurate'' solutions than when using high-order polynomials, at a lower computational cost)
-
-%Show other work of people working on optimizations for computational science kernels.
-Inter-kernel vectorisation paper from Istvan.
-Previous work in FFC.
-Spencer et-al + Shin et.
-al.
-attempt to optimize computations that could benefit from using BLAS, but in practice they don't, due to the very small dgemm operations employed.
-Saday's model-driven SIMD code vectorisation for the tensor contraction engine.
+fundamental to reduce complexity and overhead. Likewise, COFFEE uses a
+cost model and heuristics (Section~\ref{sec:pyop2-compiler}) to steer
+the optimization process.
+
+%(i.e., given a mesh, those that lead to ``less accurate'' solutions
+%than when using high-order polynomials, at a lower computational
+%cost)
+
+%Show other work of people working on optimizations for computational
+%science kernels. Inter-kernel vectorization paper from
+%Istvan. Previous work in FFC. Spencer et-al + Shin et. al. attempt to
+%optimize computations that could benefit from using BLAS, but in
+%practice they don't, due to the very small dgemm operations
+%employed. Saday's model-driven SIMD code vectorization for the tensor
+%contraction engine.
 
@@
 exploration of more sophisticated approaches for expression splitting
-also looks promising. COFFEE supports all of the problems expressible
-in Firedrake, and is already integrated with this framework. The
-generality and the applicability of the proposed code transformations
-to other domains has also been discussed.
-
-\section*{Acknowledgements}
+also looks promising. COFFEE is already integrated with Firedrake and
+supports all of the problems expressible in Firedrake. In addition, we
+have discussed The generality and the applicability of the proposed
+code transformations to other domains.
+
+\section*{Acknowledgments}
 This research is partly funded by the MAPDES project, by the

### 2014-11-03 TACO-paper-v2/coffee.tex
edits back to 25 pages

@@
 \begin{bottomstuff}
+This research is partly funded by the MAPDES project, by the
+Department of Computing at Imperial College London, by EPSRC through
+grants EP/I00677X/1, EP/I006761/1, and EP/L000407/1, by NERC grants
+NE/K008951/1 and NE/K006789/1, by the U.S.  National Science
+Foundation through grants 0811457, 0926687 and 1059417, by the
+U.S. Army through contract W911NF-10-1-000, by 
+Louisiana State University and by a
+HiPEAC collaboration grant. The authors would like to thank Carlo
+Bertolli, Lawrence Mitchell and Francis Russell for their invaluable
+suggestions and their contribution to the Firedrake project.  We thank
+the ACM TACO and HiPEAC'15 reviewers for their suggestions that led to
+a significantly improved version.
+
 Author's addresses: Fabio Luporini $\&$ Florian Rathgeber $\&$
@@
 
+\comment{
 \section*{Acknowledgments}
@@
 a significantly improved version.
+}
 
